### 2020-11-03 jasa-latex-template-v2019/JASAtemplate.tex
adding JASA template as baseline

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%  JASA LaTeX Template File
%  To make articles using JASA.cls, Version 1.1
%  September 14, 2019
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Step 1:
%% Uncomment the style that you want to use:

%%%%%%% For Preprint
%% For manuscript, 12pt, one column style

%% Comment this out if you'd rather use another style:
\documentclass[preprint]{JASA}

%%%%% Preprint Options %%%%%
%% The track changes option allows you to mark changes
%% and will produce a list of changes, their line number
%% and page number at the end of the article.
%\documentclass[preprint,trackchanges]{JASA}

%% NumberedRefs is used for numbered bibliography and citations.
%% Default is Author-Year style.
%% \documentclass[preprint,NumberedRefs]{JASA}

%%%%%%% For Reprint
%% For appearance of finished article; 2 columns, 10 pt fonts

% \documentclass[reprint]{JASA}

%%%%% Reprint Options %%%%%

%% For testing to see if author has exceeded page length request, use 12pt option
%\documentclass[reprint,12pt]{JASA}

%% NumberedRefs is used for numbered bibliography and citations.
%% Default is Author-Year style.
% \documentclass[reprint,NumberedRefs]{JASA}

%% TurnOnLineNumbers
%% Make lines be numbered in reprint style:
% \documentclass[reprint,TurnOnLineNumbers]{JASA}

\begin{document}
%% the square bracket argument will send term to running head in
%% preprint, or running foot in reprint style.

\title[]{}

% ie
%\title[JASA/Sample JASA Article]{Sample JASA Article}

%% repeat as needed
\author{}

% ie
%\author{Author One}
%\author{Author Two}
%\author{Author Three}

\affiliation{}

% ie
%\affiliation{Department1,  University1, City, State ZipCode, Country}

\altaffiliation{}

% may be added after \author{}, ie
% \altaffiliation{Also at: Department1,  University1, City, State ZipCode, Country.}

%% for corresponding author
\email{}

%% For preprint only,
%  optional, if you want want this message to appear in upper right corner of title page
% \preprint{}

%ie
%\preprint{Author, JASA}

% optional, if desired:
%\date{\today}

\begin{abstract}
% Put your abstract here. Abstracts are limited to 200 words for
% regular articles and 100 words for Letters to the Editor. Please no
% personal pronouns, also please do not use the words ``new'' and/or
% ``novel'' in the abstract. An article usually includes an abstract, a
% concise summary of the work covered at length in the main body of the
% article.     
\end{abstract}

%% pacs numbers not used

\maketitle

%  End of title page for Preprint option --------------------------------- %

%% See preprint.tex/.pdf or reprint.tex/.pdf for many examples

%% before appendix (optional) and bibliography:
\begin{acknowledgments}
%This research was supported by  ...
\end{acknowledgments}

% -------------------------------------------------------------------------------------------------------------------
%   Appendix  (optional)

%\appendix
%\section{Appendix title}

%If only one appendix, please use
%\appendix*
%\section{Appendix title}

%=======================================================

%Use \bibliography{<name of your .bib file>}+
%to make your bibliography with BibTeX.

%=======================================================

\end{document}

=================
MORE INFORMATION
=================

Note: The only figure formats allowed are the following: 
.pdf, .ps, .eps, or .jpg. Figure files must be named in this fashion:
Figure\#.xxx, where `\#' is the figure number and `xxx' is the file format
(Figure1.eps, Figure2.jpg, Figure3a.ps, Figure3b.ps, etc).

Fraktur and Blackboard (or open face or double struck) characters should be
typeset using the \verb+\mathfrak{#1}+ and \verb+\mathbb{#1}+ commands
respectively. Both are supplied by the \texttt{amssymb} package which
is called in JASA, so no need to type in \usepackage{amssymb}. For an
example of amssymb in use: \verb+$\mathbb{R}$+ gives $\mathbb{R}$ and
\verb+$\mathfrak{G}$+ gives $\mathfrak{G}$.

\section{Floats, Figures and Tables}

Figures and tables are typically ``floats'' which means that their
final position is determined by \LaTeX\ while the document is being
typeset. \LaTeX\ isn't always successful in placing floats
optimally.  Use the figure* environment to get a wide figure that spans the 
page in a two-column layout.

For help in page makeup in the reprint version, you can use 
\onecolumngrid ...\twocolumngrid to end a page early so that you
can put in something at the bottom of the page. See reprintsample.tex
for an example.

Examine the \LaTeX\ source and output for Table~\ref{tab:table1}.

\begin{ruledtabular}...\end{ruledtabular} is used to spread the table
out to the width of the column or page, and to add two lines at the
top and bottom of the table.

This is generally helpful, but multicolumn doesn't work well in
the ruledtabular environment.

If you need to use the \multicolumn command you should not use
ruledtabular, but instead type in \hline\hline... table...\hline\hline
so that there are still two lines at the top and bottom of the table.

([ht]  means make it here and not float if it will fit on the page;
 otherwise put it on the top of the next page.)

\begin{table}[ht]
\caption{\label{tab:table1}A table with more columns still fits
properly in a column. Note that several entries share the same
footnote. Inspect the \LaTeX\ input for this table to see
exactly how it is done.}

\begin{ruledtabular}
\begin{tabular}{cccccccc}
 &$r_c$ (\AA)&$r_0$ (\AA)&$\kappa r_0$&
 &$r_c$ (\AA) &$r_0$ (\AA)&$\kappa r_0$\\
\hline
Cu& 0.800 & 14.10 & 2.550 &Sn\footnotemark[1]
& 0.680 & 1.870 & 3.700 \\
Ag& 0.990 & 15.90 & 2.710 &Pb\footnotemark[2]
& 0.450 & 1.930 & 3.760 \\
Au& 1.150 & 15.90 & 2.710 &Ca\footnotemark[3]
& 0.750 & 2.170 & 3.560 \\
Mg& 0.490 & 17.60 & 3.200 &Sr\footnotemark[4]
& 0.900 & 2.370 & 3.720 \\
Zn& 0.300 & 15.20 & 2.970 &Li\footnotemark[2]
& 0.380 & 1.730 & 2.830 \\
Cd& 0.530 & 17.10 & 3.160 &Na\footnotemark[5]
& 0.760 & 2.110 & 3.120 \\
Hg& 0.550 & 17.80 & 3.220 &K\footnotemark[5]
&  1.120 & 2.620 & 3.480 \\
Al& 0.230 & 15.80 & 3.240 &Rb\footnotemark[3]
& 1.330 & 2.800 & 3.590 \\
Ga& 0.310 & 16.70 & 3.330 &Cs\footnotemark[4]
& 1.420 & 3.030 & 3.740 \\
In& 0.460 & 18.40 & 3.500 &Ba\footnotemark[5]
& 0.960 & 2.460 & 3.780 \\
Tl& 0.480 & 18.90 & 3.550 & & & & \\
\end{tabular}
\end{ruledtabular}
\footnotetext[1]{Here's the first.}
\footnotetext[2]{Here's the second.}
\footnotetext[3]{Here's the third.}
\footnotetext[4]{Here's the fourth.}
\footnotetext[5]{And etc.}
\end{table}

DCOLUMN commands

``If we do not want to break the fractional and the integral part in two columns,
the dcolumn package provides a new type of column
D{sep -in}{sep -out}{ before.after}
The first argument {sep-in} is the symbol used in the
.tex document to separate
the integral and the fractional part (usually the decimal point . or the decimal
comma ,), the second argument {sep-out}
is the symbol that we want in the
output, the third is the number of digits on the left (before) and on the right
(after) this symbol. The numbers are aligned to the decimal point and, in case
that the third argument is negative, the decimal point is aligned to the center of
the column. If the columns have a heading, it must be inserted into
the command \multicolumn{1}{c}{...}''

An example using dcolumn:
\begin{verbatim}
{\hsize= 2in
\begin{ruledtabular}
\begin{tabular}{cD {,}{.}{5.4}}
Expression           & \multicolumn {1}{c}{ Value }\\
\hline
$\pi$                  &      3,1416                 \\
$\pi^{\pi}$           &     36,46                    \\
$\pi^{\pi^{\pi}}$    & 80662,7                      \\
\end{tabular}
\end{ruledtabular}
}
\end{verbatim}
\vskip12pt
{\hsize= 2in
\begin{ruledtabular}
\begin{tabular}{cD {,}{.}{5.4}}
Expression           & \multicolumn {1}{c}{ Value }\\
\hline
$\pi$                  &      3,1416                 \\
$\pi^{\pi}$           &     36,46                    \\
$\pi^{\pi^{\pi}}$    & 80662,7                      \\
\end{tabular}
\end{ruledtabular}
}

FIGURES

>> See preprint.tex/.pdf or reprint.tex/.pdf for many examples of using
the \figline{} and fig{} commands, new for this style <<

\figline{} will center one or more figures on one line.

\fig{<name of file>}{<width>}{<letter to put underneath>}

\leftfig{<name of file>}{<width>}{<letter to put underneath>}

\rightfig{<name of file>}{<width>}{<letter to put underneath>}

\boxedfig{<name of file>}{<width>}{<letter to put underneath>}

\rotatefig{<degrees of rotation>}{<name of file>}{<width>}+\\
\verb+{<letter to put underneath>}+

Example of Figline with Narrow Caption:

\figline{
\fig{figsamp.jpg}{.7\textwidth}{}
\narrowcaption{.2\textwidth}{Here is a narrow caption.}
}

All the fig type commands used in \figline{} will produce
their letter for use in label and crossreferencing, if
the \label is used after the \fig, \fig{}{}{}\label{stuff}

Figcolumn for stacking figures:
\figcolumn{
\fig{figsamp.jpg}{.2\textwidth}{(A)}
\fig{figsamp.jpg}{.2\textwidth}{(B)}
\fig{figsamp.jpg}{.2\textwidth}{(C)}
}

MULTIMEDIA

Please note that this is for multimedia intended to appear inline
within the published article.

Here is what a multimedia entry will look like:
\multimedia{http://dx.doi.org/10.1121/1.4947423.1}{Corresponding pulse-compressed echo envelopes
and video recordings from a fluttering luna moth.
Echoes from the wings and body of the moth generally dominate the
acoustic returns, which vary greatly over consecutive ensonifications
across the wingbeat cycle. File of type ``mp4'' (15.3
MB)}\label{mmtest1}

Here we try cross referencing the multimedia entry: The multimedia
above is Mm.~\ref{mmtest1}.

SUPPLEMENTARY MATERIAL FOR PUBLICATION

ASA prefers that authors to submit related/relevant article files as
supplementary material with their submission.

An example of reference to supplementary material:

The sound files and videos for this and other figures
are included as supplementary materials\footnote{See
Supplementary materials at [URL will be inserted by AIP]
for [give a brief description of the material].}.

The contents of the footnote above will appear at the beginning of the
bibliography when the `author-year' documentclass option is used;
interleaved with other references otherwise.

FILE NAMING CONVENTIONS

Supplementary Figure or
	Supplementary Figure or Text files should be named: SuppPub\#.xxx, where ``\#'' is
	a number and ``xxx'' is the file format extension
	(SuppPub1.docx, SuppPub2.jpg, etc)

	Supplementary Multimedia files: SuppPubmm\#.xxx, where ``\#'' is a
	number and ``xxx'' is the file format extension (SuppPubmm1.mp3,
	SuppPubmm2.gif, etc)

Multimedia files must be named accordingly: MM\#.xxx, where ``\#'' is the
number and ``xxx'' is the file format extension (MM1.wav, MM2.avi, etc).

The only figure formats allowed are the following: 
.pdf, .ps, .eps, or .jpg. Figure files must be named in this fashion:
Figure\#.xxx, where ``\#'' is the figure number and ``xxx'' is the file format
(Figure1.eps, Figure2.jpg, Figure3a.ps, Figure3b.ps, etc).

TRACK CHANGES
See TrackChangeSample.tex/.pdf for examples in use.

Track changes does not work when using `reprint' option.

 Track Changes:
 To add words, \added{<word added>}
 To delete words, \deleted{<word deleted>}
 To replace words, \replace{<word to be replaced>}{<replacement word>}
 To explain why change was made: \explain{<explanation>}

 At the end of the document, use \listofchanges, which will list the
 changes and the page and line number where the change was made.

 When final version, \listofchanges will not produce anything,
 \added{} word will be printed, \deleted{} will take away the word,
 \replaced{}{} will print only the 2nd argument.
 \explain will not print anything.

FIGURES AND TABLES in APPENDIX
Figure and table numbering are continuous through the article,
and handled the same as they are in the rest of the article.

BIBLIOGRAPHY

Resources for making your bibliography entries
correctly included in this package:

JASA-ReferenceStyles.pdf

bibsamp1.tex/.pdf and bibsamp2.tex/.pdf
for examples of output;

sampbib.bib for an example of
how to make your .bib database entries in your own .bib file.

See examples in preprintsample.tex/.pdf and reprintsample.tex/.pdf

CITATIONS
Should be \citep{} unless you have used documentclass 
option `NumberedRefs' , in which case \cite{} should be used.

BIBTEX
You are highly recommended to use BibTeX to produce your bibliography:
it will be both easier and less error prone.

There are two possible bibliography styles: the default, author-year,
and the optional style, NumberedRefs, which you would call using\\

\documentclass[preprint,NumberedRefs]{JASA}

\documentclass[reprint,NumberedRefs]{JASA}

=======================================================
IMPORTANT

Use \bibliography{<name of your .bib file>}
to make your bibliography with BibTeX.

=======================================================

### 2020-11-03 jasa-latex-template-v2019/TrackChangesSample.tex
adding JASA template as baseline

%	JASA LaTeX Sample File, Preprint Sample
%
%  Beginner Latex users should refer to their favorite online documentation
%  here is one from the TeX Users Group 
%	https://www.tug.org/twg/mactex/tutorials/ltxprimer-1.0.pdf
%
%  Useful FAQ from  https://journals.aps.org/revtex/revtex-faq
%

%%%%%%% For Preprint
%% For manuscript, 12pt, one column style

%\documentclass[preprint]{JASA}

%%%%% Preprint Options %%%%%
%% The track changes option allows you to mark changes
%% and will produce a list of changes, their line number
%% and page number at the end of the article.

%% >>>>>>>>>>>>>>>>>> Use this option for Track Changes:
\documentclass[preprint,trackchanges]{JASA}

%% NumberedRefs is used for numbered bibliography and citations.
%% Default is Author-Year style.
%% \documentclass[preprint,NumberedRefs]{JASA}

%%%%%%% For Reprint
%% For appearance of finished article; 2 columns, 10 pt fonts

% \documentclass[reprint]{JASA}

%%%%% Reprint Options %%%%%

%% For testing to see if author has exceeded page length request, use 12pt option
%\documentclass[reprint,12pt]{JASA}

%% NumberedRefs is used for numbered bibliography and citations.
%% Default is Author-Year style.
% \documentclass[reprint,NumberedRefs]{JASA}

\begin{document}
\title[JASA/Sample JASA Article]{Sample JASA Article}
\author{Author One}
\author{Author Two}
\author{Author Three}
\affiliation{Department1,  University1, City, State ZipCode, Country}

\author{Author Four}
\email{author.four@university.edu}
\affiliation{Department2,  University2, City, State ZipCode, Country}

\author{Author Five}			
\altaffiliation{Also at: Another University, City, State ZipCode, Country.}
\affiliation{Department3,  University3, City, State ZipCode, Country}

\preprint{Author, JASA}		%  if you want want this message to appear in upper right corner of title page

\date{\today}

\begin{abstract}
Put your abstract here. Abstracts are limited to 200 words for
regular articles and 100 words for Letters to the Editor. Please no
personal pronouns, also please do not use the words ``new'' and/or
``novel'' in the abstract. An article usually includes an abstract, a
concise summary of the work covered at length in the main body of the
article.     
\end{abstract}

%% pacs numbers not used

\maketitle

%  End of title page for Preprint option --------------------------------- %

\section{\label{sec:1} Introduction}
This sample document demonstrates the use of JASA in manuscripts 
prepared for submission to the Journal of the Acoustical Society of America.

See JASA-TeXGuide.pdf, which is part of this package, for extensive
documentation on using commands for JASA.

You can compare the .tex version of this file with the resulting .pdf
version to give you an idea of what  commands are available and how
they work. At the top of the .tex file you'll find a listing of the
documentclass options, and an explanation of their results.
Some additional suggestions are included in the body of this
manuscript.

  Beginner Latex users should refer to their favorite online documentation. A 
  useful place to start is the primer from the TeX Users Group 
  \url{https://www.tug.org/twg/mactex/tutorials/ltxprimer-1.0.pdf}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Track Changes:
% To add words, \added{<word added>}
% To delete words, \deleted{<word deleted>}
% To replace words, \replace{<word to be replaced>}{<replacement word>}
% To explain why change was made: \explain{<explanation>}

% At the end of the document, use \listofchanges, which will list the
% changes and the page and line number where the change was made.

% When final version, \listofchanges will not produce anything,
% \added{} word will be printed, \deleted{} will take away the word,
% \replaced{}{} will print only the 2nd argument.
% \explain will not print anything.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Track Changes}
ASA prefers that the Track Changes commands only be used to
track revisions.

\subsection{Using track changes commands}
Track changes commands will work only when the option
\verb+trackchanges+ is used, ie:\\
\verb+\documentclass[preprint,trackchanges]{JASA}+,\\
and only when the `preprint' option has been used.

Using the option `trackchanges' to the `preprint' documentclass\\
(\verb+\documentclass[preprint,trackchanges]{JASA}+)\\
activates the commands, \verb+\added{}+, \verb+\deleted{}+
and \verb+\replaced{}{}+ to mark changes that we've made.

\subsection{Available track changes commands}

To add words, \verb+\added{<word added>}+

To delete words, \verb+\deleted{<word deleted>}+

To replace words, \verb+\replace{<word to be replaced>}{<replacement word>}+

To explain why change was made: \verb+\explain{<explanation>}+

\subsection{Available option for track changes commands}

Comments can be used for additional information to the author, perhaps
a date, or the editor's initials, or more text.
\vskip12pt
To add comment when adding words, \verb+\added[comment]{<word added>}+

To delete words, \verb+\deleted[comment]{<word deleted>}+

To replace words, \verb+\replace[comment]{<word to be replaced>}{<replacement word>}+

To explain why change was made: \verb+\explain[comment]{<explanation>}+

\subsection{How to embed a citation in
{\tt\string\replace\string{\string}\string{\string}}}

Here's how to embed citations within \verb+\replaced{}{}:+\\
\verb+\replaced{\hbox{\citep{xyz}}{\hbox{\citep{zyx}}+.

ie,\\
\verb+\replaced{\hbox{\citep{booksamp1}}}{\hbox{\citep{joursamp1}}}+

\subsection{End of document, `list of changes'}

At the end of the document, use \verb+\listofchanges+ is called by
\verb+\end{document}+. It will list the
changes and the page and line number where the change was made.

When final version, \verb+\listofchanges+ will not produce anything,
\verb+\added{}+ word will be printed, \verb+\deleted{}+ will take away the word,
\verb+\replaced{}{}+ will print only the 2nd argument.
\verb+\explain{}+ will not print anything.

%% only use these samples if the `trackchanges' option has been used:

\subsection{Samples of track changes}

This shows `added': \added{This was added to the text}

%%
%% Use \explain{} before change, so that it ends up on the right line.
\explain{Redundant sentence, better without it. Do you mind?\\ --~JC}

Here is an example of deleted in the body  of a paragraph.\deleted{This was deleted from the text}

We replaced \replaced{XYZ}{ZYX}

At the end of the document, use \verb+\listofchanges+ is called by
\verb+\end{document}+. It will list the
changes and the page and line number where the change was made.

If `trackchanges' is not an option, \verb+\listofchanges+ will
not produce anything. \verb+\added{xyz}+ will put `xyz' in the
text; \verb+\delete{zzz}+ will produce nothing; and 
\verb+\replace{abc}{def}+ will leave `def' in your text.

\subsection{Add comment for the change?}
If you want to add a name to identify who made the change,
or any other comment, you can use  \verb+[]+ to enter a comment,
i.e.,\\ \verb+\added[Amy, Sept 14, 2019]+\\
\verb+{This was added to the text}+.

\added[Amy, Sept 14, 2019]{This was added to the text}

\deleted[Not really necessary]{This was deleted from the text}

We replaced \replaced[(written backwards originally)]{XYZ}{ZYX}

%% Track changes does not work when using `reprint' option.

\section{\label{sec:5}Conclusion}

And in conclusion\ldots

\begin{acknowledgments}
This research was supported by  ...
\end{acknowledgments}

\end{document}

### 2020-11-03 jasa-latex-template-v2019/bibsamp1.tex
adding JASA template as baseline

%% JASA, September 15, 2019

%% This file: BibSamp1.tex, 
%% showing Author-Year form of bibliography entries made with BibTeX.

%% Bibliography style:
%% author-year style is default; JASAauthyear2.bst is called automatically

\documentclass[preprint]{JASA}

\begin{document}

\section*{Bibliography made with BibTeX, Author-Year Examples}

Examples are based on the samples seen in JASA-ReferenceStyles.pdf which you
are encouraged to examine and use as a basis for the appearance of
your bibliography.\\
These are sample Author Year bibliography entries and results.
Author Year is the default bibliography style.

To make example:\\
 pdflatex bibsamp1, bibtex bibsamp1, pdflatex bibsamp1, pdflatex
 bibsamp1.

 See matching entries in sampbib.bib for examples of making the entries.

NOTE: Click on the citations to go to their referands. Enjoy!

The command \verb+\citep{bibitem}+ should usually be used for making
citations. \verb+\citep{joursamp1}+ will produce: (Christian {\it et
al.}, 1984). If you need only the year in parens, you may use
\verb+\citet{<bibitem>}+. When using \verb+\citet{joursamp1}+, for
example, you will produce: Christian {\it et al.} (1984).

\section*{Journal references}

Normal journal cite: \citep{joursamp1}

Volume number with issue number: \citep{joursamp3}

Journal article published online, not yet printed: \citep[published
online]{sampMisc2}

\section*{Book references}

Book reference \citep{booksamp1}

Revised edition \citep{booksamp4}

Reprinted in \citep{sampReprint2}

\section*{In press}

\citep[in press]{inpress2}

\citep[in press]{inpress3}

\section*{Translation}

\citep{translation}

\section*{Website examples}

\citep{websiteauthyear}.

\section*{Tech Report example}

\citep{samptechreport6}.

\section*{Dissertation}
\citep{sampthesis}

\section*{Patent}

\citep{samppatent2}

\section*{Standards}

\citep{amstand,ansi}

\section*{In Proceedings}

\citep{sampinproceedings3}

\section*{Computer Language Documentation}
Computer language documentation,

\citep{sampcode2}

\section*{Reprint}
Sample reprint, 
\citep{sampReprint}

\section*{Sample Series}
Sample Series,

\citep{sampSeries}

\section*{Sample E-Print}
 Sample E-Print 
\citep{sampEprint}

\section*{Miscellaneous}
Sample ISO specifications: \citep{isoSamp}.

\section*{Newspaper citation}
Here is a new field available for the *.bib file,
\verb+@{newpaper}+.

\cite{newspaperSamp}

The resulting bibliography entry should follow this form:

Author, A. (year). ``Article title,'' Newspaper name Volume number,
page number (Month day).

\bibliography{sampbib}

\end{document}

### 2020-11-03 jasa-latex-template-v2019/bibsamp2.tex
adding JASA template as baseline

%% JASA, September 15, 2019

% This file: BibSamp2.tex, showing numerical form of bibliography entries made with BibTeX

%% Bibliography style:
%% author-year style is default, NumberedRefs must be used in documentclass option
%% to get numbered references. JASAnum2.bst will then be used for the bibliography
%% style, automatically.

\documentclass[preprint, NumberedRefs]{JASA}

\begin{document}

\section*{Bibliography made with BibTeX, Numerical Examples}

You must use `NumberedRefs' as a documentclass option to get
numbered references.\\
Examples are based on the samples seen in JASA-ReferenceStyles.pdf which you
are encouraged to examine and use as a basis for the appearance of your bibliography.

To make example:\\
 pdflatex bibsamp2, bibtex bibsamp2, pdflatex bibsamp2, pdflatex
 bibsamp2.\\
 See matching entries in sampbib.bib for examples of making the entries.

NOTE: Click on the citations to go to their referands. Enjoy!

\section*{Journal references}

 Normal journal cite: \cite{joursamp1}

 Sample bib with only one page: \cite{joursamp2}.

 Volume number with issue number: \cite{yang}.

Journal article published online, not yet printed: \cite{samponline}.

\section*{Book references}

Edited by \cite{sampincollection3}.

Edited by \cite{sampincollection4}.

Book reference \cite{booksamp1}.

\section*{In press}

\cite{inpress1,inpressSmith}

\section*{Translation}

\cite{translation}

\section*{Website examples}

Citing websites\cite{website}.

\section*{Tech Report examples}
\cite{samptechreport3,samptechreport4}

\section*{Dissertation}
\cite{sampthesis}

\section*{Patent}

\cite{samppatent}

\section*{Standards}

\cite{sampstandards,sampstandards2}

\section*{In Proceedings}

 \cite{sampinproceedings2,
sampinproceedings3}

\section*{Computer Language Documentation}
Computer language documentation, 
\cite{sampcode,sampcode2}

\section*{Reprint}
Sample reprint, 
\cite{sampReprint}

\section*{Sample Series}
Sample Series, 
\cite{sampSeries}

\section*{Sample E-Print, (URL)}
 Sample E-Print 
\cite{sampEprint2}

\section*{Miscellaneous}
\cite{sampMisc}

\section*{Newspaper}
A new function has been added for use in the *.bib file: \verb+\@newspaper+

Here is a citation example: \cite{newspaperSamp}

The resulting bibliography entry should look like this:

\verb+ A. Author, ``Article title,'' Newspaper name XX, xxx-xxx (Month day, year).+

\bibliography{sampbib}

\end{document}

### 2020-11-03 jasa-latex-template-v2019/preprintsample.tex
adding JASA template as baseline

%	JASA LaTeX Sample File, Preprint Sample
%
%  Beginner Latex users should refer to their favorite online documentation
%  here is one from the TeX Users Group 
%	https://www.tug.org/twg/mactex/tutorials/ltxprimer-1.0.pdf
%
%  Useful FAQ from  https://journals.aps.org/revtex/revtex-faq
%

%%%%%%% For Preprint
%% For manuscript, 12pt, one column style

\documentclass[preprint]{JASA}

%%%%% Preprint Options %%%%%
%% The track changes option allows you to mark changes
%% and will produce a list of changes, their line number
%% and page number at the end of the article.
 %\documentclass[preprint,trackchanges]{JASA}

%% NumberedRefs is used for numbered bibliography and citations.
%% Default is Author-Year style.
% \documentclass[preprint,NumberedRefs]{JASA}

%%%%%%% For Reprint
%% For appearance of finished article; 2 columns, 10 pt fonts

% \documentclass[reprint]{JASA}

%%%%% Reprint Options %%%%%

%% For testing to see if author has exceeded page length request, use 12pt option
% \documentclass[reprint,12pt]{JASA}

%% NumberedRefs is used for numbered bibliography and citations.
%% Default is Author-Year style.
% \documentclass[reprint,NumberedRefs]{JASA}

%% TurnOnLineNumbers
%% Make lines be numbered in reprint style:
% \documentclass[reprint,TurnOnLineNumbers]{JASA}

%% Optional algorithm package. You can comment this out and
%% include another package if you prefer another way to make
%% algorithms examples (but please check that the package is compatible with 
%%Editorial Manager; see JASA-EL-TeXGuide.pdf).
\usepackage{algpseudocode}

\begin{document}

\title[JASA/Sample JASA Article]{Sample JASA Article}
\author{Author One}
\author{Author Two}
\author{Author Three}
\affiliation{Department1,  University1, City, State ZipCode, Country}

\author{Author Four}
\email{author.four@university.edu}
\affiliation{Department2,  University2, City, State ZipCode, Country}

\author{Author Five}			
\altaffiliation{Also at: Department, University, City, State ZipCode, Country.}
\affiliation{Department3,  University3, City, State ZipCode, Country}

\preprint{Author, JASA}		%  if you want want this message to appear in upper right corner of title page

\date{\today}

\begin{abstract}
Put your abstract here. Abstracts are limited to 200 words for
regular articles and 100 words for Letters to the Editor. Please no
personal pronouns, also please do not use the words ``new'' and/or
``novel'' in the abstract. An article usually includes an abstract, a
concise summary of the work covered at length in the main body of the
article.     
\end{abstract}

%% pacs numbers not used

\maketitle

%  End of title page for Preprint option --------------------------------- %

\section{\label{sec:1} Introduction}
This sample document demonstrates the use of JASA in manuscripts 
prepared for submission to the Journal of the Acoustical Society of America.

See JASA-TeXGuide.pdf, which is part of this package, for extensive
documentation on using commands for JASA.

You can compare the .tex version of this file with the resulting .pdf
version to give you an idea of what  commands are available and how
they work. At the top of the .tex file you'll find a listing of the
documentclass options, and an explanation of their results.
Some additional suggestions are included in the body of this
manuscript.

  Beginner Latex users should refer to their favorite online documentation. A 
  useful place to start is the primer from the TeX Users Group 
  \url{https://www.tug.org/twg/mactex/tutorials/ltxprimer-1.0.pdf}

EXAMPLE TEXT: This is example text. This is example text. This is example text. This is example text. This is example text. This is example text. This is example text. This is example text. This is example text. This is example text. This is example text. This is example text. This is example text. This is example text. This is example text. This is example text.

The paper is organized as follows: Section~\ref{sec:2} presents
initial information, while
Section~\ref{sec:3} presents examples of mathematical expressions.

 % Sample showing how to include figure, this is a floating figure

\begin{figure}[ht]
%% \reprintcolumnwidth is the same in preprint and reprint for
%% ease of use for authors:
\includegraphics[width=\reprintcolumnwidth]{figsamp.jpg}
\caption{\label{fig:FIG1}{Caption here.}}

\raggedright
Note: The only figure formats allowed are the following: 
.pdf, .ps, .eps, or .jpg. Figure files must be named in this fashion:
Figure\#.xxx, where ``\#'' is the figure number and ``xxx'' is the file format
(Figure1.eps, Figure2.jpg, Figure3a.ps, Figure3b.ps, etc).

[For these sample pages we have used only figsamp.jpg for convenience]
\end{figure}

\section{\label{sec:2} Section Two}

An example of another first-level Section with following example text that refers to subsections using 
\verb+\ref{subsec:XXX}+ ...  EXAMPLE: Some background in
section~\ref{sec:2} and details  in subsection~\ref{subsec:2:1}.

\subsection{\label{subsec:2:1} Sample subsection}

   \subsubsection{Sample subsubsection\label{subsubsec:1}}

\paragraph{Sample paragraph}Here is text following the paragraph
heading.
Here is a figure reference: is shown in Fig.~\ref{fig:FIG1}.

\section{Inline and display math samples\label{sec:3}}

\subsection{\label{subsec:3:3} Math and equations $\alpha\beta\Delta\Gamma$}
Inline math may be typeset using the \verb+$+ delimiters. Bold math
symbols may be achieved using the \verb+bm+ package and the
\verb+\bm{#1}+ command it supplies. For instance, a bold $\alpha$ can
be typeset as \verb+$\bm{\alpha}$+ giving $\bm{\alpha}$. Fraktur and
Blackboard (or open face or double struck) characters should be
typeset using the \verb+\mathfrak{#1}+ and \verb+\mathbb{#1}+ commands
respectively. Both are supplied by the \texttt{amssymb} package which
is called in JASA, so you don't need an \verb+\usepackage{amssymb}+
command in your .tex file. For
example, \verb+$\mathbb{R}$+ gives $\mathbb{R}$ and
\verb+$\mathfrak{G}$+ gives $\mathfrak{G}$.

In \LaTeX\ there are many different ways to display equations; a
few preferred ways are noted below. Displayed math will center by
default.

Below we have numbered single-line equations; this is the most common
type of equation.
\begin{eqnarray}
\chi_+(p){\bf [}2|{\bf p}|(|{\bf p}|+p_z){\bf ]}^{-1/2}
\left(
\begin{array}{c}
|{\bf p}|+p_z\\
px+ip_y
\end{array}\right)\;,
\\
\left\{%
 1234567890abc123\alpha\beta\gamma\delta1234556\alpha\beta
 \frac{1\sum^{a}_{b}}{A^2}%
\right\}%
\label{eq:one}.
\end{eqnarray}
Note the open one in Eq.~(\ref{eq:one}).

Not all numbered equations will fit within a narrow column this
way. The equation number will move down automatically if it cannot fit
on the same line with a one-line equation.
\begin{equation}
\chi_+(p){\bf [}2|{\bf p}|(|{\bf p}|+p_z){\bf ]}^{-1/2}
\alpha\beta\gamma\delta123455678\alpha\beta\Gamma\Delta
 \frac{1\sum^{a}_{b}}{A^2}1234%
\end{equation}
% No blank lines after \end{equation}
When the \verb+\label{#1}+ command is used [ie. input for
Eq.~(\ref{eq:one})], the equation can be referred to in text without
knowing the equation number that \TeX\ will assign to it. Just
use \verb+\ref{#1}+, where \verb+#1+ is the same name that used in
the \verb+\label{#1}+ command.

Unnumbered single-line equations can be typeset
using the \verb+\[+, \verb+\]+ format:
\[g^+g^+ \rightarrow g^+g^+g^+g^+ \dots ~,~~q^+q^+\rightarrow
q^+g^+g^+ \dots ~. \]

Note the equations can be lettered with the
subequations environment:
\begin{subequations}
\begin{eqnarray}
A&=&mc, \label{eqa}
\\
B&=&mc^2, \label{eqb}
\\
C&\agt& mc^3. \label{eqc}
\end{eqnarray}
\end{subequations}
Referenced: Eqs.~(\ref{eqa}), (\ref{eqb}), and (\ref{eqc}).

\section{Floats, Figures and Tables}

 Figures and tables are typically ``floats'' which means that their
final position is determined by \LaTeX\ while the document is being
typeset. \LaTeX\ isn't always successful in placing floats
optimally.  Use the figure* environment to get a wide figure that spans the 
page in a two-column layout.

\subsection{\label{subsec:3:2} Tables}
Tables generally should be surrounded with
\verb+\begin{ruledtabular}...\end{ruledtabular}+\\
This will guarantee that they are the width of the
page or column, and have two ruled lines at the top
and bottom of the table.

\verb+[ht]+ in the code below instructs \LaTeX\ to place the table
where it appears in type, if it will fit on the page;
otherwise put it on the top of the next page.

Footnotes in a table are labeled a, b, c,
 etc.  They can
be  specified  by  using  the  \LaTeX\
\verb+\footnotemark[]+
and
\verb+\footnotetext[]+ commands.
The  footnotes  for  a  table  are  typeset  at  the
bottom  of  the  table,  rather  than  at  the  bottom  of  the
page or at the end of the references.  The arguments for
\verb+\footnotemark[]+
and
\verb+\footnotetext[]+
should be numbers
1, 2, \ldots\  The journal style will convert these to letters.

This system allows
multiple entries to refer to the same
footnote.

\begin{table}[ht]
\caption{\label{tab:table1}A table with more columns still fits
properly in a column. Note that several entries share the same
footnote. Inspect the \LaTeX\ input for this table to see
exactly how it is done.}

\begin{ruledtabular}
\begin{tabular}{cccccccc}
 &$r_c$ (\AA)\footnotemark[1]&$r_0$ (\AA)&$\kappa r_0$&
 &$r_c$ (\AA) &$r_0$ (\AA)&$\kappa r_0$\\
\hline
Cu& 0.800 & 14.10 & 2.550 &Sn\footnotemark[1]
& 0.680 & 1.870 & 3.700 \\
Ag& 0.990 & 15.90 & 2.710 &Pb\footnotemark[2]
& 0.450 & 1.930 & 3.760 \\
%Au& 1.150 & 15.90 & 2.710 &Ca\footnotemark[3]
%& 0.750 & 2.170 & 3.560 
\end{tabular}
\end{ruledtabular}
\footnotetext[1]{Here's the first.}
\footnotetext[2]{Here's the second.}
%\footnotetext[3]{Here's the third.}
\end{table}
\clearpage
\subsection{Plain Tables: When NOT to use `ruledtabular'}
There are a number of cases when `ruledtablar' should not
be used: basically for any table using complex content or 
commands.

\subsubsection{Using {\tt\string\multicolumn}}
When you'd like to use the multicolumn command in your
table, you'll find that `ruledtabular' will cause bad
formatting. In that case, Don't Use Ruledtabular, and
instead put in \verb+\hline\hline+ at the top and bottom of
the table.

\begin{table}[ht]
\caption{A table made without `ruledtabular' needs to have two hlines
added to the top and bottom of the table.}
\vskip3pt
\begin{tabular}{cccccccc}
\hline\hline
 &$r_c$ (\AA)\footnotemark[1]&$r_0$ (\AA)&$\kappa r_0$&
 &$r_c$ (\AA) &$r_0$ (\AA)&$\kappa r_0$\\
\hline
Cu& 0.800 & 14.10 & 2.550 &Sn\footnotemark[1]
& 0.680 & 1.870 & 3.700 \\
Ag& 0.990 & 15.90 & 2.710 &Pb\footnotemark[2]
& 0.450 & 1.930 & 3.760 \\
Au& 1.150 & 15.90 & 2.710 &Ca\footnotemark[3]
& 0.750 & 2.170 & 3.560\\
\hline\hline
\end{tabular}
\footnotetext[1]{This is the first table note.}
\footnotetext[2]{This is the second table note.}
\footnotetext[3]{This is the third table note.}
\end{table}

\newpage

\vskip12pt
\vskip1sp
\subsubsection{Using the {\tt\string\adjustbox\string{\string}
(tabular)\string\end\string{adjustbox\string}}
command}
There may be times when the table is too wide, or you want to have
the table be the width of the page, whether or not it appears in
preprint or reprint version of JASA. In this case you can use
\verb+\begin{adjustbox}{<width>} (tabular) \end{adjustbox}+.
(`adjustbox' will NOT work with `ruledtabular')

You can set a maximum width with \\
\verb+\begin{adjustbox}{max width=\textwidth}(tabular)\end{adjustbox}+\\  
in which case the table in the reprint version might be less than
the full text width;

Or you can set the exact width you'd like with\\
\verb+\begin{adjustbox}{width=\textwidth}(tabular)\end{adjustbox}+\\
in which case the table will be the full width of the page in
either preprint or reprint.

This way you can make a table that will fit in the correct width
whether you are using the preprint or reprint option.

\begin{table*}[b]
\caption{Top~5 rated $\widehat{\textrm{ITD}}$ estimation methods
according to the sum and product metric criteria for $\pm0.5$~JND and
$\pm1$~JND tolerance thresholds (normalized scores).}

\begin{adjustbox}{width=\textwidth}
\begin{tabular}{c| l l| l l| l l| l l}
\hline\hline
Rank \#& \multicolumn{2}{c}{sum criteria [$\pm0.5$~JND]} &
\multicolumn{2}{|c}{sum criteria [$\pm1$~JND]} &
\multicolumn{2}{|c}{product criteria [$\pm0.5$~JND]} & \multicolumn{2}{|c}{product criteria [$\pm1$~JND]}\\ 
\hline
1 & Threshold --30dB {lp} & (0.43) & Threshold --30dB {lp} & (0.71) & Threshold --30dB {lp} & (1.00) & Threshold --30dB {lp} & (1.00) \\ 
2 & MaxIACCe {lp} & (0.39) & Threshold --20dB {lp} & (0.66) & MaxIACCe {lp} & (0.39) & Threshold --20dB {lp} & (0.57) \\ 
3 & Threshold --20dB {lp} & (0.38) & CenIACCr {bb} &(0.62) & CenIACCr {lp} &(0.33) & CenIACCr {bb} & (0.37) \\ 
4 & CenIACCr {lp} &(0.37) & MaxIACCe {lp} & (0.61) & Threshold --20dB {lp} & (0.29) & MaxIACCe {lp} & (0.34) \\ 
5 & Cen-$e^2$ {lp} &(0.34) & CenIACCe {lp} & (0.61) & Cen-$e^2$ {lp}
& (0.10) & CenIACCr {lp} & (0.33) \\
\hline\hline
\end{tabular}
\end{adjustbox}

\label{tab:topMethods} 
\end{table*}

\clearpage

\subsection{Using dcolumn}
The call to \verb+\usepackage{dcolumn}+ is included in JASA.cls so you
don't need to add it explicitly.
\href{http://anorien.csc.warwick.ac.uk/mirrors/CTAN/macros/latex/required/tools/dcolumn.pdf}
{\url{http://anorien.csc.warwick.ac.uk/mirrors/CTAN/macros/latex/required/tools/dcolumn.pdf}}
will give you detailed information.
A gentler introduction may be found in this informative and well
illustrated
article: \href{https://www.tug.org/pracjourn/2007-1/mori/mori.pdf}
{\url{https://www.tug.org/pracjourn/2007-1/mori/mori.pdf}}, starting on page
20. (You may want to look at more examples in this quite comprehensive
article on making tables in \LaTeX.)

\begin{quote}
``If we do not want to break the fractional and the integral part in two columns,
the dcolumn package provides a new type of column\\
\verb+D{sep -in}{sep -out}{ before.after}+\\
The first argument \verb+{sep-in}+ is the symbol used in the
.tex document to separate
the integral and the fractional part (usually the decimal point . or the decimal
comma ,), the second argument \verb+{sep-out}+
is the symbol that we want in the
output, the third is the number of digits on the left (before) and on the right
(after) this symbol. The numbers are aligned to the decimal point and, in case
that the third argument is negative, the decimal point is aligned to the center of
the column. If the columns have a heading, it must be inserted into
the command \verb+\multicolumn{1}{c}{...}+
\end{quote}
\newpage
An example using dcolumn:
\begin{verbatim}
{\hsize= 2in
\begin{ruledtabular}
\begin{tabular}{cD {,}{.}{5.4}}
Expression           & \multicolumn {1}{c}{ Value }\\
\hline
$\pi$                  &      3,1416                 \\
$\pi^{\pi}$           &     36,46                    \\
$\pi^{\pi^{\pi}}$    & 80662,7                      \\
\end{tabular}
\end{ruledtabular}
}
\end{verbatim}
\vskip12pt
{\hsize= 2in
\begin{ruledtabular}
\begin{tabular}{cD {,}{.}{5.4}}
Expression           & \multicolumn {1}{c}{ Value }\\
\hline
$\pi$                  &      3,1416                 \\
$\pi^{\pi}$           &     36,46                    \\
$\pi^{\pi^{\pi}}$    & 80662,7                      \\
\end{tabular}
\end{ruledtabular}
}

\newpage

\begin{figure*}[p]
\parskip=6pt
\baselineskip=12pt

\subsection{Sample Figures, new commands available in this style}
\vglue-14pt
{\it Note that the publisher determines the final layout, so 
your choice of figure alignment may not be reflected in the published
article.}

\noindent
\verb+\figline{}+ will center one or more figures on one line.

\noindent
\verb+\fig{<name of file>}{<width>}{<letter to put underneath>}+

\noindent
\verb+\leftfig{<name of file>}{<width>}{<letter to put underneath>}+

\noindent
\verb+\rightfig{<name of file>}{<width>}{<letter to put underneath>}+

\noindent
\verb+\boxedfig{<name of file>}{<width>}{<letter to put underneath>}+

\noindent
\verb+\rotatefig{<degrees of rotation>}{<name of file>}{<width>}+\\
\verb+{<letter to put underneath>}+

The following illustrations show these commands in use.

\begin{verbatim}
\figline{\fig{figsamp.jpg}{4cm}{(a)}
\fig{figsamp.jpg}{4cm}{(b)}}
\figline{\fig{figsamp.jpg}{4cm}{(c)}
\fig{figsamp.jpg}{4cm}{(d)}}
\figline{\fig{figsamp.jpg}{4cm}{(e)}}
\end{verbatim}

\figline{\fig{figsamp.jpg}{4cm}{(a)}
\fig{figsamp.jpg}{4cm}{(b)}}
\figline{\fig{figsamp.jpg}{4cm}{(c)}
\fig{figsamp.jpg}{4cm}{(d)}}
\figline{\fig{figsamp.jpg}{4cm}{(e)}}

\caption{ \label{fig:pressure_field} Multiple images on one figure example (a) image 1, (b-f) ($\rho$=1000 kg/m$^3$) and speed of sound ($c$=1500 m/s).}

\end{figure*}

\begin{figure*}
\baselineskip=12pt
\begin{verbatim}
\figline{\boxedfig{figsamp.jpg}{2in}{(a)}}
\figline{\leftfig{figsamp.jpg}{2in}{(b)}\rightfig{figsamp.jpg}{2in}{(c)}}
\figline{\rotatefig{90}{figsamp.jpg}{2in}{(d)}\rotatefig{180}{figsamp.jpg}{2in}{(e)}}
\end{verbatim}

\figline{\boxedfig{figsamp.jpg}{2in}{(a)}}

\figline{\leftfig{figsamp.jpg}{2in}{(b)}\rightfig{figsamp.jpg}{2in}{(c)}}

\figline{\rotatefig{90}{figsamp.jpg}{2in}{(d)}
\rotatefig{180}{figsamp.jpg}{2in}{(e)}
}

\caption{More figure examples: (a) boxedfig, 
(b)leftfig; (c)right fig; (d) rotatefig 90 degrees;
(e) rotatefig 180 degrees. }

\end{figure*}

\clearpage
\section{Labels in Figline}
We can label and
reference separate parts of the figure when using figline.
The reference will give the illustration letter as well as the figure
number.

To label figures used in \verb+\figline{}+ type in your
label immediately after the \verb+\fig{}{}{}+ command, inside the
argument to figline. For example:

\begin{verbatim}
\figline{\fig{<name of file>}{<width>}{<letter to put underneath>}\label{<labelname>}}
\end{verbatim}

The same placement should be used for all the kinds of fig
environments used in \verb+\figline{}+:
\begin{verbatim}
\fig{}{}{}\label{}, \leftfig{}{}{}\label{}, \rightfig{}{}{}\label{}, 
\boxedfig{}{}{}\label{}, \rotatefig{}{}{}\label{},  \narrowcaption{}{}\label{}.
\end{verbatim}

\begin{figure*}
\baselineskip=12pt
\begin{verbatim}
\figline{\boxedfig{figsamp.jpg}{2in}{(a)}\label{boxedfigLetter}}
\figline{\leftfig{figsamp.jpg}{2in}{(b)}
\rightfig{figsamp.jpg}{2in}{(c)}\label{rightfigLetter}}
\figline{\rotatefig{90}{figsamp.jpg}{2in}{(d)}
\rotatefig{180}{figsamp.jpg}{2in}{(e)}\label{rotatefigLetter}}
\end{verbatim}

\figline{\boxedfig{figsamp.jpg}{2in}{(a)}\label{boxedfigLetter}}
\figline{\leftfig{figsamp.jpg}{2in}{(b)}\rightfig{figsamp.jpg}{2in}{(c)}\label{rightfigLetter}}
\figline{\rotatefig{90}{figsamp.jpg}{2in}{(d)}\rotatefig{180}{figsamp.jpg}{2in}{(e)}\label{rotatefigLetter}}

\caption{More figure examples, showing how to enter
{\tt\string\label\string{\string}} command.}

\verb+References:\ref{boxedfigLetter},\ref{rightfigLetter},\ref{rotatefigLetter}+

which produces\\

References: \ref{boxedfigLetter},\ref{rightfigLetter},\ref{rotatefigLetter}

\end{figure*}

\begin{figure*}
\baselineskip=12pt
\begin{verbatim}
\sidebysidefigures{figsamp.jpg}{Describing the first
illustration.}/{figsamp.jpg}{Describing the second illustration.}
\end{verbatim}

\sidebysidefigures{figsamp.jpg}{Describing the first
illustration.}/{figsamp.jpg}{Describing the second illustration.}
\end{figure*}

\begin{figure*}
\baselineskip=12pt
\begin{verbatim}
\figline{
\fig{figsamp.jpg}{.7\textwidth}{}
\narrowcaption{.2\textwidth}{Here is a narrow caption.}
}
\end{verbatim}
\figline{\fig{figsamp.jpg}{.7\textwidth}{} \narrowcaption{.2\textwidth}{Here is a narrow
caption.}}
\end{figure*}

\begin{figure*}
\baselineskip=12pt
\begin{verbatim}
\figline{\fig{figsamp.jpg}{.2\textwidth}{(a)}
\fig{figsamp.jpg}{.2\textwidth}{(b)}
\fig{figsamp.jpg}{.2\textwidth}{(c)}
\narrowcaption{.25\textwidth}{Caption for three illustrations. 
The caption may produce many lines, but only one paragraph.
}}
\end{verbatim}
\figline{\fig{figsamp.jpg}{.2\textwidth}{(a)}
\fig{figsamp.jpg}{.2\textwidth}{(b)}
\fig{figsamp.jpg}{.2\textwidth}{(c)}
\narrowcaption{.25\textwidth}{Caption for three illustrations. 
The caption may produce many lines, but only one paragraph.
}}
\end{figure*}

\begin{figure*}
\baselineskip=12pt
\begin{verbatim}
\vtop to 0pt{\vss
\figline{\leftfig{figsamp.jpg}{.2\textwidth}{(a)}
\leftfig{figsamp.jpg}{.2\textwidth}{(b)}\hfill}
\figline{\leftfig{figsamp.jpg}{.2\textwidth}{(c)}
\leftfig{figsamp.jpg}{.2\textwidth}{(d)}\hfill}
\vss}
\narrowcaption{.25\textwidth}{Here is a narrow caption that will can be
positioned to the right of four illustrations.
You cannot have more than one paragraph of text in a caption.
You cannot have more than one paragraph of text in a caption.
You cannot have more than one paragraph of text in a caption.
You cannot have more than one paragraph of text in a caption.
}
\vskip36pt % to give more space above the footline
\end{verbatim}
\vskip1.3in
\vtop to 0pt{\vss
\figline{\leftfig{figsamp.jpg}{.2\textwidth}{(a)}
\leftfig{figsamp.jpg}{.2\textwidth}{(b)}\hfill}
\figline{\leftfig{figsamp.jpg}{.2\textwidth}{(c)}
\leftfig{figsamp.jpg}{.2\textwidth}{(d)}\label{test}\hfill}
\vss}
\narrowcaption{.25\textwidth}{Here is a narrow caption that will can be
positioned to the right of four illustrations.
You cannot have more than one paragraph of text in a caption.
You cannot have more than one paragraph of text in a caption.
You cannot have more than one paragraph of text in a caption.
You cannot have more than one paragraph of text in a caption.
}
\vskip36pt
\end{figure*}

\clearpage

\begin{figure}[h]
\baselineskip=12pt
\begin{verbatim}
\figcolumn{
\fig{figsamp.jpg}{.2\textwidth}{(a)}
\fig{figsamp.jpg}{.2\textwidth}{(b)}
\fig{figsamp.jpg}{.2\textwidth}{(c)}
}
\end{verbatim}

\figcolumn{
\fig{figsamp.jpg}{.2\textwidth}{(a)}
\fig{figsamp.jpg}{.2\textwidth}{(b)}
\fig{figsamp.jpg}{.2\textwidth}{(c)}
}

\caption{Here are some stacking figures in a single column.
The {\tt\string\figcolumn\string{\string}} command works equally well in single
or double column figures.}
\end{figure}

\begin{nofloatfigure}
\centering
\includegraphics[width=2in]{figsamp}
\caption{\label{nofloatcaption}
Showing how you can have a caption that
continues across pages or columns.\\
This is a caption in a no float figure.
It is designed to continue across columns
or pages if it is particularly long.
This is a caption that will continue
across pages if necessary.
This is a caption that will continue
across pages if necessary.
This is a caption that will continue
across pages if necessary.
This is a caption that will continue
across pages if necessary.
This is a caption that will continue
across pages if necessary.
This is a caption that will continue
across pages if necessary.
This is a caption that will continue
across pages if necessary.
This is a caption that will continue
across pages if necessary.
This is a caption that will continue
across pages if necessary.
This is a caption that will continue
across pages if necessary.
}
\end{nofloatfigure}

\newpage
\section{Algorithm Examples}
This example uses \verb+\usepackage{algpseudocode}+ as you see above. If
you would rather use another algorithm package, please comment out
this package
 (\verb+%\usepackage{algpseudocode}+) and type in the
package name that you'd like to use (but please check that the package is compatible with Editorial Manager; see JASA-EL-TeXGuide.pdf).

For other algorithm packages, see
\url{https://en.wikibooks.org/wiki/LaTeX/Algorithms}

Here is an example of algorithmic:

\begin{algorithmic}
\If {$i\geq maxval$}
    \State $i\gets 0$
\Else
    \If {$i+k\leq maxval$}
        \State $i\gets i+k$
    \EndIf
\EndIf
\end{algorithmic}

Documentation for the package is found at 
\url{http://mirrors.rit.edu/CTAN/macros/latex/contrib/algorithm2e/doc/algorithm2e.pdf}
which shows many other examples and options.

\newpage
\subsection{Example of multimedia entry}
Please note that this is for multimedia intended to appear inline
within the published article.

Here is what a multimedia entry will look like:
\multimedia{http://dx.doi.org/10.1121/1.4947423.1}{Corresponding pulse-compressed echo envelopes
and video recordings from a fluttering luna moth.
Echoes from the wings and body of the moth generally dominate the
acoustic returns, which vary greatly over consecutive ensonifications
across the wingbeat cycle. File of type ``mp4'' (15.3
MB)}\label{mmtest1}

Here we try cross referencing the multimedia entry: The multimedia
above is Mm.~\ref{mmtest1}.

\subsection{Supplementary Material}
ASA
prefers that authors to submit related/relevant article files as
supplementary material with their submission.

\subsection{Supplementary material for publication}
Any archival supplemental materials to be published with the
manuscript (eg., supplementary figures) should be cited in-text and a footnote provided.

An example of reference to supplementary material:

The sound files and videos for this and other figures
are included as supplementary materials\footnote{See
Supplementary materials at [URL will be inserted by AIP]
for [give a brief description of the material].}.

The contents of the footnote above will appear at the beginning of the
bibliography made with BibTeX when the default `author-year' documentclass option is used;
BibTeX output will have the footnote interleaved with other
references if the NumberedRefs documentclass option is used.

\subsection{File naming conventions}
Here are the conventions for naming files:

\begin{itemize}
\item
Supplementary Figure or
	Supplementary Figure or Text files should be named: SuppPub\#.xxx, where ``\#'' is
	a number and ``xxx" is the file format extension
	(SuppPub1.docx, SuppPub2.jpg, etc)

\item
	Supplementary Multimedia files: SuppPubmm\#.xxx, where ``\#'' is a
	number and ``xxx'' is the file format extension (SuppPubmm1.mp3,
	SuppPubmm2.gif, etc)

\item
Multimedia files must be named accordingly: MM\#.xxx, where ``\#'' is the
number and ``xxx'' is the file format extension (MM1.wav, MM2.avi, etc).

\item
The only figure formats allowed are the following: 
.pdf, .ps, .eps, or .jpg. Figure files must be named in this fashion:
Figure\#.xxx, where ``\#'' is the figure number and ``xxx'' is the file format
(Figure1.eps, Figure2.jpg, Figure3a.ps, Figure3b.ps, etc).

\end{itemize}

\section{\label{sec:5}Conclusion}

And in conclusion\ldots

\begin{acknowledgments}
This research was supported by  ...
\end{acknowledgments}

 % -------------------------------------------------------------------------------------------------------------------
 %   Appendix  (optional)

\appendix
\section{Appendices}
To start the appendix, use the \verb+\appendix+ command.
This signals that all following section commands refer to appendixes
instead of regular sections. Therefore, the \verb+\appendix+ command
should be used only once---to set up the section commands to act as
appendices. Thereafter normal section commands are used. The heading
for a section can be left empty. For example,
\begin{verbatim}
\appendix
\section{}
\end{verbatim}
will produce an appendix heading that says ``APPENDIX A'' and
\begin{verbatim}
\appendix
\section{Background}
\end{verbatim}
will produce an appendix heading that says ``APPENDIX A: BACKGROUND''
(note that the colon is set automatically).

If there is only one appendix, then the letter ``A'' should not
appear. This is suppressed by using the star version of the appendix
command (\verb+\appendix*+ in the place of \verb+\appendix+).
\vskip12pt
\vskip1sp
\section{A little more on appendices}

Observe that this appendix was started by using
\begin{verbatim}
\section{A little more on appendixes}
\end{verbatim}

Note the equation number in an appendix:
\begin{equation}
E=mc^2.
\end{equation}

\subsection{\label{app:subsec}A subsection in an appendix}

You can use a subsection or subsubsection in an appendix. Note the
numbering: we are now in Appendix~\ref{app:subsec}.

\subsubsection{\label{app:subsubsec}A subsubsection in an appendix}
Note the equation numbers in this appendix, produced with the
subequations environment:
\begin{subequations}
\begin{eqnarray}
E&=&mc, \label{appa}
\\
E&=&mc^2, \label{appb}
\\
E&\agt& mc^3. \label{appc}
\end{eqnarray}
\end{subequations}
They turn out to be Eqs.~(\ref{appa}), (\ref{appb}), and (\ref{appc}).

\section{Sample appendix figure and table}
Figure and table numbering are continuous through the article,
and handled the same as they are in the rest of the article.

\begin{figure}[ht]
\includegraphics[width=1in]{figsamp.jpg}
\caption{Figure in an appendix.}
\end{figure}

\begin{table}[ht]
\caption{Here is the caption for a table in an appendix.}
\centering
\begin{ruledtabular}
\begin{tabular}{cccc}
one&two&three&four\\
\hline
C&D&E&F\\
\end{tabular}
\end{ruledtabular}
\vskip12pt
\end{table}

\subsection{Footnotes}
The contents of the footnotes will appear at the beginning of the
bibliography when BibTeX produces the .bbl file using the default
AuthorYear style; interleaved with other references if 
Numbered\-Refs option:\\
\verb+\documentclass[preprint,NumberedRefs]{JASA}+\\
and BibTeX has been used.

This example show where this cite \citep{booksamp1} will appear in the
bibliography,
depending on whether we use default author-year style
or call for the NumberedRefs documentclass option.

Here are some sample footnotes:\footnote{Here is the second footnote.
It will appear before the beginning of the bibliography in Author-Year
style (default) or it will be 
 interleaved with other references when using the Numbered\-Refs
 option.}$^,$% (comma between two sequential footnotes)
\footnote{Here is a third footnote.}

\section{Making the Bibliography Using BibTeX}
Authors are highly  recommended to use BibTeX to produce their
bibliographies. The results will be predictable and even if
it might take some time to get comfortable with  using BibTeX,
in the long run it will save you endless aggravation.

A resource for making your bibliography entries
correctly is included in this package: 
JASA-ReferenceStyles.pdf. You will also find
the files
bibsamp1.tex/.pdf and bibsamp2.tex/.pdf
for examples of output; and sampbib.bib for an example of
how to make your .bib database entries.

There are two possible bibliography styles: the default, author-year,
and the optional style, Numbered\-Refs, which you would call using\\
{\verb+\documentclass[preprint,NumberedRefs]{JASA}+ }

\verb+\citep{}+ should normally be used rather than \verb+\cite{}+.

You can also use \verb+\citet{}+ if it is more grammatically correct
to have only the year in parens (note: this is used with author-year style references).

\begin{verbatim}
\citep{bibitemName} = (bibitemName, year)
or
\citet{bibitemName} = bibitemName (year)
\end{verbatim}

Note that the citations are hyperlinked to their entries in the
bibliography:

Normal journal cite: \citep{joursamp1},
 Book reference \citet{booksamp1}, 
Computer language documentation:
\citep{sampcode2}.

Every \verb+\citep+  or \verb+\citet{}+ will produce a citation and an entry in the
bibliography. Every citation must have a matching entry in the
bibliography
database file (\verb+\filename.bib+).

Make your bibliography by doing: pdflatex filename,  bibtex filename,
pdflatex filename, pdflatex filename.

{\bfseries\itshape
Compare the results you get with\\
{\verb+\documentclass[preprint]{JASA}+ }\\
vs.\\
{\verb+\documentclass[preprint,NumberedRefs]{JASA}+ }
}

\bibliography{sampbib}
\end{document}

### 2020-11-03 jasa-latex-template-v2019/reprintsample.tex
adding JASA template as baseline

--- 2020-11-03 jasa-latex-template-v2019/preprintsample.tex
@@
 
-\documentclass[preprint]{JASA}
+% \documentclass[preprint]{JASA}
 
@@
 
-% \documentclass[reprint]{JASA}
+ \documentclass[reprint]{JASA}
 
@@
 %% include another package if you prefer another way to make
-%% algorithms examples (but please check that the package is compatible with 
-%%Editorial Manager; see JASA-EL-TeXGuide.pdf).
+%% algorithms examples.
 \usepackage{algpseudocode}
@@
 \author{Author Five}			
-\altaffiliation{Also at: Department, University, City, State ZipCode, Country.}
+\altaffiliation{Also at: Another University, City, State ZipCode, Country.}
 \affiliation{Department3,  University3, City, State ZipCode, Country}
 
-\preprint{Author, JASA}		%  if you want want this message to appear in upper right corner of title page
+\preprint{Author, JASA}		%  if you want want this message to appear in upper right corner of title page for preprint only
 
@@
 \maketitle
-
-%  End of title page for Preprint option --------------------------------- %
 
@@
 
-EXAMPLE TEXT: This is example text.
-This is example text.
-This is example text.
-This is example text.
-This is example text.
-This is example text.
-This is example text.
-This is example text.
-This is example text.
-This is example text.
-This is example text.
-This is example text.
-This is example text.
-This is example text.
-This is example text.
-This is example text.
-
 The paper is organized as follows: Section~\ref{sec:2} presents
@@
 
- % Sample showing how to include figure, this is a floating figure
-
-\begin{figure}[ht]
-%% \reprintcolumnwidth is the same in preprint and reprint for
-%% ease of use for authors:
-\includegraphics[width=\reprintcolumnwidth]{figsamp.jpg}
-\caption{\label{fig:FIG1}{Caption here.}}
-
-\raggedright
-Note: The only figure formats allowed are the following: 
-.pdf, .ps, .eps, or .jpg. Figure files must be named in this fashion:
-Figure\#.xxx, where ``\#'' is the figure number and ``xxx'' is the file format
-(Figure1.eps, Figure2.jpg, Figure3a.ps, Figure3b.ps, etc).
-
-[For these sample pages we have used only figsamp.jpg for convenience]
-\end{figure}
+{\it
+[For these sample pages we have used only figsamp.jpg for
+convenience]}
 
 \section{\label{sec:2} Section Two}
-
 An example of another first-level Section with following example text that refers to subsections using 
@@
 \subsubsection{Sample subsubsection\label{subsubsec:1}}
+
+% Sample showing how to include figure, this is a floating figure
+
+\begin{figure}
+%% \reprintcolumnwidth is the same in preprint and reprint for
+%% ease of use for authors:
+\includegraphics[width=\reprintcolumnwidth]{figsamp.jpg}
+\caption{\label{fig:FIG1}{Caption here.}}
+\raggedright
+Note: The only figure formats allowed are the following: 
+.pdf, .ps, .eps, or .jpg. Figure files must be named in this fashion:
+Figure\#.xxx, where ``\#'' is the figure number and ``xxx'' is the file format
+(Figure1.eps, Figure2.jpg, Figure3a.ps, Figure3b.ps, etc). 
+\end{figure}
 
@@
 respectively. Both are supplied by the \texttt{amssymb} package which
-is called in JASA, so you don't need an \verb+\usepackage{amssymb}+
-command in your .tex file. For
-example, \verb+$\mathbb{R}$+ gives $\mathbb{R}$ and
+is called in JASA, so no need enter \verb+\includepackage{amssymb}+.
+To see amssymb in action:
+\verb+$\mathbb{R}$+ gives $\mathbb{R}$ and
 \verb+$\mathfrak{G}$+ gives $\mathfrak{G}$.
@@
 \end{table}
-\clearpage
+
 \subsection{Plain Tables: When NOT to use `ruledtabular'}
-There are a number of cases when `ruledtablar' should not
+There are a number of cases when `ruledtabular' should not
 be used: basically for any table using complex content or 
@@
 
-\newpage
-
-\vskip12pt
-\vskip1sp
 \subsubsection{Using the {\tt\string\adjustbox\string{\string}
-(tabular)\string\end\string{adjustbox\string}}
-command}
+(tabular)\\
+\string\end\string{adjustbox\string}} command}
 There may be times when the table is too wide, or you want to have
@@
 preprint or reprint version of JASA. In this case you can use
-\verb+\begin{adjustbox}{<width>} (tabular) \end{adjustbox}+.
+\verb+\begin{adjustbox}{<width>}+\\
+\verb+(tabular) \end{adjustbox}+.\\
 (`adjustbox' will NOT work with `ruledtabular')
@@
 You can set a maximum width with \\
-\verb+\begin{adjustbox}{max width=\textwidth}(tabular)\end{adjustbox}+\\  
+\verb+\begin{adjustbox}{max width=\textwidth}(tabular)+\\
+\verb+\end{adjustbox}+\\  
 in which case the table in the reprint version might be less than
@@
 Or you can set the exact width you'd like with\\
-\verb+\begin{adjustbox}{width=\textwidth}(tabular)\end{adjustbox}+\\
+\verb+\begin{adjustbox}{width=\textwidth}(tabular)+\\
+\verb+\end{adjustbox}+\\
 in which case the table will be the full width of the page in
@@
 
-\begin{table*}[b]
-\caption{Top~5 rated $\widehat{\textrm{ITD}}$ estimation methods
-according to the sum and product metric criteria for $\pm0.5$~JND and
-$\pm1$~JND tolerance thresholds (normalized scores).}
-
-\begin{adjustbox}{width=\textwidth}
-\begin{tabular}{c| l l| l l| l l| l l}
-\hline\hline
-Rank \#& \multicolumn{2}{c}{sum criteria [$\pm0.5$~JND]} &
-\multicolumn{2}{|c}{sum criteria [$\pm1$~JND]} &
-\multicolumn{2}{|c}{product criteria [$\pm0.5$~JND]} & \multicolumn{2}{|c}{product criteria [$\pm1$~JND]}\\ 
-\hline
-1 & Threshold --30dB {lp} & (0.43) & Threshold --30dB {lp} & (0.71) & Threshold --30dB {lp} & (1.00) & Threshold --30dB {lp} & (1.00) \\ 
-2 & MaxIACCe {lp} & (0.39) & Threshold --20dB {lp} & (0.66) & MaxIACCe {lp} & (0.39) & Threshold --20dB {lp} & (0.57) \\ 
-3 & Threshold --20dB {lp} & (0.38) & CenIACCr {bb} &(0.62) & CenIACCr {lp} &(0.33) & CenIACCr {bb} & (0.37) \\ 
-4 & CenIACCr {lp} &(0.37) & MaxIACCe {lp} & (0.61) & Threshold --20dB {lp} & (0.29) & MaxIACCe {lp} & (0.34) \\ 
-5 & Cen-$e^2$ {lp} &(0.34) & CenIACCe {lp} & (0.61) & Cen-$e^2$ {lp}
-& (0.10) & CenIACCr {lp} & (0.33) \\
-\hline\hline
-\end{tabular}
-\end{adjustbox}
-
-\label{tab:topMethods} 
-\end{table*}
-
-\clearpage
-
 \subsection{Using dcolumn}
-The call to \verb+\usepackage{dcolumn}+ is included in JASA.cls so you
+\verb+\usepackage{dcolumn}+ is included in JASA.cls so you
 don't need to add it explicitly.
@@
 \end{quote}
-\newpage
+
 An example using dcolumn:
@@
 
+%% Temporarily going into one column mode is very helpful in making
+%% up pages with multiple figures and tables. The commands are
+%% \onecolumngrid ...\twocolumngrid
+
+\onecolumngrid
+
+\begin{table*}[h]
+\caption{Top~5 rated $\widehat{\textrm{ITD}}$ estimation methods
+according to the sum and product metric criteria for $\pm0.5$~JND and
+$\pm1$~JND tolerance thresholds (normalized scores).}
+
+\begin{adjustbox}{width=\textwidth}
+\begin{tabular}{c| l l| l l| l l| l l}
+\hline\hline
+Rank \#& \multicolumn{2}{c}{sum criteria [$\pm0.5$~JND]} &
+\multicolumn{2}{|c}{sum criteria [$\pm1$~JND]} &
+\multicolumn{2}{|c}{product criteria [$\pm0.5$~JND]} & \multicolumn{2}{|c}{product criteria [$\pm1$~JND]}\\ 
+\hline
+1 & Threshold --30dB {lp} & (0.43) & Threshold --30dB {lp} & (0.71) & Threshold --30dB {lp} & (1.00) & Threshold --30dB {lp} & (1.00) \\ 
+2 & MaxIACCe {lp} & (0.39) & Threshold --20dB {lp} & (0.66) & MaxIACCe {lp} & (0.39) & Threshold --20dB {lp} & (0.57) \\ 
+3 & Threshold --20dB {lp} & (0.38) & CenIACCr {bb} &(0.62) & CenIACCr {lp} &(0.33) & CenIACCr {bb} & (0.37) \\ 
+4 & CenIACCr {lp} &(0.37) & MaxIACCe {lp} & (0.61) & Threshold --20dB {lp} & (0.29) & MaxIACCe {lp} & (0.34) \\ 
+5 & Cen-$e^2$ {lp} &(0.34) & CenIACCe {lp} & (0.61) & Cen-$e^2$ {lp}
+& (0.10) & CenIACCr {lp} & (0.33) \\
+\hline\hline
+\end{tabular}
+\end{adjustbox}
+
+\label{tab:topMethods} 
+\end{table*}
+\twocolumngrid
+
 \newpage
@@
 \subsection{Sample Figures, new commands available in this style}
-\vglue-14pt
+
 {\it Note that the publisher determines the final layout, so 
@@
 
-\clearpage
+\begin{figure*}[h]
 \section{Labels in Figline}
@@
 label immediately after the \verb+\fig{}{}{}+ command, inside the
-argument to figline. For example:
+argument to figline.
 
@@
 
-\begin{figure*}
 \baselineskip=12pt
 \begin{verbatim}
-\figline{\boxedfig{figsamp.jpg}{2in}{(a)}\label{boxedfigLetter}}
-\figline{\leftfig{figsamp.jpg}{2in}{(b)}
-\rightfig{figsamp.jpg}{2in}{(c)}\label{rightfigLetter}}
-\figline{\rotatefig{90}{figsamp.jpg}{2in}{(d)}
-\rotatefig{180}{figsamp.jpg}{2in}{(e)}\label{rotatefigLetter}}
+\figline{\boxedfig{figsamp.jpg}{1in}{(a)}\label{boxedfigLetter}}
+\figline{\leftfig{figsamp.jpg}{1in}{(b)}
+\rightfig{figsamp.jpg}{1in}{(c)}\label{rightfigLetter}}
+\figline{\rotatefig{90}{figsamp.jpg}{1in}{(d)}
+\rotatefig{180}{figsamp.jpg}{1in}{(e)}\label{rotatefigLetter}}
 \end{verbatim}
 
-\figline{\boxedfig{figsamp.jpg}{2in}{(a)}\label{boxedfigLetter}}
-\figline{\leftfig{figsamp.jpg}{2in}{(b)}\rightfig{figsamp.jpg}{2in}{(c)}\label{rightfigLetter}}
-\figline{\rotatefig{90}{figsamp.jpg}{2in}{(d)}\rotatefig{180}{figsamp.jpg}{2in}{(e)}\label{rotatefigLetter}}
+\figline{\boxedfig{figsamp.jpg}{1.5in}{(a)}\label{boxedfigLetter}}
+\figline{\leftfig{figsamp.jpg}{1.5in}{(b)}\rightfig{figsamp.jpg}{1.5in}{(c)}\label{rightfigLetter}}
+\figline{\rotatefig{90}{figsamp.jpg}{1.5in}{(d)}\rotatefig{180}{figsamp.jpg}{1.5in}{(e)}\label{rotatefigLetter}}
 
@@
 \begin{verbatim}
-\figline{\fig{figsamp.jpg}{.2\textwidth}{(a)}
-\fig{figsamp.jpg}{.2\textwidth}{(b)}
-\fig{figsamp.jpg}{.2\textwidth}{(c)}
+\figline{\fig{figsamp.jpg}{.2\textwidth}{(A)}
+\fig{figsamp.jpg}{.2\textwidth}{(B)}
+\fig{figsamp.jpg}{.2\textwidth}{(C)}
 \narrowcaption{.25\textwidth}{Caption for three illustrations. 
@@
 \end{verbatim}
-\figline{\fig{figsamp.jpg}{.2\textwidth}{(a)}
-\fig{figsamp.jpg}{.2\textwidth}{(b)}
-\fig{figsamp.jpg}{.2\textwidth}{(c)}
+\figline{\fig{figsamp.jpg}{.2\textwidth}{(A)}
+\fig{figsamp.jpg}{.2\textwidth}{(B)}
+\fig{figsamp.jpg}{.2\textwidth}{(C)}
 \narrowcaption{.25\textwidth}{Caption for three illustrations. 
@@
 \end{figure*}
-
 \clearpage
@@
 \figcolumn{
-\fig{figsamp.jpg}{.2\textwidth}{(a)}
-\fig{figsamp.jpg}{.2\textwidth}{(b)}
-\fig{figsamp.jpg}{.2\textwidth}{(c)}
+\fig{figsamp.jpg}{.2\textwidth}{(A)}
+\fig{figsamp.jpg}{.2\textwidth}{(B)}
+\fig{figsamp.jpg}{.2\textwidth}{(C)}
 }
@@
 \begin{nofloatfigure}
-\centering
-\includegraphics[width=2in]{figsamp}
-\caption{\label{nofloatcaption}
-Showing how you can have a caption that
-continues across pages or columns.\\
-This is a caption in a no float figure.
-It is designed to continue across columns
-or pages if it is particularly long.
-This is a caption that will continue
-across pages if necessary.
-This is a caption that will continue
-across pages if necessary.
-This is a caption that will continue
-across pages if necessary.
-This is a caption that will continue
-across pages if necessary.
-This is a caption that will continue
-across pages if necessary.
-This is a caption that will continue
-across pages if necessary.
-This is a caption that will continue
-across pages if necessary.
-This is a caption that will continue
-across pages if necessary.
-This is a caption that will continue
-across pages if necessary.
-This is a caption that will continue
-across pages if necessary.
+%% \reprintcolumnwidth is the same in preprint and reprint for
+%% ease of use for authors:
+\includegraphics[width=\reprintcolumnwidth]{figsamp.jpg}
+\caption{\label{fig:FIG1}
+Here is a sample of a long caption that will continue
+over columns or pages because it is used in 
+{\tt\string\begin\string{nofloatfigure\string}...\string\end\string{nofloatfigure\string}}
+We want to test a very long caption, so here it is.
+We want to test a very long caption, so here it is.
+We want to test a very long caption, so here it is.
+We want to test a very long caption, so here it is.
+We want to test a very long caption, so here it is.
+We want to test a very long caption, so here it is.
+We want to test a very long caption, so here it is.
+We want to test a very long caption, so here it is.
+We want to test a very long caption, so here it is.
+We want to test a very long caption, so here it is.
 }
@@
 
-\newpage
 \section{Algorithm Examples}
@@
  (\verb+%\usepackage{algpseudocode}+) and type in the
-package name that you'd like to use (but please check that the package is compatible with Editorial Manager; see JASA-EL-TeXGuide.pdf).
+package name that you'd like to use.
 
@@
 
-\newpage
 \subsection{Example of multimedia entry}
@@
 \begin{figure}[ht]
-\includegraphics[width=1in]{figsamp.jpg}
+\includegraphics[width=2.5in]{figsamp.jpg}
 \caption{Figure in an appendix.}
@@
 \end{ruledtabular}
-\vskip12pt
 \end{table}
 
-\subsection{Footnotes}
+\section{Footnotes}
 The contents of the footnotes will appear at the beginning of the
@@
 Numbered\-Refs option:\\
-\verb+\documentclass[preprint,NumberedRefs]{JASA}+\\
+\verb+\documentclass[reprint,NumberedRefs]{JASA}+\\
 and BibTeX has been used.
@@
 Authors are highly  recommended to use BibTeX to produce their
-bibliographies. The results will be predictable and even if
-it might take some time to get comfortable with  using BibTeX,
-in the long run it will save you endless aggravation.
+bibliographies. The results will be predictable
+ even if
+it might take some time to get comfortable with  using BibTeX.
 
+%in the long run it will save you endless aggravation.
 A resource for making your bibliography entries
@@
 and the optional style, Numbered\-Refs, which you would call using\\
-{\verb+\documentclass[preprint,NumberedRefs]{JASA}+ }
+{\verb+\documentclass[reprint,NumberedRefs]{JASA}+ }
 
-\verb+\citep{}+ should normally be used rather than \verb+\cite{}+.
+Every \verb+\cite+ will produce a citation and an entry in the
+bibliography and every cite must have a matching entry in the bib
+database file.
 
-You can also use \verb+\citet{}+ if it is more grammatically correct
-to have only the year in parens (note: this is used with author-year style references).
-
-\begin{verbatim}
-\citep{bibitemName} = (bibitemName, year)
-or
-\citet{bibitemName} = bibitemName (year)
-\end{verbatim}
-
+\verb+\citep{}+ should be used rather than \verb+\cite{}+
 Note that the citations are hyperlinked to their entries in the
@@
 
-Every \verb+\citep+  or \verb+\citet{}+ will produce a citation and an entry in the
-bibliography. Every citation must have a matching entry in the
-bibliography
-database file (\verb+\filename.bib+).
-
-Make your bibliography by doing: pdflatex filename,  bibtex filename,
+Make your bibliography by calling: pdflatex filename,  bibtex filename,
 pdflatex filename, pdflatex filename.
-
-{\bfseries\itshape
-Compare the results you get with\\
-{\verb+\documentclass[preprint]{JASA}+ }\\
-vs.\\
-{\verb+\documentclass[preprint,NumberedRefs]{JASA}+ }
-}
 

### 2020-11-05 modem-paper/main.tex
initial template breakdown for modem paper
fixing makefile to include bibliography, clean up main.tex

--- 2020-11-03 jasa-latex-template-v2019/preprintsample.tex
@@
-%	JASA LaTeX Sample File, Preprint Sample
-%
-%  Beginner Latex users should refer to their favorite online documentation
-%  here is one from the TeX Users Group 
-%	https://www.tug.org/twg/mactex/tutorials/ltxprimer-1.0.pdf
-%
-%  Useful FAQ from  https://journals.aps.org/revtex/revtex-faq
-%
+\title[JASA/Sample JASA Article]{Sample JASA Article}
+\author{EeShan Bhatt}
+\author{Oscar Viquez}
+\author{Henrik Schmidt}
+\affiliation{Department of Mechanical Engineering,  Massachusetts Institute of Technology, Cambridge, MA 02139, USA}
 
-%%%%%%% For Preprint
-%% For manuscript, 12pt, one column style
+% \author{Author Four}
+% \email{author.four@university.edu}
+% \affiliation{Department2,  University2, City, State ZipCode, Country}
 
-\documentclass[preprint]{JASA}
+% \author{Author Five}
+% \altaffiliation{Also at: Department, University, City, State ZipCode, Country.}
+% \affiliation{Department3,  University3, City, State ZipCode, Country}
 
-%%%%% Preprint Options %%%%%
-%% The track changes option allows you to mark changes
-%% and will produce a list of changes, their line number
-%% and page number at the end of the article.
- %\documentclass[preprint,trackchanges]{JASA}
-
-%% NumberedRefs is used for numbered bibliography and citations.
-%% Default is Author-Year style.
-% \documentclass[preprint,NumberedRefs]{JASA}
-
-%%%%%%% For Reprint
-%% For appearance of finished article; 2 columns, 10 pt fonts
-
-% \documentclass[reprint]{JASA}
-
-%%%%% Reprint Options %%%%%
-
-%% For testing to see if author has exceeded page length request, use 12pt option
-% \documentclass[reprint,12pt]{JASA}
-
-%% NumberedRefs is used for numbered bibliography and citations.
-%% Default is Author-Year style.
-% \documentclass[reprint,NumberedRefs]{JASA}
-
-%% TurnOnLineNumbers
-%% Make lines be numbered in reprint style:
-% \documentclass[reprint,TurnOnLineNumbers]{JASA}
-
-%% Optional algorithm package. You can comment this out and
-%% include another package if you prefer another way to make
-%% algorithms examples (but please check that the package is compatible with 
-%%Editorial Manager; see JASA-EL-TeXGuide.pdf).
-\usepackage{algpseudocode}
-
-\begin{document}
-
-\title[JASA/Sample JASA Article]{Sample JASA Article}
-\author{Author One}
-\author{Author Two}
-\author{Author Three}
-\affiliation{Department1,  University1, City, State ZipCode, Country}
-
-\author{Author Four}
-\email{author.four@university.edu}
-\affiliation{Department2,  University2, City, State ZipCode, Country}
-
-\author{Author Five}			
-\altaffiliation{Also at: Department, University, City, State ZipCode, Country.}
-\affiliation{Department3,  University3, City, State ZipCode, Country}
-
-\preprint{Author, JASA}		%  if you want want this message to appear in upper right corner of title page
+\preprint{Bhatt, JASA}   %  if you want want this message to appear in upper right corner of title page
 
@@
 prepared for submission to the Journal of the Acoustical Society of America.
-
-See JASA-TeXGuide.pdf, which is part of this package, for extensive
-documentation on using commands for JASA.
-
-You can compare the .tex version of this file with the resulting .pdf
-version to give you an idea of what  commands are available and how
-they work. At the top of the .tex file you'll find a listing of the
-documentclass options, and an explanation of their results.
-Some additional suggestions are included in the body of this
-manuscript.
-
-  Beginner Latex users should refer to their favorite online documentation. A 
-  useful place to start is the primer from the TeX Users Group 
-  \url{https://www.tug.org/twg/mactex/tutorials/ltxprimer-1.0.pdf}
-
-EXAMPLE TEXT: This is example text.
-This is example text.
-This is example text.
-This is example text.
-This is example text.
-This is example text.
-This is example text.
-This is example text.
-This is example text.
-This is example text.
-This is example text.
-This is example text.
-This is example text.
-This is example text.
-This is example text.
-This is example text.
 
@@
 
-\section{Inline and display math samples\label{sec:3}}
-
-\subsection{\label{subsec:3:3} Math and equations $\alpha\beta\Delta\Gamma$}
-Inline math may be typeset using the \verb+$+ delimiters. Bold math
-symbols may be achieved using the \verb+bm+ package and the
-\verb+\bm{#1}+ command it supplies. For instance, a bold $\alpha$ can
-be typeset as \verb+$\bm{\alpha}$+ giving $\bm{\alpha}$. Fraktur and
-Blackboard (or open face or double struck) characters should be
-typeset using the \verb+\mathfrak{#1}+ and \verb+\mathbb{#1}+ commands
-respectively. Both are supplied by the \texttt{amssymb} package which
-is called in JASA, so you don't need an \verb+\usepackage{amssymb}+
-command in your .tex file. For
-example, \verb+$\mathbb{R}$+ gives $\mathbb{R}$ and
-\verb+$\mathfrak{G}$+ gives $\mathfrak{G}$.
-
-In \LaTeX\ there are many different ways to display equations; a
-few preferred ways are noted below. Displayed math will center by
-default.
-
-Below we have numbered single-line equations; this is the most common
-type of equation.
-\begin{eqnarray}
-\chi_+(p){\bf [}2|{\bf p}|(|{\bf p}|+p_z){\bf ]}^{-1/2}
-\left(
-\begin{array}{c}
-|{\bf p}|+p_z\\
-px+ip_y
-\end{array}\right)\;,
-\\
-\left\{%
- 1234567890abc123\alpha\beta\gamma\delta1234556\alpha\beta
- \frac{1\sum^{a}_{b}}{A^2}%
-\right\}%
-\label{eq:one}.
-\end{eqnarray}
-Note the open one in Eq.~(\ref{eq:one}).
-
-Not all numbered equations will fit within a narrow column this
-way. The equation number will move down automatically if it cannot fit
-on the same line with a one-line equation.
-\begin{equation}
-\chi_+(p){\bf [}2|{\bf p}|(|{\bf p}|+p_z){\bf ]}^{-1/2}
-\alpha\beta\gamma\delta123455678\alpha\beta\Gamma\Delta
- \frac{1\sum^{a}_{b}}{A^2}1234%
-\end{equation}
-% No blank lines after \end{equation}
-When the \verb+\label{#1}+ command is used [ie. input for
-Eq.~(\ref{eq:one})], the equation can be referred to in text without
-knowing the equation number that \TeX\ will assign to it. Just
-use \verb+\ref{#1}+, where \verb+#1+ is the same name that used in
-the \verb+\label{#1}+ command.
-
-Unnumbered single-line equations can be typeset
-using the \verb+\[+, \verb+\]+ format:
-\[g^+g^+ \rightarrow g^+g^+g^+g^+ \dots ~,~~q^+q^+\rightarrow
-q^+g^+g^+ \dots ~. \]
-
-Note the equations can be lettered with the
-subequations environment:
-\begin{subequations}
-\begin{eqnarray}
-A&=&mc, \label{eqa}
-\\
-B&=&mc^2, \label{eqb}
-\\
-C&\agt& mc^3. \label{eqc}
-\end{eqnarray}
-\end{subequations}
-Referenced: Eqs.~(\ref{eqa}), (\ref{eqb}), and (\ref{eqc}).
-
-\section{Floats, Figures and Tables}
-
- Figures and tables are typically ``floats'' which means that their
-final position is determined by \LaTeX\ while the document is being
-typeset. \LaTeX\ isn't always successful in placing floats
-optimally.  Use the figure* environment to get a wide figure that spans the 
-page in a two-column layout.
-
-\subsection{\label{subsec:3:2} Tables}
-Tables generally should be surrounded with
-\verb+\begin{ruledtabular}...\end{ruledtabular}+\\
-This will guarantee that they are the width of the
-page or column, and have two ruled lines at the top
-and bottom of the table.
-
-\verb+[ht]+ in the code below instructs \LaTeX\ to place the table
-where it appears in type, if it will fit on the page;
-otherwise put it on the top of the next page.
-
-Footnotes in a table are labeled a, b, c,
- etc.  They can
-be  specified  by  using  the  \LaTeX\
-\verb+\footnotemark[]+
-and
-\verb+\footnotetext[]+ commands.
-The  footnotes  for  a  table  are  typeset  at  the
-bottom  of  the  table,  rather  than  at  the  bottom  of  the
-page or at the end of the references.  The arguments for
-\verb+\footnotemark[]+
-and
-\verb+\footnotetext[]+
-should be numbers
-1, 2, \ldots\  The journal style will convert these to letters.
-
-This system allows
-multiple entries to refer to the same
-footnote.
-
-\begin{table}[ht]
-\caption{\label{tab:table1}A table with more columns still fits
-properly in a column. Note that several entries share the same
-footnote. Inspect the \LaTeX\ input for this table to see
-exactly how it is done.}
-
-\begin{ruledtabular}
-\begin{tabular}{cccccccc}
- &$r_c$ (\AA)\footnotemark[1]&$r_0$ (\AA)&$\kappa r_0$&
- &$r_c$ (\AA) &$r_0$ (\AA)&$\kappa r_0$\\
-\hline
-Cu& 0.800 & 14.10 & 2.550 &Sn\footnotemark[1]
-& 0.680 & 1.870 & 3.700 \\
-Ag& 0.990 & 15.90 & 2.710 &Pb\footnotemark[2]
-& 0.450 & 1.930 & 3.760 \\
-%Au& 1.150 & 15.90 & 2.710 &Ca\footnotemark[3]
-%& 0.750 & 2.170 & 3.560 
-\end{tabular}
-\end{ruledtabular}
-\footnotetext[1]{Here's the first.}
-\footnotetext[2]{Here's the second.}
-%\footnotetext[3]{Here's the third.}
-\end{table}
-\clearpage
-\subsection{Plain Tables: When NOT to use `ruledtabular'}
-There are a number of cases when `ruledtablar' should not
-be used: basically for any table using complex content or 
-commands.
-
-\subsubsection{Using {\tt\string\multicolumn}}
-When you'd like to use the multicolumn command in your
-table, you'll find that `ruledtabular' will cause bad
-formatting. In that case, Don't Use Ruledtabular, and
-instead put in \verb+\hline\hline+ at the top and bottom of
-the table.
-
-\begin{table}[ht]
-\caption{A table made without `ruledtabular' needs to have two hlines
-added to the top and bottom of the table.}
-\vskip3pt
-\begin{tabular}{cccccccc}
-\hline\hline
- &$r_c$ (\AA)\footnotemark[1]&$r_0$ (\AA)&$\kappa r_0$&
- &$r_c$ (\AA) &$r_0$ (\AA)&$\kappa r_0$\\
-\hline
-Cu& 0.800 & 14.10 & 2.550 &Sn\footnotemark[1]
-& 0.680 & 1.870 & 3.700 \\
-Ag& 0.990 & 15.90 & 2.710 &Pb\footnotemark[2]
-& 0.450 & 1.930 & 3.760 \\
-Au& 1.150 & 15.90 & 2.710 &Ca\footnotemark[3]
-& 0.750 & 2.170 & 3.560\\
-\hline\hline
-\end{tabular}
-\footnotetext[1]{This is the first table note.}
-\footnotetext[2]{This is the second table note.}
-\footnotetext[3]{This is the third table note.}
-\end{table}
-
-\newpage
-
-\vskip12pt
-\vskip1sp
-\subsubsection{Using the {\tt\string\adjustbox\string{\string}
-(tabular)\string\end\string{adjustbox\string}}
-command}
-There may be times when the table is too wide, or you want to have
-the table be the width of the page, whether or not it appears in
-preprint or reprint version of JASA. In this case you can use
-\verb+\begin{adjustbox}{<width>} (tabular) \end{adjustbox}+.
-(`adjustbox' will NOT work with `ruledtabular')
-
-You can set a maximum width with \\
-\verb+\begin{adjustbox}{max width=\textwidth}(tabular)\end{adjustbox}+\\  
-in which case the table in the reprint version might be less than
-the full text width;
-
-Or you can set the exact width you'd like with\\
-\verb+\begin{adjustbox}{width=\textwidth}(tabular)\end{adjustbox}+\\
-in which case the table will be the full width of the page in
-either preprint or reprint.
-
-This way you can make a table that will fit in the correct width
-whether you are using the preprint or reprint option.
-
-\begin{table*}[b]
-\caption{Top~5 rated $\widehat{\textrm{ITD}}$ estimation methods
-according to the sum and product metric criteria for $\pm0.5$~JND and
-$\pm1$~JND tolerance thresholds (normalized scores).}
-
-\begin{adjustbox}{width=\textwidth}
-\begin{tabular}{c| l l| l l| l l| l l}
-\hline\hline
-Rank \#& \multicolumn{2}{c}{sum criteria [$\pm0.5$~JND]} &
-\multicolumn{2}{|c}{sum criteria [$\pm1$~JND]} &
-\multicolumn{2}{|c}{product criteria [$\pm0.5$~JND]} & \multicolumn{2}{|c}{product criteria [$\pm1$~JND]}\\ 
-\hline
-1 & Threshold --30dB {lp} & (0.43) & Threshold --30dB {lp} & (0.71) & Threshold --30dB {lp} & (1.00) & Threshold --30dB {lp} & (1.00) \\ 
-2 & MaxIACCe {lp} & (0.39) & Threshold --20dB {lp} & (0.66) & MaxIACCe {lp} & (0.39) & Threshold --20dB {lp} & (0.57) \\ 
-3 & Threshold --20dB {lp} & (0.38) & CenIACCr {bb} &(0.62) & CenIACCr {lp} &(0.33) & CenIACCr {bb} & (0.37) \\ 
-4 & CenIACCr {lp} &(0.37) & MaxIACCe {lp} & (0.61) & Threshold --20dB {lp} & (0.29) & MaxIACCe {lp} & (0.34) \\ 
-5 & Cen-$e^2$ {lp} &(0.34) & CenIACCe {lp} & (0.61) & Cen-$e^2$ {lp}
-& (0.10) & CenIACCr {lp} & (0.33) \\
-\hline\hline
-\end{tabular}
-\end{adjustbox}
-
-\label{tab:topMethods} 
-\end{table*}
-
-\clearpage
-
-\subsection{Using dcolumn}
-The call to \verb+\usepackage{dcolumn}+ is included in JASA.cls so you
-don't need to add it explicitly.
-\href{http://anorien.csc.warwick.ac.uk/mirrors/CTAN/macros/latex/required/tools/dcolumn.pdf}
-{\url{http://anorien.csc.warwick.ac.uk/mirrors/CTAN/macros/latex/required/tools/dcolumn.pdf}}
-will give you detailed information.
-A gentler introduction may be found in this informative and well
-illustrated
-article: \href{https://www.tug.org/pracjourn/2007-1/mori/mori.pdf}
-{\url{https://www.tug.org/pracjourn/2007-1/mori/mori.pdf}}, starting on page
-20. (You may want to look at more examples in this quite comprehensive
-article on making tables in \LaTeX.)
-
-\begin{quote}
-``If we do not want to break the fractional and the integral part in two columns,
-the dcolumn package provides a new type of column\\
-\verb+D{sep -in}{sep -out}{ before.after}+\\
-The first argument \verb+{sep-in}+ is the symbol used in the
-.tex document to separate
-the integral and the fractional part (usually the decimal point . or the decimal
-comma ,), the second argument \verb+{sep-out}+
-is the symbol that we want in the
-output, the third is the number of digits on the left (before) and on the right
-(after) this symbol. The numbers are aligned to the decimal point and, in case
-that the third argument is negative, the decimal point is aligned to the center of
-the column. If the columns have a heading, it must be inserted into
-the command \verb+\multicolumn{1}{c}{...}+
-\end{quote}
-\newpage
-An example using dcolumn:
-\begin{verbatim}
-{\hsize= 2in
-\begin{ruledtabular}
-\begin{tabular}{cD {,}{.}{5.4}}
-Expression           & \multicolumn {1}{c}{ Value }\\
-\hline
-$\pi$                  &      3,1416                 \\
-$\pi^{\pi}$           &     36,46                    \\
-$\pi^{\pi^{\pi}}$    & 80662,7                      \\
-\end{tabular}
-\end{ruledtabular}
-}
-\end{verbatim}
-\vskip12pt
-{\hsize= 2in
-\begin{ruledtabular}
-\begin{tabular}{cD {,}{.}{5.4}}
-Expression           & \multicolumn {1}{c}{ Value }\\
-\hline
-$\pi$                  &      3,1416                 \\
-$\pi^{\pi}$           &     36,46                    \\
-$\pi^{\pi^{\pi}}$    & 80662,7                      \\
-\end{tabular}
-\end{ruledtabular}
-}
-
-\newpage
-
-\begin{figure*}[p]
-\parskip=6pt
-\baselineskip=12pt
-
-\subsection{Sample Figures, new commands available in this style}
-\vglue-14pt
-{\it Note that the publisher determines the final layout, so 
-your choice of figure alignment may not be reflected in the published
-article.}
-
-\noindent
-\verb+\figline{}+ will center one or more figures on one line.
-
-\noindent
-\verb+\fig{<name of file>}{<width>}{<letter to put underneath>}+
-
-\noindent
-\verb+\leftfig{<name of file>}{<width>}{<letter to put underneath>}+
-
-\noindent
-\verb+\rightfig{<name of file>}{<width>}{<letter to put underneath>}+
-
-\noindent
-\verb+\boxedfig{<name of file>}{<width>}{<letter to put underneath>}+
-
-\noindent
-\verb+\rotatefig{<degrees of rotation>}{<name of file>}{<width>}+\\
-\verb+{<letter to put underneath>}+
-
-The following illustrations show these commands in use.
-
-\begin{verbatim}
-\figline{\fig{figsamp.jpg}{4cm}{(a)}
-\fig{figsamp.jpg}{4cm}{(b)}}
-\figline{\fig{figsamp.jpg}{4cm}{(c)}
-\fig{figsamp.jpg}{4cm}{(d)}}
-\figline{\fig{figsamp.jpg}{4cm}{(e)}}
-\end{verbatim}
-
-\figline{\fig{figsamp.jpg}{4cm}{(a)}
-\fig{figsamp.jpg}{4cm}{(b)}}
-\figline{\fig{figsamp.jpg}{4cm}{(c)}
-\fig{figsamp.jpg}{4cm}{(d)}}
-\figline{\fig{figsamp.jpg}{4cm}{(e)}}
-
-\caption{ \label{fig:pressure_field} Multiple images on one figure example (a) image 1, (b-f) ($\rho$=1000 kg/m$^3$) and speed of sound ($c$=1500 m/s).}
-
-\end{figure*}
-
-\begin{figure*}
-\baselineskip=12pt
-\begin{verbatim}
-\figline{\boxedfig{figsamp.jpg}{2in}{(a)}}
-\figline{\leftfig{figsamp.jpg}{2in}{(b)}\rightfig{figsamp.jpg}{2in}{(c)}}
-\figline{\rotatefig{90}{figsamp.jpg}{2in}{(d)}\rotatefig{180}{figsamp.jpg}{2in}{(e)}}
-\end{verbatim}
-
-\figline{\boxedfig{figsamp.jpg}{2in}{(a)}}
-
-\figline{\leftfig{figsamp.jpg}{2in}{(b)}\rightfig{figsamp.jpg}{2in}{(c)}}
-
-\figline{\rotatefig{90}{figsamp.jpg}{2in}{(d)}
-\rotatefig{180}{figsamp.jpg}{2in}{(e)}
-}
-
-\caption{More figure examples: (a) boxedfig, 
-(b)leftfig; (c)right fig; (d) rotatefig 90 degrees;
-(e) rotatefig 180 degrees. }
-
-\end{figure*}
-
-\clearpage
-\section{Labels in Figline}
-We can label and
-reference separate parts of the figure when using figline.
-The reference will give the illustration letter as well as the figure
-number.
-
-To label figures used in \verb+\figline{}+ type in your
-label immediately after the \verb+\fig{}{}{}+ command, inside the
-argument to figline. For example:
-
-\begin{verbatim}
-\figline{\fig{<name of file>}{<width>}{<letter to put underneath>}\label{<labelname>}}
-\end{verbatim}
-
-The same placement should be used for all the kinds of fig
-environments used in \verb+\figline{}+:
-\begin{verbatim}
-\fig{}{}{}\label{}, \leftfig{}{}{}\label{}, \rightfig{}{}{}\label{}, 
-\boxedfig{}{}{}\label{}, \rotatefig{}{}{}\label{},  \narrowcaption{}{}\label{}.
-\end{verbatim}
-
-\begin{figure*}
-\baselineskip=12pt
-\begin{verbatim}
-\figline{\boxedfig{figsamp.jpg}{2in}{(a)}\label{boxedfigLetter}}
-\figline{\leftfig{figsamp.jpg}{2in}{(b)}
-\rightfig{figsamp.jpg}{2in}{(c)}\label{rightfigLetter}}
-\figline{\rotatefig{90}{figsamp.jpg}{2in}{(d)}
-\rotatefig{180}{figsamp.jpg}{2in}{(e)}\label{rotatefigLetter}}
-\end{verbatim}
-
-\figline{\boxedfig{figsamp.jpg}{2in}{(a)}\label{boxedfigLetter}}
-\figline{\leftfig{figsamp.jpg}{2in}{(b)}\rightfig{figsamp.jpg}{2in}{(c)}\label{rightfigLetter}}
-\figline{\rotatefig{90}{figsamp.jpg}{2in}{(d)}\rotatefig{180}{figsamp.jpg}{2in}{(e)}\label{rotatefigLetter}}
-
-\caption{More figure examples, showing how to enter
-{\tt\string\label\string{\string}} command.}
-
-\verb+References:\ref{boxedfigLetter},\ref{rightfigLetter},\ref{rotatefigLetter}+
-
-which produces\\
-
-References: \ref{boxedfigLetter},\ref{rightfigLetter},\ref{rotatefigLetter}
-
-\end{figure*}
-
-\begin{figure*}
-\baselineskip=12pt
-\begin{verbatim}
-\sidebysidefigures{figsamp.jpg}{Describing the first
-illustration.}/{figsamp.jpg}{Describing the second illustration.}
-\end{verbatim}
-
-\sidebysidefigures{figsamp.jpg}{Describing the first
-illustration.}/{figsamp.jpg}{Describing the second illustration.}
-\end{figure*}
-
-\begin{figure*}
-\baselineskip=12pt
-\begin{verbatim}
-\figline{
-\fig{figsamp.jpg}{.7\textwidth}{}
-\narrowcaption{.2\textwidth}{Here is a narrow caption.}
-}
-\end{verbatim}
-\figline{\fig{figsamp.jpg}{.7\textwidth}{} \narrowcaption{.2\textwidth}{Here is a narrow
-caption.}}
-\end{figure*}
-
-\begin{figure*}
-\baselineskip=12pt
-\begin{verbatim}
-\figline{\fig{figsamp.jpg}{.2\textwidth}{(a)}
-\fig{figsamp.jpg}{.2\textwidth}{(b)}
-\fig{figsamp.jpg}{.2\textwidth}{(c)}
-\narrowcaption{.25\textwidth}{Caption for three illustrations. 
-The caption may produce many lines, but only one paragraph.
-}}
-\end{verbatim}
-\figline{\fig{figsamp.jpg}{.2\textwidth}{(a)}
-\fig{figsamp.jpg}{.2\textwidth}{(b)}
-\fig{figsamp.jpg}{.2\textwidth}{(c)}
-\narrowcaption{.25\textwidth}{Caption for three illustrations. 
-The caption may produce many lines, but only one paragraph.
-}}
-\end{figure*}
-
-\begin{figure*}
-\baselineskip=12pt
-\begin{verbatim}
-\vtop to 0pt{\vss
-\figline{\leftfig{figsamp.jpg}{.2\textwidth}{(a)}
-\leftfig{figsamp.jpg}{.2\textwidth}{(b)}\hfill}
-\figline{\leftfig{figsamp.jpg}{.2\textwidth}{(c)}
-\leftfig{figsamp.jpg}{.2\textwidth}{(d)}\hfill}
-\vss}
-\narrowcaption{.25\textwidth}{Here is a narrow caption that will can be
-positioned to the right of four illustrations.
-You cannot have more than one paragraph of text in a caption.
-You cannot have more than one paragraph of text in a caption.
-You cannot have more than one paragraph of text in a caption.
-You cannot have more than one paragraph of text in a caption.
-}
-\vskip36pt % to give more space above the footline
-\end{verbatim}
-\vskip1.3in
-\vtop to 0pt{\vss
-\figline{\leftfig{figsamp.jpg}{.2\textwidth}{(a)}
-\leftfig{figsamp.jpg}{.2\textwidth}{(b)}\hfill}
-\figline{\leftfig{figsamp.jpg}{.2\textwidth}{(c)}
-\leftfig{figsamp.jpg}{.2\textwidth}{(d)}\label{test}\hfill}
-\vss}
-\narrowcaption{.25\textwidth}{Here is a narrow caption that will can be
-positioned to the right of four illustrations.
-You cannot have more than one paragraph of text in a caption.
-You cannot have more than one paragraph of text in a caption.
-You cannot have more than one paragraph of text in a caption.
-You cannot have more than one paragraph of text in a caption.
-}
-\vskip36pt
-\end{figure*}
-
-\clearpage
-
-\begin{figure}[h]
-\baselineskip=12pt
-\begin{verbatim}
-\figcolumn{
-\fig{figsamp.jpg}{.2\textwidth}{(a)}
-\fig{figsamp.jpg}{.2\textwidth}{(b)}
-\fig{figsamp.jpg}{.2\textwidth}{(c)}
-}
-\end{verbatim}
-
-\figcolumn{
-\fig{figsamp.jpg}{.2\textwidth}{(a)}
-\fig{figsamp.jpg}{.2\textwidth}{(b)}
-\fig{figsamp.jpg}{.2\textwidth}{(c)}
-}
-
-\caption{Here are some stacking figures in a single column.
-The {\tt\string\figcolumn\string{\string}} command works equally well in single
-or double column figures.}
-\end{figure}
-
-\begin{nofloatfigure}
-\centering
-\includegraphics[width=2in]{figsamp}
-\caption{\label{nofloatcaption}
-Showing how you can have a caption that
-continues across pages or columns.\\
-This is a caption in a no float figure.
-It is designed to continue across columns
-or pages if it is particularly long.
-This is a caption that will continue
-across pages if necessary.
-This is a caption that will continue
-across pages if necessary.
-This is a caption that will continue
-across pages if necessary.
-This is a caption that will continue
-across pages if necessary.
-This is a caption that will continue
-across pages if necessary.
-This is a caption that will continue
-across pages if necessary.
-This is a caption that will continue
-across pages if necessary.
-This is a caption that will continue
-across pages if necessary.
-This is a caption that will continue
-across pages if necessary.
-This is a caption that will continue
-across pages if necessary.
-}
-\end{nofloatfigure}
-
-\newpage
-\section{Algorithm Examples}
-This example uses \verb+\usepackage{algpseudocode}+ as you see above. If
-you would rather use another algorithm package, please comment out
-this package
- (\verb+%\usepackage{algpseudocode}+) and type in the
-package name that you'd like to use (but please check that the package is compatible with Editorial Manager; see JASA-EL-TeXGuide.pdf).
-
-For other algorithm packages, see
-\url{https://en.wikibooks.org/wiki/LaTeX/Algorithms}
-
-Here is an example of algorithmic:
-
-\begin{algorithmic}
-\If {$i\geq maxval$}
-    \State $i\gets 0$
-\Else
-    \If {$i+k\leq maxval$}
-        \State $i\gets i+k$
-    \EndIf
-\EndIf
-\end{algorithmic}
-
-Documentation for the package is found at 
-\url{http://mirrors.rit.edu/CTAN/macros/latex/contrib/algorithm2e/doc/algorithm2e.pdf}
-which shows many other examples and options.
-
-\newpage
-\subsection{Example of multimedia entry}
-Please note that this is for multimedia intended to appear inline
-within the published article.
-
-Here is what a multimedia entry will look like:
-\multimedia{http://dx.doi.org/10.1121/1.4947423.1}{Corresponding pulse-compressed echo envelopes
-and video recordings from a fluttering luna moth.
-Echoes from the wings and body of the moth generally dominate the
-acoustic returns, which vary greatly over consecutive ensonifications
-across the wingbeat cycle. File of type ``mp4'' (15.3
-MB)}\label{mmtest1}
-
-Here we try cross referencing the multimedia entry: The multimedia
-above is Mm.~\ref{mmtest1}.
-
-\subsection{Supplementary Material}
-ASA
-prefers that authors to submit related/relevant article files as
-supplementary material with their submission.
-
-\subsection{Supplementary material for publication}
-Any archival supplemental materials to be published with the
-manuscript (eg., supplementary figures) should be cited in-text and a footnote provided.
-
-An example of reference to supplementary material:
-
-The sound files and videos for this and other figures
-are included as supplementary materials\footnote{See
-Supplementary materials at [URL will be inserted by AIP]
-for [give a brief description of the material].}.
-
-The contents of the footnote above will appear at the beginning of the
-bibliography made with BibTeX when the default `author-year' documentclass option is used;
-BibTeX output will have the footnote interleaved with other
-references if the NumberedRefs documentclass option is used.
-
-\subsection{File naming conventions}
-Here are the conventions for naming files:
-
-\begin{itemize}
-\item
-Supplementary Figure or
-	Supplementary Figure or Text files should be named: SuppPub\#.xxx, where ``\#'' is
-	a number and ``xxx" is the file format extension
-	(SuppPub1.docx, SuppPub2.jpg, etc)
-
-\item
-	Supplementary Multimedia files: SuppPubmm\#.xxx, where ``\#'' is a
-	number and ``xxx'' is the file format extension (SuppPubmm1.mp3,
-	SuppPubmm2.gif, etc)
-
-\item
-Multimedia files must be named accordingly: MM\#.xxx, where ``\#'' is the
-number and ``xxx'' is the file format extension (MM1.wav, MM2.avi, etc).
-
-\item
-The only figure formats allowed are the following: 
-.pdf, .ps, .eps, or .jpg. Figure files must be named in this fashion:
-Figure\#.xxx, where ``\#'' is the figure number and ``xxx'' is the file format
-(Figure1.eps, Figure2.jpg, Figure3a.ps, Figure3b.ps, etc).
-
-\end{itemize}
-
-\section{\label{sec:5}Conclusion}
-
-And in conclusion\ldots
-
-\begin{acknowledgments}
-This research was supported by  ...
-\end{acknowledgments}
-
- % -------------------------------------------------------------------------------------------------------------------
- %   Appendix  (optional)
-
-\appendix
-\section{Appendices}
-To start the appendix, use the \verb+\appendix+ command.
-This signals that all following section commands refer to appendixes
-instead of regular sections. Therefore, the \verb+\appendix+ command
-should be used only once---to set up the section commands to act as
-appendices. Thereafter normal section commands are used. The heading
-for a section can be left empty. For example,
-\begin{verbatim}
-\appendix
-\section{}
-\end{verbatim}
-will produce an appendix heading that says ``APPENDIX A'' and
-\begin{verbatim}
-\appendix
-\section{Background}
-\end{verbatim}
-will produce an appendix heading that says ``APPENDIX A: BACKGROUND''
-(note that the colon is set automatically).
-
-If there is only one appendix, then the letter ``A'' should not
-appear. This is suppressed by using the star version of the appendix
-command (\verb+\appendix*+ in the place of \verb+\appendix+).
-\vskip12pt
-\vskip1sp
-\section{A little more on appendices}
-
-Observe that this appendix was started by using
-\begin{verbatim}
-\section{A little more on appendixes}
-\end{verbatim}
-
-Note the equation number in an appendix:
-\begin{equation}
-E=mc^2.
-\end{equation}
-
-\subsection{\label{app:subsec}A subsection in an appendix}
-
-You can use a subsection or subsubsection in an appendix. Note the
-numbering: we are now in Appendix~\ref{app:subsec}.
-
-\subsubsection{\label{app:subsubsec}A subsubsection in an appendix}
-Note the equation numbers in this appendix, produced with the
-subequations environment:
-\begin{subequations}
-\begin{eqnarray}
-E&=&mc, \label{appa}
-\\
-E&=&mc^2, \label{appb}
-\\
-E&\agt& mc^3. \label{appc}
-\end{eqnarray}
-\end{subequations}
-They turn out to be Eqs.~(\ref{appa}), (\ref{appb}), and (\ref{appc}).
-
-\section{Sample appendix figure and table}
-Figure and table numbering are continuous through the article,
-and handled the same as they are in the rest of the article.
-
-\begin{figure}[ht]
-\includegraphics[width=1in]{figsamp.jpg}
-\caption{Figure in an appendix.}
-\end{figure}
-
-\begin{table}[ht]
-\caption{Here is the caption for a table in an appendix.}
-\centering
-\begin{ruledtabular}
-\begin{tabular}{cccc}
-one&two&three&four\\
-\hline
-C&D&E&F\\
-\end{tabular}
-\end{ruledtabular}
-\vskip12pt
-\end{table}
-
-\subsection{Footnotes}
-The contents of the footnotes will appear at the beginning of the
-bibliography when BibTeX produces the .bbl file using the default
-AuthorYear style; interleaved with other references if 
-Numbered\-Refs option:\\
-\verb+\documentclass[preprint,NumberedRefs]{JASA}+\\
-and BibTeX has been used.
-
-This example show where this cite \citep{booksamp1} will appear in the
-bibliography,
-depending on whether we use default author-year style
-or call for the NumberedRefs documentclass option.
-
-Here are some sample footnotes:\footnote{Here is the second footnote.
-It will appear before the beginning of the bibliography in Author-Year
-style (default) or it will be 
- interleaved with other references when using the Numbered\-Refs
- option.}$^,$% (comma between two sequential footnotes)
-\footnote{Here is a third footnote.}
-
-\section{Making the Bibliography Using BibTeX}
-Authors are highly  recommended to use BibTeX to produce their
-bibliographies. The results will be predictable and even if
-it might take some time to get comfortable with  using BibTeX,
-in the long run it will save you endless aggravation.
-
-A resource for making your bibliography entries
-correctly is included in this package: 
-JASA-ReferenceStyles.pdf. You will also find
-the files
-bibsamp1.tex/.pdf and bibsamp2.tex/.pdf
-for examples of output; and sampbib.bib for an example of
-how to make your .bib database entries.
-
-There are two possible bibliography styles: the default, author-year,
-and the optional style, Numbered\-Refs, which you would call using\\
-{\verb+\documentclass[preprint,NumberedRefs]{JASA}+ }
-
-\verb+\citep{}+ should normally be used rather than \verb+\cite{}+.
-
-You can also use \verb+\citet{}+ if it is more grammatically correct
-to have only the year in parens (note: this is used with author-year style references).
-
-\begin{verbatim}
-\citep{bibitemName} = (bibitemName, year)
-or
-\citet{bibitemName} = bibitemName (year)
-\end{verbatim}
-
-Note that the citations are hyperlinked to their entries in the
-bibliography:
-
 Normal journal cite: \citep{joursamp1},
@@
 \bibliography{sampbib}
-\end{document}

### 2020-11-06 modem-paper/main.tex
adding figure scaffolding to paper draft

@@
-\title[JASA/Sample JASA Article]{Sample JASA Article}
+\title[JASA/Sample JASA Article]{Considering vertical sound speed uncertainty in the Beaufort Lens for more realistic range estimation between acoustic modems}
 \author{EeShan Bhatt}
 \author{Oscar Viquez}
+\author{Bradli Howard}
 \author{Henrik Schmidt}
@@
 
-% \author{Author Five}
-% \altaffiliation{Also at: Department, University, City, State ZipCode, Country.}
-% \affiliation{Department3,  University3, City, State ZipCode, Country}
-
 \preprint{Bhatt, JASA}   %  if you want want this message to appear in upper right corner of title page
@@
 \begin{abstract}
+
+{\color{red}[Draft Abstract]}
+
 Put your abstract here. Abstracts are limited to 200 words for
@@
 
-%  End of title page for Preprint option --------------------------------- %
+% =========================================================================== %
+% =========================================================================== %
+\section{\label{sec:1} Introduction}
 
-\section{\label{sec:1} Introduction}
+{\color{red}
+[Intro : EOF Approach, generally]
+
+[Motivation: ICEX (\& Irminger?)]
+}
+
 This sample document demonstrates the use of JASA in manuscripts
@@
 
- % Sample showing how to include figure, this is a floating figure
+% =========================================================================== %
+% =========================================================================== %
+\section{\label{sec:2} Methods}
 
+\subsection{\label{subsec:2:1} Environmental EOFs}
+
+{\color{red}[DETAIL : EOF Approach]}
+
+% ================================================== %
 \begin{figure}[ht]
-%% \reprintcolumnwidth is the same in preprint and reprint for
-%% ease of use for authors:
+\figline{\fig{figsamp.jpg}{\reprintcolumnwidth}{(a)}}
+\figline{\fig{figsamp.jpg}{4cm}{(b)} \fig{figsamp.jpg}{4cm}{(c)}}
+
+% \includegraphics[width=\reprintcolumnwidth]{figsamp.jpg}
+\caption{\label{fig:FIG1}{Extracting EOFs from environmental data: (a) HYCOM and ITP data, (b) EOF shapes and weight distributions.}}
+
+\end{figure}
+% ================================================== %
+
+This section would generally discuss EOFs, the equations that extract them from data collections, and perhaps the "information capacity" w.r.t.
+to the number of depth points vs number of samples.
+Should we also include, in somewhat general terms, the "uniqueness" constraint?
+As in, i.e.: sampled region is too large or too small to adequately capture the features needed for reconstruction.
+
+% =========================================================================== %
+% =========================================================================== %
+\section{\label{sec:3} Results}
+
+\subsection{\label{subsec:3:1} ICEX 2020 conditions}
+{\color{red}[CTD comparison]}
+
+% ================================================== %
+\begin{figure}[ht]
+\includegraphics[width=\reprintcolumnwidth]{figsamp.jpg}
+\caption{\label{fig:FIG2}{CTD casts vs. HYCOM-derived baseline}}
+\end{figure}
+% ================================================== %
+
+% ========================= %
+% ========================= %
+\subsection{\label{subsec:3:2} One-Way Travel Times}
+{\color{red}[OWTT comparison]}
+
+% ================================================== %
+\begin{figure}[ht]
+\includegraphics[width=\reprintcolumnwidth]{figsamp.jpg}
+\caption{\label{fig:FIG3}{Modem placement diagram}}
+\end{figure}
+% ================================================== %
+
+% ================================================== %
+\begin{figure}[ht]
+\figline{\fig{figsamp.jpg}{\reprintcolumnwidth}{(a)}}
+\figline{\fig{figsamp.jpg}{4cm}{(b)} \fig{figsamp.jpg}{4cm}{(c)}}
+% \includegraphics[width=\reprintcolumnwidth]{figsamp.jpg}
+\caption{\label{fig:FIG4}{(a) OWTT histogram, (b) "..." OF RANGE "...".}}
+\end{figure}
+% ================================================== %
+
+% =========================================================================== %
+% =========================================================================== %
+\section{\label{sec:4} Discussion}
+
+\subsection{\label{subsec:4:1} Model agreement discussion}
+{\color{red}[Model agreement discussion]}
+
+% ================================================== %
+\begin{figure}[ht]
+\includegraphics[width=\reprintcolumnwidth]{figsamp.jpg}
+\caption{\label{fig:FIG5}{Comparison of range-independent model vs. field data}}
+\end{figure}
+% ================================================== %
+
+\subsection{\label{subsec:4:2} Bootstrapping EOF}
+{\color{red}[Bootstrapping EOF]}
+
+% ================================================== %
+\begin{figure}[ht]
+\includegraphics[width=\reprintcolumnwidth]{figsamp.jpg}
+\caption{\label{fig:FIG6}{Convergence}}
+\end{figure}
+% ================================================== %
+
+% ================================================== %
+\begin{figure}[ht]
+\includegraphics[width=\reprintcolumnwidth]{figsamp.jpg}
+\caption{\label{fig:FIG7}{Bootstrap error estimate}}
+\end{figure}
+% ================================================== %
+
+% ================================================== %
+\begin{figure}[ht]
 \includegraphics[width=\reprintcolumnwidth]{figsamp.jpg}
@@
 \raggedright
+{\color{red}
 Note: The only figure formats allowed are the following:
@@
 (Figure1.eps, Figure2.jpg, Figure3a.ps, Figure3b.ps, etc).
+}
 
@@
 \end{figure}
-
-\section{\label{sec:2} Section Two}
+% ================================================== %
 
@@
 section~\ref{sec:2} and details  in subsection~\ref{subsec:2:1}.
-
-\subsection{\label{subsec:2:1} Sample subsection}
 
@@
 
-Make your bibliography by doing: pdflatex filename,  bibtex filename,
-pdflatex filename, pdflatex filename.
+Here is an example of algorithmic:
 
-{\bfseries\itshape
-Compare the results you get with\\
-{\verb+\documentclass[preprint]{JASA}+ }\\
-vs.\\
-{\verb+\documentclass[preprint,NumberedRefs]{JASA}+ }
-}
+\begin{algorithmic}
+\If {$i\geq maxval$}
+    \State $i\gets 0$
+\Else
+    \If {$i+k\leq maxval$}
+        \State $i\gets i+k$
+    \EndIf
+\EndIf
+\end{algorithmic}
+
+\begin{acknowledgments}
+This research was supported by  ...
+\end{acknowledgments}
 

### 2020-11-12 eof_play/generate_coeffs_plot.m
all files in working directory

@@ -1,0 +1,266 @@
+%% generate_coeffs.m
+% eeshan bhatt | eesh (at) mit (dot) edu
+% LAMSS - Dr. Henrik Schmidt
+% Feb 2020
+
+function [chosen_weights,chosen_eofnum,ssp_approx] = generate_coeffs_plot(z,ssp,eof_file);
+
+addpath(genpath('./helper/'));
+
+[z,sort_ind] = sort(z);
+ssp = ssp(sort_ind);
+
+
+
+
+%% get new weights based off EOF file on vehicle
+
+filename        = eof_file;
+eofs            = ncread(filename,'eofs'); eofs = double(eofs);
+baseval         = ncread(filename,'baseval'); baseval = double(baseval);
+num_eofs        = ncread(filename,'num_eofs'); num_eofs = double(num_eofs);
+num_depth       = ncread(filename,'num_depth'); num_depth = double(num_depth);
+depth           = ncread(filename,'depth'); depth = double(depth);
+default_weights = ncread(filename,'weights'); default_weights = double(default_weights);
+
+xi              = ncread(filename,'pdf_val'); xi = double(xi);
+f               = ncread(filename,'pdf_freq'); f = double(f);
+
+%% interpolate to HYCOM depth
+max_cast_depth = max(z);
+min_cast_depth = min(z);
+ind_cast_depth = find(depth > max_cast_depth, 1, 'first');
+Cq = interp1(z,ssp,depth(1:ind_cast_depth),'linear','extrap');
+
+Z = depth(1:ind_cast_depth);
+Cqq = [Cq; baseval(numel(Cq)+1:end)];
+count = 0;
+
+%% get estimate for R,Z (OPTIONAL)
+request_r = 0;
+request_z = 0;
+
+%% loop through combinations
+fprintf('\nrunning ray trace simulation... \n');
+for ne = 2:num_eofs
+    
+    combos = nchoosek(1:num_eofs,ne);
+    [num_combos,~] = size(combos);
+    
+    for cmb = 1:num_combos
+        count = count + 1;
+        
+        eof_subset = combos(cmb,:);
+        
+        % find weights for valid depths
+        ind_cast_depth = numel(Cq);
+        new_weights = eofs(1:ind_cast_depth,eof_subset) \ (Cq - baseval(1:ind_cast_depth));
+        new_weights = round(new_weights,3);
+        ssp_approx = baseval + eofs(:,eof_subset)*new_weights;
+        
+        % sum of squares compared CTD
[…]

### 2020-11-12 eof_play/helper/fix_hycom_forecast.m

@@ -1,0 +1,86 @@
+%% fix_hycom_forecast.m
+% eeshan bhatt | eesh (at) mit (dot) edu
+% LAMSS - Dr. Henrik Schmidt
+% Feb 2020
+
+%% setup
+clear; clc;
+
+warning('this is dangerous!!');
+
+loc = '../../../../missions-lamss/cruise/icex20/data/environment/ocean_files/*.nc';
+[file,path] = uigetfile(loc,'please pick the original TS file');
+tsfile = [path file];
+
+[file,path] = uigetfile(loc, 'please pick the original UV file');
+uvfile = [path file];
+
+%% duplicate files
+
+new_ts = [path 'hacked_ts.nc'];
+new_uv = [path 'hacked_uv.nc'];
+
+copyfile(tsfile,new_ts);
+copyfile(uvfile,new_uv);
+
+%% add attributes for ts file
+
+% salinity
+fileattrib(new_ts,'+w');
+ncwriteatt(new_ts,'salinity','missing_value',-30000);
+ncwriteatt(new_ts,'salinity','scale_factor',1);
+ncwriteatt(new_ts,'salinity','add_offset',0);
+
+% water_temp
+ncwriteatt(new_ts,'water_temp','missing_value',-30000);
+ncwriteatt(new_ts,'water_temp','scale_factor',1);
+ncwriteatt(new_ts,'water_temp','add_offset',0);
+
+%% add attributes for uv file
+
+% u
+fileattrib(new_uv,'+w');
+ncwriteatt(new_uv,'water_u','missing_value',-30000);
+ncwriteatt(new_uv,'water_u','scale_factor',0.001);
+ncwriteatt(new_uv,'water_u','add_offset',0);
+
+% v
+ncwriteatt(new_uv,'water_v','missing_value',-30000);
+ncwriteatt(new_uv,'water_v','scale_factor',0.001);
+ncwriteatt(new_uv,'water_v','add_offset',0);
+
+%% fix time?
+
+ncreadatt(tsfile,'time','units')
+ncreadatt(uvfile,'time','units')
+
+bool = input('\n confirm these are the same timestamp: [1/0] ');
+if bool
+    time = ncread(tsfile,'time');
+    n_yr = input('\nplease enter the year: ');
+    n_mth = input('\nplease enter the month: ');
+    n_day = input('\nplease enter the day: ');
+    n_hr = input('\nplease enter the hour: ');
+    n_min = input('\nplease enter the min: ');
[…]

### 2020-11-12 eof_play/helper/model_parabolicEquation.m

@@ -1,0 +1,122 @@
+function [rt,zt,tl] = model_parabolicEquation(ssp,z)
+% ******************************************************
+% PE
+% ******************************************************
+bool_plot = false;
+
+zs  = 33;			% source depth
+f   = 3500/8; omega = 2 * pi * f;	% frequency in Hertz
+
+d   = max(z);			% bottom depth
+nz  = numel(z);	% number of finite-difference points
+h   = d / nz; h2  = h * h;	% mesh spacing
+z   = linspace( 0, d, nz );	% grid coordinates
+
+c = ssp;
+
+% c0 = 1500;
+%x = 2 * ( z - 1300 ) / 1300;
+%c = c0 * ( 1 + eps * ( x - 1 + exp( -x ) ) );
+
+if bool_plot
+    figure(1)
+    subplot(1,5,1)
+    plot( z, c ); view( 90, 90 );
+    xlabel( 'Depth (m)' ); ylabel( 'Sound Speed (m/s)' )
+    xlim([0 1000]);
+    grid on
+end
+
+% ******************************************************
+% Gaussian starter
+% ******************************************************
+
+% \psi(0,z) = \sqrt{ k_0 } \,
+% e^{ -{ k_0^2 \over 2 } ( z - z_s )^2 } (6.100)
+
+c0 = min(ssp);
+k0 = omega / c0;
+isd = zs / d * nz;
+
+nr     = 1001;
+rmax   = 10000.0;
+deltar = rmax / nr;
+r = linspace( 0.0, rmax, nr );
+
+psi = zeros( nz, nr );
+
+% we have broadened the Gaussian to make it more
+% narrow angle
+% the usual formula has fac = 1
+fac = 10;
+psi( : , 1 ) = sqrt( k0 / fac ) * ...
+    exp( -( k0 / fac )^2 * ...
+    ( (z - zs * ones( 1, nz ) ) ).^2 )' ;
+
+% ******************************************************
+% Form marching matrix
+% ******************************************************
+% SPE: 2ik_0 { \pa \psi \over {\pa r} }
+%          + { \pa^2 \psi \over {\pa z^2} }
+%          + k_0^2 ( n^2 - 1 ) \psi = 0  (6.8)
+
+c0     = 1500;
+n      = c0 ./ c;
+
[…]

### 2020-11-12 eof_play/helper/model_rayTrace.m

@@ -1,0 +1,394 @@
+function [xxf, zzf, ttf, ddf] = model_rayTrace(xo,zo,theta0,tt,zz,cc,plotflag)
+%% A function to raytrace through a sound speed profile to a given time.
+%
+% *Val Schmidt
+% *Center for Coastal and Ocean Mapping / Joint Hydrographic Center
+% *University of New Hampshire
+% *2009
+%
+%
+% Arguments:
+%   xo:      starting horizontal coordinate of the ray in meters
+%   zo:      starting vertical coordinate of the ray in meters
+%   theta0:  starting angle of the ray in degrees
+%   tt:      max travel time of the ray in seconds
+%   zz:      vertical coordinates of the sound speed profile (+ down)
+%   cc:      sound speed measurements at zz locations
+%   plotflag: Boolean flag for auto plotting result (optional) default is 0.
+%
+% Output:
+%   xxf:       horizontal coordinates of the ray path
+%   zzf:       vertical coordinates of the ray path
+%   ttf:       actual travel time of the ray in seconds
+%   ddf:       distance the ray traveled
+%
+% Raytrace will trace the path sound will travel through a medium with an
+% isotropic horizontal sound speed profile and a vertical sound speed
+% profile given by (zz,cc) from a point (xo,zo) and departing at an angle
+% theta0 for a time of tt seconds. The path is returned in vectors x and z.
+% The actual time of travel is returned in t, which may be slightly greater
+% that tt. The sound speed profile should include a measurement at the
+% surface (zz = 0). If it does not, the profile will be extended to the
+% surface by replicating the first measurement there.
+%%
+% In addition to the above, raytrace() handles caustics (when rays bend
+% through horizontal) and reflections off the surface (z=0) and bottom
+% which is taken to be at the maximum depth measurement in the sound speed
+% profile. The bottom is assumed to be flat. Reflections up to a maximum
+% number (MAXBOUNCES) is supported. MAXBOUNCES is hard-coded into the
+% function. Decreasing MAXBOUNCES will reduce the memory required for
+% calculations when many reflections are not needed.
+%%
+% EXAMPLE:
+%
+% Trace a ray for 120 seconds starting from 0 m depth at a 5 degree down
+% angle through a sound speed profile that has a value of 1520 at the
+% surface, decreases with a gradient of -0.05 m/s /m to 750 m, and then
[…]

### 2020-11-12 eof_play/run_rt.m

@@ -1,0 +1,92 @@
+function [ray_rz_error,R,Z] = run_rt(y_ssp,yhat_ssp,z,travel_time,request_r,request_z,plot_bool)
+
+range_check = false;
+% define plot_bool, range_check
+if nargin == 4
+    plot_bool = false;
+elseif nargin == 6
+    plot_bool = false;
+    if request_r + request_z > 0
+        range_check = true;
+        travel_time = travel_time(1);
+    end
+end
+
+% fix grid spacing for raytracer
+zspace = 0:max(z);
+Yq = interp1(z,y_ssp,zspace);
+Yhatq = interp1(z,yhat_ssp,zspace);
+
+yvals = [Yq; Yhatq];
+
+% ray tracing parameters
+if range_check
+    theta = 0:0.5:90;
+else
+    theta = 0:2:70;
+end
+
+theta(theta==90)=[];
+num_theta = length(theta);
+
+z0 = 33;
+
+% empty structures to compare
+R = zeros(2,num_theta);
+Z = R;
+
+% loop over times
+for tt = 1:numel(travel_time)
+    tnow = travel_time(tt);
+    % loop over two SSPs
+    for yv = 1:2
+        [rr,zz,~,~] = model_rayTrace(0,z0,theta,tnow,zspace,yvals(yv,:),false);
+        
+        if range_check
+            for dtheta = 1:length(theta)
+                indr = find(abs(rr{dtheta} - request_r) > 0.2*request_r);
+                indz = find(abs(zz{dtheta} - request_z) > 0.2*request_z);
+                ind_rc = union(indr,indz);
+                
+                % we know rr,zz will hold b/c range_check only runs w/ one
+                % time
+                if numel(ind_rc) > 0
+                    rr{dtheta}(ind_rc) = NaN;
+                    zz{dtheta}(ind_rc) = NaN;
+                end
+                
+                R(yv,dtheta) = mean(real(rr{dtheta}),'omitnan');
+                Z(yv,dtheta) = mean(real(zz{dtheta}),'omitnan');
+            end
+        else
+            for dtheta = 1:length(theta)
+                R(yv,dtheta) = real(rr{dtheta}(end));
+                Z(yv,dtheta) = real(zz{dtheta}(end));
+            end
+        end
+    end   
+    
+% environmental error
+range_error(tt,:) = R(2,:) - R(1,:);
[…]

### 2020-11-12 src/ieeeplot.m

@@ -1,0 +1,51 @@
+%% plot w/ coeffs
+
+load bradli_ctd_eofnum.mat
+
+figure(1); clf;
+
+filename = '~/.dropboxmit/icex_2020_mat/eeof_itp_Mar2013.nc';
+num_eofs = double(ncread(filename,'num_eofs'));
+pdf_val = double(ncread(filename,'pdf_val'));
+pdf_freq = double(ncread(filename,'pdf_freq'));
+weights = double(ncread(filename,'weights'));
+baseval = double(ncread(filename,'baseval'));
+
+
+listlist = NaN(8,7);
+for cc = [1 2 4:9]
+    listlist(cc,ctd_eofnum{cc}) = ctd_weight{cc};
+end
+
+
+
+%% figure
+figure(1)
+for ne = 1:num_eofs
+    subplot(4,2,ne);
+    plot(pdf_val(ne,:),pdf_freq(ne,:)/max(pdf_freq(ne,:)))
+    grid on
+    xval = max(abs(pdf_val(ne,:)));
+    xlim([-xval xval])
+    
+    [~,ind_max] = max(pdf_freq(ne,:));
+
+    
+    hold on
+    plot([pdf_val(ne,ind_max) pdf_val(ne,ind_max)],[0 1],'r--')
+    
+    
+    Fq = interp1(pdf_val(ne,:),pdf_freq(ne,:)/max(pdf_freq(ne,:)),listlist(:,ne));
+    Fq(isnan(Fq)) = 0;
+    
+    plot(listlist(:,ne),Fq,'*')
+    
+    hold off
+    
+    default_weights(ne) = pdf_val(ne,ind_max);
+    
+    str = sprintf('eeof %d',ne);
+    title(str);
+    
+    axis tight
+end

### 2020-11-12 src/make_env_plots_3x3.m

@@ -51,16 +51,22 @@ num_check = 3;
 itp_c = cell(length(CTD),num_check);
 itp_z = cell(length(CTD),num_check);
 
+disp('finished ITP search! \n');
+
 
 %% loop through CTDs
 for p = 1:length(CTD)
     
-    %% ctd info
+    % ctd info
     ind_below = find(CTD(p).raw_z > 2);
     
     ctd_z{p} = CTD(p).raw_z(ind_below);
     ctd_c{p} = CTD(p).raw_c(ind_below);
     ctd_t(p) = datenum(CTD(p).time,'DD-mmm-yy HHMM');
+    
+    addpath(genpath('./eof_play/'));
+    eof_filepath = '~/.dropboxmit/icex_2020_mat/eeof_itp_Mar2013.nc';
+    [ctd_weight{p},ctd_eofnum{p},yhat{p}] = generate_coeffs_plot(ctd_z{p},ctd_c{p},eof_filepath);
     
     request_depth = max(CTD(p).raw_z);
     request_time = datenum(ctd_t(p));
@@ -140,12 +146,16 @@ end
 %% write to a figure
 
 figure(1); clf;
-ha = tight_subplot(3,3,[.06 .03],[.1 .05],[.1 .02]);
+ha = tight_subplot(1,4,[.06 .03],[.1 .05],[.1 .02]);
 
-for p = 1:9
+eeof_depths  = double(ncread(eof_filepath,'depth'));
+
+pcount = 0;
+for p = [1 4 5 8]
     
+    pcount = pcount + 1;
     % get figure
-    axes(ha(p));
+    axes(ha(pcount));
     
     if strcmp(CTD(p).type,'rsk')
         color = [153 51 153]./256;
@@ -159,31 +169,31 @@ for p = 1:9
         p_itp = plot(itp_c{p,lp},itp_z{p,lp},'color',[5 119 177 80]./256,'linewidth',3);
     end
     p_ctd = plot(ctd_c{p},ctd_z{p},'-','color',color,'linewidth',3);
+    
+    % plot EOF -- 
+    p_eof = plot(yhat{p},eeof_depths, 'ko');
+    
     hold off
     
     beautify_plot();
     title(datestr(ctd_t(p),'DD mmm yy HHMM'));
     
-    if mod(p,3) ~= 1
+    if p > 1
         yticklabels('');
     end
     
-    if p <= 6
-        xticklabels('');
-    end
     
-    if p == 8
-        xlabel('c [m/s]')
-    end
[…]

### 2020-11-12 modem-paper/main.tex
affiliations for MIT-WHOI

@@
 \title[JASA/Sample JASA Article]{Considering vertical sound speed uncertainty in the Beaufort Lens for more realistic range estimation between acoustic modems}
-\author{EeShan Bhatt}
+\author{EeShan C. Bhatt}
+\email{eesh@mit.edu}
+\affiliation{Department of Mechanical Engineering, Massachusetts Institute of Technology, Cambridge, MA}
+\affiliation{Applied Ocean Physics \& Engineering, Woods Hole Oceanographic Institution, Woods Hole, MA}
 \author{Oscar Viquez}
+\affiliation{Department of Mechanical Engineering, Massachusetts Institute of Technology, Cambridge, MA}
 \author{Bradli Howard}
+\affiliation{Department of Mechanical Engineering, Massachusetts Institute of Technology, Cambridge, MA}
+\affiliation{Applied Ocean Physics \& Engineering, Woods Hole Oceanographic Institution, Woods Hole, MA}
 \author{Henrik Schmidt}
-\affiliation{Department of Mechanical Engineering,  Massachusetts Institute of Technology, Cambridge, MA 02139, USA}
+\affiliation{Department of Mechanical Engineering, Massachusetts Institute of Technology, Cambridge, MA}
 

### 2020-12-31 src/main_plot_bathy.m
plot code w/ figures

@@ -1,0 +1,12 @@
+%% test_bathy
+clear; clc;
+
+% filename
+bathyfile = '~/missions-lamss/cruise/icex20/data/environment/noaa_bathy_file.nc';
+
+lat = ncread(bathyfile,'lat');
+lon = ncread(bathyfile,'lon');
+bathy = ncread(bathyfile,'Band1');
+
+
+

### 2020-12-31 src/main_plot_tobytest_by_design.m

@@ -21,7 +21,7 @@ olat = 71.17733;
 olon = -142.40413;
 
 %%  pick a toby test event!
-for iNL = 1 % num_listing
+for iNL = 2 % num_listing
     
     figure(iNL); clf;
     

### 2020-12-31 src/main_plot_tobytest_by_event.m

@@ -17,7 +17,7 @@ listing = dir(location);
 num_listing = numel(listing);
 
 %%  pick a toby test event!
-for iNL = 5
+for iNL = 4
     
     figure(1); clf;
     

### 2020-12-31 src/itp_import_buoy_loc.m
itp codes

@@ -1,4 +1,4 @@
-function [timestamp,lon,lat] = itp_import_rawloc(filename)
+function [timestamp,lon,lat] = itp_import_buoy_loc(filename)
 
 B = load(filename);
 

### 2020-12-31 src/itp_import_profiler_data.m

@@ -1,4 +1,4 @@
-function [year,day,pressure,temp,salinity,oxygen] = importfile_itp(filename, startRow, endRow)
+function [year,day,pressure,temp,salinity,oxygen] = itp_import_profiler_data(filename, startRow, endRow)
 %IMPORTFILE_ITP Import numeric data from a text file as column vectors.
 %   [YEAR,DAY,PRESSURE,TEMP,SALINITY,OXYGEN] = IMPORTFILE1(FILENAME) Reads
 %   data from text file FILENAME for the default selection.

### 2020-12-31 src/itp_import_profiler_loc.m

@@ -1,4 +1,4 @@
-function info = itp_import_loc(filename, startRow, endRow)
+function info = itp_import_profiler_loc(filename, startRow, endRow)
 %IMPORTFILE Import numeric data from a text file as a matrix.
 %   ITP103GRD0000 = IMPORTFILE(FILENAME) Reads data from text file FILENAME
 %   for the default selection.

### 2020-12-31 src/itp_import_buoy_loc.m
moved itp codes to lamss-environmental

@@ -1,15 +1,0 @@
-function [timestamp,lon,lat] = itp_import_buoy_loc(filename)
-
-B = load(filename);
-
-%% Allocate imported array to column variable names
-year = B(:,1);
-day = B(:,2);
-lon = B(:,3);
-lat = B(:,4);
-buffer = zeros(size(year));
-timestamp = datenum([year buffer day buffer buffer buffer]);
-end
-
-
-

### 2020-12-31 src/itp_import_profiler_data.m

@@ -1,102 +1,0 @@
-function [year,day,pressure,temp,salinity,oxygen] = itp_import_profiler_data(filename, startRow, endRow)
-%IMPORTFILE_ITP Import numeric data from a text file as column vectors.
-%   [YEAR,DAY,PRESSURE,TEMP,SALINITY,OXYGEN] = IMPORTFILE1(FILENAME) Reads
-%   data from text file FILENAME for the default selection.
-%
-%   [YEAR,DAY,PRESSURE,TEMP,SALINITY,OXYGEN] = IMPORTFILE1(FILENAME,
-%   STARTROW, ENDROW) Reads data from rows STARTROW through ENDROW of text
-%   file FILENAME.
-%
-% Example:
-%   [year,day,pressure,temp,salinity,oxygen] = importfile_itp('itp84grd0000.dat',2, 177);
-%
-%    See also TEXTSCAN.
-
-% Auto-generated by MATLAB on 2018/11/28 17:28:05
-
-%% Initialize variables.
-if nargin<=2
-    startRow = 4;
-    endRow = inf;
-end
-
-%% Read columns of data as text:
-% For more information, see the TEXTSCAN documentation.
-formatSpec = '%4s%11s%5s%10s%10s%s%[^\n\r]';
-
-%% Open the text file.
-fileID = fopen(filename,'r');
-
-%% Read columns of data according to the format.
-% This call is based on the structure of the file used to generate this
-% code. If an error occurs for a different file, try regenerating the code
-% from the Import Tool.
-dataArray = textscan(fileID, formatSpec, endRow(1)-startRow(1)+1, 'Delimiter', '', 'WhiteSpace', '', 'TextType', 'string', 'HeaderLines', startRow(1)-1, 'ReturnOnError', false, 'EndOfLine', '\r\n');
-for block=2:length(startRow)
-    frewind(fileID);
-    dataArrayBlock = textscan(fileID, formatSpec, endRow(block)-startRow(block)+1, 'Delimiter', '', 'WhiteSpace', '', 'TextType', 'string', 'HeaderLines', startRow(block)-1, 'ReturnOnError', false, 'EndOfLine', '\r\n');
-    for col=1:length(dataArray)
-        dataArray{col} = [dataArray{col};dataArrayBlock{col}];
-    end
-end
-
-%% Close the text file.
-fclose(fileID);
-
-%% Convert the contents of columns containing numeric text to numbers.
-% Replace non-numeric text with NaN.
[…]

### 2020-12-31 src/itp_import_profiler_loc.m

@@ -1,77 +1,0 @@
-function info = itp_import_profiler_loc(filename, startRow, endRow)
-%IMPORTFILE Import numeric data from a text file as a matrix.
-%   ITP103GRD0000 = IMPORTFILE(FILENAME) Reads data from text file FILENAME
-%   for the default selection.
-%
-%   ITP103GRD0000 = IMPORTFILE(FILENAME, STARTROW, ENDROW) Reads data from
-%   rows STARTROW through ENDROW of text file FILENAME.
-%
-% Example:
-%   itp103grd0000 = importfile('itp103grd0000.dat', 2, 2);
-%
-%    See also TEXTSCAN.
-
-% Auto-generated by MATLAB on 2019/11/14 16:25:09
-
-%% Initialize variables.
-if nargin<=2
-    startRow = 2;
-    endRow = 2;
-end
-
-%% Format for each line of text:
-%   column1: double (%f)
-%	column2: double (%f)
-%   column3: double (%f)
-%	column4: double (%f)
-%   column5: double (%f)
-% For more information, see the TEXTSCAN documentation.
-formatSpec = '%4f%11f%5f%10f%f%[^\n\r]';
-
-%% Open the text file.
-fileID = fopen(filename,'r');
-
-%% Read columns of data according to the format.
-% This call is based on the structure of the file used to generate this
-% code. If an error occurs for a different file, try regenerating the code
-% from the Import Tool.
-dataArray = textscan(fileID, formatSpec, endRow(1)-startRow(1)+1, 'Delimiter', '', 'WhiteSpace', '', 'TextType', 'string', 'EmptyValue', NaN, 'HeaderLines', startRow(1)-1, 'ReturnOnError', false, 'EndOfLine', '\r\n');
-for block=2:length(startRow)
-    frewind(fileID);
-    dataArrayBlock = textscan(fileID, formatSpec, endRow(block)-startRow(block)+1, 'Delimiter', '', 'WhiteSpace', '', 'TextType', 'string', 'EmptyValue', NaN, 'HeaderLines', startRow(block)-1, 'ReturnOnError', false, 'EndOfLine', '\r\n');
-    for col=1:length(dataArray)
-        dataArray{col} = [dataArray{col};dataArrayBlock{col}];
-    end
-end
-
-%% Close the text file.
-fclose(fileID);
-
-%% Post processing for unimportable data.
-% No unimportable data rules were applied during the import, so no post
[…]

### 2020-12-31 src/data-prep/main_collate_toby_test_data.m
data prep update

@@ -167,7 +167,7 @@ for itt = 1:nTobyTest
     fprintf('\n');
     
 end
-filename = sprintf('./data-tobytest-by-design/tobytest-all');
+filename = sprintf('tobytest-all');
 
 save(filename,'event');
 fprintf(' saved %s.mat \n',filename);

### 2020-12-31 src/data-prep/main_sort_tobytest_by_design.m

@@ -24,7 +24,7 @@ for utd = unique_tx_depth
         index_eeof  = eeof_bool == ute;
         index = and(index_depth,index_eeof);
         
-        str = sprintf('./data-tobytest-by-design/tobytest_txz%d_eeof%d',utd,ute);
+        str = sprintf('tobytest_txz%d_eeof%d',utd,ute);
         
         experiment = event(index);
         save(str,'experiment');

### 2020-12-31 src/play_toby_test.m
ITP raw location plot on white/gray map w/ Seadragon

@@ -1,105 +1,0 @@
-%% play_toby_test.m
-
-clear; clc;
-
-%% directory
-dir_toby_test = '~/.dropboxmit/icex_2020_mat/toby_test_sorted/*.mat';
-listing = dir(dir_toby_test);
-
-% macrura   modem ID = 4  -- ignore this for now?
-% h1        modem ID = 10
-% h2        modem ID = 11
-% h3        modem ID = 12
-% h4        modem ID = 13
-
-dir_group_velocity = '~/.dropboxmit/icex_ipynb/gvels_by_node.mat';
-load(dir_group_velocity);
-gvels_hydrohole = gvels_by_node.hydrohole;
-gvels_macrura = gvels_by_node.macrura;
-
-%% plot -- group velocities
-
-
-
-%% plot -- grid of tx/rx
-
-for ld = 4
-    
-    filename = [listing(ld).folder '/' listing(ld).name];
-    
-    disp(listing(ld).name);
-    
-    A = load(filename);
-    
-    % print time?
-    t0 = A.nav.toby_test.time(1);
-    tf = A.nav.toby_test.time(end);
-    
-    fprintf('%s to %s \n', h_convertTime(t0,1), h_convertTime(tf,1));
-    
-    tx_rx_owtt=cell(5,5);
-    
-    % sort h1, h2, h3, h4
-    [tx_rx_owtt] = h_sort_tx_rx(A.comms.h1,1,tx_rx_owtt);
-    [tx_rx_owtt] = h_sort_tx_rx(A.comms.h2,2,tx_rx_owtt);
-    [tx_rx_owtt] = h_sort_tx_rx(A.comms.h3,3,tx_rx_owtt);
-    [tx_rx_owtt] = h_sort_tx_rx(A.comms.h4,4,tx_rx_owtt);
-    [tx_rx_owtt] = h_sort_tx_rx(A.comms.macrura_10k,5,tx_rx_owtt);
-    
-end
-
-%% plot
-
-figure(1);clf;
-
-ha = tight_subplot(5,5,.05,.08,.08);
-
-count_rr_tt = 0;
-for rr = 1:5
-    for tt = 1:5
-        
-        count_rr_tt = count_rr_tt + 1;
-        
-        axes(ha(count_rr_tt));
-        
-        num_contacts = length(tx_rx_owtt{rr,tt});
-        
-        
-        if num_contacts > 0
-            vals = tx_rx_owtt{rr,tt};
-            histogram(vals,ceil(exp(1).*sqrt(num_contacts)))
[…]

### 2020-12-31 src/plot_itp_raw_locs.m

@@ -1,4 +1,4 @@
-%% plot_raw_locs.m
+%% main_plot_itp_raw_locs.m
 
 %% prep workspace
 clear; clc;
@@ -19,7 +19,7 @@ newcolors = distinguishable_colors(7);
 for l = 1:length(listing)
     filepath = fullfile([listing(l).folder '/' listing(l).name]);
     
-    [timestamp, lon, lat] = itp_import_rawloc(filepath);
+    [timestamp, lon, lat] = itp_import_buoy_loc(filepath);
     
     fprintf('%s | %s | %s \n',listing(l).name(1:6),datestr(min(timestamp)),datestr(max(timestamp))) 
     

### 2020-12-31 eof_play/generate_coeffs_plot.m
removing eof_play

@@ -1,266 +1,0 @@
-%% generate_coeffs.m
-% eeshan bhatt | eesh (at) mit (dot) edu
-% LAMSS - Dr. Henrik Schmidt
-% Feb 2020
-
-function [chosen_weights,chosen_eofnum,ssp_approx] = generate_coeffs_plot(z,ssp,eof_file);
-
-addpath(genpath('./helper/'));
-
-[z,sort_ind] = sort(z);
-ssp = ssp(sort_ind);
-
-
-
-
-%% get new weights based off EOF file on vehicle
-
-filename        = eof_file;
-eofs            = ncread(filename,'eofs'); eofs = double(eofs);
-baseval         = ncread(filename,'baseval'); baseval = double(baseval);
-num_eofs        = ncread(filename,'num_eofs'); num_eofs = double(num_eofs);
-num_depth       = ncread(filename,'num_depth'); num_depth = double(num_depth);
-depth           = ncread(filename,'depth'); depth = double(depth);
-default_weights = ncread(filename,'weights'); default_weights = double(default_weights);
-
-xi              = ncread(filename,'pdf_val'); xi = double(xi);
-f               = ncread(filename,'pdf_freq'); f = double(f);
-
-%% interpolate to HYCOM depth
-max_cast_depth = max(z);
-min_cast_depth = min(z);
-ind_cast_depth = find(depth > max_cast_depth, 1, 'first');
-Cq = interp1(z,ssp,depth(1:ind_cast_depth),'linear','extrap');
-
-Z = depth(1:ind_cast_depth);
-Cqq = [Cq; baseval(numel(Cq)+1:end)];
-count = 0;
-
-%% get estimate for R,Z (OPTIONAL)
-request_r = 0;
-request_z = 0;
-
-%% loop through combinations
-fprintf('\nrunning ray trace simulation... \n');
-for ne = 2:num_eofs
-    
-    combos = nchoosek(1:num_eofs,ne);
-    [num_combos,~] = size(combos);
-    
-    for cmb = 1:num_combos
-        count = count + 1;
-        
-        eof_subset = combos(cmb,:);
-        
-        % find weights for valid depths
-        ind_cast_depth = numel(Cq);
-        new_weights = eofs(1:ind_cast_depth,eof_subset) \ (Cq - baseval(1:ind_cast_depth));
-        new_weights = round(new_weights,3);
-        ssp_approx = baseval + eofs(:,eof_subset)*new_weights;
-        
-        % sum of squares compared CTD
[…]

### 2020-12-31 eof_play/helper/fix_hycom_forecast.m

@@ -1,86 +1,0 @@
-%% fix_hycom_forecast.m
-% eeshan bhatt | eesh (at) mit (dot) edu
-% LAMSS - Dr. Henrik Schmidt
-% Feb 2020
-
-%% setup
-clear; clc;
-
-warning('this is dangerous!!');
-
-loc = '../../../../missions-lamss/cruise/icex20/data/environment/ocean_files/*.nc';
-[file,path] = uigetfile(loc,'please pick the original TS file');
-tsfile = [path file];
-
-[file,path] = uigetfile(loc, 'please pick the original UV file');
-uvfile = [path file];
-
-%% duplicate files
-
-new_ts = [path 'hacked_ts.nc'];
-new_uv = [path 'hacked_uv.nc'];
-
-copyfile(tsfile,new_ts);
-copyfile(uvfile,new_uv);
-
-%% add attributes for ts file
-
-% salinity
-fileattrib(new_ts,'+w');
-ncwriteatt(new_ts,'salinity','missing_value',-30000);
-ncwriteatt(new_ts,'salinity','scale_factor',1);
-ncwriteatt(new_ts,'salinity','add_offset',0);
-
-% water_temp
-ncwriteatt(new_ts,'water_temp','missing_value',-30000);
-ncwriteatt(new_ts,'water_temp','scale_factor',1);
-ncwriteatt(new_ts,'water_temp','add_offset',0);
-
-%% add attributes for uv file
-
-% u
-fileattrib(new_uv,'+w');
-ncwriteatt(new_uv,'water_u','missing_value',-30000);
-ncwriteatt(new_uv,'water_u','scale_factor',0.001);
-ncwriteatt(new_uv,'water_u','add_offset',0);
-
-% v
-ncwriteatt(new_uv,'water_v','missing_value',-30000);
-ncwriteatt(new_uv,'water_v','scale_factor',0.001);
-ncwriteatt(new_uv,'water_v','add_offset',0);
-
-%% fix time?
-
-ncreadatt(tsfile,'time','units')
-ncreadatt(uvfile,'time','units')
-
-bool = input('\n confirm these are the same timestamp: [1/0] ');
-if bool
-    time = ncread(tsfile,'time');
-    n_yr = input('\nplease enter the year: ');
-    n_mth = input('\nplease enter the month: ');
-    n_day = input('\nplease enter the day: ');
-    n_hr = input('\nplease enter the hour: ');
-    n_min = input('\nplease enter the min: ');
[…]

### 2020-12-31 eof_play/helper/model_parabolicEquation.m

@@ -1,122 +1,0 @@
-function [rt,zt,tl] = model_parabolicEquation(ssp,z)
-% ******************************************************
-% PE
-% ******************************************************
-bool_plot = false;
-
-zs  = 33;			% source depth
-f   = 3500/8; omega = 2 * pi * f;	% frequency in Hertz
-
-d   = max(z);			% bottom depth
-nz  = numel(z);	% number of finite-difference points
-h   = d / nz; h2  = h * h;	% mesh spacing
-z   = linspace( 0, d, nz );	% grid coordinates
-
-c = ssp;
-
-% c0 = 1500;
-%x = 2 * ( z - 1300 ) / 1300;
-%c = c0 * ( 1 + eps * ( x - 1 + exp( -x ) ) );
-
-if bool_plot
-    figure(1)
-    subplot(1,5,1)
-    plot( z, c ); view( 90, 90 );
-    xlabel( 'Depth (m)' ); ylabel( 'Sound Speed (m/s)' )
-    xlim([0 1000]);
-    grid on
-end
-
-% ******************************************************
-% Gaussian starter
-% ******************************************************
-
-% \psi(0,z) = \sqrt{ k_0 } \,
-% e^{ -{ k_0^2 \over 2 } ( z - z_s )^2 } (6.100)
-
-c0 = min(ssp);
-k0 = omega / c0;
-isd = zs / d * nz;
-
-nr     = 1001;
-rmax   = 10000.0;
-deltar = rmax / nr;
-r = linspace( 0.0, rmax, nr );
-
-psi = zeros( nz, nr );
-
-% we have broadened the Gaussian to make it more
-% narrow angle
-% the usual formula has fac = 1
-fac = 10;
-psi( : , 1 ) = sqrt( k0 / fac ) * ...
-    exp( -( k0 / fac )^2 * ...
-    ( (z - zs * ones( 1, nz ) ) ).^2 )' ;
-
-% ******************************************************
-% Form marching matrix
-% ******************************************************
-% SPE: 2ik_0 { \pa \psi \over {\pa r} }
-%          + { \pa^2 \psi \over {\pa z^2} }
-%          + k_0^2 ( n^2 - 1 ) \psi = 0  (6.8)
-
-c0     = 1500;
-n      = c0 ./ c;
-
[…]

### 2020-12-31 eof_play/helper/model_rayTrace.m

@@ -1,394 +1,0 @@
-function [xxf, zzf, ttf, ddf] = model_rayTrace(xo,zo,theta0,tt,zz,cc,plotflag)
-%% A function to raytrace through a sound speed profile to a given time.
-%
-% *Val Schmidt
-% *Center for Coastal and Ocean Mapping / Joint Hydrographic Center
-% *University of New Hampshire
-% *2009
-%
-%
-% Arguments:
-%   xo:      starting horizontal coordinate of the ray in meters
-%   zo:      starting vertical coordinate of the ray in meters
-%   theta0:  starting angle of the ray in degrees
-%   tt:      max travel time of the ray in seconds
-%   zz:      vertical coordinates of the sound speed profile (+ down)
-%   cc:      sound speed measurements at zz locations
-%   plotflag: Boolean flag for auto plotting result (optional) default is 0.
-%
-% Output:
-%   xxf:       horizontal coordinates of the ray path
-%   zzf:       vertical coordinates of the ray path
-%   ttf:       actual travel time of the ray in seconds
-%   ddf:       distance the ray traveled
-%
-% Raytrace will trace the path sound will travel through a medium with an
-% isotropic horizontal sound speed profile and a vertical sound speed
-% profile given by (zz,cc) from a point (xo,zo) and departing at an angle
-% theta0 for a time of tt seconds. The path is returned in vectors x and z.
-% The actual time of travel is returned in t, which may be slightly greater
-% that tt. The sound speed profile should include a measurement at the
-% surface (zz = 0). If it does not, the profile will be extended to the
-% surface by replicating the first measurement there.
-%%
-% In addition to the above, raytrace() handles caustics (when rays bend
-% through horizontal) and reflections off the surface (z=0) and bottom
-% which is taken to be at the maximum depth measurement in the sound speed
-% profile. The bottom is assumed to be flat. Reflections up to a maximum
-% number (MAXBOUNCES) is supported. MAXBOUNCES is hard-coded into the
-% function. Decreasing MAXBOUNCES will reduce the memory required for
-% calculations when many reflections are not needed.
-%%
-% EXAMPLE:
-%
-% Trace a ray for 120 seconds starting from 0 m depth at a 5 degree down
-% angle through a sound speed profile that has a value of 1520 at the
-% surface, decreases with a gradient of -0.05 m/s /m to 750 m, and then
[…]

### 2020-12-31 eof_play/helper/snd_spd.m

@@ -1,443 +1,0 @@
-function [ssp,info] = snd_spd(D,T,S,equation,lat)
-
-% SND_SPD Various sound-speed equations.
-%
-%          1)  SND_SPD(D,T,S) returns the sound speed (m/sec) given vectors
-%             of salinity (ppt), temperature (deg C) and DEPTH !!!  (m) using
-%             the formula of Mackenzie:
-%
-%           Mackenzie, K.V. "Nine-term Equation for Sound Speed in the Oceans",
-%           J. Acoust. Soc. Am. 70 (1981), 807-812.
-%
-%
-%          2) SND_SPD(P,T,S,'del grosso',lat) returns the sound speed (m/sec) 
-%            given vectors of salinity (ppt), temperature (deg C), and 
-%            PRESSURE (dbar) using the Del Grosso equation:
-%
-%            Del Grosso, "A New Equation for the speed of sound in Natural
-%            Waters", J. Acoust. Soc. Am. 56#4 (1974).
-%
-%          3) SND_SPD(P,T,S,'chen') returns the sound speed (m/sec) given
-%            given vectors of salinity (ppt), temperature (deg C), and 
-%            PRESSURE (dbar) using the Chen and Millero equation:
-%
-%            Chen and Millero, "The Sound Speed in Seawater", J. Acoust. 
-%            Soc. Am. 62 (1977), 1129-1135
-%
-%          4) SND_SPD(P,T,S,'state') returns the sound speed (m/sec) given
-%            given vectors of salinity (ppt), temperature (deg C), and 
-%            PRESSURE (dbar) by using derivatives of the EOS80 equation of
-%            state for seawater and the adiabatic lapse rate.
-% 
-%  [SSP,Info] = SND_SPD( ... ) returns the sound speed SSP and 
-%                               the [ 3 by 1 ] CellStringArray Info with:
-%                               { Author ; Title ; Reference }
-%
-%  requires m-files:  ADIABATT.M,
-%                     SWSTATE.M,
-%
-
[…]

### 2020-12-31 eof_play/run_rt.m

@@ -1,92 +1,0 @@
-function [ray_rz_error,R,Z] = run_rt(y_ssp,yhat_ssp,z,travel_time,request_r,request_z,plot_bool)
-
-range_check = false;
-% define plot_bool, range_check
-if nargin == 4
-    plot_bool = false;
-elseif nargin == 6
-    plot_bool = false;
-    if request_r + request_z > 0
-        range_check = true;
-        travel_time = travel_time(1);
-    end
-end
-
-% fix grid spacing for raytracer
-zspace = 0:max(z);
-Yq = interp1(z,y_ssp,zspace);
-Yhatq = interp1(z,yhat_ssp,zspace);
-
-yvals = [Yq; Yhatq];
-
-% ray tracing parameters
-if range_check
-    theta = 0:0.5:90;
-else
-    theta = 0:2:70;
-end
-
-theta(theta==90)=[];
-num_theta = length(theta);
-
-z0 = 33;
-
-% empty structures to compare
-R = zeros(2,num_theta);
-Z = R;
-
-% loop over times
-for tt = 1:numel(travel_time)
-    tnow = travel_time(tt);
-    % loop over two SSPs
-    for yv = 1:2
-        [rr,zz,~,~] = model_rayTrace(0,z0,theta,tnow,zspace,yvals(yv,:),false);
-        
-        if range_check
-            for dtheta = 1:length(theta)
-                indr = find(abs(rr{dtheta} - request_r) > 0.2*request_r);
-                indz = find(abs(zz{dtheta} - request_z) > 0.2*request_z);
-                ind_rc = union(indr,indz);
-                
-                % we know rr,zz will hold b/c range_check only runs w/ one
-                % time
-                if numel(ind_rc) > 0
-                    rr{dtheta}(ind_rc) = NaN;
-                    zz{dtheta}(ind_rc) = NaN;
-                end
-                
-                R(yv,dtheta) = mean(real(rr{dtheta}),'omitnan');
-                Z(yv,dtheta) = mean(real(zz{dtheta}),'omitnan');
-            end
-        else
-            for dtheta = 1:length(theta)
-                R(yv,dtheta) = real(rr{dtheta}(end));
-                Z(yv,dtheta) = real(zz{dtheta}(end));
-            end
-        end
-    end   
-    
-% environmental error
-range_error(tt,:) = R(2,:) - R(1,:);
[…]

### 2020-12-31 src/play_owtt_range_gvel.m

@@ -1,123 +1,0 @@
-%% play_owtt_range_gvel.m
-clear;
-
-% eeshan bhatt
-
-%% load group velocity data
-dir_group_velocity = '~/.dropboxmit/icex_ipynb/gvels_by_node.mat';
-load(dir_group_velocity);
-gvels_topside = gvels_by_node.hydrohole;
-gvels_macrura = gvels_by_node.macrura;
-clear gvels_by_node
-
-%% load toby test data
-dir_toby_test = '~/.dropboxmit/icex_2020_mat/toby_test_sorted/*.mat';
-listing_toby_test = dir(dir_toby_test);
-nTobyTest = length(listing_toby_test);
-
-%% loop through toby tests
-
-for itt = 1:nTobyTest
-    
-    %% comms
-    A = load([listing_toby_test(itt).folder '/' listing_toby_test(itt).name]);
-    
-    nameString = extractBefore(listing_toby_test(itt).name,'.');
-    nameString = split(nameString,'_');
-    if length(nameString) == 4
-        nameString = sprintf('toby test %s',nameString{4});
-    else
-        nameString = sprintf('toby test %s.%s',nameString{4},nameString{5});
-    end
-    
-    % A.comms.h*.event   : src, dest, arr_time, ref_sec, travel_time 
-    % A.comms.h*.nav     : time, lat, lon, depth, x, y, z
-    % A.comms.h*.src_nav : time, lat, lon, depth, x, y, z
-    
-    comm_ids = {'h1','h2','h3','h4','macrura_10k'};
-    nIDs = length(comm_ids);
-    tx_rx_owtt=cell(5,5);
-    oneWayTravelTime = [];
-    
-    for k = 1:nIDs
-        
-        %[tx_rx_owtt] = h_sort_tx_rx(A.comms,comm_ids,tx_rx_owtt);
-        
-        [owttByReceiver{k}] = h_get_owtt(A.comms, comm_ids{k});
-        oneWayTravelTime = [oneWayTravelTime owttByReceiver{k}];
-    end
-
-    
-    %% make KDE
-    [F,XI]=ksdensity(oneWayTravelTime,'support','positive','NumPoints',1000,'function','pdf','bandwidth',2.^(-5));
-    
-    % subplot
-    subplot(3,3,itt)
-    plot(XI,F);
-    title(nameString)
-    
-end
-
-% subplot 1: map of h* lat/lon
-
-
-
[…]

### 2020-12-31 src/plot_ctd_hycom_ensemble.m
ensemble plotting script

@@ -1,4 +1,3 @@
-%% make_env_plots.m
 % eeshan bhatt
 % eesh@mit.edu
 

### 2020-12-31 src/plot_ctd_hycom_itp.m
exploration plots

@@ -1,4 +1,4 @@
-%% make_env_plots.m
+%% plot_ctd_hycom_itp.m
 % eeshan bhatt
 % eesh@mit.edu
 
@@ -39,7 +39,7 @@ for cc = 1:length(listing)
     dat_file_path = fullfile(listing(cc).folder, listing(cc).name);
     
     itp_search.filepath(cc)   = dat_file_path;
-    info                        = itp_import_loc(dat_file_path);
+    info                      = itp_import_profiler_loc(dat_file_path);
     
     itp_search.lon(cc)        = info.lon;
     itp_search.lat(cc)        = info.lat;
@@ -51,7 +51,7 @@ num_check = 3;
 itp_c = cell(length(CTD),num_check);
 itp_z = cell(length(CTD),num_check);
 
-disp('finished ITP search! \n');
+disp('finished ITP search!');
 
 
 %% loop through CTDs
@@ -66,7 +66,6 @@ for p = 1:length(CTD)
     
     addpath(genpath('./eof_play/'));
     eof_filepath = '~/.dropboxmit/icex_2020_mat/eeof_itp_Mar2013.nc';
-    [ctd_weight{p},ctd_eofnum{p},yhat{p}] = generate_coeffs_plot(ctd_z{p},ctd_c{p},eof_filepath);
     
     request_depth = max(CTD(p).raw_z);
     request_time = datenum(ctd_t(p));
@@ -125,7 +124,7 @@ for p = 1:length(CTD)
         num_obs = 0;
         while (num_obs < 200)
             count = count + 1;
-            [~,~,pressure,temp,salinity,~] = importfile_itp(itp_search.filepath(best_index(count)));
+            [~,~,pressure,temp,salinity,~] = itp_import_profiler_data(itp_search.filepath(best_index(count)));
             num_obs = length(pressure);
             
             itp_z{p,lp} = pressureToDepth(pressure,itp_search.lat(best_index(count)));
@@ -146,23 +145,25 @@ end
 %% write to a figure
 
 figure(1); clf;
-ha = tight_subplot(1,4,[.06 .03],[.1 .05],[.1 .02]);
+ha = tight_subplot(3,3,[.06 .03],[.1 .05],[.1 .02]);
 
 eeof_depths  = double(ncread(eof_filepath,'depth'));
 
 pcount = 0;
-for p = [1 4 5 8]
+for p = 1:9
     
+    % axes count -- not always equal to CTD count
[…]

### 2020-12-31 src/plot_modem_pdfs_rxtx.m

@@ -1,4 +1,4 @@
-%% make_modem_plots.m
+%% plot_modem_pdfs_rxtx.m
 
 %% prep workspace
 clear; clc;

### 2020-12-31 src/main_plot_bathy.m
delete useless script

@@ -1,12 +1,0 @@
-%% test_bathy
-clear; clc;
-
-% filename
-bathyfile = '~/missions-lamss/cruise/icex20/data/environment/noaa_bathy_file.nc';
-
-lat = ncread(bathyfile,'lat');
-lon = ncread(bathyfile,'lon');
-bathy = ncread(bathyfile,'Band1');
-
-
-

### 2020-12-31 src/main_plot_ctd_eeof_weights.m

@@ -1,51 +1,0 @@
-%% plot w/ coeffs
-
-load bradli_ctd_eofnum.mat
-
-figure(1); clf;
-
-filename = '~/.dropboxmit/icex_2020_mat/eeof_itp_Mar2013.nc';
-num_eofs = double(ncread(filename,'num_eofs'));
-pdf_val = double(ncread(filename,'pdf_val'));
-pdf_freq = double(ncread(filename,'pdf_freq'));
-weights = double(ncread(filename,'weights'));
-baseval = double(ncread(filename,'baseval'));
-
-
-listlist = NaN(8,7);
-for cc = [1 2 4:9]
-    listlist(cc,ctd_eofnum{cc}) = ctd_weight{cc};
-end
-
-
-
-%% figure
-figure(1)
-for ne = 1:num_eofs
-    subplot(4,2,ne);
-    plot(pdf_val(ne,:),pdf_freq(ne,:)/max(pdf_freq(ne,:)))
-    grid on
-    xval = max(abs(pdf_val(ne,:)));
-    xlim([-xval xval])
-    
-    [~,ind_max] = max(pdf_freq(ne,:));
-
-    
-    hold on
-    plot([pdf_val(ne,ind_max) pdf_val(ne,ind_max)],[0 1],'r--')
-    
-    
-    Fq = interp1(pdf_val(ne,:),pdf_freq(ne,:)/max(pdf_freq(ne,:)),listlist(:,ne));
-    Fq(isnan(Fq)) = 0;
-    
-    plot(listlist(:,ne),Fq,'*')
-    
-    hold off
-    
-    default_weights(ne) = pdf_val(ne,ind_max);
-    
-    str = sprintf('eeof %d',ne);
-    title(str);
-    
-    axis tight
-end

### 2021-01-01 doc/main.tex
modded template w/ useful makefile

--- 2020-11-12 modem-paper/main.tex
@@
-\title[JASA/Sample JASA Article]{Considering vertical sound speed uncertainty in the Beaufort Lens for more realistic range estimation between acoustic modems}
+\title[JASA/Sample JASA Article]{In situ estimation of sound speed profiles and horizontal group velocities for range estimation between under ice acoustic modems}%{Considering vertical sound speed uncertainty in the Beaufort Lens for more realistic range estimation between acoustic modems}
 \author{EeShan C. Bhatt}
 \email{eesh@mit.edu}
+\affiliation{MIT-WHOI Joint Program in Oceanography/Applied Ocean Science \& Engineering, Cambridge and Woods Hole, MA, USA}
 \affiliation{Department of Mechanical Engineering, Massachusetts Institute of Technology, Cambridge, MA}
-\affiliation{Applied Ocean Physics \& Engineering, Woods Hole Oceanographic Institution, Woods Hole, MA}
 \author{Oscar Viquez}
@@
 \author{Bradli Howard}
+\affiliation{MIT-WHOI Joint Program in Oceanography/Applied Ocean Science \& Engineering, Cambridge and Woods Hole, MA, USA}
 \affiliation{Department of Mechanical Engineering, Massachusetts Institute of Technology, Cambridge, MA}
-\affiliation{Applied Ocean Physics \& Engineering, Woods Hole Oceanographic Institution, Woods Hole, MA}
 \author{Henrik Schmidt}
+\email{henrik@mit.edu}
 \affiliation{Department of Mechanical Engineering, Massachusetts Institute of Technology, Cambridge, MA}
@@
 
-[Motivation: ICEX (\& Irminger?)]
+[Motivation: ICEX]
 }

### 2021-01-01 src/main_plot_tobytest_by_event.m
dashboard for each toby test finished --- use this to examine timing irregularity!

@@ -1,4 +1,6 @@
 %% main_plot_tobytest_by_event
+% explore data by each toby test
+% see notes for time divisions
 
 % eeshan bhatt
 
@@ -12,7 +14,7 @@ alpha_grey      = [0.6 0.6 0.6];
 set(0,'defaultAxesFontSize',13)
 
 %% load toby test data by event
-location = './data-tobytest-by-event/*.mat';
+location = '../data/data-tobytest-by-event/*.mat';
 listing = dir(location);
 num_listing = numel(listing);
 
@@ -96,12 +98,6 @@ for iNL = 4
         disp('toby test has eof OFF');
     end
     
-    OBJ_EOF = eb_load_eeof_file('eeof_itp_Mar2013.nc');
-    weights = [-10 -9.257 -1.023 3.312 -5.067 1.968 1.47].';
-    
-    ssp_estimate = OBJ_EOF.baseval + (OBJ_EOF.eofs * weights).*eof_bool;
-    [RT,ZT,TL] = run_pe(ssp_estimate,OBJ_EOF.depth);
-    
 end
 
 %% tetradic color wheel
@@ -124,7 +120,7 @@ end
 
 
 %% figure locations in x,y
-subplot(6,3,[3 6]);
+subplot(6,3,[2 5]);
 hold on
 for nx = 1:num_events
     plot([rx_x(nx) tx_x(nx)],[rx_y(nx) tx_y(nx)],'color',[alpha_grey alpha_color],'linewidth',7,'HandleVisibility','off');
@@ -166,7 +162,7 @@ title([event(1).tag.tstr ' to ' event(end).tag.tstr],'fontsize',lg_font_size+1);
 
 %% figure for contacts in z
 
-subplot(6,3,[2 5])
+subplot(6,3,[1 4])
 hold on
 for nz = 1:num_events
     plot([0 1 ],[tx_z(nz) rx_z(nz)],'-','color',[alpha_grey num_events./5.*alpha_color],'linewidth',7)
@@ -222,8 +218,52 @@ set(gca,'ydir','reverse')
 yticks([0 20 30 90]);
 title([event(1).tag.name ' : ' num2str(length(event)) ' contacts'])
 
+%% figure: gvel -- timeline
+subplot(6,3,[3 6])
+hold on
+for utr = unique_tag_rx
+    index = find(tag_rx == utr);
[…]

### 2021-01-01 src/h_set_xy_bounds.m
helper function for plotting limits

@@ -1,0 +1,16 @@
+%% helper function : set_xy_bounds(x1,x2);
+function [] = h_set_xy_bounds(x1,x2,y1,y2)
+% h_set_xy_bounds --- sets xlim and ylim based on inputs
+
+x_std = std([x1(:); x2(:)]);
+y_std = std([y1(:); y2(:)]);
+
+min_xval = min([min(x1(:)) min(x2(:))])-x_std/5;
+max_xval = max([max(x1(:)) max(x2(:))])+x_std/5;
+
+min_yval = min([min(y1(:)) min(y2(:))])-y_std/5;
+max_yval = max([max(y1(:)) max(y2(:))])+y_std/5;
+
+xlim([min_xval max_xval]);
+ylim([min_yval max_yval]);
+end

### 2021-08-20 doc/main.tex
benchmark commit

@@
-\title[JASA/Sample JASA Article]{In situ estimation of sound speed profiles and horizontal group velocities for range estimation between under ice acoustic modems}%{Considering vertical sound speed uncertainty in the Beaufort Lens for more realistic range estimation between acoustic modems}
+\title[JASA/draft]{Real-time Acoustic Ray Range Estimation for Underwater Vehicle Navigation}
 \author{EeShan C. Bhatt}
-\email{eesh@mit.edu}
+\email{ebhatt@whoi.edu}
 \affiliation{MIT-WHOI Joint Program in Oceanography/Applied Ocean Science \& Engineering, Cambridge and Woods Hole, MA, USA}
@@
 \begin{abstract}
-
-{\color{red}[Draft Abstract]}
-
-Put your abstract here. Abstracts are limited to 200 words for
-regular articles and 100 words for Letters to the Editor. Please no
-personal pronouns, also please do not use the words ``new'' and/or
-``novel'' in the abstract. An article usually includes an abstract, a
-concise summary of the work covered at length in the main body of the
-article.
+Almost all underwater navigation paradigms rely on the conversion of recorded travel time into range to compute a trilateration solution.
+For real-time navigation, this conversion is treated deterministically.
+For positioning in post-processing, more computationally and/or labor intensive modeling methods are often employed to reduce the uncertainty driven by multipath arrivals.
+For an autonomous underwater vehicle deployment in March 2020 in the Beaufort Sea, the total under-ice environment and complex acoustic propagation via the Beaufort Lens necessitated the real-time integration of model-based data processing.
+An embedded, stochastic, ray-based prediction of the horizontal group velocity achieved a ranging mean absolute error of 11 m compared to GPS-aided beacon-to-beacon communication events.
+In post-processing, a multipath adjustment achieved ranging performance that rivals GPS.
+To our knowledge, these results are the first field experiment to demonstrate a real-time, physics-based model to constrain ranging error in underwater navigation.
 \end{abstract}
@@
 % =========================================================================== %
+
 \section{\label{sec:1} Introduction}
 
-{\color{red}
-[Intro : EOF Approach, generally]
-
-[Motivation: ICEX]
-}
-
-This sample document demonstrates the use of JASA in manuscripts
-prepared for submission to the Journal of the Acoustical Society of America.
-
-The paper is organized as follows: Section~\ref{sec:2} presents
-initial information, while
-Section~\ref{sec:3} presents examples of mathematical expressions.
+The paper is organized as follows: Section~\ref{sec:2} presents the physical
+experimental setup and data processing simplifications; Section~\ref{sec:3} presents key results for group velocity predictions and range estimations; Section~\ref{sec:4} closes with a discussion.
 
@@
 
-\subsection{\label{subsec:2:1} Environmental EOFs}
+The operational solution for ICEX20 relied on an Integrated Communication and Navigation Network (ICNN) \citep{schneider_self-adapting_2020,randeni_construction_2020,randeni_high-resolution_2021}.
+The ICNN was initially developed via numerous virtual experiments, to push the boundaries of algorithms and interfaces between different hardware components.
+The simulation approach is a necessary testbed for robustness and produces better results than post-processing previous field data, as that restricts mission configurations to the data taken, which can hamper troubleshooting in the field.
+The simulation capabilities are largely physics-driven with a modular system of systems approach: an environmental simulator with sub-components for the ocean, Arctic ice, and ocean acoustics; a vehicle simulator with sub-components for vehicle dynamics and navigation; a topside hardware simulator and acoustic communications simulator, both with a software and hardware-in-the-loop version \citep{schneider_netsim_2018}.
+The virtual environment similarly emulates the interfaces between the real components to test the entire software pipeline.
+Both simulation capabilities are integral to mission success.
 
-{\color{red}[DETAIL : EOF Approach]}
+\subsection{Operational paradigm on the ice}
 
-% ================================================== %
-\begin{figure}[ht]
-\figline{\fig{figsamp.jpg}{\reprintcolumnwidth}{(a)}}
-\figline{\fig{figsamp.jpg}{4cm}{(b)} \fig{figsamp.jpg}{4cm}{(c)}}
+The ICNN is comprised of four ice buoys, loosely in a rectangle, roughly 2 km away from a central ice camp with a topside computer, as shown in Fig.~\ref{fig:icnnOverview}.
 
-% \includegraphics[width=\reprintcolumnwidth]{figsamp.jpg}
-\caption{\label{fig:FIG1}{Extracting EOFs from environmental data: (a) HYCOM and ITP data, (b) EOF shapes and weight distributions.}}
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=\reprintcolumnwidth]{figs/ICNN_Overview}
+	\label{fig:icnnOverview}
+	\caption{CAPTION}
+\end{figure}
 
+Each ice buoy is outfitted with a WHOI Micro-Modem \citep{singh_underwater_2006}, with a four-element receiver array and a single transmitter, and one-tenth of a millisecond resolution.
+The receiving and transmitting elements were split between shallow and deep depths (30 and 90 m, respectively) to provide better coverage due to the acoustic shadow zone in the Beaufort Lens.
+The buoys do not encompass the full operating range of the vehicle but are positioned to minimize overlap in trilateration for spherical positioning\citep{deffenbaugh_relationship_1996}.
+
+Vehicle missions in any underwater environment, but especially acoustically complex ones, must balance competing uses of the acoustic channel.
+Thus this network uses a single synchronized digital communication packet to provide both tracking and data to the operator.
+The ICNN operational paradigm is as follows:
+\begin{enumerate}
+\item The AUV, running an ice-tracking DVL and an onboard hydrodynamic model, broadcasts its perceived location on a scheduled, time-synchronized message via WHOI Micro-Modem
+\item Four ice buoys, each outfitted with a WHOI Micro-Modem, receive messages from the AUV and send that information over freewave radio to a Topside computer
+\item The topside computer converts travel times into range estimates using a stochastic embedded prediction of the horizontal group velocity via BELLHOP ray tracing code \citep{porter_bellhop_2011}
+\item The topside computer calculates a new position by trilaterating the range estimates
+\item The position differential is broadcast to the vehicle to update its navigation solution
+\end{enumerate}
+
+This paper specifically addresses the third step\textemdash the topside computer's conversion of travel time into range, which underpins the trilateration in the ICNN, and is an often overlooked element of any underwater acoustic navigation system.
+The results accordingly focus on validating acoustic range estimates from communication events between GPS-tracked beacons.
+This ``modem experiment'' was conducted while the vehicle was still being prepped to go under the ice.
+
+\subsection{Modem experiment design}
+
+Because the navigation solution on the vehicle, during a mission, can only be evaluated on the basis of the error estimates sent, a sister experiment for validating the real-time ranging approach was implemented.
+The ice buoy modems were run as ``virtual vehicles'' at a given depth and would transmit to all other ice buoys.
+The internal sound speed estimate was then toggled using a handful of weights applied to a basis function representation of the sound speed.
+This trial was configured for as many ice buoy modems as possible, with each experiment running for roughly an hour between the two sound speed modalities.
+There are three important engineering choices in the ICNN network to contextualize this dataset:
+\begin{enumerate}
+\item Each modem buoy only has one transmit depth layer (30 or 90 m)
+\item The transmit and receive layers are independent
+\item At any given time, the receive depth layers are the same across all four ice buoys
+\end{enumerate}
+
+The design enables a self-adapting network to transmit and receive at the optimal depth to maintain contact with the AUV \citep{schneider_self-adapting_2020}.
+This adaptivity was not used during the modem experiment but was instead toggled manually.
+Given the complexity of the ICNN system, this experiment did not collect an exhaustive set of data across all buoy, source depth, receive layer, and sound speed combinations, as this time was also used to verify and resolve any hardware issues in the modems themselves.
+
+Importantly, only one modem at a time could run the vehicle behavior to trigger group velocity calculations; other modems that were transmitting simultaneously recorded events as modems, not as vehicles.
+
+The dataset introduced in this thesis has parallel physical and virtual layers; given the small operational scale, on the order of kilometers, and a singular SSP, both layers are projected into a range independent coordinate system for ease of comparison.
+
+\subsubsection{Physical layer}
+The physical layer, i.e.
+the physical link through the real ocean, describes the beacon to beacon communication events through the real water column; it covers all source depths (20, 30, and 90 m) and both receive layers (30 and 90 m), where the surface expressions of the beacons are tracked by GPS.
+Events are sorted by source depth and sound speed profile given that this is the dominating factor for the location and span of the shadow zone compared to any temporal or spatial scales of variability.
+
+\begin{figure}[h!]
+  \centering
+  \includegraphics[width=.50\textwidth]{figs/modem-chart.pdf} \hfill
+  \includegraphics[width=.48\textwidth]{figs/zs20-birdseye.pdf} \\
+  \vspace{1em}
+  \includegraphics[width=.48\textwidth]{figs/zs30-birdseye.pdf} \hfill
+  \includegraphics[width=.48\textwidth]{figs/zs90-birdseye.pdf} \\
+  \caption{An overview of the modem experiment by source and receiver depth and position.
+Subfigure (a) shows a chart of the successful events between source and receiver depth.
+The black column on the left, $tx$, shows the source depth, $z_s$.
+The column on the right, $rx$, shows the receivers with the amount of good contacts.
+The orientation of the triangles\textemdash sideways, upwards, and downwards\textemdash corresponds to depths of 20, 30, and 90 m.
+Subfigures (b-d) correspond to receiver depths of 20, 30, and 90 m, respectively.
+Camp Seadragon, demarcated by the purple triangle, is situated just north of the continental shelf.}
+  \label{fig:overview}
   \end{figure}
-% ================================================== %
 
-This section would generally discuss EOFs, the equations that extract them from data collections, and perhaps the "information capacity" w.r.t.
-to the number of depth points vs number of samples.
-Should we also include, in somewhat general terms, the "uniqueness" constraint?
-As in, i.e.: sampled region is too large or too small to adequately capture the features needed for reconstruction.
+\clearpage
+
+\subsubsection{Virtual layer}
+The virtual layer describes the BELLHOP simulations for each communication event in the physical layer, where a handful of weights applied to a basis function representation can reconstitute any number of candidate SSPs \citep{bhatt_embedded_2021}.
+The SSP is the only configurable input to the group velocity prediction.
+
+During ICEX20, the SSP was toggled between the historical``baseline'' and a ``weighted'' estimate based on a CTD, i.e.
+an optimal set of weights chosen by a human decision maker via an interactive Tactical Decision Aid framework \citep{bhatt_embedded_2021}.
+For the post-processing analysis, we add the SSP from HYCOM for comparison, to mirror the information available on a submarine\textemdash modeled data, historical data, and in situ data.
+This order corresponds to increasing ducted conditions of the Beaufort Lens.
+
+Fig. \ref{fig:raytrace-zs30} shows eigenrays in a range independent coordinate system for all three sound speed environments for a source depth of 30 m.
+It is important to note that BELLHOP produces an abundance of eigenrays; the ones displayed here are chosen by travel time proximity to the recorded data.
+An important pattern is the increased complexity in multipath as ducted conditions increase.
+
+\begin{figure}[h!]
+  \centering
+  \includegraphics[width=\reprintcolumnwidth]{figs/raytrace-3env-zs-30.pdf}
+  \caption{Eigenrays for beacon to beacon events for each nominal source depth of 30 m.
+The beacons are highlighted in color/marker coding in figure \ref{fig:overview}.
+The eigenrays are curated from BELLHOP by travel time proximity and are traced in the representative colors over a total ray fan in gray.}
+  \label{fig:raytrace-zs30}
+  \end{figure}
+
+\FloatBarrier
+\subsection{Group velocity prediction methods}
+
+The variation of horizontal group velocity from any source-receiver pair is the fundamental challenge to implement GPS-like navigation in an LBL navigation paradigm, especially in an acoustically complex propagation environment.
+The use of the acoustic modem network for tracking relies on the accurate estimates of travel times between the submerged platform and range modems, supported by clock synchronization and a pre-determined scheduling of acoustic events.
+For the Beaufort Lens, the strong multipath effects make it virtually impossible to deterministically predict the modem triggering time.
+
+\subsubsection{Minimal bounce criteria}
+Instead, for each individual node $i$, an embedded stochastic tracking framework is used to provide a running estimate of the horizontal group velocity $u_{i,j}$ for the conversion from travel time to range from modem $j$, with the ultimate goal of matching the naive group velocity, i.e.
+the GPS-recorded distance between two nodes divided by the modem-recorded one way travel time between them.
+
+In the ICEX20 configuration, the acoustic tracking is running on the topside computer, which controls the integrated communication and navigation range.
+Here we assume that the group velocities $u_{i,j}$ are smoothly varying over the course of a vehicle mission, i.e., with respect to range, mission time, and the frequency of updates relative to vehicle motion.
+The group velocity is continuously tracked using predictions from the \textit{Virtual Ocean} infrastructure.
+
+When the topside tracking framework receives a modem message, with a time delay, $\Delta t$, from one of the range modems, it will request a new estimate of the group velocity and its associated uncertainty.
+The group velocity estimate is found using the extrapolated navigation for range, $\hat{r}$, and an impulse response estimate received in the form of ray travel times $dt_{j}$ and amplitudes $a_{j}$ for that range and depth.
+
+The initial call to BELLHOP is over a sparse, dynamically created depth and range grid, of which the plane wave approximation is used to bring the ray to the range and depth posited by the onboard tracking solution.
+Using only the $N_0$ rays with neither surface nor bottom bounces, it will then estimate the current group velocity $u$ from a power weighted average of the ray travel times,
+\begin{equation}
+u = \frac{\hat{r} \sum_{n=1}^{N_{0}} a_{n}^{2}}{\sum_{n=1}^{N_{0}} dt_{n}a_{n}^{2}} ~, 
+\end{equation}
+and the associated weighted standard deviation,
+\begin{equation}
+\sigma_{u} \simeq \sqrt{\frac {\sum_{n=1}^{N_{0}} (dt_{n}-\hat{r}/u)^{2}a_{n}^{2}}{ \sum_{n=1}^{N_{0}} a_{n}^{2}} } \frac{u^{2}}{\hat{r}}
+\end{equation}
+If no direct paths exist, i.e. $N_{0}=0$, then the group velocity is computed using the same algorithm for the ray arrivals with one bounce, and so on.
+
+Finally, the psuedorange is calculated simply as
+\begin{equation}
+r_{i,j} = u_{i,j} \Delta t_{i,j} 
+\end{equation}
+
+This stochastic method for group velocity calculation can run in real-time, appearing to be orders of magnitude faster than general post-processing methods which seek to determine the specific ray itself that best matches a prominent indicator from the arrival structure.
+The method assumes direct path arrival, where multipath over the shallow receiver depth is trivial for the group velocity calculation; thus we call it the minimal bounce criteria (MBC).
+The BELLHOP simulation that runs this calculation uses 3600 rays with launch angle fan of -60 to 60 degrees.
+However, this high density of launch angles creates a greedy estimator of direct path eigenrays, and this simulation pipeline almost always guarantees a direct path found.
+The MBC improves upon a deterministic estimate in that it considers the ray path, travel time, and amplitude to predict a realistic group velocity.
+
+\subsubsection{Nearest Bounce Criteria}
+
+As shown in the eigenray traces, an acoustic arrival does not always take the direct path from source to the receiver, and the path it takes is extremely dependent on the sound speed profile.
+Often, BELLHOP will produce a number of arrivals resulting from multiple propagation paths for any source-receiver pair.
+The suggested algorithm, the nearest bounce criteria (NBC), is only a slight modification from the minimal bounce criteria, and includes multipath as a new dimension of information with negligible additional computation.
+
+Given a running estimate for the horizontal group velocity $u_{i,j}$ between nodes $i$ and $j$, the navigation system has an extrapolated value for range, $\hat{r}$, in the virtual layer, and a recorded travel time, $\Delta t_{i,j}$, in the physical layer.
+Instead of using only the $N_0$ rays with neither surface nor bottom bounces to estimate group velocity, we solve for the power weighted average of the ray travel time for the $N_k$ rays with $k$ bounces,
+\begin{equation}
+t_k = \frac{\sum_{n=1}^{N_{k}} dt_{n}a_{n}^{2}}{\sum_{n=1}^{N_{k}} a_{n}^{2}} ~, 
+\end{equation}
+find the nearest matching power weighted average to recorded travel time,
+\begin{equation}
+t_{i,j,k} = \min_{k=0,1,2,...} \left| t_k - \Delta t_{i,j} \right|
+\end{equation}
+predict a group velocity,
+\begin{equation}
+u_{i,j} = \dfrac{\hat{r}}{t_{i,j,k}}
+\end{equation}
+and estimate the range as was done previously.
+\begin{equation}
+r_{i,j} = u_{i,j}\Delta t_{i,j}
+\end{equation}
+
+This method selects a different group velocity based on the multipath arrival structure, as the expected arrival is not always the first arrival and could even be masked by noise or blocked temporarily \citep{deffenbaugh_acoustic_1996}.
 
@@
 % =========================================================================== %
-\section{\label{sec:3} Results}
 
-\subsection{\label{subsec:3:1} ICEX 2020 conditions}
-{\color{red}[CTD comparison]}
+\FloatBarrier
+\section{\label{sec:4} Real-time psuedorange results}
 
-% ================================================== %
-\begin{figure}[ht]
-\includegraphics[width=\reprintcolumnwidth]{figsamp.jpg}
-\caption{\label{fig:FIG2}{CTD casts vs. HYCOM-derived baseline}}
+The modem experiment generated 811 beacon to beacon communication events with their own real-time NBC group velocity predictions.
+As expected, the algorithm is generally overestimating range as it resolves a direct path that does not represent the actual arrival.
+This section examines the absolute range error between the psuedorange and the GPS derived range with the caveat that the data collected is fairly asymmetrical across source depths and sound speed conditions.
+
+\begin{table}[h!]
+\renewcommand{\arraystretch}{1.5}
+\centering
+\begin{tabular}{r|ccccc}
+SSP & Count & Mean [m] & Median [m] & STD [m] & Max [m] \\ \hline
+Baseline & 243 & 11.38 & 11.96 & 4.23 & 23.95 \\ 
+Chosen weights & 568 & 11.36 & 11.40 & 8.12 & 26.0 \\
+\end{tabular}
+\caption[Range estimation error for \textit{in situ} events]{The range estimation error for \textit{in situ} events. The error here is presented as the absolute, not directional, error.}
+\label{tab:rangeErrorInSitu}
+\end{table}
+
+Table \ref{tab:rangeErrorInSitu} breaks down the number of events, mean, median, standard deviation, and maximum absolute range error for the \textit{in situ} calculations.
+The mean and median for both sound speeds are similar on the order of centimeters; the chosen weights show a maximum greater than that of baseline, but it is from a OWTT event that has no comparison in the baseline data set.
+The overarching statistical results show that both sound speed environments, given the smaller operational scales and shallow depths, are accurate enough to approximate the group velocity structure given a local grid.
+Of course, these statistics obscure the relationship between range error and range (or OWTT as the proxy for range).
+Analysis shows that where there is overlap in range between sound speed conditions, the range error difference is no more than a few meters.
+This is likely driven by the prominence of the duct.
+
+Where there is overlap in the physical layer, i.e. in OWTT, the range error driven by the two virtual layers is no more than a few meters.
+This is best demonstrated by all source and receiver depth pairings, except for a source at 30 m and a receiver at 90 m, where the baseline is clearly visibly closer to no range error than the weighted environment.
+This may be driven by the prominence of the duct, but it is certainly occluded by the order of magnitude difference between the number of events in each virtual layer.
+
+\FloatBarrier
+\section{\label{sec:5} Post-processing psuedorange results}
+
+Due to the operational constraint of only one modem running the vehicle behavior at a time, the previous analysis has looked at 811 total beacon to beacon events, where 243 use the baseline SSP and 568 use the weighted SSP.
+While overall statistics are presented, and individual events are compared for the same SSP between the two algorithms, the events themselves are not equally distributed with respect to depth, range, and sound speed status.
+
+The analysis is separated by source depth, and where possible, receiver depth, as the group velocities across source and receiver depths may be different.
+
+There are 1242 total beacon to beacon events in the modem experiment that would have been accepted by the ICNN for real-time ranging and navigation had the vehicle behavior been running on all modems.
+The value for the extrapolated range, $\hat{r}$, is only tracked when the modem runs the vehicle behavior; thus we replace $\hat{r}$ with the GPS-tracked range for all modem events.
+Naturally, because $\hat{r}$ converges to the correct solution, a comparison of $\hat{r}$ with the GPS-tracked range shows a normal, zero-centered distribution within the bounds of GPS drift.
+The analysis here therefore seeds ``omniscient'' knowledge of the range and leverages the post-processing pipeline to more thoroughly evaluate the acoustic range estimate for all modem events, with all three sound speed sources, and both group velocity criterion.
+Thus, the results in this section evaluate the utility of the algorithms and sound speed sources, divorced from their role in the ICNN.
+We first compare the post-processing results between the two algorithms, and then take a step-by-step approach to examine how the nearest bounce algorithm works in more detail.
+
+\FloatBarrier
+\subsection{Group velocity prediction}
+
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{figs/SLIDES-gvel-comparison-wdata-30.pdf}
+\caption{\label{fig:gvel30}{Caption here.}}
 \end{figure}
-% ================================================== %
 
-% ========================= %
-% ========================= %
-\subsection{\label{subsec:3:2} One-Way Travel Times}
-{\color{red}[OWTT comparison]}
+\subsection{Psuedorange estimations}
 
-% ================================================== %
-\begin{figure}[ht]
-\includegraphics[width=\reprintcolumnwidth]{figsamp.jpg}
-\caption{\label{fig:FIG3}{Modem placement diagram}}
+\subsubsection{Comparing the nearest and minimal bounce criteria}
+
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{figs/compare-methods-postv1v2.pdf}
+\caption{\label{fig:gvel30}{Caption here.}}
 \end{figure}
-% ================================================== %
 
-% ================================================== %
-\begin{figure}[ht]
-\figline{\fig{figsamp.jpg}{\reprintcolumnwidth}{(a)}}
-\figline{\fig{figsamp.jpg}{4cm}{(b)} \fig{figsamp.jpg}{4cm}{(c)}}
-% \includegraphics[width=\reprintcolumnwidth]{figsamp.jpg}
-\caption{\label{fig:FIG4}{(a) OWTT histogram, (b) "..." OF RANGE "...".}}
+\FloatBarrier
+\subsubsection{Psuedorange error versus one way travel time}
+
+If there was an assumed deterministic group velocity, the range error would grow with OWTT proportionally to the difference between the estimated and actual value.
+Thus, an indication that the algorithm is adapting group velocity correctly is that range error does not strictly increase with OWTT.
+
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{figs/range-error-owtt-newalgorithm.pdf}
+\caption{\label{fig:gvel30}{Caption here.}}
 \end{figure}
-% ================================================== %
+
+\FloatBarrier
+\subsection{GPS sensor drift}
+
+Systemic GPS drift drives the severe banding found in comparisons of range estimations via both group velocity algorithms, with those bands showing 100\% positive correlation between methods and being a similar size in meters.
+Thus, while this thesis is primarily motivated by achieving similar accuracy and precision of GPS, the impressive results from acoustic ranging necessitate a discussion about the limitations of GPS, particularly at polar latitudes.
+
+The ``Global Positioning System Standard Positioning Service Performance Standard'', published by the Department of Defense in Apr 2020 (CITE GPS TECHNICAL STANDARD) claims that well-designed GPS receivers have been achieving a horizontal accuracy of 3 meters and a vertical accuracy of 5 meters, 95\% of the time.
+But ``well-designed'' is an ambiguous enough term that obfuscates two confounding factors for high latitude GPS drift\textemdash operational limitations and ionospheric scattering.
+
+% operational
+There are two metrics that demonstrate the operational limitations of GPS tracking technologies\textemdash fix rate and positional dilution of precision (PDOP) \citep{swanlund_gps_2016}.
+Fix rate looks at the ratio of successful fixes to overall attempts in acquiring a location.
+In animal tracking studies in polar latitudes, fix rates have significantly improved in the last twenty years, from as low as 26\% in 1996\citep{moen_effects_1996} to 87.2\% in 2018 \citep{jung_accuracy_2018}.
+The latter study showed a precision of 4.3 $\pm$ 4.0 m and an accuracy of 6.0 $\pm$ 4.7 m.
+While significant metric to evaluate GPS robustness, accuracy and precision, not the fix rate, explains the severe banding observed; fix rate bias, where various aspects of a landscape can induce bias by impeding the GPS receiver from acquiring its location, is not relevant given the static nature and small time window of ICEX20.
+
+PDOP looks at the effect of satellite geometry on location precision, as satellite orbits are designed to optimize performance for human population density.
+A high PDOP value indicates a poor geometric configuration, such as satellites being tightly clustered above a GPS receiver, that limits the reliability of trilateration.
+The literature shows correlation between PDOP values and location error \citep{swanlund_gps_2016}.
+PDOP is generally thought of as three-dimensional positioning and can be broken down into the horizontal and vertical components, HDOP and VDOP, respectively.
+The low elevation angles in polar areas are good for horizontal positioning but are susceptible to poorer vertical positioning.
+Overall this contributes to higher noise level in observations than usually seen in most global applications.
+The sparse and limited infrastructure in the Arctic make it ideal for satellite-based navigation, with some research proposing new satellite systems to accommodate increased interest in the region \citep{reid_gnss_2016}.
+
+However, both of these metrics are functions of the satellite constellation available to any given GPS receiver.
+The Arctic area is beyond the reach of many geostationary satellites, where they are low on the horizon and only visible for brief periods.
+Given that their positions are tracked by stations on Earth, and their timing is updated twice a day via atomic clock, a reasonable hypothesis for bands of GPS error are different and/or updated satellite connections.
+
+% ionospheric
+The second confounding factor is the ionosphere, the layer of the Earth's atmosphere roughly 80 to 1000 km above the surface that contains a high concentration of ions and free elections. 
+In the ionosphere, electromagnetic signals are affected by the total amount of electrons encountered by the signal; the size of the signal delay is dependent on the carrier frequencies.
+The normal signal delay creates a GPS error of 5 to 15 meters during the daytime and 1 to 3 meters at night (CITATION).
+The literature even references the total amount of electrons encountered by the ``GPS raypath'' as a means of explaining signal fluctuations \citep{themens_nature_2015}.
+This variability is observed by seasonal and solar cycle variabilities,\citep{themens_nature_2015} severe ionospheric storms\citep{mitchell_gps_2005}, and auroral activity (increased electron precipitation)\citep{jin_gps_2014,gwal_gps_2011}.
+In fact, the \textit{aurora borealis} was observed on multiple nights during ICEX20.
+
+In a humbling twist, at polar latitudes, GNSS systems experience similar operational, technical, and environmental challenges that we observe in underwater acoustic communications; the underwater acoustics community generally seems to ignore these challenges when validating localization and navigation efforts.
+
+Figure \ref{fig:gps-drift-example} highlights the GPS and observed OWTT drift relative to the ice movement for two pairs of modem buoy connections.
+The top two panels show GPS drift in both the $x$ and $y$ directions relative to the median $x$ and $y$ distance recorded between the two buoys.
+The bottom two panels indicate the GPS drift as $\delta R = \sqrt{\delta x^2 + \delta y^2}$ and temporal drift, $\delta t$, relative to the median OWTT recorded between the two modems.
+Here we use the median, not the mean, to be more robust to outliers in the dataset.
+The dashed line is scaled by a group velocity of 1440 m/s, such that if there was no GPS drift, all events should exist on the line.
+
+The left hand side of the plot shows the connections between the North and East buoys.
+The top shows consistent change in distance with respect to time, as the $\delta X$ measurement trends from 2 to -2 and the $\delta Y$ measurement trends from -1 to 1 meters.
+This indicates that the ice movement between the North and East buoys is non-rigid, i.e., they are actually moving relative to each other.
+The width of the scatter plot is a decent indicator for the precision of the GPS signal, as it remains consistent across many colors of scattered points, which correspond to similar regions in time.
+The idea that these buoys are moving relative to each other is corroborated in the bottom plot, where some events exist on the dashed line indicating correlation between GPS movement and OWTT change.
+Some of the offsets between the vertical bands relate to different operational configurations of source and receiver depths.
+However, the large majority of events show vertical banding for the same nominal $\delta t$, indicating the amount of GPS drift.
+
+This idea of GPS drift relative to the same OWTT measurements is also indicated by events between the other two buoys, South and West, on the right hand side.
+The GPS drift, on the top, shows a point cloud centered around the origin for events four hours and 13 hours after the experiment origin; thus these buoys are moving in a more rigid ice floe.
+There is minimal impact by source and receiver depth on the spread of OWTT.
+The GPS drift relative to the OWTT, on the bottom, shows a large amount of GPS drift for very little OWTT drift.
+The OWTT drift observed here is sensitive to the acoustic scattering, multipath, or environmental microstructure.
+
+These are two subsets of the physical links, they cover all four GPS modem buoys except for Camp.
+The GPS at camp, as shown in figure \ref{fig:iceFloe}, is the least accurate due to the human activity and infrastructure occluding the physical puck.
+A more comprehensive look at all physical link pairs is in Appendix \ref{app:beacon2beacon}.
+
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=\textwidth]{figs/gps-drift-example} 
+	\caption[Examination of GPS and modem drift]{A comparison of GPS drift (top) colored by hours since the experiment origin and GPS drift against OWTT spread (bottom) demarcated by source depth and receiver depth.
+The physical link between North and East are shown on the left; South and West is on the right.
+}
+	\label{fig:gps-drift-example}
+\end{figure}
 
@@
 
-\subsection{\label{subsec:4:1} Model agreement discussion}
-{\color{red}[Model agreement discussion]}
+Range estimation is an essential step of acoustic localization and navigation.
+This work solely focuses on mitigating range estimation error via a stochastic prediction of the group velocity, knowing that this can have an outsized benefit on minimizing trilateration error.
+We hypothesize and validate that the group velocity, especially for mission durations, is smoothly varying in range and multipath structure, and sensitive to source and receiver depth.
+We employ a GPS-tracked experiment to have a ground truth comparison for real-time localization algorithms.
+
+This work, to our knowledge, is the first successful field deployment of a real-time, ray-based acoustic ranging system for underwater navigation.
+The methods here demonstrate remarkable ranging accuracy and precision for the Beaufort Lens, an acoustically complex environment.
+The \textit{in situ} prediction via BELLHOP balances the tension in simulation fidelity versus capturing the overarching ensemble physics to approximate a local group velocity.
+
+Furthermore, the improved nearest bounce criteria, validated in a high fidelity post-processing pipeline, produces range estimates that rival GPS accuracy and outclass GPS precision, especially at polar latitudes.
+It transforms multipath, widely considered as an obstacle for acoustic ranging, into a new information content to refine ranging accuracy.
+
+The scalability of the methods provided here for long-range applications remain to be seen.
+Given the NBC algorithm's improvement with increased multipath, its effectiveness is likely only challenged by the valid operational scales of a range independent propagation environment.
+Even then, the dominant SSP amongst the variability may very well produce GPS-like accuracy.
+The NBC algorithm can be easily be retooled to discriminate between surface and bottom bounces, and employ simple statistical learning or more advanced machine learning techniques to select an appropriate representative group velocity.
+Some more advanced vehicle intelligence can toggle use the range uncertainty to toggle between a more conservative group velocity estimate and the NBC, which is intrinsically quite precise.
+
+There are many avenues that this approach can be further refined and tested for field operations.
+Amongst them is defining the uncertainty grid for BELLHOP via stochastic or data-driven measures such as the distance traveled by the AUV between ICNN updates or the magnitude of the position corrections by the ICNN.
+Another is to entirely remove the seeded range and instead rely on the submerged asset's depth and recorded OWTT to find high probability fields in range, as briefly indicated in Chapter 6.
+
+As the Arctic continues to become more accessible and navigable, safe navigation infrastructure and practices will be increasingly vital for industry, tourism, commercial, and military operations.
+This work contributes a new paradigm to embed environmentally adaptive behavior into AUV navigation that has demonstrated success in an acoustically complex and operationally challenging environment.
+Thus, we believe that this work enables more accurate range estimation, localization, and/or navigation for any field experiment given known source and receiver depths.
+
+(NEED TO SKIM CONCLUSION FOR MORE MATERIAL)
+
+\FloatBarrier
+\section*{Supplementary Figures}
+
+\clearpage
+\subsection*{Ice movement during the modem experiment}
+\begin{figure}[h!]
+\includegraphics[width=\textwidth]{figs/ice-floe-drift.pdf}
+\caption{\label{fig:iceFloeDrift}{Caption here.}}
+\end{figure}
+
+\clearpage
+\FloatBarrier
+\subsection*{Group velocity prediction by source depth}
+\begin{figure}[h!]
+\includegraphics[width=0.8\textwidth]{figs/SLIDES-gvel-comparison-wdata-20.pdf} \\
+\includegraphics[width=0.8\textwidth]{figs/SLIDES-gvel-comparison-wdata-90.pdf}
+\end{figure}
 
 % ================================================== %
-\begin{figure}[ht]
-\includegraphics[width=\reprintcolumnwidth]{figsamp.jpg}
-\caption{\label{fig:FIG5}{Comparison of range-independent model vs. field data}}
-\end{figure}
+% \begin{figure}[ht]
+% \includegraphics[width=\reprintcolumnwidth]{figsamp.jpg}
+% \caption{\label{fig:FIG1}{Caption here.}}
+
+% \raggedright
+% {\color{red}
+% Note: The only figure formats allowed are the following:
+% .pdf, .ps, .eps, or .jpg. Figure files must be named in this fashion:
+% Figure\#.xxx, where ``\#'' is the figure number and ``xxx'' is the file format
+% (Figure1.eps, Figure2.jpg, Figure3a.ps, Figure3b.ps, etc).
+% }
+
+% [For these sample pages we have used only figsamp.jpg for convenience]
+% \end{figure}
 % ================================================== %
 
-\subsection{\label{subsec:4:2} Bootstrapping EOF}
-{\color{red}[Bootstrapping EOF]}
+% \subsubsection{Sample subsubsection\label{subsubsec:1}}
 
-% ================================================== %
-\begin{figure}[ht]
-\includegraphics[width=\reprintcolumnwidth]{figsamp.jpg}
-\caption{\label{fig:FIG6}{Convergence}}
-\end{figure}
-% ================================================== %
+% \paragraph{Sample paragraph}Here is text following the paragraph
+% heading.
+% Here is a figure reference: is shown in Fig.~\ref{fig:FIG1}.
 
-% ================================================== %
-\begin{figure}[ht]
-\includegraphics[width=\reprintcolumnwidth]{figsamp.jpg}
-\caption{\label{fig:FIG7}{Bootstrap error estimate}}
-\end{figure}
-% ================================================== %
+% Normal journal cite: \citep{joursamp1},
+%  Book reference \citet{booksamp1},
+% Computer language documentation:
+% \citep{sampcode2}.
 
-% ================================================== %
-\begin{figure}[ht]
-\includegraphics[width=\reprintcolumnwidth]{figsamp.jpg}
-\caption{\label{fig:FIG1}{Caption here.}}
+% Every \verb+\citep+  or \verb+\citet{}+ will produce a citation and an entry in the
+% bibliography. Every citation must have a matching entry in the
+% bibliography
+% database file (\verb+\filename.bib+).
 
-\raggedright
-{\color{red}
-Note: The only figure formats allowed are the following:
-.pdf, .ps, .eps, or .jpg. Figure files must be named in this fashion:
-Figure\#.xxx, where ``\#'' is the figure number and ``xxx'' is the file format
-(Figure1.eps, Figure2.jpg, Figure3a.ps, Figure3b.ps, etc).
-}
+% Here is an example of algorithmic:
 
-[For these sample pages we have used only figsamp.jpg for convenience]
-\end{figure}
-% ================================================== %
-
-An example of another first-level Section with following example text that refers to subsections using
-\verb+\ref{subsec:XXX}+ ...  EXAMPLE: Some background in
-section~\ref{sec:2} and details  in subsection~\ref{subsec:2:1}.
-
-   \subsubsection{Sample subsubsection\label{subsubsec:1}}
-
-\paragraph{Sample paragraph}Here is text following the paragraph
-heading.
-Here is a figure reference: is shown in Fig.~\ref{fig:FIG1}.
-
-Normal journal cite: \citep{joursamp1},
- Book reference \citet{booksamp1},
-Computer language documentation:
-\citep{sampcode2}.
-
-Every \verb+\citep+  or \verb+\citet{}+ will produce a citation and an entry in the
-bibliography. Every citation must have a matching entry in the
-bibliography
-database file (\verb+\filename.bib+).
-
-Here is an example of algorithmic:
-
-\begin{algorithmic}
-\If {$i\geq maxval$}
-    \State $i\gets 0$
-\Else
-    \If {$i+k\leq maxval$}
-        \State $i\gets i+k$
-    \EndIf
-\EndIf
-\end{algorithmic}
+% \begin{algorithmic}
+% \If {$i\geq maxval$}
+%     \State $i\gets 0$
+% \Else
+%     \If {$i+k\leq maxval$}
+%         \State $i\gets i+k$
+%     \EndIf
+% \EndIf
+% \end{algorithmic}
 
 \begin{acknowledgments}
-This research was supported by  ...
+This research was supported by the Office of Naval Research and the NDSEG Fellowship.
+We are grateful to all our collaborators in ICEX20\textemdash Bluefin Robotics, Toby Schneider, Lee Freitag \textemdash
 \end{acknowledgments}

### 2021-08-20 src/slides_gvel_comparison.m
changed for all 3 depths

@@ -28,6 +28,9 @@ indValid = ~isnan(DATA.simGvel);
 % calculate RangeAnomaly
 DATA.rangeAnomaly = DATA.owtt .* DATA.simGvel - DATA.recRange;
 
+% DEPTH
+ZS = 90;
+
 %% load post-processing sim for v2
 listing = dir('./bellhop-gvel-gridded/csv_arr/*gridded.csv');
 
@@ -88,7 +91,7 @@ tiledlayout(1,2,'TileSpacing','none','Padding','compact');
 shapeBounce = {'o','x','s','^','d'};
 
 count = 0;
-for zs = [30]
+for zs = ZS
     index1 = DATA.sourceDepth == zs;
     
     for zr = [30 90]
@@ -136,18 +139,22 @@ for zs = [30]
         title(sprintf('receiver depth = %u m',zr),'fontsize',14,'fontweight','bold');
         %text(0.95,1445,sprintf('rx depth = %u m',zr),'HorizontalAlignment','left','VerticalAlignment','bottom','fontsize',12);
         
+        xlim auto
+        % xlim([0.9 2.26])
+        ylim([1419 1449])
+        xx = xlim();
+        
         if sum(index)>1
-            text(2.25,1448,sprintf('n = %u events',sum(index)),'HorizontalAlignment','right','VerticalAlignment','top','fontsize',12);
+            text(xx(2),1449,sprintf('n = %u events',sum(index)),'HorizontalAlignment','right','VerticalAlignment','top','fontsize',12);
         else
-            text(2.25,1448,sprintf('n = %u event',sum(index)),'HorizontalAlignment','right','VerticalAlignment','top','fontsize',12);
+            text(xx(2),1449,sprintf('n = %u event',sum(index)),'HorizontalAlignment','right','VerticalAlignment','top','fontsize',12);
         end
         grid on
         
-        xlim([0.9 2.26])
-        ylim([1423 1448])
+
         
         % add in manual indicator for bottom bounce events for panel 1
-        if zr == 30
+        if ZS == 30 && zr == 30
             indManual = find(T1{4}.gvel < 1000);
             hold on
             plot(DATA.owtt(indManual),1424,'.','handlevisibility','off','markersize',15,'color',[colorSet{4}]);
@@ -172,6 +179,7 @@ nexttile(1);
 
[…]

### 2021-08-20 doc/main.tex
most figs, needs discussion, fleshing out sections, and intro

@@
 
-The ICNN is comprised of four ice buoys, loosely in a rectangle, roughly 2 km away from a central ice camp with a topside computer, as shown in Fig.~\ref{fig:icnnOverview}.
+The ICNN is comprised of four ice buoys, in a loose rectangle, roughly 2 km away from a central ice camp with a topside computer, as shown in Fig.~\ref{fig:icnnOverview}.
 
@@
 Each ice buoy is outfitted with a WHOI Micro-Modem \citep{singh_underwater_2006}, with a four-element receiver array and a single transmitter, and one-tenth of a millisecond resolution.
-The receiving and transmitting elements were split between shallow and deep depths (30 and 90 m, respectively) to provide better coverage due to the acoustic shadow zone in the Beaufort Lens.
-The buoys do not encompass the full operating range of the vehicle but are positioned to minimize overlap in trilateration for spherical positioning\citep{deffenbaugh_relationship_1996}.
+The receiving and transmitting elements were split between shallow and deeper depths (30 and 90 m, respectively) to provide better coverage due to the acoustic shadow zone in the Beaufort Lens.
+The buoys do not encompass the full operating range of the vehicle but are positioned to minimize overlap in trilateration for spherical positioning \citep{deffenbaugh_relationship_1996}.
 
 Vehicle missions in any underwater environment, but especially acoustically complex ones, must balance competing uses of the acoustic channel.
-Thus this network uses a single synchronized digital communication packet to provide both tracking and data to the operator.
+This network uses a single synchronized digital communication packet to provide both tracking and data to the operator.
 The ICNN operational paradigm is as follows:
@@
 
-The dataset introduced in this thesis has parallel physical and virtual layers; given the small operational scale, on the order of kilometers, and a singular SSP, both layers are projected into a range independent coordinate system for ease of comparison.
+The dataset introduced in this thesis has parallel physical and virtual layer; given the small operational scale, on the order of kilometers, and a singular SSP, both layers are projected into a range independent coordinate system for ease of comparison.
 
@@
 The physical layer, i.e.
-the physical link through the real ocean, describes the beacon to beacon communication events through the real water column; it covers all source depths (20, 30, and 90 m) and both receive layers (30 and 90 m), where the surface expressions of the beacons are tracked by GPS.
-Events are sorted by source depth and sound speed profile given that this is the dominating factor for the location and span of the shadow zone compared to any temporal or spatial scales of variability.
+the acoustic link through the real ocean, describes the beacon to beacon communication events through the water column; it covers all source depths (20, 30, and 90 m) and both receive layers (30 and 90 m), where the surface expressions of the beacons are tracked by GPS.
+Events are sorted by source depth and sound speed profile, given that these are the dominating factor for the location and span of the shadow zone compared to any temporal or spatial scales of variability.
 
@@
 \subsubsection{Virtual layer}
-The virtual layer describes the BELLHOP simulations for each communication event in the physical layer, where a handful of weights applied to a basis function representation can reconstitute any number of candidate SSPs \citep{bhatt_embedded_2021}.
+The virtual layer describes the BELLHOP simulations for each communication event in the physical layer, where a handful of weights applied to a basis function representation can reconstitute a number of realistic candidate SSPs \citep{leblanc_underwater_1980,lin_merging_2010}.
 The SSP is the only configurable input to the group velocity prediction.
@@
 an optimal set of weights chosen by a human decision maker via an interactive Tactical Decision Aid framework \citep{bhatt_embedded_2021}.
-For the post-processing analysis, we add the SSP from HYCOM for comparison, to mirror the information available on a submarine\textemdash modeled data, historical data, and in situ data.
+For the post-processing analysis, we add the SSP from HYCOM for comparison, to mirror the information available on a submarine\textemdash modeled data, historical data, and \textit{in situ} data.
 This order corresponds to increasing ducted conditions of the Beaufort Lens.
@@
 It is important to note that BELLHOP produces an abundance of eigenrays; the ones displayed here are chosen by travel time proximity to the recorded data.
-An important pattern is the increased complexity in multipath as ducted conditions increase.
+The important pattern is the increased complexity in multipath via surface interactions as ducted conditions increase.
 
@@
   \includegraphics[width=\reprintcolumnwidth]{figs/raytrace-3env-zs-30.pdf}
-  \caption{Eigenrays for beacon to beacon events for each nominal source depth of 30 m.
+  \caption{Eigenrays for beacon to beacon events for each sound speed with a nominal source depth of 30 m.
 The beacons are highlighted in color/marker coding in figure \ref{fig:overview}.
-The eigenrays are curated from BELLHOP by travel time proximity and are traced in the representative colors over a total ray fan in gray.}
+The eigenrays are curated from BELLHOP by travel time proximity and are traced in the representative receiver colors over a total ray fan in gray.}
   \label{fig:raytrace-zs30}
@@
 \begin{table}[h!]
-\renewcommand{\arraystretch}{1.5}
+\renewcommand{\arraystretch}{1.3}
 \centering
-\begin{tabular}{r|ccccc}
-SSP & Count & Mean [m] & Median [m] & STD [m] & Max [m] \\ \hline
+\begin{tabular}{r|c|c|c|c|c}
+SSP & Count & ~Mean [m] & ~Median [m] & ~STD [m] & ~Max [m] \\ \hline
 Baseline & 243 & 11.38 & 11.96 & 4.23 & 23.95 \\ 
@@
 \end{tabular}
-\caption[Range estimation error for \textit{in situ} events]{The range estimation error for \textit{in situ} events. The error here is presented as the absolute, not directional, error.}
+\caption[Range estimation error for \textit{in situ} events]{The range estimation error for \textit{in situ} events.
+The error here is presented as the absolute, not directional, error.
+Almost all directional errors, however, are positive.}
 \label{tab:rangeErrorInSitu}
@@
 Table \ref{tab:rangeErrorInSitu} breaks down the number of events, mean, median, standard deviation, and maximum absolute range error for the \textit{in situ} calculations.
-The mean and median for both sound speeds are similar on the order of centimeters; the chosen weights show a maximum greater than that of baseline, but it is from a OWTT event that has no comparison in the baseline data set.
-The overarching statistical results show that both sound speed environments, given the smaller operational scales and shallow depths, are accurate enough to approximate the group velocity structure given a local grid.
-Of course, these statistics obscure the relationship between range error and range (or OWTT as the proxy for range).
+The mean and median for both sound speeds are similar on the order of centimeters; the chosen weights show a maximum 2 meters greater than that of baseline, but it is from a communication event that has no comparison in the baseline data set.
 Analysis shows that where there is overlap in range between sound speed conditions, the range error difference is no more than a few meters.
 This is likely driven by the prominence of the duct.
-
-Where there is overlap in the physical layer, i.e. in OWTT, the range error driven by the two virtual layers is no more than a few meters.
-This is best demonstrated by all source and receiver depth pairings, except for a source at 30 m and a receiver at 90 m, where the baseline is clearly visibly closer to no range error than the weighted environment.
-This may be driven by the prominence of the duct, but it is certainly occluded by the order of magnitude difference between the number of events in each virtual layer.
+The overarching statistical results show that both sound speed environments, given the smaller operational scales and shallow depths, are accurate enough to approximate the group velocity structure given the local grid.
+Of course, overall statistics obscure the relationship between range error and range (or OWTT as the proxy for range).
 
@@
 
-Due to the operational constraint of only one modem running the vehicle behavior at a time, the previous analysis has looked at 811 total beacon to beacon events, where 243 use the baseline SSP and 568 use the weighted SSP.
-While overall statistics are presented, and individual events are compared for the same SSP between the two algorithms, the events themselves are not equally distributed with respect to depth, range, and sound speed status.
-
-The analysis is separated by source depth, and where possible, receiver depth, as the group velocities across source and receiver depths may be different.
-
+A post-processing analysis that replicated the real-time pipeline was run to overcome the unequal distribution of communication events with respect to depth, range, and sound speed status.
 There are 1242 total beacon to beacon events in the modem experiment that would have been accepted by the ICNN for real-time ranging and navigation had the vehicle behavior been running on all modems.
@@
 Naturally, because $\hat{r}$ converges to the correct solution, a comparison of $\hat{r}$ with the GPS-tracked range shows a normal, zero-centered distribution within the bounds of GPS drift.
-The analysis here therefore seeds ``omniscient'' knowledge of the range and leverages the post-processing pipeline to more thoroughly evaluate the acoustic range estimate for all modem events, with all three sound speed sources, and both group velocity criterion.
-Thus, the results in this section evaluate the utility of the algorithms and sound speed sources, divorced from their role in the ICNN.
-We first compare the post-processing results between the two algorithms, and then take a step-by-step approach to examine how the nearest bounce algorithm works in more detail.
+The analysis here therefore seeds ``omniscient'' knowledge of the range and leverages the post-processing pipeline to more thoroughly evaluate the acoustic range estimate for all modem events, with three relevant sound speed sources, and both group velocity criterion.
+Accordingly, the results in this section evaluate the utility of the algorithms and sound speed sources, divorced from their role in the ICNN.
 
 \FloatBarrier
-\subsection{Group velocity prediction}
+\subsection{Group velocity predictions}
 
@@
 \includegraphics[width=\textwidth]{figs/SLIDES-gvel-comparison-wdata-30.pdf}
-\caption{\label{fig:gvel30}{Caption here.}}
+\caption{\label{fig:gvel30}{A comparison of group velocity predictions for all beacon to beacon events in post-processing with a source depth of 30 m.
+The sound speed source is indicated by color.
+The minimal and nearest bounce criterion are distinguished by different marker shapes, compared to the separately colored orange dots showing the naive, data-driven group velocity calculation.}}
 \end{figure}
 
-\subsection{Psuedorange estimations}
-
-\subsubsection{Comparing the nearest and minimal bounce criteria}
+\FloatBarrier
+\subsection{Psuedorange estimations from the nearest and minimal bounce criteria}
 
@@
 \includegraphics[width=\textwidth]{figs/compare-methods-postv1v2.pdf}
-\caption{\label{fig:gvel30}{Caption here.}}
+\caption{\label{fig:compareV1V2}{Caption here.}}
 \end{figure}
@@
 \FloatBarrier
-\subsubsection{Psuedorange error versus one way travel time}
+\subsection{Psuedorange error versus one way travel time}
 
@@
 \includegraphics[width=\textwidth]{figs/range-error-owtt-newalgorithm.pdf}
-\caption{\label{fig:gvel30}{Caption here.}}
+\caption{\label{fig:gvel30}{The post-processed range error for source depths of 20, 30, and 90 m, and receiver depths of 30 and 90 m.
+The dashed gray line shows no error.
+The shaded region connects the range performance across all events.}
 \end{figure}
@@
 
-Systemic GPS drift drives the severe banding found in comparisons of range estimations via both group velocity algorithms, with those bands showing 100\% positive correlation between methods and being a similar size in meters.
-Thus, while this thesis is primarily motivated by achieving similar accuracy and precision of GPS, the impressive results from acoustic ranging necessitate a discussion about the limitations of GPS, particularly at polar latitudes.
+Systemic GPS drift drives the severe banding found in comparisons of range estimations via both group velocity algorithms, as shown in Fig.
+\ref{fig:compareV1V2}, with those bands showing 100\% positive correlation between methods and being a similar size in meters.
+Thus, while this work is primarily motivated by achieving similar accuracy and precision of GPS, the impressive results from acoustic ranging necessitate a discussion about the limitations of GPS, particularly at polar latitudes.
 
@@
 Fix rate looks at the ratio of successful fixes to overall attempts in acquiring a location.
-In animal tracking studies in polar latitudes, fix rates have significantly improved in the last twenty years, from as low as 26\% in 1996\citep{moen_effects_1996} to 87.2\% in 2018 \citep{jung_accuracy_2018}.
+In animal tracking studies in polar latitudes, fix rates have significantly improved in the last twenty years, from as low as 26\% in 1996 \citep{moen_effects_1996} to 87.2\% in 2018 \citep{jung_accuracy_2018}.
 The latter study showed a precision of 4.3 $\pm$ 4.0 m and an accuracy of 6.0 $\pm$ 4.7 m.
@@
 However, both of these metrics are functions of the satellite constellation available to any given GPS receiver.
-The Arctic area is beyond the reach of many geostationary satellites, where they are low on the horizon and only visible for brief periods.
+The Arctic is beyond the reach of many geostationary satellites, where they are low on the horizon and only visible for brief periods.
 Given that their positions are tracked by stations on Earth, and their timing is updated twice a day via atomic clock, a reasonable hypothesis for bands of GPS error are different and/or updated satellite connections.
@@
 The literature even references the total amount of electrons encountered by the ``GPS raypath'' as a means of explaining signal fluctuations \citep{themens_nature_2015}.
-This variability is observed by seasonal and solar cycle variabilities,\citep{themens_nature_2015} severe ionospheric storms\citep{mitchell_gps_2005}, and auroral activity (increased electron precipitation)\citep{jin_gps_2014,gwal_gps_2011}.
+This variability is observed by seasonal and solar cycle variabilities \citep{themens_nature_2015}, severe ionospheric storms \citep{mitchell_gps_2005}, and auroral activity (increased electron precipitation) \citep{jin_gps_2014,gwal_gps_2011}.
 In fact, the \textit{aurora borealis} was observed on multiple nights during ICEX20.
@@
 Figure \ref{fig:gps-drift-example} highlights the GPS and observed OWTT drift relative to the ice movement for two pairs of modem buoy connections.
-The top two panels show GPS drift in both the $x$ and $y$ directions relative to the median $x$ and $y$ distance recorded between the two buoys.
-The bottom two panels indicate the GPS drift as $\delta R = \sqrt{\delta x^2 + \delta y^2}$ and temporal drift, $\delta t$, relative to the median OWTT recorded between the two modems.
-Here we use the median, not the mean, to be more robust to outliers in the dataset.
-The dashed line is scaled by a group velocity of 1440 m/s, such that if there was no GPS drift, all events should exist on the line.
-
-The left hand side of the plot shows the connections between the North and East buoys.
-The top shows consistent change in distance with respect to time, as the $\delta X$ measurement trends from 2 to -2 and the $\delta Y$ measurement trends from -1 to 1 meters.
-This indicates that the ice movement between the North and East buoys is non-rigid, i.e., they are actually moving relative to each other.
-The width of the scatter plot is a decent indicator for the precision of the GPS signal, as it remains consistent across many colors of scattered points, which correspond to similar regions in time.
-The idea that these buoys are moving relative to each other is corroborated in the bottom plot, where some events exist on the dashed line indicating correlation between GPS movement and OWTT change.
-Some of the offsets between the vertical bands relate to different operational configurations of source and receiver depths.
-However, the large majority of events show vertical banding for the same nominal $\delta t$, indicating the amount of GPS drift.
-
-This idea of GPS drift relative to the same OWTT measurements is also indicated by events between the other two buoys, South and West, on the right hand side.
-The GPS drift, on the top, shows a point cloud centered around the origin for events four hours and 13 hours after the experiment origin; thus these buoys are moving in a more rigid ice floe.
-There is minimal impact by source and receiver depth on the spread of OWTT.
-The GPS drift relative to the OWTT, on the bottom, shows a large amount of GPS drift for very little OWTT drift.
-The OWTT drift observed here is sensitive to the acoustic scattering, multipath, or environmental microstructure.
-
-These are two subsets of the physical links, they cover all four GPS modem buoys except for Camp.
-The GPS at camp, as shown in figure \ref{fig:iceFloe}, is the least accurate due to the human activity and infrastructure occluding the physical puck.
-A more comprehensive look at all physical link pairs is in Appendix \ref{app:beacon2beacon}.
+The two panels indicate the GPS drift as $\delta R = \sqrt{\delta x^2 + \delta y^2}$ and temporal drift, $\delta t$, relative to the median OWTT recorded between the two modems.
+The dashed line is scaled by a group velocity of 1440 m/s, such that if there were ideal sensor measurements with no drift, all events should exist on or near the line.
 
@@
 	\centering
-	\includegraphics[width=\textwidth]{figs/gps-drift-example} 
-	\caption[Examination of GPS and modem drift]{A comparison of GPS drift (top) colored by hours since the experiment origin and GPS drift against OWTT spread (bottom) demarcated by source depth and receiver depth.
-The physical link between North and East are shown on the left; South and West is on the right.
-}
+	\includegraphics[width=\reprintcolumnwidth]{figs/gpsDrift-owttDrift.pdf} 
+	\caption{A comparison of GPS drift (y-axis) versus OWTT drift (x-axis), colored by source and receiver depth.
+The physical link between North and East are shown on the top; South and West is on the bottom.}
 	\label{fig:gps-drift-example}
 \end{figure}
+
+The top panel shows the connections between the North and East buoys.
+There is relative, i.e. non-rigid, ice movement between the North and East buoys, evidenced by events spanning the dashed line.
+But the height of the scatter plot is indicates the precision of the GPS signal, as it remains consistent across many arrival time bands.
+Naturally, some minor offsets between these vertical bands relate to different operational configurations of source and receiver depth.
+However, the large majority of events show vertical banding for the same nominal $\delta t$, indicating the amount of GPS drift.
+
+This idea of GPS drift relative to the same OWTT measurements is further indicated by events between the other two buoys, South and West, in the bottom panel.
+These buoys are moving in a more rigid ice floe and there is minimal impact by source and receiver depth on the spread of OWTT.
+The GPS drift is much larger relative to the OWTT drift, which is sensitive to acoustic scattering, multipath, and/or environmental microstructure.
+
+These are just two subsets of the physical links that cover all four GPS modem buoys, except for Camp.
+The GPS at camp is the least accurate due to the human activity and infrastructure occluding the physical puck.
 
@@
 Range estimation is an essential step of acoustic localization and navigation.
-This work solely focuses on mitigating range estimation error via a stochastic prediction of the group velocity, knowing that this can have an outsized benefit on minimizing trilateration error.
-We hypothesize and validate that the group velocity, especially for mission durations, is smoothly varying in range and multipath structure, and sensitive to source and receiver depth.
+This work solely focuses on mitigating range estimation error via a stochastic prediction of the group velocity, knowing that this can have an disproportionate benefit on minimizing trilateration error.
+We hypothesize and validate that the embedded stochastic prediction of a single group velocity is a smoothly varying function of range, source and receiver depth pairings, and multipath structure.
 We employ a GPS-tracked experiment to have a ground truth comparison for real-time localization algorithms.
+
+The real-time system achieves GPS-like navigation without taking into account multipath structure; the ranging error improves by an order of magnitude with the suggested multipath adaptability, minimizing range error to single meters.
+Post-processing analysis shows that this method of ranging is actually sensitive to GPS drift.
 
@@
 
-\FloatBarrier
+\clearpage
 \section*{Supplementary Figures}
 
-\clearpage
 \subsection*{Ice movement during the modem experiment}
@@
 \includegraphics[width=\textwidth]{figs/ice-floe-drift.pdf}
-\caption{\label{fig:iceFloeDrift}{Caption here.}}
+\caption{\label{fig:iceFloeDrift}{The ICNN accounts for ice drift throughout the mission duration.
+This is the magnitude of the ice drift recorded, at each buoy, throughout the modem experiment.
+The shaded regions reference which sound speed estimate was used.}}
 \end{figure}
@@
 \includegraphics[width=0.8\textwidth]{figs/SLIDES-gvel-comparison-wdata-20.pdf} \\
+\vspace{2em}
 \includegraphics[width=0.8\textwidth]{figs/SLIDES-gvel-comparison-wdata-90.pdf}
+\caption{A comparison of group velocity predictions for all beacon to beacon events in post-processing.
+The top panel shows a source depth of 20 m; the bottom panel shows a source depth of 90 m.
+The sound speed source is indicated by color.
+The minimal and nearest bounce criterion are distinguished by different marker shapes, compared to the separately colored orange dots showing the naive, data-driven group velocity calculation.}
 \end{figure}
@@
 \begin{acknowledgments}
-This research was supported by the Office of Naval Research and the NDSEG Fellowship.
-We are grateful to all our collaborators in ICEX20\textemdash Bluefin Robotics, Toby Schneider, Lee Freitag \textemdash
+We acknowledge the significant operational effort spearheaded by the LAMSS ICEX20 team and all our collaborators.
+Bhatt was funded by a National Defense, Science, and Engineering Graduate Fellowship.
+This work was supported by the Office of Naval Research 322-OA under ICEX20 (N00014-17-1-2474) and Task Force Ocean (N00014-19-1-2716).
+
 \end{acknowledgments}

### 2021-08-20 src/analyze_GPS_drift.m

@@ -9,7 +9,7 @@ modem_labels = {'North','South','East','West','Camp'};
 
 colorDepth = containers.Map([20 30 90],{[70 240 240]./256,[0 130 200]./256,[0 0 128]./256});
 sourceDepth = containers.Map([20 30 90],{'>','^','v'});
-alphaDepth = containers.Map([20 30 90],[.2 .2 .2]);
+alphaDepth = containers.Map([20 30 90],[.5 .4 .4]);
 
 sspGVEL = 1440/1000; % meters / millisecond
 
@@ -99,69 +99,71 @@ set(gca,'XColor','white');
 set(gca,'YColor','white');
 
 % export
-h_printThesisPNG('gps-drift');
+% h_printThesisPNG('gps-drift');
 
 %% sub-figure
 
-figure('name','gps-drift-example','renderer','painters','position',[108 108 1200 1100]); clf;
-t = tiledlayout(2,2,'TileSpacing','compact');
+figure('name','gps-drift-example','renderer','painters','position',[108 108 500 1200]); clf;
+t = tiledlayout(2,1,'TileSpacing','compact');
 
 rbounds1 = [-2.6 2.6];
 rticks1 = [-2:2];
 
-xbounds2 = [-1.1 1.1];
+xbounds2 = [-1.05 1.05];
 xticks2 = [-1:0.5:1];
 
-% panel 1
-nexttile;
-h_dxdy('North','East',200);
-%colorbar;
-xlim(rbounds1)
-ylim(rbounds1)
-xticks(rticks1);
-yticks(rticks1);
-xticklabels auto
-yticklabels auto
-xlabel('GPS \deltax [m]');
-ylabel('GPS \deltay [m]');
-title('GPS drift --- North and East buoys','fontsize',15);
-
-% panel 2
-nexttile;
-h_dxdy('South','West',200);
-xlim(rbounds1)
-ylim(rbounds1)
-xticks(rticks1);
-yticks(rticks1);
-xticklabels auto
-yticklabels auto
-xlabel('GPS \deltax [m]');
-ylabel('GPS \deltay [m]');
-title('GPS drift --- South and West buoys','fontsize',15);
-cb = colorbar;
-cb.Ticks = 0:2:14;
-cb.TickLabels = num2cell(0:2:14);
-cb.Label.String = 'time past [hours]';
+% % panel 1
+% nexttile;
+% h_dxdy('North','East',200);
+% %colorbar;
[…]

### 2021-08-21 doc/main.tex
updates, need to finish intro, proofread

@@
 \section{\label{sec:1} Introduction}
+
+Autonomous underwater vehicles (AUVs) are an increasingly capable platform to explore and sample the ocean, particularly for remote and/or dangerous environments.
+Lightweight, mobile, and inexpensive compared to crewed- and ship-based systems, these vehicles can enable a safer, data-forward style of field oceanography \citep{bellingham_robotics_2007,petillot_underwater_2019}.
+
+But navigation uncertainty is a major challenge in considering AUVs as reliable and standard tools for oceanographic research.
+While land and air-based robots utilize satellite-based global positioning system (GPS) to achieve stunning location accuracy and precision throughout the duration of their missions, AUVs cannot access GPS while underwater due to the rapid attenuation of electromagnetic waves \citep{preisig_acoustic_2007}.
+A vehicle can stall on the surface to receive a GPS fix, but this foolproof method of re-positioning is inaccessible in an ice-covered environment or other GPS-denied situations.
+A recent study tracking 22 floats under the ice in the Weddell Gyre near Antarctica showed that they were unable to surface for 8 months and position uncertainty grew to 116 $\pm$ 148 km in the lateral directions \citep{chamberlain_observing_2018}.
+
+Much work has been done in underwater vehicle navigation, which relies on any combination of dead reckoning, hydrodynamic models, inertial navigation systems, doppler velocity logs, and/or acoustic baseline navigation systems \citep{paull_auv_2014}.
+In particular, the long baseline (LBL) navigation system is the most similar in scale and style to GPS, and most appropriate for mitigating drift in underwater vehicle localization without overburdening computation on the vehicle \citep{van_uffelen_global_2021}.
+
+\subsection{Timing, Positioning, and Navigation}
+
+Underwater acoustic navigation (and GPS) is predicated on timing and positioning.
+Timing is the ability to maintain accurate and precise time, sometimes synchronized across multiple nodes.
+The advent of the WHOI micro-modem \citep{singh_underwater_2006} and synchronized chip scale atomic clocks \citep{gardner_second_2016} enabled ranging via one-way travel time (OWTT).
+
+Positioning is the ability to accurately determine one's location and is generally done via trilateration.
+Real-time navigation efforts assume a locally homogeneous sound speed profile (SSP) that enables a linear, deterministic scaling between travel time and range \citep{eustice_recent_2006,eustice_experimental_2007,webster_preliminary_2009,webster_advances_2012}.
+Short range (less than ten kilometer) navigation efforts accurately rely on a nominal sound speed value and achieve minimal navigation error \cite{eustice_experimental_2007,webster_preliminary_2009,kepper_mems_2017}.
+For mesoscale (tens to hundreds of kilometers) navigation or localization efforts, a deterministic \citep{graupe_preliminary_2019} or through-the-sensor \citep{webster_towards_2015} value for sound speed is used.
+
+Post-processing localization efforts (re-navigation for a self-determinedly moving object or re-positioning for a passively moving one), without computational or time constraints, can indulge more complex acoustic methods and/or ocean representations.
+There are many different approaches, from accounting for a depth dependent SSP \citep{van_uffelen_localization_2016}, including spatio-temporal variability in the sound speed profile to minimize error \citep{graupe_preliminary_2019, mikhalevsky_deep_2020}, comparing acoustic records with synthetic acoustic records computed through an ocean model \citep{wu_deep_2019}, or isolating a specific ray path to find a representative group velocity \citep{mikhalevsky_deep_2020}.
+The ``cold start'' algorithm developed by \citet{mikhalevsky_deep_2020} re-positioned a floating hydrophone array with an error of 85 m and a standard deviation of 32 m relative to ground truth GPS data, but implementing this algorithm for navigation remains to be seen.
+
+\subsection{Underwater positioning and navigation in the Arctic}
+
+However, the key assumptions for OWTT ranging\textemdash a locally homogeneous sound speed profile that enables a linear, deterministic scaling between travel time and range\textemdash did not hold during a vehicle deployment in ICEX16, in the Beaufort Sea, in March 2016.
+
+However, the new Arctic ... [Beaufort Lens]
+
+The literature in underwater acoustic navigation and positioning is either real-time or physics-based.
+AUV operations in the Beaufort Sea necessitate a real-time and physics-based method to overcome the unique challenges of total ice cover, multiple ice floes, and the Beaufort Lens.
 
@@
   \caption{Eigenrays for beacon to beacon events for each sound speed with a nominal source depth of 30 m.
-The beacons are highlighted in color/marker coding in figure \ref{fig:overview}.
+The beacons are highlighted in color/marker coding in Fig.
+\ref{fig:overview}.
 The eigenrays are curated from BELLHOP by travel time proximity and are traced in the representative receiver colors over a total ray fan in gray.}
@@
 
-As shown in the eigenray traces, an acoustic arrival does not always take the direct path from source to the receiver, and the path it takes is extremely dependent on the sound speed profile.
+As shown in the eigenray traces of Fig.
+\ref{fig:raytrace-zs30}, an acoustic arrival does not always take the direct path from source to the receiver, and the path it takes is extremely dependent on the sound speed profile.
 Often, BELLHOP will produce a number of arrivals resulting from multiple propagation paths for any source-receiver pair.
-The suggested algorithm, the nearest bounce criteria (NBC), is only a slight modification from the minimal bounce criteria, and includes multipath as a new dimension of information with negligible additional computation.
+The suggested algorithm, the nearest bounce criteria (NBC), is a slight modification from the minimal bounce criteria, and includes multipath as a new dimension of information with negligible additional computation.
+This metric was introduced in post-processing.
 
@@
 Chosen weights & 568 & 11.36 & 11.40 & 8.12 & 26.0 \\
+\toprule
 \end{tabular}
@@
 
+There are 1242 total beacon to beacon events in the modem experiment that would have been accepted by the ICNN for real-time ranging and navigation had the vehicle behavior been running on all modems.
 A post-processing analysis that replicated the real-time pipeline was run to overcome the unequal distribution of communication events with respect to depth, range, and sound speed status.
-There are 1242 total beacon to beacon events in the modem experiment that would have been accepted by the ICNN for real-time ranging and navigation had the vehicle behavior been running on all modems.
+
 The value for the extrapolated range, $\hat{r}$, is only tracked when the modem runs the vehicle behavior; thus we replace $\hat{r}$ with the GPS-tracked range for all modem events.
-Naturally, because $\hat{r}$ converges to the correct solution, a comparison of $\hat{r}$ with the GPS-tracked range shows a normal, zero-centered distribution within the bounds of GPS drift.
-The analysis here therefore seeds ``omniscient'' knowledge of the range and leverages the post-processing pipeline to more thoroughly evaluate the acoustic range estimate for all modem events, with three relevant sound speed sources, and both group velocity criterion.
+Because $\hat{r}$ converges to the correct solution, a comparison of $\hat{r}$ with the GPS-tracked range shows a normal, zero-centered distribution within the bounds of GPS drift.
+The analysis here therefore seeds realistic but ``omniscient'' knowledge of the range and leverages the post-processing pipeline to more thoroughly evaluate the acoustic range estimate for all modem events, with three relevant sound speed sources, and both group velocity criterion.
 Accordingly, the results in this section evaluate the utility of the algorithms and sound speed sources, divorced from their role in the ICNN.
@@
 
+The minimal and nearest bounce algorithm is applied using the three sound speed inputs shown in Fig.~\ref{fig:raytrace-zs30}; the resulting predicted group velocities for a source depth of 30 m are shown in Fig~\ref{fig:gvel30}.
+The takeaways from this source depth are representative of those from 20 and 90 m as well (see Supplementary Fig.~\ref{fig:gvelMore}).
+
 \begin{figure}[h!]
-\includegraphics[width=\textwidth]{figs/SLIDES-gvel-comparison-wdata-30.pdf}
-\caption{\label{fig:gvel30}{A comparison of group velocity predictions for all beacon to beacon events in post-processing with a source depth of 30 m.
+\includegraphics[width=\columnwidth]{figs/SLIDES-gvel-comparison-wdata-30.pdf}
+\caption{A comparison of group velocity predictions for all beacon to beacon events in post-processing with a source depth of 30 m, with group velocity on the y-axis and recorded travel time on the x-axis.
+The left panel is for a receiver depth of 30 m; the right panel for 90 m.
 The sound speed source is indicated by color.
-The minimal and nearest bounce criterion are distinguished by different marker shapes, compared to the separately colored orange dots showing the naive, data-driven group velocity calculation.}}
+The minimal and nearest bounce criterion are distinguished by different marker shapes, compared to the separately colored orange dots showing the naive, data-driven group velocity calculation.}
+\label{fig:gvel30}
+\end{figure}
+
+Because direct path dominates the predictions from the MBC, group velocity estimates are greater than those from NBC.
+The one instance this is not true is important to mention\textemdash for a travel time of 2.2 s, where the minimal multipath is dominated by a single bottom bounce, but the eigenray traces show two, three, and four bounces for HYCOM, baseline, and weighted SSPs, respectively (the red line to the furthest transmission in Fig.
+\ref{fig:raytrace-zs30}).
+The nearest bounce method never classifies anything as direct path for the 30 m source, though BELLHOP is quite adept at finding those rays.
+
+The goal of the group velocity estimation is to converge towards the naive group velocity estimation, i.e. the GPS-derived range divided by the recorded travel time.
+For a 30 m receiver depth, the NBC shows more overlap with data-derived values as it classifies multipath more correctly.
+For a 90 m receiver depth, the overlap is less accurate due to computational constraints.
+Given a limited fan of rays entering the shadow zone, there is a less reliable simulated time arrival packet.
+
+Overall, analysis shows that the higher multipath classification produces more accurate range estimates, likely driven by a tighter bundle of rays.
+However, the data is too small to draw significant conclusions.
+Discontinuities in multipath classification verify our hypothesis for its importance for a smoothly varying group velocity, as shown in the cluster where HYCOM jumps from one to two bounces amidst the baseline SSP seeing two bounces and the weighted SSP seeing three.
+
+\FloatBarrier
+\subsection{Psuedorange error versus one way travel time}
+
+When using a linear scaling to convert travel time into range, any offset between the assumed sound speed and the horizontal group velocity produces unconstrained error with increasing receiver distance.
+Figure \ref{fig:rangeError} shows the directional range error footprints for all three sound speed inputs with respect to OWTT, separated by source and receiver depth configurations.
+The weighted SSP range error generally has the smallest and most zero-centered footprint.
+The one case it does not is for the source-receiver pairings between 30 and 90 m in depth.
+The increased error for these reciprocal transmission paths is most likely driven by the computational artifacts propagating through the steep sound speed gradients of the lens, through a shadow zone.
+All other source depth pairings are significantly improved using the chosen weights compared to HYCOM or the baseline.
+Most importantly, we see the consequences of the adaptive group velocity in that range error does not strictly increase with OWTT.
+
+\begin{figure}[!ht]
+\includegraphics[width=\textwidth]{figs/range-error-owtt-newalgorithm.pdf}
+\caption{The post-processed range error for source depths of 20, 30, and 90 m, and receiver depths of 30 and 90 m.
+The dashed gray line shows no error.
+The shaded region connects the range performance across all events.}
+\label{fig:rangeError}
 \end{figure}
@@
 
+The previous section displayed the psuedorange performance using the NBC.
+Here, we explicitly compare the ranging accuracy between the nearest and minimal bounce criteria for all events in the post-processing pipeline.
+It is important to note that because the post-processing pipeline is the same for both methods, that the efficacy of the NBC compared to the MBC is 100\%, as the former simplifies to the latter for direct path classifications.
+
 \begin{figure}[h!]
 \includegraphics[width=\textwidth]{figs/compare-methods-postv1v2.pdf}
-\caption{\label{fig:compareV1V2}{Caption here.}}
+\caption{\label{fig:compareV1V2}{A comparison of range estimation error for all beacon to beacon events in post-processing.
+From left to right, the SSPs are the baseline, the chosen weights, and HYCOM.
+The colors indicate the source depth, darkening with depth, and the shapes indicate the multipath structure.
+Because this plot is square, the shaded region shows where the updated algorithm is less accurate than the \textit{in situ} algorithm.}}
 \end{figure}
 
-\FloatBarrier
-\subsection{Psuedorange error versus one way travel time}
+Figure \ref{fig:compareV1V2} is a crossmap for all three SSPs for both group velocity predictors, with the minimal bounce on the x-axis and the nearest bounce on the y-axis; the important range estimation metrics are summarized in table \ref{tab:rangeV1V2}.
 
-If there was an assumed deterministic group velocity, the range error would grow with OWTT proportionally to the difference between the estimated and actual value.
-Thus, an indication that the algorithm is adapting group velocity correctly is that range error does not strictly increase with OWTT.
+The first thing to notice is that all the events are on the right hand side of the plots\textemdash the MBC has a systemic overestimation, which is reduced via NBC as events straddle the x-axis.
+The improvement from MBC to NBC is most evident for the realistic sound speed; while the HYCOM SSP improves from a median absolute range error of 6.41 to 4.61 m, the baseline SSP improves from 10.30 to 2.27 m, and the weighted SSP improves from 13.28 to 2.12 m.
+Table \ref{tab:rangeErrorPost} shows further statistics on the absolute range error by SSP and group velocity algorithm.
+The order of magnitude improvement in the ducted SSPs demonstrate the effectiveness of the algorithm exploiting the multipath conditions.
+The gridded approach and multipath classification build in the computational cushion for minimal error and minimal bias compared to direct eigenray assessment.
 
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{figs/range-error-owtt-newalgorithm.pdf}
-\caption{\label{fig:gvel30}{The post-processed range error for source depths of 20, 30, and 90 m, and receiver depths of 30 and 90 m.
-The dashed gray line shows no error.
-The shaded region connects the range performance across all events.}
-\end{figure}
+\begin{table}[h!]
+\renewcommand{\arraystretch}{1.5}
+\centering
+\begin{tabular}{r|cc|cc|cc}\toprule
+ & \multicolumn{2}{c|}{\textbf{Baseline} } & \multicolumn{2}{c|}{\textbf{Chosen Weights} } & \multicolumn{2}{c}{\textbf{HYCOM}} \\
+ algorithm & \cellcolor[HTML]{EFEFEF}minimal & nearest & \cellcolor[HTML]{EFEFEF} minimal& nearest & \cellcolor[HTML]{EFEFEF}minimal & nearest \\ \hline
+ minimum [m] 	& \cellcolor[HTML]{EFEFEF}0.01 & 0.00	& \cellcolor[HTML]{EFEFEF}0.00 	& 0.00 	& \cellcolor[HTML]{EFEFEF}0.11 & 0.01 \\
+ 25th \% [m]   & \cellcolor[HTML]{EFEFEF}4.96 & 0.99	& \cellcolor[HTML]{EFEFEF}6.26 	& 0.95 	& \cellcolor[HTML]{EFEFEF}3.30 & 2.25 \\   
+ median [m]		& \cellcolor[HTML]{EFEFEF}10.30 & 2.27 	& \cellcolor[HTML]{EFEFEF}13.28 & 2.12 	& \cellcolor[HTML]{EFEFEF}6.41 & 4.61 \\
+ 75th \% [m]   & \cellcolor[HTML]{EFEFEF}15.81 & 5.51 	& \cellcolor[HTML]{EFEFEF}19.75 & 4.11 	& \cellcolor[HTML]{EFEFEF}10.92 & 7.46 \\
+ maximum [m]    & \cellcolor[HTML]{EFEFEF}22.52 & 14.96 & \cellcolor[HTML]{EFEFEF}1491  & 20.21 & \cellcolor[HTML]{EFEFEF}19.55 & 15.81 \\
+ \toprule
+\end{tabular}
+\caption[Comparison of post-processing range estimation algorithms across all events]{A comparison of range estimation metrics for each sound speed source and group velocity estimation algorithm for all 1283 beacon to beacon events via post-processing.
+The 0th (minimum), 25th, 50th (median), 75th, and 100th (maximum) percentiles are shown to the range resolution afforded by the WHOI Micro-Modem.
+There are a few outliers that drive the mean to be higher than the median.}
+\label{tab:rangeErrorV1V2}
+\end{table}
+
+All plots show the same 45 degree banding of source depth and multipath structure, indicating systemic GPS noise.
+Even with poorer accuracy from the GPS solution, the acoustic methods remain high precision.
+Some bands are adjacent to others of the same styling, likely due to ice drift.
 
 \FloatBarrier
-\subsection{GPS sensor drift}
+\subsection{Further analysis of the anomalous beacon to beacon connection}
+
+As shown in table \ref{tab:rangeErrorPost}, there is a striking maximum range error of 1491 m for weighted SSP in the minimal bounce criteria.
+This section isolates the events that encompass that outlier for further examination, summarized in table \ref{tab:rangeErrorPostExplanation}.
+There are 10 events from South transmitting at 30 m depth to North receiving at 30 m depth.
+The OWTT spread is from 2.1958 to 2.1963 s; the naive group velocity is 1429.3 to 1430.1 m/s; and the GPS-tracked range is from 3138.54 m to 3140.87 m.
+This example ends up being an excellent case study for how sound speed and multipath fidelity work in concert to minimize range error.
+
+\begin{table}[h!]
+\renewcommand{\arraystretch}{1.5}
+\centering
+\begin{tabular}{r|cc|cc|cc}\toprule
+ & \multicolumn{2}{c|}{\textbf{Baseline} } & \multicolumn{2}{c|}{\textbf{Chosen Weights} } & \multicolumn{2}{c}{\textbf{HYCOM}} \\
+ algorithm & \cellcolor[HTML]{EFEFEF}minimal & nearest & \cellcolor[HTML]{EFEFEF} minimal& nearest & \cellcolor[HTML]{EFEFEF}minimal & nearest \\ \hline
+ \# bounces 				& \cellcolor[HTML]{EFEFEF}1 		& 3 		& \cellcolor[HTML]{EFEFEF}1 		& 4 		& \cellcolor[HTML]{EFEFEF}0 	 & 2 \\
+ mean OWTT [s] 				& \cellcolor[HTML]{EFEFEF}2.1853 	& 2.1943 	& \cellcolor[HTML]{EFEFEF}4.1812 	& 2.1955 	& \cellcolor[HTML]{EFEFEF}2.1877 & 2.1902 \\
+ mean group velocity [m/s] 	& \cellcolor[HTML]{EFEFEF}1436.7 	& 1430.8 	& \cellcolor[HTML]{EFEFEF}750.88 	& 1430.0 	& \cellcolor[HTML]{EFEFEF}1435.1 & 1433.4 \\
+ range error [m] 			& \cellcolor[HTML]{EFEFEF}15.4 		& 2.39 	 	& \cellcolor[HTML]{EFEFEF}-1491 	& 0.77 		& \cellcolor[HTML]{EFEFEF}11.9 & 8.30 \\
+ \toprule
+\end{tabular}
+\caption[Further look at algorithm differences for one beacon to beacon pairing]{A comparison of range estimation metrics for each sound speed source and group velocity estimation algorithm on all 10 communication events between North at 30 m \& South at 30 m.}
+\label{tab:rangeErrorPostExplanation}
+\end{table}
+
+The large error is driven by the MBC unexpectedly defaulting to a bottom bounce with a much greater OWTT.
+Thus the NBC ``correctly'' classifies the multipath as 4 bounces, reducing the range error from greater than a kilometer to less than a meter.
+While there is no actual way of knowing if this is the correct multipath structure, the range error is remarkably small.
+This pattern of not choosing the minimal observed bounce structure is consistent across all SSPs; the baseline goes from 1 to 3 bounces and HYCOM goes from 0 to 2 bounces.
+Notably, the baseline and HYCOM range errors are never egregiously large, but are nonetheless improved with the NBC algorithm.
+Thus, for acoustically complex environments, the NBC has a disproportionately positive impact as the estimated SSP approaches the desired SSP.
+
+In comparison to the earlier eigenray traces, the results here distill the importance of the gridded approach to curate multipath propagation. 
+The grid gives BELLHOP tolerance for finding solutions that otherwise may not be found in a center or single point solution; the limitations of numerical computation, especially in a real-time environment, are more adeptly addressed by baking in a small amount of uncertainty than trying to prescribe an exact solution.
+This creates more data to sort through given that BELLHOP is a greedy estimator of eigenrays, but is computationally simpler than running a dense ray fan for precise eigenray finding.
+
+The NBC algorithm, coupled with sound speed fidelity, work in concert to mitigate error.
+Either one alone is not enough for the Beaufort Lens, as the environmental realism begins to match the acoustic conditions of intense multipath.
+
+\FloatBarrier
+\subsection{GPS sensor drift at polar latitudes}
 
@@
 Range estimation is an essential step of acoustic localization and navigation.
-This work solely focuses on mitigating range estimation error via a stochastic prediction of the group velocity, knowing that this can have an disproportionate benefit on minimizing trilateration error.
+Current approaches in real-time underwater acoustic navigation simplify the non-linear relationship between a sound speed profile and acoustic propagation with a deterministic sound speed.
+Some post-processing approaches attempt to leverage the acoustic arrival structure via various ray methods, but often use a singular SSP for long-term deployments.
+Thus, the conversion from travel time to range, particularly for real-time vehicle deployments, can be pre-conditioned for error growth as the OWTT/range increases.
+For GPS-denied navigation, and hostile environments like the Arctic, tolerance for error is close to none.
+This work addresses a critical need in acoustic navigation by retooling acoustic arrival methods generally deemed too complex or labor intensive for real-time, ray-based range estimation to achieve GPS-like positioning.
+
 We hypothesize and validate that the embedded stochastic prediction of a single group velocity is a smoothly varying function of range, source and receiver depth pairings, and multipath structure.
 We employ a GPS-tracked experiment to have a ground truth comparison for real-time localization algorithms.
-
-The real-time system achieves GPS-like navigation without taking into account multipath structure; the ranging error improves by an order of magnitude with the suggested multipath adaptability, minimizing range error to single meters.
-Post-processing analysis shows that this method of ranging is actually sensitive to GPS drift.
-
-This work, to our knowledge, is the first successful field deployment of a real-time, ray-based acoustic ranging system for underwater navigation.
-The methods here demonstrate remarkable ranging accuracy and precision for the Beaufort Lens, an acoustically complex environment.
-The \textit{in situ} prediction via BELLHOP balances the tension in simulation fidelity versus capturing the overarching ensemble physics to approximate a local group velocity.
-
-Furthermore, the improved nearest bounce criteria, validated in a high fidelity post-processing pipeline, produces range estimates that rival GPS accuracy and outclass GPS precision, especially at polar latitudes.
-It transforms multipath, widely considered as an obstacle for acoustic ranging, into a new information content to refine ranging accuracy.
-
-The scalability of the methods provided here for long-range applications remain to be seen.
-Given the NBC algorithm's improvement with increased multipath, its effectiveness is likely only challenged by the valid operational scales of a range independent propagation environment.
-Even then, the dominant SSP amongst the variability may very well produce GPS-like accuracy.
-The NBC algorithm can be easily be retooled to discriminate between surface and bottom bounces, and employ simple statistical learning or more advanced machine learning techniques to select an appropriate representative group velocity.
-Some more advanced vehicle intelligence can toggle use the range uncertainty to toggle between a more conservative group velocity estimate and the NBC, which is intrinsically quite precise.
+The real-time system achieves GPS-like navigation for an AUV without taking into account multipath structure; the ranging error improves by an order of magnitude with the suggested multipath adaptability, minimizing range error to single meters.
+Post-processing analysis shows that this method of ranging is sensitive to GPS drift itself; thus embedding a stochastic prediction of the horizontal group velocity has an outsized benefit to minimizing trilateration error.
 
@@
 Amongst them is defining the uncertainty grid for BELLHOP via stochastic or data-driven measures such as the distance traveled by the AUV between ICNN updates or the magnitude of the position corrections by the ICNN.
-Another is to entirely remove the seeded range and instead rely on the submerged asset's depth and recorded OWTT to find high probability fields in range, as briefly indicated in Chapter 6.
+Another is to entirely remove the seeded range and instead rely on the submerged asset's depth and recorded OWTT to find high probability fields in range.
 
-As the Arctic continues to become more accessible and navigable, safe navigation infrastructure and practices will be increasingly vital for industry, tourism, commercial, and military operations.
-This work contributes a new paradigm to embed environmentally adaptive behavior into AUV navigation that has demonstrated success in an acoustically complex and operationally challenging environment.
-Thus, we believe that this work enables more accurate range estimation, localization, and/or navigation for any field experiment given known source and receiver depths.
+This work, to our knowledge, is the first successful field deployment of a real-time, model-based acoustic ranging system for underwater navigation.
+The methods exploit the upward refracting nature and the total ice cover of the Arctic environment to achieve remarkable ranging accuracy and precision.
+It transforms multipath, widely considered as an obstacle for acoustic ranging, into a new information content to refine ranging accuracy.
+We believe that this work enables more accurate range estimation, localization, and/or navigation for any field experiment given known source and receiver depths.
 
-(NEED TO SKIM CONCLUSION FOR MORE MATERIAL)
+\subsection{Broader relevance \& future work}
+
+Performance in other acoustic environments will require introducing a different thresholded metric to sort time of arrivals.
+Given the NBC algorithm's improvement with increased multipath, its effectiveness is likely only challenged by the valid operational scales of a range independent propagation environment.
+For mesoscale operations, like that of many gliders, the group velocity criteria may need to be modified to better account for variability driven by range dependent propagation through internal waves, eddies, or even bathymetric changes like underwater volcanoes.
+BELLHOP simulations provide other relevant eigenray information, like time and angle of arrival, that is ripe for statistical and machine learning methods to classify a representative group velocity.
+A bespoke and fast ray tracing method, which was used during ICEX20 to evaluate the acoustic fit of sound speed profiles, can easily report back the number of turning points instead of the number of bounces for multipath classification.
+
+This approach will start to break down in extremely dynamic environments.
+Fast moving fronts, as seen in estuaries like the Connecticut River into the Long Island Sound, present an entirely new set of challenges not seen by internal waves or eddies.
+Buoyancy fluctuations in these regions threaten even simple AUV tasks like following a trackline.
+Acoustic communications are further complicated given a shallow environment with significant scattering \citep{lavery_measurements_2010,ross_acoustic_2012,lavery_broadband_2013}, where fast timefront modeling may only be coherent for trivial probabilities of the ocean state.
+Realistic \textit{in situ} considerations of the acoustic environment may not be possible in such environments without complete through-the-sensor integration of echosounder data and/or a hyper-realistic onboard ocean model.
+
+Many approaches to underwater navigation combine it with acoustic tomography, i.e., a joint estimation of both source and receiver locations and the ocean volume between them.
+There has been considerable success at this effort in post-processing methods, which utilize intensive\textemdash and due to the non-linearity of sound propagation, often brute force\textemdash computational methods.
+For vehicle operations, fast tomography is the ideal implementation, in that one can fully consider how sound speed structure, horizontally and vertically, influences sound propagation.
+AUVs can serve as moving sources to better image the ocean volume \citep{deffenbaugh_optimal_1997,elisseeff_ocean_2002}, where mobile tomography and navigation converge on the same set of component technologies: position estimation, sound speed parameterization estimation, ray path identification, and vehicle path optimization.
+
+But there are overwhelming challenges, operationally and computationally, for fast, mobile tomography to become a realistic endeavor.
+Addressing the spatial and temporal scales of what can be solved deterministically and what must be solved stochastically imposes a resolution constraint on the utility of gridded models that already dwarfs that of AUV operations.
+Given that AUV operations are often on small spatial and temporal scales, the added benefit of a gridded model is quite small, and in cases like the Arctic, may actually mischaracterize the ocean volume.
+
+For gliders, with longer and larger operational scales, an ocean model may provide useful information.
+Currently gliders are low power and do not have the storage or computational power to run a full-scale, realistic ocean model.
+Of course, all of this is assuming that the ocean model is accurate and has high enough resolution to resolve ocean features at small enough scales, which is in of itself an active research field.
+
+Ocean model realism (especially at finer scales) correlates with computational power and time.
+Thus, until a full-scale ocean model can be run in real-time on a small computer like a Raspberry Pi, the framework proposed in this thesis strikes a balance for the necessary environmental realism as depth-dependent with temporal updates.
+This has a cascading effect on acoustic realism, currently a range-independent scheme.
+Faster, more effective computational methods (and hardware devices) for acoustic propagation can certainly improve the real-time joint estimation problem, but are still constrained by this depth-dependent snapshot of the ocean volume.
+Until we reach that faraway goal, an updatable, range independent scheme, as employed in this work (see Section...) is a practical and effective approach.
+
+The Arctic, a generally inaccessible, dangerous, and expensive region for crewed monitoring via icebreakers,  is a prime location to invest in vehicle deployments and underwater observatories \citep{mikhalevsky_multipurpose_2015}.
+As anthropogenic climate change renders the region more accessible, safe navigation infrastructure and practices will be increasingly vital for industry, tourism, commercial, and military operations.
 
@@
 The minimal and nearest bounce criterion are distinguished by different marker shapes, compared to the separately colored orange dots showing the naive, data-driven group velocity calculation.}
+\label{fig:gvelMore}
 \end{figure}

### 2021-08-26 doc/main.tex
draft shared with Henrik + Oscar

@@
 \author{Oscar Viquez}
-\affiliation{Department of Mechanical Engineering, Massachusetts Institute of Technology, Cambridge, MA}
-\author{Bradli Howard}
-\affiliation{MIT-WHOI Joint Program in Oceanography/Applied Ocean Science \& Engineering, Cambridge and Woods Hole, MA, USA}
 \affiliation{Department of Mechanical Engineering, Massachusetts Institute of Technology, Cambridge, MA}
@@
 For positioning in post-processing, more computationally and/or labor intensive modeling methods are often employed to reduce the uncertainty driven by multipath arrivals.
-For an autonomous underwater vehicle deployment in March 2020 in the Beaufort Sea, the total under-ice environment and complex acoustic propagation via the Beaufort Lens necessitated the real-time integration of model-based data processing.
-An embedded, stochastic, ray-based prediction of the horizontal group velocity achieved a ranging mean absolute error of 11 m compared to GPS-aided beacon-to-beacon communication events.
-In post-processing, a multipath adjustment achieved ranging performance that rivals GPS.
-To our knowledge, these results are the first field experiment to demonstrate a real-time, physics-based model to constrain ranging error in underwater navigation.
+An autonomous underwater vehicle deployment in March 2020 in the Beaufort Sea faced significant operational challenges with a total ice-covered environment and complex acoustic propagation via the Beaufort Lens.
+We implement an embedded, stochastic, ray-based prediction of the horizontal group velocity and validate its performance with communication events between GPS-aided beacons, achieving a psuedorange mean absolute error of 11 m.
+In post-processing, we introduce a multipath criterion and achieve ranging performance that rivals GPS.
+To our knowledge, these results are the first field experiment to demonstrate a real-time, physics-based data processing to minimize ranging error in underwater navigation.
 \end{abstract}
@@
 
-Much work has been done in underwater vehicle navigation, which relies on any combination of dead reckoning, hydrodynamic models, inertial navigation systems, doppler velocity logs, and/or acoustic baseline navigation systems \citep{paull_auv_2014}.
+Much progress has been made in underwater vehicle navigation, which relies on any combination of dead reckoning, hydrodynamic models, inertial navigation systems, doppler velocity logs, and/or acoustic baseline navigation systems \citep{paull_auv_2014}.
 In particular, the long baseline (LBL) navigation system is the most similar in scale and style to GPS, and most appropriate for mitigating drift in underwater vehicle localization without overburdening computation on the vehicle \citep{van_uffelen_global_2021}.
@@
 Positioning is the ability to accurately determine one's location and is generally done via trilateration.
-Real-time navigation efforts assume a locally homogeneous sound speed profile (SSP) that enables a linear, deterministic scaling between travel time and range \citep{eustice_recent_2006,eustice_experimental_2007,webster_preliminary_2009,webster_advances_2012}.
-Short range (less than ten kilometer) navigation efforts accurately rely on a nominal sound speed value and achieve minimal navigation error \cite{eustice_experimental_2007,webster_preliminary_2009,kepper_mems_2017}.
-For mesoscale (tens to hundreds of kilometers) navigation or localization efforts, a deterministic \citep{graupe_preliminary_2019} or through-the-sensor \citep{webster_towards_2015} value for sound speed is used.
+Navigation assumes a degree of autonomy to move from a current position to a desired one.
+The standard for real-time navigation efforts assume a locally homogeneous sound speed profile (SSP) that enables a linear, deterministic scaling between travel time and range \citep{eustice_recent_2006,eustice_experimental_2007,webster_preliminary_2009,webster_advances_2012}.
+Short range (less than ten kilometers) navigation efforts achieve minimal navigation error with a nominal sound speed value \citep{eustice_experimental_2007,webster_preliminary_2009,kepper_mems_2017}.
+For mesoscale (tens to hundreds of kilometers) navigation efforts, a deterministic \citep{graupe_preliminary_2019} or through-the-sensor \citep{webster_towards_2015} value for sound speed is used, but error remains in the hundreds of meters.
 
 Post-processing localization efforts (re-navigation for a self-determinedly moving object or re-positioning for a passively moving one), without computational or time constraints, can indulge more complex acoustic methods and/or ocean representations.
-There are many different approaches, from accounting for a depth dependent SSP \citep{van_uffelen_localization_2016}, including spatio-temporal variability in the sound speed profile to minimize error \citep{graupe_preliminary_2019, mikhalevsky_deep_2020}, comparing acoustic records with synthetic acoustic records computed through an ocean model \citep{wu_deep_2019}, or isolating a specific ray path to find a representative group velocity \citep{mikhalevsky_deep_2020}.
-The ``cold start'' algorithm developed by \citet{mikhalevsky_deep_2020} re-positioned a floating hydrophone array with an error of 85 m and a standard deviation of 32 m relative to ground truth GPS data, but implementing this algorithm for navigation remains to be seen.
+There are many different approaches, from considering a temporally averaged mean depth dependent SSP \citep{van_uffelen_localization_2016}, including spatio-temporal variability in the sound speed profile to minimize error \citep{graupe_preliminary_2019, mikhalevsky_deep_2020}, comparing acoustic records with synthetic acoustic records computed through an ocean model \citep{wu_deep_2019}.
+The ``cold start'' algorithm developed by \citet{mikhalevsky_deep_2020} isolates a specific ray path to find a representative group velocity \citep{mikhalevsky_deep_2020} and re-positioned a floating hydrophone array with an error of 85 m and a standard deviation of 32 m relative to ground truth GPS data, but implementing this algorithm for navigation remains to be seen.
+These post-processing efforts have generally been deemed too computationally intensive and/or labor intensive for real-time efforts, where automation of identifying time arrivals is further complicated by multipath.
 
@@
 
-However, the key assumptions for OWTT ranging\textemdash a locally homogeneous sound speed profile that enables a linear, deterministic scaling between travel time and range\textemdash did not hold during a vehicle deployment in ICEX16, in the Beaufort Sea, in March 2016.
+Underwater navigation research is broadly motivated by GPS-like navigation in GPS-denied conditions.
+The Arctic, while remote, is the perfect place to test mature navigation technologies in real GPS-denied conditions.
 
-However, the new Arctic ... [Beaufort Lens]
+Literature and historical reviews of underwater navigation approaches in under-ice conditions highlight the emerging technology learning curve for vehicle operations in the Arctic, the hostile conditions brought on by a semi-permanent moving ice cover, and the need for both absolute and relative navigation with little to no tolerance for error \citep{norgren_unmanned_2014,mcfarlane_autonomous_2015,barker_scientific_2020}.
 
-The literature in underwater acoustic navigation and positioning is either real-time or physics-based.
-AUV operations in the Beaufort Sea necessitate a real-time and physics-based method to overcome the unique challenges of total ice cover, multiple ice floes, and the Beaufort Lens.
+One of the first successful AUV launches and recoveries involved equipping the vehicle with barbs which became entangled into a homing net underneath the ice \citep{bellingham_auv_1993}.
+Since then, AUVs have been used for ocean bottom \citep{jakuba_long-baseline_2008,kunz_deep_2008}, under-ice \citep{kaminski_12_2010}, and hydrographic surveys \citep{kukulya_under-ice_2010,kukulya_development_2016,stevens_linking_2016,plueddemann_autonomous_2012}.
+These efforts show a combination of short and long range positioning systems, where all acoustic navigation methods used a depth-averaged sound speed for linear scaling between recorded travel time and range.
+In many cases robust quantitative navigation performance could not be evaluated due to small operational scale and/or moving ice floes.
+But most importantly, all of these efforts experienced a nearly isovelocity or classical upward refracting Arctic acoustic environment.
 
-The paper is organized as follows: Section~\ref{sec:2} presents the physical
-experimental setup and data processing simplifications; Section~\ref{sec:3} presents key results for group velocity predictions and range estimations; Section~\ref{sec:4} closes with a discussion.
+The new Arctic\textemdash one with much less multi-year sea ice\textemdash is experiencing significant changes in the water column as well \citep{mackinnon_warm_2021}.
+The Beaufort Lens is a highly dynamic, highly stratified water mass, keeping a neutrally buoyant layer of warm Pacific water at 50 to 70 m in depth.
+Surrounded by cooler waters, it distinguishes a local maximum in the SSP, creating an upper and lower duct\textemdash the upper duct degrades signal coherence due to intensified ice interaction and the lower duct effectively traps sound for long range propagation \citep{poulsen_acoustic_2016}.
+Modeling output \citep{duda_long-range_2019,duda_effects_2021} and experimental observations \cite{kucukosmanoglu_observations_2021,badiey_azimuthal_2019,ballard_temporal_2020} indicate a high degree of spatio-temporal variability of the lower duct throughout the Chukchki and Beaufort Seas.
+
+An experiment in 2014 exploited this lower duct to provide acoustic navigation and communication at mesoscale ranges in the Canada Basin.
+\citet{freitag_long_2015} and \citet{webster_towards_2015} deployed multiple GPS-linked ice buoys with transducers that were capable of communications greater than 400 km with a range accuracy of 40 m when the sound speed was known.
+The transducers were at at 100 m depth, well-positioned in the lower duct spanning 50 to 200 m.
+The glider navigation accuracy, however, was severely degraded as acoustic packets were sent when the gliders were outside the sound channel.
+When gliders were deployed again in 2017, their navigation error was on the order of half a kilometer, and a glider re-navigation using an acoustic arrival method and the average sound-speed profile from the gliders reduced navigation error by a factor of 4-5, depending on the dive \citep{graupe_preliminary_2019}.
+Navigation error is referenced to intermittent GPS fixes, as all of these glider deployments occurred in the Marginal Ice Zone with little to no ice at the time.
+
+\subsection{Ice Exercise 2020}
+
+The Beaufort Lens is equally important for shorter ranges as well, where a shadow zone over vehicle operational ranges is created by a drastic change in the reliable acoustic path \citep{schmidt_acoustic_2016}.
+Thus the key simplifying assumption for OWTT ranging\textemdash a locally homogeneous sound speed profile that enables a linear, deterministic scaling between travel time and range\textemdash becomes more incorrect due to increased multipath in the upper duct of the Beaufort Lens.
+
+The results from this paper derive from data taken while deploying AUV Macrura, a modified Bluefin-21, during the Ice Exercise 2020 (ICEX20), in the Beaufort Sea, in March 2020.
+The challenges anticipated during ICEX20\textemdash total ice cover, moving ice floes, and a variable Beaufort Lens\textemdash crystallized the necessity of environmental awareness for the navigation solution.
+
+In this paper we present a working solution for underwater navigation that achieves GPS-like accuracy and precision.
+We embed a stochastic prediction of a single group velocity to convert recorded travel time into range.
+This prediction ingests an updatable snapshot of the SSP via acoustic message; over the course of ICEX20, this capability was used several times. 
+Ultimately, integrating environmental realism enabled a successful emergency recovery of the AUV.
+Due to a harddrive error, the AUV stalled underneath the ice surface and broadcasted its location.
+A team was able to dig a hole through the 2 meter ice sheet and immediately see the vehicle.
+Its successful recovery amidst total ice cover is shown in Fig. \ref{fig:vehicleRecovery}.
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=\reprintcolumnwidth]{figs/vehicleRecovery.jpg}
+	\caption{The successful emergency recovery of AUV Macrura, with the the auger used to drill through the ice sheet in the background. Photo courtesy of LT CDR D. Goodwin.}
+	\label{fig:vehicleRecovery}
+\end{figure}
+
+The rest of this paper is organized as follows: Section~\ref{sec:methods} presents the physical
+experimental setup and model-based data processing; 
+section \ref{sec:realtime} quickly conveys real-time results and the limitations of the data collected; section~\ref{sec:postprocess} presents key results for group velocity predictions and range estimations; section~\ref{sec:discussion} closes with a discussion on importance, future work, and broader relevance.
 
@@
 % =========================================================================== %
-\section{\label{sec:2} Methods}
+\section{\label{sec:methods} Methods}
 
-The operational solution for ICEX20 relied on an Integrated Communication and Navigation Network (ICNN) \citep{schneider_self-adapting_2020,randeni_construction_2020,randeni_high-resolution_2021}.
+The acoustic communication dataset accrued for ICEX20 relied on the Integrated Communication and Navigation Network (ICNN) \citep{schneider_self-adapting_2020,randeni_construction_2020,randeni_high-resolution_2021}.
 The ICNN was initially developed via numerous virtual experiments, to push the boundaries of algorithms and interfaces between different hardware components.
@@
 	\includegraphics[width=\reprintcolumnwidth]{figs/ICNN_Overview}
+	\caption{A cartoon overview of the Integrated Communication and Navigation Network (ICNN), which provides joint data and tracking information between AUV and a human decision maker at Topside.}
 	\label{fig:icnnOverview}
-	\caption{CAPTION}
 \end{figure}
@@
 \item Four ice buoys, each outfitted with a WHOI Micro-Modem, receive messages from the AUV and send that information over freewave radio to a Topside computer
-\item The topside computer converts travel times into range estimates using a stochastic embedded prediction of the horizontal group velocity via BELLHOP ray tracing code \citep{porter_bellhop_2011}
+\item The topside computer converts travel times into range estimates using a stochastic embedded prediction of the horizontal group velocity via BELLHOP ray tracing code \citep{porter_bellhop_2011} coupled with a Virtual Ocean \citep{schneider_netsim_2018}
 \item The topside computer calculates a new position by trilaterating the range estimates
-\item The position differential is broadcast to the vehicle to update its navigation solution
+\item The position differential, not the absolute position, is broadcast to the vehicle to update its navigation solution and be robust to latency and intermittency
 \end{enumerate}
@@
 This paper specifically addresses the third step\textemdash the topside computer's conversion of travel time into range, which underpins the trilateration in the ICNN, and is an often overlooked element of any underwater acoustic navigation system.
-The results accordingly focus on validating acoustic range estimates from communication events between GPS-tracked beacons.
+The analysis accordingly focuses on validating acoustic range estimates from communication events between GPS-tracked beacons, not the AUV itself.
+This allows the results to be decoupled from all the other moving parts in the ICNN.
 This ``modem experiment'' was conducted while the vehicle was still being prepped to go under the ice.
@@
 
-Because the navigation solution on the vehicle, during a mission, can only be evaluated on the basis of the error estimates sent, a sister experiment for validating the real-time ranging approach was implemented.
+Because the navigation solution on the vehicle during a mission can only be evaluated on the basis of the error estimates sent, a sister experiment for validating the real-time ranging approach was implemented.
 The ice buoy modems were run as ``virtual vehicles'' at a given depth and would transmit to all other ice buoys.
 The internal sound speed estimate was then toggled using a handful of weights applied to a basis function representation of the sound speed.
-This trial was configured for as many ice buoy modems as possible, with each experiment running for roughly an hour between the two sound speed modalities.
+This trial was configured for as many ice buoy modems as possible with the time allotted, with each experiment running for roughly an hour between the two sound speed modalities.
 There are three important engineering choices in the ICNN network to contextualize this dataset:
@@
 Given the complexity of the ICNN system, this experiment did not collect an exhaustive set of data across all buoy, source depth, receive layer, and sound speed combinations, as this time was also used to verify and resolve any hardware issues in the modems themselves.
-
 Importantly, only one modem at a time could run the vehicle behavior to trigger group velocity calculations; other modems that were transmitting simultaneously recorded events as modems, not as vehicles.
 
-The dataset introduced in this thesis has parallel physical and virtual layer; given the small operational scale, on the order of kilometers, and a singular SSP, both layers are projected into a range independent coordinate system for ease of comparison.
+A key construction in the ICNN are parallel physical and virtual layers; given the small operational scale, on the order of kilometers, and a singular SSP, both layers are projected into a range independent coordinate system for ease of comparison.
 
 \subsubsection{Physical layer}
-The physical layer, i.e.
-the acoustic link through the real ocean, describes the beacon to beacon communication events through the water column; it covers all source depths (20, 30, and 90 m) and both receive layers (30 and 90 m), where the surface expressions of the beacons are tracked by GPS.
+The physical layer describes the acoustic link between beacons in the water column.
+It covers all source depths (20, 30, and 90 m) and both receive layers (30 and 90 m), where the surface expressions of the beacons are tracked by GPS.
 Events are sorted by source depth and sound speed profile, given that these are the dominating factor for the location and span of the shadow zone compared to any temporal or spatial scales of variability.
@@
   \centering
-  \includegraphics[width=.50\textwidth]{figs/modem-chart.pdf} \hfill
+  \includegraphics[width=.51\textwidth]{figs/modem-chart.pdf} \hfill
   \includegraphics[width=.48\textwidth]{figs/zs20-birdseye.pdf} \\
@@
 \subsubsection{Virtual layer}
-The virtual layer describes the BELLHOP simulations for each communication event in the physical layer, where a handful of weights applied to a basis function representation can reconstitute a number of realistic candidate SSPs \citep{leblanc_underwater_1980,lin_merging_2010}.
-The SSP is the only configurable input to the group velocity prediction.
+The virtual layer describes the BELLHOP simulations for each communication event in the physical layer.
+The SSP used for the raytracing is driven by a handful of weights applied to a basis function representation
+of the SSP variability to reconstitute a number of realistic candidate SSPs \citep{leblanc_underwater_1980,lin_merging_2010}.
+The SSP is the only human configurable input to the group velocity prediction.
 
@@
 an optimal set of weights chosen by a human decision maker via an interactive Tactical Decision Aid framework \citep{bhatt_embedded_2021}.
-For the post-processing analysis, we add the SSP from HYCOM for comparison, to mirror the information available on a submarine\textemdash modeled data, historical data, and \textit{in situ} data.
+For the post-processing analysis, we add the SSP from HYCOM for comparison, to mirror the information available on a submarine\textemdash modeled data, historical data, and \textit{in situ} data (personal conversation with LT B.
+Howard and LT CDR D.
+Goodwin).
 This order corresponds to increasing ducted conditions of the Beaufort Lens.
 
-Fig. \ref{fig:raytrace-zs30} shows eigenrays in a range independent coordinate system for all three sound speed environments for a source depth of 30 m.
+Fig. \ref{fig:raytrace-zs30} shows eigenrays in a range independent coordinate system for all three sound speed environments, for a source depth of 30 m.
 It is important to note that BELLHOP produces an abundance of eigenrays; the ones displayed here are chosen by travel time proximity to the recorded data.
-The important pattern is the increased complexity in multipath via surface interactions as ducted conditions increase.
+With this filter, the chosen eigenrays show a greater amount of surface interactions as the SSP duct conditions increase.
 
@@
 The use of the acoustic modem network for tracking relies on the accurate estimates of travel times between the submerged platform and range modems, supported by clock synchronization and a pre-determined scheduling of acoustic events.
-For the Beaufort Lens, the strong multipath effects make it virtually impossible to deterministically predict the modem triggering time.
+For the Beaufort Lens in particular, the strong multipath effects make it virtually impossible to deterministically predict the modem triggering time.
 
-\subsubsection{Minimal bounce criteria}
+\subsubsection{Minimal bounce criteria (MBC)}
 Instead, for each individual node $i$, an embedded stochastic tracking framework is used to provide a running estimate of the horizontal group velocity $u_{i,j}$ for the conversion from travel time to range from modem $j$, with the ultimate goal of matching the naive group velocity, i.e.
@@
 
-The initial call to BELLHOP is over a sparse, dynamically created depth and range grid, of which the plane wave approximation is used to bring the ray to the range and depth posited by the onboard tracking solution.
+The initial call to BELLHOP is over a sparse local grid, of which the plane wave approximation is used to bring the ray to the range and depth posited by the onboard tracking solution.
+For each grid point, BELLHOP will produce a number of arrivals resulting from multiple propagation paths for any source-receiver pair.
 Using only the $N_0$ rays with neither surface nor bottom bounces, it will then estimate the current group velocity $u$ from a power weighted average of the ray travel times,
@@
 However, this high density of launch angles creates a greedy estimator of direct path eigenrays, and this simulation pipeline almost always guarantees a direct path found.
-The MBC improves upon a deterministic estimate in that it considers the ray path, travel time, and amplitude to predict a realistic group velocity.
+The MBC improves upon a deterministic estimate in that it considers the simplest ray path, travel time, and amplitude to predict a realistic group velocity.
 
@@
 \ref{fig:raytrace-zs30}, an acoustic arrival does not always take the direct path from source to the receiver, and the path it takes is extremely dependent on the sound speed profile.
-Often, BELLHOP will produce a number of arrivals resulting from multiple propagation paths for any source-receiver pair.
 The suggested algorithm, the nearest bounce criteria (NBC), is a slight modification from the minimal bounce criteria, and includes multipath as a new dimension of information with negligible additional computation.
-This metric was introduced in post-processing.
+This metric was introduced in post-processing and does not include the plane wave approximation.
 
@@
 This method selects a different group velocity based on the multipath arrival structure, as the expected arrival is not always the first arrival and could even be masked by noise or blocked temporarily \citep{deffenbaugh_acoustic_1996}.
+We manually cap the number of bounces to 4 because of the smaller operational scale and the attenuation accrued with many surface interactions.
+Bottom bounces are not encoded separately because of ray's tendency to refract upward, not due to information limitations. 
+In addition, for a more consistent comparison, the local sparse grid is standardized as 11$x$11 points over 10 m in range and 20 m in depth, which is similar to those from the vehicle logs.
+The extra size in depth addresses the computational limitations of how time-bounded eigenrays seem to stack vertically in the water column, as shown in Fig \ref{fig:rays3env-zs30}.
+Lastly, to populate the multipath criterion effectively with a multitude of arrivals per number of surface bounces, we abandon the plane-wave extension to the interpolated receiver location.
 
@@
 \FloatBarrier
-\section{\label{sec:4} Real-time psuedorange results}
+\section{\label{sec:realtime} Real-time psuedorange results}
 
@@
 As expected, the algorithm is generally overestimating range as it resolves a direct path that does not represent the actual arrival.
-This section examines the absolute range error between the psuedorange and the GPS derived range with the caveat that the data collected is fairly asymmetrical across source depths and sound speed conditions.
+This section examines the absolute range error between the psuedorange and the GPS derived range with the caveat that the data collected is fairly unbalanced across source depths and sound speed conditions.
 
@@
 This is likely driven by the prominence of the duct.
+
 The overarching statistical results show that both sound speed environments, given the smaller operational scales and shallow depths, are accurate enough to approximate the group velocity structure given the local grid.
-Of course, overall statistics obscure the relationship between range error and range (or OWTT as the proxy for range).
+Indeed, the error statistics indicated here are similar to that of the ICNN \citep{randeni_high-resolution_2021} and accurate enough to have facilitated a successful vehicle recovery.
 
 \FloatBarrier
-\section{\label{sec:5} Post-processing psuedorange results}
+\section{\label{sec:postprocess} Post-processing psuedorange results}
 
@@
 
-The value for the extrapolated range, $\hat{r}$, is only tracked when the modem runs the vehicle behavior; thus we replace $\hat{r}$ with the GPS-tracked range for all modem events.
+It is important to note that the value for the extrapolated range, $\hat{r}$, is only tracked when the modem runs the vehicle behavior; thus we replace $\hat{r}$ with the GPS-tracked range for all modem events.
 Because $\hat{r}$ converges to the correct solution, a comparison of $\hat{r}$ with the GPS-tracked range shows a normal, zero-centered distribution within the bounds of GPS drift.
-The analysis here therefore seeds realistic but ``omniscient'' knowledge of the range and leverages the post-processing pipeline to more thoroughly evaluate the acoustic range estimate for all modem events, with three relevant sound speed sources, and both group velocity criterion.
-Accordingly, the results in this section evaluate the utility of the algorithms and sound speed sources, divorced from their role in the ICNN.
+The analysis here therefore seeds realistic but ``omniscient'' knowledge of the extrapolated range and leverages the post-processing pipeline to more thoroughly evaluate the acoustic range estimate for all modem events, with three relevant sound speed sources, and both group velocity criterion.
+Accordingly, the results in this section evaluate the utility of the algorithms and sound speed sources, divorced from their role in the ICNN while maintaining real-time relevance.
 
@@
 
-The minimal and nearest bounce algorithm is applied using the three sound speed inputs shown in Fig.~\ref{fig:raytrace-zs30}; the resulting predicted group velocities for a source depth of 30 m are shown in Fig~\ref{fig:gvel30}.
+The minimal and nearest bounce algorithms are applied, both without the plane wave extensions, with the three sound speed inputs shown in Fig.~\ref{fig:raytrace-zs30}.
+The resulting predicted group velocities for a source depth of 30 m are shown in Fig~\ref{fig:gvel30}.
 The takeaways from this source depth are representative of those from 20 and 90 m as well (see Supplementary Fig.~\ref{fig:gvelMore}).
@@
 Because direct path dominates the predictions from the MBC, group velocity estimates are greater than those from NBC.
-The one instance this is not true is important to mention\textemdash for a travel time of 2.2 s, where the minimal multipath is dominated by a single bottom bounce, but the eigenray traces show two, three, and four bounces for HYCOM, baseline, and weighted SSPs, respectively (the red line to the furthest transmission in Fig.
-\ref{fig:raytrace-zs30}).
+The one instance this is not true is important to mention\textemdash for a travel time of 2.2 s, where the minimal multipath is dominated by a single bottom bounce (the red line to the furthest transmission in Fig.
+\ref{fig:raytrace-zs30}, more detail in Section \ref{subsec:anomalousBeacon}).
 The nearest bounce method never classifies anything as direct path for the 30 m source, though BELLHOP is quite adept at finding those rays.
@@
 For a 30 m receiver depth, the NBC shows more overlap with data-derived values as it classifies multipath more correctly.
-For a 90 m receiver depth, the overlap is less accurate due to computational constraints.
-Given a limited fan of rays entering the shadow zone, there is a less reliable simulated time arrival packet.
+For a 90 m receiver depth, the overlap is less accurate due to computational constraints of a limited fan of rays entering the shadow zone rendering a less reliable simulated time arrival packet.
 
 Overall, analysis shows that the higher multipath classification produces more accurate range estimates, likely driven by a tighter bundle of rays.
-However, the data is too small to draw significant conclusions.
-Discontinuities in multipath classification verify our hypothesis for its importance for a smoothly varying group velocity, as shown in the cluster where HYCOM jumps from one to two bounces amidst the baseline SSP seeing two bounces and the weighted SSP seeing three.
+However, that data is too small to draw significant conclusions.
+Discontinuities in multipath classification verify our hypothesis for its importance to a smoothly varying group velocity, as shown in the cluster for a receiver depth of 30 m, where HYCOM jumps from one to two bounces amidst the baseline SSP and weighted SSP smoothly increasing while consistently seeing two and three bounces, respectively.
 
@@
 
-When using a linear scaling to convert travel time into range, any offset between the assumed sound speed and the horizontal group velocity produces unconstrained error with increasing receiver distance.
-Figure \ref{fig:rangeError} shows the directional range error footprints for all three sound speed inputs with respect to OWTT, separated by source and receiver depth configurations.
+Fig. \ref{fig:rangeError} shows the directional range error footprints for all three sound speed inputs with respect to OWTT, separated by source and receiver depth configurations.
 The weighted SSP range error generally has the smallest and most zero-centered footprint.
 The one case it does not is for the source-receiver pairings between 30 and 90 m in depth.
-The increased error for these reciprocal transmission paths is most likely driven by the computational artifacts propagating through the steep sound speed gradients of the lens, through a shadow zone.
+The increased error for these reciprocal transmission paths is most likely driven by the computational artifacts propagating through the steep sound speed gradients of the lens and through the shadow zone.
 All other source depth pairings are significantly improved using the chosen weights compared to HYCOM or the baseline.
+
+When using a linear scaling to convert travel time into range, any offset between the assumed sound speed and the horizontal group velocity produces unconstrained error with increasing receiver distance.
 Most importantly, we see the consequences of the adaptive group velocity in that range error does not strictly increase with OWTT.
@@
 
-Figure \ref{fig:compareV1V2} is a crossmap for all three SSPs for both group velocity predictors, with the minimal bounce on the x-axis and the nearest bounce on the y-axis; the important range estimation metrics are summarized in table \ref{tab:rangeV1V2}.
+Figure \ref{fig:compareV1V2} is a crossmap for all three SSPs for both group velocity predictors, with the minimal bounce on the x-axis and the nearest bounce on the y-axis; the important range estimation metrics are summarized in table \ref{tab:rangeErrorV1V2}.
 
-The first thing to notice is that all the events are on the right hand side of the plots\textemdash the MBC has a systemic overestimation, which is reduced via NBC as events straddle the x-axis.
+All the events are on the right hand side of the plots\textemdash the MBC has a systemic overestimation, which is reduced via NBC as events straddle the x-axis.
 The improvement from MBC to NBC is most evident for the realistic sound speed; while the HYCOM SSP improves from a median absolute range error of 6.41 to 4.61 m, the baseline SSP improves from 10.30 to 2.27 m, and the weighted SSP improves from 13.28 to 2.12 m.
@@
 The order of magnitude improvement in the ducted SSPs demonstrate the effectiveness of the algorithm exploiting the multipath conditions.
-The gridded approach and multipath classification build in the computational cushion for minimal error and minimal bias compared to direct eigenray assessment.
+
+Given the computational constraints of real-time modeling, the gridded approach facilitates enough multipath classification to build in a ``ray ensemble'' of characteristic group velocities.
+This result is not always possible when aiming to find eigenrays to just an individual point, even with a higher density of launch angles.
+An important takeaway for those interested in ray-based localization is leveraging a local grid to give BELLHOP tolerance for finding solutions that otherwise may not be found in a center or single point solution.
+The limitations of numerical computation, particularly for a complex environment, are more adeptly addressed by accepting some uncertainty in position to than prescribing an exact solution.
+Even though BELLHOP is a greedy estimator for eigenrays, the additional data created is a negligible burden.
 
@@
 All plots show the same 45 degree banding of source depth and multipath structure, indicating systemic GPS noise.
-Even with poorer accuracy from the GPS solution, the acoustic methods remain high precision.
-Some bands are adjacent to others of the same styling, likely due to ice drift.
+The width of those bands indicates the high precision of the acoustic methods relative to lower precision from GPS.
+Some bands are adjacent to others of the same styling, likely due to ice drift and/or various GPS satellite constellations.
 
@@
 \subsection{Further analysis of the anomalous beacon to beacon connection}
+\label{subsec:anomalousBeacon}
 
@@
 The large error is driven by the MBC unexpectedly defaulting to a bottom bounce with a much greater OWTT.
-Thus the NBC ``correctly'' classifies the multipath as 4 bounces, reducing the range error from greater than a kilometer to less than a meter.
-While there is no actual way of knowing if this is the correct multipath structure, the range error is remarkably small.
+The NBC classifies the multipath as 4 bounces, reducing the range error from greater than a kilometer to less than a meter.
+While there is no actual way of knowing if this is the correct multipath structure, the range error is remarkably small, at 0.025\%. 
 This pattern of not choosing the minimal observed bounce structure is consistent across all SSPs; the baseline goes from 1 to 3 bounces and HYCOM goes from 0 to 2 bounces.
@@
 Thus, for acoustically complex environments, the NBC has a disproportionately positive impact as the estimated SSP approaches the desired SSP.
-
-In comparison to the earlier eigenray traces, the results here distill the importance of the gridded approach to curate multipath propagation. 
-The grid gives BELLHOP tolerance for finding solutions that otherwise may not be found in a center or single point solution; the limitations of numerical computation, especially in a real-time environment, are more adeptly addressed by baking in a small amount of uncertainty than trying to prescribe an exact solution.
-This creates more data to sort through given that BELLHOP is a greedy estimator of eigenrays, but is computationally simpler than running a dense ray fan for precise eigenray finding.
-
-The NBC algorithm, coupled with sound speed fidelity, work in concert to mitigate error.
-Either one alone is not enough for the Beaufort Lens, as the environmental realism begins to match the acoustic conditions of intense multipath.
 
@@
 
-The ``Global Positioning System Standard Positioning Service Performance Standard'', published by the Department of Defense in Apr 2020 (CITE GPS TECHNICAL STANDARD) claims that well-designed GPS receivers have been achieving a horizontal accuracy of 3 meters and a vertical accuracy of 5 meters, 95\% of the time.
+The ``Global Positioning System Standard Positioning Service Performance Standard'', published by the Department of Defense in Apr 2020 claims that well-designed GPS receivers have been achieving a horizontal accuracy of 3 meters and a vertical accuracy of 5 meters, 95\% of the time.
 But ``well-designed'' is an ambiguous enough term that obfuscates two confounding factors for high latitude GPS drift\textemdash operational limitations and ionospheric scattering.
@@
 In the ionosphere, electromagnetic signals are affected by the total amount of electrons encountered by the signal; the size of the signal delay is dependent on the carrier frequencies.
-The normal signal delay creates a GPS error of 5 to 15 meters during the daytime and 1 to 3 meters at night (CITATION).
 The literature even references the total amount of electrons encountered by the ``GPS raypath'' as a means of explaining signal fluctuations \citep{themens_nature_2015}.
@@
 
-In a humbling twist, at polar latitudes, GNSS systems experience similar operational, technical, and environmental challenges that we observe in underwater acoustic communications; the underwater acoustics community generally seems to ignore these challenges when validating localization and navigation efforts.
-
+In a humbling twist, at polar latitudes, GNSS systems experience similar operational, technical, and environmental challenges that we observe in underwater acoustic communications.
 Figure \ref{fig:gps-drift-example} highlights the GPS and observed OWTT drift relative to the ice movement for two pairs of modem buoy connections.
@@
 % =========================================================================== %
-\section{\label{sec:4} Discussion}
+\section{\label{sec:discussion} Discussion}
 
@@
 Current approaches in real-time underwater acoustic navigation simplify the non-linear relationship between a sound speed profile and acoustic propagation with a deterministic sound speed.
-Some post-processing approaches attempt to leverage the acoustic arrival structure via various ray methods, but often use a singular SSP for long-term deployments.
+Some post-processing approaches attempt to leverage the acoustic arrival structure via various ray methods, but often use a singular SSP for simplicity, even over long term missions or dynamic conditions.
 Thus, the conversion from travel time to range, particularly for real-time vehicle deployments, can be pre-conditioned for error growth as the OWTT/range increases.
-For GPS-denied navigation, and hostile environments like the Arctic, tolerance for error is close to none.
+
+For GPS-denied navigation, especially in hostile environments like the Arctic, tolerance for error is close to none.
 This work addresses a critical need in acoustic navigation by retooling acoustic arrival methods generally deemed too complex or labor intensive for real-time, ray-based range estimation to achieve GPS-like positioning.
 
-We hypothesize and validate that the embedded stochastic prediction of a single group velocity is a smoothly varying function of range, source and receiver depth pairings, and multipath structure.
+We hypothesize and validate that the embedded stochastic prediction of a single group velocity is a smoothly varying function of range, source and receiver depth pairings, as well as multipath structure.
 We employ a GPS-tracked experiment to have a ground truth comparison for real-time localization algorithms.
@@
 
-This work, to our knowledge, is the first successful field deployment of a real-time, model-based acoustic ranging system for underwater navigation.
+The literature in underwater acoustic navigation and positioning is either real-time or physics-based.
+In this paper we demonstrate a field-tested approach that, by coupling data streams with fast acoustic modeling, is both real-time and physics-based.
 The methods exploit the upward refracting nature and the total ice cover of the Arctic environment to achieve remarkable ranging accuracy and precision.
@@
 BELLHOP simulations provide other relevant eigenray information, like time and angle of arrival, that is ripe for statistical and machine learning methods to classify a representative group velocity.
-A bespoke and fast ray tracing method, which was used during ICEX20 to evaluate the acoustic fit of sound speed profiles, can easily report back the number of turning points instead of the number of bounces for multipath classification.
+A bespoke and fast ray tracing method, which was used during ICEX20 to evaluate the acoustic fit of sound speed profile parameterization \citep{bhatt_embedded_2021}, can easily report back the number of turning points instead of the number of bounces for multipath classification.
 
@@
 Buoyancy fluctuations in these regions threaten even simple AUV tasks like following a trackline.
-Acoustic communications are further complicated given a shallow environment with significant scattering \citep{lavery_measurements_2010,ross_acoustic_2012,lavery_broadband_2013}, where fast timefront modeling may only be coherent for trivial probabilities of the ocean state.
+Acoustic communications are further complicated given a shallow environment with significant scattering \citep{lavery_measurements_2010,ross_acoustic_2012,lavery_broadband_2013}, where fast acoustic modeling may only be coherent for trivial probabilities of the ocean state.
 Realistic \textit{in situ} considerations of the acoustic environment may not be possible in such environments without complete through-the-sensor integration of echosounder data and/or a hyper-realistic onboard ocean model.
@@
 But there are overwhelming challenges, operationally and computationally, for fast, mobile tomography to become a realistic endeavor.
-Addressing the spatial and temporal scales of what can be solved deterministically and what must be solved stochastically imposes a resolution constraint on the utility of gridded models that already dwarfs that of AUV operations.
+Addressing the spatial and temporal scales of what can be solved deterministically and what must be solved stochastically imposes a resolution constraint on the utility of gridded models\textemdash resolving fine features inaccurately (or with a false sense of confidence) could be more harmful than assuming range independence.
 Given that AUV operations are often on small spatial and temporal scales, the added benefit of a gridded model is quite small, and in cases like the Arctic, may actually mischaracterize the ocean volume.
-
-For gliders, with longer and larger operational scales, an ocean model may provide useful information.
+For gliders, with longer and larger operational scales, an ocean model may provide more useful information.
 Currently gliders are low power and do not have the storage or computational power to run a full-scale, realistic ocean model.
-Of course, all of this is assuming that the ocean model is accurate and has high enough resolution to resolve ocean features at small enough scales, which is in of itself an active research field.
-
-Ocean model realism (especially at finer scales) correlates with computational power and time.
-Thus, until a full-scale ocean model can be run in real-time on a small computer like a Raspberry Pi, the framework proposed in this thesis strikes a balance for the necessary environmental realism as depth-dependent with temporal updates.
-This has a cascading effect on acoustic realism, currently a range-independent scheme.
-Faster, more effective computational methods (and hardware devices) for acoustic propagation can certainly improve the real-time joint estimation problem, but are still constrained by this depth-dependent snapshot of the ocean volume.
-Until we reach that faraway goal, an updatable, range independent scheme, as employed in this work (see Section...) is a practical and effective approach.
+A lightweight representation of the key environmental and acoustic features, passed through the same manner of acoustic message from the modem experiment, may drastically improve glider navigation.
 
@@
 As anthropogenic climate change renders the region more accessible, safe navigation infrastructure and practices will be increasingly vital for industry, tourism, commercial, and military operations.
+We hope this work enables more frequent and confident monitoring for remote and/or dangerous regions of high climate security importance.
 
@@
 
-% \subsubsection{Sample subsubsection\label{subsubsec:1}}
-
-% \paragraph{Sample paragraph}Here is text following the paragraph
-% heading.
-% Here is a figure reference: is shown in Fig.~\ref{fig:FIG1}.
-
-% Normal journal cite: \citep{joursamp1},
-%  Book reference \citet{booksamp1},
-% Computer language documentation:
-% \citep{sampcode2}.
-
-% Every \verb+\citep+  or \verb+\citet{}+ will produce a citation and an entry in the
-% bibliography. Every citation must have a matching entry in the
-% bibliography
-% database file (\verb+\filename.bib+).
-
-% Here is an example of algorithmic:
-
-% \begin{algorithmic}
-% \If {$i\geq maxval$}
-%     \State $i\gets 0$
-% \Else
-%     \If {$i+k\leq maxval$}
-%         \State $i\gets i+k$
-%     \EndIf
-% \EndIf
-% \end{algorithmic}
-
 \begin{acknowledgments}

### 2021-08-26 doc/main.tex
fix pseudo* spelling

@@
 An autonomous underwater vehicle deployment in March 2020 in the Beaufort Sea faced significant operational challenges with a total ice-covered environment and complex acoustic propagation via the Beaufort Lens.
-We implement an embedded, stochastic, ray-based prediction of the horizontal group velocity and validate its performance with communication events between GPS-aided beacons, achieving a psuedorange mean absolute error of 11 m.
+We implement an embedded, stochastic, ray-based prediction of the horizontal group velocity and validate its performance with communication events between GPS-aided beacons, achieving a pseudorange mean absolute error of 11 m.
 In post-processing, we introduce a multipath criterion and achieve ranging performance that rivals GPS.
@@
 
-Finally, the psuedorange is calculated simply as
+Finally, the pseudorange is calculated simply as
 \begin{equation}
@@
 \FloatBarrier
-\section{\label{sec:realtime} Real-time psuedorange results}
+\section{\label{sec:realtime} Real-time pseudorange results}
 
@@
 As expected, the algorithm is generally overestimating range as it resolves a direct path that does not represent the actual arrival.
-This section examines the absolute range error between the psuedorange and the GPS derived range with the caveat that the data collected is fairly unbalanced across source depths and sound speed conditions.
+This section examines the absolute range error between the pseudorange and the GPS derived range with the caveat that the data collected is fairly unbalanced across source depths and sound speed conditions.
 
@@
 \FloatBarrier
-\section{\label{sec:postprocess} Post-processing psuedorange results}
+\section{\label{sec:postprocess} Post-processing pseudorange results}
 
@@
 \FloatBarrier
-\subsection{Psuedorange error versus one way travel time}
+\subsection{Pseudorange error versus one way travel time}
 
@@
 \FloatBarrier
-\subsection{Psuedorange estimations from the nearest and minimal bounce criteria}
+\subsection{Pseudorange estimations from the nearest and minimal bounce criteria}
 
-The previous section displayed the psuedorange performance using the NBC.
+The previous section displayed the pseudorange performance using the NBC.
 Here, we explicitly compare the ranging accuracy between the nearest and minimal bounce criteria for all events in the post-processing pipeline.

### 2021-08-26 doc/main.tex
fix typos, very minor edits

@@
 \citet{freitag_long_2015} and \citet{webster_towards_2015} deployed multiple GPS-linked ice buoys with transducers that were capable of communications greater than 400 km with a range accuracy of 40 m when the sound speed was known.
-The transducers were at at 100 m depth, well-positioned in the lower duct spanning 50 to 200 m.
+The transducers were at 100 m depth, well-positioned in the lower duct spanning 50 to 200 m.
 The glider navigation accuracy, however, was severely degraded as acoustic packets were sent when the gliders were outside the sound channel.
@@
 Bottom bounces are not encoded separately because of ray's tendency to refract upward, not due to information limitations. 
-In addition, for a more consistent comparison, the local sparse grid is standardized as 11$x$11 points over 10 m in range and 20 m in depth, which is similar to those from the vehicle logs.
+In addition, for a more consistent comparison, the local sparse grid is standardized as 11$\times$11 points over 10 m in range and 20 m in depth, which is similar to those from the vehicle logs.
 The extra size in depth addresses the computational limitations of how time-bounded eigenrays seem to stack vertically in the water column, as shown in Fig \ref{fig:rays3env-zs30}.
@@
 An important takeaway for those interested in ray-based localization is leveraging a local grid to give BELLHOP tolerance for finding solutions that otherwise may not be found in a center or single point solution.
-The limitations of numerical computation, particularly for a complex environment, are more adeptly addressed by accepting some uncertainty in position to than prescribing an exact solution.
+The limitations of numerical computation, particularly for a complex environment, are more adeptly addressed by accepting some uncertainty in position than by prescribing an exact solution.
 Even though BELLHOP is a greedy estimator for eigenrays, the additional data created is a negligible burden.
@@
 
-As shown in table \ref{tab:rangeErrorPost}, there is a striking maximum range error of 1491 m for weighted SSP in the minimal bounce criteria.
+As shown in table \ref{tab:rangeErrorPost}, there is a striking maximum range error of 1491 m for the weighted SSP in the minimal bounce criteria.
 This section isolates the events that encompass that outlier for further examination, summarized in table \ref{tab:rangeErrorPostExplanation}.
@@
 There is relative, i.e. non-rigid, ice movement between the North and East buoys, evidenced by events spanning the dashed line.
-But the height of the scatter plot is indicates the precision of the GPS signal, as it remains consistent across many arrival time bands.
+But the height of the scatter plot is indicative of the precision of the GPS signal, as it remains consistent across many arrival time bands.
 Naturally, some minor offsets between these vertical bands relate to different operational configurations of source and receiver depth.

### 2021-08-26 doc/main.tex
update title per Henrik's email
minor/general edits and suggestions

@@
-\title[JASA/draft]{Real-time Acoustic Ray Range Estimation for Underwater Vehicle Navigation}
+\title[JASA/draft]{Under-ice acoustic navigation using real-time model-aided range estimation}
 \author{EeShan C. Bhatt}
@@
 Autonomous underwater vehicles (AUVs) are an increasingly capable platform to explore and sample the ocean, particularly for remote and/or dangerous environments.
-Lightweight, mobile, and inexpensive compared to crewed- and ship-based systems, these vehicles can enable a safer, data-forward style of field oceanography \citep{bellingham_robotics_2007,petillot_underwater_2019}.
+Lightweight, mobile, and inexpensive compared to crewed and ship-based systems, these vehicles can enable a safer, data-forward style of field oceanography \citep{bellingham_robotics_2007,petillot_underwater_2019}.
 
@@
 There are many different approaches, from considering a temporally averaged mean depth dependent SSP \citep{van_uffelen_localization_2016}, including spatio-temporal variability in the sound speed profile to minimize error \citep{graupe_preliminary_2019, mikhalevsky_deep_2020}, comparing acoustic records with synthetic acoustic records computed through an ocean model \citep{wu_deep_2019}.
-The ``cold start'' algorithm developed by \citet{mikhalevsky_deep_2020} isolates a specific ray path to find a representative group velocity \citep{mikhalevsky_deep_2020} and re-positioned a floating hydrophone array with an error of 85 m and a standard deviation of 32 m relative to ground truth GPS data, but implementing this algorithm for navigation remains to be seen.
+The ``cold start'' algorithm developed by \citet{mikhalevsky_deep_2020} isolates a specific ray path to find a representative group velocity and re-positioned a floating hydrophone array with an error of 85 m and a standard deviation of 32 m relative to ground truth GPS data, but implementing this algorithm for navigation remains to be seen.
 These post-processing efforts have generally been deemed too computationally intensive and/or labor intensive for real-time efforts, where automation of identifying time arrivals is further complicated by multipath.
@@
 The Beaufort Lens is equally important for shorter ranges as well, where a shadow zone over vehicle operational ranges is created by a drastic change in the reliable acoustic path \citep{schmidt_acoustic_2016}.
-Thus the key simplifying assumption for OWTT ranging\textemdash a locally homogeneous sound speed profile that enables a linear, deterministic scaling between travel time and range\textemdash becomes more incorrect due to increased multipath in the upper duct of the Beaufort Lens.
+Thus the key simplifying assumption for OWTT ranging\textemdash a locally homogeneous sound speed profile that enables a linear, deterministic scaling between travel time and range\textemdash becomes more inaccurate due to increased multipath in the upper duct of the Beaufort Lens.
 
 The results from this paper derive from data taken while deploying AUV Macrura, a modified Bluefin-21, during the Ice Exercise 2020 (ICEX20), in the Beaufort Sea, in March 2020.
-The challenges anticipated during ICEX20\textemdash total ice cover, moving ice floes, and a variable Beaufort Lens\textemdash crystallized the necessity of environmental awareness for the navigation solution.
+The challenges anticipated for ICEX20\textemdash total ice cover, moving ice floes, and a variable Beaufort Lens\textemdash crystallized the necessity of environmental awareness for the navigation solution.
 
@@
 The simulation approach is a necessary testbed for robustness and produces better results than post-processing previous field data, as that restricts mission configurations to the data taken, which can hamper troubleshooting in the field.
-The simulation capabilities are largely physics-driven with a modular system of systems approach: an environmental simulator with sub-components for the ocean, Arctic ice, and ocean acoustics; a vehicle simulator with sub-components for vehicle dynamics and navigation; a topside hardware simulator and acoustic communications simulator, both with a software and hardware-in-the-loop version \citep{schneider_netsim_2018}.
+The simulation capabilities are largely physics-driven with a modular system of systems approach: an environmental simulator with sub-components for the ocean, including Arctic ice drift and ocean acoustic propagation; a vehicle simulator with sub-components for vehicle dynamics and navigation; a topside hardware simulator and acoustic communications simulator, both with a software-only configuration and a hardware-in-the-loop version \citep{schneider_netsim_2018}.
 The virtual environment similarly emulates the interfaces between the real components to test the entire software pipeline.
@@
 	\includegraphics[width=\reprintcolumnwidth]{figs/ICNN_Overview}
-	\caption{A cartoon overview of the Integrated Communication and Navigation Network (ICNN), which provides joint data and tracking information between AUV and a human decision maker at Topside.}
+	\caption{A schematic overview of the Integrated Communication and Navigation Network (ICNN), which provides joint data-transfer and tracking between AUV and a human decision maker at Topside.}
 	\label{fig:icnnOverview}
@@
 Given the complexity of the ICNN system, this experiment did not collect an exhaustive set of data across all buoy, source depth, receive layer, and sound speed combinations, as this time was also used to verify and resolve any hardware issues in the modems themselves.
-Importantly, only one modem at a time could run the vehicle behavior to trigger group velocity calculations; other modems that were transmitting simultaneously recorded events as modems, not as vehicles.
+Importantly, only one modem at a time could run the vehicle behavior to trigger group velocity calculations; other modems that were active during the experiment recorded events as tracking buoys, not as vehicles.
 
@@
 
-During ICEX20, the SSP was toggled between the historical``baseline'' and a ``weighted'' estimate based on a CTD, i.e.
+During ICEX20, the SSP was toggled between the historical``baseline'' and a ``weighted'' estimate based on a local CTD cast, i.e.
 an optimal set of weights chosen by a human decision maker via an interactive Tactical Decision Aid framework \citep{bhatt_embedded_2021}.
@@
 When the topside tracking framework receives a modem message, with a time delay, $\Delta t$, from one of the range modems, it will request a new estimate of the group velocity and its associated uncertainty.
-The group velocity estimate is found using the extrapolated navigation for range, $\hat{r}$, and an impulse response estimate received in the form of ray travel times $dt_{j}$ and amplitudes $a_{j}$ for that range and depth.
+The group velocity estimate is found using the vehicle's reported depth and the extrapolated navigation solution for range, $\hat{r}$, as inputs for the ray tracing program.
+The latter returns an impulse response estimate in the form of ray travel times $dt_{j}$ and amplitudes $a_{j}$ for that range and depth.
 
-The initial call to BELLHOP is over a sparse local grid, of which the plane wave approximation is used to bring the ray to the range and depth posited by the onboard tracking solution.
+The initial call to BELLHOP is over a sparse local grid, from which the plane wave approximation is used to bring the ray to the range and depth posited by the onboard tracking solution.
 For each grid point, BELLHOP will produce a number of arrivals resulting from multiple propagation paths for any source-receiver pair.
@@
 Because $\hat{r}$ converges to the correct solution, a comparison of $\hat{r}$ with the GPS-tracked range shows a normal, zero-centered distribution within the bounds of GPS drift.
-The analysis here therefore seeds realistic but ``omniscient'' knowledge of the extrapolated range and leverages the post-processing pipeline to more thoroughly evaluate the acoustic range estimate for all modem events, with three relevant sound speed sources, and both group velocity criterion.
+The present analysis therefore seeds realistic but ``omniscient'' knowledge of the extrapolated range and leverages the post-processing pipeline to more thoroughly evaluate the acoustic range estimate for all modem events, with three relevant sound speed sources, and both group velocity criterion.
 Accordingly, the results in this section evaluate the utility of the algorithms and sound speed sources, divorced from their role in the ICNN while maintaining real-time relevance.
@@
 The sound speed source is indicated by color.
-The minimal and nearest bounce criterion are distinguished by different marker shapes, compared to the separately colored orange dots showing the naive, data-driven group velocity calculation.}
+The minimal and nearest bounce criterion are distinguished by different marker shapes, compared to the separately colored red dots showing the naive, data-driven group velocity calculation.}
 \label{fig:gvel30}
@@
 For a 30 m receiver depth, the NBC shows more overlap with data-derived values as it classifies multipath more correctly.
-For a 90 m receiver depth, the overlap is less accurate due to computational constraints of a limited fan of rays entering the shadow zone rendering a less reliable simulated time arrival packet.
+For a 90 m receiver depth, the overlap is less accurate due to computational constraints of a limited fan of rays entering the shadow zone rendering a less reliable simulated times of arrival packet.
 
@@
 The one case it does not is for the source-receiver pairings between 30 and 90 m in depth.
-The increased error for these reciprocal transmission paths is most likely driven by the computational artifacts propagating through the steep sound speed gradients of the lens and through the shadow zone.
+The increased error for these reciprocal transmission paths is most likely driven by the computational artifacts encountered when propagating through the steep sound speed gradients of the lens and through the shadow zone.
 All other source depth pairings are significantly improved using the chosen weights compared to HYCOM or the baseline.
@@
 
-The large error is driven by the MBC unexpectedly defaulting to a bottom bounce with a much greater OWTT.
+The large error in this instance is driven by the MBC unexpectedly defaulting to a bottom bounce with a much greater OWTT.
 The NBC classifies the multipath as 4 bounces, reducing the range error from greater than a kilometer to less than a meter.
@@
 The latter study showed a precision of 4.3 $\pm$ 4.0 m and an accuracy of 6.0 $\pm$ 4.7 m.
-While significant metric to evaluate GPS robustness, accuracy and precision, not the fix rate, explains the severe banding observed; fix rate bias, where various aspects of a landscape can induce bias by impeding the GPS receiver from acquiring its location, is not relevant given the static nature and small time window of ICEX20.
+While fix rate is a significant metric to evaluate GPS robustness, it is accuracy and precision, and not the fix rate, that explain the severe banding observed; fix rate bias, where various aspects of a landscape can induce bias by impeding the GPS receiver from acquiring its location, is not relevant given the static nature and small time window of ICEX20.
 
@@
 
-These are just two subsets of the physical links that cover all four GPS modem buoys, except for Camp.
+These are just two subsets of the physical links that cover all four GPS modem buoys.
 The GPS at camp is the least accurate due to the human activity and infrastructure occluding the physical puck.
@@
 
-There are many avenues that this approach can be further refined and tested for field operations.
+There are many avenues through which this approach can be further refined and tested for field operations.
 Amongst them is defining the uncertainty grid for BELLHOP via stochastic or data-driven measures such as the distance traveled by the AUV between ICNN updates or the magnitude of the position corrections by the ICNN.

### 2021-08-26 doc/main.tex

@@
 
-The design enables a self-adapting network to transmit and receive at the optimal depth to maintain contact with the AUV \citep{schneider_self-adapting_2020}.
+The design of the ICNN enables a self-adapting network to transmit and receive at the optimal depth to maintain contact with the AUV \citep{schneider_self-adapting_2020}.
 This adaptivity was not used during the modem experiment but was instead toggled manually.
@@
 A key construction in the ICNN are parallel physical and virtual layers; given the small operational scale, on the order of kilometers, and a singular SSP, both layers are projected into a range independent coordinate system for ease of comparison.
+Furthermore, because one of the buoys was configured to act as a virtual vehicle at any given time during the modem experiment, the tracking range was supplemented with an additional modem deployed from the camp site and configured to act as a virtual buoy; the transducer attached to the camp site modem was lowered to a depth of 20 m.
 
@@
 The physical layer describes the acoustic link between beacons in the water column.
-It covers all source depths (20, 30, and 90 m) and both receive layers (30 and 90 m), where the surface expressions of the beacons are tracked by GPS.
+It covers all source depths (20, 30, and 90 m) and both buoy receive layers (30 and 90 m), where the surface expressions of the beacons are tracked by GPS.
 Events are sorted by source depth and sound speed profile, given that these are the dominating factor for the location and span of the shadow zone compared to any temporal or spatial scales of variability.
@@
 The orientation of the triangles\textemdash sideways, upwards, and downwards\textemdash corresponds to depths of 20, 30, and 90 m.
-Subfigures (b-d) correspond to receiver depths of 20, 30, and 90 m, respectively.
+Subfigures (b-d) correspond to transmitter depths of 20, 30, and 90 m, respectively.
 Camp Seadragon, demarcated by the purple triangle, is situated just north of the continental shelf.}
@@
 The suggested algorithm, the nearest bounce criteria (NBC), is a slight modification from the minimal bounce criteria, and includes multipath as a new dimension of information with negligible additional computation.
-This metric was introduced in post-processing and does not include the plane wave approximation.
+This metric was evaluated in post-processing.
 
 Given a running estimate for the horizontal group velocity $u_{i,j}$ between nodes $i$ and $j$, the navigation system has an extrapolated value for range, $\hat{r}$, in the virtual layer, and a recorded travel time, $\Delta t_{i,j}$, in the physical layer.
-Instead of using only the $N_0$ rays with neither surface nor bottom bounces to estimate group velocity, we solve for the power weighted average of the ray travel time for the $N_k$ rays with $k$ bounces,
+Instead of using only the $N_0$ rays with neither surface nor bottom bounces to estimate group velocity and subsequently moving to incremental number of bounces only when no valid direct path solutions exist, we solve for the power weighted average of the ray travel time for the $N_k$ rays with $k$ bounces,
 \begin{equation}
@@
 
-This method selects a different group velocity based on the multipath arrival structure, as the expected arrival is not always the first arrival and could even be masked by noise or blocked temporarily \citep{deffenbaugh_acoustic_1996}.
+This method selects a different group velocity based on the multipath arrival structure, as the detected arrival is not always the first arrival or the direct path and could even be masked by noise or blocked temporarily \citep{deffenbaugh_acoustic_1996}.
 We manually cap the number of bounces to 4 because of the smaller operational scale and the attenuation accrued with many surface interactions.
@@
 The extra size in depth addresses the computational limitations of how time-bounded eigenrays seem to stack vertically in the water column, as shown in Fig \ref{fig:rays3env-zs30}.
-Lastly, to populate the multipath criterion effectively with a multitude of arrivals per number of surface bounces, we abandon the plane-wave extension to the interpolated receiver location.
 
@@
 
-The modem experiment generated 811 beacon to beacon communication events with their own real-time NBC group velocity predictions.
+The modem experiment generated 811 beacon to beacon communication events with their own real-time MBC group velocity predictions.
 As expected, the algorithm is generally overestimating range as it resolves a direct path that does not represent the actual arrival.
@@
 The mean and median for both sound speeds are similar on the order of centimeters; the chosen weights show a maximum 2 meters greater than that of baseline, but it is from a communication event that has no comparison in the baseline data set.
-Analysis shows that where there is overlap in range between sound speed conditions, the range error difference is no more than a few meters.
+Analysis shows that where there is overlap in range between sound speed conditions used for the real-time MBC, the range error difference is no more than a few meters.
 This is likely driven by the prominence of the duct.
@@
 The overarching statistical results show that both sound speed environments, given the smaller operational scales and shallow depths, are accurate enough to approximate the group velocity structure given the local grid.
-Indeed, the error statistics indicated here are similar to that of the ICNN \citep{randeni_high-resolution_2021} and accurate enough to have facilitated a successful vehicle recovery.
+Indeed, the error statistics indicated here are similar to those reported for the ICNN \citep{randeni_high-resolution_2021} and accurate enough to have facilitated a successful vehicle recovery.
 
@@
 
-There are 1242 total beacon to beacon events in the modem experiment that would have been accepted by the ICNN for real-time ranging and navigation had the vehicle behavior been running on all modems.
+There are 1242 total beacon to beacon events in the modem experiment that would have been accepted by the ICNN for real-time ranging and navigation had the ranging behavior been running for all modems in the network, and not only for tracking the vehicle.
 A post-processing analysis that replicated the real-time pipeline was run to overcome the unequal distribution of communication events with respect to depth, range, and sound speed status.
@@
 
-All the events are on the right hand side of the plots\textemdash the MBC has a systemic overestimation, which is reduced via NBC as events straddle the x-axis.
+All the events are on the unshaded region of the plots\textemdash the MBC has a systemic overestimation, which is reduced via NBC as events straddle the x-axis.
 The improvement from MBC to NBC is most evident for the realistic sound speed; while the HYCOM SSP improves from a median absolute range error of 6.41 to 4.61 m, the baseline SSP improves from 10.30 to 2.27 m, and the weighted SSP improves from 13.28 to 2.12 m.
@@
 
-All plots show the same 45 degree banding of source depth and multipath structure, indicating systemic GPS noise.
+All plots in Fig. \ref{fig:compareV1V2} show the same 45 degree banding of source depth and multipath structure, indicating systemic GPS noise.
 The width of those bands indicates the high precision of the acoustic methods relative to lower precision from GPS.
@@
 The literature in underwater acoustic navigation and positioning is either real-time or physics-based.
-In this paper we demonstrate a field-tested approach that, by coupling data streams with fast acoustic modeling, is both real-time and physics-based.
+In this paper we demonstrate a field-tested approach that is both real-time and physics-based; this is achieved by coupling data streams with fast acoustic modeling.
 The methods exploit the upward refracting nature and the total ice cover of the Arctic environment to achieve remarkable ranging accuracy and precision.
@@
 
-Performance in other acoustic environments will require introducing a different thresholded metric to sort time of arrivals.
+Performance in other acoustic environments may require introducing a different thresholded metric to sort time of arrivals.
 Given the NBC algorithm's improvement with increased multipath, its effectiveness is likely only challenged by the valid operational scales of a range independent propagation environment.

### 2021-08-27 doc/main.tex
changes to file names for submission

@@
 	\centering
-	\includegraphics[width=\reprintcolumnwidth]{figs/vehicleRecovery.jpg}
+	\includegraphics[width=\reprintcolumnwidth]{figs/Fig1.jpg}
 	\caption{The successful emergency recovery of AUV Macrura, with the the auger used to drill through the ice sheet in the background. Photo courtesy of LT CDR D. Goodwin.}
@@
 	\centering
-	\includegraphics[width=\reprintcolumnwidth]{figs/ICNN_Overview}
+	\includegraphics[width=\reprintcolumnwidth]{figs/Fig2.pdf}
 	\caption{A schematic overview of the Integrated Communication and Navigation Network (ICNN), which provides joint data-transfer and tracking between AUV and a human decision maker at Topside.}
@@
   \centering
-  \includegraphics[width=.51\textwidth]{figs/modem-chart.pdf} \hfill
-  \includegraphics[width=.48\textwidth]{figs/zs20-birdseye.pdf} \\
+  \includegraphics[width=.51\textwidth]{figs/Fig3a.pdf} \hfill
+  \includegraphics[width=.48\textwidth]{figs/Fig3b.pdf} \\
   \vspace{1em}
-  \includegraphics[width=.48\textwidth]{figs/zs30-birdseye.pdf} \hfill
-  \includegraphics[width=.48\textwidth]{figs/zs90-birdseye.pdf} \\
+  \includegraphics[width=.48\textwidth]{figs/Fig3c.pdf} \hfill
+  \includegraphics[width=.48\textwidth]{figs/Fig3d.pdf} \\
   \caption{An overview of the modem experiment by source and receiver depth and position.
@@
   \centering
-  \includegraphics[width=\reprintcolumnwidth]{figs/raytrace-3env-zs-30.pdf}
+  \includegraphics[width=\reprintcolumnwidth]{figs/Fig4.pdf}
   \caption{Eigenrays for beacon to beacon events for each sound speed with a nominal source depth of 30 m.
@@
 \begin{figure}[h!]
-\includegraphics[width=\columnwidth]{figs/SLIDES-gvel-comparison-wdata-30.pdf}
+\includegraphics[width=\columnwidth]{figs/Fig5.pdf}
 \caption{A comparison of group velocity predictions for all beacon to beacon events in post-processing with a source depth of 30 m, with group velocity on the y-axis and recorded travel time on the x-axis.
@@
 \begin{figure}[!ht]
-\includegraphics[width=\textwidth]{figs/range-error-owtt-newalgorithm.pdf}
+\includegraphics[width=\textwidth]{figs/Fig6.pdf}
 \caption{The post-processed range error for source depths of 20, 30, and 90 m, and receiver depths of 30 and 90 m.
@@
 \begin{figure}[h!]
-\includegraphics[width=\textwidth]{figs/compare-methods-postv1v2.pdf}
+\includegraphics[width=\textwidth]{figs/Fig7.pdf}
 \caption{\label{fig:compareV1V2}{A comparison of range estimation error for all beacon to beacon events in post-processing.
@@
 	\centering
-	\includegraphics[width=\reprintcolumnwidth]{figs/gpsDrift-owttDrift.pdf} 
+	\includegraphics[width=\reprintcolumnwidth]{figs/Fig8.pdf} 
 	\caption{A comparison of GPS drift (y-axis) versus OWTT drift (x-axis), colored by source and receiver depth.
@@
 \begin{figure}[h!]
-\includegraphics[width=\textwidth]{figs/ice-floe-drift.pdf}
+\includegraphics[width=\textwidth]{figs/Fig9.pdf}
 \caption{\label{fig:iceFloeDrift}{The ICNN accounts for ice drift throughout the mission duration.
@@
 \begin{figure}[h!]
-\includegraphics[width=0.8\textwidth]{figs/SLIDES-gvel-comparison-wdata-20.pdf} \\
+\includegraphics[width=0.8\textwidth]{figs/Fig10a.pdf} \\
 \vspace{2em}
-\includegraphics[width=0.8\textwidth]{figs/SLIDES-gvel-comparison-wdata-90.pdf}
+\includegraphics[width=0.8\textwidth]{figs/Fig10b.pdf}
 \caption{A comparison of group velocity predictions for all beacon to beacon events in post-processing.
@@
 % ================================================== %
-
+\FloatBarrier
+\clearpage
 \begin{acknowledgments}

### 2021-12-09 rebuttal/trunk-rebuttal.tex
changes for LINKED LINE NUMBER REFERENCING BETWEEN DOCUMENTS

\documentclass{article}
	\def\papertitle{Under-ice acoustic navigation using real-time model-aided range estimation}
	\def\authors{E. Bhatt, O. V\'{i}quez, H. Schmidt}
	\def\journal{Journal of the Acoustical Society of America}
	\def\doi{}

\input{p-rebuttal-letter.tex}

Dear Editors,

We thank the reviewers and editor for their thoughtful feedback and comments in improving this work. We have significantly restructured the manuscript along several suggestions to separate the real-time navigation methodology, the real-time localization results (Mininum Bounce Criteria), and the post-processing re-localization improvement (Nearest Bounce Criteria). We have also refashioned the abstract and introduction to better elucidate the motivation for this work as a model-aided component for LBL navigation in total under-ice conditions, which could be further generalized for any acoustic environment. The rebuttal letter is organized with sections delineated by the review, shown in the table of contents.

Sincerely,

Dr. EeShan Bhatt \\
\emph{on behalf of all authors}

% table of contents
\renewcommand{\contentsname}{}
\begingroup
\let\pagebreak\relax
\setcounter{tocdepth}{1}
\tableofcontents
\endgroup

%% ------- MACRO FOR CROSS REFERENCING ----- %%
\newcommand{\llabel}[1]{\hypertarget{llineno:#1}{\linelabel{#1}}}
\newcommand{\lref}[1]{\hyperlink{llineno:#1}{\ref*{#1}}}

%% ------- COMMENTS FROM REVIEWER ONE ------ %%
\section{Reviewer One}

	\subsection*{Summary}

	\ebline{This paper demonstrates how a positioning system over ~1.5-3 km of range achieves 11 m of positioning uncertainty.  The work demonstrates how a single group velocity calculated from acoustic rays predicted by BELLHOP can be used to estimate ranges to be used in a LBL network for positioning.  The data presented are from transceivers on fixed ice-tethered buoys but the system could also be implemented on an autonomous underwater vehicle.}

	\lref{v1:test}, \lref{v1:test2}

	This summary captures some of the concepts and results presented in the paper but misses the fact that this system was in fact deployed onboard an underwater vehicle during under-ice deployments. The motivation behind this paper is to improve LBL navigation’s conversion of travel time to range for an under-ice environment that could be generalized to any acoustic environment. To do this, we introduce the Minimal Bounce Criteria method to operationalize real-time acoustic modeling into an effective horizontal group velocity estimate.

	This system was implemented on an autonomous underwater vehicle (AUV) on multiple missions, including an 11 km untethered, under-ice deployment, in March 2020. The AUV was deployed through a hydrohole from Topside and due to a disk error, stalled underneath the ice surface while transmitting its perceived location. The AUV was found within a meter away from its reported location, which serves as strong (but qualitative) evidence of the navigation performance provided by the single group velocity calculation. Because the navigation updates throughout the AUV mission have no GPS ground truth to compare to, this paper evaluates the same system performance across the GPS-navigated LBL beacons.

	The real-time error from the Nearest Bounce Criteria (NBC) achieves a mean absolute positioning error of 11 m. Upon further evaluation of the acoustic events, the Minimal Bounce Criteria (MBC) is run in the same automated pipeline in post-processing (akin to real-time computational constraints and needs) and achieves an error of roughly 3 m.

	Both the NBC and MBC pseudorange estimations are compared to the GPS data on the LBL beacons. The error itself is small and the pattern of error exhibited indicates that the GPS ground truth data has significant noise (drift) that is not replicated by either pseudorange estimation method. For the MBC in particular, the post-processing error is within the precision of the GPS sensor, suggesting that its performance rivals that of GPS.

	\subsection*{Overall Comments}

	\reviewer{}{The claims made in the abstract need to be better supported by the data presented or else tempered a bit. What is demonstrated here is not real-time navigation on a vehicle, but calculation of a handful of short fixed ranges from ice-tethered buoys. The limited data presented here begs more discussion with regards to how applicable the algorithm would be in normal vehicle operation and to navigation of the vehicle, as there were limited ranges between the beacons. What depths does the vehicle typically operate at and what  is the maximum range from the beacons?  The abstract stresses real-time results, but the real-time results are just 3 paragraphs, and the paper mostly concentrates on post-processing results.}

	We agree that there are limitations to how transferable this analysis is, which have now been made clearer in the abstract, introduction, and conclusion.

	\reviewer{}{The hypothesis is stated in the discussion section lines 515 and following: “We hypothesize and validate that the embedded stochastic prediction of a single group velocity is a smoothly varying function of range, source, and receiver depth pairings as well as multipath structure.”  I am not convinced that the few short fixed ranges presented here validate this hypothesis.  Lines 248 and following indicate how the system should track a continuously changing group velocity, but this is not demonstrated with the few fixed ranges presented here.  Also, it would be good to state this hypothesis more clearly earlier in the manuscript.}

	\reviewer{}{The introduction and conclusion sections should be tightened up to clarify the state of the art in various arenas of underwater vehicle navigation and how this work fits in.  The error metrics presented here should be specifically compared with other positioning systems or algorithms over similar spatial scales that do not integrate the ray propagation physics. Various ranging and positioning methods are referenced, but the narrative jumps between different types of AUVs and implementations and between scales of a few kilometers and hundreds of kilometers. This should be discussed in terms of range, vehicle, and propagation conditions.  Line 176 mentions that the SSP and the physics of ray refraction are often overlooked in positioning algorithms; this is, I presume, referring to real-time results or short-range results, but doesn’t specifically say this.}

	\reviewer{}{Related to the point above, the introduction should make clear what type of vehicles the proposed method of navigation is applicable to and over what ranges. The authors do start to do this when they note that a shallow receiver depth enables the estimation method (line 274), but should more directly specify under what conditions the proposed algorithm is applicable.  In the Discussion section, there is an implication that this method of using a single group velocity could be applied to mesoscale operations but should acknowledge that propagation at these larger scales results in arrivals that are spread over several seconds including multiple modal arrivals with varying group velocities.  This method would presumably result in errors much higher than 11m….how might that scale?}

	\reviewer{}{I understand what the authors are referencing when they use the term “group velocity” but I think this term should be used a bit more carefully.  Group velocities are often associated with modal propagation in which each mode has a distinct group velocity.  The goal in this work is to determine a single group velocity by taking a power weighted average of several different ray arrivals to derive range.  This is addressed somewhat in lines 395 and following, but I think it would be appropriate to expand on this discussion and move it up.  Lines 235 and following state “the variation of horizontal group velocity from any source-receiver pair is the fundamental challenge to implement GPS-like navigation in an LBL navigation paradigm…” this statement could use some clarification.  Also, perhaps change the axis on Figure 5 to read something like ‘implied group velocity.”}

	\reviewer{}{As written, this manuscript seems more suited to an ocean engineering journal.  For a JASA audience, this should be more focused on the acoustics.  Perhaps I missed it, but I didn’t see that the manuscript even mentioned the frequencies or waveforms that were used in the experiment.  I’d suggest, for example, less emphasis on which options were toggled on/off or triggered (e.g. line 186, 196, 241, 254 ) and more on, for example, why specific depths were chosen and how they provide better coverage due to the acoustic shadow zone (e.g., Line 155), or how the specific rays were eliminated by the filtering (Line 232). Lines 370-372 state “The increased error for these reciprocal transmission paths is most likely driven by the computational artifacts encountered when propagating through the steep sound speed gradients of the lens and through the shadow zone.“ The authors might consider providing more information on these artifacts.}

	\reviewer{}{There seems to be a somewhat outsized discussion of GPS drift, but I didn’t see how the GPS uncertainty was directly applied to the uncertainty values presented in the paper.  I’d suggest condensing this section.}

	\subsection*{Specific Comments}

	\reviewer{Line 226}
	{In the section describing post-processing analysis, the authors bring up that the processing was designed to mirror information available on a submarine – modeled data, historical data, and in situ data.  This is a nice rationale for the analysis, and I’d suggest highlighting this in the introduction.}

	\reviewer{}{How long was the experiment from start to finish, and how much did the sound speed profile change over the duration of the experiment?}

	\reviewer{}{What is the magnitude of the uncertainty in the position of the transceivers relative to the GPS positions measured at the surface? Presumably the transceivers particularly at 90 m depth could move from the lat/lon position of the GPS receivers at the surface.}

	\reviewer{Lines 40-42}{Note ranging via one way travel times that have been used prior to the advent of the micro-modem, for example RAFOS.}

	\reviewer{Line 192}{Clarify what you mean when you state that the transmit and receive layers are independent? Also, the discussions of physical and virtual layers are confusing. I think everything is there, but reorganization would make this more clear. Consider moving up the paragraph starting at line 203.}

	\reviewer{Line 205}{What do you mean that the coordinate system is range independent?}

	\reviewer{Line 222}{Could you provide more information on the baseline and weighted SSPs.  Where did the baseline profile come from?  }

	\reviewer{Line 256}{How is the group velocity uncertainty calculated?}

	\reviewer{Line 231}{Quantify the travel-time proximity to recorded data that was used to filter the eigenrays.}

	\reviewer{Lines 260, 303-305}{Please provide more information on the BELLHOP implementation.   Line 260 indicates a sparse grid.  From lines 303-305 do I understand that the grid is 1m in range and 2 m in depth?}

	{\color{blue}different definitions of sparse. we meant that we're not computing other locations. they think sparse means the grid points are far apart.}

	\reviewer{Line 309}{Suggest re-writing this sentence for clarity: “As expected, the algorithm is generally overestimating range as it resolves a direct path that does not represent the actual arrival.”}

	\reviewer{Line 364}{Could use a figure reference.}

	\reviewer{Line 562}{I don’t agree that many approaches to underwater navigation combine it with acoustic tomography.}
	{\color{blue}Benefit of concurrent navigation and tomography has been discussed, which is not necessarily that many approaches combine it.}

	\reviewer{Figure 1}{Figure 1 is a nice picture showing the vehicle, which motivates the work presented here, but was not used for collection of any of the data presented.  As it’s not pertinent to the data presented here, I’d suggest removing this figure or at least combining this figure with Figure 2.}

	\reviewer{Figure 3}{Figure 3 tallies the number of successful events, and I think the use of the directional arrows communicates the data well.  How many events are successful in comparison with those that are considered unsuccessful and what generally causes an event to be deemed unsuccessful?}

	\reviewer{Figure 5}{What is the point labeled u=751 m/s that appears to fall around 1424 m/s?}

	\reviewer{Table II}{Might this information be portrayed better by a box plot?}

	\subsection*{Minor Edits}

	\reviewer{Fig. 2 and 3}{Fig 2 refers to central node position as Topside and Fig. 3 as Camp Seadragon. Suggest standardizing.}

	\reviewer{Lines 311, 359}{'Data' should be plural.}

	\reviewer{Figure 6}{Suggest including receiver depth along with source depth in the title as they are both important in distinguishing between the cases.}

	\reviewer{Table I}{Table I indicates direction errors are positive.  Please note the sign convention in the caption.}

	\reviewer{Table II}{‘range estimation error metrics’ rather than ‘range estimation metrics’?}

	\reviewer{Table II and III}{Suggest removing ‘algorithm’ label.  Also substitute ‘MBC’ and ‘NBC’ for minimal and nearest to match the text.}

	\reviewer{}{There are several figures and tables which are not referenced properly (Line 305, 391, 411)}

%% ------- COMMENTS FROM REVIEWER TWO ------ %%

\section{Reviewer Two}

	\reviewer{Line 7}{'Beaufort Lens' This is not yet a scientific term, suggest finding descriptive terms to use. What comprises the 'lens'? The Pacific Summer Water is the warm layer.}

	\reviewer{Line 10}{'GPS aided', do  you mean GPS-navigated?}
	Yes.

	\reviewer{Line 12}{'rival GPS' is a fairly bold statement when GPS is used as ground truth.}
	{\color{blue} Accuracy comparable to conventional GPS; does not introduce decrease precision than is already documented by GPS puck precision; Effectively extends single GPS into the water column with comparable performance. Given more beacons, there is potential to reduce uncertainty in the same way that GPS / differential GPS works}

	\reviewer{Line 89}{Is the 'lens' the warm water? Or a combination of layers? What does 'lens' mean acoustically? Normally in ocean acoustics people refer to ducts and layers.}

	\reviewer{Line 94}{Ice Tethered profiler data shows fairly slow variability in space and time in the Beaufort, but changes in characteristics occur as the deep basin transitions to the Chukchi, so what is meant by 'high degree' exactly?}

	\reviewer{Fig 4 and Line 216}{Fig 4 and 216 - where did the baseline come from (exactly)? What database, etc.}

	\reviewer{Line 224}{why did a human have to chose an optimal set of weights? Why not just use the measured SSP, or an average of a few?}

	\reviewer{Line 231}{'abundance'. Is there a more precise way to say this? Bellhop does what it is programmed to…}

	\reviewer{Line 241}{'modem triggering time', do you mean 'detected arrival time'?}

	\reviewer{Line 249}{Use of the word 'range' for the 'navy range' type of use is intermixed with range as in distance. Suggest not using 'range' and in 'navy range' and saying 'nav' or 'comms' system or network.}

	\reviewer{Line 261}{'bring'. Do you mean interpolate?}

	\reviewer{Line 281}{Awkward sentence, either unclear or extremely obvious. "...an acoustic arrival does not always take the direct path from source to the receiver,".}

	\reviewer{}{for both criteria: a graphic that shows the method could be very useful.}

	\reviewer{Line 319}{"This is likely driven by the prominence of the duct." Do you mean: consistency of sound speed within the duct where the rays were trapped and thus traveled? Or something else?}

	\reviewer{Line 410}{'anomalous' - does this section need to be included? Or just the issue summarized and noted? Delete the table?}
	{\color{blue}Delete table, summarize issue.}

	\reviewer{Lines 585-591}{Suggest deleting, not very appropriate for an academic paper.}

	\reviewer{Fig 10}{Would be good to have some conclusions in addition to the description.}

%% ------- COMMENTS FROM EDITOR ------ %%
\section{Comments from the Editor}

	\reviewer{}
	{Both reviewers recommend that the manuscript be reconsidered after revision based on the reviews, and I concur. The key point of the manuscript seems to be to compare two different approaches to determining an effective sound speed for converting travel times to ranges for use in determining positions by trilateration in an Arctic environment (designated Minimal Bounce Criteria and Nearest Bounce Criteria). The goal is not evident in the abstract and is first more-or-less (but not entirely) clearly stated in Section II (Methods) (lines 174-176). While some discussion of the AUV operations is appropriate to motivate the analysis, the extensive discussion in the Introduction and elsewhere in the manuscript seems unnecessary and inappropriate. The use of jargon from the AUV application (e.g., physical layer, virtual layer) also seems unnecessary.}

	\reviewer{}
	{I agree with Reviewer 1 that the manuscript needs much more information on the acoustic parameters (frequencies, pulse lengths/modulation, ranges, expected travel-time measurement accuracy, etc.). In addition, the ray geometries are quite different for cases in which the source and receiver are both in the surface duct (i.e., 20 and 30 m), both in the Beaufort Duct (for the sound-speed profiles for which it is present) (i.e., 90 m) and one of the source or receiver is in the surface duct and the other is in the Beaufort Duct. The manuscript needs to include some discussion of these different cases.}

	\reviewer{}
	{The authors need to carefully review the information in the Information for Contributors available on the JASA website. In particular:
	\begin{itemize}
	\item Section IV.F. Abstract page: “Personal pronouns and explicit claims as to novelty should be assiduously avoided.”
	\item Section X.A. Introductory section: “Although some discussion of the background of the work may be advisable, a statement of the precise subject of the work must appear within the first two paragraphs.”
	\end{itemize}}

	\reviewer{}
	{As noted by Reviewer 1, the term “group velocity” has a precise definition. It seems that what the authors mean here is an effective sound speed used to convert travel times to ranges.}

	\reviewer{}
	{Why is the effective sound-speed determined from the GPS-determined range and the acoustic travel time referred to as “… the naïve group velocity …”? What is naïve about it?}

	\reviewer{}
	{Figure 3 is only referenced in the Fig. 4 caption.}

	\reviewer{}
	{I agree with both reviewers that the Section IV.E, GPS sensor drift at polar latitudes, is out of place. I find the term “GPS drift” itself odd. It seems that the authors really mean GPS accuracy.}
	{\color{blue}GPS precision}

	\reviewer{}
	{The statement (or similar statements) made in various places in the manuscript that “… we present a working solution for underwater navigation that achieves GPS-like accuracy and precision …” (lines 120-121) is not a quantitative or meaningful statement. GPS (really GNSS) applications come in different flavors, including stand-alone single frequency, stand-alone dual frequency, differential, real-time kinematic, etc., with varying accuracies.}
	{\color{blue}achieves ``x meter'' level accuracy compared to ``y meter'' level accuracy reported by the manufacturer}

	\reviewer{}
	{Unsupported generalizations about previous work, e.g., “… conversion of travel time into range … is an often overlooked element of any underwater navigation system …” (lines 174-176) are to be avoided. In this case, the generalization is in fact incorrect.}

	\reviewer{}
	{Several of the references are incomplete, i.e., Norgren et al. (2014), Poulsen et al. (2016), Schmidt and Schneider (2016), and Schneider et al. (2020).}

%% ------- REVIEWER QUESTIONS ------- %%

\section{Reviewer's Responses to Questions}

	\reviewer{Is the manuscript of good scientific quality, free from errors, misconceptions or ambiguities; does it present original work; and does it contain sufficient new results, new applications or new developments of reasonable enough significance to warrant its publication in JASA? Please indicate in your report (in detailed comments, below) any points which are objectionable or which need attention.}
	{Reviewer 1: It presents original work.  There are aspects which need clarification and are outlined in the attached report.

	Reviewer 2: The work is interesting and describes a data set collected from the Arctic, which is of high value. However, some of the writing is a bit awkward and occasionally vague or imprecise.}

	\reviewer{Is JASA an appropriate journal in which to present this work? In this regard, please consider carefully the commitment of JASA to publish work that is within the scope of Acoustics. Does the content of the manuscript, including terminology and the references cited, meet this criterion?}
	{Reviewer 1: I wouldn’t say JASA is the most appropriate journal.  As written, this work might be better suited to an ocean engineering journal. The results do, however, rely on acoustic propagation models which does make it fall within the scope of JASA, and it is particularly applicable to the Arctic Acoustics special issue.  I've made some suggestions in the attached (e.g., I didn't see that the frequency of the acoustic transmissions was  mentioned.  That is something that should be mentioned for an acoustics audience).

	Reviewer 2: Yes.}

	\reviewer{Is the manuscript a clear, concise, reasonably self-contained presentation of the material, giving adequate references to related work? Is English satisfactory? Please indicate needed changes in your report.}
	{Reviewer 1: The English is satisfactory.  This paper could benefit from some reorganization and careful editing. Several references to figures and tables were missing or incomplete. The hypothesis was presented in the discussion, but could be improved if this was more clearly stated at the outset.  Please see attached report for specific comments and suggestions.

	Reviewer 2: Overall the language is fine, details below.}

	\reviewer{Are the tables and figures clear and relevant, and are the captions adequate? Are there either too many or too few? If any of the figures are in color, is the color essential for conveying the scientific point?}
	{Reviewer 1: I would suggest combining Figures 1 and 2 or eliminating Figure 1.  Some suggestions for improving figure and table titles and captions are in the attached report.  The use of color does help convey the scientific point.

	Reviewer 2: In general they are fine except as noted.}

	\reviewer{If Supplementary material was submitted, is it relevant to the manuscript and should it be deposited in the Supplemental Depository for reference to the manuscript?}
	{Reviewer 1: The supplemental figures were relevant to the manuscript.

	Reviewer 2: NA}

	\reviewer{Does the paper make effective use of journal space, or are parts unnecessary, unimportant, or subject to condensation? If so, which?
	}
	{Reviewer 1: I'd suggest condensing the section on GPS drift (Section III E).

	Reviewer 2: The GPS section seems out of place. Can it be shortened to a few sentences and references?}

	\reviewer{
	Is the title appropriate and the abstract adequate for verbatim reproduction in abstract journals? IMPORTANT: The lead paragraph should advertise the main points of the article and must describe in terms accessible to the general reader the context and significance or the research problem studied and the importance of the results.}
	{Reviewer 1: The title and abstract are a bit misleading as there is no navigation demonstrated in the paper and the ranging results were from fixed buoys rather than a vehicle.  The abstract, and the final sentence in particular, should be tempered to reflect the scope of the results.  Also, the abstract specifies absolute error of 11m, but should note the range over which it’s measured.  The lead paragraph does not advertise the main points of the article.

	Reviewer 2: Title, yes.
	The abstract does need some work. Statements like this are risky, suggest summarizing technical contributions.
	\begin{quote}
	To our knowledge, these results are the first field experiment to demonstrate a real-time, physics-based data processing to minimize ranging error in underwater navigation.
	\end{quote}}

\end{document}

### 2021-12-12 pipeline/overview-info/main_plot_experiment_overview.m
organizing into pipelines

@@ -126,7 +126,7 @@ end
 %legend(Lgd,LgdStr,'location','BestOutside');
 legend(Lgd,LgdStr,'Position',[0.9135 0.7945 0.0632 0.1292]);
 
-h_printThesisPNG('tx-rx-depth');
+% h_printThesisPNG('tx-rx-depth');
 
 %% figure 2 : tx rx chart
 
@@ -252,6 +252,4 @@ end
     
 legend(Lgd,LgdStr,'location','WestOutside','fontsize',14);
 
-h_printThesisPNG('modem-chart');
-
-%% figure 3
+% h_printThesisPNG('modem-chart');

### 2021-12-12 src/gvel_1_data.m

@@ -1,288 +1,0 @@
-%% gvel_1_data
-% makes plots to explore *data* from ICEX20
-
-%% prep workspace
-clear; clc; close all;
-
-% load data
-A = readtable('../pipeline/bellhop-gvel-gridded/gveltable.csv');
-
-% remove crazy 11 second event, event that is nominally 1.58* seconds
-indBad1 = find(A.owtt > 4);
-indBad2 = find(strcmp(A.rxNode,'East') & A.owtt > 1.55);
-indBad = union(indBad1,indBad2);
-
-% 1.587 events, had clock errors + Bellhop can't resolve these
-A.simGvel(indBad) = NaN;
-
-% load modem marker information
-load p_modemMarkerDetails
-
-plotbool = [1 1 1 1 1 1];
-% plotbool = [0 0 0 0 0 1];
-
-%% figure --- owtt vs range
-if plotbool(1) == 1
-    
-    figure('name','owtt-range','renderer','painters','position',[108 108 1400 600]);
-    
-    % all events
-    subplot(2,1,1);
-    hold on
-    for k = 1:numel(A.simGvel)
-        
-        % plot
-        yval = A.owtt(k);
-        xval = A.recRange(k);
-        scatter(xval,yval,...
-            150,markerModemMap(A.rxNode{k}),markerShape(A.recDepth(k)),...
-            'filled','MarkerFaceAlpha',0.2,'handlevisibility','off');
-    end
-    hold off
-    grid on
-    xlabel('range [m]');
-    ylabel('travel time [s]');
-    title(sprintf('All valid events, n=%u',numel(A.simGvel)));
-    
-    
-    % only simGvel
-    indValid = ~isnan(A.simGvel);
-    subplot(2,1,2);
-    hold on
-    for k = 1:numel(indValid)
-        
-        % plot if simulation captured gvel information
-        if indValid(k)==1
-            yval = A.owtt(k);
-            xval = A.recRange(k);
-            scatter(xval,yval,...
-                150,markerModemMap(A.rxNode{k}),markerShape(A.recDepth(k)),...
-                'filled','MarkerFaceAlpha',0.2,'handlevisibility','off');
-        end
-    end
-    hold off
-    grid on
-    xlabel('range [m]');
-    ylabel('travel time [s]');
-    title(sprintf('Events with in situ group velocity estimate, n=%u',sum(indValid)));
-    
-end
-
-
-%% figure --- gvel compared to owtt & range
-
[…]

### 2021-12-12 rebuttal/rebuttal-letter/trunk-rebuttal.tex

--- 2021-12-09 rebuttal/trunk-rebuttal.tex
@@
 
-\input{p-rebuttal-letter.tex}
+\input{../p-rebuttal-letter.tex}
 
@@
 
+%\ref*{v1:test}
+% \pageref*{v1:test}
+
+% \formatLineTracking{test}{v1:test}
+
 % table of contents
@@
 \endgroup
-
-%% ------- MACRO FOR CROSS REFERENCING ----- %%
-\newcommand{\llabel}[1]{\hypertarget{llineno:#1}{\linelabel{#1}}}
-\newcommand{\lref}[1]{\hyperlink{llineno:#1}{\ref*{#1}}}
 
@@
 The data presented are from transceivers on fixed ice-tethered buoys but the system could also be implemented on an autonomous underwater vehicle.}
-
-	\lref{v1:test}, \lref{v1:test2}
 
@@
 
-	\reviewer{Line 226}
+	\lreviewer{Line 226}{}
 	{In the section describing post-processing analysis, the authors bring up that the processing was designed to mirror information available on a submarine – modeled data, historical data, and in situ data.
@@
 
-	\reviewer{}{How long was the experiment from start to finish, and how much did the sound speed profile change over the duration of the experiment?}
+	\lreviewer{}{}{How long was the experiment from start to finish, and how much did the sound speed profile change over the duration of the experiment?}
 
-	\reviewer{}{What is the magnitude of the uncertainty in the position of the transceivers relative to the GPS positions measured at the surface?
+	\lreviewer{}{}{What is the magnitude of the uncertainty in the position of the transceivers relative to the GPS positions measured at the surface?
 Presumably the transceivers particularly at 90 m depth could move from the lat/lon position of the GPS receivers at the surface.}
 
-	\reviewer{Lines 40-42}{Note ranging via one way travel times that have been used prior to the advent of the micro-modem, for example RAFOS.}
+	\lreviewer{Lines 40-42}{}{Note ranging via one way travel times that have been used prior to the advent of the micro-modem, for example RAFOS.}
 
-	\reviewer{Line 192}{Clarify what you mean when you state that the transmit and receive layers are independent?
+	\lreviewer{Line 192}{}{Clarify what you mean when you state that the transmit and receive layers are independent?
 Also, the discussions of physical and virtual layers are confusing.
@@
 
-	\reviewer{Line 205}{What do you mean that the coordinate system is range independent?}
+	\lreviewer{Line 205}{}{What do you mean that the coordinate system is range independent?}
 
-	\reviewer{Line 222}{Could you provide more information on the baseline and weighted SSPs.  Where did the baseline profile come from?  }
+	\lreviewer{Line 222}{}{Could you provide more information on the baseline and weighted SSPs.  Where did the baseline profile come from?  }
 
-	\reviewer{Line 256}{How is the group velocity uncertainty calculated?}
+	\lreviewer{Line 256}{}{How is the group velocity uncertainty calculated?}
 
-	\reviewer{Line 231}{Quantify the travel-time proximity to recorded data that was used to filter the eigenrays.}
+	\lreviewer{Line 231}{}{Quantify the travel-time proximity to recorded data that was used to filter the eigenrays.}
 
-	\reviewer{Lines 260, 303-305}{Please provide more information on the BELLHOP implementation.
+	\lreviewer{Lines 260, 303-305}{}{Please provide more information on the BELLHOP implementation.
 Line 260 indicates a sparse grid.
@@
 
-	\reviewer{Line 309}{Suggest re-writing this sentence for clarity: “As expected, the algorithm is generally overestimating range as it resolves a direct path that does not represent the actual arrival.”}
+	\lreviewer{Line 309}{}{Suggest re-writing this sentence for clarity: “As expected, the algorithm is generally overestimating range as it resolves a direct path that does not represent the actual arrival.”}
 
-	\reviewer{Line 364}{Could use a figure reference.}
+	\lreviewer{Line 364}{}{Could use a figure reference.}
 
-	\reviewer{Line 562}{I don’t agree that many approaches to underwater navigation combine it with acoustic tomography.}
+	\lreviewer{Line 562}{}{I don’t agree that many approaches to underwater navigation combine it with acoustic tomography.}
 	{\color{blue}Benefit of concurrent navigation and tomography has been discussed, which is not necessarily that many approaches combine it.}
 
-	\reviewer{Figure 1}{Figure 1 is a nice picture showing the vehicle, which motivates the work presented here, but was not used for collection of any of the data presented.
+	\lreviewer{Figure 1}{}{Figure 1 is a nice picture showing the vehicle, which motivates the work presented here, but was not used for collection of any of the data presented.
 As it’s not pertinent to the data presented here, I’d suggest removing this figure or at least combining this figure with Figure 2.}
 
-	\reviewer{Figure 3}{Figure 3 tallies the number of successful events, and I think the use of the directional arrows communicates the data well.
+	\lreviewer{Figure 3}{}{Figure 3 tallies the number of successful events, and I think the use of the directional arrows communicates the data well.
 How many events are successful in comparison with those that are considered unsuccessful and what generally causes an event to be deemed unsuccessful?}
 
-	\reviewer{Figure 5}{What is the point labeled u=751 m/s that appears to fall around 1424 m/s?}
+	\lreviewer{Figure 5}{}{What is the point labeled u=751 m/s that appears to fall around 1424 m/s?}
 
-	\reviewer{Table II}{Might this information be portrayed better by a box plot?}
+	\lreviewer{Table II}{}{Might this information be portrayed better by a box plot?}
 
@@
 
-	\reviewer{Fig. 2 and 3}{Fig 2 refers to central node position as Topside and Fig. 3 as Camp Seadragon. Suggest standardizing.}
+	\lreviewer{Fig. 2 and 3}{}{Fig 2 refers to central node position as Topside and Fig. 3 as Camp Seadragon. Suggest standardizing.}
 
-	\reviewer{Lines 311, 359}{'Data' should be plural.}
+	\lreviewer{Lines 311, 359}{}{'Data' should be plural.}
 
-	\reviewer{Figure 6}{Suggest including receiver depth along with source depth in the title as they are both important in distinguishing between the cases.}
+	\lreviewer{Figure 6}{}{Suggest including receiver depth along with source depth in the title as they are both important in distinguishing between the cases.}
 
-	\reviewer{Table I}{Table I indicates direction errors are positive.  Please note the sign convention in the caption.}
+	\lreviewer{Table I}{}{Table I indicates direction errors are positive.  Please note the sign convention in the caption.}
 
-	\reviewer{Table II}{‘range estimation error metrics’ rather than ‘range estimation metrics’?}
+	\lreviewer{Table II}{}{‘range estimation error metrics’ rather than ‘range estimation metrics’?}
 
-	\reviewer{Table II and III}{Suggest removing ‘algorithm’ label.  Also substitute ‘MBC’ and ‘NBC’ for minimal and nearest to match the text.}
+	\lreviewer{Table II and III}{}{Suggest removing ‘algorithm’ label.  Also substitute ‘MBC’ and ‘NBC’ for minimal and nearest to match the text.}
 
-	\reviewer{}{There are several figures and tables which are not referenced properly (Line 305, 391, 411)}
+	\lreviewer{}{}{There are several figures and tables which are not referenced properly (Line 305, 391, 411)}
 
@@
 
-	\reviewer{Line 7}{'Beaufort Lens' This is not yet a scientific term, suggest finding descriptive terms to use. What comprises the 'lens'? The Pacific Summer Water is the warm layer.}
+	\lreviewer{Line 7}{}{'Beaufort Lens' This is not yet a scientific term, suggest finding descriptive terms to use. What comprises the 'lens'? The Pacific Summer Water is the warm layer.}
 
-	\reviewer{Line 10}{'GPS aided', do  you mean GPS-navigated?}
+	\lreviewer{Line 10}{}{'GPS aided', do  you mean GPS-navigated?}
 	Yes.
 
-	\reviewer{Line 12}{'rival GPS' is a fairly bold statement when GPS is used as ground truth.}
+	\lreviewer{Line 12}{}{'rival GPS' is a fairly bold statement when GPS is used as ground truth.}
 	{\color{blue} Accuracy comparable to conventional GPS; does not introduce decrease precision than is already documented by GPS puck precision; Effectively extends single GPS into the water column with comparable performance.
@@
 
-	\reviewer{Line 89}{Is the 'lens' the warm water? Or a combination of layers? What does 'lens' mean acoustically? Normally in ocean acoustics people refer to ducts and layers.}
+	\lreviewer{Line 89}{}{Is the 'lens' the warm water? Or a combination of layers? What does 'lens' mean acoustically? Normally in ocean acoustics people refer to ducts and layers.}
 
-	\reviewer{Line 94}{Ice Tethered profiler data shows fairly slow variability in space and time in the Beaufort, but changes in characteristics occur as the deep basin transitions to the Chukchi, so what is meant by 'high degree' exactly?}
+	\lreviewer{Line 94}{}{Ice Tethered profiler data shows fairly slow variability in space and time in the Beaufort, but changes in characteristics occur as the deep basin transitions to the Chukchi, so what is meant by 'high degree' exactly?}
 
-	\reviewer{Fig 4 and Line 216}{Fig 4 and 216 - where did the baseline come from (exactly)? What database, etc.}
+	\lreviewer{Fig 4 and Line 216}{}{Fig 4 and 216 - where did the baseline come from (exactly)? What database, etc.}
 
-	\reviewer{Line 224}{why did a human have to chose an optimal set of weights? Why not just use the measured SSP, or an average of a few?}
+	\lreviewer{Line 224}{}{why did a human have to chose an optimal set of weights? Why not just use the measured SSP, or an average of a few?}
 
-	\reviewer{Line 231}{'abundance'. Is there a more precise way to say this? Bellhop does what it is programmed to…}
+	\lreviewer{Line 231}{}{'abundance'. Is there a more precise way to say this? Bellhop does what it is programmed to…}
 
-	\reviewer{Line 241}{'modem triggering time', do you mean 'detected arrival time'?}
+	\lreviewer{Line 241}{}{'modem triggering time', do you mean 'detected arrival time'?}
 
-	\reviewer{Line 249}{Use of the word 'range' for the 'navy range' type of use is intermixed with range as in distance.
+	\lreviewer{Line 249}{}{Use of the word 'range' for the 'navy range' type of use is intermixed with range as in distance.
 Suggest not using 'range' and in 'navy range' and saying 'nav' or 'comms' system or network.}
 
-	\reviewer{Line 261}{'bring'. Do you mean interpolate?}
+	\lreviewer{Line 261}{}{'bring'. Do you mean interpolate?}
 
-	\reviewer{Line 281}{Awkward sentence, either unclear or extremely obvious. "...an acoustic arrival does not always take the direct path from source to the receiver,".}
+	\lreviewer{Line 281}{}{Awkward sentence, either unclear or extremely obvious. "...an acoustic arrival does not always take the direct path from source to the receiver,".}
 
-	\reviewer{}{for both criteria: a graphic that shows the method could be very useful.}
+	\lreviewer{}{}{for both criteria: a graphic that shows the method could be very useful.}
 
-	\reviewer{Line 319}{"This is likely driven by the prominence of the duct." Do you mean: consistency of sound speed within the duct where the rays were trapped and thus traveled? Or something else?}
+	\lreviewer{Line 319}{}{"This is likely driven by the prominence of the duct." Do you mean: consistency of sound speed within the duct where the rays were trapped and thus traveled?
+Or something else?}
 
-	\reviewer{Line 410}{'anomalous' - does this section need to be included? Or just the issue summarized and noted? Delete the table?}
+	\lreviewer{Line 410}{}{'anomalous' - does this section need to be included? Or just the issue summarized and noted? Delete the table?}
 	{\color{blue}Delete table, summarize issue.}
 
-	\reviewer{Lines 585-591}{Suggest deleting, not very appropriate for an academic paper.}
+	\lreviewer{Lines 585-591}{}{Suggest deleting, not very appropriate for an academic paper.}
 
-	\reviewer{Fig 10}{Would be good to have some conclusions in addition to the description.}
+	\lreviewer{Fig 10}{}{Would be good to have some conclusions in addition to the description.}
 
@@
 
-	\reviewer{Is the manuscript of good scientific quality, free from errors, misconceptions or ambiguities; does it present original work; and does it contain sufficient new results, new applications or new developments of reasonable enough significance to warrant its publication in JASA?
+	\reviewer{}{Is the manuscript of good scientific quality, free from errors, misconceptions or ambiguities; does it present original work; and does it contain sufficient new results, new applications or new developments of reasonable enough significance to warrant its publication in JASA?
 Please indicate in your report (in detailed comments, below) any points which are objectionable or which need attention.}
@@
 
-	\reviewer{Is JASA an appropriate journal in which to present this work?
+	\reviewer{}{Is JASA an appropriate journal in which to present this work?
 In this regard, please consider carefully the commitment of JASA to publish work that is within the scope of Acoustics.
@@
 
-	\reviewer{Is the manuscript a clear, concise, reasonably self-contained presentation of the material, giving adequate references to related work?
+	\reviewer{}{Is the manuscript a clear, concise, reasonably self-contained presentation of the material, giving adequate references to related work?
 Is English satisfactory?
@@
 
-	\reviewer{Are the tables and figures clear and relevant, and are the captions adequate?
+	\reviewer{}{Are the tables and figures clear and relevant, and are the captions adequate?
 Are there either too many or too few?
@@
 
-	\reviewer{If Supplementary material was submitted, is it relevant to the manuscript and should it be deposited in the Supplemental Depository for reference to the manuscript?}
+	\reviewer{}{If Supplementary material was submitted, is it relevant to the manuscript and should it be deposited in the Supplemental Depository for reference to the manuscript?}
 	{Reviewer 1: The supplemental figures were relevant to the manuscript.
@@
 
-	\reviewer{Does the paper make effective use of journal space, or are parts unnecessary, unimportant, or subject to condensation? If so, which?
+	\reviewer{}{Does the paper make effective use of journal space, or are parts unnecessary, unimportant, or subject to condensation? If so, which?
 	}
@@
 
-	\reviewer{
+	\reviewer{}{
 	Is the title appropriate and the abstract adequate for verbatim reproduction in abstract journals?

### 2021-12-13 rebuttal/rebuttal-letter/trunk-rebuttal.tex
tagging from OG line numbers in manuscript document

@@
 We have significantly restructured the manuscript along several suggestions to separate the real-time navigation methodology, the real-time localization results (Mininum Bounce Criteria), and the post-processing re-localization improvement (Nearest Bounce Criteria).
+We have also introduced that language (navigation, localization, and re-localization) more explicitly and consistently to avoid confusion and better contextualize our contribution.
+
 We have also refashioned the abstract and introduction to better elucidate the motivation for this work as a model-aided component for LBL navigation in total under-ice conditions, which could be further generalized for any acoustic environment.
@@
 
-%\ref*{v1:test}
+% \ref*{v1:test}
 % \pageref*{v1:test}
@@
 
-	This summary captures some of the concepts and results presented in the paper but misses the fact that this system was in fact deployed onboard an underwater vehicle during under-ice deployments.
-The motivation behind this paper is to improve LBL navigation’s conversion of travel time to range for an under-ice environment that could be generalized to any acoustic environment.
+	This summary key concepts and results presented in the paper.
+However, this system was in fact deployed onboard an underwater vehicle during under-ice deployments.
+The motivation behind this work is to improve the standard LBL navigation’s conversion of travel time to range for an under-ice environment in a way that could be generalized to any acoustic waveguide.
 To do this, we introduce the Minimal Bounce Criteria method to operationalize real-time acoustic modeling into an effective horizontal group velocity estimate.
@@
 
-	\lreviewer{Line 226}{}
+	\lreviewer{line 226}{v1:1.8}
 	{In the section describing post-processing analysis, the authors bring up that the processing was designed to mirror information available on a submarine – modeled data, historical data, and in situ data.
@@
 
-	\lreviewer{Lines 40-42}{}{Note ranging via one way travel times that have been used prior to the advent of the micro-modem, for example RAFOS.}
+	\lreviewer{lines 40-42}{v1:1.11}{Note ranging via one way travel times that have been used prior to the advent of the micro-modem, for example RAFOS.}
 
-	\lreviewer{Line 192}{}{Clarify what you mean when you state that the transmit and receive layers are independent?
+	\lreviewer{line 192}{v1:1.12}{Clarify what you mean when you state that the transmit and receive layers are independent?
 Also, the discussions of physical and virtual layers are confusing.
@@
 
-	\lreviewer{Line 205}{}{What do you mean that the coordinate system is range independent?}
+	\lreviewer{line 205}{v1:1.13}{What do you mean that the coordinate system is range independent?}
 
-	\lreviewer{Line 222}{}{Could you provide more information on the baseline and weighted SSPs.  Where did the baseline profile come from?  }
+	\lreviewer{line 222}{v1:1.14}{Could you provide more information on the baseline and weighted SSPs. Where did the baseline profile come from?}
+	\label{1.14}
 
-	\lreviewer{Line 256}{}{How is the group velocity uncertainty calculated?}
+	\lreviewer{line 256}{v1:1.15}{How is the group velocity uncertainty calculated?}
 
-	\lreviewer{Line 231}{}{Quantify the travel-time proximity to recorded data that was used to filter the eigenrays.}
+	\lreviewer{line 231}{v1:1.16}{Quantify the travel-time proximity to recorded data that was used to filter the eigenrays.}
 
-	\lreviewer{Lines 260, 303-305}{}{Please provide more information on the BELLHOP implementation.
+	\lreviewer{lines 260, 303-305}{v1:1.17}{Please provide more information on the BELLHOP implementation.
 Line 260 indicates a sparse grid.
@@
 
-	\lreviewer{Line 309}{}{Suggest re-writing this sentence for clarity: “As expected, the algorithm is generally overestimating range as it resolves a direct path that does not represent the actual arrival.”}
+	\lreviewer{line 309}{v1:1.18}{Suggest re-writing this sentence for clarity: “As expected, the algorithm is generally overestimating range as it resolves a direct path that does not represent the actual arrival.”}
 
-	\lreviewer{Line 364}{}{Could use a figure reference.}
+	\lreviewer{line 364}{v1:1.19}{Could use a figure reference.}
 
-	\lreviewer{Line 562}{}{I don’t agree that many approaches to underwater navigation combine it with acoustic tomography.}
+	\lreviewer{line 562}{v1:1.20}{I don’t agree that many approaches to underwater navigation combine it with acoustic tomography.}
 	{\color{blue}Benefit of concurrent navigation and tomography has been discussed, which is not necessarily that many approaches combine it.}
@@
 	\lreviewer{Fig. 2 and 3}{}{Fig 2 refers to central node position as Topside and Fig. 3 as Camp Seadragon. Suggest standardizing.}
+	Thank you. We have standardizes this to Topside.
 
-	\lreviewer{Lines 311, 359}{}{'Data' should be plural.}
+	\lreviewer{lines 311, 359}{}{'Data' should be plural.}
 
-	\lreviewer{Figure 6}{}{Suggest including receiver depth along with source depth in the title as they are both important in distinguishing between the cases.}
+	\lreviewer{Figure 6}{v1:fig:rangeError}{Suggest including receiver depth along with source depth in the title as they are both important in distinguishing between the cases.}
 
-	\lreviewer{Table I}{}{Table I indicates direction errors are positive.  Please note the sign convention in the caption.}
+	\lreviewer{Table I}{v1:tab:rangeErrorInSitu}{Table I indicates direction errors are positive.  Please note the sign convention in the caption.}
 
-	\lreviewer{Table II}{}{‘range estimation error metrics’ rather than ‘range estimation metrics’?}
+	\lreviewer{Table II}{v1:tab:rangeErrorV1V2}{‘range estimation error metrics’ rather than ‘range estimation metrics’?}
 
@@
 
-	\lreviewer{Line 7}{}{'Beaufort Lens' This is not yet a scientific term, suggest finding descriptive terms to use. What comprises the 'lens'? The Pacific Summer Water is the warm layer.}
+	\lreviewer{line 7}{v1:2.1}{'Beaufort Lens' This is not yet a scientific term, suggest finding descriptive terms to use. What comprises the 'lens'? The Pacific Summer Water is the warm layer.}
 
-	\lreviewer{Line 10}{}{'GPS aided', do  you mean GPS-navigated?}
+	\lreviewer{line 10}{v1:2.2}{'GPS aided', do  you mean GPS-navigated?}
 	Yes.
 
-	\lreviewer{Line 12}{}{'rival GPS' is a fairly bold statement when GPS is used as ground truth.}
+	\lreviewer{line 12}{v1:2.3}{'rival GPS' is a fairly bold statement when GPS is used as ground truth.}
 	{\color{blue} Accuracy comparable to conventional GPS; does not introduce decrease precision than is already documented by GPS puck precision; Effectively extends single GPS into the water column with comparable performance.
@@
 
-	\lreviewer{Line 89}{}{Is the 'lens' the warm water? Or a combination of layers? What does 'lens' mean acoustically? Normally in ocean acoustics people refer to ducts and layers.}
+	\lreviewer{line 89}{v1:2.4}{Is the 'lens' the warm water? Or a combination of layers? What does 'lens' mean acoustically? Normally in ocean acoustics people refer to ducts and layers.}
 
-	\lreviewer{Line 94}{}{Ice Tethered profiler data shows fairly slow variability in space and time in the Beaufort, but changes in characteristics occur as the deep basin transitions to the Chukchi, so what is meant by 'high degree' exactly?}
+	\lreviewer{line 94}{v1:2.5}{Ice Tethered profiler data shows fairly slow variability in space and time in the Beaufort, but changes in characteristics occur as the deep basin transitions to the Chukchi, so what is meant by 'high degree' exactly?}
 
-	\lreviewer{Fig 4 and Line 216}{}{Fig 4 and 216 - where did the baseline come from (exactly)? What database, etc.}
+	\lreviewer{lig 4}{}{Fig 4 - where did the baseline come from (exactly)? What database, etc.}
+	Please see Comment \ref{1.14}.
 
-	\lreviewer{Line 224}{}{why did a human have to chose an optimal set of weights? Why not just use the measured SSP, or an average of a few?}
+	\lreviewer{line 224}{v1:2.7}{why did a human have to chose an optimal set of weights? Why not just use the measured SSP, or an average of a few?}
 
-	\lreviewer{Line 231}{}{'abundance'. Is there a more precise way to say this? Bellhop does what it is programmed to…}
+	\lreviewer{line 231}{v1:2.8}{'abundance'. Is there a more precise way to say this? Bellhop does what it is programmed to…}
 
-	\lreviewer{Line 241}{}{'modem triggering time', do you mean 'detected arrival time'?}
+	\lreviewer{line 241}{v1:2.9}{'modem triggering time', do you mean 'detected arrival time'?}
 
-	\lreviewer{Line 249}{}{Use of the word 'range' for the 'navy range' type of use is intermixed with range as in distance.
+	\lreviewer{line 249}{v1:2.10}{Use of the word 'range' for the 'navy range' type of use is intermixed with range as in distance.
 Suggest not using 'range' and in 'navy range' and saying 'nav' or 'comms' system or network.}
 
-	\lreviewer{Line 261}{}{'bring'. Do you mean interpolate?}
+	\lreviewer{line 261}{v1:2.11}{'bring'. Do you mean interpolate?}
+	Yes. We have changed this accordingly.
 
-	\lreviewer{Line 281}{}{Awkward sentence, either unclear or extremely obvious. "...an acoustic arrival does not always take the direct path from source to the receiver,".}
+	\lreviewer{line 281}{v1:2.12}{Awkward sentence, either unclear or extremely obvious. "...an acoustic arrival does not always take the direct path from source to the receiver,".}
 
@@
 
-	\lreviewer{Line 319}{}{"This is likely driven by the prominence of the duct." Do you mean: consistency of sound speed within the duct where the rays were trapped and thus traveled?
+	\lreviewer{line 319}{v1:2.14}{"This is likely driven by the prominence of the duct." Do you mean: consistency of sound speed within the duct where the rays were trapped and thus traveled?
 Or something else?}
 
-	\lreviewer{Line 410}{}{'anomalous' - does this section need to be included? Or just the issue summarized and noted? Delete the table?}
+	\lreviewer{line 410}{}{'anomalous' - does this section need to be included? Or just the issue summarized and noted? Delete the table?}
 	{\color{blue}Delete table, summarize issue.}
 
-	\lreviewer{Lines 585-591}{}{Suggest deleting, not very appropriate for an academic paper.}
+	\lreviewer{lines 585-591}{}{Suggest deleting, not very appropriate for an academic paper.}
 
-	\lreviewer{Fig 10}{}{Would be good to have some conclusions in addition to the description.}
+	\lreviewer{Fig 10}{v1:fig:gvelMore}{Would be good to have some conclusions in addition to the description.}
 
@@
 
+	We have significantly condensed the background for this section and narrowed the discussion for this section about the pattern of psuedorange error and what it might mean for the precision of the GPS puck we used at high latitudes.
+
 	\reviewer{}{

### 2021-12-13 manuscript/main.tex
changes in manuscript to make cross referencing work and easier to read
tagging from OG line numbers in manuscript document

--- 2021-08-27 doc/main.tex
@@
 For positioning in post-processing, more computationally and/or labor intensive modeling methods are often employed to reduce the uncertainty driven by multipath arrivals.
-An autonomous underwater vehicle deployment in March 2020 in the Beaufort Sea faced significant operational challenges with a total ice-covered environment and complex acoustic propagation via the Beaufort Lens.
-We implement an embedded, stochastic, ray-based prediction of the horizontal group velocity and validate its performance with communication events between GPS-aided beacons, achieving a pseudorange mean absolute error of 11 m.
-In post-processing, we introduce a multipath criterion and achieve ranging performance that rivals GPS.
+\llabel{2.1} An autonomous underwater vehicle deployment in March 2020 in the Beaufort Sea faced significant operational challenges with a total ice-covered environment and complex acoustic propagation via the Beaufort Lens.
+\llabel{2.2} We implement an embedded, stochastic, ray-based prediction of the horizontal group velocity and validate its performance with communication events between GPS-aided beacons, achieving a pseudorange mean absolute error of 11 m.
+\llabel{2.3} In post-processing, we introduce a multipath criterion and achieve ranging performance that rivals GPS.
 To our knowledge, these results are the first field experiment to demonstrate a real-time, physics-based data processing to minimize ranging error in underwater navigation.
@@
 
-\section{\label{sec:1} Introduction}
+\section{Introduction}
+\label{sec:1}
 
@@
 Lightweight, mobile, and inexpensive compared to crewed and ship-based systems, these vehicles can enable a safer, data-forward style of field oceanography \citep{bellingham_robotics_2007,petillot_underwater_2019}.
+\llabel{test}
 
@@
 A recent study tracking 22 floats under the ice in the Weddell Gyre near Antarctica showed that they were unable to surface for 8 months and position uncertainty grew to 116 $\pm$ 148 km in the lateral directions \citep{chamberlain_observing_2018}.
+\llabel{test2}
 
@@
 Timing is the ability to maintain accurate and precise time, sometimes synchronized across multiple nodes.
-The advent of the WHOI micro-modem \citep{singh_underwater_2006} and synchronized chip scale atomic clocks \citep{gardner_second_2016} enabled ranging via one-way travel time (OWTT).
+\llabel{1.11} The advent of the WHOI micro-modem \citep{singh_underwater_2006} and synchronized chip scale atomic clocks \citep{gardner_second_2016} enabled ranging via one-way travel time (OWTT).
 
@@
 The new Arctic\textemdash one with much less multi-year sea ice\textemdash is experiencing significant changes in the water column as well \citep{mackinnon_warm_2021}.
-The Beaufort Lens is a highly dynamic, highly stratified water mass, keeping a neutrally buoyant layer of warm Pacific water at 50 to 70 m in depth.
+\llabel{2.4} The Beaufort Lens is a highly dynamic, highly stratified water mass, keeping a neutrally buoyant layer of warm Pacific water at 50 to 70 m in depth.
 Surrounded by cooler waters, it distinguishes a local maximum in the SSP, creating an upper and lower duct\textemdash the upper duct degrades signal coherence due to intensified ice interaction and the lower duct effectively traps sound for long range propagation \citep{poulsen_acoustic_2016}.
-Modeling output \citep{duda_long-range_2019,duda_effects_2021} and experimental observations \cite{kucukosmanoglu_observations_2021,badiey_azimuthal_2019,ballard_temporal_2020} indicate a high degree of spatio-temporal variability of the lower duct throughout the Chukchki and Beaufort Seas.
+\llabel{2.5} Modeling output \citep{duda_long-range_2019,duda_effects_2021} and experimental observations \cite{kucukosmanoglu_observations_2021,badiey_azimuthal_2019,ballard_temporal_2020} indicate a high degree of spatio-temporal variability of the lower duct throughout the Chukchki and Beaufort Seas.
 
@@
 % =========================================================================== %
-\section{\label{sec:methods} Methods}
+\section{Methods}
+\label{sec:methods}
 
@@
 \item Each modem buoy only has one transmit depth layer (30 or 90 m)
-\item The transmit and receive layers are independent
+\item \llabel{1.12} The transmit and receive layers are independent
 \item At any given time, the receive depth layers are the same across all four ice buoys
@@
 
-A key construction in the ICNN are parallel physical and virtual layers; given the small operational scale, on the order of kilometers, and a singular SSP, both layers are projected into a range independent coordinate system for ease of comparison.
+\llabel{1.13} A key construction in the ICNN are parallel physical and virtual layers; given the small operational scale, on the order of kilometers, and a singular SSP, both layers are projected into a range independent coordinate system for ease of comparison.
 Furthermore, because one of the buoys was configured to act as a virtual vehicle at any given time during the modem experiment, the tracking range was supplemented with an additional modem deployed from the camp site and configured to act as a virtual buoy; the transducer attached to the camp site modem was lowered to a depth of 20 m.
@@
 Subfigures (b-d) correspond to transmitter depths of 20, 30, and 90 m, respectively.
-Camp Seadragon, demarcated by the purple triangle, is situated just north of the continental shelf.}
+Topside, demarcated by the purple triangle, is situated just north of the continental shelf.}
   \label{fig:overview}
@@
 
-During ICEX20, the SSP was toggled between the historical``baseline'' and a ``weighted'' estimate based on a local CTD cast, i.e.
-an optimal set of weights chosen by a human decision maker via an interactive Tactical Decision Aid framework \citep{bhatt_embedded_2021}.
-For the post-processing analysis, we add the SSP from HYCOM for comparison, to mirror the information available on a submarine\textemdash modeled data, historical data, and \textit{in situ} data (personal conversation with LT B.
+\llabel{1.14} During ICEX20, the SSP was toggled between the historical``baseline'' and a ``weighted'' estimate based on a local CTD cast, i.e.
+an optimal set of weights chosen by a human decision maker \llabel{2.7} via an interactive Tactical Decision Aid framework \citep{bhatt_embedded_2021}.
+\llabel{1.8} For the post-processing analysis, we add the SSP from HYCOM for comparison, to mirror the information available on a submarine\textemdash modeled data, historical data, and \textit{in situ} data (personal conversation with LT B.
 Howard and LT CDR D.
@@
 Fig. \ref{fig:raytrace-zs30} shows eigenrays in a range independent coordinate system for all three sound speed environments, for a source depth of 30 m.
-It is important to note that BELLHOP produces an abundance of eigenrays; the ones displayed here are chosen by travel time proximity to the recorded data.
+\llabel{1.16} It is important to note that BELLHOP produces an \llabel{2.8}abundance of eigenrays; the ones displayed here are chosen by travel time proximity to the recorded data.
 With this filter, the chosen eigenrays show a greater amount of surface interactions as the SSP duct conditions increase.
@@
 The use of the acoustic modem network for tracking relies on the accurate estimates of travel times between the submerged platform and range modems, supported by clock synchronization and a pre-determined scheduling of acoustic events.
-For the Beaufort Lens in particular, the strong multipath effects make it virtually impossible to deterministically predict the modem triggering time.
+\llabel{2.9} For the Beaufort Lens in particular, the strong multipath effects make it virtually impossible to deterministically predict the modem triggering time.
 
@@
 
-In the ICEX20 configuration, the acoustic tracking is running on the topside computer, which controls the integrated communication and navigation range.
+\llabel{2.10} In the ICEX20 configuration, the acoustic tracking is running on the topside computer, which controls the integrated communication and navigation range.
 Here we assume that the group velocities $u_{i,j}$ are smoothly varying over the course of a vehicle mission, i.e., with respect to range, mission time, and the frequency of updates relative to vehicle motion.
@@
 
-When the topside tracking framework receives a modem message, with a time delay, $\Delta t$, from one of the range modems, it will request a new estimate of the group velocity and its associated uncertainty.
+\llabel{1.15} When the topside tracking framework receives a modem message, with a time delay, $\Delta t$, from one of the range modems, it will request a new estimate of the group velocity and its associated uncertainty.
 The group velocity estimate is found using the vehicle's reported depth and the extrapolated navigation solution for range, $\hat{r}$, as inputs for the ray tracing program.
@@
 
-The initial call to BELLHOP is over a sparse local grid, from which the plane wave approximation is used to bring the ray to the range and depth posited by the onboard tracking solution.
+\llabel{2.11} The initial call to BELLHOP is over a sparse local grid, from which the plane wave approximation is used to interpolate the ray to the range and depth posited by the onboard tracking solution.
 For each grid point, BELLHOP will produce a number of arrivals resulting from multiple propagation paths for any source-receiver pair.
@@
 
-As shown in the eigenray traces of Fig.
+\llabel{2.12} As shown in the eigenray traces of Fig.
 \ref{fig:raytrace-zs30}, an acoustic arrival does not always take the direct path from source to the receiver, and the path it takes is extremely dependent on the sound speed profile.
@@
 Bottom bounces are not encoded separately because of ray's tendency to refract upward, not due to information limitations. 
-In addition, for a more consistent comparison, the local sparse grid is standardized as 11$\times$11 points over 10 m in range and 20 m in depth, which is similar to those from the vehicle logs.
+\llabel{1.17} In addition, for a more consistent comparison, the local sparse grid is standardized as 11$\times$11 points over 10 m in range and 20 m in depth, which is similar to those from the vehicle logs.
 The extra size in depth addresses the computational limitations of how time-bounded eigenrays seem to stack vertically in the water column, as shown in Fig \ref{fig:rays3env-zs30}.
@@
 The modem experiment generated 811 beacon to beacon communication events with their own real-time MBC group velocity predictions.
-As expected, the algorithm is generally overestimating range as it resolves a direct path that does not represent the actual arrival.
+\llabel{1.18} As expected, the algorithm is generally overestimating range as it resolves a direct path that does not represent the actual arrival.
 This section examines the absolute range error between the pseudorange and the GPS derived range with the caveat that the data collected is fairly unbalanced across source depths and sound speed conditions.
@@
 Analysis shows that where there is overlap in range between sound speed conditions used for the real-time MBC, the range error difference is no more than a few meters.
-This is likely driven by the prominence of the duct.
+\llabel{2.14} This is likely driven by the prominence of the duct.
 
@@
 However, that data is too small to draw significant conclusions.
-Discontinuities in multipath classification verify our hypothesis for its importance to a smoothly varying group velocity, as shown in the cluster for a receiver depth of 30 m, where HYCOM jumps from one to two bounces amidst the baseline SSP and weighted SSP smoothly increasing while consistently seeing two and three bounces, respectively.
+\llabel{1.19} Discontinuities in multipath classification verify our hypothesis for its importance to a smoothly varying group velocity, as shown in the cluster for a receiver depth of 30 m, where HYCOM jumps from one to two bounces amidst the baseline SSP and weighted SSP smoothly increasing while consistently seeing two and three bounces, respectively.
 
@@
 
-Many approaches to underwater navigation combine it with acoustic tomography, i.e., a joint estimation of both source and receiver locations and the ocean volume between them.
+\llabel{1.20} Many approaches to underwater navigation combine it with acoustic tomography, i.e., a joint estimation of both source and receiver locations and the ocean volume between them.
 There has been considerable success at this effort in post-processing methods, which utilize intensive\textemdash and due to the non-linearity of sound propagation, often brute force\textemdash computational methods.

### 2021-12-13 rebuttal/rebuttal-letter/trunk-rebuttal.tex
starting to write edits

@@
 We thank the reviewers and editor for their thoughtful feedback and comments in improving this work.
-We have significantly restructured the manuscript along several suggestions to separate the real-time navigation methodology, the real-time localization results (Mininum Bounce Criteria), and the post-processing re-localization improvement (Nearest Bounce Criteria).
-We have also introduced that language (navigation, localization, and re-localization) more explicitly and consistently to avoid confusion and better contextualize our contribution.
+We have significantly restructured the manuscript along several suggestions to separate the real-time navigation methodology, the real-time positioning results (Mininum Bounce Criteria), and the post-processing re-positioning improvement (Nearest Bounce Criteria).
+We have also introduced that language (navigation, positioning, and re-positioning) more explicitly to avoid confusion and better contextualize our contribution.
 
@@
 
-	\ebline{This paper demonstrates how a positioning system over ~1.5-3 km of range achieves 11 m of positioning uncertainty.
+	\ebline{This paper demonstrates how a positioning system over \~1.5-3 km of range achieves 11 m of positioning uncertainty.
 The work demonstrates how a single group velocity calculated from acoustic rays predicted by BELLHOP can be used to estimate ranges to be used in a LBL network for positioning.
@@
 
-	This summary key concepts and results presented in the paper.
-However, this system was in fact deployed onboard an underwater vehicle during under-ice deployments.
-The motivation behind this work is to improve the standard LBL navigation’s conversion of travel time to range for an under-ice environment in a way that could be generalized to any acoustic waveguide.
-To do this, we introduce the Minimal Bounce Criteria method to operationalize real-time acoustic modeling into an effective horizontal group velocity estimate.
-
-	This system was implemented on an autonomous underwater vehicle (AUV) on multiple missions, including an 11 km untethered, under-ice deployment, in March 2020.
-The AUV was deployed through a hydrohole from Topside and due to a disk error, stalled underneath the ice surface while transmitting its perceived location.
-The AUV was found within a meter away from its reported location, which serves as strong (but qualitative) evidence of the navigation performance provided by the single group velocity calculation.
-Because the navigation updates throughout the AUV mission have no GPS ground truth to compare to, this paper evaluates the same system performance across the GPS-navigated LBL beacons.
-
-	The real-time error from the Nearest Bounce Criteria (NBC) achieves a mean absolute positioning error of 11 m.
-Upon further evaluation of the acoustic events, the Minimal Bounce Criteria (MBC) is run in the same automated pipeline in post-processing (akin to real-time computational constraints and needs) and achieves an error of roughly 3 m.
-
-	Both the NBC and MBC pseudorange estimations are compared to the GPS data on the LBL beacons.
-The error itself is small and the pattern of error exhibited indicates that the GPS ground truth data has significant noise (drift) that is not replicated by either pseudorange estimation method.
-For the MBC in particular, the post-processing error is within the precision of the GPS sensor, suggesting that its performance rivals that of GPS.
+	This summary captures key results presented in the paper.
+However, we note that the system was successfully deployed on an underwater vehicle during ICEX20.
+The results from this work are specifically limited to the methodology behind calculating a single group velocity, which can be quantified through beacon-to-beacon communications.
 
@@
 
-	We agree that there are limitations to how transferable this analysis is, which have now been made clearer in the abstract, introduction, and conclusion.
+	We have tempered the claims in the abstract to reflect the limitations of the dataset collected and the transferability of the analysis in this paper.
+	We have also included a discussion of how these results might compare for various depths and ranges during vehicle operations (see \hl{XXX}).
+
+	With respect to the concentrating on real-time results versus post-processing results, we would point out that the AUV was deployed through a hydrohole from Topside and due to a disk error, stalled underneath the ice surface while transmitting its perceived location.
+	The AUV was found within a meter from its reported location, which serves as strong (but qualitative) evidence of the navigation performance provided by the single group velocity calculation.
+	Work currently review in the Journal of Field Robotics details the other components to vehicle navigation (DVL-aided INS and a vehicle flight model) along with the magnitude of position corrections.
+
+	Given that our vehicle navigation data only has one GPS point to compare to, we use the beacon-to-beacon communications as a natural experiment to compare the effectiveness of the real-time NBC method.
+	The real-time section is small because only one modem at a time ran the vehicle behavior to calculate these acoustic pseudoranges.
+	A robust comparison is tainted by the mismatch of experimental conditions in estimated sound speed environment and source / receiver position.
+
+	We extended the dataset to all possible communication events through the same automated pipeline to decouple the effects of the sound speed environment from source and receiver position.
+This is in effect no different than running the computation during the field; the real-time NBC data is a subset of the post-processed NBC data.
+
+	While post-processing, we began to see a pattern of range overestimation using the NBC method.
+Thus, we developed the Minimal Bounce Criteria instead and ran it through the same pipeline, subject to the same computational constraints.
+
+	By comparing the errors between these two pseudorange estimation methods, we noticed a surprising underlying pattern compared to the ground truth GPS data.
+	Further investigation suggests there is noise in the GPS measurement that is not incurred in the travel time measurement that informs the acoustic range estimation.
+
+	The manuscript has been significantly restructured to separate analyses along this narrative.
 
@@
 Also, it would be good to state this hypothesis more clearly earlier in the manuscript.}
+	We agree that a few short fixed ranges do not validate this hypothesis.
+With the overarching motivation and design of LBL navigation, we have recast this hypothesis as an assumption.
+We would note, again, that the vehicle navigation performance would better determine a continuously changing group velocity, but this data would have no way of being verified.
 
@@
 	\lreviewer{Fig. 2 and 3}{}{Fig 2 refers to central node position as Topside and Fig. 3 as Camp Seadragon. Suggest standardizing.}
-	Thank you. We have standardizes this to Topside.
+	Thank you. We have standardized this to Topside.
 
 	\lreviewer{lines 311, 359}{}{'Data' should be plural.}
+	Thank you. This has been corrected.
 
@@
 	\lreviewer{Table I}{v1:tab:rangeErrorInSitu}{Table I indicates direction errors are positive. Please note the sign convention in the caption.}
+	{\color{blue} caption is correct but text references an ``overstimation'' of range that is not obvious given absolute error shown.}
 
 	\lreviewer{Table II}{v1:tab:rangeErrorV1V2}{‘range estimation error metrics’ rather than ‘range estimation metrics’?}
+	Agreed. We have made the suggested change.
 
 	\lreviewer{Table II and III}{}{Suggest removing ‘algorithm’ label.  Also substitute ‘MBC’ and ‘NBC’ for minimal and nearest to match the text.}
+	Agreed. We have made the suggested changes.
 
 	\lreviewer{}{}{There are several figures and tables which are not referenced properly (Line 305, 391, 411)}
+	Thank you. This has been corrected.
 
@@
 	\lreviewer{line 10}{v1:2.2}{'GPS aided', do you mean GPS-navigated?}
-	Yes.
+	We have changed this to ``GPS-linked'' because navigation implies self-movement.
 

### 2021-12-13 manuscript/main.tex
starting to write edits

@@
 
+% The motivation behind this work is to improve the standard LBL navigation’s conversion of travel time to range for an under-ice environment in a way that could be generalized to any acoustic waveguide.
+To do this, we introduce the Minimal Bounce Criteria method to operationalize real-time acoustic modeling into an effective horizontal group velocity estimate.
+
+% 	This system was implemented on an autonomous underwater vehicle (AUV) on multiple missions, including an 11 km untethered, under-ice deployment, in March 2020.
+The AUV was deployed through a hydrohole from Topside and due to a disk error, stalled underneath the ice surface while transmitting its perceived location.
+The AUV was found within a meter away from its reported location, which serves as strong (but qualitative) evidence of the navigation performance provided by the single group velocity calculation.
+Because the navigation updates throughout the AUV mission have no GPS ground truth to compare to, this paper evaluates the same system performance across the GPS-navigated LBL beacons.
+
+% 	The real-time error from the Nearest Bounce Criteria (NBC) achieves a mean absolute positioning error of 11 m.
+Upon further evaluation of the acoustic events, the Minimal Bounce Criteria (MBC) is run in the same automated pipeline in post-processing (akin to real-time computational constraints and needs) and achieves an error of roughly 3 m.
+
+% 	Both the NBC and MBC pseudorange estimations are compared to the GPS data on the LBL beacons.
+The error itself is small and the pattern of error exhibited indicates that the GPS ground truth data has significant noise (drift) that is not replicated by either pseudorange estimation method.
+For the MBC in particular, the post-processing error is within the precision of the GPS sensor, suggesting that its performance rivals that of GPS.
+
 \begin{abstract}
-Almost all underwater navigation paradigms rely on the conversion of recorded travel time into range to compute a trilateration solution.
-For real-time navigation, this conversion is treated deterministically.
-For positioning in post-processing, more computationally and/or labor intensive modeling methods are often employed to reduce the uncertainty driven by multipath arrivals.
-\llabel{2.1} An autonomous underwater vehicle deployment in March 2020 in the Beaufort Sea faced significant operational challenges with a total ice-covered environment and complex acoustic propagation via the Beaufort Lens.
-\llabel{2.2} We implement an embedded, stochastic, ray-based prediction of the horizontal group velocity and validate its performance with communication events between GPS-aided beacons, achieving a pseudorange mean absolute error of 11 m.
-\llabel{2.3} In post-processing, we introduce a multipath criterion and achieve ranging performance that rivals GPS.
-To our knowledge, these results are the first field experiment to demonstrate a real-time, physics-based data processing to minimize ranging error in underwater navigation.
+The long baseline (LBL) underwater navigation paradigm relies on the conversion of recorded travel time to range to trilaterate for position.
+For real-time navigation, this conversion assumes an isovelocity sound speed.
+For re-navigation in post-processing, computationally and/or labor intensive acoustic modeling may be employed to reduce uncertainty driven by multipath arrivals.
+This work demonstrates an embedded, ray-based prediction method of the effective horizontal group speed to minimize vehicle position error.
+\llabel{2.1} This method, the Minimal Bounce Criteria (MBC), was implemented for a small scale AUV-LBL system in March 2020, in the Beaufort Sea, in total under-ice conditions and a double ducted acoustic propagation environment.
+The vehicle was successfully deployed and recovered.
+\llabel{2.2} Given the lack of GPS data throughout the vehicle's mission, however, the localization performance is quantified between GPS-linked beacon-to-beacon connections where a single beacon ran the vehicle's navigation stack.
+The real-time positioning error is roughly 11 meters at distances up to 3 km.
+In post-processing, the dataset is extended to all possible beacon-to-beacon events.
+The localization performance is reevaluated with an operationally equivalent navigation stack using modeled, historical, and data-based sound speed profiles for the MBC and for a new method, the Nearest Bounce Criteria (NBC).
+The best performing median re-positioning errors are roughly 10 and 2 meters, respectively.
+\llabel{2.3} Further investigation suggests that the GPS ground truth data have some noise that is not reciprocated in either acoustic localization method.
+
+% Almost all underwater navigation paradigms rely on the conversion of recorded travel time into range to compute a trilateration solution.
+% For real-time navigation, this conversion is treated deterministically.
+% For positioning in post-processing, more computationally and/or labor intensive modeling methods are often employed to reduce the uncertainty driven by multipath arrivals.
+% \llabel{2.1} An autonomous underwater vehicle deployment in March 2020 in the Beaufort Sea faced significant operational challenges with a total ice-covered environment and complex acoustic propagation via the Beaufort Lens.
+% \llabel{2.2} We implement an embedded, stochastic, ray-based prediction of the horizontal group velocity and validate its performance with communication events between GPS-aided beacons, achieving a pseudorange mean absolute error of 11 m.
+% \llabel{2.3} In post-processing, we introduce a multipath criterion and achieve ranging performance that rivals GPS.
+% To our knowledge, these results are the first field experiment to demonstrate a real-time, physics-based data processing to minimize ranging error in underwater navigation.
 \end{abstract}
@@
 \llabel{1.17} In addition, for a more consistent comparison, the local sparse grid is standardized as 11$\times$11 points over 10 m in range and 20 m in depth, which is similar to those from the vehicle logs.
-The extra size in depth addresses the computational limitations of how time-bounded eigenrays seem to stack vertically in the water column, as shown in Fig \ref{fig:rays3env-zs30}.
+The extra size in depth addresses the computational limitations of how time-bounded eigenrays seem to stack vertically in the water column, as shown in Fig \ref{fig:raytrace-zs30}.
 
@@
 \end{tabular}
-\caption[Range estimation error for \textit{in situ} events]{The range estimation error for \textit{in situ} events.
+\caption[Range estimation error for \textit{in situ} events]{The range estimation error metrics for \textit{in situ} events.
 The error here is presented as the absolute, not directional, error.
@@
 Overall, analysis shows that the higher multipath classification produces more accurate range estimates, likely driven by a tighter bundle of rays.
-However, that data is too small to draw significant conclusions.
+However, that data are too small to draw significant conclusions.
 \llabel{1.19} Discontinuities in multipath classification verify our hypothesis for its importance to a smoothly varying group velocity, as shown in the cluster for a receiver depth of 30 m, where HYCOM jumps from one to two bounces amidst the baseline SSP and weighted SSP smoothly increasing while consistently seeing two and three bounces, respectively.
@@
 The improvement from MBC to NBC is most evident for the realistic sound speed; while the HYCOM SSP improves from a median absolute range error of 6.41 to 4.61 m, the baseline SSP improves from 10.30 to 2.27 m, and the weighted SSP improves from 13.28 to 2.12 m.
-Table \ref{tab:rangeErrorPost} shows further statistics on the absolute range error by SSP and group velocity algorithm.
+Table \ref{tab:rangeErrorV1V2} shows further statistics on the absolute range error by SSP and group velocity algorithm.
 The order of magnitude improvement in the ducted SSPs demonstrate the effectiveness of the algorithm exploiting the multipath conditions.
@@
  & \multicolumn{2}{c|}{\textbf{Baseline} } & \multicolumn{2}{c|}{\textbf{Chosen Weights} } & \multicolumn{2}{c}{\textbf{HYCOM}} \\
- algorithm & \cellcolor[HTML]{EFEFEF}minimal & nearest & \cellcolor[HTML]{EFEFEF} minimal& nearest & \cellcolor[HTML]{EFEFEF}minimal & nearest \\ \hline
+  & \cellcolor[HTML]{EFEFEF}MBC & NBC & \cellcolor[HTML]{EFEFEF} MBC& NBC & \cellcolor[HTML]{EFEFEF}MBC & NBC \\ \hline
  minimum [m] 	& \cellcolor[HTML]{EFEFEF}0.01 & 0.00	& \cellcolor[HTML]{EFEFEF}0.00 	& 0.00 	& \cellcolor[HTML]{EFEFEF}0.11 & 0.01 \\
@@
 
-As shown in table \ref{tab:rangeErrorPost}, there is a striking maximum range error of 1491 m for the weighted SSP in the minimal bounce criteria.
+As shown in table \ref{tab:rangeErrorV1V2}, there is a striking maximum range error of 1491 m for the weighted SSP in the minimal bounce criteria.
 This section isolates the events that encompass that outlier for further examination, summarized in table \ref{tab:rangeErrorPostExplanation}.
@@
  & \multicolumn{2}{c|}{\textbf{Baseline} } & \multicolumn{2}{c|}{\textbf{Chosen Weights} } & \multicolumn{2}{c}{\textbf{HYCOM}} \\
- algorithm & \cellcolor[HTML]{EFEFEF}minimal & nearest & \cellcolor[HTML]{EFEFEF} minimal& nearest & \cellcolor[HTML]{EFEFEF}minimal & nearest \\ \hline
+  & \cellcolor[HTML]{EFEFEF}minimal & nearest & \cellcolor[HTML]{EFEFEF} MBC& NBC & \cellcolor[HTML]{EFEFEF}minimal & nearest \\ \hline
  \# bounces 				& \cellcolor[HTML]{EFEFEF}1 		& 3 		& \cellcolor[HTML]{EFEFEF}1 		& 4 		& \cellcolor[HTML]{EFEFEF}0 	 & 2 \\

### 2021-12-15 manuscript/main.tex
discussing abstract w/ Oscar

@@
 This work demonstrates an embedded, ray-based prediction method of the effective horizontal group speed to minimize vehicle position error.
-\llabel{2.1} This method, the Minimal Bounce Criteria (MBC), was implemented for a small scale AUV-LBL system in March 2020, in the Beaufort Sea, in total under-ice conditions and a double ducted acoustic propagation environment.
+\llabel{2.1} This method was implemented for a small scale AUV-LBL system in March 2020, in the Beaufort Sea, in total under-ice conditions and a double ducted acoustic propagation environment.
 The vehicle was successfully deployed and recovered.
@@
 In post-processing, the dataset is extended to all possible beacon-to-beacon events.
-The localization performance is reevaluated with an operationally equivalent navigation stack using modeled, historical, and data-based sound speed profiles for the MBC and for a new method, the Nearest Bounce Criteria (NBC).
+The localization performance is reevaluated with an operationally equivalent navigation stack using modeled, historical, and locally observed sound speed profiles for a Minimal Bounce Criteria (MBC), as deployed in real-time, and an improved algorithm, the Nearest Bounce Criteria (NBC).
 The best performing median re-positioning errors are roughly 10 and 2 meters, respectively.
-\llabel{2.3} Further investigation suggests that the GPS ground truth data have some noise that is not reciprocated in either acoustic localization method.
+\llabel{2.3} Further investigation suggests that the GPS ground truth data have some noise that is not reciprocated in the travel time data.
 
+%% OLD ABSTRACT
 % Almost all underwater navigation paradigms rely on the conversion of recorded travel time into range to compute a trilateration solution.
@@
 % To our knowledge, these results are the first field experiment to demonstrate a real-time, physics-based data processing to minimize ranging error in underwater navigation.
+
 \end{abstract}

### 2021-12-15 rebuttal/rebuttal-letter/trunk-rebuttal.tex
changes to trunk
finalized cover letter

@@
 \documentclass{article}
-	\def\papertitle{Under-ice acoustic navigation using real-time model-aided range estimation}
+	\def\papertitle{Under-ice acoustic navigation using real time model-aided range estimation}
 	\def\authors{E. Bhatt, O. V\'{i}quez, H. Schmidt}
@@
 
+% table of contents
+\renewcommand{\contentsname}{}
+\begingroup
+\let\pagebreak\relax
+\setcounter{tocdepth}{1}
+\tableofcontents
+\endgroup
+
+\section*{Cover Letter}
 Dear Editors,
@@
 We thank the reviewers and editor for their thoughtful feedback and comments in improving this work.
-We have significantly restructured the manuscript along several suggestions to separate the real-time navigation methodology, the real-time positioning results (Mininum Bounce Criteria), and the post-processing re-positioning improvement (Nearest Bounce Criteria).
-We have also introduced that language (navigation, positioning, and re-positioning) more explicitly to avoid confusion and better contextualize our contribution.
+We have refashioned the abstract and introduction to better elucidate the motivation for this work as a model-aided component for LBL navigation in total under-ice conditions, which could be further generalized for any acoustic environment.
+Based on several suggestions, we have restructured the manuscript and separated the real-time navigation methodology, the real-time positioning results (Minimum Bounce Criteria), and the post-processing re-positioning improvement (Nearest Bounce Criteria).
+Lastly, when evaluating our methodology, we have removed qualitative language like ``GPS-like'' or ``rivaling GPS'' in favor of quantitative descriptions of ranging error.
 
-We have also refashioned the abstract and introduction to better elucidate the motivation for this work as a model-aided component for LBL navigation in total under-ice conditions, which could be further generalized for any acoustic environment.
-The rebuttal letter is organized with sections delineated by the review, shown in the table of contents.
+We also more prominently establish definitions for navigation, positioning, and re-positioning to better contextualize our results and contributions, based on this discussion in Howe et al (2019) \emph{Observing the Oceans Acoustically}:
+\begin{quote}
+PNT is a combination of three distinct, constituent capabilities:
+
+\begin{enumerate}
+	\item Positioning is the ability to accurately and precisely determine one’s location referenced to a standard geodetic system;
+	\item Navigation is the ability to determine current and desired position (relative or absolute) and apply corrections to course, orientation, and speed to attain a desired position anywhere in the domain of concern;
+	\item Timing is the ability to acquire and maintain accurate and precise time anywhere in the domain of interest within user-defined timeliness parameters; it also includes time transfer.
+\end{enumerate}
+
+Navigation is real time and necessarily depends on positioning, and positioning depends on timing.
+\end{quote}
+
+Given the inescapability of not having GNSS ground truth for our AUV navigation data, and the tension of evaluating our travel time to range conversion on events between LBL beacons in real time and post-processing, we further delineate that positioning is in real time and re-positioning is post-processed; similarly, we interpret reviewer feedback using these definitions.
+
+The rebuttal letter is organized with sections delineated by the review document and email.
+We look forward to your comments and are happy to discuss any aspect of this rebuttal letter or the manuscript further.
+Thank you again for your time and consideration.
+
+Please note that the corresponding author is on parental leave for the month of January.
 
@@
 
-% table of contents
-\renewcommand{\contentsname}{}
-\begingroup
-\let\pagebreak\relax
-\setcounter{tocdepth}{1}
-\tableofcontents
-\endgroup
-
 %% ------- COMMENTS FROM REVIEWER ONE ------ %%
@@
 
-	\ebline{This paper demonstrates how a positioning system over \~1.5-3 km of range achieves 11 m of positioning uncertainty.
+	\ebline{This paper demonstrates how a positioning system over \texttt{\small $\sim$}1.5-3 km of range achieves 11 m of positioning uncertainty.
 The work demonstrates how a single group velocity calculated from acoustic rays predicted by BELLHOP can be used to estimate ranges to be used in a LBL network for positioning.
@@
 However, we note that the system was successfully deployed on an underwater vehicle during ICEX20.
-The results from this work are specifically limited to the methodology behind calculating a single group velocity, which can be quantified through beacon-to-beacon communications.
+The results from this work are specifically limited to the methodology behind calculating a single group velocity, which we attempt to quantify through beacon-to-beacon communications.
 
@@
 	We have tempered the claims in the abstract to reflect the limitations of the dataset collected and the transferability of the analysis in this paper.
+	We have amended Figure \ref{v1:fig:vehicleRecovery} to include vehicle data and illustrate that this approach was successful in a real field environment.
 	We have also included a discussion of how these results might compare for various depths and ranges during vehicle operations (see \hl{XXX}).
@@
 	The real-time section is small because only one modem at a time ran the vehicle behavior to calculate these acoustic pseudoranges.
-	A robust comparison is tainted by the mismatch of experimental conditions in estimated sound speed environment and source / receiver position.
+	A robust comparison is difficult given the mismatch of experimental conditions in estimated sound speed environment as well as source and receiver position.
 
@@
 Line 176 mentions that the SSP and the physics of ray refraction are often overlooked in positioning algorithms; this is, I presume, referring to real-time results or short-range results, but doesn’t specifically say this.}
+	{\color{blue} Suggestion to discuss by range, vehicle, and propagation conditions is sound.
+However, there are no other short range deployments in double ducted environments.
+That line says SSP and physics of ray refraction are often overlooked in *navigation* algorithms, which means only real-time results.}
 
@@
 Lines 370-372 state “The increased error for these reciprocal transmission paths is most likely driven by the computational artifacts encountered when propagating through the steep sound speed gradients of the lens and through the shadow zone.“ The authors might consider providing more information on these artifacts.}
+
+	{\color{blue}Thank you. The heart of this paper is injecting an acoustic model into a long baseline navigation.}
 
@@
 How many events are successful in comparison with those that are considered unsuccessful and what generally causes an event to be deemed unsuccessful?}
+
+	{\color{blue} successful is if a single modem can decipher message. }
 
@@
 
+	Thank you. We politely
+
 	\reviewer{}
@@
 	\end{itemize}}
+	Thank you for pointing this out.
+We have removed personal pronouns from the abstract and placed a precise subject of the work within the first two paragraphs before going into broader background material.
 

### 2021-12-20 manuscript/main.tex
finalized cover letter
minor changes

@@
 \llabel{2.3} Further investigation suggests that the GPS ground truth data have some noise that is not reciprocated in the travel time data.
-
-%% OLD ABSTRACT
-% Almost all underwater navigation paradigms rely on the conversion of recorded travel time into range to compute a trilateration solution.
-% For real-time navigation, this conversion is treated deterministically.
-% For positioning in post-processing, more computationally and/or labor intensive modeling methods are often employed to reduce the uncertainty driven by multipath arrivals.
-% \llabel{2.1} An autonomous underwater vehicle deployment in March 2020 in the Beaufort Sea faced significant operational challenges with a total ice-covered environment and complex acoustic propagation via the Beaufort Lens.
-% \llabel{2.2} We implement an embedded, stochastic, ray-based prediction of the horizontal group velocity and validate its performance with communication events between GPS-aided beacons, achieving a pseudorange mean absolute error of 11 m.
-% \llabel{2.3} In post-processing, we introduce a multipath criterion and achieve ranging performance that rivals GPS.
-% To our knowledge, these results are the first field experiment to demonstrate a real-time, physics-based data processing to minimize ranging error in underwater navigation.
-
 \end{abstract}
@@
 \label{sec:1}  
+Autonomous underwater vehicles (AUVs) are increasingly capable platforms to explore and samlpe the ocean, particularly for remote and/or dangerous environments.
+However, navigation uncertainty is a major challenge in considering AUVs as reliable and standard tools for oceanographic research.
+Compared to their land and air-based counterparts, underwater vehicles face a unique set of challenges brought on by the water column itself:  while land and air-based robots utilize the Global Navigation Satellite System (GNSS) to achieve stunning location accuracy and precision throughout the duration of their missions, AUVs cannot access GPS while underwater due to the rapid attenuation of electromagnetic waves \citep{preisig_acoustic_2007}.
+Typically, an AUV can stall on the surface to receive a GPS fix, but this foolproof method of re-positioning is inaccessible in an ice-covered environment or other GPS-denied situations.
 
-Autonomous underwater vehicles (AUVs) are an increasingly capable platform to explore and sample the ocean, particularly for remote and/or dangerous environments.
-Lightweight, mobile, and inexpensive compared to crewed and ship-based systems, these vehicles can enable a safer, data-forward style of field oceanography \citep{bellingham_robotics_2007,petillot_underwater_2019}.
-\llabel{test}
+% Long baseline navigation
+Though there many acoustic positioning methods, the long baseline (LBL) is closest in style
 
-But navigation uncertainty is a major challenge in considering AUVs as reliable and standard tools for oceanographic research.
-While land and air-based robots utilize satellite-based global positioning system (GPS) to achieve stunning location accuracy and precision throughout the duration of their missions, AUVs cannot access GPS while underwater due to the rapid attenuation of electromagnetic waves \citep{preisig_acoustic_2007}.
-A vehicle can stall on the surface to receive a GPS fix, but this foolproof method of re-positioning is inaccessible in an ice-covered environment or other GPS-denied situations.
-A recent study tracking 22 floats under the ice in the Weddell Gyre near Antarctica showed that they were unable to surface for 8 months and position uncertainty grew to 116 $\pm$ 148 km in the lateral directions \citep{chamberlain_observing_2018}.
-\llabel{test2}
+% A recent study tracking 22 floats under the ice in the Weddell Gyre near Antarctica showed that they were unable to surface for 8 months and position uncertainty grew to 116 $\pm$ 148 km in the lateral directions \citep{chamberlain_observing_2018}.
 

### 2021-12-22 rebuttal/rebuttal-letter/trunk-rebuttal.tex
minor changes
hi

@@
 
-We also more prominently establish definitions for navigation, positioning, and re-positioning to better contextualize our results and contributions, based on this discussion in Howe et al (2019) \emph{Observing the Oceans Acoustically}:
+We also more prominently establish definitions for navigation, positioning, and re-positioning to better contextualize our results and contributions, based on this discussion in Howe et al (2019) \emph{Observing the Oceans Acoustically} (and many others):
+
 \begin{quote} 
@@
 % \pageref*{v1:test}
-
 % \formatLineTracking{test}{v1:test}
@@
 	\reviewer{}{The introduction and conclusion sections should be tightened up to clarify the state of the art in various arenas of underwater vehicle navigation and how this work fits in.
+
 	The error metrics presented here should be specifically compared with other positioning systems or algorithms over similar spatial scales that do not integrate the ray propagation physics.
+
 	Various ranging and positioning methods are referenced, but the narrative jumps between different types of AUVs and implementations and between scales of a few kilometers and hundreds of kilometers.
 This should be discussed in terms of range, vehicle, and propagation conditions.
+
 	Line 176 mentions that the SSP and the physics of ray refraction are often overlooked in positioning algorithms; this is, I presume, referring to real-time results or short-range results, but doesn’t specifically say this.}
-	{\color{blue} Suggestion to discuss by range, vehicle, and propagation conditions is sound.
-However, there are no other short range deployments in double ducted environments.
-That line says SSP and physics of ray refraction are often overlooked in *navigation* algorithms, which means only real-time results.}
+
+	We have re-written the introduction and conclusion to clarify that our work adds a necessary model-aided component for ranging in LBL networks for complex acoustic environments, and that it could be extended for long range propagation as well.
+
+	Present error metrics in reference to ...
+
+	The suggestion to discuss in terms of range, vehicle, and propagation conditions is helpful, and we have included that framing explicitly in the manuscript (LINE LINE).
+However, considering a union of all those operational variables, there are no comparable deployments.
+	Previous Arctic deployments are either in the Marginal Ice Zone at late summer, with little to no ice, or recorded the classical Arctic upward refracting profile, where isovelocity scaling could be good enough.
+	The few Arctic deployments that have seen a double ducted environment were on gliders, with little to no sea ice, and only one study navigated and re-navigated the gliders acoustically; other studies navigated them with dynamic flight models and GPS and re-navigated them acoustically.
+
+	Lastly, we have removed the line in question, but would point out it was originally written as ``navigation'', not ``positioning''.
 
@@
 Also, perhaps change the axis on Figure 5 to read something like ‘implied group velocity.”}
+
+	``It is the variation of group speed from source to source in underwater navigation that is the fundamental challenge to implement a “GNSS-like” navigation system in the ocean with acoustics.''
 

### 2021-12-22 manuscript/main.tex
hi

@@
 \label{sec:1}  
-Autonomous underwater vehicles (AUVs) are increasingly capable platforms to explore and samlpe the ocean, particularly for remote and/or dangerous environments.
-However, navigation uncertainty is a major challenge in considering AUVs as reliable and standard tools for oceanographic research.
-Compared to their land and air-based counterparts, underwater vehicles face a unique set of challenges brought on by the water column itself:  while land and air-based robots utilize the Global Navigation Satellite System (GNSS) to achieve stunning location accuracy and precision throughout the duration of their missions, AUVs cannot access GPS while underwater due to the rapid attenuation of electromagnetic waves \citep{preisig_acoustic_2007}.
-Typically, an AUV can stall on the surface to receive a GPS fix, but this foolproof method of re-positioning is inaccessible in an ice-covered environment or other GPS-denied situations.
+Autonomous underwater vehicles (AUVs) are increasingly capable platforms to explore and sample the ocean, particularly for remote and/or dangerous regions.
+However, navigation uncertainty is a major challenge in considering AUVs as standard tools for oceanographic research.
+While land and air-based robots utilize information from  Global Positioning System (GPS) updates\textemdash or more broadly, any Global Navigation Satellite System (GNSS)\textemdash to achieve stunning location accuracy and precision throughout the duration of their missions, AUVs cannot access GPS while underwater due to the rapid attenuation of electromagnetic waves.
+Therefore, underwater vehicle navigation has relied on any combination of dead reckoning, hydrodynamic models, inertial navigation systems, doppler velocity logs, and acoustic baseline positioning systems \citep{paull_auv_2014}.
+Limiting navigation error and drift requires an AUV to periodically stall on the surface and obtain a GPS fix to reset its position error.
+This foolproof method of self-positioning is undesirable for stealth, adverse weather conditions, and mission efficiency, and inaccessible in a GPS-denied situation like an ice-covered environment.
 
-% Long baseline navigation
-Though there many acoustic positioning methods, the long baseline (LBL) is closest in style
+% introduce LBL and precise statement of work
+Of acoustic baseline positioning systems, long baseline (LBL) is the most GPS-like in style and scale, and most appropriate for mitigating drift in underwater vehicle localization without overburdening computation or payload size on the vehicle \citep{paull_auv_2014,van_uffelen_global_2021}.
+The state-of-the-art for LBL assumes an isovelocity, linear scaling between one way travel time (OWTT) and range \citep{eustice_recent_2006,eustice_experimental_2007,webster_preliminary_2009,webster_advances_2012}.
+This assumption is valid for short scale operations but introduces error in complex acoustic propagation environments.
+Thus, this paper demonstrates embedded acoustic model-aided data processing to convert recorded OWTTs into pseudoranges in real-time.
+This approach was necessitated by a low risk tolerance for navigation error, given total under-ice conditions and double ducted acoustic environment, in the Beaufort Sea, in March 2020, during the Ice Exercise 2020 (ICEX20).
+We quantify the success of the proposed acoustic ranging method on GPS-linked beacon-to-beacon communication events, not the AUV-LBL positioning itself, as the latter has no ground truth GPS to compare to.
 
-% A recent study tracking 22 floats under the ice in the Weddell Gyre near Antarctica showed that they were unable to surface for 8 months and position uncertainty grew to 116 $\pm$ 148 km in the lateral directions \citep{chamberlain_observing_2018}.
-
-Much progress has been made in underwater vehicle navigation, which relies on any combination of dead reckoning, hydrodynamic models, inertial navigation systems, doppler velocity logs, and/or acoustic baseline navigation systems \citep{paull_auv_2014}.
-In particular, the long baseline (LBL) navigation system is the most similar in scale and style to GPS, and most appropriate for mitigating drift in underwater vehicle localization without overburdening computation on the vehicle \citep{van_uffelen_global_2021}.
+A
 
@@
 
-Underwater acoustic navigation (and GPS) is predicated on timing and positioning.
-Timing is the ability to maintain accurate and precise time, sometimes synchronized across multiple nodes.
-\llabel{1.11} The advent of the WHOI micro-modem \citep{singh_underwater_2006} and synchronized chip scale atomic clocks \citep{gardner_second_2016} enabled ranging via one-way travel time (OWTT).
+In the interest of clarity, we formally reiterate definitions of timing, positioning, and navigation (CITE).
+Timing is the ability to maintain accurate and precise time synchronized across multiple nodes.
+Positioning is the ability to accurately determine one's location referenced to a standard geodetic system. It can be done via trilateration or triangulation.
+Navigation, implying a degree of autonomy, is the ability to distinguish between current and desired position, and apply necessary corrections to course, orientation, and/or speed to attain that desired position.
+Thus, navigation is also intrinsically real-time and is predicated on positioning, which is predicated on timing.
 
-Positioning is the ability to accurately determine one's location and is generally done via trilateration.
-Navigation assumes a degree of autonomy to move from a current position to a desired one.
-The standard for real-time navigation efforts assume a locally homogeneous sound speed profile (SSP) that enables a linear, deterministic scaling between travel time and range \citep{eustice_recent_2006,eustice_experimental_2007,webster_preliminary_2009,webster_advances_2012}.
-Short range (less than ten kilometers) navigation efforts achieve minimal navigation error with a nominal sound speed value \citep{eustice_experimental_2007,webster_preliminary_2009,kepper_mems_2017}.
-For mesoscale (tens to hundreds of kilometers) navigation efforts, a deterministic \citep{graupe_preliminary_2019} or through-the-sensor \citep{webster_towards_2015} value for sound speed is used, but error remains in the hundreds of meters.
+Given the computational and physical barriers of assessing underwater position in real-time, we define post-processing counterparts\textemdash re-navigation and re-positioning\textemdash which may be supplemented by additional data, model output, and computationally or labor intensive algorithms.
 
-Post-processing localization efforts (re-navigation for a self-determinedly moving object or re-positioning for a passively moving one), without computational or time constraints, can indulge more complex acoustic methods and/or ocean representations.
-There are many different approaches, from considering a temporally averaged mean depth dependent SSP \citep{van_uffelen_localization_2016}, including spatio-temporal variability in the sound speed profile to minimize error \citep{graupe_preliminary_2019, mikhalevsky_deep_2020}, comparing acoustic records with synthetic acoustic records computed through an ocean model \citep{wu_deep_2019}.
-The ``cold start'' algorithm developed by \citet{mikhalevsky_deep_2020} isolates a specific ray path to find a representative group velocity and re-positioned a floating hydrophone array with an error of 85 m and a standard deviation of 32 m relative to ground truth GPS data, but implementing this algorithm for navigation remains to be seen.
-These post-processing efforts have generally been deemed too computationally intensive and/or labor intensive for real-time efforts, where automation of identifying time arrivals is further complicated by multipath.
+\subsection{Background on underwater navigation, re-navigation, and re-positioning}
+
+Underwater navigation research is broadly motivated by acquiring GPS-like navigation accuracy (on the order of meters) in GPS-denied conditions.
+
+Navigation efforts are currently limited to small-scale deployments, and take advantage of 
+\llabel{1.11} the advent of the WHOI micro-modem \citep{singh_underwater_2006} and synchronized chip scale atomic clocks \citep{gardner_second_2016}.
+Short range (less than ten kilometers) navigation efforts achieve minimal error with a nominal sound speed value \citep{eustice_experimental_2007,webster_preliminary_2009,kepper_mems_2017}.
+However, these efforts were all conducted with X sound speed conditions...
+
+Of particular note are previous efforts in the Arctic's Beaufort Lens.
+At larger scales on the order of hundreds of kilometers, a deterministic \citep{graupe_preliminary_2019} or through-the-sensor \citep{webster_towards_2015} value for sound speed was used, but error remained in the hundreds of meters.
+
+The LBL system provides navigation aiding as position updates to correct for localization drift and error.
+
+It is important to note that post-processing localization efforts (re-navigation for a self-determinedly moving object or re-positioning for a passively moving one), are often applied for large-scale deployments to better account for spatial variability in the ocean state.
+
+Recently, buoyancy gliders have been re-navigated by considering a temporally averaged depth dependent SSP and CTD observations of...
+
+A well known example of long range re-positioning are RAFOS drifters that record the timestamps of the four strongest pings; transmit all data after mission completion; and are manually localized with uncertainties of 1.5 to 4.5 km (CITE ROSSBY, DUDA).
+
+Other re-positioning methods employ four-dimensional ocean models.
+\citet{wu_deep_2019} compares acoustic records with synthetic acoustic records computed through HYCOM \cite{chassignet_hycom_2007}.
+\citet{mikhalevsky_deep_2020} introduces an impressive ``cold start'' algorithm that isolates a specific ray path to find a representative group speed, re-positioning a floating hydrophone array with an error of 85 m and a standard deviation of 32 m relative to ground truth GPS data.
+Implementing these algorithms for navigation remains to be seen.
 
@@
 
-Underwater navigation research is broadly motivated by GPS-like navigation in GPS-denied conditions.
-The Arctic, while remote, is the perfect place to test mature navigation technologies in real GPS-denied conditions.
-
+U
+We refer the reader to excellent reviews of vehicle operations in the Arctic by \citet{norgren_unmanned_2014} and \citet{barker_scientific_2020}.
+Previous AUV deployments in the Arctic \cite{bellingham_auv_1993}
 Literature and historical reviews of underwater navigation approaches in under-ice conditions highlight the emerging technology learning curve for vehicle operations in the Arctic, the hostile conditions brought on by a semi-permanent moving ice cover, and the need for both absolute and relative navigation with little to no tolerance for error \citep{norgren_unmanned_2014,mcfarlane_autonomous_2015,barker_scientific_2020}.
@@
 \subsection{Ice Exercise 2020}
+
+During ICEX20, we deployed AUV \emph{Macrura}, a custom Bluefin-21, in complete under-ice conditions and a complex double ducted environment brought upon by a persistent warm water intrusion in the Pacific Summer Water layer, which we call the Beaufort Lens.
+The majority of the navigation updates were below 11 m and were small enough to enable an emergency recovery. 
+When considering operational variables such as range, vehicle class, and propagation conditions, there are no comparable deployments.
 
@@
 
+Underwater navigation research is broadly motivated by acquiring GPS-like navigation accuracy in GPS-denied conditions.
+The Arctic, while remote, is the perfect place to test mature navigation technologies in real GPS-denied conditions.
+
 Range estimation is an essential step of acoustic localization and navigation.

### 2021-12-22 manuscript/main.tex
new structure for intro?

@@
 Of acoustic baseline positioning systems, long baseline (LBL) is the most GPS-like in style and scale, and most appropriate for mitigating drift in underwater vehicle localization without overburdening computation or payload size on the vehicle \citep{paull_auv_2014,van_uffelen_global_2021}.
-The state-of-the-art for LBL assumes an isovelocity, linear scaling between one way travel time (OWTT) and range \citep{eustice_recent_2006,eustice_experimental_2007,webster_preliminary_2009,webster_advances_2012}.
-This assumption is valid for short scale operations but introduces error in complex acoustic propagation environments.
+The state-of-the-art for LBL outsources depth to a pressure sensor and assumes an isovelocity, linear scaling between one way travel time (OWTT) and range \citep{eustice_recent_2006,eustice_experimental_2007,webster_preliminary_2009,webster_advances_2012}.
+This assumption is valid for small and short scale operations but introduces error in complex acoustic propagation environments.
 Thus, this paper demonstrates embedded acoustic model-aided data processing to convert recorded OWTTs into pseudoranges in real-time.
@@
 
-A
-
-\subsection{Timing, Positioning, and Navigation}
-
 In the interest of clarity, we formally reiterate definitions of timing, positioning, and navigation (CITE).
 Timing is the ability to maintain accurate and precise time synchronized across multiple nodes.
-Positioning is the ability to accurately determine one's location referenced to a standard geodetic system. It can be done via trilateration or triangulation.
-Navigation, implying a degree of autonomy, is the ability to distinguish between current and desired position, and apply necessary corrections to course, orientation, and/or speed to attain that desired position.
+Positioning is the ability to accurately determine one's location referenced to a standard geodetic system.
+It can be done via trilateration or triangulation.
+Navigation, implying a degree of autonomy, is the ability to distinguish between current and desired position, and apply necessary corrections to course, orientation, and/or speed to attain a future desired position.
 Thus, navigation is also intrinsically real-time and is predicated on positioning, which is predicated on timing.
 
-Given the computational and physical barriers of assessing underwater position in real-time, we define post-processing counterparts\textemdash re-navigation and re-positioning\textemdash which may be supplemented by additional data, model output, and computationally or labor intensive algorithms.
+Given the computational and physical barriers of assessing underwater position in real-time, we define post-processing counterparts\textemdash re-navigation and re-positioning\textemdash which may also be supplemented by additional data, model output, and bespoke algorithms.
 
-\subsection{Background on underwater navigation, re-navigation, and re-positioning}
+\subsection{Long baseline navigation}
+LBL navigation was first introduced ...
+
+significantly revamped by ...
+
+key assumptions are ...
+
+papers by x,y,z show this much error. generally reserved for small and short scale deployments.
+
+\subsection{Acoustic methods for re-navigation and re-positioning}
+
+The LBL navigation methodology simplifies the acoustic propagation to an isovelocity ocean.
+Post-processing methods can take advantage of additional information 
+recording acoustic information but not using it onboard.
+
+It is important to note that post-processing localization efforts (re-navigation for a self-determinedly moving object or re-positioning for a passively moving one), are often applied for large-scale deployments to better account for spatial variability in the ocean state.
+
+Recently, buoyancy gliders have been re-navigated by considering a temporally averaged depth dependent SSP and CTD observations of...
+
+A well known example of long range re-positioning are RAFOS drifters that record the timestamps of the four strongest pings; transmit all data after mission completion; and are manually localized with uncertainties of 1.5 to 4.5 km (CITE ROSSBY, DUDA).
+
+Other re-positioning methods employ four-dimensional ocean models.
+\citet{wu_deep_2019} compares acoustic records with synthetic acoustic records computed through HYCOM \cite{chassignet_hycom_2007}.
+\citet{mikhalevsky_deep_2020} introduces an impressive ``cold start'' algorithm that isolates a specific ray path to find a representative group speed, re-positioning a floating hydrophone array with an error of 85 m and a standard deviation of 32 m relative to ground truth GPS data.
+Implementing these algorithms for navigation remains to be seen.
+
+\subsection{Underwater navigation in the Arctic}
+
+\subsection{The New Arctic}
+
+Of particular note are previous AUV deployments in the Arctic that used an LBL navigation system.
+A more comprehensive review of AUV deployments with a myriad of applications can be found in xx, yy.
+
+Because model-based acoustic arrival methods have generally been deemed too computationally and/or labor intensive for real-time,
 
@@
 
-It is important to note that post-processing localization efforts (re-navigation for a self-determinedly moving object or re-positioning for a passively moving one), are often applied for large-scale deployments to better account for spatial variability in the ocean state.
-
-Recently, buoyancy gliders have been re-navigated by considering a temporally averaged depth dependent SSP and CTD observations of...
-
-A well known example of long range re-positioning are RAFOS drifters that record the timestamps of the four strongest pings; transmit all data after mission completion; and are manually localized with uncertainties of 1.5 to 4.5 km (CITE ROSSBY, DUDA).
-
-Other re-positioning methods employ four-dimensional ocean models.
-\citet{wu_deep_2019} compares acoustic records with synthetic acoustic records computed through HYCOM \cite{chassignet_hycom_2007}.
-\citet{mikhalevsky_deep_2020} introduces an impressive ``cold start'' algorithm that isolates a specific ray path to find a representative group speed, re-positioning a floating hydrophone array with an error of 85 m and a standard deviation of 32 m relative to ground truth GPS data.
-Implementing these algorithms for navigation remains to be seen.
-
 \subsection{Underwater positioning and navigation in the Arctic}
 
-U
 We refer the reader to excellent reviews of vehicle operations in the Arctic by \citet{norgren_unmanned_2014} and \citet{barker_scientific_2020}.
@@
 
+% The results presented herein are from post- processed analysis, however they use only the data that was available in real-time—raw navigation data and only the acoustic transmissions that were successfully decoded by the AUG in real time.
+% The reasons for the difference in performance are most likely due to propagation conditions in and out of the duct.
+Receivers in the duct hear the source at any range, while those outside the duct are in a range-dependent environment where signals come and go depending on the ray paths.
+
 Fig. \ref{fig:rangeError} shows the directional range error footprints for all three sound speed inputs with respect to OWTT, separated by source and receiver depth configurations.

### 2022-02-13 rebuttal/rebuttal-letter/trunk-rebuttal.tex
new structure for intro?
read thru and initial responses

@@
 We have refashioned the abstract and introduction to better elucidate the motivation for this work as a model-aided component for LBL navigation in total under-ice conditions, which could be further generalized for any acoustic environment.
-Based on several suggestions, we have restructured the manuscript and separated the real-time navigation methodology, the real-time positioning results (Minimum Bounce Criteria), and the post-processing re-positioning improvement (Nearest Bounce Criteria).
-Lastly, when evaluating our methodology, we have removed qualitative language like ``GPS-like'' or ``rivaling GPS'' in favor of quantitative descriptions of ranging error.
+Based on several suggestions, we have restructured the manuscript and separated the real-time navigation methodology (used the Minimum Bounce Criteria, MBC, but is impossible to compare given the lack of GNSS), the real-time positioning results (MBC), and the post-processing re-positioning improvement (Nearest Bounce Criteria).
 
-We also more prominently establish definitions for navigation, positioning, and re-positioning to better contextualize our results and contributions, based on this discussion in Howe et al (2019) \emph{Observing the Oceans Acoustically} (and many others):
+For clarity, we also more prominently establish definitions for navigation, positioning, and re-positioning to better contextualize our results and contributions, based on this discussion in Howe et al (2019) \emph{Observing the Oceans Acoustically} (and many others):
 
@@
 
-Given the inescapability of not having GNSS ground truth for our AUV navigation data, and the tension of evaluating our travel time to range conversion on events between LBL beacons in real time and post-processing, we further delineate that positioning is in real time and re-positioning is post-processed; similarly, we interpret reviewer feedback using these definitions.
+Given the tension of evaluating our travel time to range conversion on events between LBL beacons in real time and post-processing, we further delineate that, in this work, positioning is in real time and re-positioning is post-processed.
+We have also added re-positioning trilateration results, processed with only the available field data and algorithms that could be performed in real time, to emulate navigation performance.
 
 The rebuttal letter is organized with sections delineated by the review document and email.
-We look forward to your comments and are happy to discuss any aspect of this rebuttal letter or the manuscript further.
+We look forward to your comments and the opportunity to continue improving this manuscript for eventual publication in JASA.
 Thank you again for your time and consideration.
 
-Please note that the corresponding author is on parental leave for the month of January.
+We also thank JASA for a generous extension given that the corresponding author started (and is still on) parental leave.
 
@@
 However, we note that the system was successfully deployed on an underwater vehicle during ICEX20.
-The results from this work are specifically limited to the methodology behind calculating a single group velocity, which we attempt to quantify through beacon-to-beacon communications.
+The results from this work are specifically limited to the methodology behind calculating a single group velocity, which we attempt to quantify through beacon-to-beacon communications with known GNSS timestamps.
 
@@
 
-	We have tempered the claims in the abstract to reflect the limitations of the dataset collected and the transferability of the analysis in this paper.
-	We have amended Figure \ref{v1:fig:vehicleRecovery} to include vehicle data and illustrate that this approach was successful in a real field environment.
-	We have also included a discussion of how these results might compare for various depths and ranges during vehicle operations (see \hl{XXX}).
+	{\color{blue}
 
-	With respect to the concentrating on real-time results versus post-processing results, we would point out that the AUV was deployed through a hydrohole from Topside and due to a disk error, stalled underneath the ice surface while transmitting its perceived location.
-	The AUV was found within a meter from its reported location, which serves as strong (but qualitative) evidence of the navigation performance provided by the single group velocity calculation.
-	Work currently review in the Journal of Field Robotics details the other components to vehicle navigation (DVL-aided INS and a vehicle flight model) along with the magnitude of position corrections.
+	We have both modified the data presented and tempered the claims in the abstract and throughout the manuscript.
+The reviewer brings up the important point of transferability: how applicable is the algorithm for normal vehicle operations given its performance on a handful of fixed ranges?
 
-	Given that our vehicle navigation data only has one GPS point to compare to, we use the beacon-to-beacon communications as a natural experiment to compare the effectiveness of the real-time NBC method.
-	The real-time section is small because only one modem at a time ran the vehicle behavior to calculate these acoustic pseudoranges.
-	A robust comparison is difficult given the mismatch of experimental conditions in estimated sound speed environment as well as source and receiver position.
+	% We have tempered the claims in the abstract to reflect the limitations of the dataset collected and the transferability of the analysis in this paper.
+	% We have amended Figure \ref{v1:fig:vehicleRecovery} to include vehicle data and illustrate that this approach was successful in a real field environment.
+	% We have also included a discussion of how these results might compare for various depths and ranges during vehicle operations (see \hl{XXX}).
 
-	We extended the dataset to all possible communication events through the same automated pipeline to decouple the effects of the sound speed environment from source and receiver position.
+	% With respect to the concentrating on real-time results versus post-processing results, we would point out that the AUV was deployed through a hydrohole from Topside and due to a disk error, stalled underneath the ice surface while transmitting its perceived location.
+	% The AUV was found within a meter from its reported location, which serves as strong (but qualitative) evidence of the navigation performance provided by the single group velocity calculation.
+	% Work currently review in the Journal of Field Robotics details the other components to vehicle navigation (DVL-aided INS and a vehicle flight model) along with the magnitude of position corrections.
+
+	% Given that our vehicle navigation data only has one GPS point to compare to, we use the beacon-to-beacon communications as a natural experiment to compare the effectiveness of the real-time NBC method.
+	% The real-time section is small because only one modem at a time ran the vehicle behavior to calculate these acoustic pseudoranges.
+	% A robust comparison is difficult given, as the reviewer points out, the handful of fixed ranges.
+
+	% We extended the dataset to all possible communication events through the same automated pipeline to decouple the effects of the sound speed environment from source and receiver position.
 This is in effect no different than running the computation during the field; the real-time NBC data is a subset of the post-processed NBC data.
 
-	While post-processing, we began to see a pattern of range overestimation using the NBC method.
+	% While post-processing, we began to see a pattern of range overestimation using the NBC method.
 Thus, we developed the Minimal Bounce Criteria instead and ran it through the same pipeline, subject to the same computational constraints.
 
-	By comparing the errors between these two pseudorange estimation methods, we noticed a surprising underlying pattern compared to the ground truth GPS data.
-	Further investigation suggests there is noise in the GPS measurement that is not incurred in the travel time measurement that informs the acoustic range estimation.
+	% By comparing the errors between these two pseudorange estimation methods, we noticed a surprising underlying pattern compared to the ground truth GPS data.
+	% Further investigation suggests there is noise in the GPS measurement that is not incurred in the travel time measurement that informs the acoustic range estimation.
 
-	The manuscript has been significantly restructured to separate analyses along this narrative.
+	% The manuscript has been significantly restructured to separate analyses along this narrative.
+	}
 
@@
 
-	We have re-written the introduction and conclusion to clarify that our work adds a necessary model-aided component for ranging in LBL networks for complex acoustic environments, and that it could be extended for long range propagation as well.
+	We believe we have tightened up the introduction and conclusion sections by focusing on this work as a model-aided augmentation to an LBL system deployed in under-ice conditions.
+	This focus presents the narrative more linearly, instead of it feeling like it jumps between different types of deployments.
 
@@
 
-	Lastly, we have removed the line in question, but would point out it was originally written as ``navigation'', not ``positioning''.
+	Lastly, we have removed the line in question, and would point out it was originally written as ``navigation'', not ``positioning''.
 
@@
 Lines 235 and following state “the variation of horizontal group velocity from any source-receiver pair is the fundamental challenge to implement GPS-like navigation in an LBL navigation paradigm…” this statement could use some clarification.
-Also, perhaps change the axis on Figure 5 to read something like ‘implied group velocity.”}
+Also, perhaps change the axis on Figure 5 to read something like ‘implied group velocity.”} \label{1dot5}
 
+	Thank you, these are all excellent suggestions.
 ``It is the variation of group speed from source to source in underwater navigation that is the fundamental challenge to implement a “GNSS-like” navigation system in the ocean with acoustics.''
@@
 	\lreviewer{line 222}{v1:1.14}{Could you provide more information on the baseline and weighted SSPs. Where did the baseline profile come from?}
-	\label{1.14}
+	\label{1dot14}
 
@@
 
-	\lreviewer{line 7}{v1:2.1}{'Beaufort Lens' This is not yet a scientific term, suggest finding descriptive terms to use. What comprises the 'lens'? The Pacific Summer Water is the warm layer.}
+	\lreviewer{line 7}{v1:2.1}{'Beaufort Lens' This is not yet a scientific term, suggest finding descriptive terms to use.
+What comprises the 'lens'?
+The Pacific Summer Water is the warm layer.}\label{2dot1}
+
+	The Beaufort Lens has been used in the literature (Litvak 2015, Schmidt and Schneider 2016) and in JASA/POMA (Poulsen et al 2016, Chen et al 2019).
+
+	We think of the lens as any feature that introduces significant refraction, such that the term Beaufort Lens is a convenient shorthand for the spatio-temporal variability of the warm water intrusion provided by the Pacific Summer Water Layer that creates two ducts.
+We prefer this distinction as the lens exists in the oceanographic space but the ducts are acoustic features dependent on source depth and frequency.
+With this jargon, the lens is susceptible to aberrations in the same way that the duct is susceptible to leaking.
+
+	For clarity, we explictly define Beaufort Lens in the manuscript as explained HERE LINK.
 
@@
 	\lreviewer{line 12}{v1:2.3}{'rival GPS' is a fairly bold statement when GPS is used as ground truth.}
-	{\color{blue} Accuracy comparable to conventional GPS; does not introduce decrease precision than is already documented by GPS puck precision; Effectively extends single GPS into the water column with comparable performance.
-Given more beacons, there is potential to reduce uncertainty in the same way that GPS / differential GPS works}
+
+	We meant ``rival GPS'' in that the downstream accuracy of the acoustic range estimation method is within the precision of GPS.
+However, we have amended this framing to say that our best performing acoustic method does not introduce significant additional error than is already documented by GPS puck precision, effectively extending single GPS performance into the shallow water column for a small volume of interest.
+Given more beacons, there is potential to reduce uncertainty in the same way that differential GPS works.
 
@@
 
+	Please see comment \ref{2dot1}.
+
 	\lreviewer{line 94}{v1:2.5}{Ice Tethered profiler data shows fairly slow variability in space and time in the Beaufort, but changes in characteristics occur as the deep basin transitions to the Chukchi, so what is meant by 'high degree' exactly?}
 
-	\lreviewer{lig 4}{}{Fig 4 - where did the baseline come from (exactly)? What database, etc.}
-	Please see Comment \ref{1.14}.
+	ITP data and modeling efforts by Duda et al (2019, 2021) suggest the duct is persistent and widespread but not always continuous in the Beaufort.
+Of course, it breaks down towards the shallower Chukchi basin.
+Thus ``high degree'' refers to variance of the observed ducted conditions, i.e., the duct and its acoustic effects can be non-existent, minimal, or drastic.
+	This is easily seen by parsing an integral duct parameter, $\kappa$, that measures the area spanning the local sound speed maximum from the lens to its conjugate depth, for all relevant ITP data.
+
+	\begin{figure}[!h]
+	\centering
+	\includegraphics[width=\textwidth]{ssp-ducts-lines}
+	\caption*{Adapted from Chapter 3 of Bhatt, E. (2021) An empirical Virtual Ocean framework for environmentally adaptive, embedded acoustic navigation for autonomous underwater vehicles, PhD Thesis.}
+	\end{figure}
+	\clearpage
+
+	\lreviewer{line 4}{}{Fig 4 - where did the baseline come from (exactly)? What database, etc.}
+	Please see Comment \ref{1dot14}.
 
@@
 
+	We have removed this statement, as it is not relevant to the findings of this work.
+However, the navigation solution existed in tandem with an interactive tactical decision aid (TDA) operated by a Navy submarine officer.
+A lightweight compression of the sound speed profile was needed to facilitate its update via digital acoustic message.
+Work about this TDA has been recently accepted for publication in IEEE JOE.
+
 	\lreviewer{line 231}{v1:2.8}{'abundance'. Is there a more precise way to say this? Bellhop does what it is programmed to…}
+
+	{\color{blue} Yes. BELLHOP finds all possible eigenrays given the prescribed launch angles whereas only a few or less (sometimes none) are valid for the travel time of interest}.
 
@@
 
+	{\color{blue} The authors are not sure what the reviewer means by this statement, specifically what ``navy range'' is.
+We believe we are only using ``range'' to mean distance.
+Nonetheless, we have adopted this suggestion and use ``navigation'' or ``communication'' system/network instead.}
+
 	\lreviewer{line 261}{v1:2.11}{'bring'. Do you mean interpolate?}
@@
 	\lreviewer{line 281}{v1:2.12}{Awkward sentence, either unclear or extremely obvious. "...an acoustic arrival does not always take the direct path from source to the receiver,".}
+
+	{\color{blue} Perhaps both. We have changed this accordingly}
 
@@
 
+	{\color{blue} We mean prominence, often used in peak finding, which would indicate both the considerable sound speed differential and the nearly horizontal slope to the local maxima.}
+
 	\lreviewer{line 410}{}{'anomalous' - does this section need to be included? Or just the issue summarized and noted? Delete the table?}
-	{\color{blue}Delete table, summarize issue.}
+
+	{\color{blue} Thank you for this suggestion. We have deleted the table and instead summarized the issue.}
 
 	\lreviewer{lines 585-591}{}{Suggest deleting, not very appropriate for an academic paper.}
+
+	This has been removed.
 
@@
 
-	Thank you. We politely
+	Thank you.
+The key point of this paper is a model-aided component to the LBL paradigm that we demonstrated in an Arctic environment.
+Because both reviewers and the editor felt the key points were comparing the two different approaches to determine an effective sound speed for converting travel times to ranges, we have significantly restructured the paper to make the narrative clearer.
+
+	The discussion of AUV operations and jargon from AUV application contextualizes the model-aided component to the LBL paradigm.
+However, we agree with the Editor that this emphasis can be distracting when used throughout the manuscript.
+We have attempted to use the most direct language possible in each section.
+We hope this revision presents the contributions more clearly and welcome further feedback on this matter.
 
@@
 The manuscript needs to include some discussion of these different cases.}
+
+	We have now included the information on acoustic parameters in the experiment description, see section REFREF.
+We apologize that this information was ommitted as the manuscript was adapted from the thesis.
+
+	Similarly, we have addressed the editor's point about the ray geometries in the discussion section, REFREF, but have opted to keep those figures in the supplementary material.
 
@@
 
+	We agree that group velocity is not the correct term. We have adopted the language ``effective horizontal sound speed'' instead.
+
 	\reviewer{}
 	{Why is the effective sound-speed determined from the GPS-determined range and the acoustic travel time referred to as “… the naïve group velocity …”? What is naïve about it?}
+
+	We have recast this as ``implied effective horizontal sound speed'', as per comment \ref{1dot5}.
 
@@
 
+	{\color{blue} Thank you for catching this. We have corrected it and made sure all figurs are referenced in the text directly.}
+
 	\reviewer{}
-	{I agree with both reviewers that the Section IV.E, GPS sensor drift at polar latitudes, is out of place. I find the term “GPS drift” itself odd. It seems that the authors really mean GPS accuracy.}
-	{\color{blue}GPS precision}
+	{I agree with both reviewers that the Section IV.E, GPS sensor drift at polar latitudes, is out of place.
+I find the term “GPS drift” itself odd.
+It seems that the authors really mean GPS accuracy.} \label{3dot7}
+	{\color{blue} We have talked with colleagues and think ``GPS noise'' is the best term.
+But given that all reviewers and the editor feel this section is out of place, we have significantly reduced it...
+OR REMOVE IT?}
 
@@
 GPS (really GNSS) applications come in different flavors, including stand-alone single frequency, stand-alone dual frequency, differential, real-time kinematic, etc., with varying accuracies.}
+	When evaluating our methodology, we have removed qualitative language like ``GPS-like'' or ``rivaling GPS'' in favor of quantitative descriptions of ranging error.
 {\color{blue}achieves ``x meter'' level accuracy compared to ``y meter'' level accuracy reported by the manufacturer}
@@
 
+	We have removed this generalization.
+
 	\reviewer{}
 	{Several of the references are incomplete, i.e., Norgren et al. (2014), Poulsen et al. (2016), Schmidt and Schneider (2016), and Schneider et al. (2020).}
+
+	{\color{blue}Thank you for pointing this out. We have double checked all references.}
 
@@
 	Reviewer 2: The work is interesting and describes a data set collected from the Arctic, which is of high value. However, some of the writing is a bit awkward and occasionally vague or imprecise.}
+
+	Thank you for the pointed feedback. We hope you find this revision clearer and more precise.
 
@@
 
+	Thank you.
+We hope this manuscript would be of relevance to the Arctic Acoustics special issue given two topics of interest: under-ice navigation systems for floats, gliders, and autonomous underwater vehicles, and comparisons of current Arctic acoustic conditions with historical data.
+
 	\reviewer{}{Is the manuscript a clear, concise, reasonably self-contained presentation of the material, giving adequate references to related work?
@@
 	Reviewer 2: Overall the language is fine, details below.}
+
+	Thank you. We have significantly reorganized the paper.
 
@@
 
+	Thank you.
+We have opted to keep the title the same, as we believe the results are a strong proxy for under-ice acoustic navigation, which we demonstrated during ICEX20 but there are inherently no data products to validate this waypoint by waypoint.
+
+	We have refashioned and tempered the abstract and introduction accordingly.
+
 \end{document}

### 2022-02-13 pipeline/compare-insitu-post/gvel_4_compare_inSitu_postProcessed.m
pipeline folders linking correctly

@@ -7,11 +7,11 @@ clear; clc; close all;
 %% load in situ data
 % DATA = readtable('./bellhop-gvel-gridded-simRange/gveltable.csv');
 
-DATA = readtable('../pipeline/bellhop-gvel-gridded-simRange/gveltable.csv');
+DATA = readtable('../bellhop-gvel-gridded-simRange/gveltable.csv');
 indValid1 = ~isnan(DATA.recRange) | ~isnan(DATA.simGvel);
 DATA = DATA(indValid1,:);
 
-A = load('../data/tobytest-recap-clean.mat'); % loads "event"
+A = load('../../data/tobytest-recap-clean.mat'); % loads "event"
 RECAP = h_unpack_experiment(A.event);
 indValid2 = ~isnan(RECAP.sim_gvel) | ~isnan(RECAP.sim_range);
 
@@ -34,7 +34,7 @@ EOF_BOOL(indBad) = NaN;
 DATA.rangeAnomaly = DATA.owtt .* DATA.simGvel - GPS_RANGE;
 
 %% range error for "sim" aka post processing nearest bounce criterion
-listing = dir('../pipeline/bellhop-gvel-gridded-simRange/csv_arr/*gridded.csv');
+listing = dir('../bellhop-gvel-gridded-simRange/csv_arr/*gridded.csv');
 
 for f = 1:numel(listing)
     T0 = readtable([listing(f).folder '/' listing(f).name]);

### 2022-02-13 pipeline/compare-mbc-nbc/gvel_5_compare_postprocessed.m

@@ -5,8 +5,8 @@
 clear; clc; close all;
 
 %% load in situ data
-DATA = readtable('../pipeline/bellhop-gvel-gridded/gveltable.csv');
-A = load('../data/tobytest-recap-clean.mat'); % loads "event"
+DATA = readtable('../bellhop-gvel-gridded/gveltable.csv');
+A = load('../../data/tobytest-recap-clean.mat'); % loads "event"
 RECAP = h_unpack_experiment(A.event);
 
 DATA.simGvel(isnan(DATA.simGvel)) = 0;
@@ -28,7 +28,7 @@ indValid = ~isnan(DATA.simGvel);
 DATA.rangeAnomaly = DATA.owtt .* DATA.simGvel - DATA.recRange;
 
 %% load post-processing, new algorithm
-listing = dir('../pipeline/bellhop-gvel-gridded/csv_arr/*gridded.csv');
+listing = dir('../bellhop-gvel-gridded/csv_arr/*gridded.csv');
 
 for f = 1:numel(listing)
     T0 = readtable([listing(f).folder '/' listing(f).name]);
@@ -51,7 +51,7 @@ for f = 1:numel(listing)
 end
 
 %% load post-processing, old algorithm
-listing = dir('../pipeline/bellhop-gvel-gridded/csv_arr/*old.csv');
+listing = dir('../bellhop-gvel-gridded/csv_arr/*old.csv');
 
 for f = 1:numel(listing)
     T0 = readtable([listing(f).folder '/' listing(f).name]);

### 2022-02-13 pipeline/data-prep/main_collate_tobytest_recap.m

@@ -111,6 +111,9 @@ for k = 1:num_events
         
         bad_events(end+1) = k;
     end
+    
+    % event that is 11 seconds?
+    if owtt > 4
 end
 
 event(bad_events) = [];

### 2022-02-13 pipeline/overview-info/main_plot_experiment_overview.m

@@ -1,11 +1,13 @@
-%% main_plot_ice_drift.m
-% plots ice drift from ../data/tobytest-recap-clean.mat
+%% main_plot_experiment_overview
 
 %% prep workspace
 clear; clc; close all;
 
 lg_font_size = 14;
 alpha_color  = .035;
+
+% add src path
+addpath('../../src/');
 
 % easily distinguishable colors to link modems
 load p_modemMarkerDetails
@@ -15,120 +17,131 @@ markerSize = 145;
 load p_legendDetails
 
 %% load toby test data recap all
-load '../data/tobytest-recap-clean.mat'
-RECAP = h_unpack_experiment(event);
+A = load('../../data/tobytest-recap-clean.mat'); % loads "event"
+RECAP = h_unpack_experiment(A.event);
 
 eof_bool = RECAP.eof_bool;
 eof_time = RECAP.data_time;
 
 %% figure 1 : txrx timeline chart
 
-figure('Name','txrx-timeline','Renderer', 'painters', 'Position', [0 0 1280 720]); clf;
-
-% mapping for depth
-plotval = -[27 18 9];
-plotZMap = containers.Map([90 30 20],plotval);
-
-for rmd = modem_rx_depth
-    
-    % rx nodes
-    index = RECAP.rx_z == rmd;
-    rxnodes = RECAP.tag_rx(index);
-    uniqueRxNodes = intersect(modem_labels,unique(rxnodes),'stable');
-    plotSpread = linspace(0,numel(uniqueRxNodes),numel(uniqueRxNodes));
-    plotSpread = plotSpread - mean(plotSpread);
-    RX_offsetMap = containers.Map(uniqueRxNodes,plotSpread);
-    
-    % tx nodes
-    index = RECAP.tx_z == rmd;
-    txnodes = RECAP.tag_tx(index);
-    uniqueTxNodes = intersect(modem_labels,unique(txnodes),'stable');
-    plotSpread = linspace(0,numel(uniqueTxNodes),numel(uniqueTxNodes));
-    plotSpread = plotSpread - mean(plotSpread);
-    TX_offsetMap = containers.Map(uniqueTxNodes,plotSpread);
-    
-    for node = modem_labels
-        node = node{1}; % cell array to character
-        
-        % RX DEPTHS
-        if any(strcmp(uniqueRxNodes,node))
-            rx_index = find(strcmp(RECAP.tag_rx,node));
[…]

### 2022-02-14 pipeline/analyze-gvel/slides_gvel_comparison.m
src contains access only functions and files, pipeline is editable

@@ -6,7 +6,7 @@
 clear; clc; close all;
 
 % load data
-DATA = readtable('./bellhop-gvel-gridded/gveltable.csv');
+DATA = readtable('../pipeline/bellhop-gvel-gridded/gveltable.csv');
 % only simGvel
 DATA.gvel = DATA.recRange ./ DATA.owtt;
 
@@ -32,7 +32,7 @@ DATA.rangeAnomaly = DATA.owtt .* DATA.simGvel - DATA.recRange;
 ZS = 90;
 
 %% load post-processing sim for v2
-listing = dir('./bellhop-gvel-gridded/csv_arr/*gridded.csv');
+listing = dir('../pipeline/bellhop-gvel-gridded/csv_arr/*gridded.csv');
 
 for k = 1:numel(listing)
     T0 = readtable([listing(k).folder '/' listing(k).name]);
@@ -56,7 +56,7 @@ for k = 1:numel(listing)
 end
 
 %% load post-processing sim, v1
-listing = dir('./bellhop-gvel-gridded/csv_arr/*old.csv');
+listing = dir('../pipeline/bellhop-gvel-gridded/csv_arr/*old.csv');
 
 for f = 1:numel(listing)
     T0 = readtable([listing(f).folder '/' listing(f).name]);
@@ -213,4 +213,4 @@ titlestr = sprintf('Group velocity predictions for a source depth = %u m',zs);
 sgtitle(titlestr,'fontsize',17,'fontweight','bold')
 
 %%
-h_printThesisPNG(sprintf('SLIDES-gvel-comparison-wdata-%u',ZS));
+% h_printThesisPNG(sprintf('SLIDES-gvel-comparison-wdata-%u',ZS));

### 2022-02-14 pipeline/compare-mbc-nbc/gvel_5_compare_postprocessed.m

@@ -8,7 +8,7 @@ clear; clc; close all;
 DATA = readtable('../bellhop-gvel-gridded/gveltable.csv');
 A = load('../../data/tobytest-recap-clean.mat'); % loads "event"
 RECAP = h_unpack_experiment(A.event);
-
+%% 
 DATA.simGvel(isnan(DATA.simGvel)) = 0;
 
 % remove crazy 11 second event, event that is nominally 1.58* seconds

### 2022-02-14 pipeline/data-prep/main_collate_tobytest_recap.m

@@ -112,8 +112,6 @@ for k = 1:num_events
         bad_events(end+1) = k;
     end
     
-    % event that is 11 seconds?
-    if owtt > 4
 end
 
 event(bad_events) = [];

### 2022-02-14 src/h_unpack_experiment.m

@@ -64,7 +64,7 @@ if mode(toby_test_eof_bool) == mean(toby_test_eof_bool)
     eof_bool = toby_test_eof_bool(1);
     
     % Bradli's chosen weights
-    OBJ_EOF = eb_read_eeof('../data/eeof-itp-fix-2013.nc',true);
+    OBJ_EOF = eb_read_eeof('../../data/eeof-itp-mar2013.nc',true);
     weights = [-6.112 15.368 -1.441 2.219 0.138 -0.322 -1.994].'; % chosen by Bradli for CTD at 2020 Mar 09 1202
     
     OBJ.ssp_estimate = OBJ_EOF.baseval + (OBJ_EOF.eofs * weights).*eof_bool;

### 2022-02-14 src/main_plot_gvel_range.m

@@ -1,88 +1,0 @@
-%% main_plot_gvel_range
-
-%% prep workspace
-clear; clc; close all;
-
-lg_font_size = 14;
-
-% easily distinguishable colors to link modems
-load p_modemMarkerDetails
-
-%% load toby test data recap all
-load '../data/tobytest-recap-clean.mat'
-RECAP = h_unpack_experiment(event);
-
-filter_rxz = unique(RECAP.rx_z);
-filter_node = {'North','South','East','West','Camp'};
-
-figure('Name','range-gvel-all','Renderer', 'painters', 'Position', [0 0 1280 720]); clf;
-
-%% figure
-
-subplot(1,2,1)
-hold on
-for fRNX = filter_node
-    index_rxn = strcmp(RECAP.tag_rx,fRNX{1});
-    
-    for fRXZ = filter_rxz
-        index_rxz = RECAP.rx_z == fRXZ;
-        
-        index = logical(index_rxn .* index_rxz);
-        
-        scatter(RECAP.sim_range(index),RECAP.sim_gvel(index),...
-            markerSize,markerModemMap(fRNX{1}),markerShape(fRXZ),'filled','MarkerFaceAlpha',.1)
-    end
-end
-hold off
-grid on
-h_set_xy_bounds(RECAP.sim_range,RECAP.data_range,RECAP.sim_gvel,RECAP.data_range ./ RECAP.data_owtt);
-title('In situ prediction')
-ylabel('group velocity [m/s]');
-xlabel('range [m]')
-
-subplot(1,2,2)
-hold on
-
-%scatter(RECAP.data_range,RECAP.data_range ./ RECAP.data_owtt,'o')
-
-index_gvel = ~isnan(RECAP.sim_gvel);
-for fRNX = filter_node
-    index_rxn = strcmp(RECAP.tag_rx,fRNX{1});
-    
-    for fRXZ = filter_rxz
-        index_rxz = RECAP.rx_z == fRXZ;
-        
-        index = logical(index_gvel .* index_rxn .* index_rxz);
-        
-        scatter(RECAP.data_range(index),RECAP.data_range(index) ./ RECAP.data_owtt(index),...
-            markerSize,markerModemMap(fRNX{1}),markerShape(fRXZ),'filled','MarkerFaceAlpha',.4)
-    end
-end
-hold off
-grid on
[…]

### 2022-02-14 src/plot_range_gvel_v2.m

@@ -1,200 +1,0 @@
-%% plot_range_gvel_v2.m
-
-clear; clc; close all;
-
-%% load data
-A = load('../data/tobytest-recap-clean.mat'); % loads "event"
-A = h_unpack_experiment(A.event);
-
-% remove crazy 11 second event, event that is nominally 1.58* seconds
-indBad1 = A.data_owtt > 3;
-indBad2 = strcmp(A.tag_rx,'East') & A.data_owtt > 1.55;
-indBad = indBad1 | indBad2;
-
-% 1.587 events, had clock errors + Bellhop can't resolve these
-A.sim_gvel(indBad) = NaN;
-
-colorDepth = containers.Map([20 30 90],{[70 240 240]./256,[0 130 200]./256,[0 0 128]./256});
-sourceDepth = containers.Map([30 90],{'^','v'});
-alphaDepth = containers.Map([20 30 90],[.3 .2 .1]);
-
-%% plot
-
-figure('name','gvel-by-range','renderer','painters','position',[108 108 1100 800]);
-t = tiledlayout(2,6,'TileSpacing','compact','Padding','default');
-
-%% gvel vs range for naive
-nexttile([1 4]);
-
-for zs = [20 30 90]
-    index1 = A.tx_z == zs;
-    
-    for zr = [30 90]
-        index2 = A.rx_z == zr;
-        
-        % in situ values
-        index3 = ~isnan(A.sim_gvel);
-        index = logical(index1.*index2.*index3);
-        
-        if sum(index)>=1
-            % plot
-            hold on
-            
-            plotIndex = find(index == 1);
-            for k = plotIndex
-                txNode = A.tag_tx(k);
-                txNode = txNode{1};
-                scatter(A.data_range(k),A.data_range(k)./A.data_owtt(k),...
-                    200,'filled',sourceDepth(zr),'MarkerFaceColor',colorDepth(zs),...
-                    'handlevisibility','off','MarkerFaceAlpha',alphaDepth(zs));
-            end
-            hold off
-        end
-    end
-end
-
-title('Naive calculation of group velocity');
-h_common_plot();
-
-%% inset
-nexttile([1 2]);
-
-for zs = [20 30 90]
-    index1 = A.tx_z == zs;
-    
-    for zr = [30 90]
-        index2 = A.rx_z == zr;
-        
-        % in situ values
-        index3 = ~isnan(A.sim_gvel);
-        index = logical(index1.*index2.*index3);
-        
-        if sum(index)>=1
[…]

### 2022-02-14 pipeline/3-analyze-gvel/slides_gvel_comparison.m
numbering pipelines for readability

@@ -6,7 +6,7 @@
 clear; clc; close all;
 
 % load data
-DATA = readtable('../pipeline/bellhop-gvel-gridded/gveltable.csv');
+DATA = readtable('../bellhop-gvel-gridded/gveltable.csv');
 % only simGvel
 DATA.gvel = DATA.recRange ./ DATA.owtt;
 
@@ -32,7 +32,7 @@ DATA.rangeAnomaly = DATA.owtt .* DATA.simGvel - DATA.recRange;
 ZS = 90;
 
 %% load post-processing sim for v2
-listing = dir('../pipeline/bellhop-gvel-gridded/csv_arr/*gridded.csv');
+listing = dir('../bellhop-gvel-gridded/csv_arr/*gridded.csv');
 
 for k = 1:numel(listing)
     T0 = readtable([listing(k).folder '/' listing(k).name]);
@@ -56,7 +56,7 @@ for k = 1:numel(listing)
 end
 
 %% load post-processing sim, v1
-listing = dir('../pipeline/bellhop-gvel-gridded/csv_arr/*old.csv');
+listing = dir('../bellhop-gvel-gridded/csv_arr/*old.csv');
 
 for f = 1:numel(listing)
     T0 = readtable([listing(f).folder '/' listing(f).name]);

### 2022-02-14 pipeline/5-analyze-postprocessing/gvel_3_simAllEvents.m

@@ -4,7 +4,7 @@
 clear; clc; close all;
 
 % load data
-DATA = readtable('../pipeline/bellhop-gvel-gridded/gveltable.csv');
+DATA = readtable('../bellhop-gvel-gridded/gveltable.csv');
 % only simGvel
 DATA.gvel = DATA.recRange ./ DATA.owtt;
 
@@ -26,7 +26,7 @@ indValid = ~isnan(DATA.simGvel);
 % calculate RangeAnomaly
 DATA.rangeAnomaly = DATA.owtt .* DATA.simGvel - DATA.recRange;
 %% load simulation
-listing = dir('../pipeline/bellhop-gvel-gridded/csv_arr/*gridded.csv');
+listing = dir('../bellhop-gvel-gridded/csv_arr/*gridded.csv');
 
 for k = 1:numel(listing)
     T0 = readtable([listing(k).folder '/' listing(k).name]);

### 2022-02-14 pipeline/8-analyze-anomaly-event/gvel_6_postprocessing_weird_event.m

@@ -5,8 +5,8 @@
 clear; clc;
 
 %% load in situ data
-DATA = readtable('../pipeline/bellhop-gvel-gridded/gveltable.csv');
-A = load('../data/tobytest-recap-clean.mat'); % loads "event"
+DATA = readtable('../bellhop-gvel-gridded/gveltable.csv');
+A = load('../../data/tobytest-recap-clean.mat'); % loads "event"
 RECAP = h_unpack_experiment(A.event);
 
 DATA.simGvel(isnan(DATA.simGvel)) = 0;
@@ -26,7 +26,7 @@ indValid = 1215:1224;
 DATA.rangeAnomaly = DATA.owtt .* DATA.simGvel - DATA.recRange;
 
 %% load post-processing, new algorithm
-listing = dir('../pipeline/bellhop-gvel-gridded/csv_arr/*gridded.csv');
+listing = dir('../bellhop-gvel-gridded/csv_arr/*gridded.csv');
 
 for f = 1:numel(listing)
     T0 = readtable([listing(f).folder '/' listing(f).name]);
@@ -49,7 +49,7 @@ for f = 1:numel(listing)
 end
 
 %% load post-processing, old algorithm
-listing = dir('../pipeline/bellhop-gvel-gridded/csv_arr/*old.csv');
+listing = dir('../bellhop-gvel-gridded/csv_arr/*old.csv');
 
 for f = 1:numel(listing)
     T0 = readtable([listing(f).folder '/' listing(f).name]);

### 2022-02-15 pipeline/1-overview/main_plot_experiment_overview.m
added percent good

@@ -139,13 +139,13 @@ eof_time = RECAP.data_time;
 countOscar = [44 110 18 82 61 28 137 74 ...
               71 74 121 160 76 ...
               32 10 28 34 ...
-              12 43 45 34 ...
-              91 55 150 155 38];
+              91 55 150 155 38 ...
+              12 43 45 34];
 c0 = 0;
 
 markerSize = 220;
 
-figure('Name','txrx-chart','Renderer', 'painters', 'Position', [0 0 950 650]); clf;
+figure('Name','txrx-chart','Renderer', 'painters', 'Position', [0 0 1200 600]); clf;
 plot(NaN,NaN,'w');
 
 % txrx chart maps
@@ -166,7 +166,8 @@ ylim([y1 y2]);
 yticks(plotval)
 yticklabels([]);
 hold on
-grid on
+ax = gca;
+ax.YGrid = 'on';
 
 patchxval = [xTX-1 xTX+1];
 patchyval = ylim();
@@ -179,8 +180,8 @@ p.EdgeColor = 'none';
 
 % add depth label
 for txDepth = modem_rx_depth
-    text(xTX-1,ZMap(txDepth)-.2,['   z_s = ' num2str(txDepth) ' m'],...
-        'color','w','HorizontalAlignment','left','VerticalAlignment','middle','fontsize',12,'fontweight','bold')
+    text(xTX-0.8,ZMap(txDepth)-.2,['   z_s = ' num2str(txDepth) ' m'],...
+        'color','w','HorizontalAlignment','left','VerticalAlignment','middle','fontsize',13,'fontweight','bold')
 end
 
 % mapping for offset
@@ -190,8 +191,6 @@ xOffsetMap = containers.Map(modem_labels,plotspread);
 x2OffsetMap = containers.Map([20 30 90],[0 0 .3]);
 z2OffsetMap = containers.Map([20 30 90],[-.1 -.1  .2]);
 textOffsetMap = containers.Map(modem_labels,1.1.*[-1 1 -1 1 -1]);
-
-
 
 for txDepth = modem_rx_depth
     
@@ -242,7 +241,12 @@ for txDepth = modem_rx_depth
                     
                     % add text for amount
                     text(xval,yval - z2OffsetMap(rDepth) + textOffsetMap(tNode{1}),num2str(sum(index)),... 
-                        'VerticalAlignment','middle','HorizontalAlignment','center','color',markerModemMap(rNode{1}));
[…]

### 2022-02-15 manuscript/main-old.tex
working through restructuring manuscript

--- 2021-12-22 manuscript/main.tex
@@
 The long baseline (LBL) underwater navigation paradigm relies on the conversion of recorded travel time to range to trilaterate for position.
-For real-time navigation, this conversion assumes an isovelocity sound speed.
+For real-time operations, this conversion assumes an isovelocity sound speed.
 For re-navigation in post-processing, computationally and/or labor intensive acoustic modeling may be employed to reduce uncertainty driven by multipath arrivals.
@@
 A team was able to dig a hole through the 2 meter ice sheet and immediately see the vehicle.
-Its successful recovery amidst total ice cover is shown in Fig. \ref{fig:vehicleRecovery}.
+Its total under ice mission is shown in Fig. \ref{fig:vehicleRecovery}.
 \begin{figure}[h!]
 	\centering
-	\includegraphics[width=\reprintcolumnwidth]{figs/Fig1.jpg}
+	\includegraphics[width=\columnwidth]{figs/auv-track-update.png}
 	\caption{The successful emergency recovery of AUV Macrura, with the the auger used to drill through the ice sheet in the background. Photo courtesy of LT CDR D. Goodwin.}
@@
 	\centering
-	\includegraphics[width=\reprintcolumnwidth]{figs/Fig2.pdf}
+	\includegraphics[width=\textwidth]{figs/Fig2.pdf}
 	\caption{A schematic overview of the Integrated Communication and Navigation Network (ICNN), which provides joint data-transfer and tracking between AUV and a human decision maker at Topside.}
@@
 The one instance this is not true is important to mention\textemdash for a travel time of 2.2 s, where the minimal multipath is dominated by a single bottom bounce (the red line to the furthest transmission in Fig.
-\ref{fig:raytrace-zs30}, more detail in Section \ref{subsec:anomalousBeacon}).
+\ref{fig:raytrace-zs30}, more detail in Section \ref{subsec:anomalousBeacon}
+).
 The nearest bounce method never classifies anything as direct path for the 30 m source, though BELLHOP is quite adept at finding those rays.
@@
 \subsection{Further analysis of the anomalous beacon to beacon connection}
-\label{subsec:anomalousBeacon}
+%\label{subsec:anomalousBeacon}
 

### 2022-02-15 manuscript/main.tex
working through restructuring manuscript

--- 2021-12-20 manuscript/main.tex
@@
 The long baseline (LBL) underwater navigation paradigm relies on the conversion of recorded travel time to range to trilaterate for position.
-For real-time navigation, this conversion assumes an isovelocity sound speed.
+For real-time operations, this conversion assumes an isovelocity sound speed.
 For re-navigation in post-processing, computationally and/or labor intensive acoustic modeling may be employed to reduce uncertainty driven by multipath arrivals.
-This work demonstrates an embedded, ray-based prediction method of the effective horizontal group speed to minimize vehicle position error.
+This work demonstrates a real-time ray-based prediction method of the effective horizontal group speed to minimize vehicle position error.
 \llabel{2.1} This method was implemented for a small scale AUV-LBL system in March 2020, in the Beaufort Sea, in total under-ice conditions and a double ducted acoustic propagation environment.
 The vehicle was successfully deployed and recovered.
-\llabel{2.2} Given the lack of GPS data throughout the vehicle's mission, however, the localization performance is quantified between GPS-linked beacon-to-beacon connections where a single beacon ran the vehicle's navigation stack.
-The real-time positioning error is roughly 11 meters at distances up to 3 km.
-In post-processing, the dataset is extended to all possible beacon-to-beacon events.
-The localization performance is reevaluated with an operationally equivalent navigation stack using modeled, historical, and locally observed sound speed profiles for a Minimal Bounce Criteria (MBC), as deployed in real-time, and an improved algorithm, the Nearest Bounce Criteria (NBC).
-The best performing median re-positioning errors are roughly 10 and 2 meters, respectively.
-\llabel{2.3} Further investigation suggests that the GPS ground truth data have some noise that is not reciprocated in the travel time data.
+\llabel{2.2} Given the lack of GPS data throughout the vehicle's mission, however, the localization performance between GPS-linked beacon-to-beacon connections serves as a proxy for the effectiveness of this approach.
+The real-time positioning error between beacons was roughly 11 meters at distances up to 3 km, matching the magnitude of corrections recorded by the AUV during field operations.
+In post-processing, the localization performance is reevaluated with an operationally equivalent navigation stack using modeled, historical, and locally observed sound speed profiles for a Minimal Bounce Criteria (MBC), as deployed in real-time, and an improved ray filtering algorithm, the Nearest Bounce Criteria (NBC).
+The best performing median re-positioning errors by bounce criterias are roughly 10 and 2 meters, respectively.
+\llabel{2.3} Further investigation suggests that the GNSS ground truth data have some noise that is not reciprocated in the travel time data.
+Re-trilateration suggests that this approach effectively extends the single meter accuracy of the deployed GNSS pucks into the water column.
 \end{abstract}
@@
 \label{sec:1}  
-Autonomous underwater vehicles (AUVs) are increasingly capable platforms to explore and samlpe the ocean, particularly for remote and/or dangerous environments.
-However, navigation uncertainty is a major challenge in considering AUVs as reliable and standard tools for oceanographic research.
-Compared to their land and air-based counterparts, underwater vehicles face a unique set of challenges brought on by the water column itself:  while land and air-based robots utilize the Global Navigation Satellite System (GNSS) to achieve stunning location accuracy and precision throughout the duration of their missions, AUVs cannot access GPS while underwater due to the rapid attenuation of electromagnetic waves \citep{preisig_acoustic_2007}.
-Typically, an AUV can stall on the surface to receive a GPS fix, but this foolproof method of re-positioning is inaccessible in an ice-covered environment or other GPS-denied situations.
+Autonomous underwater vehicles (AUVs) are increasingly capable platforms to explore and sample the ocean, particularly for remote and/or dangerous regions.
+However, navigation uncertainty is a major challenge in considering AUVs as standard tools for oceanographic research.
+While land and air-based robots utilize information from Global Navigation Satellite Systems (GNSS) to achieve stunning location accuracy and precision throughout the duration of their missions, AUVs cannot access GNSS while underwater due to the rapid attenuation of electromagnetic waves.
+Therefore, underwater vehicles have relied on any combination of dead reckoning, hydrodynamic models, inertial navigation systems, doppler velocity logs, and acoustic baseline positioning systems for navigation \citep{paull_auv_2014}.
+Limiting navigation error and drift requires an AUV to periodically stall on the surface and obtain a GNSS fix to reset its position error.
+This foolproof method of self-positioning is undesirable for stealth, adverse weather conditions, and mission efficiency, and inaccessible in a GPS-denied situation like an ice-covered environment.
 
-% Long baseline navigation
-Though there many acoustic positioning methods, the long baseline (LBL) is closest in style
+% introduce LBL and precise statement of work
+Of acoustic baseline navigation systems, long baseline (LBL) is the most GPS-like in style and scale, and most appropriate for mitigating drift without overburdening computation or payload size on the vehicle \citep{paull_auv_2014,van_uffelen_global_2021}.
+The state-of-the-art for LBL outsources depth to a pressure sensor and solves the two-dimensional localization problem by assuming an isovelocity, linear scaling between one way travel time (OWTT) and range \citep{eustice_recent_2006,eustice_experimental_2007,webster_preliminary_2009,webster_advances_2012}.
+This assumption is valid for short scale operations but oversimplifies propagation for larger and/or complex acoustic environments.
+This paper demonstrates an embedded model-aided data processing approach to convert recorded OWTTs into pseudorange estimates.
+This approach was necessitated by total under-ice conditions and a double ducted acoustic environment, in the Beaufort Sea, in March 2020, for an AUV deployment during the Ice Exercise 2020 (ICEX20).
+We quantify the success of the proposed acoustic ranging method using GPS-linked beacon-to-beacon communication events, as the AUV-LBL positioning has no ground truth GNSS to compare to.
 
-% A recent study tracking 22 floats under the ice in the Weddell Gyre near Antarctica showed that they were unable to surface for 8 months and position uncertainty grew to 116 $\pm$ 148 km in the lateral directions \citep{chamberlain_observing_2018}.
+%% rest of the background, JASA style
 
-Much progress has been made in underwater vehicle navigation, which relies on any combination of dead reckoning, hydrodynamic models, inertial navigation systems, doppler velocity logs, and/or acoustic baseline navigation systems \citep{paull_auv_2014}.
-In particular, the long baseline (LBL) navigation system is the most similar in scale and style to GPS, and most appropriate for mitigating drift in underwater vehicle localization without overburdening computation on the vehicle \citep{van_uffelen_global_2021}.
+% PNT
+For clarity, we introduce definitions for timing, positioning, and navigation.
 
-\subsection{Timing, Positioning, and Navigation}
+% More about LBL navigation
+While RAFOS floats have championed one way ranging for re-positioning \citep{rossby_rafos_1986,duda_evaluation_2006}, the ability to do so for navigation was facilitated by \llabel{1.11} the advent of the WHOI micro-modem \citep{singh_underwater_2006} and synchronized chip scale atomic clocks \citep{gardner_second_2016}.
+Several short range (less than ten kilometers) navigation efforts have achieved minimal error with a nominal sound speed value \citep{eustice_experimental_2007,webster_preliminary_2009,kepper_mems_2017}.
+However, these efforts were all conducted with X sound speed conditions.
+The approach in this paper, and other related works, was specifically designed to... the recently observed double ducted sound speed profile...
+For convenience, we call this phenomena the Beaufort Lens, with the lens referring to ...
 
-Underwater acoustic navigation (and GPS) is predicated on timing and positioning.
-Timing is the ability to maintain accurate and precise time, sometimes synchronized across multiple nodes.
-\llabel{1.11} The advent of the WHOI micro-modem \citep{singh_underwater_2006} and synchronized chip scale atomic clocks \citep{gardner_second_2016} enabled ranging via one-way travel time (OWTT).
+There is no comparable work in the Arctic for a short scale AUV deployment in the Beaufort Lens. 
+Both seminal and more recent AUV deployments witnessed the classical upward refracting sound speed profile that is amenable to an isovelocity simplification.
+Of note are large scale glider deployments; these, however, operated at X depths, X spatial scales, and did not need precise navigation because they experienced little to no sea ice.
+At larger scales on the order of hundreds of kilometers, a deterministic \citep{graupe_preliminary_2019} or through-the-sensor \citep{webster_towards_2015} value for sound speed was used, but error remained in the hundreds of meters.
+The sound speed value was chosen by...
 
-Positioning is the ability to accurately determine one's location and is generally done via trilateration.
-Navigation assumes a degree of autonomy to move from a current position to a desired one.
-The standard for real-time navigation efforts assume a locally homogeneous sound speed profile (SSP) that enables a linear, deterministic scaling between travel time and range \citep{eustice_recent_2006,eustice_experimental_2007,webster_preliminary_2009,webster_advances_2012}.
-Short range (less than ten kilometers) navigation efforts achieve minimal navigation error with a nominal sound speed value \citep{eustice_experimental_2007,webster_preliminary_2009,kepper_mems_2017}.
-For mesoscale (tens to hundreds of kilometers) navigation efforts, a deterministic \citep{graupe_preliminary_2019} or through-the-sensor \citep{webster_towards_2015} value for sound speed is used, but error remains in the hundreds of meters.
-
-Post-processing localization efforts (re-navigation for a self-determinedly moving object or re-positioning for a passively moving one), without computational or time constraints, can indulge more complex acoustic methods and/or ocean representations.
-There are many different approaches, from considering a temporally averaged mean depth dependent SSP \citep{van_uffelen_localization_2016}, including spatio-temporal variability in the sound speed profile to minimize error \citep{graupe_preliminary_2019, mikhalevsky_deep_2020}, comparing acoustic records with synthetic acoustic records computed through an ocean model \citep{wu_deep_2019}.
-The ``cold start'' algorithm developed by \citet{mikhalevsky_deep_2020} isolates a specific ray path to find a representative group velocity and re-positioned a floating hydrophone array with an error of 85 m and a standard deviation of 32 m relative to ground truth GPS data, but implementing this algorithm for navigation remains to be seen.
-These post-processing efforts have generally been deemed too computationally intensive and/or labor intensive for real-time efforts, where automation of identifying time arrivals is further complicated by multipath.
-
-\subsection{Underwater positioning and navigation in the Arctic}
-
-Underwater navigation research is broadly motivated by GPS-like navigation in GPS-denied conditions.
-The Arctic, while remote, is the perfect place to test mature navigation technologies in real GPS-denied conditions.
-
-Literature and historical reviews of underwater navigation approaches in under-ice conditions highlight the emerging technology learning curve for vehicle operations in the Arctic, the hostile conditions brought on by a semi-permanent moving ice cover, and the need for both absolute and relative navigation with little to no tolerance for error \citep{norgren_unmanned_2014,mcfarlane_autonomous_2015,barker_scientific_2020}.
-
-One of the first successful AUV launches and recoveries involved equipping the vehicle with barbs which became entangled into a homing net underneath the ice \citep{bellingham_auv_1993}.
-Since then, AUVs have been used for ocean bottom \citep{jakuba_long-baseline_2008,kunz_deep_2008}, under-ice \citep{kaminski_12_2010}, and hydrographic surveys \citep{kukulya_under-ice_2010,kukulya_development_2016,stevens_linking_2016,plueddemann_autonomous_2012}.
-These efforts show a combination of short and long range positioning systems, where all acoustic navigation methods used a depth-averaged sound speed for linear scaling between recorded travel time and range.
-In many cases robust quantitative navigation performance could not be evaluated due to small operational scale and/or moving ice floes.
-But most importantly, all of these efforts experienced a nearly isovelocity or classical upward refracting Arctic acoustic environment.
-
-The new Arctic\textemdash one with much less multi-year sea ice\textemdash is experiencing significant changes in the water column as well \citep{mackinnon_warm_2021}.
-\llabel{2.4} The Beaufort Lens is a highly dynamic, highly stratified water mass, keeping a neutrally buoyant layer of warm Pacific water at 50 to 70 m in depth.
-Surrounded by cooler waters, it distinguishes a local maximum in the SSP, creating an upper and lower duct\textemdash the upper duct degrades signal coherence due to intensified ice interaction and the lower duct effectively traps sound for long range propagation \citep{poulsen_acoustic_2016}.
-\llabel{2.5} Modeling output \citep{duda_long-range_2019,duda_effects_2021} and experimental observations \cite{kucukosmanoglu_observations_2021,badiey_azimuthal_2019,ballard_temporal_2020} indicate a high degree of spatio-temporal variability of the lower duct throughout the Chukchki and Beaufort Seas.
-
-An experiment in 2014 exploited this lower duct to provide acoustic navigation and communication at mesoscale ranges in the Canada Basin.
-\citet{freitag_long_2015} and \citet{webster_towards_2015} deployed multiple GPS-linked ice buoys with transducers that were capable of communications greater than 400 km with a range accuracy of 40 m when the sound speed was known.
-The transducers were at 100 m depth, well-positioned in the lower duct spanning 50 to 200 m.
-The glider navigation accuracy, however, was severely degraded as acoustic packets were sent when the gliders were outside the sound channel.
-When gliders were deployed again in 2017, their navigation error was on the order of half a kilometer, and a glider re-navigation using an acoustic arrival method and the average sound-speed profile from the gliders reduced navigation error by a factor of 4-5, depending on the dive \citep{graupe_preliminary_2019}.
-Navigation error is referenced to intermittent GPS fixes, as all of these glider deployments occurred in the Marginal Ice Zone with little to no ice at the time.
-
-\subsection{Ice Exercise 2020}
-
-The Beaufort Lens is equally important for shorter ranges as well, where a shadow zone over vehicle operational ranges is created by a drastic change in the reliable acoustic path \citep{schmidt_acoustic_2016}.
-Thus the key simplifying assumption for OWTT ranging\textemdash a locally homogeneous sound speed profile that enables a linear, deterministic scaling between travel time and range\textemdash becomes more inaccurate due to increased multipath in the upper duct of the Beaufort Lens.
-
-The results from this paper derive from data taken while deploying AUV Macrura, a modified Bluefin-21, during the Ice Exercise 2020 (ICEX20), in the Beaufort Sea, in March 2020.
-The challenges anticipated for ICEX20\textemdash total ice cover, moving ice floes, and a variable Beaufort Lens\textemdash crystallized the necessity of environmental awareness for the navigation solution.
-
-In this paper we present a working solution for underwater navigation that achieves GPS-like accuracy and precision.
-We embed a stochastic prediction of a single group velocity to convert recorded travel time into range.
-This prediction ingests an updatable snapshot of the SSP via acoustic message; over the course of ICEX20, this capability was used several times. 
-Ultimately, integrating environmental realism enabled a successful emergency recovery of the AUV.
-Due to a harddrive error, the AUV stalled underneath the ice surface and broadcasted its location.
-A team was able to dig a hole through the 2 meter ice sheet and immediately see the vehicle.
-Its successful recovery amidst total ice cover is shown in Fig. \ref{fig:vehicleRecovery}.
-\begin{figure}[h!]
-	\centering
-	\includegraphics[width=\reprintcolumnwidth]{figs/Fig1.jpg}
-	\caption{The successful emergency recovery of AUV Macrura, with the the auger used to drill through the ice sheet in the background. Photo courtesy of LT CDR D. Goodwin.}
-	\label{fig:vehicleRecovery}
-\end{figure}
-
-The rest of this paper is organized as follows: Section~\ref{sec:methods} presents the physical
-experimental setup and model-based data processing; 
-section \ref{sec:realtime} quickly conveys real-time results and the limitations of the data collected; section~\ref{sec:postprocess} presents key results for group velocity predictions and range estimations; section~\ref{sec:discussion} closes with a discussion on importance, future work, and broader relevance.
-
-% =========================================================================== %
-% =========================================================================== %
-\section{Methods}
-\label{sec:methods}
-
-The acoustic communication dataset accrued for ICEX20 relied on the Integrated Communication and Navigation Network (ICNN) \citep{schneider_self-adapting_2020,randeni_construction_2020,randeni_high-resolution_2021}.
-The ICNN was initially developed via numerous virtual experiments, to push the boundaries of algorithms and interfaces between different hardware components.
-The simulation approach is a necessary testbed for robustness and produces better results than post-processing previous field data, as that restricts mission configurations to the data taken, which can hamper troubleshooting in the field.
-The simulation capabilities are largely physics-driven with a modular system of systems approach: an environmental simulator with sub-components for the ocean, including Arctic ice drift and ocean acoustic propagation; a vehicle simulator with sub-components for vehicle dynamics and navigation; a topside hardware simulator and acoustic communications simulator, both with a software-only configuration and a hardware-in-the-loop version \citep{schneider_netsim_2018}.
-The virtual environment similarly emulates the interfaces between the real components to test the entire software pipeline.
-Both simulation capabilities are integral to mission success.
-
-\subsection{Operational paradigm on the ice}
-
-The ICNN is comprised of four ice buoys, in a loose rectangle, roughly 2 km away from a central ice camp with a topside computer, as shown in Fig.~\ref{fig:icnnOverview}.
+The navigation approach proposed includes a non-deterministic relationship between travel time and range previously limited to large scale re-navigation and re-positioning efforts.
+Such post-processing efforts can utilize additional information or computation that is inaccessible for onboard computation.
 
@@
 	\centering
-	\includegraphics[width=\reprintcolumnwidth]{figs/Fig2.pdf}
+	\includegraphics[width=\columnwidth]{figs/auv-track-update.png}
+	\caption{The under-ice mission track for AUV Macrura, including the position updates as it stalled underneath the ice overnight.
+A marker was placed on the ice at the vehicle's estimated self-location.
+It was recovered after a three day storm within a meter of the marker.}
+	\label{fig:vehicleRecovery}
+\end{figure}
+% include story later on and cite this figure?
+
+\clearpage
+\section{ICEX20 conditions and experiment design}
+
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=\reprintcolumnwidth]{figs/ssp-gvel-large.pdf}
+	\caption{Anticipated sound speed conditions: finish this using stuff from ICEX16, HYCOM, ITP}
+	\label{fig:sspExpectation}
+\end{figure}
+
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=\textwidth]{figs/Fig2.pdf}
 	\caption{A schematic overview of the Integrated Communication and Navigation Network (ICNN), which provides joint data-transfer and tracking between AUV and a human decision maker at Topside.}
@@
 
-Each ice buoy is outfitted with a WHOI Micro-Modem \citep{singh_underwater_2006}, with a four-element receiver array and a single transmitter, and one-tenth of a millisecond resolution.
-The receiving and transmitting elements were split between shallow and deeper depths (30 and 90 m, respectively) to provide better coverage due to the acoustic shadow zone in the Beaufort Lens.
-The buoys do not encompass the full operating range of the vehicle but are positioned to minimize overlap in trilateration for spherical positioning \citep{deffenbaugh_relationship_1996}.
-
-Vehicle missions in any underwater environment, but especially acoustically complex ones, must balance competing uses of the acoustic channel.
-This network uses a single synchronized digital communication packet to provide both tracking and data to the operator.
-The ICNN operational paradigm is as follows:
-\begin{enumerate}
-\item The AUV, running an ice-tracking DVL and an onboard hydrodynamic model, broadcasts its perceived location on a scheduled, time-synchronized message via WHOI Micro-Modem
-\item Four ice buoys, each outfitted with a WHOI Micro-Modem, receive messages from the AUV and send that information over freewave radio to a Topside computer
-\item The topside computer converts travel times into range estimates using a stochastic embedded prediction of the horizontal group velocity via BELLHOP ray tracing code \citep{porter_bellhop_2011} coupled with a Virtual Ocean \citep{schneider_netsim_2018}
-\item The topside computer calculates a new position by trilaterating the range estimates
-\item The position differential, not the absolute position, is broadcast to the vehicle to update its navigation solution and be robust to latency and intermittency
-\end{enumerate}
-
-This paper specifically addresses the third step\textemdash the topside computer's conversion of travel time into range, which underpins the trilateration in the ICNN, and is an often overlooked element of any underwater acoustic navigation system.
-The analysis accordingly focuses on validating acoustic range estimates from communication events between GPS-tracked beacons, not the AUV itself.
-This allows the results to be decoupled from all the other moving parts in the ICNN.
-This ``modem experiment'' was conducted while the vehicle was still being prepped to go under the ice.
-
-\subsection{Modem experiment design}
-
-Because the navigation solution on the vehicle during a mission can only be evaluated on the basis of the error estimates sent, a sister experiment for validating the real-time ranging approach was implemented.
-The ice buoy modems were run as ``virtual vehicles'' at a given depth and would transmit to all other ice buoys.
-The internal sound speed estimate was then toggled using a handful of weights applied to a basis function representation of the sound speed.
-This trial was configured for as many ice buoy modems as possible with the time allotted, with each experiment running for roughly an hour between the two sound speed modalities.
-There are three important engineering choices in the ICNN network to contextualize this dataset:
-\begin{enumerate}
-\item Each modem buoy only has one transmit depth layer (30 or 90 m)
-\item \llabel{1.12} The transmit and receive layers are independent
-\item At any given time, the receive depth layers are the same across all four ice buoys
-\end{enumerate}
-
-The design of the ICNN enables a self-adapting network to transmit and receive at the optimal depth to maintain contact with the AUV \citep{schneider_self-adapting_2020}.
-This adaptivity was not used during the modem experiment but was instead toggled manually.
-Given the complexity of the ICNN system, this experiment did not collect an exhaustive set of data across all buoy, source depth, receive layer, and sound speed combinations, as this time was also used to verify and resolve any hardware issues in the modems themselves.
-Importantly, only one modem at a time could run the vehicle behavior to trigger group velocity calculations; other modems that were active during the experiment recorded events as tracking buoys, not as vehicles.
-
-\llabel{1.13} A key construction in the ICNN are parallel physical and virtual layers; given the small operational scale, on the order of kilometers, and a singular SSP, both layers are projected into a range independent coordinate system for ease of comparison.
-Furthermore, because one of the buoys was configured to act as a virtual vehicle at any given time during the modem experiment, the tracking range was supplemented with an additional modem deployed from the camp site and configured to act as a virtual buoy; the transducer attached to the camp site modem was lowered to a depth of 20 m.
-
-\subsubsection{Physical layer}
-The physical layer describes the acoustic link between beacons in the water column.
-It covers all source depths (20, 30, and 90 m) and both buoy receive layers (30 and 90 m), where the surface expressions of the beacons are tracked by GPS.
-Events are sorted by source depth and sound speed profile, given that these are the dominating factor for the location and span of the shadow zone compared to any temporal or spatial scales of variability.
-
 \begin{figure}[h!]
   \centering
-  \includegraphics[width=.51\textwidth]{figs/Fig3a.pdf} \hfill
-  \includegraphics[width=.48\textwidth]{figs/Fig3b.pdf} \\
-  \vspace{1em}
-  \includegraphics[width=.48\textwidth]{figs/Fig3c.pdf} \hfill
-  \includegraphics[width=.48\textwidth]{figs/Fig3d.pdf} \\
+  \includegraphics[width=\textwidth]{figs/modem-chart.pdf}
   \caption{An overview of the modem experiment by source and receiver depth and position.
-Subfigure (a) shows a chart of the successful events between source and receiver depth.
 The black column on the left, $tx$, shows the source depth, $z_s$.
 The column on the right, $rx$, shows the receivers with the amount of good contacts.
-The orientation of the triangles\textemdash sideways, upwards, and downwards\textemdash corresponds to depths of 20, 30, and 90 m.
-Subfigures (b-d) correspond to transmitter depths of 20, 30, and 90 m, respectively.
-Topside, demarcated by the purple triangle, is situated just north of the continental shelf.}
+The orientation of the triangles\textemdash sideways, upwards, and downwards\textemdash corresponds to depths of 20, 30, and 90 m.}
   \label{fig:overview}
@@
 \clearpage
+\section{\label{sec:realtime} Real-time pseudorange analysis}
 
-\subsubsection{Virtual layer}
-The virtual layer describes the BELLHOP simulations for each communication event in the physical layer.
-The SSP used for the raytracing is driven by a handful of weights applied to a basis function representation
-of the SSP variability to reconstitute a number of realistic candidate SSPs \citep{leblanc_underwater_1980,lin_merging_2010}.
-The SSP is the only human configurable input to the group velocity prediction.
+\subsection{Minimal bounce criteria (MBC)}
 
-\llabel{1.14} During ICEX20, the SSP was toggled between the historical``baseline'' and a ``weighted'' estimate based on a local CTD cast, i.e.
-an optimal set of weights chosen by a human decision maker \llabel{2.7} via an interactive Tactical Decision Aid framework \citep{bhatt_embedded_2021}.
-\llabel{1.8} For the post-processing analysis, we add the SSP from HYCOM for comparison, to mirror the information available on a submarine\textemdash modeled data, historical data, and \textit{in situ} data (personal conversation with LT B.
-Howard and LT CDR D.
-Goodwin).
-This order corresponds to increasing ducted conditions of the Beaufort Lens.
+\subsection{Pseudorange error metrics}
 
-Fig. \ref{fig:raytrace-zs30} shows eigenrays in a range independent coordinate system for all three sound speed environments, for a source depth of 30 m.
-\llabel{1.16} It is important to note that BELLHOP produces an \llabel{2.8}abundance of eigenrays; the ones displayed here are chosen by travel time proximity to the recorded data.
-With this filter, the chosen eigenrays show a greater amount of surface interactions as the SSP duct conditions increase.
+\subsection{Discussion}
 
+\clearpage
+\section{\label{sec:post} Post-processed pseudorange analysis}
+
+\subsection{Inherent overestimation from the minimal bounce criteria}
 \begin{figure}[h!]
@@
 
-\FloatBarrier
-\subsection{Group velocity prediction methods}
+\subsection{Nearest bounce criteria (NBC)}
 
-The variation of horizontal group velocity from any source-receiver pair is the fundamental challenge to implement GPS-like navigation in an LBL navigation paradigm, especially in an acoustically complex propagation environment.
-The use of the acoustic modem network for tracking relies on the accurate estimates of travel times between the submerged platform and range modems, supported by clock synchronization and a pre-determined scheduling of acoustic events.
-\llabel{2.9} For the Beaufort Lens in particular, the strong multipath effects make it virtually impossible to deterministically predict the modem triggering time.
-
-\subsubsection{Minimal bounce criteria (MBC)}
-Instead, for each individual node $i$, an embedded stochastic tracking framework is used to provide a running estimate of the horizontal group velocity $u_{i,j}$ for the conversion from travel time to range from modem $j$, with the ultimate goal of matching the naive group velocity, i.e.
-the GPS-recorded distance between two nodes divided by the modem-recorded one way travel time between them.
-
-\llabel{2.10} In the ICEX20 configuration, the acoustic tracking is running on the topside computer, which controls the integrated communication and navigation range.
-Here we assume that the group velocities $u_{i,j}$ are smoothly varying over the course of a vehicle mission, i.e., with respect to range, mission time, and the frequency of updates relative to vehicle motion.
-The group velocity is continuously tracked using predictions from the \textit{Virtual Ocean} infrastructure.
-
-\llabel{1.15} When the topside tracking framework receives a modem message, with a time delay, $\Delta t$, from one of the range modems, it will request a new estimate of the group velocity and its associated uncertainty.
-The group velocity estimate is found using the vehicle's reported depth and the extrapolated navigation solution for range, $\hat{r}$, as inputs for the ray tracing program.
-The latter returns an impulse response estimate in the form of ray travel times $dt_{j}$ and amplitudes $a_{j}$ for that range and depth.
-
-\llabel{2.11} The initial call to BELLHOP is over a sparse local grid, from which the plane wave approximation is used to interpolate the ray to the range and depth posited by the onboard tracking solution.
-For each grid point, BELLHOP will produce a number of arrivals resulting from multiple propagation paths for any source-receiver pair.
-Using only the $N_0$ rays with neither surface nor bottom bounces, it will then estimate the current group velocity $u$ from a power weighted average of the ray travel times,
-\begin{equation}
-u = \frac{\hat{r} \sum_{n=1}^{N_{0}} a_{n}^{2}}{\sum_{n=1}^{N_{0}} dt_{n}a_{n}^{2}} ~, 
-\end{equation}
-and the associated weighted standard deviation,
-\begin{equation}
-\sigma_{u} \simeq \sqrt{\frac {\sum_{n=1}^{N_{0}} (dt_{n}-\hat{r}/u)^{2}a_{n}^{2}}{ \sum_{n=1}^{N_{0}} a_{n}^{2}} } \frac{u^{2}}{\hat{r}}
-\end{equation}
-If no direct paths exist, i.e. $N_{0}=0$, then the group velocity is computed using the same algorithm for the ray arrivals with one bounce, and so on.
-
-Finally, the pseudorange is calculated simply as
-\begin{equation}
-r_{i,j} = u_{i,j} \Delta t_{i,j} 
-\end{equation}
-
-This stochastic method for group velocity calculation can run in real-time, appearing to be orders of magnitude faster than general post-processing methods which seek to determine the specific ray itself that best matches a prominent indicator from the arrival structure.
-The method assumes direct path arrival, where multipath over the shallow receiver depth is trivial for the group velocity calculation; thus we call it the minimal bounce criteria (MBC).
-The BELLHOP simulation that runs this calculation uses 3600 rays with launch angle fan of -60 to 60 degrees.
-However, this high density of launch angles creates a greedy estimator of direct path eigenrays, and this simulation pipeline almost always guarantees a direct path found.
-The MBC improves upon a deterministic estimate in that it considers the simplest ray path, travel time, and amplitude to predict a realistic group velocity.
-
-\subsubsection{Nearest Bounce Criteria}
-
-\llabel{2.12} As shown in the eigenray traces of Fig.
-\ref{fig:raytrace-zs30}, an acoustic arrival does not always take the direct path from source to the receiver, and the path it takes is extremely dependent on the sound speed profile.
-The suggested algorithm, the nearest bounce criteria (NBC), is a slight modification from the minimal bounce criteria, and includes multipath as a new dimension of information with negligible additional computation.
-This metric was evaluated in post-processing.
-
-Given a running estimate for the horizontal group velocity $u_{i,j}$ between nodes $i$ and $j$, the navigation system has an extrapolated value for range, $\hat{r}$, in the virtual layer, and a recorded travel time, $\Delta t_{i,j}$, in the physical layer.
-Instead of using only the $N_0$ rays with neither surface nor bottom bounces to estimate group velocity and subsequently moving to incremental number of bounces only when no valid direct path solutions exist, we solve for the power weighted average of the ray travel time for the $N_k$ rays with $k$ bounces,
-\begin{equation}
-t_k = \frac{\sum_{n=1}^{N_{k}} dt_{n}a_{n}^{2}}{\sum_{n=1}^{N_{k}} a_{n}^{2}} ~, 
-\end{equation}
-find the nearest matching power weighted average to recorded travel time,
-\begin{equation}
-t_{i,j,k} = \min_{k=0,1,2,...} \left| t_k - \Delta t_{i,j} \right|
-\end{equation}
-predict a group velocity,
-\begin{equation}
-u_{i,j} = \dfrac{\hat{r}}{t_{i,j,k}}
-\end{equation}
-and estimate the range as was done previously.
-\begin{equation}
-r_{i,j} = u_{i,j}\Delta t_{i,j}
-\end{equation}
-
-This method selects a different group velocity based on the multipath arrival structure, as the detected arrival is not always the first arrival or the direct path and could even be masked by noise or blocked temporarily \citep{deffenbaugh_acoustic_1996}.
-We manually cap the number of bounces to 4 because of the smaller operational scale and the attenuation accrued with many surface interactions.
-Bottom bounces are not encoded separately because of ray's tendency to refract upward, not due to information limitations. 
-\llabel{1.17} In addition, for a more consistent comparison, the local sparse grid is standardized as 11$\times$11 points over 10 m in range and 20 m in depth, which is similar to those from the vehicle logs.
-The extra size in depth addresses the computational limitations of how time-bounded eigenrays seem to stack vertically in the water column, as shown in Fig \ref{fig:raytrace-zs30}.
-
-% =========================================================================== %
-% =========================================================================== %
-
-\FloatBarrier
-\section{\label{sec:realtime} Real-time pseudorange results}
-
-The modem experiment generated 811 beacon to beacon communication events with their own real-time MBC group velocity predictions.
-\llabel{1.18} As expected, the algorithm is generally overestimating range as it resolves a direct path that does not represent the actual arrival.
-This section examines the absolute range error between the pseudorange and the GPS derived range with the caveat that the data collected is fairly unbalanced across source depths and sound speed conditions.
-
-\begin{table}[h!]
-\renewcommand{\arraystretch}{1.3}
-\centering
-\begin{tabular}{r|c|c|c|c|c}
-SSP & Count & ~Mean [m] & ~Median [m] & ~STD [m] & ~Max [m] \\ \hline
-Baseline & 243 & 11.38 & 11.96 & 4.23 & 23.95 \\ 
-Chosen weights & 568 & 11.36 & 11.40 & 8.12 & 26.0 \\
-\toprule
-\end{tabular}
-\caption[Range estimation error for \textit{in situ} events]{The range estimation error metrics for \textit{in situ} events.
-The error here is presented as the absolute, not directional, error.
-Almost all directional errors, however, are positive.}
-\label{tab:rangeErrorInSitu}
-\end{table}
-
-Table \ref{tab:rangeErrorInSitu} breaks down the number of events, mean, median, standard deviation, and maximum absolute range error for the \textit{in situ} calculations.
-The mean and median for both sound speeds are similar on the order of centimeters; the chosen weights show a maximum 2 meters greater than that of baseline, but it is from a communication event that has no comparison in the baseline data set.
-Analysis shows that where there is overlap in range between sound speed conditions used for the real-time MBC, the range error difference is no more than a few meters.
-\llabel{2.14} This is likely driven by the prominence of the duct.
-
-The overarching statistical results show that both sound speed environments, given the smaller operational scales and shallow depths, are accurate enough to approximate the group velocity structure given the local grid.
-Indeed, the error statistics indicated here are similar to those reported for the ICNN \citep{randeni_high-resolution_2021} and accurate enough to have facilitated a successful vehicle recovery.
-
-\FloatBarrier
-\section{\label{sec:postprocess} Post-processing pseudorange results}
-
-There are 1242 total beacon to beacon events in the modem experiment that would have been accepted by the ICNN for real-time ranging and navigation had the ranging behavior been running for all modems in the network, and not only for tracking the vehicle.
-A post-processing analysis that replicated the real-time pipeline was run to overcome the unequal distribution of communication events with respect to depth, range, and sound speed status.
-
-It is important to note that the value for the extrapolated range, $\hat{r}$, is only tracked when the modem runs the vehicle behavior; thus we replace $\hat{r}$ with the GPS-tracked range for all modem events.
-Because $\hat{r}$ converges to the correct solution, a comparison of $\hat{r}$ with the GPS-tracked range shows a normal, zero-centered distribution within the bounds of GPS drift.
-The present analysis therefore seeds realistic but ``omniscient'' knowledge of the extrapolated range and leverages the post-processing pipeline to more thoroughly evaluate the acoustic range estimate for all modem events, with three relevant sound speed sources, and both group velocity criterion.
-Accordingly, the results in this section evaluate the utility of the algorithms and sound speed sources, divorced from their role in the ICNN while maintaining real-time relevance.
-
-\FloatBarrier
-\subsection{Group velocity predictions}
-
-The minimal and nearest bounce algorithms are applied, both without the plane wave extensions, with the three sound speed inputs shown in Fig.~\ref{fig:raytrace-zs30}.
-The resulting predicted group velocities for a source depth of 30 m are shown in Fig~\ref{fig:gvel30}.
-The takeaways from this source depth are representative of those from 20 and 90 m as well (see Supplementary Fig.~\ref{fig:gvelMore}).
+\subsection{Effective sound speed predictions}
 
@@
 
-Because direct path dominates the predictions from the MBC, group velocity estimates are greater than those from NBC.
-The one instance this is not true is important to mention\textemdash for a travel time of 2.2 s, where the minimal multipath is dominated by a single bottom bounce (the red line to the furthest transmission in Fig.
-\ref{fig:raytrace-zs30}, more detail in Section \ref{subsec:anomalousBeacon}).
-The nearest bounce method never classifies anything as direct path for the 30 m source, though BELLHOP is quite adept at finding those rays.
-
-The goal of the group velocity estimation is to converge towards the naive group velocity estimation, i.e. the GPS-derived range divided by the recorded travel time.
-For a 30 m receiver depth, the NBC shows more overlap with data-derived values as it classifies multipath more correctly.
-For a 90 m receiver depth, the overlap is less accurate due to computational constraints of a limited fan of rays entering the shadow zone rendering a less reliable simulated times of arrival packet.
-
-Overall, analysis shows that the higher multipath classification produces more accurate range estimates, likely driven by a tighter bundle of rays.
-However, that data are too small to draw significant conclusions.
-\llabel{1.19} Discontinuities in multipath classification verify our hypothesis for its importance to a smoothly varying group velocity, as shown in the cluster for a receiver depth of 30 m, where HYCOM jumps from one to two bounces amidst the baseline SSP and weighted SSP smoothly increasing while consistently seeing two and three bounces, respectively.
-
-\FloatBarrier
-\subsection{Pseudorange error versus one way travel time}
-
-Fig. \ref{fig:rangeError} shows the directional range error footprints for all three sound speed inputs with respect to OWTT, separated by source and receiver depth configurations.
-The weighted SSP range error generally has the smallest and most zero-centered footprint.
-The one case it does not is for the source-receiver pairings between 30 and 90 m in depth.
-The increased error for these reciprocal transmission paths is most likely driven by the computational artifacts encountered when propagating through the steep sound speed gradients of the lens and through the shadow zone.
-All other source depth pairings are significantly improved using the chosen weights compared to HYCOM or the baseline.
-
-When using a linear scaling to convert travel time into range, any offset between the assumed sound speed and the horizontal group velocity produces unconstrained error with increasing receiver distance.
-Most importantly, we see the consequences of the adaptive group velocity in that range error does not strictly increase with OWTT.
+\subsection{Pseudorange error metrics}
 
@@
 
-\FloatBarrier
-\subsection{Pseudorange estimations from the nearest and minimal bounce criteria}
+\subsection{Discussion}
 
-The previous section displayed the pseudorange performance using the NBC.
-Here, we explicitly compare the ranging accuracy between the nearest and minimal bounce criteria for all events in the post-processing pipeline.
-It is important to note that because the post-processing pipeline is the same for both methods, that the efficacy of the NBC compared to the MBC is 100\%, as the former simplifies to the latter for direct path classifications.
+\section{Re-positioning using the nearest bounce criteria}
 
@@
 
-Figure \ref{fig:compareV1V2} is a crossmap for all three SSPs for both group velocity predictors, with the minimal bounce on the x-axis and the nearest bounce on the y-axis; the important range estimation metrics are summarized in table \ref{tab:rangeErrorV1V2}.
+\begin{figure}[!ht]
+\includegraphics[width=\textwidth]{figs/trilateration-3-ways.pdf}
+\caption{WRITE THIS. Needs font change, legend change, x-axis change, new color palette.}
+\label{fig:trilat}
+\end{figure}
 
-All the events are on the unshaded region of the plots\textemdash the MBC has a systemic overestimation, which is reduced via NBC as events straddle the x-axis.
-The improvement from MBC to NBC is most evident for the realistic sound speed; while the HYCOM SSP improves from a median absolute range error of 6.41 to 4.61 m, the baseline SSP improves from 10.30 to 2.27 m, and the weighted SSP improves from 13.28 to 2.12 m.
-Table \ref{tab:rangeErrorV1V2} shows further statistics on the absolute range error by SSP and group velocity algorithm.
-The order of magnitude improvement in the ducted SSPs demonstrate the effectiveness of the algorithm exploiting the multipath conditions.
-
-Given the computational constraints of real-time modeling, the gridded approach facilitates enough multipath classification to build in a ``ray ensemble'' of characteristic group velocities.
-This result is not always possible when aiming to find eigenrays to just an individual point, even with a higher density of launch angles.
-An important takeaway for those interested in ray-based localization is leveraging a local grid to give BELLHOP tolerance for finding solutions that otherwise may not be found in a center or single point solution.
-The limitations of numerical computation, particularly for a complex environment, are more adeptly addressed by accepting some uncertainty in position than by prescribing an exact solution.
-Even though BELLHOP is a greedy estimator for eigenrays, the additional data created is a negligible burden.
-
-\begin{table}[h!]
-\renewcommand{\arraystretch}{1.5}
-\centering
-\begin{tabular}{r|cc|cc|cc}\toprule
- & \multicolumn{2}{c|}{\textbf{Baseline} } & \multicolumn{2}{c|}{\textbf{Chosen Weights} } & \multicolumn{2}{c}{\textbf{HYCOM}} \\
-  & \cellcolor[HTML]{EFEFEF}MBC & NBC & \cellcolor[HTML]{EFEFEF} MBC& NBC & \cellcolor[HTML]{EFEFEF}MBC & NBC \\ \hline
- minimum [m] 	& \cellcolor[HTML]{EFEFEF}0.01 & 0.00	& \cellcolor[HTML]{EFEFEF}0.00 	& 0.00 	& \cellcolor[HTML]{EFEFEF}0.11 & 0.01 \\
- 25th \% [m]   & \cellcolor[HTML]{EFEFEF}4.96 & 0.99	& \cellcolor[HTML]{EFEFEF}6.26 	& 0.95 	& \cellcolor[HTML]{EFEFEF}3.30 & 2.25 \\   
- median [m]		& \cellcolor[HTML]{EFEFEF}10.30 & 2.27 	& \cellcolor[HTML]{EFEFEF}13.28 & 2.12 	& \cellcolor[HTML]{EFEFEF}6.41 & 4.61 \\
- 75th \% [m]   & \cellcolor[HTML]{EFEFEF}15.81 & 5.51 	& \cellcolor[HTML]{EFEFEF}19.75 & 4.11 	& \cellcolor[HTML]{EFEFEF}10.92 & 7.46 \\
- maximum [m]    & \cellcolor[HTML]{EFEFEF}22.52 & 14.96 & \cellcolor[HTML]{EFEFEF}1491  & 20.21 & \cellcolor[HTML]{EFEFEF}19.55 & 15.81 \\
- \toprule
-\end{tabular}
-\caption[Comparison of post-processing range estimation algorithms across all events]{A comparison of range estimation metrics for each sound speed source and group velocity estimation algorithm for all 1283 beacon to beacon events via post-processing.
-The 0th (minimum), 25th, 50th (median), 75th, and 100th (maximum) percentiles are shown to the range resolution afforded by the WHOI Micro-Modem.
-There are a few outliers that drive the mean to be higher than the median.}
-\label{tab:rangeErrorV1V2}
-\end{table}
-
-All plots in Fig. \ref{fig:compareV1V2} show the same 45 degree banding of source depth and multipath structure, indicating systemic GPS noise.
-The width of those bands indicates the high precision of the acoustic methods relative to lower precision from GPS.
-Some bands are adjacent to others of the same styling, likely due to ice drift and/or various GPS satellite constellations.
-
-\FloatBarrier
-\subsection{Further analysis of the anomalous beacon to beacon connection}
-\label{subsec:anomalousBeacon}
-
-As shown in table \ref{tab:rangeErrorV1V2}, there is a striking maximum range error of 1491 m for the weighted SSP in the minimal bounce criteria.
-This section isolates the events that encompass that outlier for further examination, summarized in table \ref{tab:rangeErrorPostExplanation}.
-There are 10 events from South transmitting at 30 m depth to North receiving at 30 m depth.
-The OWTT spread is from 2.1958 to 2.1963 s; the naive group velocity is 1429.3 to 1430.1 m/s; and the GPS-tracked range is from 3138.54 m to 3140.87 m.
-This example ends up being an excellent case study for how sound speed and multipath fidelity work in concert to minimize range error.
-
-\begin{table}[h!]
-\renewcommand{\arraystretch}{1.5}
-\centering
-\begin{tabular}{r|cc|cc|cc}\toprule
- & \multicolumn{2}{c|}{\textbf{Baseline} } & \multicolumn{2}{c|}{\textbf{Chosen Weights} } & \multicolumn{2}{c}{\textbf{HYCOM}} \\
-  & \cellcolor[HTML]{EFEFEF}minimal & nearest & \cellcolor[HTML]{EFEFEF} MBC& NBC & \cellcolor[HTML]{EFEFEF}minimal & nearest \\ \hline
- \# bounces 				& \cellcolor[HTML]{EFEFEF}1 		& 3 		& \cellcolor[HTML]{EFEFEF}1 		& 4 		& \cellcolor[HTML]{EFEFEF}0 	 & 2 \\
- mean OWTT [s] 				& \cellcolor[HTML]{EFEFEF}2.1853 	& 2.1943 	& \cellcolor[HTML]{EFEFEF}4.1812 	& 2.1955 	& \cellcolor[HTML]{EFEFEF}2.1877 & 2.1902 \\
- mean group velocity [m/s] 	& \cellcolor[HTML]{EFEFEF}1436.7 	& 1430.8 	& \cellcolor[HTML]{EFEFEF}750.88 	& 1430.0 	& \cellcolor[HTML]{EFEFEF}1435.1 & 1433.4 \\
- range error [m] 			& \cellcolor[HTML]{EFEFEF}15.4 		& 2.39 	 	& \cellcolor[HTML]{EFEFEF}-1491 	& 0.77 		& \cellcolor[HTML]{EFEFEF}11.9 & 8.30 \\
- \toprule
-\end{tabular}
-\caption[Further look at algorithm differences for one beacon to beacon pairing]{A comparison of range estimation metrics for each sound speed source and group velocity estimation algorithm on all 10 communication events between North at 30 m \& South at 30 m.}
-\label{tab:rangeErrorPostExplanation}
-\end{table}
-
-The large error in this instance is driven by the MBC unexpectedly defaulting to a bottom bounce with a much greater OWTT.
-The NBC classifies the multipath as 4 bounces, reducing the range error from greater than a kilometer to less than a meter.
-While there is no actual way of knowing if this is the correct multipath structure, the range error is remarkably small, at 0.025\%. 
-This pattern of not choosing the minimal observed bounce structure is consistent across all SSPs; the baseline goes from 1 to 3 bounces and HYCOM goes from 0 to 2 bounces.
-Notably, the baseline and HYCOM range errors are never egregiously large, but are nonetheless improved with the NBC algorithm.
-Thus, for acoustically complex environments, the NBC has a disproportionately positive impact as the estimated SSP approaches the desired SSP.
-
-\FloatBarrier
-\subsection{GPS sensor drift at polar latitudes}
-
-Systemic GPS drift drives the severe banding found in comparisons of range estimations via both group velocity algorithms, as shown in Fig.
-\ref{fig:compareV1V2}, with those bands showing 100\% positive correlation between methods and being a similar size in meters.
-Thus, while this work is primarily motivated by achieving similar accuracy and precision of GPS, the impressive results from acoustic ranging necessitate a discussion about the limitations of GPS, particularly at polar latitudes.
-
-The ``Global Positioning System Standard Positioning Service Performance Standard'', published by the Department of Defense in Apr 2020 claims that well-designed GPS receivers have been achieving a horizontal accuracy of 3 meters and a vertical accuracy of 5 meters, 95\% of the time.
-But ``well-designed'' is an ambiguous enough term that obfuscates two confounding factors for high latitude GPS drift\textemdash operational limitations and ionospheric scattering.
-
-% operational
-There are two metrics that demonstrate the operational limitations of GPS tracking technologies\textemdash fix rate and positional dilution of precision (PDOP) \citep{swanlund_gps_2016}.
-Fix rate looks at the ratio of successful fixes to overall attempts in acquiring a location.
-In animal tracking studies in polar latitudes, fix rates have significantly improved in the last twenty years, from as low as 26\% in 1996 \citep{moen_effects_1996} to 87.2\% in 2018 \citep{jung_accuracy_2018}.
-The latter study showed a precision of 4.3 $\pm$ 4.0 m and an accuracy of 6.0 $\pm$ 4.7 m.
-While fix rate is a significant metric to evaluate GPS robustness, it is accuracy and precision, and not the fix rate, that explain the severe banding observed; fix rate bias, where various aspects of a landscape can induce bias by impeding the GPS receiver from acquiring its location, is not relevant given the static nature and small time window of ICEX20.
-
-PDOP looks at the effect of satellite geometry on location precision, as satellite orbits are designed to optimize performance for human population density.
-A high PDOP value indicates a poor geometric configuration, such as satellites being tightly clustered above a GPS receiver, that limits the reliability of trilateration.
-The literature shows correlation between PDOP values and location error \citep{swanlund_gps_2016}.
-PDOP is generally thought of as three-dimensional positioning and can be broken down into the horizontal and vertical components, HDOP and VDOP, respectively.
-The low elevation angles in polar areas are good for horizontal positioning but are susceptible to poorer vertical positioning.
-Overall this contributes to higher noise level in observations than usually seen in most global applications.
-The sparse and limited infrastructure in the Arctic make it ideal for satellite-based navigation, with some research proposing new satellite systems to accommodate increased interest in the region \citep{reid_gnss_2016}.
-
-However, both of these metrics are functions of the satellite constellation available to any given GPS receiver.
-The Arctic is beyond the reach of many geostationary satellites, where they are low on the horizon and only visible for brief periods.
-Given that their positions are tracked by stations on Earth, and their timing is updated twice a day via atomic clock, a reasonable hypothesis for bands of GPS error are different and/or updated satellite connections.
-
-% ionospheric
-The second confounding factor is the ionosphere, the layer of the Earth's atmosphere roughly 80 to 1000 km above the surface that contains a high concentration of ions and free elections. 
-In the ionosphere, electromagnetic signals are affected by the total amount of electrons encountered by the signal; the size of the signal delay is dependent on the carrier frequencies.
-The literature even references the total amount of electrons encountered by the ``GPS raypath'' as a means of explaining signal fluctuations \citep{themens_nature_2015}.
-This variability is observed by seasonal and solar cycle variabilities \citep{themens_nature_2015}, severe ionospheric storms \citep{mitchell_gps_2005}, and auroral activity (increased electron precipitation) \citep{jin_gps_2014,gwal_gps_2011}.
-In fact, the \textit{aurora borealis} was observed on multiple nights during ICEX20.
-
-In a humbling twist, at polar latitudes, GNSS systems experience similar operational, technical, and environmental challenges that we observe in underwater acoustic communications.
-Figure \ref{fig:gps-drift-example} highlights the GPS and observed OWTT drift relative to the ice movement for two pairs of modem buoy connections.
-The two panels indicate the GPS drift as $\delta R = \sqrt{\delta x^2 + \delta y^2}$ and temporal drift, $\delta t$, relative to the median OWTT recorded between the two modems.
-The dashed line is scaled by a group velocity of 1440 m/s, such that if there were ideal sensor measurements with no drift, all events should exist on or near the line.
+\clearpage
+\section{Investigating potential GNSS noise}
 
@@
 \end{figure}
-
-The top panel shows the connections between the North and East buoys.
-There is relative, i.e. non-rigid, ice movement between the North and East buoys, evidenced by events spanning the dashed line.
-But the height of the scatter plot is indicative of the precision of the GPS signal, as it remains consistent across many arrival time bands.
-Naturally, some minor offsets between these vertical bands relate to different operational configurations of source and receiver depth.
-However, the large majority of events show vertical banding for the same nominal $\delta t$, indicating the amount of GPS drift.
-
-This idea of GPS drift relative to the same OWTT measurements is further indicated by events between the other two buoys, South and West, in the bottom panel.
-These buoys are moving in a more rigid ice floe and there is minimal impact by source and receiver depth on the spread of OWTT.
-The GPS drift is much larger relative to the OWTT drift, which is sensitive to acoustic scattering, multipath, and/or environmental microstructure.
-
-These are just two subsets of the physical links that cover all four GPS modem buoys.
-The GPS at camp is the least accurate due to the human activity and infrastructure occluding the physical puck.
-
-% =========================================================================== %
-% =========================================================================== %
-\section{\label{sec:discussion} Discussion}
-
-Range estimation is an essential step of acoustic localization and navigation.
-Current approaches in real-time underwater acoustic navigation simplify the non-linear relationship between a sound speed profile and acoustic propagation with a deterministic sound speed.
-Some post-processing approaches attempt to leverage the acoustic arrival structure via various ray methods, but often use a singular SSP for simplicity, even over long term missions or dynamic conditions.
-Thus, the conversion from travel time to range, particularly for real-time vehicle deployments, can be pre-conditioned for error growth as the OWTT/range increases.
-
-For GPS-denied navigation, especially in hostile environments like the Arctic, tolerance for error is close to none.
-This work addresses a critical need in acoustic navigation by retooling acoustic arrival methods generally deemed too complex or labor intensive for real-time, ray-based range estimation to achieve GPS-like positioning.
-
-We hypothesize and validate that the embedded stochastic prediction of a single group velocity is a smoothly varying function of range, source and receiver depth pairings, as well as multipath structure.
-We employ a GPS-tracked experiment to have a ground truth comparison for real-time localization algorithms.
-The real-time system achieves GPS-like navigation for an AUV without taking into account multipath structure; the ranging error improves by an order of magnitude with the suggested multipath adaptability, minimizing range error to single meters.
-Post-processing analysis shows that this method of ranging is sensitive to GPS drift itself; thus embedding a stochastic prediction of the horizontal group velocity has an outsized benefit to minimizing trilateration error.
-
-There are many avenues through which this approach can be further refined and tested for field operations.
-Amongst them is defining the uncertainty grid for BELLHOP via stochastic or data-driven measures such as the distance traveled by the AUV between ICNN updates or the magnitude of the position corrections by the ICNN.
-Another is to entirely remove the seeded range and instead rely on the submerged asset's depth and recorded OWTT to find high probability fields in range.
-
-The literature in underwater acoustic navigation and positioning is either real-time or physics-based.
-In this paper we demonstrate a field-tested approach that is both real-time and physics-based; this is achieved by coupling data streams with fast acoustic modeling.
-The methods exploit the upward refracting nature and the total ice cover of the Arctic environment to achieve remarkable ranging accuracy and precision.
-It transforms multipath, widely considered as an obstacle for acoustic ranging, into a new information content to refine ranging accuracy.
-We believe that this work enables more accurate range estimation, localization, and/or navigation for any field experiment given known source and receiver depths.
-
-\subsection{Broader relevance \& future work}
-
-Performance in other acoustic environments may require introducing a different thresholded metric to sort time of arrivals.
-Given the NBC algorithm's improvement with increased multipath, its effectiveness is likely only challenged by the valid operational scales of a range independent propagation environment.
-For mesoscale operations, like that of many gliders, the group velocity criteria may need to be modified to better account for variability driven by range dependent propagation through internal waves, eddies, or even bathymetric changes like underwater volcanoes.
-BELLHOP simulations provide other relevant eigenray information, like time and angle of arrival, that is ripe for statistical and machine learning methods to classify a representative group velocity.
-A bespoke and fast ray tracing method, which was used during ICEX20 to evaluate the acoustic fit of sound speed profile parameterization \citep{bhatt_embedded_2021}, can easily report back the number of turning points instead of the number of bounces for multipath classification.
-
-This approach will start to break down in extremely dynamic environments.
-Fast moving fronts, as seen in estuaries like the Connecticut River into the Long Island Sound, present an entirely new set of challenges not seen by internal waves or eddies.
-Buoyancy fluctuations in these regions threaten even simple AUV tasks like following a trackline.
-Acoustic communications are further complicated given a shallow environment with significant scattering \citep{lavery_measurements_2010,ross_acoustic_2012,lavery_broadband_2013}, where fast acoustic modeling may only be coherent for trivial probabilities of the ocean state.
-Realistic \textit{in situ} considerations of the acoustic environment may not be possible in such environments without complete through-the-sensor integration of echosounder data and/or a hyper-realistic onboard ocean model.
-
-\llabel{1.20} Many approaches to underwater navigation combine it with acoustic tomography, i.e., a joint estimation of both source and receiver locations and the ocean volume between them.
-There has been considerable success at this effort in post-processing methods, which utilize intensive\textemdash and due to the non-linearity of sound propagation, often brute force\textemdash computational methods.
-For vehicle operations, fast tomography is the ideal implementation, in that one can fully consider how sound speed structure, horizontally and vertically, influences sound propagation.
-AUVs can serve as moving sources to better image the ocean volume \citep{deffenbaugh_optimal_1997,elisseeff_ocean_2002}, where mobile tomography and navigation converge on the same set of component technologies: position estimation, sound speed parameterization estimation, ray path identification, and vehicle path optimization.
-
-But there are overwhelming challenges, operationally and computationally, for fast, mobile tomography to become a realistic endeavor.
-Addressing the spatial and temporal scales of what can be solved deterministically and what must be solved stochastically imposes a resolution constraint on the utility of gridded models\textemdash resolving fine features inaccurately (or with a false sense of confidence) could be more harmful than assuming range independence.
-Given that AUV operations are often on small spatial and temporal scales, the added benefit of a gridded model is quite small, and in cases like the Arctic, may actually mischaracterize the ocean volume.
-For gliders, with longer and larger operational scales, an ocean model may provide more useful information.
-Currently gliders are low power and do not have the storage or computational power to run a full-scale, realistic ocean model.
-A lightweight representation of the key environmental and acoustic features, passed through the same manner of acoustic message from the modem experiment, may drastically improve glider navigation.
-
-The Arctic, a generally inaccessible, dangerous, and expensive region for crewed monitoring via icebreakers,  is a prime location to invest in vehicle deployments and underwater observatories \citep{mikhalevsky_multipurpose_2015}.
-As anthropogenic climate change renders the region more accessible, safe navigation infrastructure and practices will be increasingly vital for industry, tourism, commercial, and military operations.
-We hope this work enables more frequent and confident monitoring for remote and/or dangerous regions of high climate security importance.
 

### 2022-02-15 pipeline/1-overview/main_plot_isolate_zs.m
filepath fix

@@ -2,7 +2,7 @@
 % eeshan bhatt
 
 %% prep workspace
-clear; clc; close all
+clear; clc; close all; addpath('../../src');
 
 lg_font_size = 14;
 
@@ -16,11 +16,11 @@ zs = 90;
 load p_modemMarkerDetails
 
 %% load bathymetry data
-bathyfile = '../data/etopo1_bedrock.nc';
+bathyfile = '../../data/etopo1_bedrock.nc';
 plotBathy = h_unpack_bathy(bathyfile);
 
 %% load toby test data by experiment design
-A = load('../data/tobytest-recap-clean.mat'); % loads "event"
+A = load('../../data/tobytest-recap-clean.mat'); % loads "event"
 RECAP = h_unpack_experiment(A.event);
 
 %% figure : timeline

### 2022-02-15 pipeline/1-overview/main_plot_ssp_comparison_gvel.m

@@ -3,7 +3,7 @@
 
 %close all; clc; clear all;
 
-path = '../pipeline/bellhop-gvel-gridded/';
+path = '../bellhop-gvel-gridded/';
 
 file{3} = 'ssp-hycom.csv';
 file{2} = 'ssp-fixed-eeof.csv';

### 2022-02-15 pipeline/1-overview/main_plot_timeline_by_zs.m

@@ -14,7 +14,7 @@ markerSize = 145;
 load p_legendDetails
 
 %% load toby test data recap all
-load '../data/tobytest-recap-clean.mat'
+load '../../data/tobytest-recap-clean.mat'
 RECAP = h_unpack_experiment(event);
 
 figure('Name','timeline-by-zs','Renderer', 'painters', 'Position', [10 10 1200 650]); clf;

### 2022-02-15 pipeline/1-overview/plot_ssp_comparison.m
includes icex16 comparison for narrative

@@ -1,7 +1,13 @@
 %% main_plot_ssp_comparison.m
 % plots the EOF, baseval, and HYCOM comparison for the gvel calcluations
 
-%close all; clc; clear all;
+%% prep workspace
+close all; clc; clear all;
+
+%% load ICEX16
+load ../../data/icex16Comparison.mat
+
+%% load ICEX20
 
 path = '../bellhop-gvel-gridded/';
 
@@ -10,9 +16,10 @@ file{2} = 'ssp-fixed-eeof.csv';
 file{1} = 'ssp-fixed-baseval.csv';
 
 lineStyleSet = {'-','-','-'};
-lineWidthSet = [4 3 4];
+lineWidthSet = [4 4 4];
 colorSet = {[232, 153, 35]./256,[0 85 135]./256,[152 134 117]./256};
 
+%% plot
 figure('name','ssp-for-gvel','renderer','painters','position',[108 108 900 1000]);
 
 for k = 1:3
@@ -22,7 +29,14 @@ for k = 1:3
     hold off
 end
 
-% beautify plot
+hold on
+plot(NaN,NaN,'w');
+plot(historical.sspVal,historical.sspDepth,':','color',colorSet{1},'linewidth',4);
+plot(data.sspVal,data.sspDepth,':','color',colorSet{2},'linewidth',4);
+plot(hycom.sspVal,hycom.sspDepth,':','color',colorSet{3},'linewidth',4);
+hold off
+
+%% beautify plot
 grid on
 set(gca,'ydir','reverse');
 title('Sound speed estimates');
@@ -31,6 +45,9 @@ xlim([1431 1462])
 ylabel('depth [m]');
 xlabel('c [m/s]');
 
-legend('Baseline','Chosen Weights','HYCOM','location','southwest','fontsize',13);
+legend('ICEX20: Baseline','ICEX20: Chosen Weights','ICEX20: HYCOM','',...
+       'ICEX16: Historical', 'ICEX16: Data','ICEX16: HYCOM',...
+       'location','southwest','fontsize',13);
 
-% h_printThesisPNG('ssp-gvel-large');
+%% export
+h_printThesisPNG('ssp-gvel-icex20-icex16');

### 2022-02-15 pipeline/1-overview/plot_ssp_comparison.m
added calculation for isovelocity case

@@ -48,6 +48,16 @@ xlabel('c [m/s]');
 legend('ICEX20: Baseline','ICEX20: Chosen Weights','ICEX20: HYCOM','',...
        'ICEX16: Historical', 'ICEX16: Data','ICEX16: HYCOM',...
        'location','southwest','fontsize',13);
+   
+%% export "isovelocity" assumption
+
+T = readtable([path file{2}]);
+ssp = T.Var2;
+z = T.Var1;
+
+ind200m = find(z>=200,1,'first');
+meanSSP = mean(ssp(1:ind200m));
+
 
 %% export
-h_printThesisPNG('ssp-gvel-icex20-icex16');
+% h_printThesisPNG('ssp-gvel-icex20-icex16');

### 2022-02-15 pipeline/1-overview/plot_ssp_comparison.m
added mean ssp +/- std ssp for isovelocity case

@@ -57,6 +57,9 @@ z = T.Var1;
 
 ind200m = find(z>=200,1,'first');
 meanSSP = mean(ssp(1:ind200m));
+stdSSP = std(ssp(1:ind200m));
+
+fprintf('Isovelocity SSP = %4.1f +/- %4.1f m/s \n',meanSSP,stdSSP)
 
 
 %% export

### 2022-02-15 pipeline/3-analyze-gvel/slides_gvel_comparison.m
helper flow for pipeline3

@@ -3,85 +3,21 @@
 % predictions
 
 %% prep workspace
-clear; clc; close all;
+clear; clc; close all; addpath('../../src/');
 
-% load data
-DATA = readtable('../bellhop-gvel-gridded/gveltable.csv');
-% only simGvel
-DATA.gvel = DATA.recRange ./ DATA.owtt;
-
-DATA.simGvel(isnan(DATA.simGvel)) = 0;
-
-% remove crazy 11 second event, event that is nominally 1.58* seconds
-indBad1 = find(DATA.owtt > 4);
-indBad2 = find(strcmp(DATA.rxNode,'East') & DATA.owtt > 1.55);
-indBad3 = find(strcmp(DATA.rxNode,'Camp'));
-indBad = union(indBad1,indBad2);
-indBad = union(indBad,indBad3);
-
-% 1.587 events, had clock errors + Bellhop can't resolve these
-DATA.simGvel(indBad) = NaN;
-% only simGvel
-
-indValid = ~isnan(DATA.simGvel);
-
-% calculate RangeAnomaly
-DATA.rangeAnomaly = DATA.owtt .* DATA.simGvel - DATA.recRange;
+% unpack Bellhop gvel table
+[DATA,INDEX] = h_unpack_bellhop('../bellhop-gvel-gridded/gveltable.csv');
 
 % DEPTH
 ZS = 90;
 
-%% load post-processing sim for v2
-listing = dir('../bellhop-gvel-gridded/csv_arr/*gridded.csv');
+%% load post-processing sim for NBC
+listing2 = dir('../bellhop-gvel-gridded/csv_arr/*gridded.csv');
+[T2,colorSet] = h_get_nbc(listing2,DATA,INDEX);
 
-for k = 1:numel(listing)
-    T0 = readtable([listing(k).folder '/' listing(k).name]);
-    T0.index = T0.index + 1;
-    b = split(listing(k).name,'.');
-    tName{k} = b{1};
-    
-    % assign gvel for each index by closest time comparison
-    for j = 1:numel(T0.index)
-        delay = DATA.owtt(j);
-        tableDelay = table2array(T0(j,2:6));
-        [~,here] = min(abs(tableDelay - delay));
-        T0.gvel(j) = DATA.recRange(j)./tableDelay(here);
-        T0.owtt(j) = tableDelay(here);
-        T0.numBounces(j) = here-1;
-        if sum(j == indBad) == 1
[…]

### 2022-02-15 src/h_get_mbc.m

@@ -1,0 +1,26 @@
+function [T] = h_get_mbc(listing,DATA)
+%h_get_mbc computes the minimal bounce criterion from processed Bellhop files
+%   T is indexed by Bellhop SSP environment
+%   1 = artifact-baseval
+%   2 = artifact-eeof
+%   3 = fixed-baseval
+%   4 = fixed-eeof
+%   5 = hycom
+
+for f = 1:numel(listing)
+    T0 = readtable([listing(f).folder '/' listing(f).name]);
+    T0.index = T0.index + 1;
+    b = split(listing(f).name,'.');
+    tName{f} = b{1};
+    
+    % assign gvel by minimum bounce
+    for k = 1:numel(T0.index)
+        T0.gvel(k) = DATA.recRange(k)./T0.owtt(k);
+    end
+    
+    T0.rangeAnomaly = DATA.owtt .* T0.gvel - DATA.recRange;
+    T{f} = T0;
+end
+
+end
+

### 2022-02-15 src/h_get_nbc.m

@@ -1,0 +1,37 @@
+function [T,colorSet] = h_get_nbc(listing,DATA,INDEX)
+%h_get_nbc computes the nearest bounce criterion from processed Bellhop files
+%   T is indexed by Bellhop SSP environment
+%   1 = artifact-baseval
+%   2 = artifact-eeof
+%   3 = fixed-baseval
+%   4 = fixed-eeof
+%   5 = hycom
+
+indBad = INDEX.bad;
+
+for k = 1:numel(listing)
+    T0 = readtable([listing(k).folder '/' listing(k).name]);
+    T0.index = T0.index + 1;
+    b = split(listing(k).name,'.');
+    tName{k} = b{1};
+    
+    % assign gvel for each index by closest time comparison
+    for j = 1:numel(T0.index)
+        delay = DATA.owtt(j);
+        tableDelay = table2array(T0(j,2:6));
+        [~,here] = min(abs(tableDelay - delay));
+        T0.gvel(j) = DATA.recRange(j)./tableDelay(here);
+        T0.owtt(j) = tableDelay(here);
+        T0.numBounces(j) = here-1;
+        if sum(j == indBad) == 1
+            T0.gvel(j) = NaN;
+        end
+    end
+    T{k} = T0;
+end
+
+load p_sspColorDetails;
+colorSet = {[0 0 0],[0 0 0],colorSet{:}};
+
+end
+

### 2022-02-15 src/h_unpack_bellhop.m

@@ -1,0 +1,30 @@
+function [DATA,INDEX] = h_unpack_bellhop(filepath)
+%h_unpack_bellhop loads & filters BELLHOP runs
+
+% load data
+DATA = readtable(filepath);
+% only simGvel
+DATA.gvel = DATA.recRange ./ DATA.owtt;
+
+DATA.simGvel(isnan(DATA.simGvel)) = 0;
+
+% remove crazy 11 second event, event that is nominally 1.58* seconds
+indBad1 = find(DATA.owtt > 4);
+indBad2 = find(strcmp(DATA.rxNode,'East') & DATA.owtt > 1.55);
+indBad3 = find(strcmp(DATA.rxNode,'Camp'));
+indBad = union(indBad1,indBad2);
+indBad = union(indBad,indBad3);
+
+% 1.587 events, had clock errors
+DATA.simGvel(indBad) = NaN;
+% only simGvel
+
+indValid = ~isnan(DATA.simGvel);
+
+INDEX.valid = indValid;
+INDEX.bad   = indBad;
+
+% calculate RangeAnomaly
+DATA.rangeAnomaly = DATA.owtt .* DATA.simGvel - DATA.recRange;
+end
+

### 2022-02-15 pipeline/4-analyze-range-error/plot_insitu_range_error_by_modem.m
clean up pipeline4

@@ -1,19 +1,16 @@
-%% gvel_1_data
-% makes plots to explore *data* from ICEX20
-
 %% prep workspace
 clear; clc; close all;
 
 % load data
-A = readtable('../pipeline/bellhop-gvel-gridded/gveltable.csv');
+DATA = readtable('../bellhop-gvel-gridded/gveltable.csv');
 
 % remove crazy 11 second event, event that is nominally 1.58* seconds
-indBad1 = find(A.owtt > 4);
-indBad2 = find(strcmp(A.rxNode,'East') & A.owtt > 1.55);
+indBad1 = find(DATA.owtt > 4);
+indBad2 = find(strcmp(DATA.rxNode,'East') & DATA.owtt > 1.55);
 indBad = union(indBad1,indBad2);
 
-% 1.587 events, had clock errors + Bellhop can't resolve these
-A.simGvel(indBad) = NaN;
+% 1.587 events, had clock errors
+DATA.simGvel(indBad) = NaN;
 
 % load modem marker information
 load p_modemMarkerDetails
@@ -23,13 +20,13 @@ load p_modemMarkerDetails
 figure('name','rangeanomaly-by-owtt','renderer','painters','position',[108 108 1300 800]);
 t = tiledlayout(2,3,'TileSpacing','none','Padding','compact');
 
-index3 = ~isnan(A.simGvel);
+index3 = ~isnan(DATA.simGvel);
 count = 0;
 for zr = [30 90]
-    index2 = A.recDepth == zr;
+    index2 = DATA.recDepth == zr;
     
     for zs = [20 30 90]
-        index1 = A.sourceDepth == zs;
+        index1 = DATA.sourceDepth == zs;
         
         count = count + 1;
         nexttile;
@@ -41,8 +38,8 @@ for zr = [30 90]
             hold on
             plot([0 4],[0 0],'--','linewidth',3,'color',[0 0 0 0.6],'handlevisibility','off');
             
-            xval = A.owtt(index);
-            yval = A.simGvel(index) .* A.owtt(index) - A.recRange(index);
+            xval = DATA.owtt(index);
+            yval = DATA.simGvel(index) .* DATA.owtt(index) - DATA.recRange(index);
             
             % remove nans
             xval = xval(~isnan(yval));
@@ -61,8 +58,8 @@ for zr = [30 90]
             
             plotIndex = find(index == 1);
             for k = plotIndex.'
[…]

### 2022-02-15 pipeline/4-analyze-range-error/plot_insitu_range_error_by_ssp.m

@@ -1,22 +1,12 @@
-%% plot_data_range_error_v2.m
-
+%% prep workspace
 clear; clc; close all;
 
 %% load data
-A = load('../data/tobytest-recap-clean.mat'); % loads "event"
+A = load('../../data/tobytest-recap-clean.mat'); % loads "event"
 A = h_unpack_experiment(A.event);
+A = h_clean_simgvel(A);
 
-% remove crazy 11 second event, event that is nominally 1.58* seconds
-indBad1 = A.data_owtt > 3;
-indBad2 = strcmp(A.tag_rx,'East') & A.data_owtt > 1.55;
-indBad = indBad1 | indBad2;
-
-% 1.587 events, had clock errors + Bellhop can't resolve these
-A.sim_gvel(indBad) = NaN;
-
-colorSet = {[232, 153, 35]./256,[0 85 135]./256,[152 134 117]./256};
-
-
+load p_sspColorDetails
 %% figure 1 -- all events in 3x2 grid
 
 figure('name','rangeanomaly-by-owtt','renderer','painters','position',[108 108 1100 900]);
@@ -120,7 +110,7 @@ legend(l2,'Baseline','Chosen Weights','location','northwest');
 
 % title
 sgtitle('In situ range error by source (20,30,90 m) and receiver (30,90 m) depths','fontsize',17,'fontweight','bold')
-h_printThesisPNG('range-error-owtt-data-v2');
+% h_printThesisPNG('range-error-owtt-data-v2');
 
 %% find mean, median, etc for baseval vs eof
 

### 2022-02-15 pipeline/4-analyze-range-error/plot_post_range_error_nbc.m

@@ -1,61 +1,13 @@
 %% prep workspace
 clear; clc; close all;
 
-% load data
-DATA = readtable('../pipeline/bellhop-gvel-gridded/gveltable.csv');
-% only simGvel
-DATA.gvel = DATA.recRange ./ DATA.owtt;
-
-DATA.simGvel(isnan(DATA.simGvel)) = 0;
-
-% remove crazy 11 second event, event that is nominally 1.58* seconds
-indBad1 = find(DATA.owtt > 4);
-indBad2 = find(strcmp(DATA.rxNode,'East') & DATA.owtt > 1.55);
-indBad3 = find(strcmp(DATA.rxNode,'Camp'));
-indBad = union(indBad1,indBad2);
-indBad = union(indBad,indBad3);
-
-% 1.587 events, had clock errors + Bellhop can't resolve these
-DATA.simGvel(indBad) = NaN;
-% only simGvel
-
-indValid = ~isnan(DATA.simGvel);
-
-% calculate RangeAnomaly
-DATA.rangeAnomaly = DATA.owtt .* DATA.simGvel - DATA.recRange;
+%% unpack Bellhop gvel table
+[DATA,INDEX] = h_unpack_bellhop('../bellhop-gvel-gridded/gveltable.csv');
 
 %% load simulation
-listing = dir('../pipeline/bellhop-gvel-gridded/csv_arr/*gridded.csv');
+listing = dir('../bellhop-gvel-gridded/csv_arr/*gridded.csv');
+[T,colorSet] = h_get_nbc(listing,DATA,INDEX);
 
-for k = 1:numel(listing)
-    T0 = readtable([listing(k).folder '/' listing(k).name]);
-    T0.index = T0.index + 1;
-    b = split(listing(k).name,'.');
-    tName{k} = b{1};
-    
-    % assign gvel for each index by closest time comparison
-    for j = 1:numel(T0.index)
-        delay = DATA.owtt(j);
-        tableDelay = table2array(T0(j,2:6));
-        [~,here] = min(abs(tableDelay - delay));
-        T0.gvel(j) = DATA.recRange(j)./tableDelay(here);
-        T0.owtt(j) = tableDelay(here);
-        T0.numBounces(j) = here-1;
-        if sum(j == indBad) == 1
-            T0.gvel(j) = NaN;
-        end
-    end
-    T{k} = T0;
-end
-
-% 1 = artifact-baseval
-% 2 = artifact-eeof
-% 3 = fixed-baseval
-% 4 = fixed-eeof
[…]

### 2022-02-15 pipeline/5-analyze-postprocessing/analyze_hist_nbc.m
pipeline5, 4 plots in 1 script

@@ -1,64 +1,45 @@
-%% gvel_3_simAllEvents
-
 %% prep workspace
 clear; clc; close all;
 
-% load data
-DATA = readtable('../bellhop-gvel-gridded/gveltable.csv');
-% only simGvel
-DATA.gvel = DATA.recRange ./ DATA.owtt;
+%% load data
 
-DATA.simGvel(isnan(DATA.simGvel)) = 0;
+[DATA,INDEX] = h_unpack_bellhop('../bellhop-gvel-gridded/gveltable.csv');
 
-% remove crazy 11 second event, event that is nominally 1.58* seconds
-indBad1 = find(DATA.owtt > 4);
-indBad2 = find(strcmp(DATA.rxNode,'East') & DATA.owtt > 1.55);
-indBad3 = find(strcmp(DATA.rxNode,'Camp'));
-indBad = union(indBad1,indBad2);
-indBad = union(indBad,indBad3);
-
-% 1.587 events, had clock errors + Bellhop can't resolve these
-DATA.simGvel(indBad) = NaN;
-% only simGvel
-
-indValid = ~isnan(DATA.simGvel);
-
-% calculate RangeAnomaly
-DATA.rangeAnomaly = DATA.owtt .* DATA.simGvel - DATA.recRange;
 %% load simulation
 listing = dir('../bellhop-gvel-gridded/csv_arr/*gridded.csv');
+[T,colorSet] = h_get_nbc(listing,DATA,INDEX);
 
-for k = 1:numel(listing)
-    T0 = readtable([listing(k).folder '/' listing(k).name]);
-    T0.index = T0.index + 1;
-    b = split(listing(k).name,'.');
-    tName{k} = b{1};
-    
-    % assign gvel for each index by closest time comparison
-    for j = 1:numel(T0.index)
-        delay = DATA.owtt(j);
-        tableDelay = table2array(T0(j,2:6));
-        [~,here] = min(abs(tableDelay - delay));
-        T0.gvel(j) = DATA.recRange(j)./tableDelay(here);
-        T0.owtt(j) = tableDelay(here);
-        T0.numBounces(j) = here-1;
-        if sum(j == indBad) == 1
-            T0.gvel(j) = NaN;
-        end
-    end
-    T{k} = T0;
-end
-
-% 1 = artifact-baseval
-% 2 = artifact-eeof
-% 3 = fixed-baseval
-% 4 = fixed-eeof
-% 5 = hycom
[…]

### 2022-02-15 pipeline/6-compare-insitu-post/gvel_4_compare_inSitu_postProcessed.m
remove compare in situ v post

@@ -1,198 +1,0 @@
-%% gvel_4_compare_inSitu_postProcessed
-% compares range anomaly performance between in situ estimation & post
-% processing of *NEW* algorithm
-
-clear; clc; close all;
-
-%% load in situ data
-% DATA = readtable('./bellhop-gvel-gridded-simRange/gveltable.csv');
-
-DATA = readtable('../bellhop-gvel-gridded-simRange/gveltable.csv');
-indValid1 = ~isnan(DATA.recRange) | ~isnan(DATA.simGvel);
-DATA = DATA(indValid1,:);
-
-A = load('../../data/tobytest-recap-clean.mat'); % loads "event"
-RECAP = h_unpack_experiment(A.event);
-indValid2 = ~isnan(RECAP.sim_gvel) | ~isnan(RECAP.sim_range);
-
-EOF_BOOL = double(RECAP.eof_bool(indValid2)).';
-GPS_RANGE = RECAP.data_range(indValid2).';
-SIM_RANGE = RECAP.sim_range(indValid2).';
-
-%% remove weird event
-indBad1 = find(DATA.owtt > 4);
-indBad2 = find(strcmp(DATA.rxNode,'East') & DATA.owtt > 1.55);
-indBad = union(indBad1,indBad2);
-
-% 1.587 events, had clock errors + Bellhop can't resolve these
-DATA.recRange(indBad) = NaN;
-DATA.simGvel(indBad) = NaN;
-
-EOF_BOOL(indBad) = NaN;
-
-%% range error for "data" aka ICEX20 implementation
-DATA.rangeAnomaly = DATA.owtt .* DATA.simGvel - GPS_RANGE;
-
-%% range error for "sim" aka post processing nearest bounce criterion
-listing = dir('../bellhop-gvel-gridded-simRange/csv_arr/*gridded.csv');
-
-for f = 1:numel(listing)
-    T0 = readtable([listing(f).folder '/' listing(f).name]);
-    T0.index = T0.index + 1;
-    b = split(listing(f).name,'.');
-    tName{f} = b{1};
-    
-    % assign gvel for each index by closest time comparison
-    for k = 1:numel(T0.index)
-        delay = DATA.owtt(k);
-        tableDelay = table2array(T0(k,2:6));
-        [~,here] = min(abs(tableDelay - delay));
[…]

### 2022-02-15 pipeline/7-compare-post-mbc-nbc/plot_crossmap_post_mbc_nbc.m
pipeline for 7, includes range anomaly calc in h_get_nbc

@@ -1,79 +1,17 @@
-%% gvel_5_compare_inSitu_postProcessed
-% compares range anomaly performance between post processing of old and new
-% algorithms
-
+%% prep workspace
 clear; clc; close all;
 
 %% load in situ data
-DATA = readtable('../bellhop-gvel-gridded/gveltable.csv');
-A = load('../../data/tobytest-recap-clean.mat'); % loads "event"
-RECAP = h_unpack_experiment(A.event);
-%% 
-DATA.simGvel(isnan(DATA.simGvel)) = 0;
+[DATA,INDEX] = h_unpack_bellhop('../bellhop-gvel-gridded/gveltable.csv');
+indValid = INDEX.valid;
 
-% remove crazy 11 second event, event that is nominally 1.58* seconds
-indBad1 = find(DATA.owtt > 4);
-indBad2 = find(strcmp(DATA.rxNode,'East') & DATA.owtt > 1.55);
-indBad3 = find(strcmp(DATA.rxNode,'Camp'));
-indBad = union(indBad1,indBad2);
-indBad = union(indBad,indBad3);
+%% calculate minimal bounce critera
+listingOld = dir('../bellhop-gvel-gridded/csv_arr/*old.csv');
+SIM_OLD = h_get_mbc(listingOld,DATA);
 
-% 1.587 events, had clock errors + Bellhop can't resolve these
-DATA.simGvel(indBad) = NaN;
-% only simGvel
-
-indValid = ~isnan(DATA.simGvel);
-
-% calculate RangeAnomaly
-DATA.rangeAnomaly = DATA.owtt .* DATA.simGvel - DATA.recRange;
-
-%% load post-processing, new algorithm
-listing = dir('../bellhop-gvel-gridded/csv_arr/*gridded.csv');
-
-for f = 1:numel(listing)
-    T0 = readtable([listing(f).folder '/' listing(f).name]);
-    T0.index = T0.index + 1;
-    b = split(listing(f).name,'.');
-    tName{f} = b{1};
-    
-    % assign gvel for each index by closest time comparison
-    for k = 1:numel(T0.index)
-        delay = DATA.owtt(k);
-        tableDelay = table2array(T0(k,2:6));
-        [~,here] = min(abs(tableDelay - delay));
-        T0.gvel(k) = DATA.recRange(k)./tableDelay(here);
[…]

### 2022-02-15 pipeline/7-compare-post-mbc-nbc/slides_post_rangeError_crossmap.m

@@ -1,81 +1,19 @@
 %% slides_post_rangeError_crossmap.m
 
-%% gvel_5_compare_inSitu_postProcessed
-% compares range anomaly performance between post processing of old and new
-% algorithms
-
+% prep workspace
 clear; clc; close all;
 
 %% load in situ data
-DATA = readtable('../pipeline/bellhop-gvel-gridded/gveltable.csv');
-A = load('../data/tobytest-recap-clean.mat'); % loads "event"
-RECAP = h_unpack_experiment(A.event);
+[DATA,INDEX] = h_unpack_bellhop('../bellhop-gvel-gridded/gveltable.csv');
+indValid = INDEX.valid;
 
-DATA.simGvel(isnan(DATA.simGvel)) = 0;
+%% calculate minimal bounce critera
+listingOld = dir('../bellhop-gvel-gridded/csv_arr/*old.csv');
+SIM_OLD = h_get_mbc(listingOld,DATA);
 
-% remove crazy 11 second event, event that is nominally 1.58* seconds
-indBad1 = find(DATA.owtt > 4);
-indBad2 = find(strcmp(DATA.rxNode,'East') & DATA.owtt > 1.55);
-indBad3 = find(strcmp(DATA.rxNode,'Camp'));
-indBad = union(indBad1,indBad2);
-indBad = union(indBad,indBad3);
-
-% 1.587 events, had clock errors + Bellhop can't resolve these
-DATA.simGvel(indBad) = NaN;
-% only simGvel
-
-indValid = ~isnan(DATA.simGvel);
-
-% calculate RangeAnomaly
-DATA.rangeAnomaly = DATA.owtt .* DATA.simGvel - DATA.recRange;
-
-%% load post-processing, new algorithm
-listing = dir('../pipeline/bellhop-gvel-gridded/csv_arr/*gridded.csv');
-
-for f = 1:numel(listing)
-    T0 = readtable([listing(f).folder '/' listing(f).name]);
-    T0.index = T0.index + 1;
-    b = split(listing(f).name,'.');
-    tName{f} = b{1};
-    
-    % assign gvel for each index by closest time comparison
-    for k = 1:numel(T0.index)
-        delay = DATA.owtt(k);
-        tableDelay = table2array(T0(k,2:6));
-        [~,here] = min(abs(tableDelay - delay));
[…]

### 2022-02-15 src/h_get_nbc.m

@@ -27,8 +27,12 @@ for k = 1:numel(listing)
             T0.gvel(j) = NaN;
         end
     end
+    
+    T0.rangeAnomaly = DATA.owtt .* T0.gvel - DATA.recRange;
+
     T{k} = T0;
 end
+
 
 load p_sspColorDetails;
 colorSet = {[0 0 0],[0 0 0],colorSet{:}};

### 2022-02-15 pipeline/5-analyze-postprocessing/analyze_hist_nbc.m
range anomaly included in helper function

@@ -1,6 +1,9 @@
 %% prep workspace
 clear; clc; close all;
 
+% load modem marker information
+load p_modemMarkerDetails
+
 %% load data
 
 [DATA,INDEX] = h_unpack_bellhop('../bellhop-gvel-gridded/gveltable.csv');
@@ -8,38 +11,6 @@ clear; clc; close all;
 %% load simulation
 listing = dir('../bellhop-gvel-gridded/csv_arr/*gridded.csv');
 [T,colorSet] = h_get_nbc(listing,DATA,INDEX);
-
-% for k = 1:numel(listing)
-%     T0 = readtable([listing(k).folder '/' listing(k).name]);
-%     T0.index = T0.index + 1;
-%     b = split(listing(k).name,'.');
-%     tName{k} = b{1};
-%     
-%     % assign gvel for each index by closest time comparison
-%     for j = 1:numel(T0.index)
-%         delay = DATA.owtt(j);
-%         tableDelay = table2array(T0(j,2:6));
-%         [~,here] = min(abs(tableDelay - delay));
-%         T0.gvel(j) = DATA.recRange(j)./tableDelay(here);
-%         T0.owtt(j) = tableDelay(here);
-%         T0.numBounces(j) = here-1;
-%         if sum(j == INDEX.bad) == 1
-%             T0.gvel(j) = NaN;
-%         end
-%     end
-%     T{k} = T0;
-% end
-% 
-% % 1 = artifact-baseval
-% % 2 = artifact-eeof
-% % 3 = fixed-baseval
-% % 4 = fixed-eeof
-% % 5 = hycom
-
-% load modem marker information
-load p_modemMarkerDetails
-
- % colorSet = {[0 0 0],[0 0 0],[232, 153, 35]./256,[0 85 135]./256,[152 134 117]./256};
 
 %% plot all data group velocity
 
@@ -237,7 +208,7 @@ edges = [-14:2:22];
 count = 0;
 for s = [5 3 4]
     count = count + 1;
-    T{s}.rangeAnomaly = T{s}.gvel .* DATA.owtt - DATA.recRange;
+    %T{s}.rangeAnomaly = T{s}.gvel .* DATA.owtt - DATA.recRange;
     
     h(count,:) = histcounts(T{s}.rangeAnomaly,edges,'normalization','count');
 end
@@ -301,9 +272,9 @@ count = 0;
 for s = [5 3 4]
     count = count + 1;
[…]

### 2022-02-15 pipeline/7-analyze-anomaly-event/statistics_weird_event.m
finished pipeline cleaning

@@ -1,0 +1,38 @@
+%% prep workspace
+
+clear; clc; close all;
+
+%% load in situ data
+[DATA,INDEX] = h_unpack_bellhop('../bellhop-gvel-gridded/gveltable.csv');
+
+%% load post-processing, NBC
+listingNew = dir('../bellhop-gvel-gridded/csv_arr/*gridded.csv');
+SIM_NEW = h_get_nbc(listingNew,DATA,INDEX);
+
+%% load post-processing, old algorithm
+listingOld = dir('../bellhop-gvel-gridded/csv_arr/*old.csv');
+SIM_OLD = h_get_mbc(listingOld,DATA);
+
+%% calculate statistics
+
+basevalStat = h_calc_step(SIM_OLD,SIM_NEW,3,INDEX.valid);
+
+eeofStat = h_calc_step(SIM_OLD,SIM_NEW,4,INDEX.valid);
+
+hycomStat = h_calc_step(SIM_OLD,SIM_NEW,5,INDEX.valid);
+
+%% helper function
+
+function [F] = h_calc_step(SIM_OLD,SIM_NEW,fileIndex,indValid)
+
+F.owtt_1 = mean(SIM_OLD{fileIndex}.owtt(indValid));
+F.gvel_1 = mean(SIM_OLD{fileIndex}.gvel(indValid));
+F.rang_1 = mean(SIM_OLD{fileIndex}.rangeAnomaly(indValid));
+F.nbnc_1 = SIM_OLD{fileIndex}.n_bnc(indValid).';
+
+F.owtt_2 = mean(SIM_NEW{fileIndex}.owtt(indValid));
+F.gvel_2 = mean(SIM_NEW{fileIndex}.gvel(indValid));
+F.rang_2 = mean(SIM_NEW{fileIndex}.rangeAnomaly(indValid));
+F.nbnc_2 = SIM_NEW{fileIndex}.numBounces(indValid).';
+
+end

### 2022-02-15 pipeline/8-analyze-anomaly-event/gvel_6_postprocessing_weird_event.m

@@ -1,96 +1,0 @@
-%% gvel_6_compare_inSitu_postProcessed
-% compares range anomaly performance between post processing of old and new
-% algorithms
-
-clear; clc;
-
-%% load in situ data
-DATA = readtable('../bellhop-gvel-gridded/gveltable.csv');
-A = load('../../data/tobytest-recap-clean.mat'); % loads "event"
-RECAP = h_unpack_experiment(A.event);
-
-DATA.simGvel(isnan(DATA.simGvel)) = 0;
-
-% remove crazy 11 second event, event that is nominally 1.58* seconds
-indBad1 = find(DATA.owtt > 3);
-indBad2 = find(strcmp(DATA.rxNode,'East') & DATA.owtt > 1.55);
-indBad = union(indBad1,indBad2);
-
-% 1.587 events, had clock errors + Bellhop can't resolve these
-DATA.simGvel(indBad) = NaN;
-% only simGvel
-
-indValid = 1215:1224;
-
-% calculate RangeAnomaly
-DATA.rangeAnomaly = DATA.owtt .* DATA.simGvel - DATA.recRange;
-
-%% load post-processing, new algorithm
-listing = dir('../bellhop-gvel-gridded/csv_arr/*gridded.csv');
-
-for f = 1:numel(listing)
-    T0 = readtable([listing(f).folder '/' listing(f).name]);
-    T0.index = T0.index + 1;
-    b = split(listing(f).name,'.');
-    tName{f} = b{1};
-    
-    % assign gvel for each index by closest time comparison
-    for k = 1:numel(T0.index)
-        delay = DATA.owtt(k);
-        tableDelay = table2array(T0(k,2:6));
-        [~,here] = min(abs(tableDelay - delay));
-        T0.gvel(k) = DATA.recRange(k)./tableDelay(here);
-        T0.owtt(k) = tableDelay(here);
-        T0.numBounces(k) = here-1;
-    end
-    
-    T0.rangeAnomaly = DATA.owtt .* T0.gvel - DATA.recRange;
-    SIM_NEW{f} = T0;
-end
-
-%% load post-processing, old algorithm
-listing = dir('../bellhop-gvel-gridded/csv_arr/*old.csv');
-
[…]

### 2022-02-17 pipeline/3-analyze-gvel/compare_gvel_gps_mbc_nbc.m
printing figs

@@ -132,7 +132,7 @@ end
 % legend for data
 plot(NaN,NaN,'w');
 plot(NaN,NaN,'w');
-plot(NaN,NaN,'.','color',,'markersize',15);
+plot(NaN,NaN,'.','color','markersize',15);
 
 lgdstr = {'\bf{SOUND SPEED SOURCE}','HYCOM','Baseline','Chosen Weights','','\bf{MINIMAL BOUNCE CRITERIA}',...
     'Minimal bounce','','\bf{NEAREST BOUNCE CRITERIA}',...
@@ -147,4 +147,4 @@ titlestr = sprintf('Group velocity predictions for a source depth = %u m',zs);
 sgtitle(titlestr,'fontsize',17,'fontweight','bold')
 
 %%
-% h_printThesisPNG(sprintf('SLIDES-gvel-comparison-wdata-%u',ZS));
+% h_printThesisPNG(sprintf(SLIDES-gvel-comparison-wdata-%u',ZS));

### 2022-02-17 pipeline/6-compare-post-mbc-nbc/plot_crossmap_post_mbc_nbc.m

@@ -50,7 +50,7 @@ eeof.F = h_cross_plot(eeof.xVal,eeof.yVal,eeof.zs,eeof.numBounces);
 
 nexttile;
 h_view_plot(eeof.xVal,eeof.yVal,eeof.zs,eeof.numBounces);
-title({sprintf('\\fontsize{16} Post-processed range estimation comparison for all %u beacon to beacon events',sum(indValid)),'\fontsize{13} SSP = Chosen Weights'})
+title({sprintf('\\fontsize{16} Pseudorange error comparison for all %u beacon to beacon events',sum(indValid)),'\fontsize{13} SSP = Chosen Weights'})
 yticklabels([]);
 xlabel({'minimal bounce criterion error [m]'});
 
@@ -87,12 +87,12 @@ end
 hold off
 lgdstr = {' 20 m',' 30 m',' 90 m','','','direct path','1 bounce','2 bounces','3 bounces','4 bounces'};
 
-lg1 = legend(lgdstr,'location','southwest','NumColumns',2,'fontsize',10);
+lg1 = legend(lgdstr,'location','south','NumColumns',2,'fontsize',10);
 title(lg1,'   source depth & multipath structure');
 
 %% export
 
-% h_printThesisPNG('compare-methods-postv1v2');
+h_printThesisPNG('compare-methods-postv1v2');
 
 %% figure helper function
 function [F] = h_cross_plot(xVal,yVal,zs,numBounces)

### 2022-02-18 manuscript/main.tex
figure layout

@@
 	\centering
-	\includegraphics[width=\reprintcolumnwidth]{figs/ssp-gvel-large.pdf}
+	\includegraphics[width=\reprintcolumnwidth]{figs/ssp-gvel-icex20-icex16.pdf}
 	\caption{Anticipated sound speed conditions: finish this using stuff from ICEX16, HYCOM, ITP}
@@
 
-\subsection{Discussion}
+\subsection{Inherent overestimation from the minimal bounce criteria}
 
-\clearpage
-\section{\label{sec:post} Post-processed pseudorange analysis}
+\subsubsection{Source depth of 20 m}
+\begin{figure}[h!]
+  \centering
+  \includegraphics[width=\reprintcolumnwidth]{figs/Fig4.pdf}
+  \caption{Eigenrays for beacon to beacon events for each sound speed with a nominal source depth of 30 m.
+The beacons are highlighted in color/marker coding in Fig.
+\ref{fig:overview}.
+The eigenrays are curated from BELLHOP by travel time proximity and are traced in the representative receiver colors over a total ray fan in gray.}
+  \label{fig:raytrace-zs20}
+\end{figure}
 
-\subsection{Inherent overestimation from the minimal bounce criteria}
+\subsubsection{Source depth of 30 m}
 \begin{figure}[h!]
@@
 \end{figure}
+
+\subsubsection{Source depth of 90 m}
+\begin{figure}[h!]
+  \centering
+  \includegraphics[width=\reprintcolumnwidth]{figs/Fig4.pdf}
+  \caption{Eigenrays for beacon to beacon events for each sound speed with a nominal source depth of 30 m.
+The beacons are highlighted in color/marker coding in Fig.
+\ref{fig:overview}.
+The eigenrays are curated from BELLHOP by travel time proximity and are traced in the representative receiver colors over a total ray fan in gray.}
+  \label{fig:raytrace-zs90}
+\end{figure}
+
+\clearpage
+\section{\label{sec:post} Post-processed pseudorange analysis}
 
@@
 
-\subsection{Discussion}
-
 \section{Re-positioning using the nearest bounce criteria}
@@
 \begin{figure}[!ht]
-\includegraphics[width=\textwidth]{figs/trilateration-3-ways.pdf}
-\caption{WRITE THIS. Needs font change, legend change, x-axis change, new color palette.}
+\includegraphics[width=\textwidth]{figs/trilat-stat.pdf}
+\caption{WRITE THIS.}
 \label{fig:trilat}
 \end{figure}
+
+\clearpage
+\section{Re-navigating AUV data}
+
+FIG A -- error for MBC/NBC (Oscar Fig 6.17). Think about whether to include separate in situ column. Explain why this section uses error whereas previous section uses correction.
+
+I think if we go down this route, we can take out the GNSS noise section entirely.
+The POMA paper would go over GNSS noise at high latitudes, crossmap, and look at individual drifts.
+I do not think the crossmap needs to stay in this publication; given small numerical value of the re-positioning trilateration, I think the same point is made (in a better way, actually).
 
@@
 The shaded regions reference which sound speed estimate was used.}}
-\end{figure}
-
-\clearpage
-\FloatBarrier
-\subsection*{Group velocity prediction by source depth}
-\begin{figure}[h!]
-\includegraphics[width=0.8\textwidth]{figs/Fig10a.pdf} \\
-\vspace{2em}
-\includegraphics[width=0.8\textwidth]{figs/Fig10b.pdf}
-\caption{A comparison of group velocity predictions for all beacon to beacon events in post-processing.
-The top panel shows a source depth of 20 m; the bottom panel shows a source depth of 90 m.
-The sound speed source is indicated by color.
-The minimal and nearest bounce criterion are distinguished by different marker shapes, compared to the separately colored orange dots showing the naive, data-driven group velocity calculation.}
-\label{fig:gvelMore}
 \end{figure}

### 2022-02-19 rebuttal/rebuttal-letter/trunk-rebuttal.tex
new draft of intro

@@
 
-	The Beaufort Lens has been used in the literature (Litvak 2015, Schmidt and Schneider 2016) and in JASA/POMA (Poulsen et al 2016, Chen et al 2019).
+	The Beaufort Lens has been used in the literature (Litvak 2015, Schmidt and Schneider 2016) and in JASA/POMA specifically (Poulsen et al 2016, Chen et al 2019).
 

### 2022-02-19 manuscript/main.tex
new draft of intro

@@
 The long baseline (LBL) underwater navigation paradigm relies on the conversion of recorded travel time to range to trilaterate for position.
-For real-time operations, this conversion assumes an isovelocity sound speed.
+For real-time operations, this conversion has assumed an isovelocity sound speed.
 For re-navigation in post-processing, computationally and/or labor intensive acoustic modeling may be employed to reduce uncertainty driven by multipath arrivals.
-This work demonstrates a real-time ray-based prediction method of the effective horizontal group speed to minimize vehicle position error.
-\llabel{2.1} This method was implemented for a small scale AUV-LBL system in March 2020, in the Beaufort Sea, in total under-ice conditions and a double ducted acoustic propagation environment.
+This work demonstrates a real-time ray-based prediction method of the effective sound speed along a path from source to receiver to minimize vehicle position error.
+\llabel{2.1} This method was implemented for a small scale AUV-LBL system in March 2020, in the Beaufort Sea, in total ice-covered conditions and a double ducted acoustic propagation environment.
 The vehicle was successfully deployed and recovered.
-\llabel{2.2} Given the lack of GPS data throughout the vehicle's mission, however, the localization performance between GPS-linked beacon-to-beacon connections serves as a proxy for the effectiveness of this approach.
-The real-time positioning error between beacons was roughly 11 meters at distances up to 3 km, matching the magnitude of corrections recorded by the AUV during field operations.
-In post-processing, the localization performance is reevaluated with an operationally equivalent navigation stack using modeled, historical, and locally observed sound speed profiles for a Minimal Bounce Criteria (MBC), as deployed in real-time, and an improved ray filtering algorithm, the Nearest Bounce Criteria (NBC).
-The best performing median re-positioning errors by bounce criterias are roughly 10 and 2 meters, respectively.
-\llabel{2.3} Further investigation suggests that the GNSS ground truth data have some noise that is not reciprocated in the travel time data.
-Re-trilateration suggests that this approach effectively extends the single meter accuracy of the deployed GNSS pucks into the water column.
+\llabel{2.2} Given the lack of GPS data throughout the vehicle's mission, however, the pseudorange performance is first evaluated on connections between GPS-linked beacons.
+The real-time ranging error between beacons is roughly 11 meters at distances up to 3 km.
+But a consistent overestimation in the real-time method, the Minimum Bounce Criteria (MBC), provides insights for improved eigenray filtering, which we call the Nearest Bounce Criteria (NBC).
+An operationally equivalent pipeline is used to re-position the LBL beacons and re-navigate the AUV, using a modeled, historical, and a locally observed sound speed profile.
+The median re-positioning errors for the MBC and NBC are roughly 10 and 3 meters, respectively.
+The improved trilateration performance for re-positioning and re-navigation suggests that this approach effectively extends the single meter accuracy of the deployed GNSS units into the water column.
+
 \end{abstract}
@@
 Limiting navigation error and drift requires an AUV to periodically stall on the surface and obtain a GNSS fix to reset its position error.
-This foolproof method of self-positioning is undesirable for stealth, adverse weather conditions, and mission efficiency, and inaccessible in a GPS-denied situation like an ice-covered environment.
+This foolproof method of self-positioning is undesirable for stealth, adverse weather conditions, and mission efficiency, and inaccessible in a GNSS-denied situation like an ice-covered environment.
 
 % introduce LBL and precise statement of work
-Of acoustic baseline navigation systems, long baseline (LBL) is the most GPS-like in style and scale, and most appropriate for mitigating drift without overburdening computation or payload size on the vehicle \citep{paull_auv_2014,van_uffelen_global_2021}.
-The state-of-the-art for LBL outsources depth to a pressure sensor and solves the two-dimensional localization problem by assuming an isovelocity, linear scaling between one way travel time (OWTT) and range \citep{eustice_recent_2006,eustice_experimental_2007,webster_preliminary_2009,webster_advances_2012}.
+Of underwater acoustic navigation systems, long baseline (LBL) is the most GPS-like in style and scale, and most appropriate for mitigating drift without overburdening computation or payload size on the vehicle \citep{paull_auv_2014,van_uffelen_global_2021}.
+The state-of-the-art for LBL outsources depth to a pressure sensor and solves the two-dimensional localization problem with an isovelocity, linear scaling between one way travel time (OWTT) and range \citep{eustice_recent_2006,eustice_experimental_2007,webster_preliminary_2009,webster_advances_2012}.
 This assumption is valid for short scale operations but oversimplifies propagation for larger and/or complex acoustic environments.
-This paper demonstrates an embedded model-aided data processing approach to convert recorded OWTTs into pseudorange estimates.
-This approach was necessitated by total under-ice conditions and a double ducted acoustic environment, in the Beaufort Sea, in March 2020, for an AUV deployment during the Ice Exercise 2020 (ICEX20).
-We quantify the success of the proposed acoustic ranging method using GPS-linked beacon-to-beacon communication events, as the AUV-LBL positioning has no ground truth GNSS to compare to.
+To achieve single meter, GNSS-like performance in a GNSS-denied environment, we demonstrate an embedded ray-based data processing algorithm to convert recorded OWTTs into pseudorange estimates.
+This methodology was integrated onto the AUV Macrura, deployed and recovered in the Beaufort Sea, in March 2020, during the Ice Exercise 2020 (ICEX20).
+A physics-driven methodology that received an \textit{in situ} sound speed profile (SSP) was necessary despite the small operational domain because of the relatively high-risk mission environment\textemdash total under-ice conditions and a variable double ducted acoustic environment.
 
@@
 % PNT
-For clarity, we introduce definitions for timing, positioning, and navigation.
+For clarity, we delineate specific definitions for timing, positioning, and navigation from \citep{howe_observing_2019}.
+\begin{enumerate}
+	\item Timing is the ability to acquire and maintain accurate and precise time anywhere in the domain of interest within user-defined timeliness parameters
+	\item Positioning is the ability to accurately and precisely determine one's location referenced to a standard geodetic system
+	\item Navigation is the ability to determine current and desired position (relative or absolute) and apply corrections to course, orientation, and speed to attain a desired position anywhere in the domain of concern
+\end{enumerate}
+Thus, navigation is inherently in real time and depends on positioning; positioning depends on timing.
+We also suggest re-navigation and re-positioning as post-processed corollaries, which may include knowledge or processing capabilities not available \textit{in situ}.
 
 % More about LBL navigation
-While RAFOS floats have championed one way ranging for re-positioning \citep{rossby_rafos_1986,duda_evaluation_2006}, the ability to do so for navigation was facilitated by \llabel{1.11} the advent of the WHOI micro-modem \citep{singh_underwater_2006} and synchronized chip scale atomic clocks \citep{gardner_second_2016}.
-Several short range (less than ten kilometers) navigation efforts have achieved minimal error with a nominal sound speed value \citep{eustice_experimental_2007,webster_preliminary_2009,kepper_mems_2017}.
-However, these efforts were all conducted with X sound speed conditions.
-The approach in this paper, and other related works, was specifically designed to... the recently observed double ducted sound speed profile...
-For convenience, we call this phenomena the Beaufort Lens, with the lens referring to ...
+While RAFOS floats championed one way ranging for re-positioning \citep{rossby_rafos_1986,duda_evaluation_2006}, the ability to do so for navigation was facilitated by \llabel{1.11} the advent of the WHOI micro-modem \citep{singh_underwater_2006} and synchronized chip scale atomic clocks \citep{gardner_second_2016}.
+AUV navigation efforts have achieved root mean square (RMS) localization error on the order of tens of meters relative to GNSS surface position over less than ten kilometers in shallow \citep{eustice_experimental_2007,kepper_mems_2017,claus_closed-loop_2018} and deep water \citep{,kunz2008deep,jakuba2008long,webster_preliminary_2009}.
+However, these efforts all used a nominal sound speed for travel time conversion and the vehicles were limited to shallower isovelocity regimes. 
+Localization algorithms that do consider environmental or acoustic uncertainty tend to focus on longer and larger experiments, where spatio-temporal variability cannot be ignored.
+These methods have been reserved for post-processing as they can be labor intensive, computationally heavy, and/or require additional information like contemporaneous data.
+For example, gliders navigating with kinematic flight models and equipped with acoustic modems were later unambiguously associated with predicted ray arrivals, resulting in roughly a kilometer error and hundred meters uncertainty over basin scale propagation \cite{van_uffelen_estimating_2013}.
+A follow up study investigated how a single temporally and spatially averaged SSP could mitigate position error for a four month glider mission \cite{van_uffelen_localization_2016}.
+\citet{wu_assessment_2018} cross correlate three days of real acoustic records with synthetic ones generated through ocean model snapshots from HYCOM \cite{chassignet_hycom_2007}.
+While potentially applicable for various ocean states, this is reliant on model realism and impractical for real-time operations.
+\cite{mikhalevsky_deep_2020} introduces a ``cold start'' algorithm that does not require prior knowledge of track, position, or sound speed information.
+The algorithm inputs a four-dimensional ocean model, constrained by tomography data, into a range dependent ray code to isolate the last path detected in a full multipath pattern.
+Then, a representative group speed is solved for alongside position in a least squares fashion. 
+Their results re-position a floating hydrophone array with an error of 58 m and a standard deviation of 32 m based on six sources 129--450 km away.
+This approach remains to be realized for real-time navigation.
 
-There is no comparable work in the Arctic for a short scale AUV deployment in the Beaufort Lens. 
-Both seminal and more recent AUV deployments witnessed the classical upward refracting sound speed profile that is amenable to an isovelocity simplification.
-Of note are large scale glider deployments; these, however, operated at X depths, X spatial scales, and did not need precise navigation because they experienced little to no sea ice.
-At larger scales on the order of hundreds of kilometers, a deterministic \citep{graupe_preliminary_2019} or through-the-sensor \citep{webster_towards_2015} value for sound speed was used, but error remained in the hundreds of meters.
-The sound speed value was chosen by...
+Compared to the previous work mentioned, the approach in this paper integrates real-time model-aided data processing to estimate a representative sound speed along a path from source to receiver for small scale navigation, taking advantage of climatology, \textit{in situ} data, and fast acoustic modeling.
+An environmentally- and physically-driven relationship between recorded travel time and estimated range is necessary because of the multipath uncertainty brought upon by the increasingly observed double ducted environment in the Beaufort Sea, which some refer to as the ``Beaufort Lens'' \citep{litvak2015acoustics,chen_spectral_2019,chen2020temporal}.
 
-The navigation approach proposed includes a non-deterministic relationship between travel time and range previously limited to large scale re-navigation and re-positioning efforts.
-Such post-processing efforts can utilize additional information or computation that is inaccessible for onboard computation.
+Given that a lens introduces significant ray refraction, the Beaufort Lens is a convenient shorthand for the spatio-temporal variability of the local temperature and sound speed maxima driven by Pacific Summer Water, generally around 50 to 60 m in depth.
+From top to bottom, there is less saline surface water via ice melt; warm saline Pacific Summer Water; cold Pacific Winter Water; warm saline Atlantic water, and then Arctic Deep Water \citep{duda_acoustic_2017,ballard_temporal_2020}.
+With this intense stratification, the lens creates a unique double ducted environment that has drastic consequences for acoustic communication and navigation compared to historically observed monotonically increasing sound speed profile \citep{schmidt2016acoustic}.
+Transmissions in the upper duct, between the surface ice and the lens, experience minimal attenuation but degrade in signal coherence with repeated reflections under the ice.
+In lower duct, between the lens and its conjugate depth in the Atlantic water, sound above 350 Hz is trapped effectively for long range propagation \citep{poulsen2016acoustic}.
+
+Thorough reviews of uncrewed vehicle operations in polar environments can be found in \citep{norgren_unmanned_2014} and \citep{barker_scientific_2020}; there is no comparable work in the Arctic for a short range AUV deployment in the Beaufort Lens.
+Seminal \citep{brooke1981arcs,jackson1983autonomous,light1989autonomous,bellingham1995auv,hayes2002determining} and more recent AUV deployments \citep{jakuba2008long,kunz_deep_2008,kukulya2010under,plueddemann_autonomous_2012,timmermans2013scales,fossum2021adaptive} witnessed the classical upward refracting sound speed profile that is amenable to an isovelocity assumption.
+
+Of note, despite different platforms and scales, are recent glider deployments in the Canada Basin.
+In 2014, in partially ice-covered conditions, a long range LBL system with WHOI micro-modems at 100 m exploited the lower duct for long range communication with two gliders \citep{freitag_long_2015,webster2015towards}.
+The sound speed value measured at the time of reception was used to estimate range in post-processing.
+The beacon-to-beacon performance was excellent, achieving contact at ranges greater than 200 km with a position uncertainty of 40 m.
+The beacon-to-glider performance, however, deteriorated due to missed contacts outside the duct.
+In 2017, gliders were deployed in a region with no ice coverage. 
+Ranges were linearly scaled by a statistical description of sound speed observations taken during the experiment, 1450 $\pm$ 6.5 m/s \citep{graupe2019preliminary}.
+This resulted in an error of 550 m, which was reduced by a factor between 4 and 5, depending on the dive, using a post-processing acoustic arrival matching method.
+Both cases exploit the lower duct for high fidelity communication at long ranges.
+Unintuitively, the smaller nature of our deployment during ICEX20 is not a simplifying factor.
+For source depths typical to vehicle operations, 30 to 200 m, a shadow zone spans from 2 to 6 kilometers in range \citep{schmidt2016acoustic}.
+
+The paper is organized as follows. Section \ref{sec:icex20} details the experimental conditions during ICEX20.
+Given that there is no GNSS ground truth for the vehicle position while underway, we first evaluate the real-time ranging performance of GPS-linked beacon-to-beacon communication events in section \ref{sec:realtime}.
+Section \ref{sec:post} uses insights from field data to introduce a new ray filtering algorithm to improve range estimation.
+Section \ref{sec:trilat} emulates the real-time processing pipeline to re-position beacon-to-beacon events and re-navigate AUV Macrura.
+
+\clearpage
+\section{ICEX20 conditions and experiment design}\label{sec:icex20}
 
@@
 	\centering
-	\includegraphics[width=\columnwidth]{figs/auv-track-update.png}
+	\includegraphics[width=0.7\columnwidth]{figs/auv-track-update.png} \hfill
+	\includegraphics[width=0.28\columnwidth]{figs/Fig1.jpg}
 	\caption{The under-ice mission track for AUV Macrura, including the position updates as it stalled underneath the ice overnight.
@@
 % include story later on and cite this figure?
-
-\clearpage
-\section{ICEX20 conditions and experiment design}
 
@@
   \includegraphics[width=\reprintcolumnwidth]{figs/Fig4.pdf}
-  \caption{Eigenrays for beacon to beacon events for each sound speed with a nominal source depth of 30 m.
+  \caption{Eigenrays for beacon to beacon events for each sound speed with a nominal source depth of 20 m.
 The beacons are highlighted in color/marker coding in Fig.
@@
   \includegraphics[width=\reprintcolumnwidth]{figs/Fig4.pdf}
-  \caption{Eigenrays for beacon to beacon events for each sound speed with a nominal source depth of 30 m.
+  \caption{Eigenrays for beacon to beacon events for each sound speed with a nominal source depth of 90 m.
 The beacons are highlighted in color/marker coding in Fig.
@@
 
-\section{Re-positioning using the nearest bounce criteria}
+\clearpage
+\section{Trilateration for ICEX20 field data}\label{sec:trilat}
 
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{figs/Fig7.pdf}
-\caption{\label{fig:compareV1V2}{A comparison of range estimation error for all beacon to beacon events in post-processing.
-From left to right, the SSPs are the baseline, the chosen weights, and HYCOM.
-The colors indicate the source depth, darkening with depth, and the shapes indicate the multipath structure.
-Because this plot is square, the shaded region shows where the updated algorithm is less accurate than the \textit{in situ} algorithm.}}
-\end{figure}
+\subsection{Re-positioning beacon to beacon events}
 
@@
 
-\clearpage
-\section{Re-navigating AUV data}
+\subsection{Re-navigating AUV Macrura}
 
@@
 
-\clearpage
-\section{Investigating potential GNSS noise}
+\subsection{Investigating potential GNSS noise}
 
@@
 	\centering
-	\includegraphics[width=\reprintcolumnwidth]{figs/Fig8.pdf} 
+	\includegraphics[width=\reprintcolumnwidth]{figs/gps-drift-example.pdf} 
 	\caption{A comparison of GPS drift (y-axis) versus OWTT drift (x-axis), colored by source and receiver depth.
@@
 	\label{fig:gps-drift-example}
-\end{figure}
-
-\clearpage
-\section*{Supplementary Figures}
-
-\subsection*{Ice movement during the modem experiment}
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{figs/Fig9.pdf}
-\caption{\label{fig:iceFloeDrift}{The ICNN accounts for ice drift throughout the mission duration.
-This is the magnitude of the ice drift recorded, at each buoy, throughout the modem experiment.
-The shaded regions reference which sound speed estimate was used.}}
 \end{figure}

### 2022-02-19 pipeline/1-overview/plot_ssp_comparison.m
convert this from PNG to PDF

@@ -31,9 +31,9 @@ end
 
 hold on
 plot(NaN,NaN,'w');
-plot(historical.sspVal,historical.sspDepth,':','color',colorSet{1},'linewidth',4);
-plot(data.sspVal,data.sspDepth,':','color',colorSet{2},'linewidth',4);
-plot(hycom.sspVal,hycom.sspDepth,':','color',colorSet{3},'linewidth',4);
+plot(historical.sspVal,historical.sspDepth,':','color',[colorSet{1} 0.8],'linewidth',4);
+plot(data.sspVal,data.sspDepth,':','color',[colorSet{2} 0.8],'linewidth',4);
+plot(hycom.sspVal,hycom.sspDepth,':','color',[colorSet{3} 0.8],'linewidth',4);
 hold off
 
 %% beautify plot
@@ -63,4 +63,4 @@ fprintf('Isovelocity SSP = %4.1f +/- %4.1f m/s \n',meanSSP,stdSSP)
 
 
 %% export
-% h_printThesisPNG('ssp-gvel-icex20-icex16');
+h_printThesisPNG('ssp-gvel-icex20-icex16');

### 2022-02-19 pipeline/9-trilateration/modem_trilat_hist.m
reroute source file

@@ -3,7 +3,7 @@
 clear; clc; close all;
 
 %% load trilat mat from Oscar
-load ../../data/trilat_plus_hists.mat
+load ../../data/modem_trilat_plus_hists.mat
 trilat = trilateration_results;
 clear trilateration_results
 

### 2022-02-19 pipeline/9-trilateration/auv_trilat_hist.m
consistent trilat hists

@@ -1,0 +1,52 @@
+%% prep workspace
+
+clear; clc; close all;
+
+%% load trilat mat from Oscar
+load ../../data/auv_ops_trilat_plus_hists.mat
+trilat = trilateration_results;
+clear trilateration_results
+
+colors = {[153 51 153]./256,[51 153 153]./256};
+
+Y{2}.error = trilat.mbc.error;
+Y{1}.error = trilat.nbc.error;
+
+N = numel(Y{1}.error);
+
+
+%% figure
+figure('name','trilat-histogram','renderer','painters','position',[108 108 1200 1000]);
+t = tiledlayout(2,1,'TileSpacing','compact','Padding','Compact');
+
+%% tile 1 - histogram of corrections
+nexttile(1);
+
+edges = [0:5:65];
+Y{2}.count = trilat.mbc.correction;
+Y{1}.count = trilat.nbc.correction;
+h_hist_boxplot(Y,edges,colors,max(edges));
+
+xlabel('RMS of position correction [m]')
+set(gca,'fontsize',13)
+title(sprintf('Distribution of AUV re-navigation corrections (n=%u)',N),'fontsize',15)
+
+legend off
+
+%% tile 2 - histogram of error
+nexttile(2);
+
+edges = [0:3:21];
+Y{1}.count = trilat.nbc.error;
+Y{2}.count = trilat.mbc.error;
+
+h_hist_boxplot(Y,edges,colors,max(edges));
+xlabel('RMS of position error [m]')
+set(gca,'fontsize',13)
+title({'',...
+    sprintf('Distribution of AUV re-navigation error (n=%u)',N)}...
+    ,'fontsize',15)
+
+%% export
+
+h_printThesisPNG('auv-trilat-stat');

### 2022-02-19 pipeline/9-trilateration/h_hist_boxplot.m

@@ -1,0 +1,72 @@
+function [] = h_hist_boxplot(Y,EDGES,COLORS,THRESHOLD)
+%h_hist_boxplot makes a beautiful histogram w/ a boxplot underlaid
+
+diffEdges = mean(diff(EDGES))./2;
+
+% histogram
+for s = 1:numel(Y)
+    h(s,:) = histcounts(Y{s}.count,EDGES,'normalization','probability');
+end
+B = bar(EDGES(1:end-1)+diffEdges,h,0.9,'FaceColor','flat','EdgeColor','none');
+
+for s = 1:2
+    B(s).CData = COLORS{s};
+    B(s).FaceAlpha = 0.8;
+end
+
+grid on
+ylabel('probability');
+xticks(EDGES);
+xlim([-1 max(EDGES)]);
+
+% box plot
+hold on
+buff = -0.15;
+
+
+% add bottom "box" plot
+hold on
+buff = -0.15;
+maxY = ceil(max(h(:))*10)/10;
+ylim([buff maxY]);
+yticks([0:0.1:maxY]);
+
+kbuff = .05;
+for s = 1:2
+    
+    yval = Y{s}.count;
+    
+    yval(yval>THRESHOLD) = NaN;
+    
+    meanVal = mean(yval,'omitnan');
+    xQuant1 = quantile(yval,[0 1]);
+    xQuant2 = quantile(yval,[.25 .5 .75]);
+    
+    plot(xQuant1,ones(2,1).*-s*kbuff,'handlevisibility','off','color',[COLORS{s} 0.4]);
+    plot(xQuant1,-s*kbuff,'.','handlevisibility','off','color',COLORS{s},'MarkerSize',15);
+    plot(meanVal,-s*kbuff,'o','handlevisibility','off','color',COLORS{s},'MarkerSize',12);
+    
+    plot(xQuant2(1),-s*kbuff,'<','handlevisibility','off','color',COLORS{s});
+    plot(xQuant2(3),-s*kbuff,'>','handlevisibility','off','color',COLORS{s});
+    plot(xQuant2(2),-s*kbuff,'d','handlevisibility','off','color',COLORS{s});
+
+end
+
+plot(NaN,NaN,'w');
+plot(NaN,NaN,'k.','markersize',15);
+plot(NaN,NaN,'k<');
+plot(NaN,NaN,'ko','markersize',15);
+plot(NaN,NaN,'kd');
+plot(NaN,NaN,'k>');
+
+% add legend
[…]

### 2022-02-19 pipeline/9-trilateration/modem_trilat_hist.m

@@ -7,74 +7,22 @@ load ../../data/modem_trilat_plus_hists.mat
 trilat = trilateration_results;
 clear trilateration_results
 
-edges = [0:3:42];
+edges = [0:3:48];
 Y{2}.count = trilat.mbc.correction;
 Y{1}.count = trilat.nbc.correction;
+N = numel(Y{1}.count);
 
 colors = {[153 51 153]./256,[51 153 153]./256};
 
-%% figure 1 - histogram
+%% figure
 figure('name','trilat-histogram','renderer','painters','position',[108 108 1200 500]);
 
-for s = 1:2
-    h(s,:) = histcounts(Y{s}.count,edges,'normalization','probability');
-end
-B = bar(edges(1:end-1)+1.5,h,0.9,'FaceColor','flat','EdgeColor','none');
+h_hist_boxplot(Y,edges,colors,48);
 
-for s = 1:2
-    B(s).CData = colors{s};
-    B(s).FaceAlpha = 0.8;
-end
-
-grid on
-xlabel('Re-navigation RMS error [m]')
-ylabel('probability');
-xticks(edges);
-xlim([-1 39]);
-%set(gca,'fontsize',12)
-title('Distribution of trilateration corrections')
-
-% add bottom "box" plot
-hold on
-buff = -0.15;
-ylim([buff 0.6]);
-yticks([0:0.1:0.6]);
-
-kbuff = .05;
-for s = 1:2
-    
-    yval = Y{s}.count;
-    
-    yval(yval>40) = NaN;
-    
-    meanVal = mean(yval,'omitnan');
-    xQuant1 = quantile(yval,[0 1]);
-    xQuant2 = quantile(yval,[.25 .5 .75]);
-    
-    plot(xQuant1,ones(2,1).*-s*kbuff,'handlevisibility','off','color',[colors{s} 0.4]);
-    plot(xQuant1,-s*kbuff,'.','handlevisibility','off','color',colors{s},'MarkerSize',15);
-    plot(meanVal,-s*kbuff,'o','handlevisibility','off','color',colors{s},'MarkerSize',12);
-    
-    plot(xQuant2(1),-s*kbuff,'<','handlevisibility','off','color',colors{s});
-    plot(xQuant2(3),-s*kbuff,'>','handlevisibility','off','color',colors{s});
-    plot(xQuant2(2),-s*kbuff,'d','handlevisibility','off','color',colors{s});
-
-end
-
-plot(NaN,NaN,'w');
[…]

### 2022-02-19 manuscript/main.tex
finalized figs

@@
 \begin{figure}[!ht]
-\includegraphics[width=\textwidth]{figs/trilat-stat.pdf}
+\includegraphics[width=\textwidth]{figs/beacon-trilat-stat.pdf}
 \caption{WRITE THIS.}
-\label{fig:trilat}
+\label{fig:trilat-beacon}
 \end{figure}
@@
 
-FIG A -- error for MBC/NBC (Oscar Fig 6.17). Think about whether to include separate in situ column. Explain why this section uses error whereas previous section uses correction.
-
-I think if we go down this route, we can take out the GNSS noise section entirely.
-The POMA paper would go over GNSS noise at high latitudes, crossmap, and look at individual drifts.
-I do not think the crossmap needs to stay in this publication; given small numerical value of the re-positioning trilateration, I think the same point is made (in a better way, actually).
+\begin{figure}[!ht]
+\includegraphics[width=\textwidth]{figs/auv-trilat-stat.pdf}
+\caption{WRITE THIS.}
+\label{fig:trilat-auv}
+\end{figure}
 

### 2022-02-19 manuscript/main.tex
ported over bad revision draft into new structure

@@
 The improved trilateration performance for re-positioning and re-navigation suggests that this approach effectively extends the single meter accuracy of the deployed GNSS units into the water column.
+\llabel{2.3}
 
@@
 % PNT
-For clarity, we delineate specific definitions for timing, positioning, and navigation from \citep{howe_observing_2019}.
+For clarity, we delineate specific definitions for timing, positioning, and navigation from \citet{howe_observing_2019}.
 \begin{enumerate}
@@
 % More about LBL navigation
-While RAFOS floats championed one way ranging for re-positioning \citep{rossby_rafos_1986,duda_evaluation_2006}, the ability to do so for navigation was facilitated by \llabel{1.11} the advent of the WHOI micro-modem \citep{singh_underwater_2006} and synchronized chip scale atomic clocks \citep{gardner_second_2016}.
-AUV navigation efforts have achieved root mean square (RMS) localization error on the order of tens of meters relative to GNSS surface position over less than ten kilometers in shallow \citep{eustice_experimental_2007,kepper_mems_2017,claus_closed-loop_2018} and deep water \citep{,kunz2008deep,jakuba2008long,webster_preliminary_2009}.
+While RAFOS floats championed one way ranging for re-positioning \citep{rossby_rafos_1986,duda_evaluation_2006}, the ability to do so for navigation was facilitated by \llabel{1.11} the advent of the WHOI Micro-Modem \citep{singh_underwater_2006} and synchronized chip scale atomic clocks \citep{gardner_second_2016}.
+AUV navigation efforts have achieved root mean square (RMS) localization error on the order of tens of meters relative to GNSS surface position over less than ten kilometers in shallow \citep{eustice_experimental_2007,kepper_mems_2017,claus_closed-loop_2018} and deep water \citep{kunz2008deep,jakuba2008long,webster_preliminary_2009}.
 However, these efforts all used a nominal sound speed for travel time conversion and the vehicles were limited to shallower isovelocity regimes.
+
 Localization algorithms that do consider environmental or acoustic uncertainty tend to focus on longer and larger experiments, where spatio-temporal variability cannot be ignored.
@@
 While potentially applicable for various ocean states, this is reliant on model realism and impractical for real-time operations.
-\cite{mikhalevsky_deep_2020} introduces a ``cold start'' algorithm that does not require prior knowledge of track, position, or sound speed information.
+\citet{mikhalevsky_deep_2020} introduces a ``cold start'' algorithm that does not require prior knowledge of track, position, or sound speed information.
 The algorithm inputs a four-dimensional ocean model, constrained by tomography data, into a range dependent ray code to isolate the last path detected in a full multipath pattern.
 Then, a representative group speed is solved for alongside position in a least squares fashion. 
-Their results re-position a floating hydrophone array with an error of 58 m and a standard deviation of 32 m based on six sources 129--450 km away.
-This approach remains to be realized for real-time navigation.
+This approach is able to re-position a floating hydrophone array with an error of 58 m and a standard deviation of 32 m based on six sources 129--450 km away but remains to seen for real-time navigation.
 
-Compared to the previous work mentioned, the approach in this paper integrates real-time model-aided data processing to estimate a representative sound speed along a path from source to receiver for small scale navigation, taking advantage of climatology, \textit{in situ} data, and fast acoustic modeling.
-An environmentally- and physically-driven relationship between recorded travel time and estimated range is necessary because of the multipath uncertainty brought upon by the increasingly observed double ducted environment in the Beaufort Sea, which some refer to as the ``Beaufort Lens'' \citep{litvak2015acoustics,chen_spectral_2019,chen2020temporal}.
+The ICEX20 AUV deployment necessitated an environmentally- and physically-driven relationship between recorded travel time and estimated range due to the multipath uncertainty brought upon by an increasingly observed double ducted environment in the Beaufort Sea, which some refer to as the ``Beaufort Lens'' \citep{litvak2015acoustics,chen_spectral_2019,chen2020temporal}.
 
-Given that a lens introduces significant ray refraction, the Beaufort Lens is a convenient shorthand for the spatio-temporal variability of the local temperature and sound speed maxima driven by Pacific Summer Water, generally around 50 to 60 m in depth.
-From top to bottom, there is less saline surface water via ice melt; warm saline Pacific Summer Water; cold Pacific Winter Water; warm saline Atlantic water, and then Arctic Deep Water \citep{duda_acoustic_2017,ballard_temporal_2020}.
-With this intense stratification, the lens creates a unique double ducted environment that has drastic consequences for acoustic communication and navigation compared to historically observed monotonically increasing sound speed profile \citep{schmidt2016acoustic}.
+Given that a lens introduces significant ray refraction, the Beaufort Lens is a convenient shorthand for the spatio-temporal variability of the local temperature and sound speed maxima generally around 50 to 60 m in depth.
+A neutrally buoyant layer of warm Pacific Summer Water creates a unique double ducted environment \textemdash the upper duct degrades signal coherence due to intensified ice interaction and the lower duct effectively traps sound for long range propagation \citep{poulsen_acoustic_2016}.
+\llabel{2.5} Modeling output \citep{duda_long-range_2019,duda_effects_2021} and experimental observations \citep{ballard_temporal_2020,Bhatt2021} suggest the duct is persistent and widespread but not necessarily continuous; it and its acoustic effects can be non-existent, minimal, or drastic.
 Transmissions in the upper duct, between the surface ice and the lens, experience minimal attenuation but degrade in signal coherence with repeated reflections under the ice.
-In lower duct, between the lens and its conjugate depth in the Atlantic water, sound above 350 Hz is trapped effectively for long range propagation \citep{poulsen2016acoustic}.
+In lower duct, between the lens and its conjugate depth in the Atlantic water, roughly 200 m, sound above 350 Hz is trapped effectively for long range propagation \citep{poulsen2016acoustic}.
 
@@
 Of note, despite different platforms and scales, are recent glider deployments in the Canada Basin.
-In 2014, in partially ice-covered conditions, a long range LBL system with WHOI micro-modems at 100 m exploited the lower duct for long range communication with two gliders \citep{freitag_long_2015,webster2015towards}.
+In 2014, in partially ice-covered conditions, a long range LBL system with WHOI Micro-Modems at 100 m exploited the lower duct for long range communication with two gliders \citep{freitag_long_2015,webster2015towards}.
 The sound speed value measured at the time of reception was used to estimate range in post-processing.
@@
 
-The paper is organized as follows. Section \ref{sec:icex20} details the experimental conditions during ICEX20.
+Compared to the previous small scale navigation efforts, the approach in this paper integrates real-time model-aided data processing to estimate a representative sound speed along a path from source to receiver, leveraging climatology, \textit{in situ} data, and fast acoustic modeling.
+The paper is organized as follows.
+Section \ref{sec:icex20} details the experimental conditions during ICEX20.
 Given that there is no GNSS ground truth for the vehicle position while underway, we first evaluate the real-time ranging performance of GPS-linked beacon-to-beacon communication events in section \ref{sec:realtime}.
 Section \ref{sec:post} uses insights from field data to introduce a new ray filtering algorithm to improve range estimation.
-Section \ref{sec:trilat} emulates the real-time processing pipeline to re-position beacon-to-beacon events and re-navigate AUV Macrura.
+Section \ref{sec:trilat} emulates the real-time processing pipeline to re-position beacon-to-beacon events and re-navigate AUV \emph{Macrura}.
 
 \clearpage
-\section{ICEX20 conditions and experiment design}\label{sec:icex20}
+\section{Overview of the ICEX20 experiment}\label{sec:icex20}
+
+The results from this paper derive from data taken while deploying the AUV \emph{Macrura}, a custom Bluefin-21, during the Ice Exercise 2020 (ICEX20), in the Beaufort Sea, from March 8 to 11, in double-ducted and ice-covered conditions.
+The AUV deployment was supported by the Integrated Communication and Navigation Network (ICNN) \citep{schneider_self-adapting_2020,randeni_construction_2020,randeni_high-resolution_2021}.
+The ICNN was initially developed via numerous virtual experiments to push the boundaries of algorithms and interfaces between different hardware components.
+The simulation approach serves as a testbed for robustness to produce better results than post-processing previous field data.
+The simulation capabilities are largely physics-driven with a modular system of systems approach\textemdash an environmental simulator with sub-components for the ocean, including Arctic ice drift and ocean acoustic propagation; a vehicle simulator with sub-components for vehicle dynamics and navigation; a topside hardware simulator and acoustic communications simulator, both with a software-only configuration and a hardware-in-the-loop version \citep{schneider_netsim_2018}.
+The virtual environment similarly emulates the interfaces between the real components to test the entire software pipeline.
+Both simulation capabilities are integral to mission success.
+
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=\reprintcolumnwidth]{figs/Fig2.pdf}
+	\caption{A schematic overview of the Integrated Communication and Navigation Network (ICNN), which provides joint data-transfer and tracking between AUV and a human decision maker at Topside.}
+	\label{fig:icnnOverview}
+\end{figure}
+
+The ICNN is comprised of four ice buoys, in a loose rectangle, roughly 2 km away from a central ice camp with a topside computer, as shown in Fig.~\ref{fig:icnnOverview}.
+The AUV and each ice buoy are outfitted with a WHOI Micro-Modem \citep{singh_underwater_2006}, with a four-element receiver array and a single transmitter, one-tenth of a millisecond resolution.
+Acoustic messages were sent with a 10 kHz carrier frequency, 5 kHz bandwidth, and phase-shift keying modulation on a thirty-second cycle, giving room for two-way communication throughout the mission volume.
+The receive and transmit elements were split between shallow and deeper depths (30 and 90 m, respectively) to provide better coverage across the shadow zone.
+While each buoy only has one transmit depth, all buoys have both receive depths but the active receive layer is consistent across all buoys. \llabel{1.12}
+The design of the ICNN enables a self-adapting network to transmit and receive at the optimal depth to maintain contact with the AUV \citep{schneider_self-adapting_2020}.
+The buoys do not encompass the full horizontal range of the vehicle but are positioned to minimize overlap in trilateration for spherical positioning \citep{deffenbaugh_relationship_1996}.
+
+To balance competing uses of the acoustic channel, the network uses a single synchronized digital communication packet to provide both tracking and data to the operator.
+\begin{enumerate}
+\item The AUV, running an ice-tracking DVL and an onboard hydrodynamic model, broadcasts its perceived location on a scheduled, time-synchronized message via WHOI Micro-Modem
+\item Four ice buoys, each outfitted with a WHOI Micro-Modem, receive messages from the AUV and send that information over freewave radio to a Topside computer
+\item The topside computer converts travel times into range estimates using a stochastic embedded prediction of the horizontal group velocity via BELLHOP ray tracing code \citep{porter_bellhop_2011} coupled with an updatable Virtual Ocean \citep{schneider_netsim_2018,bhatt_embedded_2021}
+\item The topside computer calculates a new position by trilaterating the range estimates
+\item The position differential, not the absolute position, is broadcast to the vehicle to update its navigation solution and be robust to latency and intermittency
+\end{enumerate}
+
+In the face of GNSS-denied navigation, this approach was anecdotally successful, as shown in \ref{fig:vehicleRecovery}.
+The AUV \emph{Macrura} was deployed through a hydrohole from an ice camp but recovered through an emergency hydrohole.
+A random disk error stalled the AUV underneath the ice but did not prevent it from transmitting its location.
+Due to an incoming storm, a team placed a physical marker on the ice at the location.
+Three days later, \emph{Macrura} was found within a meter of the marker.
+We view the emergency recovery as qualitative proof of the robustness of this navigation approach.
+Nonetheless, this paper specifically addresses the third and fourth steps\textemdash the conversion of travel time into range and its effect on trilateration.
+By focusing on range estimates between GPS-tracked beacons, and re-running the trilateration pipeline, the results are decoupled from all other mechanisms in the ICNN.
 
@@
 \end{figure}
-% include story later on and cite this figure?
+
+An important component to our navigation solution is an accurate estimation of the sound speed profile.
+Previous field experience, during the Ice Exercise 2016 (ICEX16), demonstrated the negative effects of the Beaufort Lens on tracking and communication \citep{schmidt2016acoustic}.
+Fig. \ref{fig:sspExpectation} shows historical, modeled, and \textit{in situ} data for both ICEX16 and ICEX20.
+These three input streams were selected to mirror the information available on a submarine (personal conversation with LT B. Howard and LT CDR D. Goodwin). \llabel{1.8}
+The SSP information was compressed with a basis representation and sent via lightweight digital acoustic message to the AUV \citep{bhatt_embedded_2021}. \llabel{2.7}
+All modeled data comes from HYCOM \cite{chassignet_hycom_2007}, which does not seem to capture the forcing mechanisms that cause the Beaufort Lens.
+For ICEX16, the data-driven profile was sourced from nearby Ice Tethered Profilers (ITP) after the field experiment \cite{krishfield2008automated,toole2011ice} and exhibits a fairly low lens; the historical profile is from the World Ocean Atlas.
+For ICEX20, the chosen weights (data-driven) profile derives from a basis representation estimation of initial CTD casts taken on site, showing an intense warm water intrusion; the baseline (historical) profile, showing moderate ducted conditions, comes from the average of March 2013 data.
+This month best matched sea ice and sound speed conditions at the beginning of ICEX20 \citep{bhatt_embedded_2021}.
+It is important to note that all profiles that do show the Beaufort Lens do so with different local sound speed maxima at different depths, reflective of the wide range of lens properties observed for all ITP data in the region.
+The variability of the lens height and prominence is the main reason an updatable SSP was integrated into the ICNN solution.
+
+During ICEX20, the HYCOM profile was available but never deployed. \llabel{1.14}
+For post-processing comparison, we introduce both the HYCOM profile and an isovelocity case, 1441.8 $\pm$ 3.7 m/s, as the mean and standard deviation of the observed sound speed profile over the first 200 m.
 
@@
 	\includegraphics[width=\reprintcolumnwidth]{figs/ssp-gvel-icex20-icex16.pdf}
-	\caption{Anticipated sound speed conditions: finish this using stuff from ICEX16, HYCOM, ITP}
+	\caption{Sound speed conditions for historical (baseline), data (chosen weights), and HYCOM model for both ICEX16 and ICEX20.}
 	\label{fig:sspExpectation}
@@
 
-\begin{figure}[h!]
-	\centering
-	\includegraphics[width=\textwidth]{figs/Fig2.pdf}
-	\caption{A schematic overview of the Integrated Communication and Navigation Network (ICNN), which provides joint data-transfer and tracking between AUV and a human decision maker at Topside.}
-	\label{fig:icnnOverview}
-\end{figure}
+\clearpage
+\section{\label{sec:realtime} Real-time pseudorange analysis}
+
+Because the vehicle's navigation solution during a mission can only be evaluated on the basis of the error estimates sent, a sister experiment for validating the real-time ranging approach was implemented.
+Ice buoy modems were run as ``virtual vehicles'' at a fixed depth, receiving position updates from the other beacons as well as a camp site modem lowered to 20 m. \llabel{1.13}
+Figure \ref{fig:overview} shows successful events sorted by source depth.
+In this analysis, we assume there is insignificant displacement between the GNSS puck surface expression and subsurface modem; this is supported by unusually low ice drift rates, on average, just 0.7 cm/s.
 
@@
   \includegraphics[width=\textwidth]{figs/modem-chart.pdf}
-  \caption{An overview of the modem experiment by source and receiver depth and position.
+  \caption{An overview of the modem experiment by source and receiver depth and position with the number of successful decodings and the percent success rate, defined as the ratio of decoded to detected.
 The black column on the left, $tx$, shows the source depth, $z_s$.
@@
 
-\clearpage
-\section{\label{sec:realtime} Real-time pseudorange analysis}
+\subsection{Minimal bounce criteria (MBC)}
 
-\subsection{Minimal bounce criteria (MBC)}
+The fundamental challenge to implement GNSS-like navigation, especially in an acoustically complex propagation environment, is characterizing a single sound speed to compensate for the effects of ray refraction and reflection.
+The use of the acoustic modem network for tracking relies on the accurate estimates of travel times between the submerged platform and LBL beacons, supported by clock synchronization and a pre-determined scheduling of acoustic events.
+\llabel{2.9} For the Beaufort Lens in particular, the strong multipath effects make it virtually impossible to deterministically predict the modem's detected arrival time.
+
+Instead, for each individual node $i$, an embedded stochastic tracking framework is used to provide a running estimate of the horizontal group velocity $u_{i,j}$ for the conversion from travel time to range from modem $j$, with the ultimate goal of matching the naive group velocity, i.e.
+the GPS-recorded distance between two nodes divided by the modem-recorded one way travel time between them.
+
+\llabel{2.10} In the ICEX20 configuration, the acoustic tracking is running on the topside computer, which controls the ICNN.
+Here we assume that the group velocities $u_{i,j}$ are smoothly varying over the course of a vehicle mission, i.e., with respect to range, mission time, and the frequency of updates relative to vehicle motion.
+The group velocity is tracked on a thirty-second cycle using predictions from the \textit{Virtual Ocean} infrastructure.
+
+\llabel{1.15} When the topside tracking framework receives a modem message, with a time delay, $\Delta t$, from one of the range modems, it will request a new estimate of the group velocity and its associated uncertainty.
+The group velocity estimate is found using the vehicle's reported depth and the extrapolated navigation solution for range, $\hat{r}$, as inputs for the ray tracing program.
+The latter returns an impulse response estimate in the form of ray travel times $dt_{j}$ and amplitudes $a_{j}$ for that range and depth.
+
+\llabel{2.11} The initial call to BELLHOP is over a local grid centered at the range and depth posited by the onboard tracking solution.
+The grid is initialized as 11$\times$11 points spanning 10 m horizontally and 20 m vertically.
+The horizontal dimension reflects the accumulated position error given a thirty-second communication cycle; the vertical dimension reflects how, computationally, eigenrays of the same timefront seem to stack vertically in the water column.
+\llabel{1.17}
+For each grid point, BELLHOP produces a number of arrivals resulting from multiple propagation paths for any source-receiver pair.
+Using only the $N_0$ rays with neither surface nor bottom bounces, it will then estimate the current group velocity $u$ from a power weighted average of the ray travel times,
+\begin{equation}
+u = \frac{\hat{r} \sum_{n=1}^{N_{0}} a_{n}^{2}}{\sum_{n=1}^{N_{0}} dt_{n}a_{n}^{2}} ~, 
+\end{equation}
+and the associated weighted standard deviation,
+\begin{equation}
+\sigma_{u} \simeq \sqrt{\frac {\sum_{n=1}^{N_{0}} (dt_{n}-\hat{r}/u)^{2}a_{n}^{2}}{ \sum_{n=1}^{N_{0}} a_{n}^{2}} } \frac{u^{2}}{\hat{r}}
+\end{equation}
+If no direct paths exist, i.e. $N_{0}=0$, then the group velocity is computed using the same algorithm for the ray arrivals with one bounce, and so on.
+
+Finally, the pseudorange is calculated simply as
+\begin{equation}
+r_{i,j} = u_{i,j} \Delta t_{i,j} 
+\end{equation}
+
+This stochastic, ensemble method for group velocity calculation can run in real-time, appearing to be orders of magnitude faster than other post-processing methods which seek to determine the specific ray itself that best matches a prominent indicator from the arrival structure.
+The BELLHOP simulation that runs this calculation uses 3600 rays with launch angle fan of -60 to 60 degrees, a depth dependent sound speed profile, and a range dependent bathymetry. \llabel{1.13}
+Even though the NBC looks for just the least complex multipath, the high density of launch angles almost always guarantees a direct path.
+The MBC improves upon a deterministic estimate in that it considers the source and receiver depth to predict a more realistic group velocity.
 
@@
 
-\subsection{Inherent overestimation from the minimal bounce criteria}
+The sister modem experiment generated 811 beacon to beacon communication events with their own real-time MBC group velocity predictions.
+Given the complexity of the ICNN system, this experiment did not collect an exhaustive set of data across all buoy, source depth, receive depth, and sound speed combinations.
+\llabel{1.18} The algorithm generally overestimates pseudoranges because it resolves the effective sound speed for the most direct path.
+
+\begin{figure}[h!]
+  \centering
+  \includegraphics[width=\reprintcolumnwidth]{figs/range-error-insitu.pdf}
+  \caption{The real-time range error by source (20, 30, and 90 m) and receiver (30 and 90 m) pairings for both sound speed estimates used during ICEX20.
+The amount of communication events is notated in the top right of each panel.
+The dashed line indicates no range error, and the boundary drawn indicates scope of the range error as a function of one way travel time.}
+  \label{fig:range-error-insitu}
+\end{figure}
+
+Figure \ref{fig:range-error-insitu} shows the range error boundary for both SSP inputs used in ICEX20.
+A promising sign that the MBC method adapts sound speed realistically is no obvious error growth as travel time increases.
+The baseline SSP (n=243 events) has an absolute pseudorange error of 11.38 $\pm$ 4.23 m; the weighted SSP (n=568), 11.36 $\pm$ 8.12 m.
+The discrepancy between these two is largely due to outlier events only contained in the weighted SSP set.
+Where there is overlap between sound speed conditions used for the real-time MBC, the pseudorange error difference is no more than a few meters.
+The overarching results show that sound speed estimates derived from eigenrays for a local grid, as opposed to a singular point, are accurate enough to support vehicle navigation.
+Nonetheless, the consistent overestimation of pseudorange invites further analysis into acoustic arrival matching.
+
+\subsection{Eigenray identification for beacon-to-beacon events}
+
+Fig. \ref{fig:raytrace-zs30} shows eigenrays in a range independent coordinate system for all three sound speed environments, for a source depth of 30 m.
+\llabel{1.16} It is important to note that BELLHOP produces an \llabel{2.8}abundance of eigenrays; the ones displayed here are chosen by travel time proximity to the recorded data.
+With this filter, the chosen eigenrays show a greater amount of surface interactions as the SSP duct conditions increase.
+
+\llabel{2.14} This is likely driven by the prominence of the duct.
 
 \subsubsection{Source depth of 20 m}
-\begin{figure}[h!]
+\begin{figure}[ht!]
   \centering
-  \includegraphics[width=\reprintcolumnwidth]{figs/Fig4.pdf}
+  \includegraphics[width=\reprintcolumnwidth]{figs/raytrace-3env-zs-20.pdf}
   \caption{Eigenrays for beacon to beacon events for each sound speed with a nominal source depth of 20 m.
@@
 \subsubsection{Source depth of 30 m}
-\begin{figure}[h!]
+\begin{figure}[ht!]
   \centering
-  \includegraphics[width=\reprintcolumnwidth]{figs/Fig4.pdf}
+  \includegraphics[width=\reprintcolumnwidth]{figs/raytrace-3env-zs-30.pdf}
   \caption{Eigenrays for beacon to beacon events for each sound speed with a nominal source depth of 30 m.
@@
 \subsubsection{Source depth of 90 m}
-\begin{figure}[h!]
+\begin{figure}[ht!]
   \centering
-  \includegraphics[width=\reprintcolumnwidth]{figs/Fig4.pdf}
+  \includegraphics[width=\reprintcolumnwidth]{figs/raytrace-3env-zs-90.pdf}
   \caption{Eigenrays for beacon to beacon events for each sound speed with a nominal source depth of 90 m.
@@
 
+There are 1242 total beacon-to-beacon events in the modem experiment if real-time ranging can be substituted for post-processed ranging.
+Thus, a post-processing analysis that emulates the real-time navigation engine was run to overcome the unequal distribution of communication events with respect to depth, range, and sound speed status.
+
+It is important to note that the value for the extrapolated range, $\hat{r}$, is only tracked when the modem runs the vehicle behavior; thus we replace $\hat{r}$ with the GPS-tracked range for all modem events.
+Because $\hat{r}$ converges to the correct solution, a comparison of $\hat{r}$ with the GPS-tracked range shows a normal, zero-centered distribution within the bounds of GPS drift.
+The present analysis therefore seeds realistic but ``omniscient'' knowledge of the extrapolated range and leverages the post-processing pipeline to more thoroughly evaluate the acoustic range estimate for all modem events, with three relevant sound speed sources, and both group velocity criterion.
+Accordingly, the results in this section evaluate the utility of the algorithms and sound speed sources, divorced from their role in the ICNN while maintaining real-time relevance.
+
 \subsection{Nearest bounce criteria (NBC)}
+
+\llabel{2.12} As shown in the eigenray traces of Fig. \ref{fig:raytrace-zs30}, the extent of ray bending and repeated reflections is extremely dependent on the sound speed profile.
+An isovelocity approach would completely miss this nuance; our field-tested approach that only resolved the simplest path is unlikely to resolve the one that triggers modem detection.
+Based on this insight, a new algorithm, the nearest bounce criteria (NBC), is a slight modification from the MBC and includes multipath as a new dimension of information to exploit.
+This metric, while run in post-processing, adds a negligible amount of computation for real-time efficacy.
+
+Given a running estimate for the horizontal group velocity $u_{i,j}$ between nodes $i$ and $j$, the navigation system has an extrapolated value for range, $\hat{r}$, in the virtual layer, and a recorded travel time, $\Delta t_{i,j}$, in the physical layer.
+Instead of using only the $N_0$ rays with neither surface nor bottom bounces to estimate group velocity and subsequently moving to incremental number of bounces only when no valid direct path solutions exist, we solve for the power weighted average of the ray travel time for the $N_k$ rays with $k$ bounces,
+\begin{equation}
+t_k = \frac{\sum_{n=1}^{N_{k}} dt_{n}a_{n}^{2}}{\sum_{n=1}^{N_{k}} a_{n}^{2}} ~, 
+\end{equation}
+find the nearest matching power weighted average to recorded travel time,
+\begin{equation}
+t_{i,j,k} = \min_{k=0,1,2,...} \left| t_k - \Delta t_{i,j} \right|
+\end{equation}
+predict a group velocity,
+\begin{equation}
+u_{i,j} = \dfrac{\hat{r}}{t_{i,j,k}}
+\end{equation}
+and estimate the range as was done previously.
+\begin{equation}
+r_{i,j} = u_{i,j}\Delta t_{i,j}
+\end{equation}
+
+This method selects a different group velocity based on the multipath arrival structure, as the detected arrival is not always the first arrival or the direct path and could even be masked by noise or blocked temporarily \citep{deffenbaugh_acoustic_1996}.
+We manually cap the number of bounces to 4 because of the smaller operational scale and the attenuation accrued with many surface interactions.
+Bottom bounces are not encoded separately because of ray's tendency to refract upward, not due to information limitations.
 
@@
 
+The minimal and nearest bounce algorithms are applied with the three sound speed inputs shown in Fig.~\ref{fig:raytrace-zs30}.
+The resulting predicted group velocities for all source depths are shown in Fig~\ref{fig:gvel-post}.
+
 \begin{figure}[h!]
-\includegraphics[width=\columnwidth]{figs/Fig5.pdf}
+\includegraphics[width=\columnwidth]{figs/gvel-txrxdepth-wIso-wGPS.pdf}
 \caption{A comparison of group velocity predictions for all beacon to beacon events in post-processing with a source depth of 30 m, with group velocity on the y-axis and recorded travel time on the x-axis.
@@
 The minimal and nearest bounce criterion are distinguished by different marker shapes, compared to the separately colored red dots showing the naive, data-driven group velocity calculation.}
-\label{fig:gvel30}
+\label{fig:gvel-post}
 \end{figure}
+
+The goal of the group velocity estimation is to converge towards the implied sound speed, i.e. the GNSS-derived range divided by the recorded travel time.
+For a 30 m receiver depth, the NBC shows more overlap with data-derived values as it classifies multipath more correctly.
+For a 90 m receiver depth, the overlap is less accurate due to computational constraints of a limited fan of rays entering the shadow zone rendering a less reliable simulated times of arrival packet.
+
+Overall, analysis shows that the higher multipath classification produces more accurate sound speed predictions, likely driven by a tighter and/or sparser bundle of rays.
+However, that data are too small to draw significant conclusions.
+\llabel{1.19} Discontinuities in multipath classification verify our hypothesis for its importance to a smoothly varying group velocity, as shown in the cluster for a receiver depth of 30 m, where HYCOM jumps from one to two bounces amidst the baseline SSP and weighted SSP smoothly increasing while consistently seeing two and three bounces, respectively.
 
@@
 \begin{figure}[!ht]
-\includegraphics[width=\textwidth]{figs/Fig6.pdf}
+\includegraphics[width=\textwidth]{figs/range-error-allMethods.pdf}
 \caption{The post-processed range error for source depths of 20, 30, and 90 m, and receiver depths of 30 and 90 m.
@@
 \end{figure}
+
+Fig. \ref{fig:rangeError} shows the directional range error footprints for all three sound speed inputs with respect to OWTT, separated by source and receiver depth configurations.
+The weighted SSP range error generally has the smallest and most zero-centered footprint.
+The one case it does not is for the source-receiver pairings between 30 and 90 m in depth.
+The increased error for these reciprocal transmission paths is most likely driven by the computational artifacts encountered when propagating through the steep sound speed gradients of the lens and through the shadow zone.
+All other source depth pairings are significantly improved using the chosen weights compared to HYCOM or the baseline.
+
+When using a linear scaling to convert travel time into range, any offset between the assumed sound speed and the horizontal group velocity produces unconstrained error with increasing receiver distance.
+Most importantly, we see the consequences of the adaptive group velocity in that range error does not strictly increase with OWTT.
+
+The improvement from MBC to NBC is most evident for the realistic sound speed; while the HYCOM SSP improves from a median absolute range error of 6.41 to 4.61 m, the baseline SSP improves from 10.30 to 2.27 m, and the weighted SSP improves from 13.28 to 2.12 m.
+Table \ref{tab:rangeErrorV1V2} shows further statistics on the absolute range error by SSP and group velocity algorithm.
+The order of magnitude improvement in the ducted SSPs demonstrate the effectiveness of the algorithm exploiting the multipath conditions.
+
+\begin{table}[h!]
+\renewcommand{\arraystretch}{1.5}
+\centering
+\begin{tabular}{r|cc|cc|cc}\toprule
+ & \multicolumn{2}{c|}{\textbf{Baseline} } & \multicolumn{2}{c|}{\textbf{Chosen Weights} } & \multicolumn{2}{c}{\textbf{HYCOM}} \\
+  & \cellcolor[HTML]{EFEFEF}MBC & NBC & \cellcolor[HTML]{EFEFEF} MBC& NBC & \cellcolor[HTML]{EFEFEF}MBC & NBC \\ \hline
+ minimum [m] 	& \cellcolor[HTML]{EFEFEF}0.01 & 0.00	& \cellcolor[HTML]{EFEFEF}0.00 	& 0.00 	& \cellcolor[HTML]{EFEFEF}0.11 & 0.01 \\
+ 25th \% [m]   & \cellcolor[HTML]{EFEFEF}4.96 & 0.99	& \cellcolor[HTML]{EFEFEF}6.26 	& 0.95 	& \cellcolor[HTML]{EFEFEF}3.30 & 2.25 \\   
+ median [m]		& \cellcolor[HTML]{EFEFEF}10.30 & 2.27 	& \cellcolor[HTML]{EFEFEF}13.28 & 2.12 	& \cellcolor[HTML]{EFEFEF}6.41 & 4.61 \\
+ 75th \% [m]   & \cellcolor[HTML]{EFEFEF}15.81 & 5.51 	& \cellcolor[HTML]{EFEFEF}19.75 & 4.11 	& \cellcolor[HTML]{EFEFEF}10.92 & 7.46 \\
+ maximum [m]    & \cellcolor[HTML]{EFEFEF}22.52 & 14.96 & \cellcolor[HTML]{EFEFEF}1491  & 20.21 & \cellcolor[HTML]{EFEFEF}19.55 & 15.81 \\
+ \toprule
+\end{tabular}
+\caption[Comparison of post-processing range estimation algorithms across all events]{A comparison of range estimation metrics for each sound speed source and group velocity estimation algorithm for all 1283 beacon to beacon events via post-processing.
+The 0th (minimum), 25th, 50th (median), 75th, and 100th (maximum) percentiles are shown to the range resolution afforded by the WHOI Micro-Modem.
+There are a few outliers that drive the mean to be higher than the median.}
+\label{tab:rangeErrorV1V2}
+\end{table}
+
+As shown in table \ref{tab:rangeErrorV1V2}, there is a striking maximum range error of 1491 m for the weighted SSP in the minimal bounce criteria.
+There are 10 events from South transmitting at 30 m depth to North receiving at 30 m depth.
+The OWTT spread is from 2.1958 to 2.1963 s; the naive group velocity is 1429.3 to 1430.1 m/s; and the GPS-tracked range is from 3138.54 m to 3140.87 m.
+This example ends up being an excellent case study for how sound speed and multipath fidelity work in concert to minimize range error.
+The large error in this instance is driven by the MBC unexpectedly defaulting to a bottom bounce with a much greater OWTT.
+The NBC classifies the multipath as 4 bounces, reducing the range error from greater than a kilometer to less than a meter.
+While there is no actual way of knowing if this is the correct multipath structure, the range error is remarkably small, at 0.025\%. 
+This pattern of not choosing the minimal observed bounce structure is consistent across all SSPs; the baseline goes from 1 to 3 bounces and HYCOM goes from 0 to 2 bounces.
+Notably, the baseline and HYCOM range errors are never egregiously large, but are nonetheless improved with the NBC algorithm.
+Thus, for acoustically complex environments, the NBC has a disproportionately positive impact as the estimated SSP approaches the desired SSP.
 
@@
 
+The fact that the bulk of the re-localization error exists within the precision of the GNSS unit deployed invites a further look into GNSS noise.
+In the Arctic, GNSS performance worsens due to poor constellation coverage, larger ionospheric effects, and multipath interference.
+The National Security Implications of Climate Change for U.S. Naval Forces \citep{NAP12914} details some of the limitations of the Global Positioning System (GPS) at polar latitudes.
+Radio infrastructure that provides position corrections and references does not regularly extend to polar regions.
+The effect is minor for surface platform navigation \textemdash roughly 15 m of horizontal precision has been displayed at the North Pole\textemdash but is significant enough to register against the modem's detected travel times.
+Figure \ref{fig:gps-drift-example} zooms in on the GNSS and OWTT noise relative to the ice movement for two pairs of modem buoy connections.
+The two panels indicate the GPS drift as $\delta R = \sqrt{\delta x^2 + \delta y^2}$ and temporal drift, $\delta t$, relative to the median OWTT recorded between the two modems.
+The dashed line is scaled by a group velocity of 1440 m/s, such that if there were ideal sensor measurements with no drift, all events should exist on or near the line.
+
 \begin{figure}[h!]
@@
 \end{figure}
+
+The top panel shows the connections between the North and East buoys.
+There is relative, i.e. non-rigid, ice movement between the North and East buoys, evidenced by events spanning the dashed line.
+But the height of the scatter plot is indicative of the precision of the GPS signal, as it remains consistent across many arrival time bands.
+Naturally, some minor offsets between these vertical bands relate to different operational configurations of source and receiver depth.
+However, the large majority of events show vertical banding for the same nominal $\delta t$, indicating the amount of GPS drift.
+
+This idea of GPS drift relative to the same OWTT measurements is further indicated by events between the other two buoys, South and West, in the bottom panel.
+These buoys are moving in a more rigid ice floe and there is minimal impact by source and receiver depth on the spread of OWTT.
+The GPS drift is much larger relative to the OWTT drift, which is sensitive to acoustic scattering, multipath, and/or environmental microstructure.
+
+These are just two subsets of the physical links that cover all four GPS modem buoys.
+The GPS at camp is the least accurate due to the human activity and infrastructure occluding the physical puck.
+
+\section{Discussion}
+
+Given the computational constraints of real-time modeling, the gridded approach facilitates enough multipath classification to build in a ``ray ensemble'' of characteristic group velocities.
+This result is not always possible when aiming to find eigenrays to just an individual point, even with a higher density of launch angles.
+An important takeaway for those interested in ray-based localization is leveraging a local grid to give BELLHOP tolerance for finding solutions that otherwise may not be found in a center or single point solution.
+The limitations of numerical computation, particularly for a complex environment, are more adeptly addressed by accepting some uncertainty in position than by prescribing an exact solution.
+Even though BELLHOP is a greedy estimator for eigenrays, the additional data created is a negligible burden.
+
+Underwater navigation research is broadly motivated by acquiring GPS-like navigation accuracy in GPS-denied conditions.
+The Arctic, while remote, is the perfect place to test mature navigation technologies in real GPS-denied conditions.
+
+Range estimation is an essential step of acoustic localization and navigation.
+Current approaches in real-time underwater acoustic navigation simplify the non-linear relationship between a sound speed profile and acoustic propagation with a deterministic sound speed.
+Some post-processing approaches attempt to leverage the acoustic arrival structure via various ray methods, but often use a singular SSP for simplicity, even over long term missions or dynamic conditions.
+Thus, the conversion from travel time to range, particularly for real-time vehicle deployments, can be pre-conditioned for error growth as the OWTT/range increases.
+
+For GPS-denied navigation, especially in hostile environments like the Arctic, tolerance for error is close to none.
+This work addresses a critical need in acoustic navigation by retooling acoustic arrival methods generally deemed too complex or labor intensive for real-time, ray-based range estimation to achieve GPS-like positioning.
+
+We hypothesize and validate that the embedded stochastic prediction of a single group velocity is a smoothly varying function of range, source and receiver depth pairings, as well as multipath structure.
+We employ a GPS-tracked experiment to have a ground truth comparison for real-time localization algorithms.
+The real-time system achieves GPS-like navigation for an AUV without taking into account multipath structure; the ranging error improves by an order of magnitude with the suggested multipath adaptability, minimizing range error to single meters.
+Post-processing analysis shows that this method of ranging is sensitive to GPS drift itself; thus embedding a stochastic prediction of the horizontal group velocity has an outsized benefit to minimizing trilateration error.
+
+There are many avenues through which this approach can be further refined and tested for field operations.
+Amongst them is defining the uncertainty grid for BELLHOP via stochastic or data-driven measures such as the distance traveled by the AUV between ICNN updates or the magnitude of the position corrections by the ICNN.
+Another is to entirely remove the seeded range and instead rely on the submerged asset's depth and recorded OWTT to find high probability fields in range.
+
+The literature in underwater acoustic navigation and positioning is either real-time or physics-based.
+In this paper we demonstrate a field-tested approach that is both real-time and physics-based; this is achieved by coupling data streams with fast acoustic modeling.
+The methods exploit the upward refracting nature and the total ice cover of the Arctic environment to achieve remarkable ranging accuracy and precision.
+It transforms multipath, widely considered as an obstacle for acoustic ranging, into a new information content to refine ranging accuracy.
+We believe that this work enables more accurate range estimation, localization, and/or navigation for any field experiment given known source and receiver depths.
+
+\subsection{Broader relevance \& future work}
+
+Performance in other acoustic environments may require introducing a different thresholded metric to sort time of arrivals.
+Given the NBC algorithm's improvement with increased multipath, its effectiveness is likely only challenged by the valid operational scales of a range independent propagation environment.
+For mesoscale operations, like that of many gliders, the group velocity criteria may need to be modified to better account for variability driven by range dependent propagation through internal waves, eddies, or even bathymetric changes like underwater volcanoes.
+BELLHOP simulations provide other relevant eigenray information, like time and angle of arrival, that is ripe for statistical and machine learning methods to classify a representative group velocity.
+A bespoke and fast ray tracing method, which was used during ICEX20 to evaluate the acoustic fit of sound speed profile parameterization \citep{bhatt_embedded_2021}, can easily report back the number of turning points instead of the number of bounces for multipath classification.
+
+This approach will start to break down in extremely dynamic environments.
+Fast moving fronts, as seen in estuaries like the Connecticut River into the Long Island Sound, present an entirely new set of challenges not seen by internal waves or eddies.
+Buoyancy fluctuations in these regions threaten even simple AUV tasks like following a trackline.
+Acoustic communications are further complicated given a shallow environment with significant scattering \citep{lavery_measurements_2010,ross_acoustic_2012,lavery_broadband_2013}, where fast acoustic modeling may only be coherent for trivial probabilities of the ocean state.
+Realistic \textit{in situ} considerations of the acoustic environment may not be possible in such environments without complete through-the-sensor integration of echosounder data and/or a hyper-realistic onboard ocean model.
+
+\llabel{1.20} Many approaches to underwater navigation combine it with acoustic tomography, i.e., a joint estimation of both source and receiver locations and the ocean volume between them.
+There has been considerable success at this effort in post-processing methods, which utilize intensive\textemdash and due to the non-linearity of sound propagation, often brute force\textemdash computational methods.
+For vehicle operations, fast tomography is the ideal implementation, in that one can fully consider how sound speed structure, horizontally and vertically, influences sound propagation.
+AUVs can serve as moving sources to better image the ocean volume \citep{deffenbaugh_optimal_1997,elisseeff_ocean_2002}, where mobile tomography and navigation converge on the same set of component technologies: position estimation, sound speed parameterization estimation, ray path identification, and vehicle path optimization.
+
+But there are overwhelming challenges, operationally and computationally, for fast, mobile tomography to become a realistic endeavor.
+Addressing the spatial and temporal scales of what can be solved deterministically and what must be solved stochastically imposes a resolution constraint on the utility of gridded models\textemdash resolving fine features inaccurately (or with a false sense of confidence) could be more harmful than assuming range independence.
+Given that AUV operations are often on small spatial and temporal scales, the added benefit of a gridded model is quite small, and in cases like the Arctic, may actually mischaracterize the ocean volume.
+For gliders, with longer and larger operational scales, an ocean model may provide more useful information.
+Currently gliders are low power and do not have the storage or computational power to run a full-scale, realistic ocean model.
+A lightweight representation of the key environmental and acoustic features, passed through the same manner of acoustic message from the modem experiment, may drastically improve glider navigation.
 

### 2022-02-20 manuscript/main.tex
completely filled in new format -- likely has redundant paragraphs

@@
 \author{Henrik Schmidt}
-\email{henrik@mit.edu}
 \affiliation{Department of Mechanical Engineering, Massachusetts Institute of Technology, Cambridge, MA}
@@
 \date{\today}
-
-% The motivation behind this work is to improve the standard LBL navigation’s conversion of travel time to range for an under-ice environment in a way that could be generalized to any acoustic waveguide.
-To do this, we introduce the Minimal Bounce Criteria method to operationalize real-time acoustic modeling into an effective horizontal group velocity estimate.
-
-% 	This system was implemented on an autonomous underwater vehicle (AUV) on multiple missions, including an 11 km untethered, under-ice deployment, in March 2020.
-The AUV was deployed through a hydrohole from Topside and due to a disk error, stalled underneath the ice surface while transmitting its perceived location.
-The AUV was found within a meter away from its reported location, which serves as strong (but qualitative) evidence of the navigation performance provided by the single group velocity calculation.
-Because the navigation updates throughout the AUV mission have no GPS ground truth to compare to, this paper evaluates the same system performance across the GPS-navigated LBL beacons.
-
-% 	The real-time error from the Nearest Bounce Criteria (NBC) achieves a mean absolute positioning error of 11 m.
-Upon further evaluation of the acoustic events, the Minimal Bounce Criteria (MBC) is run in the same automated pipeline in post-processing (akin to real-time computational constraints and needs) and achieves an error of roughly 3 m.
-
-% 	Both the NBC and MBC pseudorange estimations are compared to the GPS data on the LBL beacons.
-The error itself is small and the pattern of error exhibited indicates that the GPS ground truth data has significant noise (drift) that is not replicated by either pseudorange estimation method.
-For the MBC in particular, the post-processing error is within the precision of the GPS sensor, suggesting that its performance rivals that of GPS.
 
@@
 \llabel{2.3}
-
 \end{abstract}
@@
 The results from this paper derive from data taken while deploying the AUV \emph{Macrura}, a custom Bluefin-21, during the Ice Exercise 2020 (ICEX20), in the Beaufort Sea, from March 8 to 11, in double-ducted and ice-covered conditions.
-The AUV deployment was supported by the Integrated Communication and Navigation Network (ICNN) \citep{schneider_self-adapting_2020,randeni_construction_2020,randeni_high-resolution_2021}.
+The AUV deployment was supported by the Integrated Communication and Navigation Network (ICNN) \citep{schneider_self-adapting_2020,randeni_construction_2020,randeni_high-resolution_2021}, a specialized implementation of the LBL solution.
 The ICNN was initially developed via numerous virtual experiments to push the boundaries of algorithms and interfaces between different hardware components.
@@
 The AUV and each ice buoy are outfitted with a WHOI Micro-Modem \citep{singh_underwater_2006}, with a four-element receiver array and a single transmitter, one-tenth of a millisecond resolution.
-Acoustic messages were sent with a 10 kHz carrier frequency, 5 kHz bandwidth, and phase-shift keying modulation on a thirty-second cycle, giving room for two-way communication throughout the mission volume.
+Acoustic messages were sent with a 10 kHz carrier frequency, 5 kHz bandwidth, and phase-shift keying (PSK)modulation on a thirty-second cycle, giving room for two-way communication throughout the mission volume.
 The receive and transmit elements were split between shallow and deeper depths (30 and 90 m, respectively) to provide better coverage across the shadow zone.
@@
 Figure \ref{fig:overview} shows successful events sorted by source depth.
-In this analysis, we assume there is insignificant displacement between the GNSS puck surface expression and subsurface modem; this is supported by unusually low ice drift rates, on average, just 0.7 cm/s.
+In this analysis, we assume there is insignificant displacement between the GNSS puck surface expression and subsurface modem; this is supported by unusually low observed ice drift rates, on average, just 0.7 cm/s.
 
@@
 
-Instead, for each individual node $i$, an embedded stochastic tracking framework is used to provide a running estimate of the horizontal group velocity $u_{i,j}$ for the conversion from travel time to range from modem $j$, with the ultimate goal of matching the naive group velocity, i.e.
+Instead, for each individual modem $i$, an embedded stochastic tracking framework is used to provide a running estimate of the horizontal group velocity $u_{i,j}$ for the conversion from travel time to range from modem $j$, with the ultimate goal of matching the naive group velocity, i.e.
 the GPS-recorded distance between two nodes divided by the modem-recorded one way travel time between them.
@@
 \llabel{2.11} The initial call to BELLHOP is over a local grid centered at the range and depth posited by the onboard tracking solution.
+Compared to a point solver, solving for arrivals on a grid minimizes the impact of numerical resolution in the propagation model.
+The resulting ray traces may represent paths that would reach a receiver in a real application but might not be recognized by the numerical solver as a viable detection.
 The grid is initialized as 11$\times$11 points spanning 10 m horizontally and 20 m vertically.
@@
 
-This stochastic, ensemble method for group velocity calculation can run in real-time, appearing to be orders of magnitude faster than other post-processing methods which seek to determine the specific ray itself that best matches a prominent indicator from the arrival structure.
-The BELLHOP simulation that runs this calculation uses 3600 rays with launch angle fan of -60 to 60 degrees, a depth dependent sound speed profile, and a range dependent bathymetry. \llabel{1.13}
-Even though the NBC looks for just the least complex multipath, the high density of launch angles almost always guarantees a direct path.
-The MBC improves upon a deterministic estimate in that it considers the source and receiver depth to predict a more realistic group velocity.
+Thus the NBC method assumes the signal detected by the modem will be dominated set of paths with the least number of boundary interactions.
+Importantly, this stochastic, ensemble method for group velocity calculation can run in real-time, appearing to be orders of magnitude faster than other post-processing methods which seek to determine the specific ray itself that best matches a prominent indicator from the arrival structure.
+The BELLHOP simulation that runs this calculation uses 3600 rays with launch angle fan of -60 to 60 degrees, a representative depth dependent sound speed profile, and a range dependent bathymetry.
+\llabel{1.13}
 
@@
 The overarching results show that sound speed estimates derived from eigenrays for a local grid, as opposed to a singular point, are accurate enough to support vehicle navigation.
+While the NBC looks for just the least complex multipath, the high density of launch angles almost always guarantees a direct path.
 Nonetheless, the consistent overestimation of pseudorange invites further analysis into acoustic arrival matching.
@@
 
-Fig. \ref{fig:raytrace-zs30} shows eigenrays in a range independent coordinate system for all three sound speed environments, for a source depth of 30 m.
-\llabel{1.16} It is important to note that BELLHOP produces an \llabel{2.8}abundance of eigenrays; the ones displayed here are chosen by travel time proximity to the recorded data.
-With this filter, the chosen eigenrays show a greater amount of surface interactions as the SSP duct conditions increase.
-
-\llabel{2.14} This is likely driven by the prominence of the duct.
+Accounting for ice movement between beacons creates nominal ranges with small variability.
+Figs. \ref{fig:raytrace-zs20,fig:raytrace-zs30,fig:raytrace-zs90} each show eigenrays for three sound speed environments for source depths of 20, 30, and 90 m, respectively.
+Eigenrays were initially found using the built-in BELLHOP protocol with a launch angle step of 0.05 degrees from -60 to 60 degrees.
+Separately, recorded travel times between beacons were clustered with 1 millisecond boundaries such that some source-receiver pairs had multiple, distinct travel times to approximate.
+The BELLHOP eigenray returns\llabel{2.8} were then filtered such that one was selected per travel time cluster, in the hopes that the eigenray will converge to the receiver locations for the most realistic sound speed input.
+\llabel{1.16}
+It should be noted that bottom bounces were recovered but filtered out.
+The three source depths create distinguishable ray geometries with respect to the three sound speed inputs.
 
@@
 
+For a source at 20 m in depth, shown in Fig. \ref{fig:raytrace-zs20}, reliable eigenrays are found for all sound speed inputs.
+Rays refract upwards and neatly intersect with the shallow and deep receiver locations between 1.5 and 1.8 km in range.
+However, the ray paths for the shallow receivers change both in the number of surface interactions and where the surface interactions occur with respect to range across the SSPs.
+As the Beaufort Lens strengthens, the chosen paths to the second farthest shallow buoy (North, in red) interact with the surface more and become distinct.
+The weighted SSP shows the most interesting effects for the deeper receivers.
+The ray paths all interact with the surface and the eigenrays for the Northern (red) and Western (green) buoys are in fact the same ray.
+
 \subsubsection{Source depth of 30 m}
@@
 \end{figure}
+
+The ray geometries from the 30 m source, shown in Fig. \ref{fig:raytrace-zs20}, show a similar degradation of eigenray identification with increased ducting.
+Receptions span 1.5 to 3.2 km.
+Once again, eigenrays for HYCOM and the baseline SSP are visually appropriate.
+Rays for the weighted SSP show how the surface channel intensifies ice interactions and how the shadow zone denies reliable acoustic paths.
+Pointedly, the increasing number of surface reflections to the farthest shallow buoy (North, in red) crystallize the MBC's tendency for overestimation.
+For the HYCOM, baseline, and weighted SSP inputs, the most appropriate eigenrays show 2, 3, and 4 surface interactions.
 
@@
 
+Lastly, Fig. \ref{fig:raytrace-zs90} shows ray geometries from the 90 m source, elucidating a different extent of the shadow zone. 
+While the receiver locations are similar to that of the 30 m source depth, the deeper source depth effectively negates the upper duct and places the upper (and some of the lower) receivers in unreliable acoustic paths.
+The HYCOM eigenrays still show the most reliable acoustic paths that deteriorate with increasing ducted conditions.
+The lack of direct paths from the observed SSP further points out the shortcomings of the MBC approach.
+
+The goal of the MBC algorithm was to provide a reliable, physically intuitive interpretation of the acoustic propagation without taking on the additional burden of regularly identifying specific paths that may connect any given source-receiver pair in the network.
+Its performance was adequate for vehicle navigation and would have likely sufficed if it were not for the prominence of the duct observed relative that of other model and data products. \llabel{2.14}
+
 \clearpage
@@
 
-There are 1242 total beacon-to-beacon events in the modem experiment if real-time ranging can be substituted for post-processed ranging.
+From all events recorded during the modem test experiment, there are 1242 successfully decoded beacon-to-beacon events.
+Only these events are used to evaluate ranging accuracy, as the ICNN was not configured to use receptions with failure flags.
 Thus, a post-processing analysis that emulates the real-time navigation engine was run to overcome the unequal distribution of communication events with respect to depth, range, and sound speed status.
@@
 
-Given a running estimate for the horizontal group velocity $u_{i,j}$ between nodes $i$ and $j$, the navigation system has an extrapolated value for range, $\hat{r}$, in the virtual layer, and a recorded travel time, $\Delta t_{i,j}$, in the physical layer.
+Given a running estimate for the horizontal group velocity $u_{i,j}$ between nodes $i$ and $j$, the navigation system has an extrapolated value for range, $\hat{r}$, and a recorded travel time, $\Delta t_{i,j}$.
 Instead of using only the $N_0$ rays with neither surface nor bottom bounces to estimate group velocity and subsequently moving to incremental number of bounces only when no valid direct path solutions exist, we solve for the power weighted average of the ray travel time for the $N_k$ rays with $k$ bounces,
@@
 
-Overall, analysis shows that the higher multipath classification produces more accurate sound speed predictions, likely driven by a tighter and/or sparser bundle of rays.
+As the environmental and ray filtering method become better representations of the real ocean, the lower the expected mismatch is between the implied and estimated effective sound speeds.
+Analysis shows that the higher multipath classification produces more accurate sound speed predictions, likely driven by a tighter and/or sparser bundle of rays.
 However, that data are too small to draw significant conclusions.
@@
 
+To overcome potentially intermittent acoustic communication, the operational paradigm of the ICNN computes corrections relative to the trilaterated position estimates transmitted by the vehicle, rather than transmitting the updated positions themselves.
+The reliability of the correction is directly linked to how accurately the travel time measurements are converted to pseudoranges.
+This section aims to resolve that tension by reevaluating the trilateration results with respect to the MBC and NBC algorithms.
+The MBC/NBC sound speed estimates were tracked independently for each transmitter-receiver pair; although the sound speed was expected to be locally smooth near a given receiver, no such assumption was enforced between distinct acoustic links.
+
 \subsection{Re-positioning beacon to beacon events}
+
+When the beacons ran as virtual vehicles, the ICNN did not have access to that buoy's GPS data stream except for what was sent via digital acoustic message.
+The static nature of the experiment means that the initial estimate transmitted to the ICNN was in fact a ground truth position.
+Therefore, a distribution of corrections from the ICNN, as shown in Fig. \ref{fig:trilat-beacon}, reflects positioning accuracy.
+The NBC clearly outperforms the MBC, with almost 80\% of the corrections below 6 meters and the median within the deployed GNSS puck precision of 3 meters.
+By contrast, the MBC shows roughly 20\% within the GNSS puck precision, and separate peaks from 9--12 meters and 21--27 meters.
+This trimodal nature reflects the distribution of reflections on the ice surface.
 
@@
 \includegraphics[width=\textwidth]{figs/beacon-trilat-stat.pdf}
-\caption{WRITE THIS.}
+\caption{Trilateration events; however, 264 entries are 2-beacon solutions, 22 are 3-beacon, and 2 are 4-beacon.
+When limited to only two ranging circles, the trilateration chooses the closer of the two intersection points.
+The x-axis is cut off at the upper limit for the NBC algorithm.}
 \label{fig:trilat-beacon}
@@
 
+In several events, the MBC is unable to accurately estimate the effective sound speed for one of the acoustic links, leading to a large positioning error.
+The NBC, however, better resolves an approximation of the acoustic path.
+For example, in some trilateration solutions for the Eastern buoy, the MBC shows a correction of more than a kilometer; the NBC is two order of magnitudes less.
+
 \subsection{Re-navigating AUV Macrura}
+
+Up to this point, pseudorange estimation and localization have been evaluated on GPS-linked beacon-to-beacon connections to validate the NBC algorithm.
+This analysis ports the MBC and NBC algorithms to re-navigate the AUV \emph{Macrura}.
+
+The AUV dataset clearly exhibits instances where a receiver detects the same transmission more than once.
+This is not surprising considering the complex multipath provided by the Beaufort Lens.
+The 11 hour vehicle mission contains 3260 transmissions, 12938 total detections, and 4704 successful receptions.
+Allowing receptions with PSK errors would almost double the number of recorded multipath arrivals exploited for positioning, if a real-time solution could correctly parse paths from different arrivals in the same thirty-second cycle.
+Thus it remains a future endeavor to explore how failure mode information from acoustic modems could be used to identify unsuccessful but otherwise trustworthy arrivals to augment trilateration samples.
+
+The following performance analysis is constrained to what the vehicle acted on in real-time.
+AUV \emph{Macrura} and the ICNN ran an adaptive depth behavior to maintain acoustic communication on the insight that cross-layer links were more likely to fail than same-layer ones.
+Accordingly, the vehicle dove deeper than 50 meters about 20\% of the time it was underway.
+
+In contrast to the modem tests, where position correction illustrated re-positioning accuracy, the re-navigation corrections are less valuable in the absence of GNSS ground truth.
+The correction magnitude necessarily depends on the vehicle's internal navigation estimate, which is prone to larger errors from sensor drifts, ocean currents, and other error not captured in the hydrodynamic model.
+Thus, larger corrections are not necessarily indicative of worse performance.
+Navigation accuracy may be better described by trilateration error, the RMS of the remaining pseudorange errors from each acoustic link.
+
+Fig. \ref{fig:trilat-auv} shows the correction magnitudes and trilateration errors for events with three or more receptions during AUV operations.
+Whereas the MBC has a fairly bimodal nature, with peaks centered around 10--15 and 35--40 m, the NBC favors smaller corrections, from 5--20 m, and has a long tail.
+The distribution of corrections are much larger than the distribution of RMS error.
+It is apparent that, while both methods are quite successful, there is strong evidence that the NBC achieves single meter accuracy.
 
@@
 \includegraphics[width=\textwidth]{figs/auv-trilat-stat.pdf}
-\caption{WRITE THIS.}
+\caption{Distribution of vehicle re-navigation corrections (top) and error (bottom) computed in post-processing for the Minimum and Nearest Bounce Criterion.}
 \label{fig:trilat-auv}
@@
 
-The fact that the bulk of the re-localization error exists within the precision of the GNSS unit deployed invites a further look into GNSS noise.
+The fact that the bulk of the best performing re-navigation error exists within the precision of the GNSS unit deployed invites a further look into GNSS noise.
 In the Arctic, GNSS performance worsens due to poor constellation coverage, larger ionospheric effects, and multipath interference.
@@
 
-\subsection{Broader relevance \& future work}
-
 Performance in other acoustic environments may require introducing a different thresholded metric to sort time of arrivals.
@@
 This work was supported by the Office of Naval Research 322-OA under ICEX20 (N00014-17-1-2474) and Task Force Ocean (N00014-19-1-2716).
-
 \end{acknowledgments}

### 2022-02-20 rebuttal/rebuttal-letter/trunk-rebuttal.tex
some minor comments resolved

@@
 We have refashioned the abstract and introduction to better elucidate the motivation for this work as a model-aided component for LBL navigation in total under-ice conditions, which could be further generalized for any acoustic environment.
-Based on several suggestions, we have restructured the manuscript and separated the real-time navigation methodology (used the Minimum Bounce Criteria, MBC, but is impossible to compare given the lack of GNSS), the real-time positioning results (MBC), and the post-processing re-positioning improvement (Nearest Bounce Criteria).
+Based on several suggestions, we have restructured the manuscript and separated the real time navigation methodology (used the Minimum Bounce Criteria, MBC, but is impossible to compare given the lack of GNSS), the real-time positioning results (MBC), and the post-processing re-positioning improvement (Nearest Bounce Criteria).
 
@@
 	\lreviewer{line 205}{v1:1.13}{What do you mean that the coordinate system is range independent?}
+	We mean that the propagation is range independent, with one sound speed profile. It does actually take into account bathymetry.
 
@@
 	\lreviewer{line 241}{v1:2.9}{'modem triggering time', do you mean 'detected arrival time'?}
+	Yes. We have changed this accordingly.
 
@@
 
-	{\color{blue} The authors are not sure what the reviewer means by this statement, specifically what ``navy range'' is.
-We believe we are only using ``range'' to mean distance.
-Nonetheless, we have adopted this suggestion and use ``navigation'' or ``communication'' system/network instead.}
+	We have adopted this suggestion and use navigation'' or communication system instead.
 

### 2022-02-20 manuscript/main.tex
initial clean up of refs

@@
 While land and air-based robots utilize information from Global Navigation Satellite Systems (GNSS) to achieve stunning location accuracy and precision throughout the duration of their missions, AUVs cannot access GNSS while underwater due to the rapid attenuation of electromagnetic waves.
-Therefore, underwater vehicles have relied on any combination of dead reckoning, hydrodynamic models, inertial navigation systems, doppler velocity logs, and acoustic baseline positioning systems for navigation \citep{paull_auv_2014}.
+Therefore, underwater vehicles have relied on any combination of dead reckoning, hydrodynamic models, inertial navigation systems, doppler velocity logs, and acoustic baseline positioning systems for navigation \citep{Paull2014}.
 Limiting navigation error and drift requires an AUV to periodically stall on the surface and obtain a GNSS fix to reset its position error.
@@
 % introduce LBL and precise statement of work
-Of underwater acoustic navigation systems, long baseline (LBL) is the most GPS-like in style and scale, and most appropriate for mitigating drift without overburdening computation or payload size on the vehicle \citep{paull_auv_2014,van_uffelen_global_2021}.
-The state-of-the-art for LBL outsources depth to a pressure sensor and solves the two-dimensional localization problem with an isovelocity, linear scaling between one way travel time (OWTT) and range \citep{eustice_recent_2006,eustice_experimental_2007,webster_preliminary_2009,webster_advances_2012}.
+Of underwater acoustic navigation systems, long baseline (LBL) is the most GPS-like in style and scale, and most appropriate for mitigating drift without overburdening computation or payload size on the vehicle \citep{VanUffelen2021}.
+The state-of-the-art for LBL outsources depth to a pressure sensor and solves the two-dimensional localization problem with an isovelocity, linear scaling between one way travel time (OWTT) and range \citep{Eustice2006,Eustice2007,Webster2009,Webster2012}.
 This assumption is valid for short scale operations but oversimplifies propagation for larger and/or complex acoustic environments.
@@
 % PNT
-For clarity, we delineate specific definitions for timing, positioning, and navigation from \citet{howe_observing_2019}.
+For clarity, we delineate specific definitions for timing, positioning, and navigation from \citet{howe2019observing}.
 \begin{enumerate}
@@
 % More about LBL navigation
-While RAFOS floats championed one way ranging for re-positioning \citep{rossby_rafos_1986,duda_evaluation_2006}, the ability to do so for navigation was facilitated by \llabel{1.11} the advent of the WHOI Micro-Modem \citep{singh_underwater_2006} and synchronized chip scale atomic clocks \citep{gardner_second_2016}.
-AUV navigation efforts have achieved root mean square (RMS) localization error on the order of tens of meters relative to GNSS surface position over less than ten kilometers in shallow \citep{eustice_experimental_2007,kepper_mems_2017,claus_closed-loop_2018} and deep water \citep{kunz2008deep,jakuba2008long,webster_preliminary_2009}.
+While RAFOS floats championed one way ranging for re-positioning \citep{Rossby1986,Duda2006}, the ability to do so for navigation was facilitated by \llabel{1.11} the advent of the WHOI Micro-Modem \citep{Singh2006} and synchronized clocks.
+AUV navigation efforts have achieved root mean square (RMS) localization error on the order of tens of meters relative to GNSS surface position over less than ten kilometers in shallow \citep{Eustice2007,Kepper2017,Claus2018} and deep water \citep{Kunz2008,Jakuba2008,Webster2009}.
 However, these efforts all used a nominal sound speed for travel time conversion and the vehicles were limited to shallower isovelocity regimes.
@@
 These methods have been reserved for post-processing as they can be labor intensive, computationally heavy, and/or require additional information like contemporaneous data.
-For example, gliders navigating with kinematic flight models and equipped with acoustic modems were later unambiguously associated with predicted ray arrivals, resulting in roughly a kilometer error and hundred meters uncertainty over basin scale propagation \cite{van_uffelen_estimating_2013}.
-A follow up study investigated how a single temporally and spatially averaged SSP could mitigate position error for a four month glider mission \cite{van_uffelen_localization_2016}.
-\citet{wu_assessment_2018} cross correlate three days of real acoustic records with synthetic ones generated through ocean model snapshots from HYCOM \cite{chassignet_hycom_2007}.
+For example, gliders navigating with kinematic flight models and equipped with acoustic modems were later unambiguously associated with predicted ray arrivals, resulting in roughly a kilometer error and hundred meters uncertainty over basin scale propagation \cite{VanUffelen2013}.
+A follow up study investigated how a single temporally and spatially averaged SSP could mitigate position error for a four month glider mission \cite{vanuffelen2016localization}.
+\citet{Wu2019} cross correlate three days of real acoustic records with synthetic ones generated through ocean model snapshots from HYCOM \cite{Chassignet2007}.
 While potentially applicable for various ocean states, this is reliant on model realism and impractical for real-time operations.
-\citet{mikhalevsky_deep_2020} introduces a ``cold start'' algorithm that does not require prior knowledge of track, position, or sound speed information.
+\citet{Mikhalevsky2020} introduces a ``cold start'' algorithm that does not require prior knowledge of track, position, or sound speed information.
 The algorithm inputs a four-dimensional ocean model, constrained by tomography data, into a range dependent ray code to isolate the last path detected in a full multipath pattern.
@@
 
-The ICEX20 AUV deployment necessitated an environmentally- and physically-driven relationship between recorded travel time and estimated range due to the multipath uncertainty brought upon by an increasingly observed double ducted environment in the Beaufort Sea, which some refer to as the ``Beaufort Lens'' \citep{litvak2015acoustics,chen_spectral_2019,chen2020temporal}.
+The ICEX20 AUV deployment necessitated an environmentally- and physically-driven relationship between recorded travel time and estimated range due to the multipath uncertainty brought upon by an increasingly observed double ducted environment in the Beaufort Sea, which some refer to as the ``Beaufort Lens'' \citep{litvak2015acoustics,Chen2019,Chen2020}.
 
 Given that a lens introduces significant ray refraction, the Beaufort Lens is a convenient shorthand for the spatio-temporal variability of the local temperature and sound speed maxima generally around 50 to 60 m in depth.
-A neutrally buoyant layer of warm Pacific Summer Water creates a unique double ducted environment \textemdash the upper duct degrades signal coherence due to intensified ice interaction and the lower duct effectively traps sound for long range propagation \citep{poulsen_acoustic_2016}.
-\llabel{2.5} Modeling output \citep{duda_long-range_2019,duda_effects_2021} and experimental observations \citep{ballard_temporal_2020,Bhatt2021} suggest the duct is persistent and widespread but not necessarily continuous; it and its acoustic effects can be non-existent, minimal, or drastic.
+A neutrally buoyant layer of warm Pacific Summer Water creates a unique double ducted environment \textemdash the upper duct degrades signal coherence due to intensified ice interaction and the lower duct effectively traps sound for long range propagation \citep{Poulsen2017}.
+\llabel{2.5} Modeling output \citep{Duda2019,Duda2021} and experimental observations \citep{Ballard2020,Bhatt2021} suggest the duct is persistent and widespread but not necessarily continuous; it and its acoustic effects can be non-existent, minimal, or drastic.
 Transmissions in the upper duct, between the surface ice and the lens, experience minimal attenuation but degrade in signal coherence with repeated reflections under the ice.
-In lower duct, between the lens and its conjugate depth in the Atlantic water, roughly 200 m, sound above 350 Hz is trapped effectively for long range propagation \citep{poulsen2016acoustic}.
+In lower duct, between the lens and its conjugate depth in the Atlantic water, roughly 200 m, sound above 350 Hz is trapped effectively for long range propagation \citep{Poulsen2017}.
 
-Thorough reviews of uncrewed vehicle operations in polar environments can be found in \citep{norgren_unmanned_2014} and \citep{barker_scientific_2020}; there is no comparable work in the Arctic for a short range AUV deployment in the Beaufort Lens.
-Seminal \citep{brooke1981arcs,jackson1983autonomous,light1989autonomous,bellingham1995auv,hayes2002determining} and more recent AUV deployments \citep{jakuba2008long,kunz_deep_2008,kukulya2010under,plueddemann_autonomous_2012,timmermans2013scales,fossum2021adaptive} witnessed the classical upward refracting sound speed profile that is amenable to an isovelocity assumption.
+Thorough reviews of uncrewed vehicle operations in polar environments can be found in \citep{Norgren2014} and \citep{Barker2020}; there is no comparable work in the Arctic for a short range AUV deployment in the Beaufort Lens.
+Seminal \citep{brooke1981arcs,jackson1983autonomous,light1989autonomous,bellingham1995auv,hayes2002determining} and more recent AUV deployments \citep{Jakuba2008,Kunz2008,}\citep{jakuba_2008_long,kunz_deep_2008,kukulya2010under,plueddemann_autonomous_2012,timmermans2013scales,fossum2021adaptive} witnessed the classical upward refracting sound speed profile that is amenable to an isovelocity assumption.
 
 Of note, despite different platforms and scales, are recent glider deployments in the Canada Basin.
-In 2014, in partially ice-covered conditions, a long range LBL system with WHOI Micro-Modems at 100 m exploited the lower duct for long range communication with two gliders \citep{freitag_long_2015,webster2015towards}.
+In 2014, in partially ice-covered conditions, a long range LBL system with WHOI Micro-Modems at 100 m exploited the lower duct for long range communication with two gliders \citep{Freitag2016,Webster2015}.
 The sound speed value measured at the time of reception was used to estimate range in post-processing.
@@
 In 2017, gliders were deployed in a region with no ice coverage. 
-Ranges were linearly scaled by a statistical description of sound speed observations taken during the experiment, 1450 $\pm$ 6.5 m/s \citep{graupe2019preliminary}.
+Ranges were linearly scaled by a statistical description of sound speed observations taken during the experiment, 1450 $\pm$ 6.5 m/s \citep{Graupe2019}.
 This resulted in an error of 550 m, which was reduced by a factor between 4 and 5, depending on the dive, using a post-processing acoustic arrival matching method.
@@
 Unintuitively, the smaller nature of our deployment during ICEX20 is not a simplifying factor.
-For source depths typical to vehicle operations, 30 to 200 m, a shadow zone spans from 2 to 6 kilometers in range \citep{schmidt2016acoustic}.
+For source depths typical to vehicle operations, 30 to 200 m, a shadow zone spans from 2 to 6 kilometers in range \citep{Schmidt2016}.
 
@@
 The results from this paper derive from data taken while deploying the AUV \emph{Macrura}, a custom Bluefin-21, during the Ice Exercise 2020 (ICEX20), in the Beaufort Sea, from March 8 to 11, in double-ducted and ice-covered conditions.
-The AUV deployment was supported by the Integrated Communication and Navigation Network (ICNN) \citep{schneider_self-adapting_2020,randeni_construction_2020,randeni_high-resolution_2021}, a specialized implementation of the LBL solution.
+The AUV deployment was supported by the Integrated Communication and Navigation Network (ICNN) \citep{Schneider2020,Randeni2020,randeni_high-resolution_2021}, a specialized implementation of the LBL solution.
 The ICNN was initially developed via numerous virtual experiments to push the boundaries of algorithms and interfaces between different hardware components.
 The simulation approach serves as a testbed for robustness to produce better results than post-processing previous field data.
-The simulation capabilities are largely physics-driven with a modular system of systems approach\textemdash an environmental simulator with sub-components for the ocean, including Arctic ice drift and ocean acoustic propagation; a vehicle simulator with sub-components for vehicle dynamics and navigation; a topside hardware simulator and acoustic communications simulator, both with a software-only configuration and a hardware-in-the-loop version \citep{schneider_netsim_2018}.
+The simulation capabilities are largely physics-driven with a modular system of systems approach\textemdash an environmental simulator with sub-components for the ocean, including Arctic ice drift and ocean acoustic propagation; a vehicle simulator with sub-components for vehicle dynamics and navigation; a topside hardware simulator and acoustic communications simulator, both with a software-only configuration and a hardware-in-the-loop version \citep{Schneider2018}.
 The virtual environment similarly emulates the interfaces between the real components to test the entire software pipeline.
@@
 The ICNN is comprised of four ice buoys, in a loose rectangle, roughly 2 km away from a central ice camp with a topside computer, as shown in Fig.~\ref{fig:icnnOverview}.
-The AUV and each ice buoy are outfitted with a WHOI Micro-Modem \citep{singh_underwater_2006}, with a four-element receiver array and a single transmitter, one-tenth of a millisecond resolution.
+The AUV and each ice buoy are outfitted with a WHOI Micro-Modem \citep{Singh2006}, with a four-element receiver array and a single transmitter, one-tenth of a millisecond resolution.
 Acoustic messages were sent with a 10 kHz carrier frequency, 5 kHz bandwidth, and phase-shift keying (PSK)modulation on a thirty-second cycle, giving room for two-way communication throughout the mission volume.
@@
 While each buoy only has one transmit depth, all buoys have both receive depths but the active receive layer is consistent across all buoys. \llabel{1.12}
-The design of the ICNN enables a self-adapting network to transmit and receive at the optimal depth to maintain contact with the AUV \citep{schneider_self-adapting_2020}.
-The buoys do not encompass the full horizontal range of the vehicle but are positioned to minimize overlap in trilateration for spherical positioning \citep{deffenbaugh_relationship_1996}.
+The design of the ICNN enables a self-adapting network to transmit and receive at the optimal depth to maintain contact with the AUV \citep{Schneider2020}.
+The buoys do not encompass the full horizontal range of the vehicle but are positioned to minimize overlap in trilateration for spherical positioning \citep{Deffenbaugh1996}.
 
@@
 \item Four ice buoys, each outfitted with a WHOI Micro-Modem, receive messages from the AUV and send that information over freewave radio to a Topside computer
-\item The topside computer converts travel times into range estimates using a stochastic embedded prediction of the horizontal group velocity via BELLHOP ray tracing code \citep{porter_bellhop_2011} coupled with an updatable Virtual Ocean \citep{schneider_netsim_2018,bhatt_embedded_2021}
+\item The topside computer converts travel times into range estimates using a stochastic embedded prediction of the horizontal group velocity via BELLHOP ray tracing code \citep{Porter2011} coupled with an updatable Virtual Ocean \citep{Schneider2018,bhatt_embedded_2022}
 \item The topside computer calculates a new position by trilaterating the range estimates
@@
 An important component to our navigation solution is an accurate estimation of the sound speed profile.
-Previous field experience, during the Ice Exercise 2016 (ICEX16), demonstrated the negative effects of the Beaufort Lens on tracking and communication \citep{schmidt2016acoustic}.
+Previous field experience, during the Ice Exercise 2016 (ICEX16), demonstrated the negative effects of the Beaufort Lens on tracking and communication \citep{Schmidt2016}.
 Fig. \ref{fig:sspExpectation} shows historical, modeled, and \textit{in situ} data for both ICEX16 and ICEX20.
 These three input streams were selected to mirror the information available on a submarine (personal conversation with LT B. Howard and LT CDR D. Goodwin). \llabel{1.8}
-The SSP information was compressed with a basis representation and sent via lightweight digital acoustic message to the AUV \citep{bhatt_embedded_2021}. \llabel{2.7}
-All modeled data comes from HYCOM \cite{chassignet_hycom_2007}, which does not seem to capture the forcing mechanisms that cause the Beaufort Lens.
-For ICEX16, the data-driven profile was sourced from nearby Ice Tethered Profilers (ITP) after the field experiment \cite{krishfield2008automated,toole2011ice} and exhibits a fairly low lens; the historical profile is from the World Ocean Atlas.
+The SSP information was compressed with a basis representation and sent via lightweight digital acoustic message to the AUV \citep{bhatt_embedded_2022}. \llabel{2.7}
+All modeled data comes from HYCOM \cite{Chassignet2007}, which does not seem to capture the forcing mechanisms that cause the Beaufort Lens.
+For ICEX16, the data-driven profile was sourced from nearby Ice Tethered Profilers (ITP) after the field experiment \cite{Krishfield2008,Toole2011} and exhibits a fairly low lens; the historical profile is from the World Ocean Atlas.
 For ICEX20, the chosen weights (data-driven) profile derives from a basis representation estimation of initial CTD casts taken on site, showing an intense warm water intrusion; the baseline (historical) profile, showing moderate ducted conditions, comes from the average of March 2013 data.
-This month best matched sea ice and sound speed conditions at the beginning of ICEX20 \citep{bhatt_embedded_2021}.
+This month best matched sea ice and sound speed conditions at the beginning of ICEX20 \citep{bhatt_embedded_2022}.
 It is important to note that all profiles that do show the Beaufort Lens do so with different local sound speed maxima at different depths, reflective of the wide range of lens properties observed for all ITP data in the region.
@@
 
-This method selects a different group velocity based on the multipath arrival structure, as the detected arrival is not always the first arrival or the direct path and could even be masked by noise or blocked temporarily \citep{deffenbaugh_acoustic_1996}.
+This method selects a different group velocity based on the multipath arrival structure, as the detected arrival is not always the first arrival or the direct path and could even be masked by noise or blocked temporarily \citep{deffenbaugh1996acoustic}.
 We manually cap the number of bounces to 4 because of the smaller operational scale and the attenuation accrued with many surface interactions.
@@
 BELLHOP simulations provide other relevant eigenray information, like time and angle of arrival, that is ripe for statistical and machine learning methods to classify a representative group velocity.
-A bespoke and fast ray tracing method, which was used during ICEX20 to evaluate the acoustic fit of sound speed profile parameterization \citep{bhatt_embedded_2021}, can easily report back the number of turning points instead of the number of bounces for multipath classification.
+A bespoke and fast ray tracing method, which was used during ICEX20 to evaluate the acoustic fit of sound speed profile parameterization \citep{bhatt_embedded_2022}, can easily report back the number of turning points instead of the number of bounces for multipath classification.
 
-This approach will start to break down in extremely dynamic environments.
-Fast moving fronts, as seen in estuaries like the Connecticut River into the Long Island Sound, present an entirely new set of challenges not seen by internal waves or eddies.
-Buoyancy fluctuations in these regions threaten even simple AUV tasks like following a trackline.
-Acoustic communications are further complicated given a shallow environment with significant scattering \citep{lavery_measurements_2010,ross_acoustic_2012,lavery_broadband_2013}, where fast acoustic modeling may only be coherent for trivial probabilities of the ocean state.
-Realistic \textit{in situ} considerations of the acoustic environment may not be possible in such environments without complete through-the-sensor integration of echosounder data and/or a hyper-realistic onboard ocean model.
+This approach will start to break down in extremely dynamic environments, like fast moving fronts.
+Realistic \textit{in situ} considerations of the acoustic environment may not be possible without complete through-the-sensor integration of acoustic dat and/or hyper realistic ocean models.
 
@@
 For vehicle operations, fast tomography is the ideal implementation, in that one can fully consider how sound speed structure, horizontally and vertically, influences sound propagation.
-AUVs can serve as moving sources to better image the ocean volume \citep{deffenbaugh_optimal_1997,elisseeff_ocean_2002}, where mobile tomography and navigation converge on the same set of component technologies: position estimation, sound speed parameterization estimation, ray path identification, and vehicle path optimization.
+AUVs can serve as moving sources to better image the ocean volume \citep{Deffenbaugh1997,Elisseeff2002}, where mobile tomography and navigation converge on the same set of component technologies: position estimation, sound speed parameterization estimation, ray path identification, and vehicle path optimization.
 

### 2022-02-20 pipeline/9-trilateration/auv_trilat_hist.m
changing xlabels

@@ -27,7 +27,7 @@ Y{2}.count = trilat.mbc.correction;
 Y{1}.count = trilat.nbc.correction;
 h_hist_boxplot(Y,edges,colors,max(edges));
 
-xlabel('RMS of position correction [m]')
+xlabel('RMS correction [m]')
 set(gca,'fontsize',13)
 title(sprintf('Distribution of AUV re-navigation corrections (n=%u)',N),'fontsize',15)
 
@@ -41,7 +41,7 @@ Y{1}.count = trilat.nbc.error;
 Y{2}.count = trilat.mbc.error;
 
 h_hist_boxplot(Y,edges,colors,max(edges));
-xlabel('RMS of position error [m]')
+xlabel('RMS error [m]')
 set(gca,'fontsize',13)
 title({'',...
     sprintf('Distribution of AUV re-navigation error (n=%u)',N)}...

### 2022-02-20 pipeline/9-trilateration/modem_trilat_hist.m

@@ -20,7 +20,7 @@ figure('name','trilat-histogram','renderer','painters','position',[108 108 1200 
 h_hist_boxplot(Y,edges,colors,48);
 
 %% dress up figure
-xlabel('RMS of position correction [m]')
+xlabel('RMS correction [m]')
 set(gca,'fontsize',13)
 title(sprintf('Distribution of beacon re-positioning corrections (n=%u)',N),'fontsize',15);
 

### 2022-02-20 manuscript/main.tex
including trilat re-nav stat in abstract

@@
 \begin{abstract}
-The long baseline (LBL) underwater navigation paradigm relies on the conversion of recorded travel time to range to trilaterate for position.
+The long baseline (LBL) underwater navigation paradigm relies on the conversion of recorded travel times into pseudoranges to trilaterate position.
 For real-time operations, this conversion has assumed an isovelocity sound speed.
@@
 The real-time ranging error between beacons is roughly 11 meters at distances up to 3 km.
-But a consistent overestimation in the real-time method, the Minimum Bounce Criteria (MBC), provides insights for improved eigenray filtering, which we call the Nearest Bounce Criteria (NBC).
+But a consistent overestimation in the real-time method provides insights for improved eigenray filtering by the number of surface bounces.
 An operationally equivalent pipeline is used to re-position the LBL beacons and re-navigate the AUV, using a modeled, historical, and a locally observed sound speed profile.
-The median re-positioning errors for the MBC and NBC are roughly 10 and 3 meters, respectively.
-The improved trilateration performance for re-positioning and re-navigation suggests that this approach effectively extends the single meter accuracy of the deployed GNSS units into the water column.
-\llabel{2.3}
+The median re-navigation error is 1.84 $\pm$ 2.19 RMS m.
+The improved trilateration performance for suggests that this approach effectively extends the single meter accuracy of the deployed GNSS units into the water column. \llabel{2.3}
 \end{abstract}
@@
 \llabel{2.11} The initial call to BELLHOP is over a local grid centered at the range and depth posited by the onboard tracking solution.
-Compared to a point solver, solving for arrivals on a grid minimizes the impact of numerical resolution in the propagation model.
-The resulting ray traces may represent paths that would reach a receiver in a real application but might not be recognized by the numerical solver as a viable detection.
-The grid is initialized as 11$\times$11 points spanning 10 m horizontally and 20 m vertically.
+The grid, compared to a point solver, adds redundancy in resolving the actual multipath structure for a reliable acoustic path without taxing onboard computational time and memory.
+It is initialized as 11$\times$11 points spanning 10 m horizontally and 20 m vertically.
 The horizontal dimension reflects the accumulated position error given a thirty-second communication cycle; the vertical dimension reflects how, computationally, eigenrays of the same timefront seem to stack vertically in the water column.

### 2022-02-20 manuscript/main.tex
reread thru intro

@@
 However, navigation uncertainty is a major challenge in considering AUVs as standard tools for oceanographic research.
-While land and air-based robots utilize information from Global Navigation Satellite Systems (GNSS) to achieve stunning location accuracy and precision throughout the duration of their missions, AUVs cannot access GNSS while underwater due to the rapid attenuation of electromagnetic waves.
+While land and air-based robots utilize information from Global Navigation Satellite Systems (GNSS) to achieve stunning location accuracy and precision throughout the duration of their missions, AUVs cannot access GNSS fixes while underwater.
 Therefore, underwater vehicles have relied on any combination of dead reckoning, hydrodynamic models, inertial navigation systems, doppler velocity logs, and acoustic baseline positioning systems for navigation \citep{Paull2014}.
@@
 \end{enumerate}
-Thus, navigation is inherently in real time and depends on positioning; positioning depends on timing.
+Thus, navigation is inherently in real-time and depends on positioning; positioning depends on timing.
 We also suggest re-navigation and re-positioning as post-processed corollaries, which may include knowledge or processing capabilities not available \textit{in situ}.
@@
 % More about LBL navigation
-While RAFOS floats championed one way ranging for re-positioning \citep{Rossby1986,Duda2006}, the ability to do so for navigation was facilitated by \llabel{1.11} the advent of the WHOI Micro-Modem \citep{Singh2006} and synchronized clocks.
+While RAFOS floats championed one way ranging for re-positioning \citep{Rossby1986,Duda2006}, the ability to do so for navigation was facilitated by \llabel{1.11} the advent of the WHOI Micro-Modem \citep{Singh2006} and synchronized clocks \citep{rypkema_one-way_2017}.
 AUV navigation efforts have achieved root mean square (RMS) localization error on the order of tens of meters relative to GNSS surface position over less than ten kilometers in shallow \citep{Eustice2007,Kepper2017,Claus2018} and deep water \citep{Kunz2008,Jakuba2008,Webster2009}.
@@
 Localization algorithms that do consider environmental or acoustic uncertainty tend to focus on longer and larger experiments, where spatio-temporal variability cannot be ignored.
-These methods have been reserved for post-processing as they can be labor intensive, computationally heavy, and/or require additional information like contemporaneous data.
+These methods have also been reserved for post-processing as they can be labor intensive, computationally heavy, and/or require additional information like contemporaneous data.
 For example, gliders navigating with kinematic flight models and equipped with acoustic modems were later unambiguously associated with predicted ray arrivals, resulting in roughly a kilometer error and hundred meters uncertainty over basin scale propagation \cite{VanUffelen2013}.
@@
 
-The ICEX20 AUV deployment necessitated an environmentally- and physically-driven relationship between recorded travel time and estimated range due to the multipath uncertainty brought upon by an increasingly observed double ducted environment in the Beaufort Sea, which some refer to as the ``Beaufort Lens'' \citep{litvak2015acoustics,Chen2019,Chen2020}.
+The ICEX20 AUV deployment necessitated an environmentally- and physically-driven relationship between recorded travel times and estimated pseudoranges due to the multipath uncertainty brought upon by an increasingly observed double ducted environment in the Beaufort Sea, which some refer to as the ``Beaufort Lens'' \citep{litvak2015acoustics,Chen2019,Chen2020}.
 
-Given that a lens introduces significant ray refraction, the Beaufort Lens is a convenient shorthand for the spatio-temporal variability of the local temperature and sound speed maxima generally around 50 to 60 m in depth.
+Given that a lens introduces significant ray refraction, the Beaufort Lens is a shorthand for the spatio-temporal variability of the local temperature and sound speed maxima generally around 50 to 60 m in depth.
 A neutrally buoyant layer of warm Pacific Summer Water creates a unique double ducted environment \textemdash the upper duct degrades signal coherence due to intensified ice interaction and the lower duct effectively traps sound for long range propagation \citep{Poulsen2017}.
-\llabel{2.5} Modeling output \citep{Duda2019,Duda2021} and experimental observations \citep{Ballard2020,Bhatt2021} suggest the duct is persistent and widespread but not necessarily continuous; it and its acoustic effects can be non-existent, minimal, or drastic.
+\llabel{2.5} Modeling output \citep{Duda2019,Duda2021} and experimental observations \citep{Ballard2020,Bhatt2021} suggest that, in the Beaufort Sea, the duct is persistent and widespread but not necessarily continuous; it and its acoustic effects can be non-existent, minimal, or drastic.
 Transmissions in the upper duct, between the surface ice and the lens, experience minimal attenuation but degrade in signal coherence with repeated reflections under the ice.
-In lower duct, between the lens and its conjugate depth in the Atlantic water, roughly 200 m, sound above 350 Hz is trapped effectively for long range propagation \citep{Poulsen2017}.
+In lower duct, between the lens and its conjugate depth in the Atlantic water, roughly 200 m, sound above 350 Hz is trapped near losslessly for long range propagation \citep{Poulsen2017}.
 
@@
 In 2014, in partially ice-covered conditions, a long range LBL system with WHOI Micro-Modems at 100 m exploited the lower duct for long range communication with two gliders \citep{Freitag2016,Webster2015}.
-The sound speed value measured at the time of reception was used to estimate range in post-processing.
+The sound speed value measured at the time of reception was used to estimate pseudorange in post-processing.
 The beacon-to-beacon performance was excellent, achieving contact at ranges greater than 200 km with a position uncertainty of 40 m.
-The beacon-to-glider performance, however, deteriorated due to missed contacts outside the duct.
-In 2017, gliders were deployed in a region with no ice coverage. 
-Ranges were linearly scaled by a statistical description of sound speed observations taken during the experiment, 1450 $\pm$ 6.5 m/s \citep{Graupe2019}.
+The beacon-to-glider performance, however, deteriorated due to missed contacts outside the duct, and was not described quantitatively.
+In 2017, gliders were deployed in a region with no ice coverage \citep{Graupe2019}.
+Ranges were linearly scaled by a statistical description of sound speed observations taken during the experiment, 1450 $\pm$ 6.5 m/s.
 This resulted in an error of 550 m, which was reduced by a factor between 4 and 5, depending on the dive, using a post-processing acoustic arrival matching method.
@@
 The paper is organized as follows.
-Section \ref{sec:icex20} details the experimental conditions during ICEX20.
+Section \ref{sec:icex20} details the experimental approach and conditions during ICEX20.
 Given that there is no GNSS ground truth for the vehicle position while underway, we first evaluate the real-time ranging performance of GPS-linked beacon-to-beacon communication events in section \ref{sec:realtime}.

### 2022-02-20 manuscript/main.tex
fix bibs for intro

@@
 Thorough reviews of uncrewed vehicle operations in polar environments can be found in \citep{Norgren2014} and \citep{Barker2020}; there is no comparable work in the Arctic for a short range AUV deployment in the Beaufort Lens.
-Seminal \citep{brooke1981arcs,jackson1983autonomous,light1989autonomous,bellingham1995auv,hayes2002determining} and more recent AUV deployments \citep{Jakuba2008,Kunz2008,}\citep{jakuba_2008_long,kunz_deep_2008,kukulya2010under,plueddemann_autonomous_2012,timmermans2013scales,fossum2021adaptive} witnessed the classical upward refracting sound speed profile that is amenable to an isovelocity assumption.
+Seminal \citep{brooke1981arcs,jackson1983autonomous,light1989autonomous,bellingham1995auv,hayes2002determining} and more recent AUV deployments \citep{Jakuba2008,Kunz2008,kukulya2010under,plueddemann_autonomous_2012,timmermans2013scales,fossum2021adaptive} witnessed the classical upward refracting sound speed profile that is amenable to an isovelocity assumption.
 
@@
 
-The results from this paper derive from data taken while deploying the AUV \emph{Macrura}, a custom Bluefin-21, during the Ice Exercise 2020 (ICEX20), in the Beaufort Sea, from March 8 to 11, in double-ducted and ice-covered conditions.
+The results from this paper derive from data taken while deploying the AUV \emph{Macrura}, a custom Bluefin-21, during the Ice Exercise 2020 (ICEX20), in the Beaufort Sea, from March 8th to 11th.
 The AUV deployment was supported by the Integrated Communication and Navigation Network (ICNN) \citep{Schneider2020,Randeni2020,randeni_high-resolution_2021}, a specialized implementation of the LBL solution.

### 2022-02-20 manuscript/main.tex
thru section 2

@@
 To achieve single meter, GNSS-like performance in a GNSS-denied environment, we demonstrate an embedded ray-based data processing algorithm to convert recorded OWTTs into pseudorange estimates.
-This methodology was integrated onto the AUV Macrura, deployed and recovered in the Beaufort Sea, in March 2020, during the Ice Exercise 2020 (ICEX20).
+This methodology was integrated onto the AUV \emph{Macrura}, deployed and recovered in the Beaufort Sea, in March 2020, during the Ice Exercise 2020 (ICEX20).
 A physics-driven methodology that received an \textit{in situ} sound speed profile (SSP) was necessary despite the small operational domain because of the relatively high-risk mission environment\textemdash total under-ice conditions and a variable double ducted acoustic environment.
@@
 % PNT
-For clarity, we delineate specific definitions for timing, positioning, and navigation from \citet{howe2019observing}.
+For consistency, we delineate specific definitions for timing, positioning, and navigation from \citet{howe2019observing}.
 \begin{enumerate}
@@
 A neutrally buoyant layer of warm Pacific Summer Water creates a unique double ducted environment \textemdash the upper duct degrades signal coherence due to intensified ice interaction and the lower duct effectively traps sound for long range propagation \citep{Poulsen2017}.
-\llabel{2.5} Modeling output \citep{Duda2019,Duda2021} and experimental observations \citep{Ballard2020,Bhatt2021} suggest that, in the Beaufort Sea, the duct is persistent and widespread but not necessarily continuous; it and its acoustic effects can be non-existent, minimal, or drastic.
+Modeling output \citep{Duda2019,Duda2021} and experimental observations \citep{Ballard2020,Bhatt2021} suggest that, in the Beaufort Sea, the duct is persistent and widespread but not necessarily continuous; it and its acoustic effects can be non-existent, minimal, or drastic.
 Transmissions in the upper duct, between the surface ice and the lens, experience minimal attenuation but degrade in signal coherence with repeated reflections under the ice.
@@
 The AUV deployment was supported by the Integrated Communication and Navigation Network (ICNN) \citep{Schneider2020,Randeni2020,randeni_high-resolution_2021}, a specialized implementation of the LBL solution.
-The ICNN was initially developed via numerous virtual experiments to push the boundaries of algorithms and interfaces between different hardware components.
-The simulation approach serves as a testbed for robustness to produce better results than post-processing previous field data.
+The ICNN was initially developed via numerous virtual experiments to ensure robust algorithms and interfaces between different hardware components. 
 The simulation capabilities are largely physics-driven with a modular system of systems approach\textemdash an environmental simulator with sub-components for the ocean, including Arctic ice drift and ocean acoustic propagation; a vehicle simulator with sub-components for vehicle dynamics and navigation; a topside hardware simulator and acoustic communications simulator, both with a software-only configuration and a hardware-in-the-loop version \citep{Schneider2018}.
 The virtual environment similarly emulates the interfaces between the real components to test the entire software pipeline.
-Both simulation capabilities are integral to mission success.
+
+\subsection{The Integrated Communication and Navigation Network}
 
@@
 
-The ICNN is comprised of four ice buoys, in a loose rectangle, roughly 2 km away from a central ice camp with a topside computer, as shown in Fig.~\ref{fig:icnnOverview}.
-The AUV and each ice buoy are outfitted with a WHOI Micro-Modem \citep{Singh2006}, with a four-element receiver array and a single transmitter, one-tenth of a millisecond resolution.
-Acoustic messages were sent with a 10 kHz carrier frequency, 5 kHz bandwidth, and phase-shift keying (PSK)modulation on a thirty-second cycle, giving room for two-way communication throughout the mission volume.
-The receive and transmit elements were split between shallow and deeper depths (30 and 90 m, respectively) to provide better coverage across the shadow zone.
+The ICNN is comprised of four ice buoys in a loose rectangle, roughly 2 km away from a central ice camp with a topside computer, as shown in Fig.~\ref{fig:icnnOverview}.
+The AUV and each ice buoy are outfitted with a WHOI Micro-Modem \citep{Singh2006}, with a four-element receiver array, a single transmitter, and one-tenth of a millisecond resolution.
+Acoustic messages were sent with a 10 kHz carrier frequency, 5 kHz bandwidth, and phase-shift keying (PSK)modulation on a time-division multiple access schedule with a thirty-second cycle, giving room for two-way communication throughout the mission volume.
+The receive and transmit elements were split between shallow and deeper depths\textemdash 30 and 90 m\textemdash to provide better coverage across the shadow zone.
 While each buoy only has one transmit depth, all buoys have both receive depths but the active receive layer is consistent across all buoys. \llabel{1.12}
@@
 \item Four ice buoys, each outfitted with a WHOI Micro-Modem, receive messages from the AUV and send that information over freewave radio to a Topside computer
-\item The topside computer converts travel times into range estimates using a stochastic embedded prediction of the horizontal group velocity via BELLHOP ray tracing code \citep{Porter2011} coupled with an updatable Virtual Ocean \citep{Schneider2018,bhatt_embedded_2022}
+\item The topside computer converts travel times into pseudorange estimates using a stochastic embedded prediction of the horizontal group velocity via BELLHOP ray tracing code \citep{Porter2011} using a sound speed profile provided by an updatable Virtual Ocean \citep{Schneider2018,bhatt_embedded_2022}
 \item The topside computer calculates a new position by trilaterating the range estimates
@@
 
-In the face of GNSS-denied navigation, this approach was anecdotally successful, as shown in \ref{fig:vehicleRecovery}.
+In the face of GNSS-denied navigation, this approach was anecdotally successful, as shown in Fig. \ref{fig:vehicleRecovery}.
 The AUV \emph{Macrura} was deployed through a hydrohole from an ice camp but recovered through an emergency hydrohole.
-A random disk error stalled the AUV underneath the ice but did not prevent it from transmitting its location.
+A random disk error stalled \emph{Macrura} underneath the ice but did not prevent it from transmitting its location.
 Due to an incoming storm, a team placed a physical marker on the ice at the location.
@@
 We view the emergency recovery as qualitative proof of the robustness of this navigation approach.
-Nonetheless, this paper specifically addresses the third and fourth steps\textemdash the conversion of travel time into range and its effect on trilateration.
-By focusing on range estimates between GPS-tracked beacons, and re-running the trilateration pipeline, the results are decoupled from all other mechanisms in the ICNN.
+Nonetheless, this paper specifically addresses the third and fourth steps\textemdash the conversion of travel times into pseudoranges and its effect on trilateration.
+By focusing on pseudorange estimates between GPS-tracked beacons, and re-running the trilateration pipeline, the results are decoupled from all other mechanisms in the ICNN.
 
@@
 	\includegraphics[width=0.28\columnwidth]{figs/Fig1.jpg}
-	\caption{The under-ice mission track for AUV Macrura, including the position updates as it stalled underneath the ice overnight.
+	\caption{The under-ice mission track for AUV \emph{Macrura}, including the position updates as it stalled underneath the ice overnight.
 A marker was placed on the ice at the vehicle's estimated self-location.
@@
 
-An important component to our navigation solution is an accurate estimation of the sound speed profile.
+\subsection{ICEX20 sound speed conditions}
+
+An important component to our navigation solution is an accurate estimation of a representative SSP for the ocean volume.
 Previous field experience, during the Ice Exercise 2016 (ICEX16), demonstrated the negative effects of the Beaufort Lens on tracking and communication \citep{Schmidt2016}.
-Fig. \ref{fig:sspExpectation} shows historical, modeled, and \textit{in situ} data for both ICEX16 and ICEX20.
+Fig. \ref{fig:sspExpectation} shows historical, modeled, and \textit{in situ} sound speed data for both ICEX16 and ICEX20.
 These three input streams were selected to mirror the information available on a submarine (personal conversation with LT B. Howard and LT CDR D. Goodwin). \llabel{1.8}
-The SSP information was compressed with a basis representation and sent via lightweight digital acoustic message to the AUV \citep{bhatt_embedded_2022}. \llabel{2.7}
+In the field, the SSP information was shared with the vehicle via basis representation compression on a lightweight digital acoustic message \citep{bhatt_embedded_2022}. \llabel{2.7}
 All modeled data comes from HYCOM \cite{Chassignet2007}, which does not seem to capture the forcing mechanisms that cause the Beaufort Lens.
 For ICEX16, the data-driven profile was sourced from nearby Ice Tethered Profilers (ITP) after the field experiment \cite{Krishfield2008,Toole2011} and exhibits a fairly low lens; the historical profile is from the World Ocean Atlas.
-For ICEX20, the chosen weights (data-driven) profile derives from a basis representation estimation of initial CTD casts taken on site, showing an intense warm water intrusion; the baseline (historical) profile, showing moderate ducted conditions, comes from the average of March 2013 data.
+For ICEX20, the chosen weights (data-driven) profile derives from an estimate of initial CTD casts taken on site, showing an intense warm water intrusion; the baseline (historical) profile, showing moderate ducted conditions, comes from the average of March 2013 ITP data.
 This month best matched sea ice and sound speed conditions at the beginning of ICEX20 \citep{bhatt_embedded_2022}.
 It is important to note that all profiles that do show the Beaufort Lens do so with different local sound speed maxima at different depths, reflective of the wide range of lens properties observed for all ITP data in the region.
+\llabel{2.5}
 The variability of the lens height and prominence is the main reason an updatable SSP was integrated into the ICNN solution.
@@
 
-\subsection{Re-navigating AUV Macrura}
+\subsection{Re-navigating AUV \emph{Macrura}}
 

### 2022-02-20 pipeline/1-overview/plot_modem_chart.m
added tx counts to modem chart

@@ -136,12 +136,14 @@ eof_time = RECAP.data_time;
 % for each tx depth (if multiple, bottom first)
 % for each rx depth
 % in order of valid data in modem labels -- North, South, East, West, Camp
-countOscar = [44 110 18 82 61 28 137 74 ...
+countRX = [44 110 18 82 61 28 137 74 ...
               71 74 121 160 76 ...
               32 10 28 34 ...
               91 55 150 155 38 ...
               12 43 45 34];
 c0 = 0;
+
+countTX = containers.Map(modem_labels,[303 94 417 95 396]);
 
 markerSize = 220;
 
@@ -190,7 +192,7 @@ plotspread = linspace(0,N,5) - N/2;
 xOffsetMap = containers.Map(modem_labels,plotspread);
 x2OffsetMap = containers.Map([20 30 90],[0 0 .3]);
 z2OffsetMap = containers.Map([20 30 90],[-.1 -.1  .2]);
-textOffsetMap = containers.Map(modem_labels,1.1.*[-1 1 -1 1 -1]);
+textOffsetMap = containers.Map(modem_labels,1.3.*[-1 1 -1 1 -1]);
 
 for txDepth = modem_rx_depth
     
@@ -205,8 +207,13 @@ for txDepth = modem_rx_depth
     
     for tNode = unique_tx_nodes
         
+        % plot tx node in TX
         scatter(xTX+0.5,ZMap(txDepth) + zOffsetMap(tNode{1}) + z2OffsetMap(txDepth),...
             markerSize,markerModemMap(tNode{1}),markerShape(txDepth),'filled')
+        
+        % plot count TX
+        text(xTX+0.5,ZMap(txDepth) + zOffsetMap(tNode{1}) + z2OffsetMap(txDepth) - z2OffsetMap(txDepth) + textOffsetMap(tNode{1}),...
+            num2str(countTX(tNode{1})),'VerticalAlignment','middle','HorizontalAlignment','center','color',markerModemMap(tNode{1}),'fontweight','bold');
         
         tx_node_index = find(strcmp(RECAP.tag_tx,tNode{1}));
         
@@ -243,9 +250,9 @@ for txDepth = modem_rx_depth
                     text(xval,yval - z2OffsetMap(rDepth) + textOffsetMap(tNode{1}),num2str(sum(index)),... 
                        'VerticalAlignment','middle','HorizontalAlignment','center','color',markerModemMap(rNode{1}),'fontweight','bold');
                      
[…]

### 2022-02-20 manuscript/main.tex
thru real-time results

@@
 
-During ICEX20, the HYCOM profile was available but never deployed. \llabel{1.14}
-For post-processing comparison, we introduce both the HYCOM profile and an isovelocity case, 1441.8 $\pm$ 3.7 m/s, as the mean and standard deviation of the observed sound speed profile over the first 200 m.
-
 \begin{figure}[h!]
@@
 
+During ICEX20, the HYCOM profile was available but never deployed. \llabel{1.14}
+For post-processing comparison, we introduce both the HYCOM profile and an isovelocity case, 1441.8 $\pm$ 3.7 m/s, as the mean and standard deviation of the observed sound speed profile over the first 200 m.
+
 \clearpage
@@
 Figure \ref{fig:overview} shows successful events sorted by source depth.
-In this analysis, we assume there is insignificant displacement between the GNSS puck surface expression and subsurface modem; this is supported by unusually low observed ice drift rates, on average, just 0.7 cm/s.
+In this analysis, we assume there is insignificant displacement between the GNSS puck surface expression and subsurface modem; this is supported by unusually low observed ice drift rates, just 0.7 cm/s on average throughout the mission.
 
@@
   \includegraphics[width=\textwidth]{figs/modem-chart.pdf}
-  \caption{An overview of the modem experiment by source and receiver depth and position with the number of successful decodings and the percent success rate, defined as the ratio of decoded to detected.
-The black column on the left, $tx$, shows the source depth, $z_s$.
-The column on the right, $rx$, shows the receivers with the amount of good contacts.
+  \caption{An overview of the modem experiment by source and receiver depth and position.
+The black column on the left, $tx$, shows the source depth, $z_s$, with the total number of transmissions.
+The column on the right, $rx$, shows the receivers with the number of successful decodings and the percent success rate, defined as the ratio of decoded to detected.
 The orientation of the triangles\textemdash sideways, upwards, and downwards\textemdash corresponds to depths of 20, 30, and 90 m.}
@@
 
-Instead, for each individual modem $i$, an embedded stochastic tracking framework is used to provide a running estimate of the horizontal group velocity $u_{i,j}$ for the conversion from travel time to range from modem $j$, with the ultimate goal of matching the naive group velocity, i.e.
-the GPS-recorded distance between two nodes divided by the modem-recorded one way travel time between them.
+Instead, for each individual modem $i$, an embedded stochastic tracking framework is used to provide a running estimate of the effective sound speed $c_{i,j}$ for the conversion from travel time to range from modem $j$, with the ultimate goal of matching the implied horizontal effective sound speed, i.e., the GPS-recorded distance between two nodes divided by the modem-recorded one way travel time between them.
 
 \llabel{2.10} In the ICEX20 configuration, the acoustic tracking is running on the topside computer, which controls the ICNN.
-Here we assume that the group velocities $u_{i,j}$ are smoothly varying over the course of a vehicle mission, i.e., with respect to range, mission time, and the frequency of updates relative to vehicle motion.
-The group velocity is tracked on a thirty-second cycle using predictions from the \textit{Virtual Ocean} infrastructure.
+Here we assume that the effective sound speeds $c_{i,j}$ are smoothly varying over the course of a vehicle mission, i.e., with respect to range, mission time, and the thirty-second frequency.
 
-\llabel{1.15} When the topside tracking framework receives a modem message, with a time delay, $\Delta t$, from one of the range modems, it will request a new estimate of the group velocity and its associated uncertainty.
-The group velocity estimate is found using the vehicle's reported depth and the extrapolated navigation solution for range, $\hat{r}$, as inputs for the ray tracing program.
-The latter returns an impulse response estimate in the form of ray travel times $dt_{j}$ and amplitudes $a_{j}$ for that range and depth.
+\llabel{1.15} When the topside tracking framework receives a message, with a time delay, $\Delta t$, it will request a new estimate for $c_{i,j}$ along with its standard deviation.
+The effective sound speed is predicted using the vehicle's reported depth and the extrapolated navigation solution for range, $\hat{r}$, as inputs to the ray tracing program, which returns an impulse response estimate in the form of ray travel times $dt_{j}$ and amplitudes $a_{j}$.
 
 \llabel{2.11} The initial call to BELLHOP is over a local grid centered at the range and depth posited by the onboard tracking solution.
-The grid, compared to a point solver, adds redundancy in resolving the actual multipath structure for a reliable acoustic path without taxing onboard computational time and memory.
-It is initialized as 11$\times$11 points spanning 10 m horizontally and 20 m vertically.
-The horizontal dimension reflects the accumulated position error given a thirty-second communication cycle; the vertical dimension reflects how, computationally, eigenrays of the same timefront seem to stack vertically in the water column.
+The grid, compared to a point solver, adds redundancy in resolving the actual multipath structure for a reliable acoustic path without overtaxing onboard computational time and memory.
+It is initialized as 11 $\times$ 11 points spanning 10 m horizontally and 20 m vertically.
+The horizontal dimension reflects the accumulated vehicle position error given a thirty-second communication cycle; the vertical dimension reflects how, computationally, eigenrays of the same timefront seem to stack vertically in the water column.
 \llabel{1.17}
-For each grid point, BELLHOP produces a number of arrivals resulting from multiple propagation paths for any source-receiver pair.
-Using only the $N_0$ rays with neither surface nor bottom bounces, it will then estimate the current group velocity $u$ from a power weighted average of the ray travel times,
+For each grid point, BELLHOP produces a number of arrivals resulting from multiple propagation paths.
+Using only the $N_0$ rays with neither surface nor bottom bounces, it will then estimate the current effective sound speed $c$ from a power weighted average of the ray travel times,
 \begin{equation}
-u = \frac{\hat{r} \sum_{n=1}^{N_{0}} a_{n}^{2}}{\sum_{n=1}^{N_{0}} dt_{n}a_{n}^{2}} ~, 
+c = \frac{\hat{r} \sum_{n=1}^{N_{0}} a_{n}^{2}}{\sum_{n=1}^{N_{0}} dt_{n}a_{n}^{2}} ~, 
 \end{equation}
@@
 \begin{equation}
-\sigma_{u} \simeq \sqrt{\frac {\sum_{n=1}^{N_{0}} (dt_{n}-\hat{r}/u)^{2}a_{n}^{2}}{ \sum_{n=1}^{N_{0}} a_{n}^{2}} } \frac{u^{2}}{\hat{r}}
+\sigma_{c} \simeq \sqrt{\frac {\sum_{n=1}^{N_{0}} (dt_{n}-\hat{r}/u)^{2}a_{n}^{2}}{ \sum_{n=1}^{N_{0}} a_{n}^{2}} } \frac{u^{2}}{\hat{r}}
 \end{equation}
-If no direct paths exist, i.e. $N_{0}=0$, then the group velocity is computed using the same algorithm for the ray arrivals with one bounce, and so on.
+If no direct paths exist, i.e. $N_{0}=0$, then the effective speed is computed using the same algorithm for the ray arrivals with one bounce, and so on.
 
@@
 \begin{equation}
-r_{i,j} = u_{i,j} \Delta t_{i,j} 
+r_{i,j} = c_{i,j} \Delta t_{i,j} 
 \end{equation}
 
-Thus the NBC method assumes the signal detected by the modem will be dominated set of paths with the least number of boundary interactions.
+Thus the MBC method assumes the signal detected by the modem will be dominated by a set of paths with the least number of boundary interactions.
 Importantly, this stochastic, ensemble method for group velocity calculation can run in real-time, appearing to be orders of magnitude faster than other post-processing methods which seek to determine the specific ray itself that best matches a prominent indicator from the arrival structure.
@@
   \includegraphics[width=\reprintcolumnwidth]{figs/range-error-insitu.pdf}
-  \caption{The real-time range error by source (20, 30, and 90 m) and receiver (30 and 90 m) pairings for both sound speed estimates used during ICEX20.
+  \caption{The real-time range error by source (rows: 20, 30, and 90 m) and receiver (columns: 30 and 90 m) pairings for both sound speed estimates used during ICEX20.
 The amount of communication events is notated in the top right of each panel.
@@
 
-Figure \ref{fig:range-error-insitu} shows the range error boundary for both SSP inputs used in ICEX20.
-A promising sign that the MBC method adapts sound speed realistically is no obvious error growth as travel time increases.
+Fig. \ref{fig:range-error-insitu} shows the range error boundary for both SSP inputs used in ICEX20.
+A promising sign that the MBC method adapts sound speed somewhat intelligently is the lack of error growth as travel time increases.
 The baseline SSP (n=243 events) has an absolute pseudorange error of 11.38 $\pm$ 4.23 m; the weighted SSP (n=568), 11.36 $\pm$ 8.12 m.
@@
 Accounting for ice movement between beacons creates nominal ranges with small variability.
-Figs. \ref{fig:raytrace-zs20,fig:raytrace-zs30,fig:raytrace-zs90} each show eigenrays for three sound speed environments for source depths of 20, 30, and 90 m, respectively.
-Eigenrays were initially found using the built-in BELLHOP protocol with a launch angle step of 0.05 degrees from -60 to 60 degrees.
+Figs. \ref{fig:raytrace-zs20}, \ref{fig:raytrace-zs30}, and \ref{fig:raytrace-zs90} show eigenrays for three sound speed environments for source depths of 20, 30, and 90 m, respectively.
+Eigenrays were initially found using the built-in BELLHOP protocol with a launch angle fan of 2400 rays between -60 and 60 degrees.
 Separately, recorded travel times between beacons were clustered with 1 millisecond boundaries such that some source-receiver pairs had multiple, distinct travel times to approximate.
@@
 It should be noted that bottom bounces were recovered but filtered out.
-The three source depths create distinguishable ray geometries with respect to the three sound speed inputs.
+The three source depths create distinct ray geometries with respect to the three sound speed inputs.
 
@@
   \includegraphics[width=\reprintcolumnwidth]{figs/raytrace-3env-zs-20.pdf}
-  \caption{Eigenrays for beacon to beacon events for each sound speed with a nominal source depth of 20 m.
+  \caption{Eigenrays for beacon-to-beacon events for each sound speed with a nominal source depth of 20 m over a total ray fan in gray.
 The beacons are highlighted in color/marker coding in Fig.
-\ref{fig:overview}.
-The eigenrays are curated from BELLHOP by travel time proximity and are traced in the representative receiver colors over a total ray fan in gray.}
+\ref{fig:overview}.}
   \label{fig:raytrace-zs20}
@@
   \includegraphics[width=\reprintcolumnwidth]{figs/raytrace-3env-zs-30.pdf}
-  \caption{Eigenrays for beacon to beacon events for each sound speed with a nominal source depth of 30 m.
+  \caption{Eigenrays for beacon-to-beacon events for each sound speed with a nominal source depth of 30 m over a total ray fan in gray.
 The beacons are highlighted in color/marker coding in Fig.
-\ref{fig:overview}.
-The eigenrays are curated from BELLHOP by travel time proximity and are traced in the representative receiver colors over a total ray fan in gray.}
+\ref{fig:overview}.}
   \label{fig:raytrace-zs30}
@@
   \includegraphics[width=\reprintcolumnwidth]{figs/raytrace-3env-zs-90.pdf}
-  \caption{Eigenrays for beacon to beacon events for each sound speed with a nominal source depth of 90 m.
+  \caption{Eigenrays for beacon-to-beacon events for each sound speed with a nominal source depth of 90 m over a total ray fan in gray.
 The beacons are highlighted in color/marker coding in Fig.
-\ref{fig:overview}.
-The eigenrays are curated from BELLHOP by travel time proximity and are traced in the representative receiver colors over a total ray fan in gray.}
+\ref{fig:overview}.}
   \label{fig:raytrace-zs90}
@@
 The goal of the MBC algorithm was to provide a reliable, physically intuitive interpretation of the acoustic propagation without taking on the additional burden of regularly identifying specific paths that may connect any given source-receiver pair in the network.
-Its performance was adequate for vehicle navigation and would have likely sufficed if it were not for the prominence of the duct observed relative that of other model and data products. \llabel{2.14}
+\llabel{2.14}
+Its performance was adequate for vehicle navigation and would have likely sufficed if it were not for the prominence of the duct observed relative that of other model and data products.
 
@@
 From all events recorded during the modem test experiment, there are 1242 successfully decoded beacon-to-beacon events.
-Only these events are used to evaluate ranging accuracy, as the ICNN was not configured to use receptions with failure flags.
 Thus, a post-processing analysis that emulates the real-time navigation engine was run to overcome the unequal distribution of communication events with respect to depth, range, and sound speed status.
@@
 Because $\hat{r}$ converges to the correct solution, a comparison of $\hat{r}$ with the GPS-tracked range shows a normal, zero-centered distribution within the bounds of GPS drift.
-The present analysis therefore seeds realistic but ``omniscient'' knowledge of the extrapolated range and leverages the post-processing pipeline to more thoroughly evaluate the acoustic range estimate for all modem events, with three relevant sound speed sources, and both group velocity criterion.
+The analysis therefore seeds realistic but ``omniscient'' knowledge of the extrapolated range and emulates the post-processing pipeline to more thoroughly evaluate the acoustic pseudorange estimate for all modem events.
+Sound speed inputs are the isovelocity sound speed in addition to the modeled, baseline, and weighted SSPs from Fig. \ref{fig:sspExpectation}.
+The analysis replicates the MBC but also introduces a new filtering algorithm, the nearest bounce criteria (NBC), based on insights gleaned from the eigenray analysis.
 Accordingly, the results in this section evaluate the utility of the algorithms and sound speed sources, divorced from their role in the ICNN while maintaining real-time relevance.

### 2022-02-20 pipeline/4-analyze-range-error/plot_insitu_range_error_by_ssp.m

@@ -1,5 +1,6 @@
 %% prep workspace
 clear; clc; close all;
+addpath('../../src');
 
 %% load data
 A = load('../../data/tobytest-recap-clean.mat'); % loads "event"
@@ -70,13 +71,15 @@ for zs = [20 30 90]
             set(gca,'fontsize',13)
             
             % for all grids
-            title(sprintf('source depth = %u m',zs),'fontsize',14,'fontweight','bold');
-            text(2.2,25,sprintf('rx depth = %u m',zr),'HorizontalAlignment','right','VerticalAlignment','bottom','fontsize',11);
+            %title(sprintf('source depth = %u m',zs),'fontsize',14,'fontweight','bold');
+            title('  ','fontsize',8);
+            text(2.2,24,sprintf('tx depth = %u m',zs),'HorizontalAlignment','right','VerticalAlignment','bottom','fontsize',11);
+            text(2.2,24,sprintf('rx depth = %u m',zr),'HorizontalAlignment','right','VerticalAlignment','top','fontsize',11);
             
             if sum(index)~=1
-                text(2.2,25-3.*eof_status,sprintf('%s = %u events',eof_str,sum(index)),'HorizontalAlignment','right','VerticalAlignment','top','fontsize',11);
+                text(2.2,21-3.*eof_status,sprintf('%s = %u events',eof_str,sum(index)),'HorizontalAlignment','right','VerticalAlignment','top','fontsize',11);
             else
-                text(2.2,25,sprintf('%s = %u event',eof_str,sum(index)),'HorizontalAlignment','right','VerticalAlignment','top','fontsize',11);
+                text(2.2,21,sprintf('%s = %u event',eof_str,sum(index)),'HorizontalAlignment','right','VerticalAlignment','top','fontsize',11);
             end
             grid on
             
@@ -108,7 +111,7 @@ hold off
 legend(l2,'Baseline','Chosen Weights','location','northwest');
 
 % title
-sgtitle('In situ range error by source (20,30,90 m) and receiver (30,90 m) depths','fontsize',17,'fontweight','bold')
+sgtitle('Real-time pseudorange error by source and receiver depths','fontsize',17,'fontweight','bold')
 h_printThesisPNG('range-error-insitu');
 
 %% find mean, median, etc for baseval vs eof

### 2022-02-20 pipeline/3-analyze-gvel/plot_gvel_by_txrx_depth.m
adding in isovelicity lines

@@ -23,6 +23,7 @@ tiledlayout(2,3,'TileSpacing','none','Padding','none');
 shapeBounce = {'o','x','s','^','d'};
 
 count = 0;
+
 for zr = [30 90]
     index2 = DATA.recDepth == zr;
     
@@ -31,6 +32,8 @@ for zr = [30 90]
         
         count = count + 1;
         nexttile;
+        
+        hold on
         
         index = logical(index1.*index2.*INDEX.valid);
         
@@ -44,15 +47,9 @@ for zr = [30 90]
         sgps.MarkerEdgeColor = 'none';
         sgps.MarkerFaceColor = [200, 78, 0]./256;
         sgps.MarkerFaceAlpha = 0.2;
-
-        %p = patch([0 5 5 0 0],...
-            %[iso.avg-iso.std,iso.avg-iso.std,iso.avg+iso.std,iso.avg+iso.std,iso.avg-iso.std],'k','handlevisibility','off');
-        %p.FaceAlpha = 0.07;
-        %p.EdgeColor = 'none';
         
         if sum(index)>=1
             % plot
-            hold on
             for s = [5 3 4]
                 xval = DATA.owtt(index);
                 yval = T2{s}.gvel(index);
@@ -144,33 +141,7 @@ lgd = legend(lgdstr,'fontsize',12,'location','WestOutside');
 hold off
 legend boxoff
 
-% % add legend
-% nexttile(1);
-% 
-% % add legend 1 -- color
-% hold on
-% plot(NaN,NaN,'w');
-% plot(NaN,NaN,'k','linewidth',4);
-% 
-% for s = [5 3 4]
-%     plot(NaN,NaN,'color',colorSet{s},'linewidth',5);
-% end
-% plot(NaN,NaN,'w');
-% plot(NaN,NaN,'w');
-% 
-% % add legend 2 -- shape
-% for nb = 0:4
-%     scatter(NaN,NaN,shapeBounce{nb+1},'MarkerEdgeColor','k');
-% end
-% 
-% lgdstr = {'\bf{SOUND SPEED SOURCE}','Isovelocity','HYCOM','Baseline','Chosen Weights','','\bf{MULTIPATH STRUCTURE}','direct path','1 bounce','2 bounces','3 bounces','4 bounces'};
-% lgd = legend(lgdstr,'numcolumns',1,'fontsize',11,'location','WestOutside');
-% %title(lgd,'SSP Source & Multipath ID');
-% legend boxoff
[…]

### 2022-02-21 manuscript/main.tex
thru re-positioning results

@@
 For post-processing comparison, we introduce both the HYCOM profile and an isovelocity case, 1441.8 $\pm$ 3.7 m/s, as the mean and standard deviation of the observed sound speed profile over the first 200 m.
+This is a contrived value taken in the style of \citet{Graupe2019} for the sake of comparison; the default value in the LAMSS simulator, which was not environmentally informed nor used during ICEX20, was 1430 m/s.
 
@@
 
-\subsection{Eigenray identification for beacon-to-beacon events}
+\subsection{Eigenray identification for beacon-to-beacon events} \label{sec:eigenrays}
 
@@
 \llabel{2.14}
+While it was unlikely to resolve multipath arrivals that triggered successful modem detection, an isovelocity approach would have provided no adaptivity against source and receiver depth differences.
 Its performance was adequate for vehicle navigation and would have likely sufficed if it were not for the prominence of the duct observed relative that of other model and data products.
@@
 
-\llabel{2.12} As shown in the eigenray traces of Fig. \ref{fig:raytrace-zs30}, the extent of ray bending and repeated reflections is extremely dependent on the sound speed profile.
-An isovelocity approach would completely miss this nuance; our field-tested approach that only resolved the simplest path is unlikely to resolve the one that triggers modem detection.
+\llabel{2.12} The extent of ray bending and repeated reflections is extremely dependent on the degree of the Beaufort Lens observed.
 Based on this insight, a new algorithm, the nearest bounce criteria (NBC), is a slight modification from the MBC and includes multipath as a new dimension of information to exploit.
@@
 
-Given a running estimate for the horizontal group velocity $u_{i,j}$ between nodes $i$ and $j$, the navigation system has an extrapolated value for range, $\hat{r}$, and a recorded travel time, $\Delta t_{i,j}$.
-Instead of using only the $N_0$ rays with neither surface nor bottom bounces to estimate group velocity and subsequently moving to incremental number of bounces only when no valid direct path solutions exist, we solve for the power weighted average of the ray travel time for the $N_k$ rays with $k$ bounces,
+Given a running estimate for the effective sound speed $c_{i,j}$ between nodes $i$ and $j$, the navigation system has an extrapolated value for range, $\hat{r}$, and a recorded travel time, $\Delta t_{i,j}$.
+Instead of using only the $N_0$ rays with neither surface nor bottom bounces to estimate conversion speed, and subsequently moving to incremental number of bounces only when no valid direct path solutions exist, we solve for the power weighted average of the ray travel time for the $N_k$ rays with $k$ bounces,
 \begin{equation}
@@
 \end{equation}
-predict a group velocity,
+predict an effective sound speed,
 \begin{equation}
-u_{i,j} = \dfrac{\hat{r}}{t_{i,j,k}}
+c_{i,j} = \dfrac{\hat{r}}{t_{i,j,k}}
 \end{equation}
@@
 \begin{equation}
-r_{i,j} = u_{i,j}\Delta t_{i,j}
+r_{i,j} = c_{i,j}\Delta t_{i,j}
 \end{equation}
 
-This method selects a different group velocity based on the multipath arrival structure, as the detected arrival is not always the first arrival or the direct path and could even be masked by noise or blocked temporarily \citep{deffenbaugh1996acoustic}.
-We manually cap the number of bounces to 4 because of the smaller operational scale and the attenuation accrued with many surface interactions.
+This method selects a different effective sound speed based on the multipath arrival structure, as the detected arrival is not always the first arrival or the direct path and could even be masked by noise or blocked temporarily \citep{deffenbaugh1996acoustic}.
+We manually cap the number of bounces at four because of the smaller operational scale and the attenuation accrued with many surface interactions.
 Bottom bounces are not encoded separately because of ray's tendency to refract upward, not due to information limitations.
@@
 
-The minimal and nearest bounce algorithms are applied with the three sound speed inputs shown in Fig.~\ref{fig:raytrace-zs30}.
-The resulting predicted group velocities for all source depths are shown in Fig~\ref{fig:gvel-post}.
+The minimal and nearest bounce algorithms are applied with the three sound speed inputs shown in Figs.~\ref{fig:raytrace-zs20},~\ref{fig:raytrace-zs30},~\ref{fig:raytrace-zs90}.
+The resulting predicted effective sound speeds are shown in Fig. \ref{fig:gvel-post} for all source depths versus one way travel time.
 
@@
 \includegraphics[width=\columnwidth]{figs/gvel-txrxdepth-wIso-wGPS.pdf}
-\caption{A comparison of group velocity predictions for all beacon to beacon events in post-processing with a source depth of 30 m, with group velocity on the y-axis and recorded travel time on the x-axis.
-The left panel is for a receiver depth of 30 m; the right panel for 90 m.
+\caption{A post-processing comparison of effective sound speed predictions for all beacon-to-beacon events.
+The rows share receiver depth; the columns share source depth.
+The recorded travel time is on the x-axis and the predicted effective sound speed is on the y-axis.
 The sound speed source is indicated by color.
-The minimal and nearest bounce criterion are distinguished by different marker shapes, compared to the separately colored red dots showing the naive, data-driven group velocity calculation.}
+The isovelocity is shown as the mean $\pm$ the standard deviation.
+The minimal and nearest bounce criterion are distinguished by different marker shapes, compared to the separately colored red dots showing the implied calculation.}
 \label{fig:gvel-post}
@@
 
-The goal of the group velocity estimation is to converge towards the implied sound speed, i.e. the GNSS-derived range divided by the recorded travel time.
-For a 30 m receiver depth, the NBC shows more overlap with data-derived values as it classifies multipath more correctly.
-For a 90 m receiver depth, the overlap is less accurate due to computational constraints of a limited fan of rays entering the shadow zone rendering a less reliable simulated times of arrival packet.
+The goal of the effective sound speed prediction is to converge towards the implied sound speed, i.e. the GNSS-derived range divided by the recorded travel time.
+As the environmental and ray filtering method become better representations of the real ocean, the lower the expected mismatch is between the implied and estimated effective sound speeds.
 
-As the environmental and ray filtering method become better representations of the real ocean, the lower the expected mismatch is between the implied and estimated effective sound speeds.
-Analysis shows that the higher multipath classification produces more accurate sound speed predictions, likely driven by a tighter and/or sparser bundle of rays.
-However, that data are too small to draw significant conclusions.
-\llabel{1.19} Discontinuities in multipath classification verify our hypothesis for its importance to a smoothly varying group velocity, as shown in the cluster for a receiver depth of 30 m, where HYCOM jumps from one to two bounces amidst the baseline SSP and weighted SSP smoothly increasing while consistently seeing two and three bounces, respectively.
+The various sound speed inputs\textemdash isovelocity aside\textemdash not only modify the predicted effective sound speed, as seen by the colored vertical offsets, but often classify a distinct number of bounces.
+HYCOM sees the most direct and one bounce multipath structures, lending a bias for faster speeds; the weighted SSP sees the most double and triple bounces, favoring slower speeds; the baseline sound speed exists in between.
+Very rarely is the multipath structure classified as a direct path, where the MBC and NBC would prediction overlap.
+In fact, the higher the multipath classification, the more accurate the sound speed prediction is, likely driven by a tighter or even sparser bundle of rays.
+\llabel{1.19} Discontinuities in multipath classification provide initial evidence for its importance to a smoothly varying group velocity, as shown in the cluster of 30 to 30 m transmissions, where HYCOM jumps from one to two classified bounces amidst the baseline SSP and weighted SSP smoothly increasing while consistently seeing two and three classified bounces, respectively.
+Of course, the prediction deteriorates with cross-layer transmissions across the duct, but not to the same degree at which eigenrays could not be found for the weighted SSP in section \ref{sec:eigenrays}.
+The evidence suggests that the grid based method provides a useful amount of redundancy to resolve similar enough eigenrays.
+
+It is useful to think about in what case the isovelocity\textemdash or any isovelocity framing\textemdash would have been appropriate.
+The transmissions from shallow to shallow receiver would may have matched the default configuration of 1430 m/s.
+The isovelocity contrived for this paper, 1441.8 m/s, best matches the transmissions from 90 to 90 m.
+The one from \citet{Graupe2019}, 1450 m/s, would have had a systemic overestimation.
+Given that implied sound speeds just for beacon-to-beacon events span 1420 to 1445 m/s, it is safe to say that a nominal sound speed would sacrifice pseudorange accuracy somewhere, and that an adaptive approach is necessary even for short range operations in the Beaufort Lens.
 
@@
 
+Pseudorange estimation plays an important role in trilateration.
+Fig. \ref{fig:rangeError} shows the directional pseudorange error ``footprints'' for the four sound speed inputs with the NBC approach, separated by source and receiver depth configurations.
+
 \begin{figure}[!ht]
 \includegraphics[width=\textwidth]{figs/range-error-allMethods.pdf}
-\caption{The post-processed range error for source depths of 20, 30, and 90 m, and receiver depths of 30 and 90 m.
+\caption{The post-processed pseudorange error; the rows share receiver depth; the columns share source depth.
 The dashed gray line shows no error.
@@
 
-Fig. \ref{fig:rangeError} shows the directional range error footprints for all three sound speed inputs with respect to OWTT, separated by source and receiver depth configurations.
 The weighted SSP range error generally has the smallest and most zero-centered footprint.
 The one case it does not is for the source-receiver pairings between 30 and 90 m in depth.
-The increased error for these reciprocal transmission paths is most likely driven by the computational artifacts encountered when propagating through the steep sound speed gradients of the lens and through the shadow zone.
+The increased error for these is most likely driven by the computational artifacts encountered when propagating through the steep sound speed gradients of the lens and through the shadow zone.
 All other source depth pairings are significantly improved using the chosen weights compared to HYCOM or the baseline.
 
-When using a linear scaling to convert travel time into range, any offset between the assumed sound speed and the horizontal group velocity produces unconstrained error with increasing receiver distance.
-Most importantly, we see the consequences of the adaptive group velocity in that range error does not strictly increase with OWTT.
+When using a linear scaling to convert travel time into range, any offset between the assumed sound speed and the horizontal group velocity produces unconstrained error with increasing receiver distance, whereas an adaptive estimate would exhibit no such trend.
+This is easily observed in the same-layer links, i.e., 30 to 30 m and 90 to 90 m.
+In cross-layer links, the isovelocity does not perform better but tends to exaggerate or flip the footprint created adaptively.
 
-The improvement from MBC to NBC is most evident for the realistic sound speed; while the HYCOM SSP improves from a median absolute range error of 6.41 to 4.61 m, the baseline SSP improves from 10.30 to 2.27 m, and the weighted SSP improves from 13.28 to 2.12 m.
-Table \ref{tab:rangeErrorV1V2} shows further statistics on the absolute range error by SSP and group velocity algorithm.
-The order of magnitude improvement in the ducted SSPs demonstrate the effectiveness of the algorithm exploiting the multipath conditions.
+The improvement from MBC to NBC is most evident for the data-driven sound speed; while the HYCOM SSP improves from a median absolute range error of 6.41 to 4.61 m, the baseline SSP improves from 10.30 to 2.27 m, and the weighted SSP improves from 13.28 to 2.12 m.
+In comparison, the isovelocity has a median error of 13.09 m.
+The order of magnitude improvement in the ducted SSPs demonstrate the effectiveness of the NBC algorithm exploiting the observed multipath conditions.
 
-\begin{table}[h!]
-\renewcommand{\arraystretch}{1.5}
-\centering
-\begin{tabular}{r|cc|cc|cc}\toprule
- & \multicolumn{2}{c|}{\textbf{Baseline} } & \multicolumn{2}{c|}{\textbf{Chosen Weights} } & \multicolumn{2}{c}{\textbf{HYCOM}} \\
-  & \cellcolor[HTML]{EFEFEF}MBC & NBC & \cellcolor[HTML]{EFEFEF} MBC& NBC & \cellcolor[HTML]{EFEFEF}MBC & NBC \\ \hline
- minimum [m] 	& \cellcolor[HTML]{EFEFEF}0.01 & 0.00	& \cellcolor[HTML]{EFEFEF}0.00 	& 0.00 	& \cellcolor[HTML]{EFEFEF}0.11 & 0.01 \\
- 25th \% [m]   & \cellcolor[HTML]{EFEFEF}4.96 & 0.99	& \cellcolor[HTML]{EFEFEF}6.26 	& 0.95 	& \cellcolor[HTML]{EFEFEF}3.30 & 2.25 \\   
- median [m]		& \cellcolor[HTML]{EFEFEF}10.30 & 2.27 	& \cellcolor[HTML]{EFEFEF}13.28 & 2.12 	& \cellcolor[HTML]{EFEFEF}6.41 & 4.61 \\
- 75th \% [m]   & \cellcolor[HTML]{EFEFEF}15.81 & 5.51 	& \cellcolor[HTML]{EFEFEF}19.75 & 4.11 	& \cellcolor[HTML]{EFEFEF}10.92 & 7.46 \\
- maximum [m]    & \cellcolor[HTML]{EFEFEF}22.52 & 14.96 & \cellcolor[HTML]{EFEFEF}1491  & 20.21 & \cellcolor[HTML]{EFEFEF}19.55 & 15.81 \\
- \toprule
-\end{tabular}
-\caption[Comparison of post-processing range estimation algorithms across all events]{A comparison of range estimation metrics for each sound speed source and group velocity estimation algorithm for all 1283 beacon to beacon events via post-processing.
-The 0th (minimum), 25th, 50th (median), 75th, and 100th (maximum) percentiles are shown to the range resolution afforded by the WHOI Micro-Modem.
-There are a few outliers that drive the mean to be higher than the median.}
-\label{tab:rangeErrorV1V2}
-\end{table}
-
-As shown in table \ref{tab:rangeErrorV1V2}, there is a striking maximum range error of 1491 m for the weighted SSP in the minimal bounce criteria.
-There are 10 events from South transmitting at 30 m depth to North receiving at 30 m depth.
-The OWTT spread is from 2.1958 to 2.1963 s; the naive group velocity is 1429.3 to 1430.1 m/s; and the GPS-tracked range is from 3138.54 m to 3140.87 m.
-This example ends up being an excellent case study for how sound speed and multipath fidelity work in concert to minimize range error.
-The large error in this instance is driven by the MBC unexpectedly defaulting to a bottom bounce with a much greater OWTT.
-The NBC classifies the multipath as 4 bounces, reducing the range error from greater than a kilometer to less than a meter.
-While there is no actual way of knowing if this is the correct multipath structure, the range error is remarkably small, at 0.025\%. 
-This pattern of not choosing the minimal observed bounce structure is consistent across all SSPs; the baseline goes from 1 to 3 bounces and HYCOM goes from 0 to 2 bounces.
-Notably, the baseline and HYCOM range errors are never egregiously large, but are nonetheless improved with the NBC algorithm.
-Thus, for acoustically complex environments, the NBC has a disproportionately positive impact as the estimated SSP approaches the desired SSP.
+There is one example that helpfully illustrates the improvement brought upon by bounce classification.
+For transmissions between North and South at 30 m, the OWTT spread is 2.1958 to 2.1963 s; the GNSS-tracked distance is 3138.54 to 3140.87 m; and the implied effective sound speed is 1429.3 to 1430.1 m/s.
+For these transmissions, the weighted SSP and the MBC approach produce a pseudorange error of -1491 m, as the effective sound speed is dominated by bottom bounce arrivals with much greater travel times.
+The NBC approach categorizes this same record as a quadruple surface bounce, reducing the pseudorange error to less than a meter.
+Comparatively, the NBC approach for HYCOM and the baseline SSP produce pseudorange errors of 8.30 and 2.39 m, respectively. 
+There is strong evidence to suggest that the sound speed and multipath fidelity codependently improve localization accuracy.
 

### 2022-02-21 pipeline/4-analyze-range-error/plot_post_range_error_nbc.m

@@ -116,4 +116,4 @@ title(lg,'Sound Speed Inputs');
 
 % title
 sgtitle('Range error by source and receiver depths','fontsize',17,'fontweight','bold')
-h_printThesisPNG('range-error-allMethods')
+% h_printThesisPNG('range-error-allMethods')

### 2022-02-21 pipeline/6-compare-post-mbc-nbc/plot_crossmap_post_mbc_nbc.m

@@ -13,16 +13,24 @@ SIM_OLD = h_get_mbc(listingOld,DATA);
 listingNew = dir('../bellhop-gvel-gridded/csv_arr/*gridded.csv');
 SIM_NEW = h_get_nbc(listingNew,DATA,INDEX);
 
+%% load iso
+load isovelocity-ssp.mat
+
+%% isovelocity
+iso.xVal = iso.avg .* DATA.owtt(INDEX.valid) - DATA.recRange(INDEX.valid);
+iso.yVal = iso.xVal;
+iso.F = h_cross_plot_stat(iso.xVal,iso.yVal);
+
 %% plot things
 
-global shapeBounce colorDepth;
+% global shapeBounce colorDepth;
 
 % marker shape & color
-shapeBounce = {'o','x','s','^','d'};
-colorDepth = containers.Map([20 30 90],{[70 240 240]./256,[0 130 200]./256,[0 0 128]./256});
+% shapeBounce = {'o','x','s','^','d'};
+% colorDepth = containers.Map([20 30 90],{[70 240 240]./256,[0 130 200]./256,[0 0 128]./256});
 
-figure('name','compare-method-postv1-postv2','renderer','painters','position',[108 108 1470 490]);
-t = tiledlayout(1,3,'Padding','compact','TileSpacing','Compact');
+% figure('name','compare-method-postv1-postv2','renderer','painters','position',[108 108 1470 490]);
+% t = tiledlayout(1,3,'Padding','compact','TileSpacing','Compact');
 
 %% figure --- baseval
 
@@ -31,13 +39,13 @@ baseval.xVal = SIM_OLD{3}.rangeAnomaly(indValid);
 baseval.yVal = SIM_NEW{3}.rangeAnomaly(indValid);
 baseval.zs = DATA.sourceDepth(indValid);
 baseval.numBounces = SIM_NEW{3}.numBounces(indValid);
-baseval.F = h_cross_plot(baseval.xVal,baseval.yVal,baseval.zs,baseval.numBounces);
+baseval.F = h_cross_plot_stat(baseval.xVal,baseval.yVal);
 
-nexttile;
-h_view_plot(baseval.xVal,baseval.yVal,baseval.zs,baseval.numBounces);
-title('SSP = Baseline','fontsize',13);
-xlabel({'minimal bounce criterion error [m]'});
[…]

### 2022-02-21 manuscript/main.tex
thru GPS section

@@
 The ICNN is comprised of four ice buoys in a loose rectangle, roughly 2 km away from a central ice camp with a topside computer, as shown in Fig.~\ref{fig:icnnOverview}.
+Each ice buoy is outfitted with a Garmin GPS 18x, with a pulse-per-second rising edge aligned to 1 microsecond and a spec sheet accuracy of 3 m, 95\% of the time.
 The AUV and each ice buoy are outfitted with a WHOI Micro-Modem \citep{Singh2006}, with a four-element receiver array, a single transmitter, and one-tenth of a millisecond resolution.
@@
 Ice buoy modems were run as ``virtual vehicles'' at a fixed depth, receiving position updates from the other beacons as well as a camp site modem lowered to 20 m. \llabel{1.13}
-Figure \ref{fig:overview} shows successful events sorted by source depth.
+Fig. \ref{fig:overview} shows successful events sorted by source depth.
 In this analysis, we assume there is insignificant displacement between the GNSS puck surface expression and subsurface modem; this is supported by unusually low observed ice drift rates, just 0.7 cm/s on average throughout the mission.
@@
 This section aims to resolve that tension by reevaluating the trilateration results with respect to the MBC and NBC algorithms.
-The MBC/NBC sound speed estimates were tracked independently for each transmitter-receiver pair; although the sound speed was expected to be locally smooth near a given receiver, no such assumption was enforced between distinct acoustic links.
+The MBC/NBC effective speed predictions were tracked independently for each source-receiver pair; although the sound speed was expected to be locally smooth near a given receiver, no such assumption was enforced between distinct acoustic links.
 
@@
 The NBC, however, better resolves an approximation of the acoustic path.
-For example, in some trilateration solutions for the Eastern buoy, the MBC shows a correction of more than a kilometer; the NBC is two order of magnitudes less.
+For example, in some trilateration solutions for the Eastern buoy, the MBC shows a correction of more than a kilometer; the NBC is two orders of magnitudes less.
 
@@
 
-The AUV dataset clearly exhibits instances where a receiver detects the same transmission more than once.
+In comparison to the modem experiment, the AUV data clearly exhibit instances where a receiver detects the same transmission more than once.
 This is not surprising considering the complex multipath provided by the Beaufort Lens.
-The 11 hour vehicle mission contains 3260 transmissions, 12938 total detections, and 4704 successful receptions.
+The 11 hour vehicle mission contains 3,260 transmissions, 12,938 total detections, and 4,704 successful receptions.
 Allowing receptions with PSK errors would almost double the number of recorded multipath arrivals exploited for positioning, if a real-time solution could correctly parse paths from different arrivals in the same thirty-second cycle.
@@
 Thus, larger corrections are not necessarily indicative of worse performance.
-Navigation accuracy may be better described by trilateration error, the RMS of the remaining pseudorange errors from each acoustic link.
+Navigation accuracy is better described by trilateration error, the RMS of the remaining pseudorange errors from each acoustic link.
 
@@
 The effect is minor for surface platform navigation \textemdash roughly 15 m of horizontal precision has been displayed at the North Pole\textemdash but is significant enough to register against the modem's detected travel times.
-Figure \ref{fig:gps-drift-example} zooms in on the GNSS and OWTT noise relative to the ice movement for two pairs of modem buoy connections.
+Fig. \ref{fig:gps-drift-example} zooms in on the GNSS and OWTT noise relative to the ice movement for two pairs of modem buoy connections.
 The two panels indicate the GPS drift as $\delta R = \sqrt{\delta x^2 + \delta y^2}$ and temporal drift, $\delta t$, relative to the median OWTT recorded between the two modems.
@@
 	\includegraphics[width=\reprintcolumnwidth]{figs/gps-drift-example.pdf} 
-	\caption{A comparison of GPS drift (y-axis) versus OWTT drift (x-axis), colored by source and receiver depth.
-The physical link between North and East are shown on the top; South and West is on the bottom.}
+	\caption{A comparison of GPS drift (y-axis) versus OWTT drift (x-axis), colored by source and receiver depth.}
 	\label{fig:gps-drift-example}
@@
 The top panel shows the connections between the North and East buoys.
-There is relative, i.e. non-rigid, ice movement between the North and East buoys, evidenced by events spanning the dashed line.
-But the height of the scatter plot is indicative of the precision of the GPS signal, as it remains consistent across many arrival time bands.
-Naturally, some minor offsets between these vertical bands relate to different operational configurations of source and receiver depth.
-However, the large majority of events show vertical banding for the same nominal $\delta t$, indicating the amount of GPS drift.
-
-This idea of GPS drift relative to the same OWTT measurements is further indicated by events between the other two buoys, South and West, in the bottom panel.
-These buoys are moving in a more rigid ice floe and there is minimal impact by source and receiver depth on the spread of OWTT.
-The GPS drift is much larger relative to the OWTT drift, which is sensitive to acoustic scattering, multipath, and/or environmental microstructure.
-
-These are just two subsets of the physical links that cover all four GPS modem buoys.
-The GPS at camp is the least accurate due to the human activity and infrastructure occluding the physical puck.
+The clusters scaling along the 1440 m/s guideline suggest relative ice movement picked up by both GNSS and OWTT.
+But the vertical distribution across many arrival time bands is indicative of the GNSS fluctuations in precision, or noise.
+Some minor offsets between these vertical bands relate to different operational configurations of source and receiver depth.
+The idea of GNSS noise relative to OWTT is further indicated by events between two other buoys, South and West.
+The relatively thin time window suggests these buoys are moving in a more rigid ice floe and that there is minimal impact by source and receiver depth on the spread of OWTT.
+Yet, the vertical distribution, spanning almost 4 meters, cannot be explained by time differentials due to acoustic scattering, multipath, and/or environmental microstructure.
 

### 2022-02-21 manuscript/main.tex
GNSS instead of GPS on plot

@@
 	\centering
-	\includegraphics[width=\reprintcolumnwidth]{figs/gps-drift-example.pdf} 
-	\caption{A comparison of GPS drift (y-axis) versus OWTT drift (x-axis), colored by source and receiver depth.}
+	\includegraphics[width=\columnwidth]{figs/gps-drift-example.pdf} 
+	\caption{A comparison of GPS drift (y-axis) versus OWTT drift (x-axis) for corners of the ICNN network with different source depths.}
 	\label{fig:gps-drift-example}

### 2022-02-21 pipeline/8-gps-check/analyze_GPS_drift.m

@@ -154,8 +154,8 @@ xlim(xbounds2);
 xticks(xticks2);
 ylim(xbounds2.*sspGVEL);
 yticks(xticks2.*sspGVEL);
-title('GPS noise between North and East buoys','fontsize',15);
-ylabel('GPS \deltaR [m]');
+title('GNSS noise between North and East buoys','fontsize',15);
+ylabel('GNSS \deltaR [m]');
 xlabel('\deltat [ms]');
 
 % panel 4
@@ -166,25 +166,28 @@ xlim(xbounds2);
 xticks(xticks2);
 ylim(xbounds2.*sspGVEL);
 yticks(xticks2.*sspGVEL);
-title('GPS noise between South and West buoys','fontsize',15);
+title('GNSS noise between South and West buoys','fontsize',15);
 %ylabel('GPS \deltaR [m]');
 xlabel('\deltat [ms]');
 
 %% legend
 % add legend
 hold on
+
+plot(NaN,NaN,'w');
 for s = [20 30 90]
     plot(NaN,NaN,'color',colorDepth(s),'linewidth',6);
 end
 
+plot(NaN,NaN,'w');
 for r = [20 30 90]
     plot(NaN,NaN,sourceDepth(r),'color','k')
 end
 hold off
-lgdstr = {' 20 m',' 30 m',' 90 m',' 20 m',' 30 m',' 90 m'};
+lgdstr = {'\bf{tx depth (color)}',' 20 m',' 30 m',' 90 m','\bf{rx depth (shape)}', ' 20 m',' 30 m',' 90 m'};
 
 lg1 = legend(lgdstr,'location','eastOutside','NumColumns',1,'fontsize',10);
-title(lg1,{'tx depth (color)','rx depth (shape)'});
+%title(lg1,{'tx depth (color)','rx depth (shape)'});
 
 %% export
 h_printThesisPNG('gps-drift-example');

### 2022-02-21 manuscript/main.tex
revision done -- needs linking to rebuttal letter

@@
 
+The Arctic, while remote, is the perfect place to demonstrate mature navigation technologies in real GNSS-denied conditions.
 Thorough reviews of uncrewed vehicle operations in polar environments can be found in \citep{Norgren2014} and \citep{Barker2020}; there is no comparable work in the Arctic for a short range AUV deployment in the Beaufort Lens.
@@
 
-This method selects a different effective sound speed based on the multipath arrival structure, as the detected arrival is not always the first arrival or the direct path and could even be masked by noise or blocked temporarily \citep{deffenbaugh1996acoustic}.
+Whereas the MBC outputs a scalar, this method first outputs a vector of effective sound speeds based on the number of reflections.
+Then a single value is selected that best matches the recorded travel time, as the detected arrival is not always the first arrival or the direct path and could even be masked by noise or blocked temporarily \citep{deffenbaugh1996acoustic}.
 We manually cap the number of bounces at four because of the smaller operational scale and the attenuation accrued with many surface interactions.
@@
 Yet, the vertical distribution, spanning almost 4 meters, cannot be explained by time differentials due to acoustic scattering, multipath, and/or environmental microstructure.
+This conclusion corroborates the vertical spread of implied effective speeds in Fig. \ref{fig:gvel-post}.
 
+\clearpage
 \section{Discussion}
 
-Given the computational constraints of real-time modeling, the gridded approach facilitates enough multipath classification to build in a ``ray ensemble'' of characteristic group velocities.
-This result is not always possible when aiming to find eigenrays to just an individual point, even with a higher density of launch angles.
-An important takeaway for those interested in ray-based localization is leveraging a local grid to give BELLHOP tolerance for finding solutions that otherwise may not be found in a center or single point solution.
-The limitations of numerical computation, particularly for a complex environment, are more adeptly addressed by accepting some uncertainty in position than by prescribing an exact solution.
-Even though BELLHOP is a greedy estimator for eigenrays, the additional data created is a negligible burden.
+Underwater navigation research is broadly motivated by acquiring GNSS-like navigation in GNSS-denied conditions.
+Accurate range estimation is essential to mitigating error.
+Current approaches for underwater acoustic navigation simplify the non-linear relationship between a SSP and timefronts with a determinstic sound speed.
+Thus, the conversion of travel time into distance can be pre-conditioned for error and error growth over the course of a vehicle mission.
+This work introduces a lightweight stochastic prediction of an effective sound speed along the path between source and receiver, retooling arrival methods generally deemed too complex or labor intensive for real-time.
+We assume that the effective sound speed would be a locally smoothly varying function with respect to operational conditions\textemdash horizontal and vertical differences and rate of difference between source and receiver.
+The field-tested approach, the minimal bounce criteria, facilitated a successful AUV recovery in a total ice-covered, double ducted environment.
+The accuracy of the MBC was validated against GPS-linked beacon-to-beacon communications.
+Given a consistent bias towards overestimation, an improved algorithm, the nearest bounce criteria, was developed on the insight that multipath structure may play an outsized role in maintaining a smoothly varying effective sound speed.
+The NBC was developed with field data and reevaluated on vehicle data, achieving a position accuracy and precision that rivals that of the deployed GNSS puck.
 
-Underwater navigation research is broadly motivated by acquiring GPS-like navigation accuracy in GPS-denied conditions.
-The Arctic, while remote, is the perfect place to test mature navigation technologies in real GPS-denied conditions.
-
-Range estimation is an essential step of acoustic localization and navigation.
-Current approaches in real-time underwater acoustic navigation simplify the non-linear relationship between a sound speed profile and acoustic propagation with a deterministic sound speed.
-Some post-processing approaches attempt to leverage the acoustic arrival structure via various ray methods, but often use a singular SSP for simplicity, even over long term missions or dynamic conditions.
-Thus, the conversion from travel time to range, particularly for real-time vehicle deployments, can be pre-conditioned for error growth as the OWTT/range increases.
-
-For GPS-denied navigation, especially in hostile environments like the Arctic, tolerance for error is close to none.
-This work addresses a critical need in acoustic navigation by retooling acoustic arrival methods generally deemed too complex or labor intensive for real-time, ray-based range estimation to achieve GPS-like positioning.
-
-We hypothesize and validate that the embedded stochastic prediction of a single group velocity is a smoothly varying function of range, source and receiver depth pairings, as well as multipath structure.
-We employ a GPS-tracked experiment to have a ground truth comparison for real-time localization algorithms.
-The real-time system achieves GPS-like navigation for an AUV without taking into account multipath structure; the ranging error improves by an order of magnitude with the suggested multipath adaptability, minimizing range error to single meters.
-Post-processing analysis shows that this method of ranging is sensitive to GPS drift itself; thus embedding a stochastic prediction of the horizontal group velocity has an outsized benefit to minimizing trilateration error.
+A key insight for both approaches was seeking an eigenray ensemble around an estimated location instead of seeking to unambigously match arrivals.
+The ensemble diversified the simulated multipath possibilities to better capture the actual multipath recorded.
+In this way, the solution exploits multipath, generally viewed as a source of uncertainty, as a new dimension of information to improve localization accuracy.
+Based on the navigation and re-navigation results of our AUV deployment in the ice-covered Beaufort Sea, we conclude that embedding a model-aided prediction of the effective sound speed has an outsized benefit to minimizing trilateration error, and that our approach sufficiently resolves the acoustic timefronts for an unpredictable, complex propagation environment like the double ducted Beaufort Lens.
 
@@
 
-The literature in underwater acoustic navigation and positioning is either real-time or physics-based.
-In this paper we demonstrate a field-tested approach that is both real-time and physics-based; this is achieved by coupling data streams with fast acoustic modeling.
-The methods exploit the upward refracting nature and the total ice cover of the Arctic environment to achieve remarkable ranging accuracy and precision.
-It transforms multipath, widely considered as an obstacle for acoustic ranging, into a new information content to refine ranging accuracy.
-We believe that this work enables more accurate range estimation, localization, and/or navigation for any field experiment given known source and receiver depths.
+The relatively simple nature of this approach suggests it is transferable to other environments, spatio-temporal scales, and platforms.
+While it is likely a particular quirk of the Beaufort Lens that filtering for reflection alone can produce a horizontal effective speed that compensates for ray refraction and reflection, it is trivial to filter along other ways, like number of turning points, to create a more diverse and informed set of multipath timefronts.
+Though the majority of re-navigation results are within single-meter accuracy, future work can examine how constellations of more LBL beacons can extend the operational domain without adding an undesirable amount of error.
+One possibility is that, during a mission, ICNN-like LBL implementions use a comparison of the GNSS self-position and acoustic positioning to invert for the ocean volume, linking how vertical and horizontal sound speed structure impact transmission integrity.
+\llabel{1.20} A fast tomographic estimate \citep{Deffenbaugh1997,Elisseeff2002}, along with its uncertainty, can be continuously communicated to assets underway to maintain contact or enable adaptive sampling.
+In this sense navigation and tomography converge on the same set of component technologies\textemdash position estimation, sound speed parameterization estimation, ray path identification, and vehicle path optimization.
 
-Performance in other acoustic environments may require introducing a different thresholded metric to sort time of arrivals.
-Given the NBC algorithm's improvement with increased multipath, its effectiveness is likely only challenged by the valid operational scales of a range independent propagation environment.
-For mesoscale operations, like that of many gliders, the group velocity criteria may need to be modified to better account for variability driven by range dependent propagation through internal waves, eddies, or even bathymetric changes like underwater volcanoes.
-BELLHOP simulations provide other relevant eigenray information, like time and angle of arrival, that is ripe for statistical and machine learning methods to classify a representative group velocity.
-A bespoke and fast ray tracing method, which was used during ICEX20 to evaluate the acoustic fit of sound speed profile parameterization \citep{bhatt_embedded_2022}, can easily report back the number of turning points instead of the number of bounces for multipath classification.
+Spatio-temporal variability is a serious challenge for accurate real-time ranging.
+On one hand, the effectiveness of eigenray filtering algorithm is likely only challenged by the valid operational scales of a range independent propagation environment.
+Longer range experiments may provide more time for eigenray filtering.
+A bootstrapping approach that filters eigenrays for several randomly generated internal wave spectrums may compensate for otherwise unknowable spatio-temporal variability.
+The model-aided component to the eigenray filtering is compatible with vertical slices from any physically driven ocean model.
+But in the long run, more accurate and higher resolution global circulation models are needed to properly resolve features that alter ducted propagation at the scales discernible to an acoustic modem.
+Through-the-sensor methods can resolve local features but would require a degree of information sharing not readily supported on the acoustic channel for large scale variability.
+But addressing the spatial and temporal scales of what can be solved deterministically and what must be solved stochastically imposes a resolution constraint that is at odds with computational overhead for real-time operations.
+Resolving features inaccurately, or with a false sense of confidence, could be more harmful than contextualizing the limitations of a range independent propagation over realistic bathymetry.
+Given that AUV operations are often on smaller spatial and temporal scales, the added benefit of a gridded model is quite small, and for features like the Beaufort Lens, not well resolved.
 
-This approach will start to break down in extremely dynamic environments, like fast moving fronts.
-Realistic \textit{in situ} considerations of the acoustic environment may not be possible without complete through-the-sensor integration of acoustic dat and/or hyper realistic ocean models.
-
-\llabel{1.20} Many approaches to underwater navigation combine it with acoustic tomography, i.e., a joint estimation of both source and receiver locations and the ocean volume between them.
-There has been considerable success at this effort in post-processing methods, which utilize intensive\textemdash and due to the non-linearity of sound propagation, often brute force\textemdash computational methods.
-For vehicle operations, fast tomography is the ideal implementation, in that one can fully consider how sound speed structure, horizontally and vertically, influences sound propagation.
-AUVs can serve as moving sources to better image the ocean volume \citep{Deffenbaugh1997,Elisseeff2002}, where mobile tomography and navigation converge on the same set of component technologies: position estimation, sound speed parameterization estimation, ray path identification, and vehicle path optimization.
-
-But there are overwhelming challenges, operationally and computationally, for fast, mobile tomography to become a realistic endeavor.
-Addressing the spatial and temporal scales of what can be solved deterministically and what must be solved stochastically imposes a resolution constraint on the utility of gridded models\textemdash resolving fine features inaccurately (or with a false sense of confidence) could be more harmful than assuming range independence.
-Given that AUV operations are often on small spatial and temporal scales, the added benefit of a gridded model is quite small, and in cases like the Arctic, may actually mischaracterize the ocean volume.
-For gliders, with longer and larger operational scales, an ocean model may provide more useful information.
-Currently gliders are low power and do not have the storage or computational power to run a full-scale, realistic ocean model.
-A lightweight representation of the key environmental and acoustic features, passed through the same manner of acoustic message from the modem experiment, may drastically improve glider navigation.
+The methods presented in this paper, including the software projects \citep{benjamin2010nested,schneider2010unified,schneider2015dynamic}, are open source and platform agnostic.
+Large AUVs, often large enough to support long duration and/or deep sea missions, would benefit from including diurnal or tidal effects for ranging.
+Gliders, though generally low power and memory, have been equipped with acoustic modems.
+Their inability to maintain position within a region of reliable acoustic path makes the impact of an environmentally adaptive pseudorange estimation disproportionately positive.
+The exact adjustments to the ensemble eigenray filtering are predicated on the expected sound speed conditions and acoustic arrival structure; the problem is ripe application for other simulation testbeds or machine learning methods.
+The continued development of embedded acoustic processing on heterogenous platforms is fundamental to support a universal underwater navigation scheme comparable to GNSS.
 

### 2022-02-21 manuscript/main.tex
fixing broken bibs

@@
 While land and air-based robots utilize information from Global Navigation Satellite Systems (GNSS) to achieve stunning location accuracy and precision throughout the duration of their missions, AUVs cannot access GNSS fixes while underwater.
-Therefore, underwater vehicles have relied on any combination of dead reckoning, hydrodynamic models, inertial navigation systems, doppler velocity logs, and acoustic baseline positioning systems for navigation \citep{Paull2014}.
+Therefore, underwater vehicles have relied on any combination of dead reckoning, hydrodynamic models, inertial navigation systems, doppler velocity logs, and acoustic baseline positioning systems for navigation \citep{paull_auv_2014}.
 Limiting navigation error and drift requires an AUV to periodically stall on the surface and obtain a GNSS fix to reset its position error.
@@
 % introduce LBL and precise statement of work
-Of underwater acoustic navigation systems, long baseline (LBL) is the most GPS-like in style and scale, and most appropriate for mitigating drift without overburdening computation or payload size on the vehicle \citep{VanUffelen2021}.
-The state-of-the-art for LBL outsources depth to a pressure sensor and solves the two-dimensional localization problem with an isovelocity, linear scaling between one way travel time (OWTT) and range \citep{Eustice2006,Eustice2007,Webster2009,Webster2012}.
+Of underwater acoustic navigation systems, long baseline (LBL) is the most GPS-like in style and scale, and most appropriate for mitigating drift without overburdening computation or payload size on the vehicle \citep{van_uffelen_global_2021}.
+The state-of-the-art for LBL outsources depth to a pressure sensor and solves the two-dimensional localization problem with an isovelocity, linear scaling between one way travel time (OWTT) and range \citep{Eustice2006,Eustice2007,webster_advances_2012,Webster2009}.
 This assumption is valid for short scale operations but oversimplifies propagation for larger and/or complex acoustic environments.
@@
 % PNT
-For consistency, we delineate specific definitions for timing, positioning, and navigation from \citet{howe2019observing}.
+For consistency, we delineate specific definitions for timing, positioning, and navigation from \citet{Howe2019}.
 \begin{enumerate}
@@
 % More about LBL navigation
-While RAFOS floats championed one way ranging for re-positioning \citep{Rossby1986,Duda2006}, the ability to do so for navigation was facilitated by \llabel{1.11} the advent of the WHOI Micro-Modem \citep{Singh2006} and synchronized clocks \citep{rypkema_one-way_2017}.
-AUV navigation efforts have achieved root mean square (RMS) localization error on the order of tens of meters relative to GNSS surface position over less than ten kilometers in shallow \citep{Eustice2007,Kepper2017,Claus2018} and deep water \citep{Kunz2008,Jakuba2008,Webster2009}.
+While RAFOS floats championed one way ranging for re-positioning \citep{Rossby1986,duda_evaluation_2006}, the ability to do so for navigation was facilitated by \llabel{1.11} the advent of the WHOI Micro-Modem \citep{Singh2006} and synchronized clocks \citep{rypkema_one-way_2017}.
+AUV navigation efforts have achieved root mean square (RMS) localization error on the order of tens of meters relative to GNSS surface position over less than ten kilometers in shallow \citep{Eustice2007,Claus2018,kepper_mems_2017} and deep water \citep{Kunz2008,jakuba_long-baseline_2008,Webster2009}.
 However, these efforts all used a nominal sound speed for travel time conversion and the vehicles were limited to shallower isovelocity regimes.
@@
 A follow up study investigated how a single temporally and spatially averaged SSP could mitigate position error for a four month glider mission \cite{vanuffelen2016localization}.
-\citet{Wu2019} cross correlate three days of real acoustic records with synthetic ones generated through ocean model snapshots from HYCOM \cite{Chassignet2007}.
+\citet{Wu2019} cross correlate three days of real acoustic records with synthetic ones generated through ocean model snapshots from HYCOM \cite{chassignet_hycom_2007}.
 While potentially applicable for various ocean states, this is reliant on model realism and impractical for real-time operations.
-\citet{Mikhalevsky2020} introduces a ``cold start'' algorithm that does not require prior knowledge of track, position, or sound speed information.
-The algorithm inputs a four-dimensional ocean model, constrained by tomography data, into a range dependent ray code to isolate the last path detected in a full multipath pattern.
+A ``cold start'' algorithm that does not require prior knowledge of track, position, or sound speed information inputs a four-dimensional ocean model, constrained by tomography data, into a range dependent ray code to isolate the last path detected in a full multipath pattern \citep{Mikhalevsky2020}.
 Then, a representative group speed is solved for alongside position in a least squares fashion. 
@@
 
-The ICEX20 AUV deployment necessitated an environmentally- and physically-driven relationship between recorded travel times and estimated pseudoranges due to the multipath uncertainty brought upon by an increasingly observed double ducted environment in the Beaufort Sea, which some refer to as the ``Beaufort Lens'' \citep{litvak2015acoustics,Chen2019,Chen2020}.
+The ICEX20 AUV deployment necessitated an environmentally- and physically-driven relationship between recorded travel times and estimated pseudoranges due to the multipath uncertainty brought upon by an increasingly observed double ducted environment in the Beaufort Sea, which some refer to as the ``Beaufort Lens'' \citep{litvak2015,chen_spectral_2019,chen_temporal_2020}.
 
 Given that a lens introduces significant ray refraction, the Beaufort Lens is a shorthand for the spatio-temporal variability of the local temperature and sound speed maxima generally around 50 to 60 m in depth.
-A neutrally buoyant layer of warm Pacific Summer Water creates a unique double ducted environment \textemdash the upper duct degrades signal coherence due to intensified ice interaction and the lower duct effectively traps sound for long range propagation \citep{Poulsen2017}.
-Modeling output \citep{Duda2019,Duda2021} and experimental observations \citep{Ballard2020,Bhatt2021} suggest that, in the Beaufort Sea, the duct is persistent and widespread but not necessarily continuous; it and its acoustic effects can be non-existent, minimal, or drastic.
+A neutrally buoyant layer of warm Pacific Summer Water creates a unique double ducted environment \textemdash the upper duct degrades signal coherence due to intensified ice interaction and the lower duct effectively traps sound for long range propagation \citep{Poulsen2016}.
+Modeling output \citep{duda_long-range_2019,duda_effects_2021} and experimental observations \citep{Ballard2020,Bhatt2021} suggest that, in the Beaufort Sea, the duct is persistent and widespread but not necessarily continuous; it and its acoustic effects can be non-existent, minimal, or drastic.
 Transmissions in the upper duct, between the surface ice and the lens, experience minimal attenuation but degrade in signal coherence with repeated reflections under the ice.
-In lower duct, between the lens and its conjugate depth in the Atlantic water, roughly 200 m, sound above 350 Hz is trapped near losslessly for long range propagation \citep{Poulsen2017}.
+In lower duct, between the lens and its conjugate depth in the Atlantic water, roughly 200 m, sound above 350 Hz is trapped near losslessly for long range propagation \citep{Poulsen2016}.
 
 The Arctic, while remote, is the perfect place to demonstrate mature navigation technologies in real GNSS-denied conditions.
-Thorough reviews of uncrewed vehicle operations in polar environments can be found in \citep{Norgren2014} and \citep{Barker2020}; there is no comparable work in the Arctic for a short range AUV deployment in the Beaufort Lens.
-Seminal \citep{brooke1981arcs,jackson1983autonomous,light1989autonomous,bellingham1995auv,hayes2002determining} and more recent AUV deployments \citep{Jakuba2008,Kunz2008,kukulya2010under,plueddemann_autonomous_2012,timmermans2013scales,fossum2021adaptive} witnessed the classical upward refracting sound speed profile that is amenable to an isovelocity assumption.
+Thorough reviews of uncrewed vehicle operations in polar environments can be found in \citet{norgren2014unmanned} and \citet{Barker2020}; there is no comparable work in the Arctic for a short range AUV deployment in the Beaufort Lens.
+Seminal \citep{brooke1981arcs,jackson1983autonomous,light1989autonomous,bellingham1995auv,hayes2002determining} and more recent AUV deployments \citep{,jakuba_long-baseline_2008,Kunz2008,Kukulya2010,Plueddemann2012,timmermans2013scales,fossum2021adaptive} witnessed the classical upward refracting sound speed profile that is amenable to an isovelocity assumption.
 
@@
 Unintuitively, the smaller nature of our deployment during ICEX20 is not a simplifying factor.
-For source depths typical to vehicle operations, 30 to 200 m, a shadow zone spans from 2 to 6 kilometers in range \citep{Schmidt2016}.
+For source depths typical to vehicle operations, 30 to 200 m, a shadow zone spans from 2 to 6 kilometers in range \citep{schmidt_acoustic_2016}.
 
@@
 The results from this paper derive from data taken while deploying the AUV \emph{Macrura}, a custom Bluefin-21, during the Ice Exercise 2020 (ICEX20), in the Beaufort Sea, from March 8th to 11th.
-The AUV deployment was supported by the Integrated Communication and Navigation Network (ICNN) \citep{Schneider2020,Randeni2020,randeni_high-resolution_2021}, a specialized implementation of the LBL solution.
+The AUV deployment was supported by the Integrated Communication and Navigation Network (ICNN) \citep{schneider_self-adapting_2020,Randeni2020,randeni_high-resolution_2021} a specialized implementation of the LBL solution.
 The ICNN was initially developed via numerous virtual experiments to ensure robust algorithms and interfaces between different hardware components. 
-The simulation capabilities are largely physics-driven with a modular system of systems approach\textemdash an environmental simulator with sub-components for the ocean, including Arctic ice drift and ocean acoustic propagation; a vehicle simulator with sub-components for vehicle dynamics and navigation; a topside hardware simulator and acoustic communications simulator, both with a software-only configuration and a hardware-in-the-loop version \citep{Schneider2018}.
+The simulation capabilities are largely physics-driven with a modular system of systems approach\textemdash an environmental simulator with sub-components for the ocean, including Arctic ice drift and ocean acoustic propagation; a vehicle simulator with sub-components for vehicle dynamics and navigation; a topside hardware simulator and acoustic communications simulator, both with a software-only configuration and a hardware-in-the-loop version \citep{schneider_netsim_2018}.
 The virtual environment similarly emulates the interfaces between the real components to test the entire software pipeline.
@@
 While each buoy only has one transmit depth, all buoys have both receive depths but the active receive layer is consistent across all buoys. \llabel{1.12}
-The design of the ICNN enables a self-adapting network to transmit and receive at the optimal depth to maintain contact with the AUV \citep{Schneider2020}.
-The buoys do not encompass the full horizontal range of the vehicle but are positioned to minimize overlap in trilateration for spherical positioning \citep{Deffenbaugh1996}.
+The design of the ICNN enables a self-adapting network to transmit and receive at the optimal depth to maintain contact with the AUV \citep{schneider_self-adapting_2020}.
+The buoys do not encompass the full horizontal range of the vehicle but are positioned to minimize overlap in trilateration for spherical positioning \citep{deffenbaugh1996posit}.
 
@@
 \item Four ice buoys, each outfitted with a WHOI Micro-Modem, receive messages from the AUV and send that information over freewave radio to a Topside computer
-\item The topside computer converts travel times into pseudorange estimates using a stochastic embedded prediction of the horizontal group velocity via BELLHOP ray tracing code \citep{Porter2011} using a sound speed profile provided by an updatable Virtual Ocean \citep{Schneider2018,bhatt_embedded_2022}
+\item The topside computer converts travel times into pseudorange estimates using a stochastic embedded prediction of the horizontal group velocity via BELLHOP ray tracing code \citep{Porter2011} using a sound speed profile provided by an updatable Virtual Ocean \citep{schneider_netsim_2018,bhatt_embedded_2022}
 \item The topside computer calculates a new position by trilaterating the range estimates
@@
 An important component to our navigation solution is an accurate estimation of a representative SSP for the ocean volume.
-Previous field experience, during the Ice Exercise 2016 (ICEX16), demonstrated the negative effects of the Beaufort Lens on tracking and communication \citep{Schmidt2016}.
+Previous field experience, during the Ice Exercise 2016 (ICEX16), demonstrated the negative effects of the Beaufort Lens on tracking and communication \citep{schmidt_acoustic_2016}.
 Fig. \ref{fig:sspExpectation} shows historical, modeled, and \textit{in situ} sound speed data for both ICEX16 and ICEX20.
@@
 In the field, the SSP information was shared with the vehicle via basis representation compression on a lightweight digital acoustic message \citep{bhatt_embedded_2022}. \llabel{2.7}
-All modeled data comes from HYCOM \cite{Chassignet2007}, which does not seem to capture the forcing mechanisms that cause the Beaufort Lens.
-For ICEX16, the data-driven profile was sourced from nearby Ice Tethered Profilers (ITP) after the field experiment \cite{Krishfield2008,Toole2011} and exhibits a fairly low lens; the historical profile is from the World Ocean Atlas.
+All modeled data comes from HYCOM \cite{chassignet_hycom_2007}, which does not seem to capture the forcing mechanisms that cause the Beaufort Lens.
+For ICEX16, the data-driven profile was sourced from nearby Ice Tethered Profilers (ITP) after the field experiment \cite{Krishfield2008,toole_ice-tethered_2011} and exhibits a fairly low lens; the historical profile is from the World Ocean Atlas.
 For ICEX20, the chosen weights (data-driven) profile derives from an estimate of initial CTD casts taken on site, showing an intense warm water intrusion; the baseline (historical) profile, showing moderate ducted conditions, comes from the average of March 2013 ITP data.
@@
 Whereas the MBC outputs a scalar, this method first outputs a vector of effective sound speeds based on the number of reflections.
-Then a single value is selected that best matches the recorded travel time, as the detected arrival is not always the first arrival or the direct path and could even be masked by noise or blocked temporarily \citep{deffenbaugh1996acoustic}.
+Then a single value is selected that best matches the recorded travel time, as the detected arrival is not always the first arrival or the direct path and could even be masked by noise or blocked temporarily \citep{deffenbaugh_acoustic_1996}.
 We manually cap the number of bounces at four because of the smaller operational scale and the attenuation accrued with many surface interactions.
@@
 One possibility is that, during a mission, ICNN-like LBL implementions use a comparison of the GNSS self-position and acoustic positioning to invert for the ocean volume, linking how vertical and horizontal sound speed structure impact transmission integrity.
-\llabel{1.20} A fast tomographic estimate \citep{Deffenbaugh1997,Elisseeff2002}, along with its uncertainty, can be continuously communicated to assets underway to maintain contact or enable adaptive sampling.
+\llabel{1.20} A fast tomographic estimate \citep{deffenbaugh_optimal_1997,Elisseeff2002}, along with its uncertainty, can be continuously communicated to assets underway to maintain contact or enable adaptive sampling.
 In this sense navigation and tomography converge on the same set of component technologies\textemdash position estimation, sound speed parameterization estimation, ray path identification, and vehicle path optimization.
@@
 
-The methods presented in this paper, including the software projects \citep{benjamin2010nested,schneider2010unified,schneider2015dynamic}, are open source and platform agnostic.
+\llabel{1.1}
+The methods presented in this paper, including the software projects \citep{,Benjamin2010,schneider2015dynamic,schneider_unified_2010}, are open source and platform agnostic.
 Large AUVs, often large enough to support long duration and/or deep sea missions, would benefit from including diurnal or tidal effects for ranging.

### 2022-02-21 rebuttal/rebuttal-letter/trunk-rebuttal.tex
added tags + more complete answers to rebuttal letter

@@
 We have refashioned the abstract and introduction to better elucidate the motivation for this work as a model-aided component for LBL navigation in total under-ice conditions, which could be further generalized for any acoustic environment.
-Based on several suggestions, we have restructured the manuscript and separated the real time navigation methodology (used the Minimum Bounce Criteria, MBC, but is impossible to compare given the lack of GNSS), the real-time positioning results (MBC), and the post-processing re-positioning improvement (Nearest Bounce Criteria).
+Based on several suggestions, we have restructured the manuscript and separated the real-time from the post-processed range estimation results.
+Our original revision aimed to contextualize the range estimation results as a proxy for navigation.
+Given several points from the reviewers and editor about the lack of this transferability, we have added AUV post-processed navigation results as well.
+The real-time results are also covered but have no GNSS ground truth to compare to.
 
@@
 The rebuttal letter is organized with sections delineated by the review document and email.
-We look forward to your comments and the opportunity to continue improving this manuscript for eventual publication in JASA.
+We look forward to your comments and the opportunity to continue improving this manuscript for potential publication in JASA.
 Thank you again for your time and consideration.
@@
 
-Dr. EeShan Bhatt \\
-\emph{on behalf of all authors}
+Dr. EeShan Bhatt \emph{on behalf of all authors}
 
@@
 What depths does the vehicle typically operate at and what is the maximum range from the beacons?
-The abstract stresses real-time results, but the real-time results are just 3 paragraphs, and the paper mostly concentrates on post-processing results.}
+The abstract stresses real-time results, but the real-time results are just 3 paragraphs, and the paper mostly concentrates on post-processing results.} \label{1dot1}
 
-	{\color{blue}
+	We have adjusted the claims made in the abstract (and throughout the paper) and hope we have presented data to better support them as well. 
+	While we initially aimed to contextualize the calculation of a handful of short fixed ranges between ice-tethered buoys as the proxy for navigation, we agree with the reviewer that this is a severe shortcoming.
+Accordingly, we added the same analysis for the AUV run in ICEX20.
 
-	We have both modified the data presented and tempered the claims in the abstract and throughout the manuscript.
-The reviewer brings up the important point of transferability: how applicable is the algorithm for normal vehicle operations given its performance on a handful of fixed ranges?
-
-	% We have tempered the claims in the abstract to reflect the limitations of the dataset collected and the transferability of the analysis in this paper.
-	% We have amended Figure \ref{v1:fig:vehicleRecovery} to include vehicle data and illustrate that this approach was successful in a real field environment.
-	% We have also included a discussion of how these results might compare for various depths and ranges during vehicle operations (see \hl{XXX}).
-
-	% With respect to the concentrating on real-time results versus post-processing results, we would point out that the AUV was deployed through a hydrohole from Topside and due to a disk error, stalled underneath the ice surface while transmitting its perceived location.
-	% The AUV was found within a meter from its reported location, which serves as strong (but qualitative) evidence of the navigation performance provided by the single group velocity calculation.
-	% Work currently review in the Journal of Field Robotics details the other components to vehicle navigation (DVL-aided INS and a vehicle flight model) along with the magnitude of position corrections.
-
-	% Given that our vehicle navigation data only has one GPS point to compare to, we use the beacon-to-beacon communications as a natural experiment to compare the effectiveness of the real-time NBC method.
-	% The real-time section is small because only one modem at a time ran the vehicle behavior to calculate these acoustic pseudoranges.
-	% A robust comparison is difficult given, as the reviewer points out, the handful of fixed ranges.
-
-	% We extended the dataset to all possible communication events through the same automated pipeline to decouple the effects of the sound speed environment from source and receiver position.
-This is in effect no different than running the computation during the field; the real-time NBC data is a subset of the post-processed NBC data.
-
-	% While post-processing, we began to see a pattern of range overestimation using the NBC method.
-Thus, we developed the Minimal Bounce Criteria instead and ran it through the same pipeline, subject to the same computational constraints.
-
-	% By comparing the errors between these two pseudorange estimation methods, we noticed a surprising underlying pattern compared to the ground truth GPS data.
-	% Further investigation suggests there is noise in the GPS measurement that is not incurred in the travel time measurement that informs the acoustic range estimation.
-
-	% The manuscript has been significantly restructured to separate analyses along this narrative.
-	}
+	We have also addressed valid questions about transferability to other vehicle operations in the discussion (line \ref{v1:1.1}).
 
@@
 Also, it would be good to state this hypothesis more clearly earlier in the manuscript.}
+
 	We agree that a few short fixed ranges do not validate this hypothesis.
-With the overarching motivation and design of LBL navigation, we have recast this hypothesis as an assumption.
-We would note, again, that the vehicle navigation performance would better determine a continuously changing group velocity, but this data would have no way of being verified.
+With the overarching motivation and design of LBL navigation, however, we have recast this hypothesis as an assumption (line \ref{v1:2.10}).
 
@@
 
-	We believe we have tightened up the introduction and conclusion sections by focusing on this work as a model-aided augmentation to an LBL system deployed in under-ice conditions.
+	We believe we have tightened up the introduction and conclusion sections by focusing on this work as a model-aided augmentation to an LBL system.
+	Then we specifically talk about AUV-LBL systems deployed in under-ice conditions.
 	This focus presents the narrative more linearly, instead of it feeling like it jumps between different types of deployments.
 
-	Present error metrics in reference to ...
+	The error metrics are presented as RMS m.
+Other positioning systems at similar scales are now covered in the introduction (line \ref{v1:1.3a}).
+We have also added an isovelocity case for fairer comparison to other smaller spatial scale AUV deployments and the proposed method in this manuscript.
 
-	The suggestion to discuss in terms of range, vehicle, and propagation conditions is helpful, and we have included that framing explicitly in the manuscript (LINE LINE).
-However, considering a union of all those operational variables, there are no comparable deployments.
-	Previous Arctic deployments are either in the Marginal Ice Zone at late summer, with little to no ice, or recorded the classical Arctic upward refracting profile, where isovelocity scaling could be good enough.
-	The few Arctic deployments that have seen a double ducted environment were on gliders, with little to no sea ice, and only one study navigated and re-navigated the gliders acoustically; other studies navigated them with dynamic flight models and GPS and re-navigated them acoustically.
+	Thank you for the suggestion to discuss in terms of range, vehicle, and propagation conditions.
+We have not done this explicitly but taken care to illustrate the differences, and what parts we are comparing.
 
-	Lastly, we have removed the line in question, and would point out it was originally written as ``navigation'', not ``positioning''.
+	Lastly, yes, we did mean real-time navigation, but have removed the line in question.
 
@@
 This method would presumably result in errors much higher than 11m….how might that scale?}
+
+	{\color{blue} COME BACK TO THIS.}
 
@@
 	Thank you, these are all excellent suggestions.
-``It is the variation of group speed from source to source in underwater navigation that is the fundamental challenge to implement a “GNSS-like” navigation system in the ocean with acoustics.''
+Instead of ``group velocity'' we use a combination of ``effective sound speed'', noted by $c$.
+The statement in question has been rewritten to talk about a single sound speed compensating for ray refraction and reflection, and is on line \ref{v1:1.5}.
 
@@
 line 186, 196, 241, 254 ) and more on, for example, why specific depths were chosen and how they provide better coverage due to the acoustic shadow zone (e.g., Line 155), or how the specific rays were eliminated by the filtering (Line 232).
-Lines 370-372 state “The increased error for these reciprocal transmission paths is most likely driven by the computational artifacts encountered when propagating through the steep sound speed gradients of the lens and through the shadow zone.“ The authors might consider providing more information on these artifacts.}
+Lines 370-372 state “The increased error for these reciprocal transmission paths is most likely driven by the computational artifacts encountered when propagating through the steep sound speed gradients of the lens and through the shadow zone.“ The authors might consider providing more information on these artifacts.}\label{1dot6}
 
-	{\color{blue}Thank you. The heart of this paper is injecting an acoustic model into a long baseline navigation.}
+	Agreed on all points.
+The frequencies are now included in the ICNN description (line \ref{v1:1.6a}).
+The discussion of which options were toggled on and off has been removed in favor of a in depth discussion of shadow zone, computational artifacts, and eigenray filtering in Section III.C, starting line \ref{v1:1.6b}.
 
@@
 I’d suggest condensing this section.}
+
+	This section has been significantly condensed and is now Section V.C, starting line \ref{v1:1.7}.
+The GPS uncertainty is now directly compared to the uncertainty from beacon-to-beacon re-positioning and AUV re-navigation.
 
@@
 
+	Thank you. This is now included in the ICEX20 experiment description, Section II.
+
 	\lreviewer{}{}{How long was the experiment from start to finish, and how much did the sound speed profile change over the duration of the experiment?}
+
+	This is a great question!
+The experiment was four days.
+The last few CTDs showed a significantly weaker duct than those at the beginning.
+The local sound speed maxima went from 1447 m/s at 40 m in depth to 1442 m/s at 60 m in depth.
+Related work on the environmental adaptivity has just been accepted in IEEE JOE.
 
@@
 
+	We assume negligible uncertainty in the position of the modems relative to the surface expression given that the ice drift rate was unusually low, roughly 0.7 cm/s.
+This is now included on line \ref{v1:1.10}.
+
 	\lreviewer{lines 40-42}{v1:1.11}{Note ranging via one way travel times that have been used prior to the advent of the micro-modem, for example RAFOS.}
+
+	Thank you. We have included this point and make the distinction that RAFOS floats are re-positioned.
 
@@
 
+	We have removed the jargon of the physical and virtual layers in favor of a more streamlined explanation. The transmit and receive layers are activated independently.
+
 	\lreviewer{line 205}{v1:1.13}{What do you mean that the coordinate system is range independent?}
-	We mean that the propagation is range independent, with one sound speed profile. It does actually take into account bathymetry.
+	We meant that the analysis of beacon-to-beacon connections is presented in r/z, not x/y/z.
+We found this framing awkward as it is more common to say the propagation has a depth dependent sound speed profile with range dependent bathymetry.
 
@@
 
+	In anticipation of ICEX20, all of the ITP data for the Beaufort Sea was assimilated.
+For operation reasons, we designed a system to transmit sound speed updates through digital acoustic message.
+This system relied on a simple basis function representation of the sound speed and its perturbations.
+Given the wide range of observed ducted conditions across all ITP data, several basis sets were created, corresponding to March of different years.
+During ICEX20, the data from Mar 2013 was found to be the closest match for observed conditions.
+The baseline is the mean sound speed from Mar 2013.
+The weighted is the basis set projection for the CTD taken during ICEX20.
+
 	\lreviewer{line 256}{v1:1.15}{How is the group velocity uncertainty calculated?}
 
+	We meant standard deviation, and have changed this accordingly. It is equation 2.
+
 	\lreviewer{line 231}{v1:1.16}{Quantify the travel-time proximity to recorded data that was used to filter the eigenrays.}
+
+	It was a one millisecond filter to cluster distinct arrivals.
 
@@
 
-	{\color{blue}different definitions of sparse. we meant that we're not computing other locations. they think sparse means the grid points are far apart.}
+	This was not clear. We meant sparse as in not full-field, which we agree is misleading. The grid is 121 points with 1 m spacing in range and 2 spacing m in depth.
 
@@
 
+	This has been rewritten as ``The algorithm generally overestimates pseudoranges because it resolves the effective sound speed for the most direct path.''
+
 	\lreviewer{line 364}{v1:1.19}{Could use a figure reference.}
 
+	We have added a figure reference.
+
 	\lreviewer{line 562}{v1:1.20}{I don’t agree that many approaches to underwater navigation combine it with acoustic tomography.}
-	{\color{blue}Benefit of concurrent navigation and tomography has been discussed, which is not necessarily that many approaches combine it.}
 
-	\lreviewer{Figure 1}{}{Figure 1 is a nice picture showing the vehicle, which motivates the work presented here, but was not used for collection of any of the data presented.
+	We agree, ``many'' misdirects the current state of the art. The benefit of concurrent navigation and tomography has been discussed, but it remains to be used as a standard technique.
+
+	\lreviewer{Figure 1}{v1:fig:vehicleRecovery}{Figure 1 is a nice picture showing the vehicle, which motivates the work presented here, but was not used for collection of any of the data presented.
 As it’s not pertinent to the data presented here, I’d suggest removing this figure or at least combining this figure with Figure 2.}
 
-	\lreviewer{Figure 3}{}{Figure 3 tallies the number of successful events, and I think the use of the directional arrows communicates the data well.
+	Thank you. We have kept this image but added the vehicle track, as it is necessary for the re-navigation analysis as mentioned in Comment \ref{1dot1}.
+
+	\lreviewer{Figure 3}{v1:fig:overview}{Figure 3 tallies the number of successful events, and I think the use of the directional arrows communicates the data well.
 How many events are successful in comparison with those that are considered unsuccessful and what generally causes an event to be deemed unsuccessful?}
 
-	{\color{blue} successful is if a single modem can decipher message. }
+	Thank you.
+We have added a successful decoding percentage, defined as the ratio of decoded over detected messages.
+We have also added the total number of transmissions, so one can see the ratio of detected over transmitted.
+An event is deemed unsuccessful given any error in the data frame, even if the error has nothing to do with the information necessary for positioning.
 
-	\lreviewer{Figure 5}{}{What is the point labeled u=751 m/s that appears to fall around 1424 m/s?}
+	\lreviewer{Figure 5}{v1:fig:gvel-post}{What is the point labeled u=751 m/s that appears to fall around 1424 m/s?}
+
+	That point is an effective sound speed prediction dominated by a bottom bounce.
+The figure has been changed to include all source depths.
+Keeping this point in would visually compress the other data.
+However, it is now discussed narratively, in line \ref{v1:1.23}.
 
 	\lreviewer{Table II}{}{Might this information be portrayed better by a box plot?}
+
+	This table described pseudorange statistics.
+Given that these are no longer serving as a proxy for navigation, it has been removed.
+The statistics for re-positioning and re-navigation are now included as histogram-box plots (Figs 11 and 12).
 
@@
 	\lreviewer{Fig. 2 and 3}{}{Fig 2 refers to central node position as Topside and Fig. 3 as Camp Seadragon. Suggest standardizing.}
+
 	Thank you. We have standardized this to Topside.
@@
 
-	\lreviewer{Table I}{v1:tab:rangeErrorInSitu}{Table I indicates direction errors are positive. Please note the sign convention in the caption.}
-	{\color{blue} caption is correct but text references an ``overstimation'' of range that is not obvious given absolute error shown.}
+	The source depths have been moved from the column titles to the individual sub-panels.
 
-	\lreviewer{Table II}{v1:tab:rangeErrorV1V2}{‘range estimation error metrics’ rather than ‘range estimation metrics’?}
-	Agreed. We have made the suggested change.
+	\lreviewer{Table I}{}{Table I indicates direction errors are positive. Please note the sign convention in the caption.}
+
+	We have removed the table but discuss this in the text at line \ref{v1:1.28}, and have corrected the convention.
+
+	\lreviewer{Table II}{}{‘range estimation error metrics’ rather than ‘range estimation metrics’?}
+	Agreed. While we have removed the table, we have adopted this language.
 
 	\lreviewer{Table II and III}{}{Suggest removing ‘algorithm’ label.  Also substitute ‘MBC’ and ‘NBC’ for minimal and nearest to match the text.}
-	Agreed. We have made the suggested changes.
+	Agreed. While we have removed the table in favor of histogram/box plot, we have adopted this language.
 
 	\lreviewer{}{}{There are several figures and tables which are not referenced properly (Line 305, 391, 411)}
-	Thank you. This has been corrected.
+	Thank you. This has been corrected and we apologize for the typographical mistakes.
 
@@
 
-	The Beaufort Lens has been used in the literature (Litvak 2015, Schmidt and Schneider 2016) and in JASA/POMA specifically (Poulsen et al 2016, Chen et al 2019).
+	We have removed the ``Beaufort Lens'' language from the abstract as it may not be self explanatory.
+However, the Beaufort Lens has been used in the literature (Litvak 2015, Schmidt and Schneider 2016) and in JASA/POMA specifically (Poulsen et al 2016, Chen et al 2019).
 
@@
 We prefer this distinction as the lens exists in the oceanographic space but the ducts are acoustic features dependent on source depth and frequency.
-With this jargon, the lens is susceptible to aberrations in the same way that the duct is susceptible to leaking.
+With this jargon, the lens is susceptible to aberrations from eddies or filaments in the same way that the duct is susceptible to leaking.
 
-	For clarity, we explictly define Beaufort Lens in the manuscript as explained HERE LINK.
+	For clarity, we explictly define the Beaufort Lens in the manuscript as discussed here at line \ref{v1:2.1a}.
 
@@
 
-	\lreviewer{line 12}{v1:2.3}{'rival GPS' is a fairly bold statement when GPS is used as ground truth.}
+	\lreviewer{line 12}{v1:2.3}{'rival GPS' is a fairly bold statement when GPS is used as ground truth.} \label{2dot3}
 
@@
 Given more beacons, there is potential to reduce uncertainty in the same way that differential GPS works.
+While this refrain is used multiple times in the revised manuscript, please see Section V in particular.
 
@@
 
-	{\color{blue} Yes. BELLHOP finds all possible eigenrays given the prescribed launch angles whereas only a few or less (sometimes none) are valid for the travel time of interest}.
+	Yes, thank you. BELLHOP finds all possible eigenrays given the prescribed launch angles whereas only a few or less (sometimes none) are valid for the travel time of interest.
 
@@
 
-	We have adopted this suggestion and use navigation'' or communication system instead.
+	We have adopted this suggestion and use navigation or communication system instead.
 
@@
 
-	{\color{blue} Perhaps both. We have changed this accordingly}
+	We have removed this sentence in favor of discussing ray refraction and reflection.
 
 	\lreviewer{}{}{for both criteria: a graphic that shows the method could be very useful.}
+
+	This is a good point.
+We hope that the separation of MBC and NBC, as well as the extended discussion in Section III-C, better illustrate both criteria.
+A visual on its own would be quite similar to the eigenray plots added.
+However, we are happy to include a graphic showing the method if reviewers find the narrative explanation lacking.
 
@@
 
-	{\color{blue} We mean prominence, often used in peak finding, which would indicate both the considerable sound speed differential and the nearly horizontal slope to the local maxima.}
+	We mean prominence, often used in peak finding, which would indicate both the considerable sound speed differential and the nearly horizontal slope to the local maxima.
+This is now explained more clearly and introduced early at line \ref{v1:2.5}.
 
@@
 
-	\lreviewer{Fig 10}{v1:fig:gvelMore}{Would be good to have some conclusions in addition to the description.}
+	\lreviewer{Fig 10}{v1:fig:gvel-post}{Would be good to have some conclusions in addition to the description.}
+
+	This figure has been condensed into the main body of the text alongside conclusions around the effective sound speed estimation.
 
@@
 
-	Thank you.
-The key point of this paper is a model-aided component to the LBL paradigm that we demonstrated in an Arctic environment.
+	Thank you. The key point of this paper is a model-aided component to the LBL paradigm that we demonstrated in the Beaufort Sea. 
 	Because both reviewers and the editor felt the key points were comparing the two different approaches to determine an effective sound speed for converting travel times to ranges, we have significantly restructured the paper to make the narrative clearer.
@@
 	{I agree with Reviewer 1 that the manuscript needs much more information on the acoustic parameters (frequencies, pulse lengths/modulation, ranges, expected travel-time measurement accuracy, etc.).
+
 	In addition, the ray geometries are quite different for cases in which the source and receiver are both in the surface duct (i.e., 20 and 30 m), both in the Beaufort Duct (for the sound-speed profiles for which it is present) (i.e., 90 m) and one of the source or receiver is in the surface duct and the other is in the Beaufort Duct.
@@
 
-	We have now included the information on acoustic parameters in the experiment description, see section REFREF.
-We apologize that this information was ommitted as the manuscript was adapted from the thesis.
+	We have now included the information on acoustic parameters in the experiment description. We apologize that this information was ommitted as the manuscript was adapted from a thesis.
 
-	Similarly, we have addressed the editor's point about the ray geometries in the discussion section, REFREF, but have opted to keep those figures in the supplementary material.
+	Similarly, we have addressed the editor's point about the ray geometries in Section III-C (see comments, but have opted to keep those figures in the supplementary material.
+
+	Comment \ref{1dot6} is relevant to both of these points.
 
@@
 
-	{\color{blue} Thank you for catching this. We have corrected it and made sure all figurs are referenced in the text directly.}
+	Thank you for catching this. We have corrected it and made sure all figures are referenced in the text directly.
 
@@
 It seems that the authors really mean GPS accuracy.} \label{3dot7}
-	{\color{blue} We have talked with colleagues and think ``GPS noise'' is the best term.
-But given that all reviewers and the editor feel this section is out of place, we have significantly reduced it...
-OR REMOVE IT?}
+
+	We have talked with colleagues and think ``GPS noise'' is the best term.
+But given that all reviewers and the editor feel this section is out of place, we have significantly reduced it.
+Please see comment \ref{2dot3}.
 
@@
 GPS (really GNSS) applications come in different flavors, including stand-alone single frequency, stand-alone dual frequency, differential, real-time kinematic, etc., with varying accuracies.}
-	When evaluating our methodology, we have removed qualitative language like ``GPS-like'' or ``rivaling GPS'' in favor of quantitative descriptions of ranging error.
-{\color{blue}achieves ``x meter'' level accuracy compared to ``y meter'' level accuracy reported by the manufacturer}
+
+	When evaluating our methodology and results, we have removed qualitative language like ``GPS-like'' or ``rivaling GPS'' in favor of quantitative descriptions of ranging error, i.e., achieving ``x`` accuracy compared to ''y`` from the GPS unit''.
 
@@
 
-	We have removed this generalization.
+	Thank you. We have removed this generalization.
 
@@
 
-	{\color{blue}Thank you for pointing this out. We have double checked all references.}
+	Thank you for pointing this out. We have double checked all references.
 
@@
 Please indicate in your report (in detailed comments, below) any points which are objectionable or which need attention.}
-	{Reviewer 1: It presents original work.  There are aspects which need clarification and are outlined in the attached report.
+	\textbf{Reviewer 1:} It presents original work.  There are aspects which need clarification and are outlined in the attached report.
 
-	Reviewer 2: The work is interesting and describes a data set collected from the Arctic, which is of high value. However, some of the writing is a bit awkward and occasionally vague or imprecise.}
+	\textbf{Reviewer 2:} The work is interesting and describes a data set collected from the Arctic, which is of high value.
+However, some of the writing is a bit awkward and occasionally vague or imprecise.
 
@@
 Does the content of the manuscript, including terminology and the references cited, meet this criterion?}
-	{Reviewer 1: I wouldn’t say JASA is the most appropriate journal.
+
+	{\textbf{Reviewer 1:} I wouldn’t say JASA is the most appropriate journal.
 As written, this work might be better suited to an ocean engineering journal.
@@
 
-	Reviewer 2: Yes.}
+	\textbf{Reviewer 2:} Yes.}
 
@@
 Please indicate needed changes in your report.}
-	{Reviewer 1: The English is satisfactory.
+	{\textbf{Reviewer 1:} The English is satisfactory.
 This paper could benefit from some reorganization and careful editing.
@@
 
-	Reviewer 2: Overall the language is fine, details below.}
+	\textbf{Reviewer 2:} Overall the language is fine, details below.}
 
@@
 If any of the figures are in color, is the color essential for conveying the scientific point?}
-	{Reviewer 1: I would suggest combining Figures 1 and 2 or eliminating Figure 1.
+	{\textbf{Reviewer 1:} I would suggest combining Figures 1 and 2 or eliminating Figure 1.
 Some suggestions for improving figure and table titles and captions are in the attached report.
@@
 
-	Reviewer 2: In general they are fine except as noted.}
+	\textbf{Reviewer 2:} In general they are fine except as noted.}
 
 	\reviewer{}{If Supplementary material was submitted, is it relevant to the manuscript and should it be deposited in the Supplemental Depository for reference to the manuscript?}
-	{Reviewer 1: The supplemental figures were relevant to the manuscript.
+	{\textbf{Reviewer 1:} The supplemental figures were relevant to the manuscript.
 
-	Reviewer 2: NA}
+	\textbf{Reviewer 2:} NA}
 
@@
 	}
-	{Reviewer 1: I'd suggest condensing the section on GPS drift (Section III E).
+	{\textbf{Reviewer 1}: I'd suggest condensing the section on GPS drift (Section III E).
 
-	Reviewer 2: The GPS section seems out of place. Can it be shortened to a few sentences and references?}
+	\textbf{Reviewer 2}: The GPS section seems out of place. Can it be shortened to a few sentences and references?}
 
@@
 IMPORTANT: The lead paragraph should advertise the main points of the article and must describe in terms accessible to the general reader the context and significance or the research problem studied and the importance of the results.}
-	{Reviewer 1: The title and abstract are a bit misleading as there is no navigation demonstrated in the paper and the ranging results were from fixed buoys rather than a vehicle.
+	{\textbf{Reviewer 1}: The title and abstract are a bit misleading as there is no navigation demonstrated in the paper and the ranging results were from fixed buoys rather than a vehicle.
 The abstract, and the final sentence in particular, should be tempered to reflect the scope of the results.
@@
 
-	Reviewer 2: Title, yes.
+	\textbf{Reviewer 2}: Title, yes.
 	The abstract does need some work. Statements like this are risky, suggest summarizing technical contributions.
@@
 
-	Thank you.
-We have opted to keep the title the same, as we believe the results are a strong proxy for under-ice acoustic navigation, which we demonstrated during ICEX20 but there are inherently no data products to validate this waypoint by waypoint.
+	Thank you. We have opted to keep the title the same, as we have now integrated results from the AUV instead of allowing beacon results to serve as a proxy for navigation writ large.
 

### 2022-02-21 manuscript/main.tex
added tags + more complete answers to rebuttal letter

@@
 While RAFOS floats championed one way ranging for re-positioning \citep{Rossby1986,duda_evaluation_2006}, the ability to do so for navigation was facilitated by \llabel{1.11} the advent of the WHOI Micro-Modem \citep{Singh2006} and synchronized clocks \citep{rypkema_one-way_2017}.
+\llabel{1.3a}
 AUV navigation efforts have achieved root mean square (RMS) localization error on the order of tens of meters relative to GNSS surface position over less than ten kilometers in shallow \citep{Eustice2007,Claus2018,kepper_mems_2017} and deep water \citep{Kunz2008,jakuba_long-baseline_2008,Webster2009}.
@@
 
+\llabel{2.1a}
 Given that a lens introduces significant ray refraction, the Beaufort Lens is a shorthand for the spatio-temporal variability of the local temperature and sound speed maxima generally around 50 to 60 m in depth.
+\llabel{2.4}
 A neutrally buoyant layer of warm Pacific Summer Water creates a unique double ducted environment \textemdash the upper duct degrades signal coherence due to intensified ice interaction and the lower duct effectively traps sound for long range propagation \citep{Poulsen2016}.
@@
 The AUV and each ice buoy are outfitted with a WHOI Micro-Modem \citep{Singh2006}, with a four-element receiver array, a single transmitter, and one-tenth of a millisecond resolution.
-Acoustic messages were sent with a 10 kHz carrier frequency, 5 kHz bandwidth, and phase-shift keying (PSK)modulation on a time-division multiple access schedule with a thirty-second cycle, giving room for two-way communication throughout the mission volume.
+\llabel{1.6a} Acoustic messages were sent with a 10 kHz carrier frequency, 5 kHz bandwidth, and phase-shift keying (PSK) modulation on a time-division multiple access schedule with a thirty-second cycle, giving room for two-way communication throughout the mission volume.
 The receive and transmit elements were split between shallow and deeper depths\textemdash 30 and 90 m\textemdash to provide better coverage across the shadow zone.
@@
 Fig. \ref{fig:overview} shows successful events sorted by source depth.
+\llabel{1.10}
 In this analysis, we assume there is insignificant displacement between the GNSS puck surface expression and subsurface modem; this is supported by unusually low observed ice drift rates, just 0.7 cm/s on average throughout the mission.
@@
 
-The fundamental challenge to implement GNSS-like navigation, especially in an acoustically complex propagation environment, is characterizing a single sound speed to compensate for the effects of ray refraction and reflection.
+\llabel{1.5} The fundamental challenge to implement GNSS-like navigation, especially in an acoustically complex propagation environment, is characterizing a single sound speed to compensate for the effects of ray refraction and reflection.
 The use of the acoustic modem network for tracking relies on the accurate estimates of travel times between the submerged platform and LBL beacons, supported by clock synchronization and a pre-determined scheduling of acoustic events.
@@
 A promising sign that the MBC method adapts sound speed somewhat intelligently is the lack of error growth as travel time increases.
+\llabel{1.28}
 The baseline SSP (n=243 events) has an absolute pseudorange error of 11.38 $\pm$ 4.23 m; the weighted SSP (n=568), 11.36 $\pm$ 8.12 m.
@@
 
-\subsection{Eigenray identification for beacon-to-beacon events} \label{sec:eigenrays}
+\subsection{Eigenray identification for beacon-to-beacon events} \label{sec:eigenrays} \llabel{1.6b}
 
@@
 The goal of the MBC algorithm was to provide a reliable, physically intuitive interpretation of the acoustic propagation without taking on the additional burden of regularly identifying specific paths that may connect any given source-receiver pair in the network.
+While it was unlikely to resolve multipath arrivals that triggered successful modem detection, an isovelocity approach would have provided no adaptivity against source and receiver depth differences.
 \llabel{2.14}
-While it was unlikely to resolve multipath arrivals that triggered successful modem detection, an isovelocity approach would have provided no adaptivity against source and receiver depth differences.
 Its performance was adequate for vehicle navigation and would have likely sufficed if it were not for the prominence of the duct observed relative that of other model and data products.
@@
 In fact, the higher the multipath classification, the more accurate the sound speed prediction is, likely driven by a tighter or even sparser bundle of rays.
-\llabel{1.19} Discontinuities in multipath classification provide initial evidence for its importance to a smoothly varying group velocity, as shown in the cluster of 30 to 30 m transmissions, where HYCOM jumps from one to two classified bounces amidst the baseline SSP and weighted SSP smoothly increasing while consistently seeing two and three classified bounces, respectively.
+\llabel{1.19} Discontinuities in multipath classification provide initial evidence for its importance to a smoothly varying group velocity, as shown in the cluster of 30 to 30 m transmissions in \ref{fig:gvel-post}, where HYCOM jumps from one to two classified bounces amidst the baseline SSP and weighted SSP smoothly increasing while consistently seeing two and three classified bounces, respectively.
 Of course, the prediction deteriorates with cross-layer transmissions across the duct, but not to the same degree at which eigenrays could not be found for the weighted SSP in section \ref{sec:eigenrays}.
@@
 The one from \citet{Graupe2019}, 1450 m/s, would have had a systemic overestimation.
-Given that implied sound speeds just for beacon-to-beacon events span 1420 to 1445 m/s, it is safe to say that a nominal sound speed would sacrifice pseudorange accuracy somewhere, and that an adaptive approach is necessary even for short range operations in the Beaufort Lens.
+In addition, over the course of the four day experiment, the local maxima of the Beaufort Lens changed from roughly 1447 m/s at 40 m to 1442 m/s at 60 m.
+Given that implied sound speeds just for beacon-to-beacon events span 1420 to 1445 m/s, it is safe to say that a nominal sound speed would sacrifice pseudorange accuracy somewhere, and that an adaptive approach is necessary even for short and/or small scale operations in the Beaufort Lens.
 
@@
 
-There is one example that helpfully illustrates the improvement brought upon by bounce classification.
+\llabel{1.23} There is one example that helpfully illustrates the improvement brought upon by bounce classification.
 For transmissions between North and South at 30 m, the OWTT spread is 2.1958 to 2.1963 s; the GNSS-tracked distance is 3138.54 to 3140.87 m; and the implied effective sound speed is 1429.3 to 1430.1 m/s.
@@
 \subsection{Investigating potential GNSS noise}
+\llabel{1.7}
 

### 2022-02-21 manuscript/main.tex
made figs bigger, easy to see details

@@
 	\centering
-	\includegraphics[width=\reprintcolumnwidth]{figs/Fig2.pdf}
+	\includegraphics[width=0.8\columnwidth]{figs/Fig2.pdf}
 	\caption{A schematic overview of the Integrated Communication and Navigation Network (ICNN), which provides joint data-transfer and tracking between AUV and a human decision maker at Topside.}
@@
   \centering
-  \includegraphics[width=\textwidth]{figs/modem-chart.pdf}
+  \includegraphics[width=\columnwidth]{figs/modem-chart.pdf}
   \caption{An overview of the modem experiment by source and receiver depth and position.
@@
   \centering
-  \includegraphics[width=\reprintcolumnwidth]{figs/range-error-insitu.pdf}
+  \includegraphics[width=\columnwidth]{figs/range-error-insitu.pdf}
   \caption{The real-time range error by source (rows: 20, 30, and 90 m) and receiver (columns: 30 and 90 m) pairings for both sound speed estimates used during ICEX20.
@@
   \centering
-  \includegraphics[width=\reprintcolumnwidth]{figs/raytrace-3env-zs-20.pdf}
+  \includegraphics[width=\columnwidth]{figs/raytrace-3env-zs-20.pdf}
   \caption{Eigenrays for beacon-to-beacon events for each sound speed with a nominal source depth of 20 m over a total ray fan in gray.
@@
   \centering
-  \includegraphics[width=\reprintcolumnwidth]{figs/raytrace-3env-zs-30.pdf}
+  \includegraphics[width=\columnwidth]{figs/raytrace-3env-zs-30.pdf}
   \caption{Eigenrays for beacon-to-beacon events for each sound speed with a nominal source depth of 30 m over a total ray fan in gray.
@@
   \centering
-  \includegraphics[width=\reprintcolumnwidth]{figs/raytrace-3env-zs-90.pdf}
+  \includegraphics[width=\columnwidth]{figs/raytrace-3env-zs-90.pdf}
   \caption{Eigenrays for beacon-to-beacon events for each sound speed with a nominal source depth of 90 m over a total ray fan in gray.
@@
 \begin{figure}[!ht]
-\includegraphics[width=\textwidth]{figs/range-error-allMethods.pdf}
+\includegraphics[width=\columnwidth]{figs/range-error-allMethods.pdf}
 \caption{The post-processed pseudorange error; the rows share receiver depth; the columns share source depth.

### 2022-02-21 rebuttal/rebuttal-letter/trunk-rebuttal.tex
oscar feedback pt 1

@@
 \documentclass{article}
-	\def\papertitle{Under-ice acoustic navigation using real time model-aided range estimation}
+	\def\papertitle{Under-ice acoustic navigation using real-time model-aided range estimation}
 	\def\authors{E. Bhatt, O. V\'{i}quez, H. Schmidt}
@@
 
-Navigation is real time and necessarily depends on positioning, and positioning depends on timing.
+Navigation is real-time and necessarily depends on positioning, and positioning depends on timing.
 \end{quote}
 
-Given the tension of evaluating our travel time to range conversion on events between LBL beacons in real time and post-processing, we further delineate that, in this work, positioning is in real time and re-positioning is post-processed.
-We have also added re-positioning trilateration results, processed with only the available field data and algorithms that could be performed in real time, to emulate navigation performance.
+Trilateration of beacon events was done in real-time, on the beacons themselves.
+In this work, we further delineate that positioning is in real-time and re-positioning is post-processed.
+We have also added re-positioning and re-navigation trilateration results to emulate our system performance given the navigation results have no ground truth GNSS to compare to.
 
@@
 However, we note that the system was successfully deployed on an underwater vehicle during ICEX20.
-The results from this work are specifically limited to the methodology behind calculating a single group velocity, which we attempt to quantify through beacon-to-beacon communications with known GNSS timestamps.
+The results from this work are specifically limited to the methodology behind calculating a single group velocity, which we attempt to validate through beacon-to-beacon communications with known GNSS timestamps.
 
@@
 
-	We have adjusted the claims made in the abstract (and throughout the paper) and hope we have presented data to better support them as well. 
-	While we initially aimed to contextualize the calculation of a handful of short fixed ranges between ice-tethered buoys as the proxy for navigation, we agree with the reviewer that this is a severe shortcoming.
+	We have adjusted the claims made in the abstract (and throughout the paper) to better emphasize what was deployed in the field, as part of normal vehicle operations, and how the proposed improvements compare to the fielded approach.
+We have presented data that we feel better support the claims about single meter accuracy as well.
+	While we initially aimed to contextualize the calculation of a handful of short fixed ranges between ice-tethered buoys as the proxy for navigation, we agree with the reviewer that this representation of our work is incomplete in portraying its portability to vehicle operations.
 Accordingly, we added the same analysis for the AUV run in ICEX20.
@@
 
-	{\color{blue} COME BACK TO THIS.}
+	Within the data collected during ICEX20, we actually do see instances of multiple arrivals detected at a single receiver, spread over multiple seconds\textemdash this can be exemplified by arrivals consistent with bottom-bounce paths.
+
+	The point of the Nearest Bounce Criteria, the improved approach proposed in this paper as part of the post-processing analysis, is precisely to expose how tracking the results of the acoustic model by modal variations can impact the ranging and positioning results; this is done using field data as the foundation.
+In our case, at short ranges under the Arctic ice, we have opted to track boundary interactions as the basis for modal variation; however, a different arrival classification basis may be needed for longer ranges in the SOFAR channel, for example.
 
@@
 	Thank you, these are all excellent suggestions.
-Instead of ``group velocity'' we use a combination of ``effective sound speed'', noted by $c$.
+Instead of ``group velocity'' we use ``effective sound speed'', noted by $c$.
 The statement in question has been rewritten to talk about a single sound speed compensating for ray refraction and reflection, and is on line \ref{v1:1.5}.
@@
 	We meant that the analysis of beacon-to-beacon connections is presented in r/z, not x/y/z.
-We found this framing awkward as it is more common to say the propagation has a depth dependent sound speed profile with range dependent bathymetry.
+We found this framing awkward as it is more common to say the propagation has a depth dependent sound speed profile.
+Range-dependent models may assume no change in the sound speed profile, but account for a range dependent bathymetry; or they may account for changes in both bathymetry and SSP over range.
+This work includes a range dependent bathymetry.
 
@@
 The baseline is the mean sound speed from Mar 2013.
-The weighted is the basis set projection for the CTD taken during ICEX20.
+The weighted SSP is the basis set projection for the CTD taken during ICEX20.
 
@@
 We have also added the total number of transmissions, so one can see the ratio of detected over transmitted.
-An event is deemed unsuccessful given any error in the data frame, even if the error has nothing to do with the information necessary for positioning.
+An event is deemed unsuccessful given any error in the transmitted headers or any data frame, even if the error has nothing to do with the information necessary for positioning.
 
@@
 
-	Thank you. We have significantly reorganized the paper.
+	Thank you. We have significantly reorganized the paper to improve clarity as suggested.
 
@@
 
-	We have significantly condensed the background for this section and narrowed the discussion for this section about the pattern of psuedorange error and what it might mean for the precision of the GPS puck we used at high latitudes.
+	We have significantly condensed the background for this section.
+In addition, we narrowed the discussion to the how trilateration error compares to that of GNSS at high latitudes and our particular GNSS unit.
 

### 2022-02-21 pipeline/3-analyze-gvel/plot_gvel_by_txrx_depth.m

@@ -1,5 +1,5 @@
 %% prep workspace
-clear; clc; close all;
+clear; clc; close all; addpath('../../src/');
 
 % load modem marker information
 load p_modemMarkerDetails
@@ -107,7 +107,7 @@ end
 %% add legend
 
 % add legend
-nexttile(4);
+nexttile(1);
 
 % add legend 1 -- color
 hold on

### 2022-02-21 manuscript/main.tex
oscar feedback pt 3(?) and dropping c from NBC/MBC

@@
 However, navigation uncertainty is a major challenge in considering AUVs as standard tools for oceanographic research.
-While land and air-based robots utilize information from Global Navigation Satellite Systems (GNSS) to achieve stunning location accuracy and precision throughout the duration of their missions, AUVs cannot access GNSS fixes while underwater.
+While land and air-based robots utilize information from Global Navigation Satellite Systems (GNSS) to achieve single-meter location accuracy and precision throughout the duration of their missions, AUVs cannot access GNSS fixes while underwater.
 Therefore, underwater vehicles have relied on any combination of dead reckoning, hydrodynamic models, inertial navigation systems, doppler velocity logs, and acoustic baseline positioning systems for navigation \citep{paull_auv_2014}.
@@
 The state-of-the-art for LBL outsources depth to a pressure sensor and solves the two-dimensional localization problem with an isovelocity, linear scaling between one way travel time (OWTT) and range \citep{Eustice2006,Eustice2007,webster_advances_2012,Webster2009}.
-This assumption is valid for short scale operations but oversimplifies propagation for larger and/or complex acoustic environments.
+This assumption is valid for short scale operations but oversimplifies propagation at larger scales or complex sound speed structure.
 To achieve single meter, GNSS-like performance in a GNSS-denied environment, we demonstrate an embedded ray-based data processing algorithm to convert recorded OWTTs into pseudorange estimates.
@@
 
-Localization algorithms that do consider environmental or acoustic uncertainty tend to focus on longer and larger experiments, where spatio-temporal variability cannot be ignored.
+Localization algorithms that do consider environmental or acoustic uncertainty tend to focus on longer duration and larger range experiments, where spatio-temporal variability cannot be ignored.
 These methods have also been reserved for post-processing as they can be labor intensive, computationally heavy, and/or require additional information like contemporaneous data.
@@
 Transmissions in the upper duct, between the surface ice and the lens, experience minimal attenuation but degrade in signal coherence with repeated reflections under the ice.
-In lower duct, between the lens and its conjugate depth in the Atlantic water, roughly 200 m, sound above 350 Hz is trapped near losslessly for long range propagation \citep{Poulsen2016}.
+In the lower duct, between the lens and its conjugate depth in the Atlantic water (roughly 200 m), sound above 350 Hz is trapped near losslessly for long range propagation \citep{Poulsen2016}.
 
@@
 Thorough reviews of uncrewed vehicle operations in polar environments can be found in \citet{norgren2014unmanned} and \citet{Barker2020}; there is no comparable work in the Arctic for a short range AUV deployment in the Beaufort Lens.
-Seminal \citep{brooke1981arcs,jackson1983autonomous,light1989autonomous,bellingham1995auv,hayes2002determining} and more recent AUV deployments \citep{,jakuba_long-baseline_2008,Kunz2008,Kukulya2010,Plueddemann2012,timmermans2013scales,fossum2021adaptive} witnessed the classical upward refracting sound speed profile that is amenable to an isovelocity assumption.
+Seminal Arctic AUV deployments \citep{brooke1981arcs,jackson1983autonomous,light1989autonomous,bellingham1995auv,hayes2002determining} and more recent ones \citep{,jakuba_long-baseline_2008,Kunz2008,Kukulya2010,Plueddemann2012,timmermans2013scales,fossum2021adaptive} witnessed the classical upward refracting sound speed profile that is amenable to an isovelocity assumption.
 
 Of note, despite different platforms and scales, are recent glider deployments in the Canada Basin.
-In 2014, in partially ice-covered conditions, a long range LBL system with WHOI Micro-Modems at 100 m exploited the lower duct for long range communication with two gliders \citep{Freitag2016,Webster2015}.
+In 2014, in partially ice-covered conditions, a long range LBL system with WHOI Micro-Modems at 100 m depth exploited the lower duct for long range communication with two gliders \citep{Freitag2016,Webster2015}.
 The sound speed value measured at the time of reception was used to estimate pseudorange in post-processing.
@@
 Both cases exploit the lower duct for high fidelity communication at long ranges.
-Unintuitively, the smaller nature of our deployment during ICEX20 is not a simplifying factor.
-For source depths typical to vehicle operations, 30 to 200 m, a shadow zone spans from 2 to 6 kilometers in range \citep{schmidt_acoustic_2016}.
+Unintuitively, the smaller scale nature of our deployment during ICEX20 is not a simplifying factor.
+For source depths typical to vehicle operations, 30 to 200 m, the Beaufort Lens introduces a shadow zone that spans from 2 to 6 kilometers in range \citep{schmidt_acoustic_2016}.
 
-Compared to the previous small scale navigation efforts, the approach in this paper integrates real-time model-aided data processing to estimate a representative sound speed along a path from source to receiver, leveraging climatology, \textit{in situ} data, and fast acoustic modeling.
+Compared to previous small scale navigation efforts, the approach in this paper integrates real-time model-aided data processing to estimate a representative sound speed along a path from source to receiver, leveraging climatology, \textit{in situ} data, and fast acoustic modeling.
 The paper is organized as follows.
@@
 
-The results from this paper derive from data taken while deploying the AUV \emph{Macrura}, a custom Bluefin-21, during the Ice Exercise 2020 (ICEX20), in the Beaufort Sea, from March 8th to 11th.
+The results from this paper derive from data collected while deploying the AUV \emph{Macrura}, a custom Bluefin-21, during the Ice Exercise 2020 (ICEX20).
+The experiment was conducted in the Beaufort Sea, from March 8th to 11th, at roughly 71.2\textdegree N.
 The AUV deployment was supported by the Integrated Communication and Navigation Network (ICNN) \citep{schneider_self-adapting_2020,Randeni2020,randeni_high-resolution_2021} a specialized implementation of the LBL solution.
@@
 Each ice buoy is outfitted with a Garmin GPS 18x, with a pulse-per-second rising edge aligned to 1 microsecond and a spec sheet accuracy of 3 m, 95\% of the time.
-The AUV and each ice buoy are outfitted with a WHOI Micro-Modem \citep{Singh2006}, with a four-element receiver array, a single transmitter, and one-tenth of a millisecond resolution.
+They are also each equipped with a WHOI Micro-Modem \citep{Singh2006}, with a four-element receiver array, a single transmitter, and one-tenth of a millisecond resolution.
 \llabel{1.6a} Acoustic messages were sent with a 10 kHz carrier frequency, 5 kHz bandwidth, and phase-shift keying (PSK) modulation on a time-division multiple access schedule with a thirty-second cycle, giving room for two-way communication throughout the mission volume.
+Thus the ICNN is dependent on the successful decoding of acoustic transmissions.
 The receive and transmit elements were split between shallow and deeper depths\textemdash 30 and 90 m\textemdash to provide better coverage across the shadow zone.
@@
 A random disk error stalled \emph{Macrura} underneath the ice but did not prevent it from transmitting its location.
-Due to an incoming storm, a team placed a physical marker on the ice at the location.
-Three days later, \emph{Macrura} was found within a meter of the marker.
-We view the emergency recovery as qualitative proof of the robustness of this navigation approach.
-Nonetheless, this paper specifically addresses the third and fourth steps\textemdash the conversion of travel times into pseudoranges and its effect on trilateration.
+Due to an incoming storm, a team drilled an exploratory hole at \emph{Macrura's} self-reported location, and were able to drill a separate hole 1 m away to tie the vehicle off to a physical marker on the ice.
+Three days later, \emph{Macrura} was recovered\textemdash the ice camp had moved over 19 km.
+The AUV's relative position to camp had also changed, from roughly 45\textdegree at 1000 m to 90\textdegree at 1100 m.
+Drama aside, we view the emergency recovery as qualitative proof of the robustness of this navigation approach.
+Nonetheless, this paper specifically addresses the third and fourth steps\textemdash the conversion of travel times into pseudoranges and its quantative effect on trilateration.
 By focusing on pseudorange estimates between GPS-tracked beacons, and re-running the trilateration pipeline, the results are decoupled from all other mechanisms in the ICNN.
@@
 All modeled data comes from HYCOM \cite{chassignet_hycom_2007}, which does not seem to capture the forcing mechanisms that cause the Beaufort Lens.
-For ICEX16, the data-driven profile was sourced from nearby Ice Tethered Profilers (ITP) after the field experiment \cite{Krishfield2008,toole_ice-tethered_2011} and exhibits a fairly low lens; the historical profile is from the World Ocean Atlas.
-For ICEX20, the chosen weights (data-driven) profile derives from an estimate of initial CTD casts taken on site, showing an intense warm water intrusion; the baseline (historical) profile, showing moderate ducted conditions, comes from the average of March 2013 ITP data.
+For ICEX16, the data-driven profile was sourced from nearby Ice Tethered Profilers (ITP) after the field experiment \cite{Krishfield2008,toole_ice-tethered_2011} and exhibits a fairly deep lens; the historical profile is from the World Ocean Atlas.
+For ICEX20, the chosen weights (data-driven) profile derives from initial CTD casts taken on site, showing an intense warm water intrusion; the baseline (historical) profile, showing moderate ducted conditions, comes from the average of March 2013 ITP data.
 This month best matched sea ice and sound speed conditions at the beginning of ICEX20 \citep{bhatt_embedded_2022}.
@@
 For post-processing comparison, we introduce both the HYCOM profile and an isovelocity case, 1441.8 $\pm$ 3.7 m/s, as the mean and standard deviation of the observed sound speed profile over the first 200 m.
-This is a contrived value taken in the style of \citet{Graupe2019} for the sake of comparison; the default value in the LAMSS simulator, which was not environmentally informed nor used during ICEX20, was 1430 m/s.
+This is a contrived value taken in the style of \citet{Graupe2019} for the sake of comparison; the default value in the LAMSS simulator, which was not environmentally informed and used when no updates were available from the ICNN, was 1430 m/s.
 
@@
 
-\subsection{Minimal bounce criteria (MBC)}
+\subsection{Minimal bounce (MB) criteria}
 
 \llabel{1.5} The fundamental challenge to implement GNSS-like navigation, especially in an acoustically complex propagation environment, is characterizing a single sound speed to compensate for the effects of ray refraction and reflection.
-The use of the acoustic modem network for tracking relies on the accurate estimates of travel times between the submerged platform and LBL beacons, supported by clock synchronization and a pre-determined scheduling of acoustic events.
+The use of the acoustic modem network for tracking relies on the accurate estimation of travel times between the submerged platform and LBL beacons, supported by clock synchronization and a pre-determined scheduling of acoustic events.
 \llabel{2.9} For the Beaufort Lens in particular, the strong multipath effects make it virtually impossible to deterministically predict the modem's detected arrival time.
 
-Instead, for each individual modem $i$, an embedded stochastic tracking framework is used to provide a running estimate of the effective sound speed $c_{i,j}$ for the conversion from travel time to range from modem $j$, with the ultimate goal of matching the implied horizontal effective sound speed, i.e., the GPS-recorded distance between two nodes divided by the modem-recorded one way travel time between them.
+Instead, for each individual receiver $i$, an embedded stochastic tracking framework is used to provide a running estimate of the effective sound speed $c_{i,j}$ for the conversion from travel time to range from modem $j$, with the ultimate goal of matching the implied horizontal effective sound speed, i.e., the GPS-recorded distance between two nodes divided by the modem-recorded one way travel time between them.
 
-\llabel{2.10} In the ICEX20 configuration, the acoustic tracking is running on the topside computer, which controls the ICNN.
-Here we assume that the effective sound speeds $c_{i,j}$ are smoothly varying over the course of a vehicle mission, i.e., with respect to range, mission time, and the thirty-second frequency.
+\llabel{2.10} In the ICEX20 configuration, the acoustic tracker is running on the topside computer, which controls the ICNN.
+Here we assume that the effective sound speeds $c_{i,j}$ are smoothly varying over the course of a vehicle mission, i.e., with respect to range from signal origin at transmitter $j$, mission time, and the thirty-second frequency.
 
@@
 For each grid point, BELLHOP produces a number of arrivals resulting from multiple propagation paths.
-Using only the $N_0$ rays with neither surface nor bottom bounces, it will then estimate the current effective sound speed $c$ from a power weighted average of the ray travel times,
+Using only the $N_0$ rays with neither surface nor bottom bounces, the tracking system will then estimate the current effective sound speed $c$ from a power weighted average of the ray travel times,
 \begin{equation}
@@
 
-Thus the MBC method assumes the signal detected by the modem will be dominated by a set of paths with the least number of boundary interactions.
+Thus the minimal bounce (MB) criteria assumes the signal detected by the modem will be dominated by a set of paths with the least number of boundary interactions.
 Importantly, this stochastic, ensemble method for group velocity calculation can run in real-time, appearing to be orders of magnitude faster than other post-processing methods which seek to determine the specific ray itself that best matches a prominent indicator from the arrival structure.
@@
 
-The sister modem experiment generated 811 beacon to beacon communication events with their own real-time MBC group velocity predictions.
-Given the complexity of the ICNN system, this experiment did not collect an exhaustive set of data across all buoy, source depth, receive depth, and sound speed combinations.
+The sister modem experiment generated 811 beacon to beacon communication events with their own effective sound speed predictions from a real-time implementation of the MB method.
+Given the complexity of the ICNN system, this experiment did not collect an exhaustive set of data across all buoy, source depth, receive depth, and model sound speed combinations.
 \llabel{1.18} The algorithm generally overestimates pseudoranges because it resolves the effective sound speed for the most direct path.
@@
 Fig. \ref{fig:range-error-insitu} shows the range error boundary for both SSP inputs used in ICEX20.
-A promising sign that the MBC method adapts sound speed somewhat intelligently is the lack of error growth as travel time increases.
+A promising sign that the MB method adapts sound speed somewhat intelligently is the lack of error growth as travel time increases.
 \llabel{1.28}
@@
 The discrepancy between these two is largely due to outlier events only contained in the weighted SSP set.
-Where there is overlap between sound speed conditions used for the real-time MBC, the pseudorange error difference is no more than a few meters.
+Where there is overlap between sound speed conditions used for the real-time MB approach, the pseudorange error difference is no more than a few meters.
 The overarching results show that sound speed estimates derived from eigenrays for a local grid, as opposed to a singular point, are accurate enough to support vehicle navigation.
-While the NBC looks for just the least complex multipath, the high density of launch angles almost always guarantees a direct path.
+While the MB looks for just the least complex multipath, the high density of launch angles almost always guarantees a direct path for the beacon-to-beacon configurations.
 Nonetheless, the consistent overestimation of pseudorange invites further analysis into acoustic arrival matching.
@@
 Rays for the weighted SSP show how the surface channel intensifies ice interactions and how the shadow zone denies reliable acoustic paths.
-Pointedly, the increasing number of surface reflections to the farthest shallow buoy (North, in red) crystallize the MBC's tendency for overestimation.
+Pointedly, the increasing number of surface reflections to the farthest shallow buoy (North, in red) crystallize the MB criteria's tendency for overestimation.
 For the HYCOM, baseline, and weighted SSP inputs, the most appropriate eigenrays show 2, 3, and 4 surface interactions.
@@
 The HYCOM eigenrays still show the most reliable acoustic paths that deteriorate with increasing ducted conditions.
-The lack of direct paths from the observed SSP further points out the shortcomings of the MBC approach.
+The lack of direct paths from the observed SSP further points out the shortcomings of the MB approach.
 
-The goal of the MBC algorithm was to provide a reliable, physically intuitive interpretation of the acoustic propagation without taking on the additional burden of regularly identifying specific paths that may connect any given source-receiver pair in the network.
+The goal of the MB algorithm was to provide a reliable, physically intuitive interpretation of the acoustic propagation without taking on the additional burden of regularly identifying specific paths that may connect any given source-receiver pair in the network.
 While it was unlikely to resolve multipath arrivals that triggered successful modem detection, an isovelocity approach would have provided no adaptivity against source and receiver depth differences.
@@
 
-It is important to note that the value for the extrapolated range, $\hat{r}$, is only tracked when the modem runs the vehicle behavior; thus we replace $\hat{r}$ with the GPS-tracked range for all modem events.
+It is important to note that the value for the extrapolated range, $\hat{r}$, is only tracked by topside for a modem claiming to be the vehicle; thus we replace $\hat{r}$ with the GPS-tracked range for all modem events.
 Because $\hat{r}$ converges to the correct solution, a comparison of $\hat{r}$ with the GPS-tracked range shows a normal, zero-centered distribution within the bounds of GPS drift.
@@
 Sound speed inputs are the isovelocity sound speed in addition to the modeled, baseline, and weighted SSPs from Fig. \ref{fig:sspExpectation}.
-The analysis replicates the MBC but also introduces a new filtering algorithm, the nearest bounce criteria (NBC), based on insights gleaned from the eigenray analysis.
+The analysis replicates the MB criteria but also introduces a new filtering algorithm, the nearest bounce (NB), based on insights gleaned from the eigenray analysis.
 Accordingly, the results in this section evaluate the utility of the algorithms and sound speed sources, divorced from their role in the ICNN while maintaining real-time relevance.
 
-\subsection{Nearest bounce criteria (NBC)}
+\subsection{Nearest bounce (NB) criteria}
 
 \llabel{2.12} The extent of ray bending and repeated reflections is extremely dependent on the degree of the Beaufort Lens observed.
-Based on this insight, a new algorithm, the nearest bounce criteria (NBC), is a slight modification from the MBC and includes multipath as a new dimension of information to exploit.
+Based on this insight, a new algorithm, the nearest bounce (NB) criteria, is a slight modification from the MB and includes multipath as a new dimension of information to exploit.
 This metric, while run in post-processing, adds a negligible amount of computation for real-time efficacy.
@@
 
-Whereas the MBC outputs a scalar, this method first outputs a vector of effective sound speeds based on the number of reflections.
-Then a single value is selected that best matches the recorded travel time, as the detected arrival is not always the first arrival or the direct path and could even be masked by noise or blocked temporarily \citep{deffenbaugh_acoustic_1996}.
+Whereas the MB outputs a scalar, this method first outputs a vector of effective sound speeds based on the number of reflections.
+Then a single value is selected in a nearest-neighbor fashion that best matches the recorded travel time, as the detected arrival is not always the first arrival or the direct path and could even be masked by noise or blocked temporarily \citep{deffenbaugh_acoustic_1996}.
 We manually cap the number of bounces at four because of the smaller operational scale and the attenuation accrued with many surface interactions.
@@
 HYCOM sees the most direct and one bounce multipath structures, lending a bias for faster speeds; the weighted SSP sees the most double and triple bounces, favoring slower speeds; the baseline sound speed exists in between.
-Very rarely is the multipath structure classified as a direct path, where the MBC and NBC would prediction overlap.
+Very rarely is the multipath structure classified as a direct path, i.e., where the NB defaults to the MB prediction.
 In fact, the higher the multipath classification, the more accurate the sound speed prediction is, likely driven by a tighter or even sparser bundle of rays.
-\llabel{1.19} Discontinuities in multipath classification provide initial evidence for its importance to a smoothly varying group velocity, as shown in the cluster of 30 to 30 m transmissions in \ref{fig:gvel-post}, where HYCOM jumps from one to two classified bounces amidst the baseline SSP and weighted SSP smoothly increasing while consistently seeing two and three classified bounces, respectively.
+\llabel{1.19} Discontinuities in multipath classification provide initial evidence for its importance to a smoothly varying group velocity, as shown in the cluster of 30 to 30 m transmissions in Fig.
+\ref{fig:gvel-post}, where HYCOM jumps from one to two classified bounces amidst the baseline SSP and weighted SSP smoothly increasing while consistently seeing two and three classified bounces, respectively.
 Of course, the prediction deteriorates with cross-layer transmissions across the duct, but not to the same degree at which eigenrays could not be found for the weighted SSP in section \ref{sec:eigenrays}.
@@
 It is useful to think about in what case the isovelocity\textemdash or any isovelocity framing\textemdash would have been appropriate.
-The transmissions from shallow to shallow receiver would may have matched the default configuration of 1430 m/s.
+The transmissions from shallow to shallow receiver may have matched the default configuration of 1430 m/s.
 The isovelocity contrived for this paper, 1441.8 m/s, best matches the transmissions from 90 to 90 m.
@@
 Pseudorange estimation plays an important role in trilateration.
-Fig. \ref{fig:rangeError} shows the directional pseudorange error ``footprints'' for the four sound speed inputs with the NBC approach, separated by source and receiver depth configurations.
+Fig. \ref{fig:rangeError} shows the directional pseudorange error ``footprints'' for the four sound speed inputs with the NB approach, separated by source and receiver depth configurations.
 
@@
 
-The improvement from MBC to NBC is most evident for the data-driven sound speed; while the HYCOM SSP improves from a median absolute range error of 6.41 to 4.61 m, the baseline SSP improves from 10.30 to 2.27 m, and the weighted SSP improves from 13.28 to 2.12 m.
+The improvement from MB to NB is most evident for the data-driven sound speed; while the HYCOM SSP improves from a median absolute range error of 6.41 to 4.61 m, the baseline SSP improves from 10.30 to 2.27 m, and the weighted SSP improves from 13.28 to 2.12 m.
 In comparison, the isovelocity has a median error of 13.09 m.
-The order of magnitude improvement in the ducted SSPs demonstrate the effectiveness of the NBC algorithm exploiting the observed multipath conditions.
+The order of magnitude improvement in the ducted SSPs demonstrate the effectiveness of the NB algorithm exploiting the observed multipath conditions.
 
@@
 For transmissions between North and South at 30 m, the OWTT spread is 2.1958 to 2.1963 s; the GNSS-tracked distance is 3138.54 to 3140.87 m; and the implied effective sound speed is 1429.3 to 1430.1 m/s.
-For these transmissions, the weighted SSP and the MBC approach produce a pseudorange error of -1491 m, as the effective sound speed is dominated by bottom bounce arrivals with much greater travel times.
-The NBC approach categorizes this same record as a quadruple surface bounce, reducing the pseudorange error to less than a meter.
-Comparatively, the NBC approach for HYCOM and the baseline SSP produce pseudorange errors of 8.30 and 2.39 m, respectively. 
+For these transmissions, the weighted SSP and the MB approach produce a pseudorange error of -1491 m, as the effective sound speed is dominated by bottom bounce arrivals with much greater travel times.
+The NB approach categorizes this same record as a quadruple surface bounce, reducing the pseudorange error to less than a meter.
+Comparatively, the NB approach for HYCOM and the baseline SSP produce pseudorange errors of 8.30 and 2.39 m, respectively. 
 There is strong evidence to suggest that the sound speed and multipath fidelity codependently improve localization accuracy.
@@
 The reliability of the correction is directly linked to how accurately the travel time measurements are converted to pseudoranges.
-This section aims to resolve that tension by reevaluating the trilateration results with respect to the MBC and NBC algorithms.
-The MBC/NBC effective speed predictions were tracked independently for each source-receiver pair; although the sound speed was expected to be locally smooth near a given receiver, no such assumption was enforced between distinct acoustic links.
+This section aims to resolve that tension by reevaluating the trilateration results with respect to the MB and NB algorithms.
+The MB/NB effective speed predictions were tracked independently for each source-receiver pair; although the sound speed was expected to be locally smooth near a given receiver, no such assumption was enforced between distinct acoustic links.
 
@@
 Therefore, a distribution of corrections from the ICNN, as shown in Fig. \ref{fig:trilat-beacon}, reflects positioning accuracy.
-The NBC clearly outperforms the MBC, with almost 80\% of the corrections below 6 meters and the median within the deployed GNSS puck precision of 3 meters.
-By contrast, the MBC shows roughly 20\% within the GNSS puck precision, and separate peaks from 9--12 meters and 21--27 meters.
+The NB clearly outperforms the MB, with almost 80\% of the corrections below 6 meters and the median within the deployed GNSS puck precision of 3 meters.
+By contrast, the MB shows roughly 20\% within the GNSS puck precision, and separate peaks from 9--12 meters and 21--27 meters.
 This trimodal nature reflects the distribution of reflections on the ice surface.
@@
 When limited to only two ranging circles, the trilateration chooses the closer of the two intersection points.
-The x-axis is cut off at the upper limit for the NBC algorithm.}
+The x-axis is cut off at the upper limit for the NB algorithm.}
 \label{fig:trilat-beacon}
@@
 
-In several events, the MBC is unable to accurately estimate the effective sound speed for one of the acoustic links, leading to a large positioning error.
-The NBC, however, better resolves an approximation of the acoustic path.
-For example, in some trilateration solutions for the Eastern buoy, the MBC shows a correction of more than a kilometer; the NBC is two orders of magnitudes less.
+In several events, the MB is unable to accurately estimate the effective sound speed for one of the acoustic links, leading to a large positioning error.
+The NB, however, better resolves an approximation of the acoustic path.
+For example, in some trilateration solutions for the Eastern buoy, the MB shows a correction of more than a kilometer; the NB is two orders of magnitudes less.
 
@@
 
-Up to this point, pseudorange estimation and localization have been evaluated on GPS-linked beacon-to-beacon connections to validate the NBC algorithm.
-This analysis ports the MBC and NBC algorithms to re-navigate the AUV \emph{Macrura}.
+Up to this point, pseudorange estimation and localization have been evaluated on GPS-linked beacon-to-beacon connections to validate the NB algorithm.
+This analysis ports the MB and NB algorithms to re-navigate the AUV \emph{Macrura}.
 
@@
 Fig. \ref{fig:trilat-auv} shows the correction magnitudes and trilateration errors for events with three or more receptions during AUV operations.
-Whereas the MBC has a fairly bimodal nature, with peaks centered around 10--15 and 35--40 m, the NBC favors smaller corrections, from 5--20 m, and has a long tail.
+Whereas the MB has a fairly bimodal nature, with peaks centered around 10--15 and 35--40 m, the NB favors smaller corrections, from 5--20 m, and has a long tail.
 The distribution of corrections are much larger than the distribution of RMS error.
-It is apparent that, while both methods are quite successful, there is strong evidence that the NBC achieves single meter accuracy.
+It is apparent that, while both methods are quite successful, there is strong evidence that the NB achieves single meter accuracy.
 
@@
 \includegraphics[width=\textwidth]{figs/auv-trilat-stat.pdf}
-\caption{Distribution of vehicle re-navigation corrections (top) and error (bottom) computed in post-processing for the Minimum and Nearest Bounce Criterion.}
+\caption{Distribution of vehicle re-navigation corrections (top) and error (bottom) computed in post-processing for the Minimal and Nearest Bounce Criterion.}
 \label{fig:trilat-auv}
@@
 The fact that the bulk of the best performing re-navigation error exists within the precision of the GNSS unit deployed invites a further look into GNSS noise.
-In the Arctic, GNSS performance worsens due to poor constellation coverage, larger ionospheric effects, and multipath interference.
-The National Security Implications of Climate Change for U.S. Naval Forces \citep{NAP12914} details some of the limitations of the Global Positioning System (GPS) at polar latitudes.
+In the Arctic, GNSS performance worsens due to poor constellation coverage, larger ionospheric effects, and multipath interference \citep{NAP12914,swanlund_gps_2016,Themens2015,Gwal2011,Jung2018,Reid2016}.
 Radio infrastructure that provides position corrections and references does not regularly extend to polar regions.
@@
 The field-tested approach, the minimal bounce criteria, facilitated a successful AUV recovery in a total ice-covered, double ducted environment.
-The accuracy of the MBC was validated against GPS-linked beacon-to-beacon communications.
+The accuracy of the MB was validated against GPS-linked beacon-to-beacon communications.
 Given a consistent bias towards overestimation, an improved algorithm, the nearest bounce criteria, was developed on the insight that multipath structure may play an outsized role in maintaining a smoothly varying effective sound speed.
-The NBC was developed with field data and reevaluated on vehicle data, achieving a position accuracy and precision that rivals that of the deployed GNSS puck.
+The NB was developed with field data and reevaluated on vehicle data, achieving a position accuracy and precision that rivals that of the deployed GNSS puck.
 

### 2022-02-21 pipeline/4-analyze-range-error/plot_post_range_error_nbc.m

@@ -1,5 +1,5 @@
 %% prep workspace
-clear; clc; close all;
+clear; clc; close all; addpath('../../src/');
 
 %% unpack Bellhop gvel table
 [DATA,INDEX] = h_unpack_bellhop('../bellhop-gvel-gridded/gveltable.csv');
@@ -96,7 +96,7 @@ for zr = [30 90]
         %         end
         if count == 4
             xlabel('one way travel time [s]');
-            ylabel('sound speed [m/s]');
+            ylabel('pseudorange error [m]');
         else
             yticklabels([]);
             xticklabels([]);
@@ -115,5 +115,5 @@ lg = legend('Isovelocity','HYCOM','Baseline','Chosen Weights','location','southe
 title(lg,'Sound Speed Inputs');
 
 % title
-sgtitle('Range error by source and receiver depths','fontsize',17,'fontweight','bold')
-% h_printThesisPNG('range-error-allMethods')
+sgtitle('Pseudorange error footprints','fontsize',17,'fontweight','bold')
+h_printThesisPNG('range-error-allMethods')

### 2022-02-22 manuscript/main.tex
final oscar edits

@@
 For transmissions between North and South at 30 m, the OWTT spread is 2.1958 to 2.1963 s; the GNSS-tracked distance is 3138.54 to 3140.87 m; and the implied effective sound speed is 1429.3 to 1430.1 m/s.
-For these transmissions, the weighted SSP and the MB approach produce a pseudorange error of -1491 m, as the effective sound speed is dominated by bottom bounce arrivals with much greater travel times.
+For these transmissions, the weighted SSP and the MB approach produce a pseudorange error of -1491 m, as the effective sound speed predicted by the minimum bounce criteria is dominated by bottom bounce arrivals with much greater travel times.
 The NB approach categorizes this same record as a quadruple surface bounce, reducing the pseudorange error to less than a meter.
@@
 Fig. \ref{fig:gps-drift-example} zooms in on the GNSS and OWTT noise relative to the ice movement for two pairs of modem buoy connections.
-The two panels indicate the GPS drift as $\delta R = \sqrt{\delta x^2 + \delta y^2}$ and temporal drift, $\delta t$, relative to the median OWTT recorded between the two modems.
+The two panels indicate the GPS noise as $\delta R = \sqrt{\delta x^2 + \delta y^2}$ and temporal drift, $\delta t$, relative to the median OWTT recorded between the two modems.
 The dashed line is scaled by a group velocity of 1440 m/s, such that if there were ideal sensor measurements with no drift, all events should exist on or near the line.
@@
 	\includegraphics[width=\columnwidth]{figs/gps-drift-example.pdf} 
-	\caption{A comparison of GPS drift (y-axis) versus OWTT drift (x-axis) for corners of the ICNN network with different source depths.}
+	\caption{A comparison of GPS noise (y-axis) versus OWTT drift (x-axis) for corners of the ICNN network with different source depths.}
 	\label{fig:gps-drift-example}
@@
 
-The top panel shows the connections between the North and East buoys.
+The left panel shows the connections between the North and East buoys.
 The clusters scaling along the 1440 m/s guideline suggest relative ice movement picked up by both GNSS and OWTT.
@@
 Thus, the conversion of travel time into distance can be pre-conditioned for error and error growth over the course of a vehicle mission.
-This work introduces a lightweight stochastic prediction of an effective sound speed along the path between source and receiver, retooling arrival methods generally deemed too complex or labor intensive for real-time.
+This work introduces a lightweight stochastic prediction of an effective speed along the acoustic path between source and receiver, retooling arrival methods generally deemed too complex or labor intensive for real-time.
 We assume that the effective sound speed would be a locally smoothly varying function with respect to operational conditions\textemdash horizontal and vertical differences and rate of difference between source and receiver.
@@
 Given a consistent bias towards overestimation, an improved algorithm, the nearest bounce criteria, was developed on the insight that multipath structure may play an outsized role in maintaining a smoothly varying effective sound speed.
-The NB was developed with field data and reevaluated on vehicle data, achieving a position accuracy and precision that rivals that of the deployed GNSS puck.
+The NB was developed with field data and reevaluated on vehicle data, achieving a position accuracy and precision that compares with that of the deployed GNSS puck.
 
@@
 The relatively simple nature of this approach suggests it is transferable to other environments, spatio-temporal scales, and platforms.
-While it is likely a particular quirk of the Beaufort Lens that filtering for reflection alone can produce a horizontal effective speed that compensates for ray refraction and reflection, it is trivial to filter along other ways, like number of turning points, to create a more diverse and informed set of multipath timefronts.
+While it is likely a particular quirk of the Beaufort Lens that filtering for reflection alone can produce a horizontal effective speed that compensates for ray refraction and reflection, the algorithm can be reconditioned to filter against other metrics, like number of turning points, to create a more diverse and informed set of multipath timefronts.
 Though the majority of re-navigation results are within single-meter accuracy, future work can examine how constellations of more LBL beacons can extend the operational domain without adding an undesirable amount of error.
 One possibility is that, during a mission, ICNN-like LBL implementions use a comparison of the GNSS self-position and acoustic positioning to invert for the ocean volume, linking how vertical and horizontal sound speed structure impact transmission integrity.
-\llabel{1.20} A fast tomographic estimate \citep{deffenbaugh_optimal_1997,Elisseeff2002}, along with its uncertainty, can be continuously communicated to assets underway to maintain contact or enable adaptive sampling.
+\llabel{1.20} A fast tomographic estimate \citep{deffenbaugh_optimal_1997,Elisseeff2002}, along with its uncertainty, could be continuously communicated to assets underway to maintain contact or enable adaptive sampling.
 In this sense navigation and tomography converge on the same set of component technologies\textemdash position estimation, sound speed parameterization estimation, ray path identification, and vehicle path optimization.
@@
 Resolving features inaccurately, or with a false sense of confidence, could be more harmful than contextualizing the limitations of a range independent propagation over realistic bathymetry.
-Given that AUV operations are often on smaller spatial and temporal scales, the added benefit of a gridded model is quite small, and for features like the Beaufort Lens, not well resolved.
+Given that AUV operations are often on smaller spatial and temporal scales, the added benefit of an ocean model is quite small, and for features like the Beaufort Lens, not well resolved.
 
 \llabel{1.1}
-The methods presented in this paper, including the software projects \citep{,Benjamin2010,schneider2015dynamic,schneider_unified_2010}, are open source and platform agnostic.
+The methods involved in this paper include open source software projects \citep{,Benjamin2010,schneider2015dynamic,schneider_unified_2010} that are platform agnostic.
 Large AUVs, often large enough to support long duration and/or deep sea missions, would benefit from including diurnal or tidal effects for ranging.
@@
 Their inability to maintain position within a region of reliable acoustic path makes the impact of an environmentally adaptive pseudorange estimation disproportionately positive.
-The exact adjustments to the ensemble eigenray filtering are predicated on the expected sound speed conditions and acoustic arrival structure; the problem is ripe application for other simulation testbeds or machine learning methods.
+The exact adjustments to the ensemble eigenray filtering are predicated on the expected sound speed conditions and acoustic arrival structure; the problem is ripe for other simulation testbeds or machine learning methods.
 The continued development of embedded acoustic processing on heterogenous platforms is fundamental to support a universal underwater navigation scheme comparable to GNSS.

### 2022-03-29 pipeline/4-analyze-range-error/stat_4ssp.m
pipeline4

@@ -1,0 +1,34 @@
+%% prep workspace
+clear; clc; close all;
+
+%% unpack Bellhop gvel table
+[DATA,INDEX] = h_unpack_bellhop('../bellhop-gvel-gridded/gveltable.csv');
+
+%% load simulation
+listing = dir('../bellhop-gvel-gridded/csv_arr/*gridded.csv');
+[T,colorSet] = h_get_nbc(listing,DATA,INDEX);
+
+%% isovelocity case - loads "iso"
+load isovelocity-ssp.mat
+
+%% statistics
+
+for s = [0 5 3 4]
+    
+    if s == 0
+        yval = iso.avg .* DATA.owtt(INDEX.valid) - DATA.recRange(INDEX.valid);
+    else
+        %yval = T{s}.gvel(INDEX.valid) .* DATA.owtt(INDEX.valid) - DATA.recRange(INDEX.valid);
+        yval = T{s}.gvel .* DATA.owtt - DATA.recRange;
+
+    end
+    
+    % print statistics
+    fprintf('%d \n',s);
+    fprintf('median = %3.2f \n', median(yval,'omitnan'));
+    
+    
+    
+    fprintf('\n \n');
+end
+    

### 2022-03-29 manuscript/jasa-num-figs/manuscript.tex

--- 2022-02-22 manuscript/main.tex
@@
+%	JASA LaTeX Sample File, Preprint Sample
+%
+%  Beginner Latex users should refer to their favorite online documentation
+%  here is one from the TeX Users Group
+%	https://www.tug.org/twg/mactex/tutorials/ltxprimer-1.0.pdf
+%
+%  Useful FAQ from  https://journals.aps.org/revtex/revtex-faq
+%
+
+%%%%%%% For Preprint
+%% For manuscript, 12pt, one column style
+
+\documentclass[preprint,TurnOnLineNumbers]{JASA}
+
+%%%%% Preprint Options %%%%%
+%% The track changes option allows you to mark changes
+%% and will produce a list of changes, their line number
+%% and page number at the end of the article.
+ %\documentclass[preprint,trackchanges]{JASA}
+
+%% NumberedRefs is used for numbered bibliography and citations.
+%% Default is Author-Year style.
+% \documentclass[preprint,NumberedRefs]{JASA}
+
+%%%%%%% For Reprint
+%% For appearance of finished article; 2 columns, 10 pt fonts
+
+% \documentclass[reprint]{JASA}
+
+%%%%% Reprint Options %%%%%
+
+%% For testing to see if author has exceeded page length request, use 12pt option
+% \documentclass[reprint,12pt]{JASA}
+
+%% NumberedRefs is used for numbered bibliography and citations.
+%% Default is Author-Year style.
+% \documentclass[reprint,NumberedRefs]{JASA}
+
+%% TurnOnLineNumbers
+%% Make lines be numbered in reprint style:
+% \documentclass[reprint,TurnOnLineNumbers]{JASA}
+
+%% Optional algorithm package. You can comment this out and
+%% include another package if you prefer another way to make
+%% algorithms examples (but please check that the package is compatible with
+%%Editorial Manager; see JASA-EL-TeXGuide.pdf).
+
+%\usepackage{algpseudocode}
+\usepackage{placeins}
+\usepackage{colortbl}
+
+%% MACROS FOR CROSS REFERENCING %%
+\newcommand{\llabel}[1]{\hypertarget{llineno:#1}{\linelabel{#1}}}
+
+\begin{document}
+
 \title[JASA/draft]{Under-ice acoustic navigation using real-time model-aided range estimation}
-\author{EeShan C. Bhatt}
+\author{EeShan C. Bhatt \footnote{Also at MIT-WHOI Joint Program in Oceanography/Applied Ocean Science \& Engineering, Cambridge, MA}}
 \email{ebhatt@whoi.edu}
-\affiliation{MIT-WHOI Joint Program in Oceanography/Applied Ocean Science \& Engineering, Cambridge and Woods Hole, MA, USA}
 \affiliation{Department of Mechanical Engineering, Massachusetts Institute of Technology, Cambridge, MA}
@@
 
-% \author{Author Four}
-% \email{author.four@university.edu}
-% \affiliation{Department2,  University2, City, State ZipCode, Country}
-
 \preprint{Bhatt, JASA}   %  if you want want this message to appear in upper right corner of title page
@@
 \begin{abstract}
-The long baseline (LBL) underwater navigation paradigm relies on the conversion of recorded travel times into pseudoranges to trilaterate position.
-For real-time operations, this conversion has assumed an isovelocity sound speed.
-For re-navigation in post-processing, computationally and/or labor intensive acoustic modeling may be employed to reduce uncertainty driven by multipath arrivals.
-This work demonstrates a real-time ray-based prediction method of the effective sound speed along a path from source to receiver to minimize vehicle position error.
-\llabel{2.1} This method was implemented for a small scale AUV-LBL system in March 2020, in the Beaufort Sea, in total ice-covered conditions and a double ducted acoustic propagation environment.
-The vehicle was successfully deployed and recovered.
-\llabel{2.2} Given the lack of GPS data throughout the vehicle's mission, however, the pseudorange performance is first evaluated on connections between GPS-linked beacons.
+The long baseline (LBL) underwater navigation paradigm relies on the conversion of travel times into pseudoranges to trilaterate position. 
+For real-time operations, this conversion assumes an isovelocity sound speed.
+For re-navigation, computationally and/or labor intensive acoustic modeling may be employed to reduce uncertainty.
+This work demonstrates a real-time ray-based prediction method of the effective sound speed along a path from source to receiver to minimize pseudorange error.
+\llabel{2.1} This method was implemented for an AUV-LBL system in the Beaufort Sea, in total ice-covered conditions and a double-ducted propagation environment.
+\llabel{2.2} Given the lack of GNSS data throughout the vehicle's mission, the pseudorange performance is first evaluated on connections between GNSS-linked beacons.
 The real-time ranging error between beacons is roughly 11 meters at distances up to 3 km.
-But a consistent overestimation in the real-time method provides insights for improved eigenray filtering by the number of surface bounces.
-An operationally equivalent pipeline is used to re-position the LBL beacons and re-navigate the AUV, using a modeled, historical, and a locally observed sound speed profile.
-The median re-navigation error is 1.84 $\pm$ 2.19 RMS m.
-The improved trilateration performance for suggests that this approach effectively extends the single meter accuracy of the deployed GNSS units into the water column. \llabel{2.3}
+A consistent overestimation in the real-time method provides insights for improved eigenray filtering by the number of bounces.
+An operationally equivalent pipeline is used to re-position the LBL beacons and re-navigate the AUV, using modeled, historical, and a locally observed sound speed profiles.
+The median re-navigation error is 1.84 $\pm$ 2.19 m.
+\llabel{2.3} The improved performance for suggests that this approach extends the single meter accuracy of the deployed GNSS units into the water column.
 \end{abstract}
@@
     \centering
-	\includegraphics[width=0.8\columnwidth]{figs/Fig2.pdf}
+    \includegraphics[width=0.8\columnwidth]{Fig1.pdf}
     \caption{A schematic overview of the Integrated Communication and Navigation Network (ICNN), which provides joint data-transfer and tracking between AUV and a human decision maker at Topside.}
@@
     \centering
-	\includegraphics[width=0.7\columnwidth]{figs/auv-track-update.png} \hfill
-	\includegraphics[width=0.28\columnwidth]{figs/Fig1.jpg}
+    \includegraphics[width=0.7\columnwidth]{Fig2a.pdf} \hfill
+    \includegraphics[width=0.28\columnwidth]{Fig2b.jpg}
     \caption{The under-ice mission track for AUV \emph{Macrura}, including the position updates as it stalled underneath the ice overnight.
@@
     \centering
-	\includegraphics[width=\reprintcolumnwidth]{figs/ssp-gvel-icex20-icex16.pdf}
+    \includegraphics[width=\reprintcolumnwidth]{Fig3.pdf}
     \caption{Sound speed conditions for historical (baseline), data (chosen weights), and HYCOM model for both ICEX16 and ICEX20.}
@@
 Because the vehicle's navigation solution during a mission can only be evaluated on the basis of the error estimates sent, a sister experiment for validating the real-time ranging approach was implemented.
-Ice buoy modems were run as ``virtual vehicles'' at a fixed depth, receiving position updates from the other beacons as well as a camp site modem lowered to 20 m. \llabel{1.13}
+Ice buoy modems were run as ``virtual vehicles'' at a fixed depth, receiving position updates from the other beacons as well as a camp site modem lowered to 20 m. %\llabel{1.13}
 Fig. \ref{fig:overview} shows successful events sorted by source depth.
@@
   \centering
-  \includegraphics[width=\columnwidth]{figs/modem-chart.pdf}
+  \includegraphics[width=\columnwidth]{Fig4.pdf}
   \caption{An overview of the modem experiment by source and receiver depth and position.
@@
   \centering
-  \includegraphics[width=\columnwidth]{figs/range-error-insitu.pdf}
+  \includegraphics[width=\columnwidth]{Fig5.pdf}
   \caption{The real-time range error by source (rows: 20, 30, and 90 m) and receiver (columns: 30 and 90 m) pairings for both sound speed estimates used during ICEX20.
@@
   \centering
-  \includegraphics[width=\columnwidth]{figs/raytrace-3env-zs-20.pdf}
+  \includegraphics[width=\columnwidth]{Fig6.pdf}
   \caption{Eigenrays for beacon-to-beacon events for each sound speed with a nominal source depth of 20 m over a total ray fan in gray.
@@
   \centering
-  \includegraphics[width=\columnwidth]{figs/raytrace-3env-zs-30.pdf}
+  \includegraphics[width=\columnwidth]{Fig7.pdf}
   \caption{Eigenrays for beacon-to-beacon events for each sound speed with a nominal source depth of 30 m over a total ray fan in gray.
@@
   \centering
-  \includegraphics[width=\columnwidth]{figs/raytrace-3env-zs-90.pdf}
+  \includegraphics[width=\columnwidth]{Fig8.pdf}
   \caption{Eigenrays for beacon-to-beacon events for each sound speed with a nominal source depth of 90 m over a total ray fan in gray.
@@
 \begin{figure}[h!]
-\includegraphics[width=\columnwidth]{figs/gvel-txrxdepth-wIso-wGPS.pdf}
+\includegraphics[width=\columnwidth]{Fig9.pdf}
 \caption{A post-processing comparison of effective sound speed predictions for all beacon-to-beacon events.
@@
 \begin{figure}[!ht]
-\includegraphics[width=\columnwidth]{figs/range-error-allMethods.pdf}
+\includegraphics[width=\columnwidth]{Fig10.pdf}
 \caption{The post-processed pseudorange error; the rows share receiver depth; the columns share source depth.
@@
 \begin{figure}[!ht]
-\includegraphics[width=\textwidth]{figs/beacon-trilat-stat.pdf}
+\includegraphics[width=\textwidth]{Fig11.pdf}
 \caption{Trilateration events; however, 264 entries are 2-beacon solutions, 22 are 3-beacon, and 2 are 4-beacon.
@@
 \begin{figure}[!ht]
-\includegraphics[width=\textwidth]{figs/auv-trilat-stat.pdf}
+\includegraphics[width=\textwidth]{Fig12.pdf}
 \caption{Distribution of vehicle re-navigation corrections (top) and error (bottom) computed in post-processing for the Minimal and Nearest Bounce Criterion.}
@@
     \centering
-	\includegraphics[width=\columnwidth]{figs/gps-drift-example.pdf} 
+    \includegraphics[width=\columnwidth]{Fig13.pdf} 
     \caption{A comparison of GPS noise (y-axis) versus OWTT drift (x-axis) for corners of the ICNN network with different source depths.}
@@
 \bibliography{sampbib}
+
+\end{document}

### 2022-03-29 pipeline/4-analyze-range-error/plot_insitu_range_error_by_ssp.m
making both plots structured same dimensions

@@ -9,15 +9,16 @@ A = h_clean_simgvel(A);
 
 load p_sspColorDetails
 %% figure 1 -- all events in 3x2 grid
-figure('name','rangeanomaly-by-owtt','renderer','painters','position',[108 108 1100 900]);
-t = tiledlayout(3,2,'TileSpacing','none','Padding','compact');
+figure('name','rangeanomaly-by-owtt','renderer','painters','position',[108 108 1300 800]);
+t = tiledlayout(2,3,'TileSpacing','none','Padding','compact');
 
 index3 = ~isnan(A.sim_gvel);
 count = 0;
-for zs = [20 30 90]
-    index1 = A.tx_z == zs;
+for zr = [30 90]
     
-    for zr = [30 90]
+    for zs = [20 30 90]
+        
+        index1 = A.tx_z == zs;
         index2 = A.rx_z == zr;
         
         count = count + 1;
@@ -31,7 +32,7 @@ for zs = [20 30 90]
             else
                 eof_str = 'baseline';
             end
-        
+            
             index = logical(index1.*index2.*index3.*index4);
             
             if sum(index)>=1
@@ -73,42 +74,40 @@ for zs = [20 30 90]
             % for all grids
             %title(sprintf('source depth = %u m',zs),'fontsize',14,'fontweight','bold');
             title('  ','fontsize',8);
-            text(2.2,24,sprintf('tx depth = %u m',zs),'HorizontalAlignment','right','VerticalAlignment','bottom','fontsize',11);
-            text(2.2,24,sprintf('rx depth = %u m',zr),'HorizontalAlignment','right','VerticalAlignment','top','fontsize',11);
+            text(2.25,23,sprintf('source depth = %u m',zs),'HorizontalAlignment','right','VerticalAlignment','bottom','fontsize',11,'fontweight','bold');
+            text(2.25,23,sprintf('receiver depth = %u m',zr),'HorizontalAlignment','right','VerticalAlignment','top','fontsize',11,'fontweight','bold');
             
-            if sum(index)~=1
-                text(2.2,21-3.*eof_status,sprintf('%s = %u events',eof_str,sum(index)),'HorizontalAlignment','right','VerticalAlignment','top','fontsize',11);
+            if eof_status==0
+                text(2.25,18,sprintf('%s = %u events',eof_str,sum(index)),'HorizontalAlignment','right','VerticalAlignment','bottom','fontsize',11);
             else
[…]

### 2022-03-29 pipeline/4-analyze-range-error/plot_post_range_error_nbc.m

@@ -82,18 +82,7 @@ for zr = [30 90]
         ylim([-12 31]);
         yticks(-10:5:30);
         xticks(1:0.2:2.2);
-        
-        %         if mod(count,3)~=1
-        %             yticklabels([])
-        %         else
-        %             ylabel('range error [m]');
-        %         end
-        %
-        %         if count >=4
-        %             xlabel('one way travel time [s]');
-        %         else
-        %             xticklabels([]);
-        %         end
+
         if count == 4
             xlabel('one way travel time [s]');
             ylabel('pseudorange error [m]');
@@ -116,4 +105,4 @@ title(lg,'Sound Speed Inputs');
 
 % title
 sgtitle('Pseudorange error footprints','fontsize',17,'fontweight','bold')
-h_printThesisPNG('range-error-allMethods')
+%h_printThesisPNG('range-error-allMethods')

### 2022-03-31 pipeline/1-overview/plot_ssp_comparison.m
changes to final revision

@@ -40,7 +40,9 @@ hold off
 grid on
 set(gca,'ydir','reverse');
 title('Sound speed estimates');
-ylim([0 600])
+ylim([0 550])
+ytick([0 30 100:100:600]);
+yticklabels({'0','30','100','200','300','400','500','600'});
 xlim([1431 1462])
 ylabel('depth [m]');
 xlabel('c [m/s]');
@@ -48,6 +50,11 @@ xlabel('c [m/s]');
 legend('ICEX20: Baseline','ICEX20: Chosen Weights','ICEX20: HYCOM','',...
        'ICEX16: Historical', 'ICEX16: Data','ICEX16: HYCOM',...
        'location','southwest','fontsize',13);
+   
+%text(1431,20,'20 ','horizontalalignment','right','verticalalignment','middle','fontsize',9);
+%text(1431,30,'30 ','horizontalalignment','right','verticalalignment','middle','fontsize',9);
+%text(1431,90,'90 ','horizontalalignment','right','verticalalignment','middle','fontsize',9);
+
    
 %% export "isovelocity" assumption
 

### 2022-03-31 pipeline/3-analyze-gvel/plot_gvel_by_txrx_depth.m

@@ -12,6 +12,8 @@ load isovelocity-ssp.mat
 listing = dir('../bellhop-gvel-gridded/csv_arr/*gridded.csv');
 [T2,colorSet] = h_get_nbc(listing,DATA,INDEX);
 
+impliedColor = [161, 183, 13]./256;
+
 %% load post-processing sim for MBC
 listing1 = dir('../bellhop-gvel-gridded/csv_arr/*old.csv');
 T1 = h_get_mbc(listing1,DATA);
@@ -43,10 +45,10 @@ for zr = [30 90]
         yline(iso.avg-iso.std,'k:','linewidth',2,'handlevisibility','off');
         
         % implied effective sound speed (GPS)
-        sgps = scatter(DATA.owtt(index),DATA.gvel(index),50,'o','filled','handlevisibility','off');
+        sgps = scatter(DATA.owtt(index),DATA.gvel(index),60,'o','filled','handlevisibility','off');
         sgps.MarkerEdgeColor = 'none';
-        sgps.MarkerFaceColor = [200, 78, 0]./256;
-        sgps.MarkerFaceAlpha = 0.2;
+        sgps.MarkerFaceColor = impliedColor;
+        sgps.MarkerFaceAlpha = 0.3;
         
         if sum(index)>=1
             % plot
@@ -131,7 +133,7 @@ end
 % legend for data
 plot(NaN,NaN,'w');
 plot(NaN,NaN,'w');
-plot(NaN,NaN,'.','color',[200, 78, 0]./256,'markersize',15);
+plot(NaN,NaN,'.','color',impliedColor,'markersize',15);
 
 lgdstr = {'\bf{SOUND SPEED SOURCE}','Isovelocity','HYCOM','Baseline','Chosen Weights','','\bf{MINIMAL BOUNCE CRITERIA}',...
     'Minimal bounce','','\bf{NEAREST BOUNCE CRITERIA}',...

### 2022-03-31 pipeline/9-trilateration/auv_trilat_hist.m

@@ -27,7 +27,7 @@ Y{2}.count = trilat.mbc.correction;
 Y{1}.count = trilat.nbc.correction;
 h_hist_boxplot(Y,edges,colors,max(edges));
 
-xlabel('RMS correction [m]')
+xlabel('correction [m rms]')
 set(gca,'fontsize',13)
 title(sprintf('Distribution of AUV re-navigation corrections (n=%u)',N),'fontsize',15)
 
@@ -41,7 +41,7 @@ Y{1}.count = trilat.nbc.error;
 Y{2}.count = trilat.mbc.error;
 
 h_hist_boxplot(Y,edges,colors,max(edges));
-xlabel('RMS error [m]')
+xlabel('error [m rms]')
 set(gca,'fontsize',13)
 title({'',...
     sprintf('Distribution of AUV re-navigation error (n=%u)',N)}...

### 2022-03-31 pipeline/9-trilateration/modem_trilat_hist.m

@@ -20,7 +20,7 @@ figure('name','trilat-histogram','renderer','painters','position',[108 108 1200 
 h_hist_boxplot(Y,edges,colors,48);
 
 %% dress up figure
-xlabel('RMS correction [m]')
+xlabel('correction [m rms]')
 set(gca,'fontsize',13)
 title(sprintf('Distribution of beacon re-positioning corrections (n=%u)',N),'fontsize',15);
 

### 2022-03-31 manuscript/manuscript.tex
overleaf text

--- 2022-03-29 manuscript/jasa-num-figs/manuscript.tex
@@
 % \documentclass[preprint,NumberedRefs]{JASA}
-
 %%%%%%% For Reprint
@@
 \title[JASA/draft]{Under-ice acoustic navigation using real-time model-aided range estimation}
-\author{EeShan C. Bhatt \footnote{Also at MIT-WHOI Joint Program in Oceanography/Applied Ocean Science \& Engineering, Cambridge, MA}}
+\author{EeShan C. Bhatt}
 \email{ebhatt@whoi.edu}
@@
 The long baseline (LBL) underwater navigation paradigm relies on the conversion of travel times into pseudoranges to trilaterate position. 
-For real-time operations, this conversion assumes an isovelocity sound speed.
+For real-time autonomous underwater vehicle (AUV) operations, this conversion assumes an isovelocity sound speed.
 For re-navigation, computationally and/or labor intensive acoustic modeling may be employed to reduce uncertainty.
@@
 \llabel{2.1} This method was implemented for an AUV-LBL system in the Beaufort Sea, in total ice-covered conditions and a double-ducted propagation environment.
-\llabel{2.2} Given the lack of GNSS data throughout the vehicle's mission, the pseudorange performance is first evaluated on connections between GNSS-linked beacons.
-The real-time ranging error between beacons is roughly 11 meters at distances up to 3 km.
+\llabel{2.2} Given the lack of Global Navigation Satellite System (GNSS) data throughout the vehicle's mission, the pseudorange performance is first evaluated on acoustic transmissions between GNSS-linked beacons.
+The mean real-time absolute range error between beacons is roughly 11 m at distances up to 3 km.
 A consistent overestimation in the real-time method provides insights for improved eigenray filtering by the number of bounces.
-An operationally equivalent pipeline is used to re-position the LBL beacons and re-navigate the AUV, using modeled, historical, and a locally observed sound speed profiles.
-The median re-navigation error is 1.84 $\pm$ 2.19 m.
-\llabel{2.3} The improved performance for suggests that this approach extends the single meter accuracy of the deployed GNSS units into the water column.
+An operationally equivalent pipeline is used to re-position the LBL beacons and re-navigate the AUV, using modeled, historical, and locally observed sound speed profiles.
+The best re-navigation error is 1.84 $\pm$ 2.19 m rms.
+\llabel{2.3} The improved performance suggests this approach extends the single meter accuracy of the deployed GNSS units into the water column.
 \end{abstract}
@@
 This foolproof method of self-positioning is undesirable for stealth, adverse weather conditions, and mission efficiency, and inaccessible in a GNSS-denied situation like an ice-covered environment.
-
 % introduce LBL and precise statement of work
-Of underwater acoustic navigation systems, long baseline (LBL) is the most GPS-like in style and scale, and most appropriate for mitigating drift without overburdening computation or payload size on the vehicle \citep{van_uffelen_global_2021}.
-The state-of-the-art for LBL outsources depth to a pressure sensor and solves the two-dimensional localization problem with an isovelocity, linear scaling between one way travel time (OWTT) and range \citep{Eustice2006,Eustice2007,webster_advances_2012,Webster2009}.
-This assumption is valid for short scale operations but oversimplifies propagation at larger scales or complex sound speed structure.
+Of underwater acoustic navigation systems, long baseline (LBL) is the most GNSS-like in style and scale, and most appropriate for mitigating drift without overburdening computation or payload size on the vehicle \citep{van_uffelen_global_2021}.
+The state-of-the-art for LBL outsources depth to a pressure sensor and solves the two-dimensional localization problem with an isovelocity, linear scaling between one-way travel time (OWTT) and range \citep{Eustice2006,Eustice2007,webster_advances_2012,Webster2009}.
+This assumption is valid for small-scale operations but oversimplifies propagation at larger scales or complex sound speed structure.
 To achieve single meter, GNSS-like performance in a GNSS-denied environment, we demonstrate an embedded ray-based data processing algorithm to convert recorded OWTTs into pseudorange estimates.
+An \textit{in situ} sound speed profile (SSP) is used despite the small operational domain because of the relatively high-risk mission environment\textemdash total under-ice conditions and a variable double-ducted acoustic environment.
 This methodology was integrated onto the AUV \emph{Macrura}, deployed and recovered in the Beaufort Sea, in March 2020, during the Ice Exercise 2020 (ICEX20).
-A physics-driven methodology that received an \textit{in situ} sound speed profile (SSP) was necessary despite the small operational domain because of the relatively high-risk mission environment\textemdash total under-ice conditions and a variable double ducted acoustic environment.
 
@@
 % More about LBL navigation
-While RAFOS floats championed one way ranging for re-positioning \citep{Rossby1986,duda_evaluation_2006}, the ability to do so for navigation was facilitated by \llabel{1.11} the advent of the WHOI Micro-Modem \citep{Singh2006} and synchronized clocks \citep{rypkema_one-way_2017}.
+While RAFOS floats championed one-way ranging for re-positioning \citep{Rossby1986,duda_evaluation_2006}, the ability to do so for navigation was facilitated by \llabel{1.11} the advent of the WHOI Micro-Modem \citep{Singh2006} and synchronized clocks \citep{rypkema_one-way_2017}.
 \llabel{1.3a}
-AUV navigation efforts have achieved root mean square (RMS) localization error on the order of tens of meters relative to GNSS surface position over less than ten kilometers in shallow \citep{Eustice2007,Claus2018,kepper_mems_2017} and deep water \citep{Kunz2008,jakuba_long-baseline_2008,Webster2009}.
+AUV navigation efforts have achieved root mean square (rms) localization error on the order of tens of meters relative to GNSS surface position over less than ten kilometers in shallow \citep{Eustice2007,Claus2018,kepper_mems_2017} and deep water \citep{Kunz2008,jakuba_long-baseline_2008,Webster2009}.
 However, these efforts all used a nominal sound speed for travel time conversion and the vehicles were limited to shallower isovelocity regimes.
@@
 These methods have also been reserved for post-processing as they can be labor intensive, computationally heavy, and/or require additional information like contemporaneous data.
-For example, gliders navigating with kinematic flight models and equipped with acoustic modems were later unambiguously associated with predicted ray arrivals, resulting in roughly a kilometer error and hundred meters uncertainty over basin scale propagation \cite{VanUffelen2013}.
-A follow up study investigated how a single temporally and spatially averaged SSP could mitigate position error for a four month glider mission \cite{vanuffelen2016localization}.
-\citet{Wu2019} cross correlate three days of real acoustic records with synthetic ones generated through ocean model snapshots from HYCOM \cite{chassignet_hycom_2007}.
-While potentially applicable for various ocean states, this is reliant on model realism and impractical for real-time operations.
-A ``cold start'' algorithm that does not require prior knowledge of track, position, or sound speed information inputs a four-dimensional ocean model, constrained by tomography data, into a range dependent ray code to isolate the last path detected in a full multipath pattern \citep{Mikhalevsky2020}.
-Then, a representative group speed is solved for alongside position in a least squares fashion. 
-This approach is able to re-position a floating hydrophone array with an error of 58 m and a standard deviation of 32 m based on six sources 129--450 km away but remains to seen for real-time navigation.
+For example, measured basin scale acoustic arrivals on gliders equipped with acoustic modems have been later unambiguously associated with predicted ray arrivals, differing by 914 m rms from flight model positions, with an estimated uncertainty of 106 m rms \cite{VanUffelen2013}.
+A follow up study investigated how accounting for glider velocity between acoustic receptions could mitigate position error during a four-month glider mission, but a single SSP was used \cite{vanuffelen2016localization}.
+\citet{Wu2019}, given three days of real acoustic records, generate synthetic records through ocean model snapshots, and cross correlate the two to estimate absolute range between a bottom-moored transmitter and a bottom-moored receiver with 150 depths.
+While potentially applicable for various deep ocean states, this is reliant on model realism and impractical for real-time operations.
+Lastly, a ``cold start'' algorithm that does not require prior knowledge of track, position, or sound speed information isolates the last path detected in a full multipath pattern \citep{Mikhalevsky2020}.
+Then, a representative group speed is solved for together with position in a least squares fashion. 
+This approach re-positioned a bottom-moored vertical hydrophone array with an error of 58 m and a standard deviation of 32 m based on six sources 129--450 km away; accuracy can be improved by incorporating a four-dimensional sound speed field from a general circulation model and this approach remains to be seen for navigation.
 
-The ICEX20 AUV deployment necessitated an environmentally- and physically-driven relationship between recorded travel times and estimated pseudoranges due to the multipath uncertainty brought upon by an increasingly observed double ducted environment in the Beaufort Sea, which some refer to as the ``Beaufort Lens'' \citep{litvak2015,chen_spectral_2019,chen_temporal_2020}.
+The ICEX20 AUV deployment necessitated an environmentally- and physically-driven relationship between recorded travel times and estimated pseudoranges due to the multipath uncertainty brought upon by an increasingly observed double-ducted environment in the Beaufort Sea, which some refer to as the ``Beaufort Lens'' \citep{litvak2015,chen_spectral_2019,chen_temporal_2020}.
 
@@
 \llabel{2.4}
-A neutrally buoyant layer of warm Pacific Summer Water creates a unique double ducted environment \textemdash the upper duct degrades signal coherence due to intensified ice interaction and the lower duct effectively traps sound for long range propagation \citep{Poulsen2016}.
+A neutrally buoyant layer of warm Pacific Summer Water creates a unique double-ducted environment \textemdash the upper duct degrades signal coherence due to intensified ice interaction and the lower duct effectively traps sound for long range propagation \citep{Poulsen2016}.
 Modeling output \citep{duda_long-range_2019,duda_effects_2021} and experimental observations \citep{Ballard2020,Bhatt2021} suggest that, in the Beaufort Sea, the duct is persistent and widespread but not necessarily continuous; it and its acoustic effects can be non-existent, minimal, or drastic.
-Transmissions in the upper duct, between the surface ice and the lens, experience minimal attenuation but degrade in signal coherence with repeated reflections under the ice.
+Transmissions in the upper duct, between the surface ice and the lens, degrade in signal coherence with repeated reflections under the ice.
 In the lower duct, between the lens and its conjugate depth in the Atlantic water (roughly 200 m), sound above 350 Hz is trapped near losslessly for long range propagation \citep{Poulsen2016}.
@@
 In 2017, gliders were deployed in a region with no ice coverage \citep{Graupe2019}.
-Ranges were linearly scaled by a statistical description of sound speed observations taken during the experiment, 1450 $\pm$ 6.5 m/s.
+Ranges were linearly scaled by a statistical description of SSPs taken during the experiment, 1450 $\pm$ 6.5 m/s.
 This resulted in an error of 550 m, which was reduced by a factor between 4 and 5, depending on the dive, using a post-processing acoustic arrival matching method.
@@
 
-Compared to previous small scale navigation efforts, the approach in this paper integrates real-time model-aided data processing to estimate a representative sound speed along a path from source to receiver, leveraging climatology, \textit{in situ} data, and fast acoustic modeling.
+Compared to previous small-scale navigation efforts, the approach in this paper integrates real-time model-aided data processing to estimate a representative sound speed along a path from source to receiver, leveraging climatology, \textit{in situ} data, and fast acoustic modeling.
 The paper is organized as follows.
 Section \ref{sec:icex20} details the experimental approach and conditions during ICEX20.
-Given that there is no GNSS ground truth for the vehicle position while underway, we first evaluate the real-time ranging performance of GPS-linked beacon-to-beacon communication events in section \ref{sec:realtime}.
+Given that there is no GNSS ground truth for the vehicle position while underway, we first evaluate the real-time ranging performance of GNSS-linked beacon-to-beacon communication events in section \ref{sec:realtime}.
 Section \ref{sec:post} uses insights from field data to introduce a new ray filtering algorithm to improve range estimation.
@@
 The experiment was conducted in the Beaufort Sea, from March 8th to 11th, at roughly 71.2\textdegree N.
-The AUV deployment was supported by the Integrated Communication and Navigation Network (ICNN) \citep{schneider_self-adapting_2020,Randeni2020,randeni_high-resolution_2021} a specialized implementation of the LBL solution.
+The AUV deployment was supported by the Integrated Communication and Navigation Network (ICNN) \citep{schneider_self-adapting_2020,Randeni2020}, a specialized implementation of the LBL solution.
 The ICNN was initially developed via numerous virtual experiments to ensure robust algorithms and interfaces between different hardware components. 
@@
 
-\begin{figure}[h!]
+\begin{figure}[!ht]
     \centering
     \includegraphics[width=0.8\columnwidth]{Fig1.pdf}
-    \caption{A schematic overview of the Integrated Communication and Navigation Network (ICNN), which provides joint data-transfer and tracking between AUV and a human decision maker at Topside.}
+    \caption{A schematic overview of the Integrated Communication and Navigation Network (ICNN), which provides joint data transfer and tracking between an AUV and a human decision maker at the topside camp.}
     \label{fig:icnnOverview}
@@
 The ICNN is comprised of four ice buoys in a loose rectangle, roughly 2 km away from a central ice camp with a topside computer, as shown in Fig.~\ref{fig:icnnOverview}.
-Each ice buoy is outfitted with a Garmin GPS 18x, with a pulse-per-second rising edge aligned to 1 microsecond and a spec sheet accuracy of 3 m, 95\% of the time.
-They are also each equipped with a WHOI Micro-Modem \citep{Singh2006}, with a four-element receiver array, a single transmitter, and one-tenth of a millisecond resolution.
-\llabel{1.6a} Acoustic messages were sent with a 10 kHz carrier frequency, 5 kHz bandwidth, and phase-shift keying (PSK) modulation on a time-division multiple access schedule with a thirty-second cycle, giving room for two-way communication throughout the mission volume.
+Each ice buoy is outfitted with a Garmin GPS 18x, with a pulse-per-second rising edge aligned to 1 microsecond and a specified accuracy of 3 m, 95\% of the time.
+They are also each equipped with a WHOI Micro-Modem \citep{Singh2006,gallimore2010whoi}, with a four-element receiver array, a single transmitter, and a real-time clock featuring a precision of 2 ppm and a resolution of one-tenth of a millisecond.
+\llabel{1.6a} Acoustic messages were sent with a 10 kHz carrier frequency, 5 kHz bandwidth, and phase-shift keying (PSK) modulation on a time-division multiple access schedule with a thirty-second cycle, giving room for two-way communication throughout the operational volume.
 Thus the ICNN is dependent on the successful decoding of acoustic transmissions.
 The receive and transmit elements were split between shallow and deeper depths\textemdash 30 and 90 m\textemdash to provide better coverage across the shadow zone.
-While each buoy only has one transmit depth, all buoys have both receive depths but the active receive layer is consistent across all buoys. \llabel{1.12}
+While each buoy only has one transmit depth, all buoys have both receive depths but the depth of the active receive layer is synchronized across all buoys. \llabel{1.12}
 The design of the ICNN enables a self-adapting network to transmit and receive at the optimal depth to maintain contact with the AUV \citep{schneider_self-adapting_2020}.
@@
 \begin{enumerate}
-\item The AUV, running an ice-tracking DVL and an onboard hydrodynamic model, broadcasts its perceived location on a scheduled, time-synchronized message via WHOI Micro-Modem
-\item Four ice buoys, each outfitted with a WHOI Micro-Modem, receive messages from the AUV and send that information over freewave radio to a Topside computer
-\item The topside computer converts travel times into pseudorange estimates using a stochastic embedded prediction of the horizontal group velocity via BELLHOP ray tracing code \citep{Porter2011} using a sound speed profile provided by an updatable Virtual Ocean \citep{schneider_netsim_2018,bhatt_embedded_2022}
+\item The AUV, running an ice-tracking doppler velocity log (DVL) and an onboard hydrodynamic model, broadcasts its perceived location on a scheduled, time-synchronized message via WHOI Micro-Modem
+\item Four ice buoys, each outfitted with a WHOI Micro-Modem, receive messages from the AUV and send that information over freewave radio to a topside computer
+\item The topside computer converts travel times into pseudorange estimates using a stochastic embedded prediction of the effective horizontal sound speed via BELLHOP ray tracing code \citep{Porter2011} using a sound speed profile provided by an updatable Virtual Ocean \citep{schneider_netsim_2018,bhatt_embedded_2022}
 \item The topside computer calculates a new position by trilaterating the range estimates
-\item The position differential, not the absolute position, is broadcast to the vehicle to update its navigation solution and be robust to latency and intermittency
+\item To combat latency and intermittency, the position differential, not the absolute position, is broadcast from a 20 m source at topside to update the vehicle's navigation solution
 \end{enumerate}
 
-In the face of GNSS-denied navigation, this approach was anecdotally successful, as shown in Fig. \ref{fig:vehicleRecovery}.
-The AUV \emph{Macrura} was deployed through a hydrohole from an ice camp but recovered through an emergency hydrohole.
+The AUV \emph{Macrura} was deployed through a hydrohole from an ice camp but recovered through an emergency hydrohole, as shown in Fig. \ref{fig:vehicleRecovery}.
 A random disk error stalled \emph{Macrura} underneath the ice but did not prevent it from transmitting its location.
-Due to an incoming storm, a team drilled an exploratory hole at \emph{Macrura's} self-reported location, and were able to drill a separate hole 1 m away to tie the vehicle off to a physical marker on the ice.
-Three days later, \emph{Macrura} was recovered\textemdash the ice camp had moved over 19 km.
-The AUV's relative position to camp had also changed, from roughly 45\textdegree at 1000 m to 90\textdegree at 1100 m.
-Drama aside, we view the emergency recovery as qualitative proof of the robustness of this navigation approach.
-Nonetheless, this paper specifically addresses the third and fourth steps\textemdash the conversion of travel times into pseudoranges and its quantative effect on trilateration.
-By focusing on pseudorange estimates between GPS-tracked beacons, and re-running the trilateration pipeline, the results are decoupled from all other mechanisms in the ICNN.
+An exploratory hole at \emph{Macrura's} self-reported position confirmed its presence.
+Due to an incoming storm, the vehicle was tied off to a physical marker on the ice, and three days later, \emph{Macrura} was recovered.
+The ice camp had moved roughly 19 km and the AUV's relative position to camp had also changed, from roughly 45\textdegree at 1000 m to 90\textdegree at 1100 m.
+While the emergency recovery serves as qualitative proof of navigation, this paper specifically addresses the third and fourth steps\textemdash the conversion of travel times into pseudoranges and its quantitative effect on trilateration.
+By focusing on pseudorange estimates between GNSS-tracked beacons, and re-running the trilateration pipeline, the results are decoupled from all other mechanisms in the ICNN.
 
-\begin{figure}[h!]
+\begin{figure}[!ht]
     \centering
-    \includegraphics[width=0.7\columnwidth]{Fig2a.pdf} \hfill
-    \includegraphics[width=0.28\columnwidth]{Fig2b.jpg}
-    \caption{The under-ice mission track for AUV \emph{Macrura}, including the position updates as it stalled underneath the ice overnight.
+    \includegraphics[width=0.7\columnwidth,trim={0 100 0 0},clip]{Fig2a.pdf} \hfill
+    \includegraphics[width=0.25\columnwidth]{Fig2b.jpg}
+    \caption{The under-ice mission track for AUV \emph{Macrura} (left), including the position updates as it stalled underneath the ice overnight (middle).
 A marker was placed on the ice at the vehicle's estimated self-location.
-It was recovered after a three day storm within a meter of the marker.}
+It was recovered after a three day storm within a meter of the marker (right).}
     \label{fig:vehicleRecovery}
@@
 
-An important component to our navigation solution is an accurate estimation of a representative SSP for the ocean volume.
-Previous field experience, during the Ice Exercise 2016 (ICEX16), demonstrated the negative effects of the Beaufort Lens on tracking and communication \citep{schmidt_acoustic_2016}.
+An important component to the navigation solution is an accurate estimation of a representative SSP for the ocean volume.
+Previous field experience, during Ice Exercise 2016 (ICEX16), demonstrated the negative effects of the Beaufort Lens on tracking and communication \citep{schmidt_acoustic_2016}.
 Fig. \ref{fig:sspExpectation} shows historical, modeled, and \textit{in situ} sound speed data for both ICEX16 and ICEX20.
@@
 In the field, the SSP information was shared with the vehicle via basis representation compression on a lightweight digital acoustic message \citep{bhatt_embedded_2022}. \llabel{2.7}
-All modeled data comes from HYCOM \cite{chassignet_hycom_2007}, which does not seem to capture the forcing mechanisms that cause the Beaufort Lens.
-For ICEX16, the data-driven profile was sourced from nearby Ice Tethered Profilers (ITP) after the field experiment \cite{Krishfield2008,toole_ice-tethered_2011} and exhibits a fairly deep lens; the historical profile is from the World Ocean Atlas.
+All modeled data comes from the Hybrid Coordinate Ocean Model (HYCOM) \cite{chassignet_hycom_2007}, which does not seem to capture the forcing mechanisms that cause the Beaufort Lens.
+For ICEX16, the data-driven profile was sourced from nearby Ice Tethered Profilers (ITP) after the field experiment \cite{Krishfield2008,toole_ice-tethered_2011} and exhibits a fairly deep lens; the historical profile is from the World Ocean Atlas \cite{woa09}.
 For ICEX20, the chosen weights (data-driven) profile derives from initial CTD casts taken on site, showing an intense warm water intrusion; the baseline (historical) profile, showing moderate ducted conditions, comes from the average of March 2013 ITP data.
@@
 
-\begin{figure}[h!]
+\begin{figure}[!ht]
     \centering
     \includegraphics[width=\reprintcolumnwidth]{Fig3.pdf}
-    \caption{Sound speed conditions for historical (baseline), data (chosen weights), and HYCOM model for both ICEX16 and ICEX20.}
+    \caption{Sound speed profiles for historical (baseline), data (chosen weights), and HYCOM model for both ICEX16 and ICEX20.}
     \label{fig:sspExpectation}
@@
 For post-processing comparison, we introduce both the HYCOM profile and an isovelocity case, 1441.8 $\pm$ 3.7 m/s, as the mean and standard deviation of the observed sound speed profile over the first 200 m.
-This is a contrived value taken in the style of \citet{Graupe2019} for the sake of comparison; the default value in the LAMSS simulator, which was not environmentally informed and used when no updates were available from the ICNN, was 1430 m/s.
+This is a contrived value taken in the style of \citet{Graupe2019} for the sake of comparison; the default value in the ICNN, which was not environmentally informed and used when no updates were available, was 1430 m/s.
 
@@
 
-Because the vehicle's navigation solution during a mission can only be evaluated on the basis of the error estimates sent, a sister experiment for validating the real-time ranging approach was implemented.
-Ice buoy modems were run as ``virtual vehicles'' at a fixed depth, receiving position updates from the other beacons as well as a camp site modem lowered to 20 m. %\llabel{1.13}
+Because the vehicle's navigation solution during a mission can only be evaluated on the basis of the error estimates sent, a separate experiment for validating the real-time ranging approach was conducted.
+Ice buoy modems were run as ``virtual vehicles'' at a fixed depth, receiving position updates from the other beacons as well as from a camp site modem lowered to 20 m. %\llabel{1.13}
 Fig. \ref{fig:overview} shows successful events sorted by source depth.
@@
 
-\begin{figure}[h!]
+\begin{figure}[!ht]
   \centering
@@
 
-Instead, for each individual receiver $i$, an embedded stochastic tracking framework is used to provide a running estimate of the effective sound speed $c_{i,j}$ for the conversion from travel time to range from modem $j$, with the ultimate goal of matching the implied horizontal effective sound speed, i.e., the GPS-recorded distance between two nodes divided by the modem-recorded one way travel time between them.
+Instead, for each individual receiver $i$, an embedded stochastic tracking framework is used to provide a running estimate of the effective sound speed $c_{i,j}$ for the conversion from travel time to range from modem $j$, with the ultimate goal of matching the implied horizontal effective sound speed, i.e., the GNSS-recorded distance between two nodes divided by the modem-recorded OWTT between them.
 
 \llabel{2.10} In the ICEX20 configuration, the acoustic tracker is running on the topside computer, which controls the ICNN.
-Here we assume that the effective sound speeds $c_{i,j}$ are smoothly varying over the course of a vehicle mission, i.e., with respect to range from signal origin at transmitter $j$, mission time, and the thirty-second frequency.
+Here we assume that the effective sound speeds $c_{i,j}$ are smoothly varying over the course of a vehicle mission, i.e., with respect to range from signal origin at transmitter $j$, mission time, and the thirty-second interval between transmissions.
 
-\llabel{1.15} When the topside tracking framework receives a message, with a time delay, $\Delta t$, it will request a new estimate for $c_{i,j}$ along with its standard deviation.
-The effective sound speed is predicted using the vehicle's reported depth and the extrapolated navigation solution for range, $\hat{r}$, as inputs to the ray tracing program, which returns an impulse response estimate in the form of ray travel times $dt_{j}$ and amplitudes $a_{j}$.
+\llabel{1.15} When the topside tracking framework receives a message, with a time delay, $\Delta t$, it requests a new estimate for $c_{i,j}$ and its standard deviation.
+The effective sound speed is computed using the vehicle's reported depth and the extrapolated navigation solution for range, $\hat{r}$, as inputs to the ray tracing program, which returns an impulse response estimate in the form of ray travel times $dt_{j}$ and amplitudes $a_{j}$.
 
@@
 It is initialized as 11 $\times$ 11 points spanning 10 m horizontally and 20 m vertically.
-The horizontal dimension reflects the accumulated vehicle position error given a thirty-second communication cycle; the vertical dimension reflects how, computationally, eigenrays of the same timefront seem to stack vertically in the water column.
+The horizontal dimension reflects the accumulated vehicle position error given a thirty-second communication cycle; the vertical dimension reflects how, computationally, eigenrays of the same time front seem to stack vertically in the water column.
 \llabel{1.17}
-For each grid point, BELLHOP produces a number of arrivals resulting from multiple propagation paths.
+For each grid point, BELLHOP produces several arrivals resulting from multiple propagation paths.
 Using only the $N_0$ rays with neither surface nor bottom bounces, the tracking system will then estimate the current effective sound speed $c$ from a power weighted average of the ray travel times,
@@
 \begin{equation}
-\sigma_{c} \simeq \sqrt{\frac {\sum_{n=1}^{N_{0}} (dt_{n}-\hat{r}/u)^{2}a_{n}^{2}}{ \sum_{n=1}^{N_{0}} a_{n}^{2}} } \frac{u^{2}}{\hat{r}}
+\sigma_{c} \simeq \sqrt{\frac {\sum_{n=1}^{N_{0}} (dt_{n}-\hat{r}/c)^{2}a_{n}^{2}}{ \sum_{n=1}^{N_{0}} a_{n}^{2}} } \frac{c^{2}}{\hat{r}}
 \end{equation}
@@
 Thus the minimal bounce (MB) criteria assumes the signal detected by the modem will be dominated by a set of paths with the least number of boundary interactions.
-Importantly, this stochastic, ensemble method for group velocity calculation can run in real-time, appearing to be orders of magnitude faster than other post-processing methods which seek to determine the specific ray itself that best matches a prominent indicator from the arrival structure.
-The BELLHOP simulation that runs this calculation uses 3600 rays with launch angle fan of -60 to 60 degrees, a representative depth dependent sound speed profile, and a range dependent bathymetry.
+Importantly, this stochastic, ensemble method for effective sound speed calculation is automated and computationally lightweight for real-time use.
+The BELLHOP simulation that runs this calculation uses 3600 rays with launch angle fan of -60\textdegree to 60\textdegree, a representative depth-dependent sound speed profile, and a range-dependent bathymetry.
 \llabel{1.13}
@@
 
-The sister modem experiment generated 811 beacon to beacon communication events with their own effective sound speed predictions from a real-time implementation of the MB method.
+The modem experiment generated 811 beacon-to-beacon communication events with effective sound speed predictions from a real-time MB implementation.
 Given the complexity of the ICNN system, this experiment did not collect an exhaustive set of data across all buoy, source depth, receive depth, and model sound speed combinations.
-\llabel{1.18} The algorithm generally overestimates pseudoranges because it resolves the effective sound speed for the most direct path.
 
-\begin{figure}[h!]
+\begin{figure}[!ht]
   \centering
   \includegraphics[width=\columnwidth]{Fig5.pdf}
-  \caption{The real-time range error by source (rows: 20, 30, and 90 m) and receiver (columns: 30 and 90 m) pairings for both sound speed estimates used during ICEX20.
-The amount of communication events is notated in the top right of each panel.
-The dashed line indicates no range error, and the boundary drawn indicates scope of the range error as a function of one way travel time.}
+  \caption{The real-time range pseudorange error relative to GNSS-derived range by source (columns: 20, 30, and 90 m) and receiver (rows: 30 and 90 m) pairings for both sound speed estimates used during ICEX20.
+The number of transmissions is noted in the top right of each panel under the source and receiver depths.
+The dashed line indicates no range error, and the boundary drawn indicates scope of the range error as a function of one-way travel time.}
   \label{fig:range-error-insitu}
@@
 Fig. \ref{fig:range-error-insitu} shows the range error boundary for both SSP inputs used in ICEX20.
-A promising sign that the MB method adapts sound speed somewhat intelligently is the lack of error growth as travel time increases.
+The algorithm generally overestimates pseudoranges because it resolves the effective sound speed for the most direct path.
 \llabel{1.28}
@@
 The overarching results show that sound speed estimates derived from eigenrays for a local grid, as opposed to a singular point, are accurate enough to support vehicle navigation.
+A promising sign that the MB method adapts sound speed somewhat intelligently is the lack of error growth as travel time increases.
 While the MB looks for just the least complex multipath, the high density of launch angles almost always guarantees a direct path for the beacon-to-beacon configurations.
@@
 Accounting for ice movement between beacons creates nominal ranges with small variability.
-Figs. \ref{fig:raytrace-zs20}, \ref{fig:raytrace-zs30}, and \ref{fig:raytrace-zs90} show eigenrays for three sound speed environments for source depths of 20, 30, and 90 m, respectively.
-Eigenrays were initially found using the built-in BELLHOP protocol with a launch angle fan of 2400 rays between -60 and 60 degrees.
-Separately, recorded travel times between beacons were clustered with 1 millisecond boundaries such that some source-receiver pairs had multiple, distinct travel times to approximate.
-The BELLHOP eigenray returns\llabel{2.8} were then filtered such that one was selected per travel time cluster, in the hopes that the eigenray will converge to the receiver locations for the most realistic sound speed input.
-\llabel{1.16}
-It should be noted that bottom bounces were recovered but filtered out.
+Figs. \ref{fig:raytrace-zs20}, \ref{fig:raytrace-zs30}, and \ref{fig:raytrace-zs90} show eigenrays for three SSPs, for source depths of 20, 30, and 90 m, respectively.
+Eigenrays were initially found using the built-in BELLHOP protocol with a launch angle fan of 2400 rays between -60\textdegree and 60\textdegree.
+Separately, recorded travel times between beacons were clustered with 1 millisecond boundaries such that some source-receiver pairs had multiple distinct travel times.
+The BELLHOP eigenray returns\llabel{2.8} were then filtered such that one was selected per travel time cluster per SSP; \llabel{1.16} bottom bounces were recovered but filtered out.
 The three source depths create distinct ray geometries with respect to the three sound speed inputs.
@@
   \caption{Eigenrays for beacon-to-beacon events for each sound speed with a nominal source depth of 20 m over a total ray fan in gray.
-The beacons are highlighted in color/marker coding in Fig.
+The Beaufort Lens strengthens from HYCOM to baseline to chosen weights SSP.
+The beacons are highlighted in the color/marker encoding from Fig.
 \ref{fig:overview}.}
@@
 
-For a source at 20 m in depth, shown in Fig. \ref{fig:raytrace-zs20}, reliable eigenrays are found for all sound speed inputs.
-Rays refract upwards and neatly intersect with the shallow and deep receiver locations between 1.5 and 1.8 km in range.
+For a source at 20 m depth, shown in Fig. \ref{fig:raytrace-zs20}, reliable eigenrays are found for all sound speed inputs.
+Rays refract upwards and intersect with the shallow and deep receiver locations between 1.5 and 1.8 km in range.
 However, the ray paths for the shallow receivers change both in the number of surface interactions and where the surface interactions occur with respect to range across the SSPs.
@@
 The weighted SSP shows the most interesting effects for the deeper receivers.
-The ray paths all interact with the surface and the eigenrays for the Northern (red) and Western (green) buoys are in fact the same ray.
+The ray paths all interact with the surface, and the eigenrays for the Northern (red) and Western (green) buoys are in fact the same ray.
 
@@
   \includegraphics[width=\columnwidth]{Fig7.pdf}
-  \caption{Eigenrays for beacon-to-beacon events for each sound speed with a nominal source depth of 30 m over a total ray fan in gray.
-The beacons are highlighted in color/marker coding in Fig.
-\ref{fig:overview}.}
+  \caption{Same as Fig. \ref{fig:raytrace-zs20} except for a nominal source depth of 30 m.}
   \label{fig:raytrace-zs30}
@@
 
-The ray geometries from the 30 m source, shown in Fig. \ref{fig:raytrace-zs20}, show a similar degradation of eigenray identification with increased ducting.
+The ray geometries from the 30 m source, in Fig. \ref{fig:raytrace-zs30}, show increased ducting and a corresponding degradation of eigenray identification.
 Receptions span 1.5 to 3.2 km.
-Once again, eigenrays for HYCOM and the baseline SSP are visually appropriate.
-Rays for the weighted SSP show how the surface channel intensifies ice interactions and how the shadow zone denies reliable acoustic paths.
-Pointedly, the increasing number of surface reflections to the farthest shallow buoy (North, in red) crystallize the MB criteria's tendency for overestimation.
-For the HYCOM, baseline, and weighted SSP inputs, the most appropriate eigenrays show 2, 3, and 4 surface interactions.
+Eigenrays for HYCOM and the baseline SSP intersect with both shallow and deep receivers.
+Eigenrays for the weighted SSP show how the surface channel intensifies ice interactions for shallow receivers and how the shadow zone denies reliable acoustic paths to the deeper receivers.
+The increasing number of surface reflections to the farthest shallow buoy (North, in red) demonstrates the MB criteria's tendency for overestimation.
+For the HYCOM, baseline, and weighted SSP inputs, the most appropriate eigenrays show two, three, and four surface interactions.
 
@@
   \includegraphics[width=\columnwidth]{Fig8.pdf}
-  \caption{Eigenrays for beacon-to-beacon events for each sound speed with a nominal source depth of 90 m over a total ray fan in gray.
-The beacons are highlighted in color/marker coding in Fig.
-\ref{fig:overview}.}
+  \caption{Same as Fig. \ref{fig:raytrace-zs20} except for a nominal source depth of 90 m.}
   \label{fig:raytrace-zs90}
@@
 
-Lastly, Fig. \ref{fig:raytrace-zs90} shows ray geometries from the 90 m source, elucidating a different extent of the shadow zone. 
-While the receiver locations are similar to that of the 30 m source depth, the deeper source depth effectively negates the upper duct and places the upper (and some of the lower) receivers in unreliable acoustic paths.
-The HYCOM eigenrays still show the most reliable acoustic paths that deteriorate with increasing ducted conditions.
-The lack of direct paths from the observed SSP further points out the shortcomings of the MB approach.
+Lastly, Fig. \ref{fig:raytrace-zs90} shows ray geometries from the 90 m source, uncovering a different shadow zone extent.
+While the receiver locations are similar to those from the 30 m source depth, the deeper source depth effectively negates the upper duct and places the upper (and some of the lower) receivers in unreliable acoustic paths.
+The HYCOM eigenrays show the most reliable acoustic paths, but these deteriorate with increasing ducted conditions.
+The lack of direct paths for the observed SSP further points out the shortcomings of the MB approach.
 
-The goal of the MB algorithm was to provide a reliable, physically intuitive interpretation of the acoustic propagation without taking on the additional burden of regularly identifying specific paths that may connect any given source-receiver pair in the network.
-While it was unlikely to resolve multipath arrivals that triggered successful modem detection, an isovelocity approach would have provided no adaptivity against source and receiver depth differences.
-\llabel{2.14}
-Its performance was adequate for vehicle navigation and would have likely sufficed if it were not for the prominence of the duct observed relative that of other model and data products.
+The goal of the MB algorithm was to provide a reliable, physically intuitive interpretation of the acoustic propagation, capturing the information provided by the acoustic model without taking on the additional burden of identifying specific eigenrays that may connect any given source-receiver pair.
+The MB algorithm exploited the effect of source and receiver depth in the acoustic model, and its performance was sufficiently adequate for vehicle navigation.
+However, the algorithm assumes that the most likely detected arrival is the most direct path modeled, which was not generally the case under the observed ducted conditions.
 
@@
 
-From all events recorded during the modem test experiment, there are 1242 successfully decoded beacon-to-beacon events.
+From all events recorded during the modem test experiment, there are 1,242 successfully decoded beacon-to-beacon events.
 Thus, a post-processing analysis that emulates the real-time navigation engine was run to overcome the unequal distribution of communication events with respect to depth, range, and sound speed status.
 
-It is important to note that the value for the extrapolated range, $\hat{r}$, is only tracked by topside for a modem claiming to be the vehicle; thus we replace $\hat{r}$ with the GPS-tracked range for all modem events.
-Because $\hat{r}$ converges to the correct solution, a comparison of $\hat{r}$ with the GPS-tracked range shows a normal, zero-centered distribution within the bounds of GPS drift.
+It is important to note that the value for the extrapolated range, $\hat{r}$, is only tracked by topside for a modem claiming to be the vehicle.
+In this section, $\hat{r}$ is replaced with the GNSS-tracked range for all modem events.
 The analysis therefore seeds realistic but ``omniscient'' knowledge of the extrapolated range and emulates the post-processing pipeline to more thoroughly evaluate the acoustic pseudorange estimate for all modem events.
-Sound speed inputs are the isovelocity sound speed in addition to the modeled, baseline, and weighted SSPs from Fig. \ref{fig:sspExpectation}.
+Sound speed inputs are the isovelocity sound speed in addition to the ICEX20 modeled, baseline, and weighted SSPs from Fig. \ref{fig:sspExpectation}.
 The analysis replicates the MB criteria but also introduces a new filtering algorithm, the nearest bounce (NB), based on insights gleaned from the eigenray analysis.
-Accordingly, the results in this section evaluate the utility of the algorithms and sound speed sources, divorced from their role in the ICNN while maintaining real-time relevance.
+Accordingly, the results in this section evaluate the utility of the algorithms and SSP sources, divorced from their role in the ICNN while maintaining relevance for real-time computation.
 
@@
 
-\llabel{2.12} The extent of ray bending and repeated reflections is extremely dependent on the degree of the Beaufort Lens observed.
+\llabel{2.12} The extent of ray bending and repeated reflections is dependent on the observed Beaufort Lens.
 Based on this insight, a new algorithm, the nearest bounce (NB) criteria, is a slight modification from the MB and includes multipath as a new dimension of information to exploit.
@@
 Given a running estimate for the effective sound speed $c_{i,j}$ between nodes $i$ and $j$, the navigation system has an extrapolated value for range, $\hat{r}$, and a recorded travel time, $\Delta t_{i,j}$.
-Instead of using only the $N_0$ rays with neither surface nor bottom bounces to estimate conversion speed, and subsequently moving to incremental number of bounces only when no valid direct path solutions exist, we solve for the power weighted average of the ray travel time for the $N_k$ rays with $k$ bounces,
+Instead of using only the $N_0$ rays with neither surface nor bottom bounces to estimate effective sound speed and iterating on additional bounces only if no valid direct path solutions exist, we solve for the power-weighted average of the ray travel time for the $N_k$ rays with $k$ bounces,
 \begin{equation}
@@
 \end{equation}
-predict an effective sound speed,
+compute an effective sound speed,
 \begin{equation}
@@
 Then a single value is selected in a nearest-neighbor fashion that best matches the recorded travel time, as the detected arrival is not always the first arrival or the direct path and could even be masked by noise or blocked temporarily \citep{deffenbaugh_acoustic_1996}.
-We manually cap the number of bounces at four because of the smaller operational scale and the attenuation accrued with many surface interactions.
-Bottom bounces are not encoded separately because of ray's tendency to refract upward, not due to information limitations.
+The number of bounces is limited to four because of the small operational scale and the attenuation accrued with many surface interactions.
+Bottom bounces are not encoded separately because the ray paths refract upward, not due to information limitations.
 
@@
 The minimal and nearest bounce algorithms are applied with the three sound speed inputs shown in Figs.~\ref{fig:raytrace-zs20},~\ref{fig:raytrace-zs30},~\ref{fig:raytrace-zs90}.
-The resulting predicted effective sound speeds are shown in Fig. \ref{fig:gvel-post} for all source depths versus one way travel time.
+The resulting predicted effective sound speeds are shown in Fig. \ref{fig:gvel-post} for all source depths versus one-way travel time.
 
-\begin{figure}[h!]
+\begin{figure}[!ht]
 \includegraphics[width=\columnwidth]{Fig9.pdf}
@@
 The recorded travel time is on the x-axis and the predicted effective sound speed is on the y-axis.
-The sound speed source is indicated by color.
-The isovelocity is shown as the mean $\pm$ the standard deviation.
-The minimal and nearest bounce criterion are distinguished by different marker shapes, compared to the separately colored red dots showing the implied calculation.}
+The sound speed source is indicated by color and the isovelocity case is shown as the mean $\pm$ the standard deviation.
+The effective sound speed using the minimal and nearest bounce criterion are distinguished by different marker shapes, compared to the separately colored green dots showing that from the implied calculation.}
 \label{fig:gvel-post}
@@
 
-The goal of the effective sound speed prediction is to converge towards the implied sound speed, i.e. the GNSS-derived range divided by the recorded travel time.
+The goal of the effective sound speed prediction is to converge towards the implied sound speed, i.e., the GNSS-derived range divided by the recorded OWTT.
 As the environmental and ray filtering method become better representations of the real ocean, the lower the expected mismatch is between the implied and estimated effective sound speeds.
@@
 The various sound speed inputs\textemdash isovelocity aside\textemdash not only modify the predicted effective sound speed, as seen by the colored vertical offsets, but often classify a distinct number of bounces.
-HYCOM sees the most direct and one bounce multipath structures, lending a bias for faster speeds; the weighted SSP sees the most double and triple bounces, favoring slower speeds; the baseline sound speed exists in between.
+HYCOM SSP fosters the most direct and one bounce multipath structures, lending a bias for faster speeds; the weighted SSP fosters the most double and triple bounces, favoring slower speeds; the baseline SSP exists in between.
 Very rarely is the multipath structure classified as a direct path, i.e., where the NB defaults to the MB prediction.
-In fact, the higher the multipath classification, the more accurate the sound speed prediction is, likely driven by a tighter or even sparser bundle of rays.
-\llabel{1.19} Discontinuities in multipath classification provide initial evidence for its importance to a smoothly varying group velocity, as shown in the cluster of 30 to 30 m transmissions in Fig.
-\ref{fig:gvel-post}, where HYCOM jumps from one to two classified bounces amidst the baseline SSP and weighted SSP smoothly increasing while consistently seeing two and three classified bounces, respectively.
+In fact, the higher the multipath classification, the more accurate the sound speed prediction is, likely driven by a tighter (or smaller) bundle of rays.
+\llabel{1.19} Discontinuities in multipath classification provide initial evidence for its importance to a smoothly varying effective sound speed, as shown in the cluster of 30 to 30 m transmissions in Fig.
+\ref{fig:gvel-post}, where HYCOM results jump from one to two classified bounces while the baseline SSP and weighted SSP results smoothly increase and exhibit two and three classified bounces, respectively.
 Of course, the prediction deteriorates with cross-layer transmissions across the duct, but not to the same degree at which eigenrays could not be found for the weighted SSP in section \ref{sec:eigenrays}.
@@
 The isovelocity contrived for this paper, 1441.8 m/s, best matches the transmissions from 90 to 90 m.
-The one from \citet{Graupe2019}, 1450 m/s, would have had a systemic overestimation.
-In addition, over the course of the four day experiment, the local maxima of the Beaufort Lens changed from roughly 1447 m/s at 40 m to 1442 m/s at 60 m.
-Given that implied sound speeds just for beacon-to-beacon events span 1420 to 1445 m/s, it is safe to say that a nominal sound speed would sacrifice pseudorange accuracy somewhere, and that an adaptive approach is necessary even for short and/or small scale operations in the Beaufort Lens.
+Over the course of the four day experiment, the local maxima of the Beaufort Lens changed from roughly 1447 m/s at 40 m to 1442 m/s at 60 m.
+Given that implied sound speeds just for beacon-to-beacon events span 1420 to 1445 m/s, it is safe to say that a nominal sound speed would sacrifice pseudorange accuracy somewhere, and that an adaptive approach is necessary even for short duration and/or small-scale operations.
 
@@
 \includegraphics[width=\columnwidth]{Fig10.pdf}
-\caption{The post-processed pseudorange error; the rows share receiver depth; the columns share source depth.
+\caption{The post-processed pseudorange error relative to GNSS-derived ranges for all beacon-to-beacon events.
+The rows share receiver depth; the columns share source depth.
 The dashed gray line shows no error.
-The shaded region connects the range performance across all events.}
+The boxed regions connect the pseudorange error across all events.}
 \label{fig:rangeError}
@@
 
-The weighted SSP range error generally has the smallest and most zero-centered footprint.
-The one case it does not is for the source-receiver pairings between 30 and 90 m in depth.
-The increased error for these is most likely driven by the computational artifacts encountered when propagating through the steep sound speed gradients of the lens and through the shadow zone.
-All other source depth pairings are significantly improved using the chosen weights compared to HYCOM or the baseline.
+The weighted SSP range error generally has the smallest and most zero-centered footprint compared to HYCOM or baseline SSP, except for cross-layer source-receiver pairings between 30 and 90 m in depth.
+The increased error for these is most likely driven by the computational artifacts encountered by propagating through or near the shadow zone (see Fig. \ref{fig:raytrace-zs30}).
+In comparison, source-receiver pairings between 20 and 90 m in depth are closer in range and just outside or on the edge of the shadow zone (see Fig.
+\ref{fig:raytrace-zs20}), such that the corresponding ray fan is dense enough to resolve eigenrays.
+All other source depth pairings are significantly improved using the chosen weights compared to HYCOM or baseline SSP.
 
-When using a linear scaling to convert travel time into range, any offset between the assumed sound speed and the horizontal group velocity produces unconstrained error with increasing receiver distance, whereas an adaptive estimate would exhibit no such trend.
+When using a range-independent scaling to convert travel time into range, any offset between the assumed and actual sound speed produces unconstrained error with increasing receiver distance, whereas an adaptive estimate should exhibit no such trend.
 This is easily observed in the same-layer links, i.e., 30 to 30 m and 90 to 90 m.
-In cross-layer links, the isovelocity does not perform better but tends to exaggerate or flip the footprint created adaptively.
+In cross-layer links, the isovelocity case tends to exaggerate or flip the footprint created adaptively.
 
-The improvement from MB to NB is most evident for the data-driven sound speed; while the HYCOM SSP improves from a median absolute range error of 6.41 to 4.61 m, the baseline SSP improves from 10.30 to 2.27 m, and the weighted SSP improves from 13.28 to 2.12 m.
-In comparison, the isovelocity has a median error of 13.09 m.
-The order of magnitude improvement in the ducted SSPs demonstrate the effectiveness of the NB algorithm exploiting the observed multipath conditions.
+The improvement from MB to NB is most evident for the data-driven sound speed; while the HYCOM SSP median absolute range error improves from 6.41 to 4.61 m, the baseline SSP one improves from 10.30 to 2.27 m, and the weighted SSP one improves from 13.28 to 2.12 m.
+In comparison, the isovelocity SSP has a median error of 13.09 m.
+The order of magnitude improvement in the ducted SSPs demonstrates the effectiveness of the NB algorithm exploiting the observed multipath conditions.
 
@@
 Comparatively, the NB approach for HYCOM and the baseline SSP produce pseudorange errors of 8.30 and 2.39 m, respectively. 
-There is strong evidence to suggest that the sound speed and multipath fidelity codependently improve localization accuracy.
+There is strong evidence to suggest that the sound speed and multipath fidelity co-dependently improve localization accuracy.
 
@@
 
-\subsection{Re-positioning beacon to beacon events}
+\subsection{Re-positioning beacon-to-beacon events}
 
-When the beacons ran as virtual vehicles, the ICNN did not have access to that buoy's GPS data stream except for what was sent via digital acoustic message.
+When the beacons ran as virtual vehicles, the ICNN did not have access to that buoy's GNSS data stream except for what was sent via digital acoustic message.
 The static nature of the experiment means that the initial estimate transmitted to the ICNN was in fact a ground truth position.
 Therefore, a distribution of corrections from the ICNN, as shown in Fig. \ref{fig:trilat-beacon}, reflects positioning accuracy.
-The NB clearly outperforms the MB, with almost 80\% of the corrections below 6 meters and the median within the deployed GNSS puck precision of 3 meters.
-By contrast, the MB shows roughly 20\% within the GNSS puck precision, and separate peaks from 9--12 meters and 21--27 meters.
-This trimodal nature reflects the distribution of reflections on the ice surface.
+The NB, with the median below 3 m rms and mean below 6 m rms, clearly outperforms the MB, with the median around 10 m rms and the mean above 15 m rms. 
+The 75th percentiles are approximately 5 m rms and 25 m rms, for the NB and MB, respectively.
+The MB shows only about 20\% within the GNSS puck precision; the separate peaks from 9--12 meters and 21--27 meters reflect the distribution of number of surface reflections.
 
@@
 \includegraphics[width=\textwidth]{Fig11.pdf}
-\caption{Trilateration events; however, 264 entries are 2-beacon solutions, 22 are 3-beacon, and 2 are 4-beacon.
-When limited to only two ranging circles, the trilateration chooses the closer of the two intersection points.
-The x-axis is cut off at the upper limit for the NB algorithm.}
+\caption{A histogram (upper) and box plot (lower) of rms correction for trilateration events; 264 entries are two-beacon solutions, 22 are three-beacon, and 2 are four-beacon.
+For the beacon-to-beacon transmissions, the correction reflects the rms distance between the trilateration solution and the initial position reported by the transmitting buoy.
+When limited to only two beacons, the trilateration chooses the intersection point closer to the initial position estimate.
+The precision of the GNSS units used is 3 m.
+The mean and standard deviation re-positioning corrections for the NB and MB are 5.25 $\pm$ 7.60 m rms and 15.47 $\pm$ 10.22 m rms, respectively.}
 \label{fig:trilat-beacon}
@@
 
-Up to this point, pseudorange estimation and localization have been evaluated on GPS-linked beacon-to-beacon connections to validate the NB algorithm.
+Up to this point, pseudorange estimation and localization have been evaluated on GNSS-linked beacon-to-beacon connections to validate the NB algorithm.
 This analysis ports the MB and NB algorithms to re-navigate the AUV \emph{Macrura}.
@@
 Allowing receptions with PSK errors would almost double the number of recorded multipath arrivals exploited for positioning, if a real-time solution could correctly parse paths from different arrivals in the same thirty-second cycle.
-Thus it remains a future endeavor to explore how failure mode information from acoustic modems could be used to identify unsuccessful but otherwise trustworthy arrivals to augment trilateration samples.
+It remains a future endeavor to explore how failure mode information from acoustic modems could be used to identify unsuccessful but otherwise trustworthy arrivals to augment trilateration samples.
 
@@
 Accordingly, the vehicle dove deeper than 50 meters about 20\% of the time it was underway.
-
+The upper panel of Fig. \ref{fig:trilat-auv} shows the correction magnitudes for events with three or more receptions during AUV operations.
+Whereas the MB has a fairly bimodal nature, with peaks centered around 10--15 and 35--40 m, the NB favors smaller corrections, from 5--20 m, and has a long tail.
 In contrast to the modem tests, where position correction illustrated re-positioning accuracy, the re-navigation corrections are less valuable in the absence of GNSS ground truth.
@@
 Thus, larger corrections are not necessarily indicative of worse performance.
-Navigation accuracy is better described by trilateration error, the RMS of the remaining pseudorange errors from each acoustic link.
 
-Fig. \ref{fig:trilat-auv} shows the correction magnitudes and trilateration errors for events with three or more receptions during AUV operations.
-Whereas the MB has a fairly bimodal nature, with peaks centered around 10--15 and 35--40 m, the NB favors smaller corrections, from 5--20 m, and has a long tail.
-The distribution of corrections are much larger than the distribution of RMS error.
-It is apparent that, while both methods are quite successful, there is strong evidence that the NB achieves single meter accuracy.
+Navigation accuracy is better described by trilateration error, the rms of the remaining pseudorange errors from each acoustic link, shown in the lower panel of Fig. \ref{fig:trilat-auv}.
+These errors represent the uncertainty inherent from the overlap of acoustic ranging estimations and are almost an order of magnitude smaller than the distribution of corrections.
+The AUV re-navigation statistics are obtained from three times as many localization events compared to the beacon-to-beacon dataset.
+The AUV set also exhibits more events with at least three beacons and more diverse range-depth pairings.
+There is strong evidence that both methods achieve single meter accuracy and that the MB method, with more than 70\% of rms error under 3 m, classifies multipath structure effectively enough to extend GNSS accuracy into the water column.
 
@@
 \includegraphics[width=\textwidth]{Fig12.pdf}
-\caption{Distribution of vehicle re-navigation corrections (top) and error (bottom) computed in post-processing for the Minimal and Nearest Bounce Criterion.}
+\caption{A distribution of AUV re-navigation rms corrections (top) and rms error (bottom) computed in post-processing for the Minimal and Nearest Bounce Criterion.
+The re-navigation corrections represent the difference between the initial reported position and the trilateration solution; the mean and standard deviation for the NB and MB are 20.89 $\pm$ 14.11 m rms and 28.83 $\pm$ 16.79 m rms, respectively.
+The re-navigation error represents the remaining uncertainty of the trilateration solution; the mean and standard deviation for the NB and MB are 1.84 $\pm$ 2.19 and 4.53 $\pm$ 4.26 m rms, respectively.
+The legend of the bottom plot applies to both.}
 \label{fig:trilat-auv}
@@
 The effect is minor for surface platform navigation \textemdash roughly 15 m of horizontal precision has been displayed at the North Pole\textemdash but is significant enough to register against the modem's detected travel times.
-Fig. \ref{fig:gps-drift-example} zooms in on the GNSS and OWTT noise relative to the ice movement for two pairs of modem buoy connections.
-The two panels indicate the GPS noise as $\delta R = \sqrt{\delta x^2 + \delta y^2}$ and temporal drift, $\delta t$, relative to the median OWTT recorded between the two modems.
-The dashed line is scaled by a group velocity of 1440 m/s, such that if there were ideal sensor measurements with no drift, all events should exist on or near the line.
+Fig. \ref{fig:gps-drift-example} zooms in on the GNSS and OWTT noise relative to the ice movement for two representative pairs of modem buoy connections.
+The two panels indicate the GNSS noise as $\delta R = \sqrt{\delta x^2 + \delta y^2}$ and temporal spread, $\delta t$, relative to the median OWTT recorded between the two modems.
+The dashed line is scaled by an effective sound speed of 1440 m/s, such that if there were ideal sensor measurements with no drift, all events should exist on or near the line.
 
-\begin{figure}[h!]
+\begin{figure}[!ht]
     \centering
     \includegraphics[width=\columnwidth]{Fig13.pdf} 
-    \caption{A comparison of GPS noise (y-axis) versus OWTT drift (x-axis) for corners of the ICNN network with different source depths.}
+    \caption{A comparison of GNSS noise (y-axis) versus OWTT spread (x-axis) between corners of the ICNN network with different source depths.
+The dashed line normalizes an effective sound speed of 1440 m/s such that the vertical clustering indicates GNSS noise.}
     \label{fig:gps-drift-example}
@@
 Accurate range estimation is essential to mitigating error.
-Current approaches for underwater acoustic navigation simplify the non-linear relationship between a SSP and timefronts with a determinstic sound speed.
+Current approaches for underwater acoustic navigation simplify the non-linear relationship between a SSP and time fronts with a determinstic sound speed.
 Thus, the conversion of travel time into distance can be pre-conditioned for error and error growth over the course of a vehicle mission.
+
 This work introduces a lightweight stochastic prediction of an effective speed along the acoustic path between source and receiver, retooling arrival methods generally deemed too complex or labor intensive for real-time.
 We assume that the effective sound speed would be a locally smoothly varying function with respect to operational conditions\textemdash horizontal and vertical differences and rate of difference between source and receiver.
-The field-tested approach, the minimal bounce criteria, facilitated a successful AUV recovery in a total ice-covered, double ducted environment.
-The accuracy of the MB was validated against GPS-linked beacon-to-beacon communications.
-Given a consistent bias towards overestimation, an improved algorithm, the nearest bounce criteria, was developed on the insight that multipath structure may play an outsized role in maintaining a smoothly varying effective sound speed.
+The field-tested approach, the minimal bounce criteria, facilitated a successful AUV recovery in a total ice-covered, double-ducted environment.
+The accuracy of the MB was validated against GNSS-linked beacon-to-beacon communications.
+Given a consistent bias towards overestimation, an improved algorithm, the nearest bounce criteria, was developed on the insight that multipath structure may play an important role in maintaining a smoothly varying effective sound speed.
 The NB was developed with field data and reevaluated on vehicle data, achieving a position accuracy and precision that compares with that of the deployed GNSS puck.
 
-A key insight for both approaches was seeking an eigenray ensemble around an estimated location instead of seeking to unambigously match arrivals.
+A key insight for both approaches was seeking an eigenray ensemble around an estimated location instead of seeking to unambiguously match arrivals.
 The ensemble diversified the simulated multipath possibilities to better capture the actual multipath recorded.
 In this way, the solution exploits multipath, generally viewed as a source of uncertainty, as a new dimension of information to improve localization accuracy.
-Based on the navigation and re-navigation results of our AUV deployment in the ice-covered Beaufort Sea, we conclude that embedding a model-aided prediction of the effective sound speed has an outsized benefit to minimizing trilateration error, and that our approach sufficiently resolves the acoustic timefronts for an unpredictable, complex propagation environment like the double ducted Beaufort Lens.
+Based on the navigation and re-navigation results of our ice-covered AUV deployment, we conclude that embedding a model-aided prediction of the effective sound speed has an outsized benefit to minimizing trilateration error, and that our approach sufficiently resolves the OWTT for an unpredictable and complex propagation environment like the double-ducted Beaufort Lens.
 
@@
 
-The relatively simple nature of this approach suggests it is transferable to other environments, spatio-temporal scales, and platforms.
-While it is likely a particular quirk of the Beaufort Lens that filtering for reflection alone can produce a horizontal effective speed that compensates for ray refraction and reflection, the algorithm can be reconditioned to filter against other metrics, like number of turning points, to create a more diverse and informed set of multipath timefronts.
+The relatively simple nature of this approach invites a discussion about how transferable it is to other environments, spatio-temporal scales, and platforms.
+While it is a particular quirk of the Beaufort Lens that filtering for reflections alone can produce a horizontal effective speed that compensates for ray refraction and reflection, the algorithm can be reconditioned to filter against other metrics, like number of turning points, to classify a more diverse and informed set of multipath features.
 Though the majority of re-navigation results are within single-meter accuracy, future work can examine how constellations of more LBL beacons can extend the operational domain without adding an undesirable amount of error.
-One possibility is that, during a mission, ICNN-like LBL implementions use a comparison of the GNSS self-position and acoustic positioning to invert for the ocean volume, linking how vertical and horizontal sound speed structure impact transmission integrity.
+One possibility is that, during a mission, ICNN-like LBL implementations use a comparison of the GNSS self-position and acoustic positioning to invert for the ocean volume, linking how vertical and horizontal sound speed structure impact transmission integrity.
 \llabel{1.20} A fast tomographic estimate \citep{deffenbaugh_optimal_1997,Elisseeff2002}, along with its uncertainty, could be continuously communicated to assets underway to maintain contact or enable adaptive sampling.
@@
 Spatio-temporal variability is a serious challenge for accurate real-time ranging.
-On one hand, the effectiveness of eigenray filtering algorithm is likely only challenged by the valid operational scales of a range independent propagation environment.
-Longer range experiments may provide more time for eigenray filtering.
-A bootstrapping approach that filters eigenrays for several randomly generated internal wave spectrums may compensate for otherwise unknowable spatio-temporal variability.
-The model-aided component to the eigenray filtering is compatible with vertical slices from any physically driven ocean model.
-But in the long run, more accurate and higher resolution global circulation models are needed to properly resolve features that alter ducted propagation at the scales discernible to an acoustic modem.
+For example, a recent modeling study showed that eddies and filaments create lateral variability in the Beaufort Lens \citep{duda_effects_2021}.
+The effectiveness of the eigenray filtering algorithm is likely challenged by the valid operational scale of a range-independent propagation environment.
+One approach to addressing this limitation is to bootstrap filtered eigenrays for several perturbations of the fitted sound speed profile, which may compensate for otherwise unknowable spatio-temporal variability.
+Another may be to include range-dependent ocean snapshots to the eigenray filtering process; this would be necessary for longer range experiments, which present more spatio-temporal variability but also additional time for onboard computation.
+More accurate and higher resolution global circulation models are needed to properly resolve features that alter ducted propagation at the scales discernible to an acoustic modem.
 Through-the-sensor methods can resolve local features but would require a degree of information sharing not readily supported on the acoustic channel for large scale variability.
-But addressing the spatial and temporal scales of what can be solved deterministically and what must be solved stochastically imposes a resolution constraint that is at odds with computational overhead for real-time operations.
+Addressing the spatial and temporal scales of what can be solved deterministically and what must be solved stochastically imposes a resolution constraint that is at odds with computational overhead for real-time operations.
 Resolving features inaccurately, or with a false sense of confidence, could be more harmful than contextualizing the limitations of a range independent propagation over realistic bathymetry.
@@
 The methods involved in this paper include open source software projects \citep{,Benjamin2010,schneider2015dynamic,schneider_unified_2010} that are platform agnostic.
-Large AUVs, often large enough to support long duration and/or deep sea missions, would benefit from including diurnal or tidal effects for ranging.
+Large AUVs, often powerful enough to support long duration and/or deep sea missions, would benefit from including diurnal or tidal effects for ranging; eigenray filtering would be simpler given sound speed homogeneity at depth.
 Gliders, though generally low power and memory, have been equipped with acoustic modems.
-Their inability to maintain position within a region of reliable acoustic path makes the impact of an environmentally adaptive pseudorange estimation disproportionately positive.
-The exact adjustments to the ensemble eigenray filtering are predicated on the expected sound speed conditions and acoustic arrival structure; the problem is ripe for other simulation testbeds or machine learning methods.
-The continued development of embedded acoustic processing on heterogenous platforms is fundamental to support a universal underwater navigation scheme comparable to GNSS.
+Their inability to maintain position within a region of reliable acoustic path makes them ripe for environmentally and acoustically adaptive range estimation, where the resource-heavy computation could occur on the LBL network.
+Ship-based computers can provide the same functionality for short and ultra-short baseline paradigms in large and/or complex acoustic environments. 
+The exact adjustments to the ensemble eigenray filtering are predicated on the expected sound speed conditions and acoustic arrival structure; the problem is well-suited for other simulation testbeds and machine learning classification.
+The continued development of embedded acoustic processing on and across heterogeneous platforms is fundamental to support a universal underwater navigation scheme comparable to GNSS.
 
@@
 \begin{acknowledgments}
-We acknowledge the significant operational effort spearheaded by the LAMSS ICEX20 team and all our collaborators.
-Bhatt was funded by a National Defense, Science, and Engineering Graduate Fellowship.
+We thank the editors and reviewers for their thoughtful and detailed feedback in improving this work.
+We also thank all those in the LAMSS ICEX20 team and our collaborators from General Dynamics and the WHOI AComms Group.
+The Ice-Tethered Profiler data were collected and made available by the Ice-Tethered Profiler Program (Toole et al., 2011; Krishfield et al., 2008) based at the Woods Hole Oceanographic Institution (\href{https://www2.whoi.edu/site/itp/}{www2.whoi.edu/site/itp}).
+
+Bhatt was a student at the MIT-WHOI Joint Program in Oceanography/Applied Ocean Science \& Engineering, funded by a National Defense, Science, and Engineering Graduate Fellowship.
 This work was supported by the Office of Naval Research 322-OA under ICEX20 (N00014-17-1-2474) and Task Force Ocean (N00014-19-1-2716).

### 2022-03-31 pipeline/9-trilateration/auv_trilat_hist.m
easy output of stats, same buffer style on plots

@@ -25,7 +25,7 @@ nexttile(1);
 edges = [0:5:65];
 Y{2}.count = trilat.mbc.correction;
 Y{1}.count = trilat.nbc.correction;
-h_hist_boxplot(Y,edges,colors,max(edges));
+statBackCorr = h_hist_boxplot(Y,edges,colors,max(edges));
 
 xlabel('correction [m rms]')
 set(gca,'fontsize',13)
@@ -40,7 +40,7 @@ edges = [0:3:21];
 Y{1}.count = trilat.nbc.error;
 Y{2}.count = trilat.mbc.error;
 
-h_hist_boxplot(Y,edges,colors,max(edges));
+statBackErr = h_hist_boxplot(Y,edges,colors,max(edges));
 xlabel('error [m rms]')
 set(gca,'fontsize',13)
 title({'',...

### 2022-03-31 pipeline/9-trilateration/h_hist_boxplot.m

@@ -1,4 +1,4 @@
-function [] = h_hist_boxplot(Y,EDGES,COLORS,THRESHOLD)
+function [statBack] = h_hist_boxplot(Y,EDGES,COLORS,THRESHOLD)
 %h_hist_boxplot makes a beautiful histogram w/ a boxplot underlaid
 
 diffEdges = mean(diff(EDGES))./2;
@@ -17,7 +17,7 @@ end
 grid on
 ylabel('probability');
 xticks(EDGES);
-xlim([-1 max(EDGES)]);
+xlim([-1 max(EDGES)+1]);
 
 % box plot
 hold on
@@ -26,12 +26,12 @@ buff = -0.15;
 
 % add bottom "box" plot
 hold on
-buff = -0.15;
 maxY = ceil(max(h(:))*10)/10;
+buff = -maxY/5;
 ylim([buff maxY]);
 yticks([0:0.1:maxY]);
 
-kbuff = .05;
+kbuff = buff/3;
 for s = 1:2
     
     yval = Y{s}.count;
@@ -42,13 +42,18 @@ for s = 1:2
     xQuant1 = quantile(yval,[0 1]);
     xQuant2 = quantile(yval,[.25 .5 .75]);
     
-    plot(xQuant1,ones(2,1).*-s*kbuff,'handlevisibility','off','color',[COLORS{s} 0.4]);
-    plot(xQuant1,-s*kbuff,'.','handlevisibility','off','color',COLORS{s},'MarkerSize',15);
-    plot(meanVal,-s*kbuff,'o','handlevisibility','off','color',COLORS{s},'MarkerSize',12);
+    plot(xQuant1,ones(2,1).*s*kbuff,'handlevisibility','off','color',[COLORS{s} 0.4]);
+    plot(xQuant1,s*kbuff,'.','handlevisibility','off','color',COLORS{s},'MarkerSize',15);
+    plot(meanVal,s*kbuff,'o','handlevisibility','off','color',COLORS{s},'MarkerSize',12);
     
-    plot(xQuant2(1),-s*kbuff,'<','handlevisibility','off','color',COLORS{s});
-    plot(xQuant2(3),-s*kbuff,'>','handlevisibility','off','color',COLORS{s});
-    plot(xQuant2(2),-s*kbuff,'d','handlevisibility','off','color',COLORS{s});
+    plot(xQuant2(1),s*kbuff,'<','handlevisibility','off','color',COLORS{s});
[…]

### 2022-03-31 pipeline/9-trilateration/modem_trilat_hist.m

@@ -17,7 +17,7 @@ colors = {[153 51 153]./256,[51 153 153]./256};
 %% figure
 figure('name','trilat-histogram','renderer','painters','position',[108 108 1200 500]);
 
-h_hist_boxplot(Y,edges,colors,48);
+statBack = h_hist_boxplot(Y,edges,colors,48);
 
 %% dress up figure
 xlabel('correction [m rms]')

### 2022-03-31 manuscript/manuscript.tex
abstract < 200 words

@@
 For real-time autonomous underwater vehicle (AUV) operations, this conversion assumes an isovelocity sound speed.
-For re-navigation, computationally and/or labor intensive acoustic modeling may be employed to reduce uncertainty.
-This work demonstrates a real-time ray-based prediction method of the effective sound speed along a path from source to receiver to minimize pseudorange error.
-\llabel{2.1} This method was implemented for an AUV-LBL system in the Beaufort Sea, in total ice-covered conditions and a double-ducted propagation environment.
+For re-navigation, computationally and/or labor-intensive acoustic modeling may be employed to reduce uncertainty.
+This work demonstrates a real-time ray-based prediction of the effective sound speed along a path from source to receiver.
+\llabel{2.1} This method was implemented for an AUV-LBL system in the Beaufort Sea, in an ice-covered and a double-ducted propagation environment.
 \llabel{2.2} Given the lack of Global Navigation Satellite System (GNSS) data throughout the vehicle's mission, the pseudorange performance is first evaluated on acoustic transmissions between GNSS-linked beacons.
